### 2020-05-05 main.tex
Initial Overleaf Import

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[maxcitenames=1]{biblatex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{tcolorbox}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{thmtools}
\usepackage{thm-restate}
\usepackage{enumerate}

\title{SynBio ML}
\author{u6015325 }
\date{January 2020}

\bibliography{ref.bib}

\begin{document}

\maketitle

\section{Machine Learning Algorithm Description}

We optimise the translation initiation rate (TIR) by designing sequential experiments. The goal is to identify the set of RBS sequences with top TIR scores with as fewer rounds as possible.

\subsection{Settings}

For the first round experiment, we design 180 RBS sequences based on the consensus sequence:

\begin{enumerate}
    \item 60 RBS sequences designed by ``1 by 1 changing" based on the consensus sequence.
    \item 60 RBS sequences by random design, including uniformly random (equal probability of choosing each letter for each base); random based on the position probability matrix (PPM).
    \item 60 RBS sequences by sequentially machine learning design. We use an bandit optimazation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) \cite{srinivas2012information}.   
\end{enumerate}{}

\subsection{Data Pre-processing}

We design the first round bandit recommendation based on the data from \textcite{jervis2018machine}. The data contains 113 non-repeated records for 56 unique RBS sequences with the TIR label. The label is between 0 - 100,000 and skewed, which is shown in Figure xx. We normalise the label to 0 - 1 using the min-max normalisation.

\begin{figure}[t]
    \centering
    \includegraphics[scale=0.7]{plots/TIR_histogram.pdf}
    \caption{TIR Histogram.}
    \label{fig: TIR Histogram.}
\end{figure}

The RBS sequence is 20 bps, we focus on -8 to -13 bps and fix others as the same as the consensus sequence, i.e. TTTAAGA + NNNNNN + TATACAT. For each base, there are 4 possibilities: A, C, G, T. So totally the feature space is $4^6$ = 4096.

\subsection{Algorithms}

\subsubsection{Random sampling}

\subsubsection{Gaussian Process Upper Confidence Bound}

We use Gaussian Process Upper Confidence Bound (GPUCB)] algrithm \cite{srinivas2012information}, with sum of dot product kernel and spectrum kernel of sequences. The algorithm basically includes two parts:

\begin{enumerate}
    \item the Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar. So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity. 
    \item bandit algorithms (Upper Confidence Bound), which recommends sequences to test for next round.
\end{enumerate}

\section{Results}

\begin{enumerate}
    \item RMSE of predictions.

    \item Similarity of recommendations. 
    \begin{figure}[t]
    \centering
    \includegraphics[scale=0.7]{plots/similarity_first_round_recommendation.pdf}
    \caption{TIR Histogram.}
    \label{fig: TIR Histogram.}
\end{figure}
\end{enumerate}{}

\printbibliography

\end{document}

### 2020-05-05 notebooks/rec_design/Clustering/HSN_util.py
clustering implement

@@ -1,0 +1,183 @@
+import seaborn as sns
+import matplotlib.pyplot as plt
+import numpy as np
+import re
+import torch
+
+
+def inverse_tanh_cpu(y):
+    return 0.5 * (np.log(1 + y) - np.log(1 - y))
+
+def inverse_tanh_cuda(y):
+    return 0.5 * (torch.log(1 + y) - torch.log(1-y))
+
+def batch_iterator(X, batch_size = 256):
+    l = len(X)
+    for idx in range(0, l, batch_size):
+        yield X[idx:idx + batch_size]
+
+
+def preprocess(text_string):
+    """
+    Accepts a text string and replaces:
+    1) urls with URLHERE
+    2) lots of whitespace with one instance
+    3) mentions with MENTIONHERE
+
+    This allows us to get standardized counts of urls and mentions
+    Without caring about specific people mentioned
+    """
+
+    space_pattern = '\s+'
+    giant_url_regex = ('http[s]?://(?:[a-zA-Z]|[0-9]|[$-_@.&+]|''[!*\(\),]|(?:%[0-9a-fA-F][0-9a-fA-F]))+')
+    mention_regex = '@[\w\-]+'
+    emoji_regex = '&#[0-9]+'
+    seperator_regex = ';+|:+'
+    quotation_regex = '"+|\-\-+'
+    exclamation_regex = '!+'
+    dots_regex = '\.\.+'
+    dot_regex = ' \.'
+    sdot_regex = '\.'
+    comma_regex = ','
+    RT_regex = 'RT'
+
+
+    parsed_text = re.sub(quotation_regex, ' ', text_string)
+    parsed_text = re.sub(RT_regex, ' ', text_string)
+    parsed_text = re.sub(dot_regex, '.', parsed_text)
+    parsed_text = re.sub(comma_regex, ' ,', parsed_text)
+    parsed_text = re.sub('!', ' !', parsed_text)
+    # parsed_text = re.sub('\?', ' \?', parsed_text)
+    # parsed_text = re.sub('\\', ' ', parsed_text)
+    parsed_text = re.sub(dots_regex, '', parsed_text)
+    parsed_text = re.sub(exclamation_regex, '', parsed_text)
+    parsed_text = re.sub(giant_url_regex, '', parsed_text)
+    parsed_text = re.sub(mention_regex, '', parsed_text)
+    parsed_text = re.sub(emoji_regex,'', parsed_text)
[…]

### 2020-05-05 notebooks/rec_design/Clustering/load2d3dplotly.py

@@ -1,0 +1,124 @@
+import plotly
+import plotly.graph_objs as go
+import numpy as np
+import pandas as pd
+import argparse
+from HSN_util import *
+
+parser = argparse.ArgumentParser(description='Show 3d scatter plot')
+parser.add_argument('npz_path', help = "the path to the npz file which contains both label and coordinates")
+
+args = parser.parse_args()
+
+data = np.load(args.npz_path)
+
+embed = data['coord']
+# y = data['label']
+text = data['text']
+y_km = data['ykm']
+print(y_km)
+# docs = data['docs']
+
+
+dim = embed.shape[1]
+
+print(args.npz_path[:-4])
+
+# l = np.array(['Harm', 'Harmless', 'Impurity','InternationalConspiracy', 'LiteralHigh','LiteralLow','MetaphoricalHigh', 'MetaphoricalLow', 'NonHuman', 'Threat', 'Threat1', 'Threat2'])
+
+assert (dim == 2 or dim == 3), "invalid dim"
+
+if dim == 2:
+    trace_list = []
+    for i in range(len(set(y_km))):
+        #trace_list.append(go.Scatter(x = embed[y_km==i,0], y = embed[y_km==i,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), opacity = 0.6), text = text, name = str(i), hoverinfo='text'))
+        trace_list.append(go.Scatter(x = embed[y_km==i,0], y = embed[y_km==i,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), colorscale = plotly.colors.qualitative.Plotly, opacity = 0.8), text = text, name = str(i), hoverinfo='text'))
+    # trace_pred = go.Scatter(x = embed[:,0], y = embed[:,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color = y_pred, colorscale = 'Viridis', opacity = 0.6), text = text, name = "Predict Label") 
+else:
[…]

### 2020-05-05 codes/embedding.py
tidy up.

@@ -1,6 +1,6 @@
 import numpy as np
 from sklearn.feature_extraction.text import CountVectorizer
-from sklearn.preprocessing import LabelEncoder
+from sklearn.preprocessing import LabelEncoder, normalize
 
 # 2nd Nov 2019
 # Mengyan Zhang
@@ -59,7 +59,10 @@ class Embedding():
             # loop through each individual sequence, from the 5' to 3' end
             for b in range(self.num_bases):
                 embedded_data[i, b * 4 + base_dict[seq[b]]] = 1
-        return embedded_data
+
+        embedded_data -= np.nanmean(embedded_data, axis = 0)
+        normalised_embedded_data = normalize(embedded_data, norm = 'l2')
+        return normalised_embedded_data
 
     def kmer(self, size = 3):
         """k-merization embedding.

### 2020-05-05 notebooks/rec_design/HSN_util.py

@@ -1,183 +1,0 @@
-import seaborn as sns
-import matplotlib.pyplot as plt
-import numpy as np
-import re
-import torch
-
-
-def inverse_tanh_cpu(y):
-    return 0.5 * (np.log(1 + y) - np.log(1 - y))
-
-def inverse_tanh_cuda(y):
-    return 0.5 * (torch.log(1 + y) - torch.log(1-y))
-
-def batch_iterator(X, batch_size = 256):
-    l = len(X)
-    for idx in range(0, l, batch_size):
-        yield X[idx:idx + batch_size]
-
-
-def preprocess(text_string):
-    """
-    Accepts a text string and replaces:
-    1) urls with URLHERE
-    2) lots of whitespace with one instance
-    3) mentions with MENTIONHERE
-
-    This allows us to get standardized counts of urls and mentions
-    Without caring about specific people mentioned
-    """
-
-    space_pattern = '\s+'
-    giant_url_regex = ('http[s]?://(?:[a-zA-Z]|[0-9]|[$-_@.&+]|''[!*\(\),]|(?:%[0-9a-fA-F][0-9a-fA-F]))+')
-    mention_regex = '@[\w\-]+'
-    emoji_regex = '&#[0-9]+'
-    seperator_regex = ';+|:+'
-    quotation_regex = '"+|\-\-+'
-    exclamation_regex = '!+'
-    dots_regex = '\.\.+'
-    dot_regex = ' \.'
-    sdot_regex = '\.'
-    comma_regex = ','
-    RT_regex = 'RT'
-
-
-    parsed_text = re.sub(quotation_regex, ' ', text_string)
-    parsed_text = re.sub(RT_regex, ' ', text_string)
-    parsed_text = re.sub(dot_regex, '.', parsed_text)
-    parsed_text = re.sub(comma_regex, ' ,', parsed_text)
-    parsed_text = re.sub('!', ' !', parsed_text)
-    # parsed_text = re.sub('\?', ' \?', parsed_text)
-    # parsed_text = re.sub('\\', ' ', parsed_text)
-    parsed_text = re.sub(dots_regex, '', parsed_text)
-    parsed_text = re.sub(exclamation_regex, '', parsed_text)
-    parsed_text = re.sub(giant_url_regex, '', parsed_text)
-    parsed_text = re.sub(mention_regex, '', parsed_text)
-    parsed_text = re.sub(emoji_regex,'', parsed_text)
[…]

### 2020-05-05 notebooks/rec_design/load2d3dplotly.py

@@ -1,124 +1,0 @@
-import plotly
-import plotly.graph_objs as go
-import numpy as np
-import pandas as pd
-import argparse
-from HSN_util import *
-
-parser = argparse.ArgumentParser(description='Show 3d scatter plot')
-parser.add_argument('npz_path', help = "the path to the npz file which contains both label and coordinates")
-
-args = parser.parse_args()
-
-data = np.load(args.npz_path)
-
-embed = data['coord']
-# y = data['label']
-text = data['text']
-y_km = data['ykm']
-print(y_km)
-# docs = data['docs']
-
-
-dim = embed.shape[1]
-
-print(args.npz_path[:-4])
-
-# l = np.array(['Harm', 'Harmless', 'Impurity','InternationalConspiracy', 'LiteralHigh','LiteralLow','MetaphoricalHigh', 'MetaphoricalLow', 'NonHuman', 'Threat', 'Threat1', 'Threat2'])
-
-assert (dim == 2 or dim == 3), "invalid dim"
-
-if dim == 2:
-    trace_list = []
-    for i in range(len(set(y_km))):
-        #trace_list.append(go.Scatter(x = embed[y_km==i,0], y = embed[y_km==i,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), opacity = 0.6), text = text, name = str(i), hoverinfo='text'))
-        trace_list.append(go.Scatter(x = embed[y_km==i,0], y = embed[y_km==i,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), colorscale = plotly.colors.qualitative.Plotly, opacity = 0.8), text = text, name = str(i), hoverinfo='text'))
-    # trace_pred = go.Scatter(x = embed[:,0], y = embed[:,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color = y_pred, colorscale = 'Viridis', opacity = 0.6), text = text, name = "Predict Label") 
-else:
[…]

### 2020-05-07 main.tex
Updated bibliography
Text updated 07052020

@@
 \usepackage[utf8]{inputenc}
-\usepackage[maxcitenames=1]{biblatex}
+\usepackage[maxcitenames=1,style=numeric]{biblatex}
 \usepackage{amsmath}
@@
 
-\title{SynBio ML}
+\title{Machine Learning based TIR predictor}
 \author{u6015325 }
-\date{January 2020}
+\date{\today{}}
 
@@
 \maketitle
+
+\section{Introduction}
+
+One of the main tenets of synthetic biology is design, evaluation and standarization of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
+This can be done in variety of ways, most of each involve designing the DNA sequence in CAD software and then physically testing it in a laboratory.
+Alternative to this is computer modelling and prediction of part behoviour based on the designed DNA sequence or design of DNA sequence based on expected function \cite{Yeoh2019,Nielsen2016}.
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen1994,Vellanoweth1992,Chen2013,Reeve2014}.\\
+According to Reeve \emph{et al.} there are three main RBS calculators, all prediciting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+Predictions from all of these models are relatively good ($R^2 >0.8$), there come with a number of caveats: i) they rely on calculations of free energies that can be hard to calculate ii) in general, the models' accuracy is improved by increasing the number of phenomenons taking place during the translation, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
+\\
+Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments.
+\cite{Freemont2019} New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughputs in data generation.
+These new datasets can be combined with data-hungry machine learning algorithms to generate new models and predictors for use in synthetic biology.
+\\
+In this article we present the first in our knowledge TIR predictor built using machine learning approach.
+
+\section{Ribosome Binding Site design}
 
@@
 
+\section{Methods}
+
+\newpage
+
 \printbibliography

### 2020-05-07 notebooks/rec_design/Clustering/load2d3dplotly.py
update plots with different shapes.

@@ -16,6 +16,9 @@ embed = data['coord']
 # y = data['label']
 text = data['text']
 y_km = data['ykm']
+known_seq = data['known_seq']
+ucb_rec = data['ucb_rec']
+
 print(y_km)
 # docs = data['docs']
 
@@ -28,11 +31,40 @@ print(args.npz_path[:-4])
 
 assert (dim == 2 or dim == 3), "invalid dim"
 
+idxes = np.asarray(range(len(y_km)))
+
 if dim == 2:
     trace_list = []
     for i in range(len(set(y_km))):
         #trace_list.append(go.Scatter(x = embed[y_km==i,0], y = embed[y_km==i,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), opacity = 0.6), text = text, name = str(i), hoverinfo='text'))
-        trace_list.append(go.Scatter(x = embed[y_km==i,0], y = embed[y_km==i,1], mode = 'markers', marker = dict(size = 14, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), colorscale = plotly.colors.qualitative.Plotly, opacity = 0.8), text = text, name = str(i), hoverinfo='text'))
+        known_seq_k = []
+        ucb_rec_k = []
+        others_k = []
+        
+        for k in idxes[y_km==i]:
+            if k in known_seq:
+                known_seq_k.append(k)
+            elif k in ucb_rec:
+                ucb_rec_k.append(k)
+            else:
+                others_k.append(k)
+
+        known_seq_k = np.asarray(known_seq_k)
+        ucb_rec_k = np.asarray(ucb_rec_k)
+        others_k = np.asarray(others_k)
+
+        if others_k.shape[0] > 0:
+            trace_list.append(go.Scatter(x = embed[others_k,0], y = embed[others_k,1], mode = 'markers', marker = dict(size = 8, symbol=0, line = dict(width = 0), color =  np.concatenate((y_km[y_km == i], np.array([j for j in range(len(set(y_km)))]))), 
+                                colorscale = plotly.colors.qualitative.Plotly, opacity = 0.5), text = text[others_k], name = str(i), hoverinfo='text'))
+
+        if known_seq_k.shape[0] > 0:
[…]

### 2020-05-08 main.tex
Writing updated until 8/5/2020

@@
 Alternative to this is computer modelling and prediction of part behoviour based on the designed DNA sequence or design of DNA sequence based on expected function \cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen1994,Vellanoweth1992,Chen2013,Reeve2014}.\\
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.\\
 According to Reeve \emph{et al.} there are three main RBS calculators, all prediciting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
@@
 \cite{Freemont2019} New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughputs in data generation.
-These new datasets can be combined with data-hungry machine learning algorithms to generate new models and predictors for use in synthetic biology.
+These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology \cite{Camacho2018}.
+In the past few years there was a significant uptick of Machine Learning based approaches to synthetic biology.
+Jervis \emph{et al.} used support vector machine and neural network to optimize production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
+Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyze time-series multiomics data to predict metabolic pathway bahaviour \cite{Costello2018}.
+There were also succesful attempts at using deap learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.
+However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential.
 \\
-In this article we present the first in our knowledge TIR predictor built using machine learning approach.
+Here we present a machine learning workflow for building a RBS trasnaltion rate predictor based on fluoresence data from cultures.
+RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
+We have used Gaussian Process - Upper Confidence Bound and Bandits algorithms to analyze and optimize the initiation rates of the designed RBS.
 
 \section{Ribosome Binding Site design}
+There is a number of qualities that impact the translation rate, but chief among them are concerned with how the ribosome recognizes and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
+In \emph{E.
+coli} the RBS is usually located in the 20 bases downstream of the start codon.
+The RBS usually has a distingushable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E.
+coli} is AGGAGG.
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).
+\\
+
+%\begin{figure}[t]
+%    \centering
+%    \includegraphics[scale=1.0]{.pdf}
+%    \caption{Anatomy of the randomized sequence.}
+%    \label{fig: Anatomy of the randomized sequence.}
+%\end{figure}
 

### 2020-05-11 methods.tex
Changed gitignore, added methods.tex

--- 2020-05-05 main.tex
@@
 
-\title{SynBio ML}
+\title{Machine Learning based TIR predictor - Methods}
 \author{u6015325 }
-\date{January 2020}
+\date{\today{}}
 
 \bibliography{ref.bib}
-
+\DeclareUnicodeCharacter{2212}{-}
 \begin{document}
@@
 
-\section{Machine Learning Algorithm Description}
-
-We optimise the translation initiation rate (TIR) by designing sequential experiments. The goal is to identify the set of RBS sequences with top TIR scores with as fewer rounds as possible.
-
-\subsection{Settings}
-
-For the first round experiment, we design 180 RBS sequences based on the consensus sequence:
-
-\begin{enumerate}
-    \item 60 RBS sequences designed by ``1 by 1 changing" based on the consensus sequence.
-    \item 60 RBS sequences by random design, including uniformly random (equal probability of choosing each letter for each base); random based on the position probability matrix (PPM).
-    \item 60 RBS sequences by sequentially machine learning design. We use an bandit optimazation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) \cite{srinivas2012information}.   
-\end{enumerate}{}
-
-\subsection{Data Pre-processing}
-
-We design the first round bandit recommendation based on the data from \textcite{jervis2018machine}.
-The data contains 113 non-repeated records for 56 unique RBS sequences with the TIR label.
-The label is between 0 - 100,000 and skewed, which is shown in Figure xx.
-We normalise the label to 0 - 1 using the min-max normalisation.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.7]{plots/TIR_histogram.pdf}
-    \caption{TIR Histogram.}
-    \label{fig: TIR Histogram.}
-\end{figure}
-
-The RBS sequence is 20 bps, we focus on -8 to -13 bps and fix others as the same as the consensus sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT.
-For each base, there are 4 possibilities: A, C, G, T.
-So totally the feature space is $4^6$ = 4096.
-
-\subsection{Algorithms}
-
-\subsubsection{Random sampling}
-
-\subsubsection{Gaussian Process Upper Confidence Bound}
-
-We use Gaussian Process Upper Confidence Bound (GPUCB)] algrithm \cite{srinivas2012information}, with sum of dot product kernel and spectrum kernel of sequences.
-The algorithm basically includes two parts:
-
-\begin{enumerate}
-    \item the Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
-    \item bandit algorithms (Upper Confidence Bound), which recommends sequences to test for next round.
-\end{enumerate}
-
-\section{Results}
-
-\begin{enumerate}
-    \item RMSE of predictions.
-
-    \item Similarity of recommendations. 
-    \begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.7]{plots/similarity_first_round_recommendation.pdf}
-    \caption{TIR Histogram.}
-    \label{fig: TIR Histogram.}
-\end{figure}
-\end{enumerate}{}
-
-\printbibliography
+\section{Method 1}
+\subsection{Method1a}
 

### 2020-05-13 main.tex
Updates from Overleaf
fix u+2212 error
input method.tex into main.tex
Updated 13/05/20

@@
 \bibliography{ref.bib}
-
+\DeclareUnicodeCharacter{2212}{-}
 \begin{document}
@@
 RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-We have used Gaussian Process - Upper Confidence Bound and Bandits algorithms to analyze and optimize the initiation rates of the designed RBS.
+We have used Gaussian Process-Upper Confidence Bound and Bandits algorithms to analyze and optimize the initiation rates of the designed RBS.
 
@@
 
-%\begin{figure}[t]
-%    \centering
-%    \includegraphics[scale=1.0]{.pdf}
-%    \caption{Anatomy of the randomized sequence.}
-%    \label{fig: Anatomy of the randomized sequence.}
-%\end{figure}
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
+    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomized core sequence in red.
+The start codon of the GFP CDS is also shown with numbers showing relative nucleotide positions.
+At the bottom a simplified diagram of the genetic device is shown.}
+    \label{fig: Anatomy of the randomized sequence.}
+\end{figure}
+
+\input{methods}
 
@@
 
-\section{Methods}
+\input{methods}
 
@@
 \printbibliography
-
 \end{document}

### 2020-05-13 main.tex
Updated 13/05/20

--- 2020-05-07 main.tex
@@
 \usepackage{enumerate}
+\usepackage{xcolor}
 
@@
 
-One of the main tenets of synthetic biology is design, evaluation and standarization of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
+One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
 This can be done in variety of ways, most of each involve designing the DNA sequence in CAD software and then physically testing it in a laboratory.
-Alternative to this is computer modelling and prediction of part behoviour based on the designed DNA sequence or design of DNA sequence based on expected function \cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen1994,Vellanoweth1992,Chen2013,Reeve2014}.\\
-According to Reeve \emph{et al.} there are three main RBS calculators, all prediciting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+Alternative to this is computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function \cite{Yeoh2019,Nielsen2016}.
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.\\
+According to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Predictions from all of these models are relatively good ($R^2 >0.8$), there come with a number of caveats: i) they rely on calculations of free energies that can be hard to calculate ii) in general, the models' accuracy is improved by increasing the number of phenomenons taking place during the translation, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
@@
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments.
-\cite{Freemont2019} New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughputs in data generation.
-These new datasets can be combined with data-hungry machine learning algorithms to generate new models and predictors for use in synthetic biology.
+\cite{Freemont2019} New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
+These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology \cite{Camacho2018}.
+In the past few years there was a significant uptick of Machine Learning based approaches to synthetic biology.
+Jervis \emph{et al.} used support vector machine and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
+Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway bahaviour \cite{Costello2018}.
+There were also successful attempts at using deep learning techniques for analysis of big data-sets \cite{Alipanahi2015,Angermueller2016}.
+However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential.
 \\
-In this article we present the first in our knowledge TIR predictor built using machine learning approach.
+Here we present a machine learning workflow for building a RBS translation rate predictor based on fluorescence data from cultures.
+RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
+We have used Gaussian Process-Upper Confidence Bound and Bandits algorithms to analyse and optimise the initiation rates of the designed RBS.
 
-\section{Ribosome Binding Site design}
+\section{Genetic device design}
+There is a number of qualities that impact the translation rate, but chief among them are concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
+In \emph{E.
+coli} the RBS is usually located in the 20 bases downstream of the start codon.
+The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E.
+coli} is AGGAGG.
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).\\
+The RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant.
+By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring the device.
+Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1.
+By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
 
-\section{Machine Learning Algorithm Description}
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
+    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
+The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.
+At the bottom a simplified SBOL diagram of the genetic device is shown.}
+    \label{fig: Anatomy of the randomized sequence.}
+\end{figure}
 
-We optimise the translation initiation rate (TIR) by designing sequential experiments. The goal is to identify the set of RBS sequences with top TIR scores with as fewer rounds as possible.
+\input{methods}
 
-\subsection{Settings}
+\section{Workflow Description}
 
-For the first round experiment, we design 180 RBS sequences based on the consensus sequence:
+Our experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with as fewer rounds as possible.
+We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds).
+The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
+After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
+The results are then fed back to the algorithm for it to recommend the next round of designs.
+
+\subsection{First Round of Designs}
+
+The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}.
+In our design we focus on randomizing of the core -8 to -13 (relative to the GFP) bps of the RBS and fix others to be the same as the consensus sequence, i.e.
+TTTAAGA + NNNNNN + TATACAT.
+Since for each of the 6 position there are 4 possibilities: A, C, G, T the total feature space is $4^6$ = 4096.\\
+For the first round of experiments, we design 180 (two times 90) RBS sequences based on the consensus sequence:
 
 \begin{enumerate}
-    \item 60 RBS sequences designed by ``1 by 1 changing" based on the consensus sequence.
-    \item 60 RBS sequences by random design, including uniformly random (equal probability of choosing each letter for each base); random based on the position probability matrix (PPM).
-    \item 60 RBS sequences by sequentially machine learning design. We use an bandit optimazation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) \cite{srinivas2012information}.   
-\end{enumerate}{}
-
-\subsection{Data Pre-processing}
-
+    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
+This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
+    \item 30 RBS sequences fully and uniformly randomised (equal probability of choosing each letter for each base) 
+    \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
+    \item 60 RBS sequences by sequentially machine learning design.
+We use an bandit optimazation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) \cite{srinivas2012information}.
 We design the first round bandit recommendation based on the data from \textcite{jervis2018machine}.
@@
 We normalise the label to 0 - 1 using the min-max normalisation.
+\end{enumerate}{}
 
@@
 \end{figure}
-
-The RBS sequence is 20 bps, we focus on -8 to -13 bps and fix others as the same as the consensus sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT.
-For each base, there are 4 possibilities: A, C, G, T.
-So totally the feature space is $4^6$ = 4096.
-
-\subsection{Algorithms}
-
-\subsubsection{Random sampling}
-
-\subsubsection{Gaussian Process Upper Confidence Bound}
-
-We use Gaussian Process Upper Confidence Bound (GPUCB)] algrithm \cite{srinivas2012information}, with sum of dot product kernel and spectrum kernel of sequences.
-The algorithm basically includes two parts:
-
-\begin{enumerate}
-    \item the Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
-    \item bandit algorithms (Upper Confidence Bound), which recommends sequences to test for next round.
-\end{enumerate}
 

### 2020-05-13 methods.tex
Minor changes to methods.tex
Updated 13/05/20

@@
-\documentclass{article}
-\usepackage[utf8]{inputenc}
-\usepackage[maxcitenames=1,style=numeric]{biblatex}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{tcolorbox}
-\usepackage{graphicx}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage{subfig}
-\usepackage{hyperref}
-\usepackage{cleveref}
-\usepackage{thmtools}
-\usepackage{thm-restate}
-\usepackage{enumerate}
+\section{Method 1}
+This will be about the DNA assembly.\\
+\subsection{Method 1a}
+This will be very specifically talking about the plasmid.
 
-\title{Machine Learning based TIR predictor - Methods}
-\author{u6015325 }
-\date{\today{}}
+\subsection{Algorithms}
 
-\bibliography{ref.bib}
-\DeclareUnicodeCharacter{2212}{-}
-\begin{document}
+\subsubsection{Random sampling}
 
-\maketitle
+\subsubsection{Gaussian Process Upper Confidence Bound}
 
-\section{Method 1}
-\subsection{Method1a}
+We use Gaussian Process Upper Confidence Bound (GPUCB)] algrithm \cite{srinivas2012information}, with sum of dot product kernel and spectrum kernel of sequences.
+The algorithm basically includes two parts:
 
-\end{document}
+\begin{enumerate}
+    \item the Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
+So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
+    \item bandit algorithms (Upper Confidence Bound), which recommends sequences to test for next round.
+\end{enumerate}

### 2020-05-13 main.tex
Updated main.tex and methods.tex 13/05/20

@@
 After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
-The results are then fed back to the algorithm for it to recommend the next round of designs.
-
-\subsection{First Round of Designs}
-
+The results are then fed back to the algorithm for it to recommend the next round of designs.\\
 The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}.
@@
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item 30 RBS sequences fully and uniformly randomised (equal probability of choosing each letter for each base) 
+    \item 30 RBS sequences fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
     \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
-    \item 60 RBS sequences by sequentially machine learning design.
-We use an bandit optimazation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) \cite{srinivas2012information}.
-We design the first round bandit recommendation based on the data from \textcite{jervis2018machine}.
-The data contains 113 non-repeated records for 56 unique RBS sequences with the TIR label.
-The label is between 0 - 100,000 and skewed, which is shown in Figure xx.
-We normalise the label to 0 - 1 using the min-max normalisation.
+    \item 60 RBS sequences recommended by initial machine learning approach.
+For this, and subsequent design rounds, we use bandit optimisation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) (see \textbf{Methods} section) \cite{srinivas2012information}.
+Since we don't have access to any prior data fitting our design we approximated the recommendations based on the data from \textcite{jervis2018machine}.
+This data contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR label.
+The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
+We have normalised the label to 0 - 1 using the min-max normalisation.
 \end{enumerate}{}
@@
 \end{figure}
+
+Based on the results from round one, we used the previously described algorithm to recommend another batch of 90 RBS.
+They have been then constructed and tested using the exact same methodology as the first round.
+In total, three more rounds have been performed, each with designs recommended by the algorithm.
 
@@
 
-\input{methods}
-
 \newpage

### 2020-05-14 methods.tex
Updated 14/05/20

@@
-\section{Method 1}
+\section{Method}
 This will be about the DNA assembly.\\
-\subsection{Method 1a}
+\subsection{SynBio methods}
 This will be very specifically talking about the plasmid.
@@
 
-\subsubsection{Random sampling}
+\subsubsection{Probability-based Random sampling}
 
-\subsubsection{Gaussian Process Upper Confidence Bound}
+\subsubsection{Machine learning experimental design}
 
-We use Gaussian Process Upper Confidence Bound (GPUCB)] algrithm \cite{srinivas2012information}, with sum of dot product kernel and spectrum kernel of sequences.
-The algorithm basically includes two parts:
+\begin{itemize}
+    \item \textit{Gaussian process regression (GPR)}.
+Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
+So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
+    \item \textit{Upper confidence bound (UCB)}. recommends sequences to test for next round.
+\end{itemize}
 
-\begin{enumerate}
-    \item the Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
-    \item bandit algorithms (Upper Confidence Bound), which recommends sequences to test for next round.
-\end{enumerate}
+Combine the two parts together, we apply \textit{Gaussian Process Upper Confidence Bound (GPUCB)} algorithm, which has been theoretically analysed by \textcite{srinivas2012information},
+
+\textbf{Choice of kernel function:} \textit{Spectrum kernel.}

### 2020-05-14 main.tex
Reconstruction and writing (from Maciej)
Updated 14/05/20
Updated Results 140520

@@
 
+\topmargin -.5in
+\textheight 9in
+\oddsidemargin -.25in
+\evensidemargin -.25in
+\textwidth 7in
+
 \title{Machine Learning based TIR predictor}
-\author{u6015325 }
+\author{}
 \date{\today{}}
@@
 
+\subsection{Sequence embedding}
+First problem to be tackled was how to embed the sequence to give features (sequences) a numerical form.
+As we will be using a Gaussian Process for regression we have investigated use of different types of kernels for embedding \cite{Ben-Hur2008}.
+We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}.
+Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
+More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+\\
+\\
+\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}
+
+\subsection{Initial predictions}
+
+\subsection{Second and subsequent rounds predictions}
+
 \begin{enumerate}

### 2020-05-14 methods.tex
Updated Results 140520

@@
 
+To automatically design the RBS sequences in batch using machine learning, we need to consider two parts: 
+1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence interval for the prediction. 
+2) Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence interval. 
+Such online learning approach provides the $\textit{exploitation-exploration balance}$. 
+We show the corresponding methods we use in the below.
+
+We consider our experimental design problem as the problem of sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence point, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values of each points in the selected set $\mathcal{S}_t$, i.e.
+$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume the noise is under Gaussian distribution with some unknown mean and variance).
+The noise influenced by the accuracy RBS calculator and other experimental interference (e.g. \textcolor{red}{To be added}). 
+Our goal is to pick RBS sequence with the largest possible TIR score after the total number of rounds $N$.
+
 \begin{itemize}
     \item \textit{Gaussian process regression (GPR)}.
-Gaussian Process regression model, which predicts the label (TIR) and how uncertain we are about our prediction (confidence width); Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
+We model $f$ as a sample from a \textit{Gaussian process} $GP(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
+We will introduce the kernel function we choose in the below.
+    A Gaussian process regression model \cite{Rasmussen2004},
+
     \item \textit{Upper confidence bound (UCB)}. recommends sequences to test for next round.
@@
 
-\textbf{Choice of kernel function:} \textit{Spectrum kernel.}
+\textbf{Choice of kernel function:} \textit{Spectrum kernel.} Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
+So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.

### 2020-05-14 main.tex
Updated results 140520

@@
 Since we don't have access to any prior data fitting our design we approximated the recommendations based on the data from \textcite{jervis2018machine}.
-This data contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR label.
-The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
-We have normalised the label to 0 - 1 using the min-max normalisation.
 \end{enumerate}{}
@@
 
-\subsection{Initial predictions}
+\subsection{Initial bandit recommended designs}
+As mentioned previously, for the first round of bandit-guided designs we used data-set available from \textcite{jervis2018machine}.
+This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR label.
+The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
+We have normalised the label to 0 - 1 using the min-max normalisation.
 

### 2020-05-14 methods.tex
Updated results 140520

@@
     \item \textit{Gaussian process regression (GPR)}.
+    A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach to regression which provide uncertainty measurements on predictions. 
     We model $f$ as a sample from a \textit{Gaussian process} $GP(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
 We will introduce the kernel function we choose in the below.
-    A Gaussian process regression model \cite{Rasmussen2004},
 
-    \item \textit{Upper confidence bound (UCB)}. recommends sequences to test for next round.
+    \item \textit{Upper confidence bound (UCB)}.
+Bandit algorithms \cite{lattimore2018bandit} provide various approaches to sequentially design an where an agent adaptively chooses one or more options among several actions based on certain policies.
+We consider a type of bandit algorithms, the upper confidence bound algorithm, which is based on the \textit{optimism in the face of uncertainty}.
+The UCB algorithm, as the name suggested, basically selects RBS sequences with the maximum upper confidence bound, i.e.
+$\operatorname{argmax}_{\mathbf{x_i \in \mathcal{D}}} \left( \mu_t(\mathbf{x_i}) + \beta_t \sigma_t(\mathbf{x_i})\right)$, where $\beta_T$ is a hyperparmeter balancing the exploitation and exploration, $\mu_t(\mathbf{x}_i), \sigma_t(\matbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
 \end{itemize}

### 2020-05-14 methods.tex
Draft for GPUCB description.

@@
 The UCB algorithm, as the name suggested, basically selects RBS sequences with the maximum upper confidence bound, i.e.
-$\operatorname{argmax}_{\mathbf{x_i \in \mathcal{D}}} \left( \mu_t(\mathbf{x_i}) + \beta_t \sigma_t(\mathbf{x_i})\right)$, where $\beta_T$ is a hyperparmeter balancing the exploitation and exploration, $\mu_t(\mathbf{x}_i), \sigma_t(\matbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
+$\operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_t(\mathbf{x}_i) + \beta_t \sigma_t(\mathbf{x}_i)\right)$,
+    where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
+    $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$. 
 \end{itemize}

### 2020-05-14 main.tex
Draft for GPUCB description.
Updated Results (Initial Bandit Design) 140520

@@
 The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
-We have normalised the label to 0 - 1 using the min-max normalisation.
+\\
+First, we have normalised the label to 0 - 1 using the min-max normalisation.
+Next, we have \textcolor{red}{transformed (not sure what would be the best term here} the sequences with the sum of kernels as described previously.
+Finally, Gaussian Process was used to process the data and allow us to recommend 60 RBS sequences.
 

### 2020-05-14 methods.tex
Finish spectrum kernel draft; add flowchart; add appendix (list things to be added).

@@
 
-\subsubsection{Probability-based Random sampling}
+\subsubsection{ Position probability matrix based sampling}
 
@@
     A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach to regression which provide uncertainty measurements on predictions. 
-    We model $f$ as a sample from a \textit{Gaussian process} $GP(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
-We will introduce the kernel function we choose in the below.
+    We model $f$ as a sample from a \textit{Gaussian process} $GP(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.\\
+    We choose to use \textit{spectrum kernel} \cite{leslie2001spectrum} to specify the kernel function of $GP$.  
+    The spectrum kernel is widely used for classifying protein sequences \cite{leslie2001spectrum, ben2008support}, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.
+More precisely, $k_\ell^{\text{spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\Phi_{\ell}^{\mathrm{spec}}(\mathbf{x}), \Phi_{\ell}^{\mathrm{spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle$, where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$.
+Denote the number of letters in the alphabet as $|\Sigma|$.
+$\Phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+    %Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
+So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
 
@@
 
-Combine the two parts together, we apply \textit{Gaussian Process Upper Confidence Bound (GPUCB)} algorithm, which has been theoretically analysed by \textcite{srinivas2012information},
+Combine the two parts together, we apply \textit{Gaussian Process Upper Confidence Bound (GPUCB)} algorithm, which has been theoretically analysed by \textcite{srinivas2012information}.
+We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}.
 
-\textbf{Choice of kernel function:} \textit{Spectrum kernel.} Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.35]{plots/flowchart.pdf}
+    \caption{Flowchart of machine learning based experimental design.}
+    \label{fig: flowchart of machine learning based experimental design.}
+\end{figure}

### 2020-05-18 main.tex
Finish spectrum kernel draft; add flowchart; add appendix (list things to be added).
Changed the Introduction to include the DBTL cycle.

@@
 One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-This can be done in variety of ways, most of each involve designing the DNA sequence in CAD software and then physically testing it in a laboratory.
-Alternative to this is computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function \cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.\\
+This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given part or organism are continually improved by going through a number of turns of the said cycle.
+This normally involves designing the DNA sequence in CAD software and then physically testing it in a laboratory.
+Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used\cite{Yeoh2019,Nielsen2016}.
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
+The biggest limitation for this approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve designs based on exprimental designs.\\
 According to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
@@
 Jervis \emph{et al.} used support vector machine and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
-Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway bahaviour \cite{Costello2018}.
+Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
 There were also successful attempts at using deep learning techniques for analysis of big data-sets \cite{Alipanahi2015,Angermueller2016}.
@@
 \\
-Here we present a machine learning workflow for building a RBS translation rate predictor based on fluorescence data from cultures.
+Here we present a machine learning workflow for building a RBS translation rate predictor based on fluorescence data from cultures as a way of improving the Learn capabilities of the DBTL cycle.
 RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
@@
 \printbibliography
+
+\clearpage
+
+\appendix
+\input{appendix}
 \end{document}

### 2020-05-18 main.tex
Reorganized the results section.

@@
 Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
-The biggest limitation for this approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve designs based on exprimental designs.\\
+The biggest limitation for this approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve designs based on experimental designs.\\
 According to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
@@
 
-\section{Genetic device design}
-There is a number of qualities that impact the translation rate, but chief among them are concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
+\input{methods}
+
+\section{Results}
+
+Our experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with minimising the number of DBTL cycle turns that we had to do.
+We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds).
+The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
+After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
+The results are then fed back to the algorithm for it to recommend the next round of designs.\\
+
+\subsection{DESIGN of the genetic device}
+There is a number of qualities that impact the protein expression rate, with many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
 In \emph{E.
@@
 By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
+The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}.
+In our design we focus on randomizing of the core -8 to -13 (relative to the GFP) bps of the RBS and fix others to be the same as the consensus sequence, i.e.
+TTTAAGA + NNNNNN + TATACAT.
+Since for each of the 6 position there are 4 possibilities: A, C, G, T the total feature space is $4^6$ = 4096.\\
+Since there was no prior data that we could have used to guide our design for the first round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that give good cover of the experimental space:
+
+\begin{enumerate}
+    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
+This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
+    \item 30 RBS sequences fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
+    \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
+    \item 60 RBS sequences recommended by initial machine learning approach.
+For this, and subsequent design rounds, we used bandit recommendation algorithm using predictions made by the Gaussian Process Upper Confidence Bound (GPUCB) algorithm (see \textbf{Methods} section) \cite{srinivas2012information}.
+Since we don't have access to any prior data fitting our design we approximated the recommendations based on the data from \textcite{jervis2018machine}.
+\end{enumerate}{}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.7]{plots/TIR_histogram.pdf}
+    \caption{TIR Histogram.}
+    \label{fig: TIR Histogram.}
+\end{figure}
+
+Based on the results from round one, we used the bandit algorithm to recommend another batch of 90 RBS based on the predictions made by the GPUCB algorithm (see \textbf{LEARN} and section \textbf{Methods}).
+They have been then constructed and tested using the exact same methodology as the first round.
+In total, three more rounds have been performed, each with designs recommended by the algorithm.
 
@@
 
-\input{methods}
+\subsection{BUILD and TEST}
 
-\section{Workflow Description}
-
-Our experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with as fewer rounds as possible.
-We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds).
-The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
-After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
-The results are then fed back to the algorithm for it to recommend the next round of designs.\\
-The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}.
-In our design we focus on randomizing of the core -8 to -13 (relative to the GFP) bps of the RBS and fix others to be the same as the consensus sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT.
-Since for each of the 6 position there are 4 possibilities: A, C, G, T the total feature space is $4^6$ = 4096.\\
-For the first round of experiments, we design 180 (two times 90) RBS sequences based on the consensus sequence:
-
-\begin{enumerate}
-    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
-This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item 30 RBS sequences fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
-    \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
-    \item 60 RBS sequences recommended by initial machine learning approach.
-For this, and subsequent design rounds, we use bandit optimisation algorithm called Gaussian Process Upper Confidence Bound (GPUCB) (see \textbf{Methods} section) \cite{srinivas2012information}.
-Since we don't have access to any prior data fitting our design we approximated the recommendations based on the data from \textcite{jervis2018machine}.
-\end{enumerate}{}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.7]{plots/TIR_histogram.pdf}
-    \caption{TIR Histogram.}
-    \label{fig: TIR Histogram.}
-\end{figure}
-
-Based on the results from round one, we used the previously described algorithm to recommend another batch of 90 RBS.
-They have been then constructed and tested using the exact same methodology as the first round.
-In total, three more rounds have been performed, each with designs recommended by the algorithm.
-
-\section{Results}
-
-\subsection{Sequence embedding}
+\subsection{LEARN - prediction of RBS performance}
 First problem to be tackled was how to embed the sequence to give features (sequences) a numerical form.
@@
 
-\subsection{Second and subsequent rounds predictions}
-
 \begin{enumerate}
@@
 
+\section{Summary}
+
 \newpage

### 2020-05-18 main.tex
Updated the DESIGN and BUILD and TEST sections with bandit and data explanation.

@@
 \usepackage{xcolor}
+\usepackage{textgreek}
 
@@
 RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-We have used Gaussian Process-Upper Confidence Bound and Bandits algorithms to analyse and optimise the initiation rates of the designed RBS.
+We have used Gaussian Process-Upper Confidence Bound (for predictions / LEARN) and Bandits (for recommendations / DESIGN) algorithms to analyse and optimise the initiation rates of the designed RBS.
 
@@
 Since we don't have access to any prior data fitting our design we approximated the recommendations based on the data from \textcite{jervis2018machine}.
+As mentioned previously, for the first round of bandit-guided designs we used data-set available from \textcite{jervis2018machine}.
+This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR label.
+The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
+First, we have normalised the label to 0 - 1 using the min-max normalisation.
+Next, the data has been processed using the GPUCB and bandit algorithms as described \textcolor{red}{elsewhere}.
 \end{enumerate}{}
@@
 \end{figure}
-
-Based on the results from round one, we used the bandit algorithm to recommend another batch of 90 RBS based on the predictions made by the GPUCB algorithm (see \textbf{LEARN} and section \textbf{Methods}).
-They have been then constructed and tested using the exact same methodology as the first round.
-In total, three more rounds have been performed, each with designs recommended by the algorithm.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
-    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
-The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.
-At the bottom a simplified SBOL diagram of the genetic device is shown.}
-    \label{fig: Anatomy of the randomized sequence.}
-\end{figure}
-
-\subsection{BUILD and TEST}
-
-\subsection{LEARN - prediction of RBS performance}
-First problem to be tackled was how to embed the sequence to give features (sequences) a numerical form.
-As we will be using a Gaussian Process for regression we have investigated use of different types of kernels for embedding \cite{Ben-Hur2008}.
-We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}.
-Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
-More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-\\
-\\
-\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}
-
-\subsection{Initial bandit recommended designs}
-As mentioned previously, for the first round of bandit-guided designs we used data-set available from \textcite{jervis2018machine}.
-This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR label.
-The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
-\\
-First, we have normalised the label to 0 - 1 using the min-max normalisation.
-Next, we have \textcolor{red}{transformed (not sure what would be the best term here} the sequences with the sum of kernels as described previously.
-Finally, Gaussian Process was used to process the data and allow us to recommend 60 RBS sequences.
 
@@
 
+Based on the results from round one, we used the multi-armed bandit algorithm to recommend another batch of 90 RBS based on the predictions made by the GPUCB algorithm (see \textbf{LEARN} and section \textbf{Methods}).
+The bandit algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
+As such, it is very useful in solving problems like ours - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
+In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space which balances two approaches: exploration and exploitation.
+Exploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
+The two approaches can be controlled with the \textbeta\enspace parameter.
+We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm toward the exploration with exploitation taking increasing role in later iterations.
+One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR).
+As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.\\
+\\For rounds beyond the first initial one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs.
+In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
+    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
+The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.
+At the bottom a simplified SBOL diagram of the genetic device is shown.}
+    \label{fig: Anatomy of the randomized sequence.}
+\end{figure}
+
+\subsection{BUILD and TEST}
+
+For the machine learning to work the analysed data-set needs two elements: high relative volume and high quality of data.
+These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and you have to cover 5-10\% of that space.
+And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations you need to ensure that the obtained results represent the real value as close as possible.\\
+
+\subsection{LEARN - prediction of RBS performance}
+First problem to be tackled was how to embed the sequence to give features (sequences) a numerical form.
+As we will be using a Gaussian Process for regression we have investigated use of different types of kernels for embedding \cite{Ben-Hur2008}.
+We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}.
+Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
+More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+\\
+\\
+\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}
+
+\subsection{EXIT}
 \section{Summary}

### 2020-05-19 main.tex
Updated introduction, Design and Build sections.

@@
 One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given part or organism are continually improved by going through a number of turns of the said cycle.
+This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
 This normally involves designing the DNA sequence in CAD software and then physically testing it in a laboratory.
 Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used\cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins among others) or empirically obtained values describing a relevant to design value, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
-The biggest limitation for this approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve designs based on experimental designs.\\
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
+The biggest limitation for this approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve designs based on experimental results.\\
 According to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
-Predictions from all of these models are relatively good ($R^2 >0.8$), there come with a number of caveats: i) they rely on calculations of free energies that can be hard to calculate ii) in general, the models' accuracy is improved by increasing the number of phenomenons taking place during the translation, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
+Predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
 \\
-Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments.
-\cite{Freemont2019} New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
-These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology \cite{Camacho2018}.
-In the past few years there was a significant uptick of Machine Learning based approaches to synthetic biology.
+Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}.
+New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
+These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly imrpoving the DBTL cycle's performance \cite{Camacho2018}.
+In the past few years there was a significant uptake of Machine Learning based approaches in synthetic biology.
 Jervis \emph{et al.} used support vector machine and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
@@
 \\
-Here we present a machine learning workflow for building a RBS translation rate predictor based on fluorescence data from cultures as a way of improving the Learn capabilities of the DBTL cycle.
+Here we present how machine learning algorithms can be used as a part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimization of protein level expression.
 RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-We have used Gaussian Process-Upper Confidence Bound (for predictions / LEARN) and Bandits (for recommendations / DESIGN) algorithms to analyse and optimise the initiation rates of the designed RBS.
+We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
 
@@
 
-Our experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with minimising the number of DBTL cycle turns that we had to do.
+Our overall experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with minimising the number of DBTL cycle turns that we had to do.
 We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds).
@@
 \subsection{DESIGN of the genetic device}
-There is a number of qualities that impact the protein expression rate, with many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
+There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
 In \emph{E.
-coli} the RBS is usually located in the 20 bases downstream of the start codon.
+coli} the RBS is usually located in the 20 bases upstream of the start codon.
 The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E.
@@
 Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).\\
-The RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant.
-By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring the device.
+In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant.
+By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring plasmid with the device.
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1.
 By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
-The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}.
-In our design we focus on randomizing of the core -8 to -13 (relative to the GFP) bps of the RBS and fix others to be the same as the consensus sequence, i.e.
+The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}.
+In our design we focus on randomising of the core -8 to -13 (relative to the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
 TTTAAGA + NNNNNN + TATACAT.
-Since for each of the 6 position there are 4 possibilities: A, C, G, T the total feature space is $4^6$ = 4096.\\
-Since there was no prior data that we could have used to guide our design for the first round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that give good cover of the experimental space:
+Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
+Since there was no prior data that we could have used to guide our design for the first round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:
 
@@
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item 30 RBS sequences fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
+    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
     \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
     \item 60 RBS sequences recommended by initial machine learning approach.
-For this, and subsequent design rounds, we used bandit recommendation algorithm using predictions made by the Gaussian Process Upper Confidence Bound (GPUCB) algorithm (see \textbf{Methods} section) \cite{srinivas2012information}.
-Since we don't have access to any prior data fitting our design we approximated the recommendations based on the data from \textcite{jervis2018machine}.
-As mentioned previously, for the first round of bandit-guided designs we used data-set available from \textcite{jervis2018machine}.
-This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR label.
+For this, and subsequent design rounds, we used multi-armed bandit recommendation algorithm using predictions made by the Gaussian Process Upper Confidence Bound (GPUCB) algorithm (see \textbf{Methods} section) \cite{srinivas2012information}.
+Since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine}.
+This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
 The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
 First, we have normalised the label to 0 - 1 using the min-max normalisation.
-Next, the data has been processed using the GPUCB and bandit algorithms as described \textcolor{red}{elsewhere}.
+Next, the data has been processed using the GPUCB and bandit algorithms as described below.
 \end{enumerate}{}
@@
 
-Based on the results from round one, we used the multi-armed bandit algorithm to recommend another batch of 90 RBS based on the predictions made by the GPUCB algorithm (see \textbf{LEARN} and section \textbf{Methods}).
 The bandit algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-As such, it is very useful in solving problems like ours - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
+As such, it is very useful in solving problems like the one presented here - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
 In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space which balances two approaches: exploration and exploitation.
@@
 As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.\\
-\\For rounds beyond the first initial one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs.
+\\
+For rounds beyond the first initial one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs.
 In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.
@@
 These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and you have to cover 5-10\% of that space.
-And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations you need to ensure that the obtained results represent the real value as close as possible.\\
+And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations one need to ensure that the obtained results represent the real value as close as possible.\\
+To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
+This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
+Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
+In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly.
+The plasmids were then transformed and the resulting transformants were tested both using microplate reader and flow cytometry.\\
 

### 2020-05-20 main.tex
Moved Learn section to the beginning; updated description of the GPR and bandits.

@@
 
+\textcolor{red}{DBTL cartoon}
+
+\subsection{LEARN - prediction of RBS performance}
+First problem to be tackled was how to embed the sequence to give our RBS sequences numerical form.
+As we will be using a Gaussian Process for regression we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}.
+The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimension space, which makes the regression process easier.
+We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}.
+Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
+More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.\\
+The predictions were made using the Gaussian Process regression (GPR) algorithm.
+In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on  perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
+Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic.
+\\
+\textcolor{red}{Figure X} shows \\
+
+\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}
+
 \subsection{DESIGN of the genetic device}
@@
     \item 60 RBS sequences recommended by initial machine learning approach.
-For this, and subsequent design rounds, we used multi-armed bandit recommendation algorithm using predictions made by the Gaussian Process Upper Confidence Bound (GPUCB) algorithm (see \textbf{Methods} section) \cite{srinivas2012information}.
+For this, and subsequent design rounds, we used multi-armed bandit recommendation algorithm using predictions made by the Gaussian Process regression (see \textbf{Methods} section) \cite{srinivas2012information}.
 Since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine}.
@@
 As such, it is very useful in solving problems like the one presented here - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
-In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space which balances two approaches: exploration and exploitation.
-Exploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
+In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space.
+In our case we use the Upper Confidence Bound \textcolor{red}{flavour} of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR).
+Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation.vExploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
 The two approaches can be controlled with the \textbeta\enspace parameter.
-We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm toward the exploration with exploitation taking increasing role in later iterations.
+We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations.
 One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR).
@@
 These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and you have to cover 5-10\% of that space.
-And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations one need to ensure that the obtained results represent the real value as close as possible.\\
+And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
 To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
@@
 
-\subsection{LEARN - prediction of RBS performance}
-First problem to be tackled was how to embed the sequence to give features (sequences) a numerical form.
-As we will be using a Gaussian Process for regression we have investigated use of different types of kernels for embedding \cite{Ben-Hur2008}.
-We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}.
-Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
-More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-\\
-\\
-\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}
+\textcolor{red}{what graphs will we show to present the results?}
 
 \subsection{EXIT}
+There are two potential points where the exit from the cycle should be considered: i) the optimum solution has been found or ii) depletion of available resources (time and/or money).
+
 \section{Summary}

### 2020-06-18 SynBio_ML_Writing/main.tex

--- 2020-05-20 main.tex
@@
 New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
-These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly imrpoving the DBTL cycle's performance \cite{Camacho2018}.
+These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018}.
 In the past few years there was a significant uptake of Machine Learning based approaches in synthetic biology.
@@
 
-Our overall experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with minimising the number of DBTL cycle turns that we had to do.
-We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds).
+Our overall experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
+We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (designated round zero since only preexisting, literature data is used for its design) or with RBS sequences recommended by the algorithm (subsequent rounds).
 The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
-After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
-The results are then fed back to the algorithm for it to recommend the next round of designs.\\
-
-\textcolor{red}{DBTL cartoon}
+After constructing, the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
+The results are then fed back to the algorithm for it to recommend the next round of designs.\\ \textcolor{red}{Figure X} shows the DBTL cycle with the machine learning process emphasised.
+Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase.
+In LEARN phase we use the Gaussian Process algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
+Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the learn phase, which we find a more natural starting point for a machine learning centred workflow.\\
 
@@
 As we will be using a Gaussian Process for regression we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}.
-The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimension space, which makes the regression process easier.
-We compared performance of Dot Product, RBS and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure X}.
+The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier.
+We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}.
 Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
-More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.\\
+More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+This approach allowed for good balance between computational complexity and performance.\\
 The predictions were made using the Gaussian Process regression (GPR) algorithm.
@@
 \\
-\textcolor{red}{Figure X} shows \\
-
-\textcolor{red}{Do we need a figure showing comparison of different kernel's performance? I think it would be useful}
+For the zeroth round, since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine} and the Gaussian Process regression method (see \textbf{Methods} section) \cite{srinivas2012information}.
+This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
+First, we have normalised the label to 0 - 1 using the min-max normalisation.
+The predictions are shown in \textcolor{red}{Figure Xb}\\
+The predictions for next rounds were predicted similarly, but with use of data obtained in previous rounds, visualisation of prediction for the last round of designs is shown in \textcolor{red}{Figure Xc} \\
 
@@
 Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
-Since there was no prior data that we could have used to guide our design for the first round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:
+Since there was no prior data that we could have used to guide our design for the zeroth round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:
 
@@
     \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
-    \item 60 RBS sequences recommended by initial machine learning approach.
-For this, and subsequent design rounds, we used multi-armed bandit recommendation algorithm using predictions made by the Gaussian Process regression (see \textbf{Methods} section) \cite{srinivas2012information}.
-Since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine}.
-This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
-First, we have normalised the label to 0 - 1 using the min-max normalisation.
-Next, the data has been processed using the GPUCB and bandit algorithms as described below.
+    \item 60 RBS sequences recommended by our multi armed bandit recommendation algorithm based on the initial literature, as described below.
 \end{enumerate}{}
@@
 In our case we use the Upper Confidence Bound \textcolor{red}{flavour} of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR).
-Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation.vExploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
+Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation.
+Exploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
 The two approaches can be controlled with the \textbeta\enspace parameter.
@@
 \\
-For rounds beyond the first initial one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs.
+For rounds beyond the initial, zeroth one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs.
 In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.
@@
 
-For the machine learning to work the analysed data-set needs two elements: high relative volume and high quality of data.
+For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
 These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and you have to cover 5-10\% of that space.
@@
 The plasmids were then transformed and the resulting transformants were tested both using microplate reader and flow cytometry.\\
+\textcolor{red}{Figure X} shows the results of the zeroth round.  \\
 
-\textcolor{red}{what graphs will we show to present the results?}
+\textcolor{red}{Figure X} shows the results of the first and subsequent rounds.\\
 
@@
 There are two potential points where the exit from the cycle should be considered: i) the optimum solution has been found or ii) depletion of available resources (time and/or money).
+In our case we have performed a total of four rounds, with performance of the RBSes increasing as seen in \textcolor{red}{Figure X}.\\
 

### 2020-07-12 codes/kernels_for_GPK.py
add wd shift kernel

@@ -272,8 +272,10 @@ class Spectrum_Kernel(Kernel):
 
     def distance(self, X, Y=None, eval_gradient=False, print_flag = False, plot_flag = False):
         """
-        Compute the distance between X and Y based on spectrum kernel:
-            d_{l}^{spectrum}(x, y) = sqrt(||phi(x) - phi(y)||^2)
+        
+        Compute the distance between x and y:
+        d(x,y) = sqrt(k(x,x) + k(y,y) - 2k(x,y))
+            
         for each pair of rows x in X and y in Y.
         when Y is None, Y is set to be equal to X.
 
@@ -301,33 +303,20 @@ class Spectrum_Kernel(Kernel):
             hyperparameter of the kernel. Only returned when eval_gradient
             is True.
         """
+        K = self.__call__(X,Y)
+        K_diag = K.diagonal()
 
-        if type(X[0,]) is not str and type(X[0,]) is not np.str_: 
-            X = inverse_label(X)           
+        K1 = np.zeros((len(K_diag), len(K_diag)))
+        K2 = np.zeros((len(K_diag), len(K_diag)))
 
-        if Y is None:
-            Y = X
-        elif type(Y[0,]) is not str  and type(Y[0,]) is not np.str_:
-            Y = inverse_label(Y)
+        for i, diag in enumerate(K_diag):
+            K1[i,:] = diag
+            K2[:,i] = diag
 
-        phi_X, phi_Y = Phi(X, Y, self.l_list, weight_flag = self.weight_flag, 
-                            padding_flag=self.padding_flag, gap_flag=self.gap_flag)
-
-        distance_matrix = np.zeros((phi_X.shape[0], phi_Y.shape[0]))
-
-        for i in range(phi_X.shape[0]):
-            for j in range(phi_Y.shape[0]):
-                if j >= i:
-                    #distance_matrix[i,j] = np.sqrt(np.sum(np.power((phi_X[i,:]- phi_Y[j,:]), 2)))
-                    distance_matrix[i,j] = np.linalg.norm(x = (phi_X[i,:] - phi_Y[j,:]), ord= 2)
-
-        for i in range(phi_X.shape[0]):
-            for j in range(phi_Y.shape[0]):
-                if j < i:
-                    distance_matrix[i,j] = distance_matrix[j,i]
-
-        return  distance_matrix, phi_X, phi_Y
+        distance_matrix = np.sqrt(K1+K2-2*K)
+        return  distance_matrix
         #return phi_X, phi_Y
+
     
[…]

### 2020-07-12 codes/kernels_for_GPK.py
add mixed spec kernel

@@ -552,6 +552,71 @@ class Sum_Spectrum_Kernel(Spectrum_Kernel):
             C.append(i[split_idx[-1]:])
         return np.asarray(A), np.asarray(B), np.asarray(C)  
 
+class Mixed_Spectrum_Kernel(Spectrum_Kernel):
+
+    def __call__(self, X, Y=None, eval_gradient=False, print_flag = False, plot_flag = False):
+        """
+        Compute the weighted degree kernel between X and Y:
+            k_{l}^{weighted degree}(x, y) 
+            = sum_{d=1}^{l}  beta_d k_d^{spectrum} (x, y)
+        for each pair of rows x in X and y in Y.
+        when Y is None, Y is set to be equal to X.
+
+        Parameters
+        ----------
+        X : array of shape (n_samples_X, ) or (n_sample_X, n_num_features)
+            each row is a sequence (string)
+            Left argument of the returned kernel k(X, Y)
+
+        Y : array of shape (n_samples_Y, ) or (n_sample_Y, n_num_features)
+            each row is a sequence (string)
+            Right argument of the returned kernel k(X, Y). If None, k(X, X)
+            if evaluated instead.
+
+        eval_gradient : bool (optional, default=False)
+            Determines whether the gradient with respect to the kernel
+            hyperparameter is determined. Only supported when Y is None.
+        Returns
+        -------
+        kernel_matrix : array of shape (n_samples_X, n_samples_Y)
+            Kernel k(X, Y)
+
+        K_gradient : array (opt.), shape (n_samples_X, n_samples_X, n_dims)
+            The gradient of the kernel k(X, X) with respect to the
+            hyperparameter of the kernel. Only returned when eval_gradient
+            is True.
+        """
+        if type(X[0,]) is not str and type(X[0,]) is not np.str_: 
+            X = inverse_label(X)  
+
+        if Y is None:
+            Y = X
+        elif type(Y[0,]) is not str  and type(Y[0,]) is not np.str_:
+            Y = inverse_label(Y)
+
+        K = np.zeros((X.shape[0], Y.shape[0]))
+        # assume all seq has the same total length
+        L = len(X[0])
+
+        assert len(self.l_list) == 1
+        l = self.l_list[0]
+
+        for d in range(1, l+1):
+            beta = 2 * float(l - d + 1)/float(l ** 2 + l)
+            K += beta * Spectrum_Kernel(l_list=[d]).__call__(X, Y)
+
[…]

### 2020-07-12 appendix.tex
add kernel choices into appendix

\section*{Appendix}

\section{Choices of Kernels}

\begin{itemize}
    \item Spectrum Kernel.
    \item Mixed Spectrum Kernel.
    \item Weighted Degree Kernel. 
     $\overline{\mathbf{x}}, \overline{\mathbf{x}}^{\prime}$ are two sequences over an alphabet $\Sigma,$ e.g., RBS sequences. By $|\Sigma|,$ we denote the number of letters in the alphabet. $\phi_{\ell}^{\text {Spec}}$ is a mapping of the sequence $\overline{\mathbf{x}}$ into a $|\Sigma|^{\ell}$ dimensional feature space. Each dimension corresponds to one of the $|\Sigma|^{\ell}$ possible strings $s$ of length $\ell$ and is the count of the
    number of occurrences of $s$ in $\overline{\mathbf{x}}$. 
    For sequences with fixed length $L$ and weighted degree kernel consider substrings starting at each position $l = 1, ..., L$, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$, \\
    \textit{Vector input}: 
    \begin{align}
        k_\ell^{WD}(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime) 
        &= \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d k_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]}, \overline{\mathbf{x}}_{[l:l+d]}^\prime)\\
        &= \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d \phi_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]})^T \phi_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]}^\prime)\\
        &= \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d \mathbb{I}(\overline{\mathbf{x}}_{[l:l+d]} = \overline{\mathbf{x}}_{[l:l+d]}^\prime),
    \end{align}
    where $\mathbb{I}(\text{true}) = 1$ and 0 otherwise. 
    \textit{Matrix input}: $\mathbf{X}, \mathbf{X}^\prime$ are two metrics which includes $n$ sequences, and $\Phi_d^{Spec}(\mathbf{X}) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$.
    \begin{align}
        K_\ell^{WD}(\mathbf{X}, \mathbf{X}^\prime)
        = \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d \Phi_d^{Spec}(\mathbf{X}_{[:, l:l+d]}) \Phi_d^{Spec}(\mathbf{X}_{[:, l:l+d]}^\prime)^T
    \end{align}

    \item Weighted Degree Kernel With Shift.
    \begin{align}
        k_\ell^{WDS}(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime) 
        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
        \left(k_d^{Spec}(\overline{\mathbf{x}}_{[l+s:l+s+d]}, \overline{\mathbf{x}}_{[l:l+d]}^\prime) + (k_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]}, \overline{\mathbf{x}}_{[l+s:l+s+d]}^\prime)\right)\\
        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
        \left(\mathbb{I}(\overline{\mathbf{x}}_{[l+s:l+s+d]} = \overline{\mathbf{x}}_{[l:l+d]}^\prime) + (\mathbb{I}(\overline{\mathbf{x}}_{[l:l+d]}= \overline{\mathbf{x}}_{[l+s:l+s+d]}^\prime)\right),
    \end{align}
    where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
    sequence, which we set to 1. $S(l)$ determines the shift
    range at position $l$.
\end{itemize}

\textbf{From kernel to distance}:
$$d(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime) = \sqrt{k(\overline{\mathbf{x}}, \overline{\mathbf{x}}) + k(\overline{\mathbf{x}}^\prime, \overline{\mathbf{x}}^\prime) - 2 k(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime)} $$

\section{Design Choices}

\begin{itemize}
    \item Embedding.
    \item Kernel.
    \item Regression model.
    \item Bandit model.
\end{itemize}

\section{Intuition behind UCB and visualisation}

\begin{itemize}
    \item Exploitation and exploration explanation.
    \item Visualise coverage by clustering plot.
    \item Table for in-clustering mean and variance.
\end{itemize}

\section{Result analysis}

\begin{itemize}
    \item violinplot.
    \item regression performance plot, table.
    \item kernel matrix plot.
\end{itemize}

\section{Figures}

### 2020-07-13 appendix.tex
Updates from Overleaf

@@
 
+We consider several string kernels below.
+
 \begin{itemize}
-    \item Spectrum Kernel.
-    \item Mixed Spectrum Kernel.
-    \item Weighted Degree Kernel. 
-     $\overline{\mathbf{x}}, \overline{\mathbf{x}}^{\prime}$ are two sequences over an alphabet $\Sigma,$ e.g., RBS sequences.
-By $|\Sigma|,$ we denote the number of letters in the alphabet.
-$\phi_{\ell}^{\text {Spec}}$ is a mapping of the sequence $\overline{\mathbf{x}}$ into a $|\Sigma|^{\ell}$ dimensional feature space.
-Each dimension corresponds to one of the $|\Sigma|^{\ell}$ possible strings $s$ of length $\ell$ and is the count of the
-    number of occurrences of $s$ in $\overline{\mathbf{x}}$. 
+    \item \textit{Spectrum Kernel.}
+    \begin{align}
+        k_\ell^{\text{Spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
+    \end{align}
+     where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. Denote the number of letters in the alphabet as $|\Sigma|$. 
+    $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+Let $\mathbf{X}, \mathbf{X}^\prime$ be two metrics which includes $n$ sequences, and $\Phi_d^{Spec}(\mathbf{X}) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, the spectrum kernel over metrics is
+    \begin{align}
+         K_\ell^{\text{Spec}}(\textbf{X}, \textbf{X}^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(\mathbf{X}) \Phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{X}^{\prime}\right)^T.
+    \end{align}
+
+    \item \textit{Sum of Spectrum Kernel,} considering weighted sum over different part of the string.
+
+    \item \textit{Mixed Spectrum Kernel,} considering weighted sum over different substring length, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$,
+        \begin{align}
+            k_\ell^{MixedSpec}(\mathbf{x}, \mathbf{x}^\prime) 
+            = \sum_{d=1}^{\ell} \beta_d k_d^{Spec}(\mathbf{x}, \mathbf{x}^\prime)
+        \end{align}
+    \item \textit{Weighted Degree Kernel,} considering positional information. WD kernel counts the match of kmers at corresponding positions in two sequences.
     For sequences with fixed length $L$ and weighted degree kernel consider substrings starting at each position $l = 1, ..., L$, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$, \\
-    \textit{Vector input}: 
     \begin{align}
-        k_\ell^{WD}(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime) 
-        &= \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d k_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]}, \overline{\mathbf{x}}_{[l:l+d]}^\prime)\\
-        &= \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d \phi_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]})^T \phi_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]}^\prime)\\
-        &= \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d \mathbb{I}(\overline{\mathbf{x}}_{[l:l+d]} = \overline{\mathbf{x}}_{[l:l+d]}^\prime),
+        k_\ell^{WD}(\mathbf{x}, \mathbf{x}^\prime) 
+        &= \sum_{d=1}^{\ell} beta_d \sum_{l=1}^{L-d+1} \gamma_l k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l:l+d]}^\prime)\\
+        &= \sum_{d=1}^{\ell} beta_d \sum_{l=1}^{L-d+1} \gamma_l \phi_d^{Spec}(\mathbf{x}_{[l:l+d]})^T \phi_d^{Spec}(\mathbf{x}_{[l:l+d]}^\prime)\\
+        &= \sum_{d=1}^{\ell} beta_d \sum_{l=1}^{L-d+1} \gamma_l \mathbb{I}(\mathbf{x}_{[l:l+d]} = \mathbf{x}_{[l:l+d]}^\prime),
     \end{align}
     where $\mathbb{I}(\text{true}) = 1$ and 0 otherwise.
-    \textit{Matrix input}: $\mathbf{X}, \mathbf{X}^\prime$ are two metrics which includes $n$ sequences, and $\Phi_d^{Spec}(\mathbf{X}) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$.
+
+    \item \textit{Weighted Degree Kernel With Shift.}
     \begin{align}
-        K_\ell^{WD}(\mathbf{X}, \mathbf{X}^\prime)
-        = \sum_{d=1}^{\ell} \sum_{l=1}^{L-d+1} \beta_d \Phi_d^{Spec}(\mathbf{X}_{[:, l:l+d]}) \Phi_d^{Spec}(\mathbf{X}_{[:, l:l+d]}^\prime)^T
-    \end{align}
-
-    \item Weighted Degree Kernel With Shift.
-    \begin{align}
-        k_\ell^{WDS}(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime) 
+        k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
         &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(k_d^{Spec}(\overline{\mathbf{x}}_{[l+s:l+s+d]}, \overline{\mathbf{x}}_{[l:l+d]}^\prime) + (k_d^{Spec}(\overline{\mathbf{x}}_{[l:l+d]}, \overline{\mathbf{x}}_{[l+s:l+s+d]}^\prime)\right)\\
+        \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
         &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(\mathbb{I}(\overline{\mathbf{x}}_{[l+s:l+s+d]} = \overline{\mathbf{x}}_{[l:l+d]}^\prime) + (\mathbb{I}(\overline{\mathbf{x}}_{[l:l+d]}= \overline{\mathbf{x}}_{[l+s:l+s+d]}^\prime)\right),
+        \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
     \end{align}
@@
 \textbf{From kernel to distance}:
-$$d(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime) = \sqrt{k(\overline{\mathbf{x}}, \overline{\mathbf{x}}) + k(\overline{\mathbf{x}}^\prime, \overline{\mathbf{x}}^\prime) - 2 k(\overline{\mathbf{x}}, \overline{\mathbf{x}}^\prime)} $$
+$$d(\mathbf{x}, \mathbf{x}^\prime) = \sqrt{k(\mathbf{x}, \mathbf{x}) + k(\mathbf{x}^\prime, \mathbf{x}^\prime) - 2 k(\mathbf{x}, \mathbf{x}^\prime)} $$
 

### 2020-07-15 codes/data_generating.py
generate data

@@ -5,6 +5,15 @@
 import numpy as np
 import pandas as pd
 import os
+import argparse
+
+parser = argparse.ArgumentParser(description='Generate valid data.')
+parser.add_argument('normalize_flag', default=True, help = 'indicates whether normalize labels. True for normalized label.')
+parser.add_argument('Format', default='Seq', help = 'Seq for rows as sequences; Sample for rows as samples')
+
+args = parser.parse_args()
+normalize_flag = bool(args.normalize_flag)
+data_format = args.Format
 
 sheet_name = '4h'
 Log_flag = False # indicates whether take log label
@@ -12,7 +21,7 @@ Norm_method = 'mean' # indicates how to normalize label (one of 'mean', 'minmax'
 Folder_path = os.getcwd() # folder path might need to change for different devices
 
 First_round_results_path = '/data/First_round_results/Results - First and Second Plate 3 reps.xlsx'
-
+Generated_File_Path = '/data/firstRound_' + sheet_name + '_norm' + str(normalize_flag) + '_format' + data_format + '.csv'
 def normalize(df, col_name):
     # take log FC -- possibly provide Gaussian distribution?
     if Log_flag:
@@ -33,50 +42,63 @@ def normalize(df, col_name):
 # Columns: 
 # RBS: 20-base RBS sequences
 # Name: Group and index 
-# Rep1, Rep2, Rep3: GFPOD for the 4h (using derivatives) for three biological replicates.
+# Rep1 - Rep5: GFPOD for the 4h (using derivatives) for three biological replicates.
 # AVERAGE: average value of replicates
 # STD: standard divation of replicates
 # PERC: STD/AVERAGE	
-# Group: one of 'bps', 'uni random', 'prob random', 'bandit' 
+# Group: one of 'bps', 'uni random', 'prob random', 'bandit'
+# Usable: Yes if valid; otherwise No
+# Replicates: the idx of replicates are valid to use
 
 Path_new = Folder_path + First_round_results_path
 df_new = pd.read_excel(Path_new, sheet_name= sheet_name)
-df_new['RBS'] = df_new['RBS'].str.upper() # convert to upper case
[…]

### 2020-08-07 codes/kernels_for_GPK.py
kernel normalisation, batch ucb

@@ -351,23 +351,41 @@ class Spectrum_Kernel(Kernel):
 
     
     def normalisation(self, kernel):
+        # https://jmlr.csail.mit.edu/papers/volume12/kloft11a/kloft11a.pdf
+        # Sec 4.2.2
+        # First calculate zero mean kernel
+        # Then calculate unit variance
+        # The the variance (defined as the 1/n trace - kernel mean) is 1 
 
-        spherical_kernel = np.zeros_like(kernel)
-        d = min(kernel.shape[0], kernel.shape[1])
-        for i in range(d):
-            for j in range(d):
-                spherical_kernel[i,j] = kernel[i,j]/np.sqrt(kernel[i,i] * kernel[j,j])
-        kernel = spherical_kernel
+        # zero mean
+        standardized_kernel = np.zeros_like(kernel)
+        kernel_mean = np.mean(kernel, axis = (0,1))
+        kernel_i_mean = []
+        kernel_j_mean = []
 
-        # standardized_kernel = np.zeros_like(kernel)
-        # kernel_mean = np.mean(kernel, axis = (0,1))
+        for i in range(kernel.shape[0]):
+            kernel_i_mean.append(kernel[i,:].mean())
+        for j in range(kernel.shape[1]):
+            kernel_j_mean.append(kernel[:,j].mean())
+                
+        for i in range(kernel.shape[0]):
+            for j in range(kernel.shape[1]):    
+                standardized_kernel[i,j] = kernel[i,j] - kernel_i_mean[i] - kernel_j_mean[j] + kernel_mean
         # n = kernel.shape[0]
         # kernel_trace = np.trace(kernel)
 
         # for i in range(kernel.shape[0]):
         #     for j in range(kernel.shape[1]):
         #         standardized_kernel[i,j] = kernel[i,j]/(1.0/n * kernel_trace  - kernel_mean)
-        # kernel = standardized_kernel
+        kernel = standardized_kernel
+        
+        # unit variance
+        spherical_kernel = np.zeros_like(kernel)
+        d = min(kernel.shape[0], kernel.shape[1])
+        for i in range(d):
+            for j in range(d):
+                spherical_kernel[i,j] = kernel[i,j]/np.sqrt(kernel[i,i] * kernel[j,j])
+        kernel = spherical_kernel
 
         return kernel        
 
@@ -742,7 +760,7 @@ class WeightedDegree_Kernel(Spectrum_Kernel):
[…]

### 2020-08-07 codes/regression.py

@@ -108,7 +108,7 @@ class GPR_Predictor():
         """
         
         train_df = pd.melt(self.df.loc[self.train_idx], id_vars=['RBS', 'RBS6', 'AVERAGE', 'STD', 'Group'], value_vars=['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5'])
-        train_df = train_df.dropna()
+        train_df = train_df.dropna(subset=['RBS', 'AVERAGE', 'value'])
         self.train_df = train_df.rename(columns = {'value': 'label'})
 
         #test_df = pd.melt(df.loc[test_idx], id_vars=['RBS', 'RBS6', 'AVERAGE', 'STD', 'Group'], value_vars=['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5'])
@@ -143,17 +143,18 @@ class GPR_Predictor():
             self.Train_test_split(random_state) # update train and test idx using random split
 
         X_train, X_test, y_train_sample, y_train_ave, y_test_ave, y_train_std, y_test_std=self.Generate_train_test_data()
-
+        print('X train shape: ', X_train.shape)
+        print('X test shape: ', X_test.shape)
         if self.kernel_name == 'WD_Kernel_Shift':
-            gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, s = self.s, normalise_flag = self.normalise_kernel), alpha = self.alpha)
+            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, s = self.s, normalise_flag = self.normalise_kernel), alpha = self.alpha)
         elif self.kernel_name == 'Sum_Spectrum_Kernel':
-            gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, b = self.b), alpha = self.alpha)
+            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, b = self.b), alpha = self.alpha)
         else:
-            gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list), alpha = self.alpha)
+            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list), alpha = self.alpha)
 
-        gp_reg.fit(X_train,y_train_sample)
-        y_train_pred_mean, y_train_pred_std = gp_reg.predict(X_train, return_std=True)
[…]

### 2020-08-08 codes/kernels_for_GPK.py
fix kernel normalisation

@@ -1,9 +1,16 @@
 import numpy as np
+import pandas as pd
 from sklearn.feature_extraction.text import CountVectorizer
-from sklearn.metrics.pairwise import check_pairwise_arrays
+#from sklearn.metrics.pairwise import check_pairwise_arrays
 from sklearn import preprocessing
 #from strkernel.mismatch_kernel import MismatchKernel, preprocess
 import matplotlib.pyplot as plt
+
+import os
+import sys
+module_path = os.path.abspath(os.path.join('../..'))
+if module_path not in sys.path:
+    sys.path.append(module_path)
 
 # To be able to normalise the kernel matrix
 # We cannot use Pairwise kernel for GP model
@@ -14,8 +21,14 @@ from sklearn.gaussian_process.kernels import Kernel, Hyperparameter
 
 BASES = ['A','C','G','T']
 
+# global features for all known and design data
+# used for normalisation
+Path = '../../data/known_design.csv'
+df = pd.read_csv(Path)
+FEATURES = np.asarray(df['RBS'])
 
-def Phi(X, Y, l_list = [3], j_X=0, j_Y=0, d=None, normalise_flag = True, weight_flag= False, padding_flag = False, gap_flag = False):
+
+def Phi(X, Y, l_list = [3], j_X=0, j_Y=0, d=None, normalise_phi_flag = False, weight_flag= False, padding_flag = False, gap_flag = False):
         """Calculate spectrum features for spectrum kernel.
 
         Phi is a mapping of the matrix X into a |alphabet|^l
@@ -96,7 +109,7 @@ def Phi(X, Y, l_list = [3], j_X=0, j_Y=0, d=None, normalise_flag = True, weight_
         embedded_X = embedded[: num_X, :].astype(float)
         embedded_Y = embedded[-num_Y: , :].astype(float)
         
-        if normalise_flag:
+        if normalise_phi_flag:
             # centering
             
             embedded_center_X = np.nanmean(embedded_X, axis = 0)
@@ -196,6 +209,9 @@ class Spectrum_Kernel(Kernel):
     l : list, default [3]
             list of number of l-mers (length of 'word')
             For example [2,3] means extracting all substrings of length 2 and 3 
+    features: array
+        list of features for all training and testing data
+        used for normalisation over all data, refer to function [normalisation]
     weight_flag: Boolean, default False
[…]

### 2020-08-08 codes/regression.py

@@ -141,16 +141,24 @@ class GPR_Predictor():
         
         if self.train_idx is None and self.test_idx is None :
             self.Train_test_split(random_state) # update train and test idx using random split
+        else:
+            self.test_size = len(self.test_idx)/(len(self.train_idx)+ len(self.test_idx))
 
         X_train, X_test, y_train_sample, y_train_ave, y_test_ave, y_train_std, y_test_std=self.Generate_train_test_data()
         print('X train shape: ', X_train.shape)
         print('X test shape: ', X_test.shape)
+
+        self.features = np.concatenate((X_train,  X_test), axis = 0)
+
         if self.kernel_name == 'WD_Kernel_Shift':
-            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, s = self.s, normalise_flag = self.normalise_kernel), alpha = self.alpha)
+            print('create kernel instance')
+            kernel_instance = self.kernel(l_list = self.l_list, features = self.features, test_size = self.test_size, s = self.s, normalise_kernel_flag = self.normalise_kernel)
+            print('finish creating kernel instance')
+            self.gp_reg = GaussianProcessRegressor(kernel = kernel_instance, alpha = self.alpha)
         elif self.kernel_name == 'Sum_Spectrum_Kernel':
-            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, b = self.b), alpha = self.alpha)
+            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, features = self.features, test_size = self.test_size, b = self.b), alpha = self.alpha)
         else:
-            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list), alpha = self.alpha)
+            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, features = self.features, test_size = self.test_size,), alpha = self.alpha)
 
         self.gp_reg.fit(X_train,y_train_sample)
         y_train_pred_mean, y_train_pred_std = self.gp_reg.predict(X_train, return_std=True)
@@ -264,10 +272,6 @@ class GPR_Predictor():
                 for l_idx, l_list in enumerate(l_lists):
                     for s_idx, s in enumerate(s_list):
                         
-                        if self.kernel_name == 'WD_Kernel_Shift':
[…]

### 2020-08-08 codes/data_generating.py
update pre-processed data with 5 replicates

@@ -8,16 +8,17 @@ import os
 import argparse
 
 parser = argparse.ArgumentParser(description='Generate valid data.')
-parser.add_argument('normalize_flag', default=True, help = 'indicates whether normalize labels. True for normalized label.')
+parser.add_argument('normalize_flag', default= 'True', help = 'indicates whether normalize labels. True for normalized label.')
 parser.add_argument('Format', default='Seq', help = 'Seq for rows as sequences; Sample for rows as samples')
 
 args = parser.parse_args()
-normalize_flag = bool(args.normalize_flag)
-data_format = args.Format
+normalize_flag = str(args.normalize_flag) # str True or False
+data_format = str(args.Format)
 
-sheet_name = '4h'
+sheet_name = 'Microplate'
 Log_flag = True # indicates whether take log label
 Norm_method = 'mean' # indicates how to normalize label (one of 'mean', 'minmax', None)
+Use_partial_rep = False
 Folder_path = os.getcwd() # folder path might need to change for different devices
 
 First_round_results_path = '/data/First_round_results/Results - First and Second Plate 3 reps.xlsx'
@@ -56,26 +57,31 @@ df_new = pd.read_excel(Path_new, sheet_name= sheet_name)
 # select only usable seq
 df_new_usable = df_new[df_new['Usable'] == 'Yes']
 
-# select only valid replicates
-df_new_valid = pd.DataFrame()
-df_new_valid[['Name', 'Group']] = df_new_usable[['Name', 'Group']]
-df_new_valid['RBS'] = df_new_usable['RBS'].str.upper() # convert to upper case
-df_new_valid['RBS6'] = df_new_usable['RBS'].str[7:13] # extract core part
-for idx, row in df_new_usable.iterrows():
-    vad_reps = []
-    for rep_idx in row['Replicates'].split(','):
-        rep_name = 'Rep' + str(rep_idx)
-        df_new_valid.loc[idx, rep_name] = row[rep_name]
-        vad_reps.append(row[rep_name])
-    df_new_valid.loc[idx, 'AVERAGE'] = np.mean(vad_reps)
-    df_new_valid.loc[idx, 'STD'] = np.std(vad_reps)
+if Use_partial_rep:
+    # select only valid replicates
+    df_new_valid = pd.DataFrame()
[…]

### 2020-08-08 codes/kernels_for_GPK.py
centering normalisation are commented, lead to non positive definite; add data visualisation

@@ -241,14 +241,14 @@ class Spectrum_Kernel(Kernel):
         self.test_size = test_size
         self.features = features
         if normalise_kernel_flag:
-            print('calculating kernel_all')
+            # print('calculating kernel_all')
             self.normalise_kernel_flag = False # only for calculate kernel_all
             self.kernel_all = self.__call__(self.features, self.features)
             self.kernel_all_mean0 = self.kernel_all.mean(axis = 0)
             self.kernel_all_mean1 = self.kernel_all.mean(axis = 1)
             self.kernel_all_mean = self.kernel_all.mean()
         
-            print('kernel_all shape: ', self.kernel_all.shape)
+            # print('kernel_all shape: ', self.kernel_all.shape)
         
 
         self.normalise_kernel_flag = normalise_kernel_flag
@@ -391,7 +391,7 @@ class Spectrum_Kernel(Kernel):
         # Then calculate unit variance
         # The the variance (defined as the 1/n trace - kernel mean) is 1 
         # normalise over the whole matrix (train, test)
-
+        '''
         # zero mean
         standardized_kernel = np.zeros_like(kernel)
         # kernel_mean = np.mean(kernel, axis = (0,1))
@@ -413,32 +413,29 @@ class Spectrum_Kernel(Kernel):
         #     for j in range(kernel.shape[1]):
         #         standardized_kernel[i,j] = kernel[i,j]/(1.0/n * kernel_trace  - kernel_mean)
         kernel = standardized_kernel
-        
+        print('After centering')
+        print(kernel.shape)
+        print(kernel)
+        '''
         # unit variance
         s0, s1 = kernel.shape
         spherical_kernel = np.zeros_like(kernel)
         if s0 == s1: # kernel over two same inputs
+            # print('1')
             for i in range(s0):
                 for j in range(s1):
                     spherical_kernel[i,j] = kernel[i,j]/np.sqrt(kernel[i,i] * kernel[j,j])
-        elif self.test_size < 0.5: # small dim is the testing data
-            if s0<s1: # k(test, train) i -> i+s1
-                for i in range(s0):
-                    for j in range(s1):
-                        spherical_kernel[i,j] = kernel[i,j]/np.sqrt(self.kernel_all[i+s1, i+s1] * self.kernel_all[j,j])
-            else: # k(train, test) j -> j+s0
-                for i in range(s0):
[…]

### 2020-08-08 codes/regression.py

@@ -282,9 +282,10 @@ class GPR_Predictor():
                             
                             self.features = np.concatenate((X_train,  X_test), axis = 0)
                             if self.kernel_name == 'WD_Kernel_Shift':
-                                gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = l_list, features = self.features, test_size = self.test_size, s = s), alpha = alpha)
+                                kernel_instance = self.kernel(l_list = l_list, features = self.features, test_size = self.test_size, s = s, normalise_kernel_flag = self.normalise_kernel)
+                                gp_reg = GaussianProcessRegressor(kernel = kernel_instance, alpha = alpha)
                             else:
-                                gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = l_list, features = self.features, test_size = self.test_size,), alpha = alpha)
+                                gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = l_list, features = self.features, test_size = self.test_size), alpha = alpha)
                         
                             gp_reg.fit(X_train, y_train_sample) # train with samples
                             y_train_predict = gp_reg.predict(X_train)

### 2020-08-08 appendix.tex
update kernel normalisation

@@
         k_\ell^{WD}(\mathbf{x}, \mathbf{x}^\prime) 
-        &= \sum_{d=1}^{\ell} beta_d \sum_{l=1}^{L-d+1} \gamma_l k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l:l+d]}^\prime)\\
-        &= \sum_{d=1}^{\ell} beta_d \sum_{l=1}^{L-d+1} \gamma_l \phi_d^{Spec}(\mathbf{x}_{[l:l+d]})^T \phi_d^{Spec}(\mathbf{x}_{[l:l+d]}^\prime)\\
-        &= \sum_{d=1}^{\ell} beta_d \sum_{l=1}^{L-d+1} \gamma_l \mathbb{I}(\mathbf{x}_{[l:l+d]} = \mathbf{x}_{[l:l+d]}^\prime),
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l:l+d]}^\prime)\\
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \phi_d^{Spec}(\mathbf{x}_{[l:l+d]})^T \phi_d^{Spec}(\mathbf{x}_{[l:l+d]}^\prime)\\
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \mathbb{I}(\mathbf{x}_{[l:l+d]} = \mathbf{x}_{[l:l+d]}^\prime),
     \end{align}
@@
 $$d(\mathbf{x}, \mathbf{x}^\prime) = \sqrt{k(\mathbf{x}, \mathbf{x}) + k(\mathbf{x}^\prime, \mathbf{x}^\prime) - 2 k(\mathbf{x}, \mathbf{x}^\prime)} $$
+
+\subsection{Normalisation of Kernel}
+
+As part of data pre-processing,
+the range of all features should be normalised so that each feature contributes approximately proportionately to the predictive model. 
+The kernel matrix is represented by the inner product of the underlying feature vectors, it need to be normalised before using in the downstream regression models. 
+Up-scaling (down-scaling) features can be understood as down-scaling (up-scaling) regularizers such that they penalise the features less (more).
+
+Here we consider two approaches for kernel normalisation: centering and unit norm. 
+We will show how to convert the normalisation in terms of feature vectors to normalisation in terms of kernel matrices. 
+As defined before, consider $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$.
+We denote $\phi(\mathbf{x}_i)$ as a column feature vector of sequence $\mathbf{x}_i$, 
+where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$. Assume there are totally $n$ sequences in the data $\mathbf{X}$ ($n'$ sequences in the data $\mathbf{X}'$). 
+We illustrate centering and unit norm normalisation below.
+
+\begin{itemize}
+    \item Centering. 
+    Define the mean vector $\bar{\Phi}(\mathbf{X}) = \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s) \in \mathbb{R}^d$, the centered feature vector $\phi^C(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    \begin{align}
+        \phi^{C}(\mathbf{x}_i) = \phi(\mathbf{x}_i) - \bar{\Phi}(\mathbf{X}) = \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s).
+    \end{align}
+    The corresponding centering kernel value between $\mathbf{x}_i$ and $\mathbf{x}_j$ is then 
+    \begin{align}
+        k^C(\mathbf{x}_i, \mathbf{x}_j) &= <\phi^C(\mathbf{x}_i), \phi^C(\mathbf{x}_j)>\\
+        &= \left( \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left( \phi(\mathbf{x}_j) - \frac{1}{n} \sum_{s' = 1}^n \phi(\mathbf{x}_{s'})\right)\\
+        &= \phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j) - \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \phi(\mathbf{x}_j) - \phi(\mathbf{x}_i)^T \left(\frac{1}{n} \sum_{s' = 1}^n \phi(\mathbf{x}_{s'})\right) + \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left(\frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        &= k(\mathbf{x}_i, \mathbf{x}_j) - \frac{1}{n} \sum_{s=1}^n k(\mathbf{x}_s, \mathbf{x}_j) - \frac{1}{n'} \sum_{s'=1}^{n'} k(\mathbf{x}_i, \mathbf{x}_{s'}) + \frac{1}{n^2} \sum_{s = 1}^n \sum_{s'=1}^{n'} k(\mathbf{x}_s, \mathbf{x}_{s'})
+    \end{align}
+
+    \item Unit Norm.
+Define the ($l_2$) norm of a feature vector $||\phi(\mathbf{x})|| = \sqrt{\sum_{m = 1}^d \phi_d(\mathbf{x})^2} = \sqrt{k(\mathbf{x}, \mathbf{x})} \in \mathbb{R}^+$, then the unit norm feature vector $\phi^{UN}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    \begin{align}
+        \phi^{UN}(\mathbf{x}_i) = \frac{\phi(\mathbf{x}_i)}{||\phi(\mathbf{x}_i)||}.
+    \end{align}
+    The corresponding unit norm kernel value between $\mathbf{x}_i$ and $\mathbf{x}_j$ is then 
+    \begin{align}
+        k^{UN}(\mathbf{x}_i, \mathbf{x}_j) &= <\frac{\phi(\mathbf{x}_i)}{||\phi(\mathbf{x}_i)||}, \frac{\phi(\mathbf{x}_j)}{||\phi(\mathbf{x}_j)||}>\\
+        &= \frac{\phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j)}{||\phi(\mathbf{x}_i)|| \times ||\phi(\mathbf{x}_j)||}\\
+        &= \frac{k(\mathbf{x}_i, \mathbf{x}_j)}{\sqrt{k(\mathbf{x}_i, \mathbf{x}_i)  k(\mathbf{x}_j, \mathbf{x}_j)}}
+    \end{align}
+
+     \item Unit Variance. 
+    After the centering and unit norm normalisation, the kernel matrix is unit variance as well. 
+    In the following, we show transformations of the unit variance (with centering) normalisation.
+    Define the variance vector ${Var}(\Phi(\mathbf{X})) = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \bar{\Phi}(\mathbf{X})||^2 = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \sum_{s'=1}^n \left(\phi(\mathbf{x}_s')\right)||^2 = \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \in \mathbb{R}$, the unit variance feature vector $\phi^{UV}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    \begin{align}
+        \phi^{UV}(\mathbf{x}_i) = \frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(\mathbf{X}))}}.
+    \end{align}
+    The corresponding kernel representation is 
+    \begin{align}
+        k^{UV}(\mathbf{x}_i, \mathbf{x}_j) &= <\frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(\mathbf{X}))}}, \frac{\phi(\mathbf{x}_j)}{\sqrt{Var(\Phi(\mathbf{X}'))}}>\\
+        &= \frac{\phi(\mathbf{x}_i)^T \mathbf{x}_j}{\sqrt{Var(\Phi(\mathbf{X})) Var(\Phi(\mathbf{X'}))}}\\
+        &= \frac{k(\mathbf{x}_i, \mathbf{x}_j)}{\sqrt{ \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \frac{1}{n} \sum_{s'=1}^{n'}  k^C(\mathbf{x}_{s'}, \mathbf{x}_{s'})}}
+    \end{align}
+    After centering and unit norm, $ \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s) = k(\mathbf{x}_i, \mathbf{x}_i)$, which implies that after centering and unit norm, the kernel matrix is already unit variance normalised.
+\end{itemize}
+For the Gaussian Process regression, we make of use of two kernel matrices: the kernel function between the training data itself, i.e. $K(\mathbf{X}_{train}, \mathbf{X}_{train})$; 
+the kernel function taking the training data and testing data as inputs, i.e. $K(\mathbf{X}_{test}, \mathbf{X}_{train})$. 
+%It is straightforward to normalise a square kernel which same input, i.e. $n = n'$ and $k(\mathbf{x}_i, \mathbf{x}_i),  k(\mathbf{x}_j, \mathbf{x}_j)$ taken from the diagonal of the matrix. 
+%The second one (between train and test) is a little bit tricky. The different is not only that $n \neq n'$. 
+We will state two ways of normalise those two kind of matrices:
+\begin{itemize}
+    \item Normalise training and testing data separately.
+    This approach is preferred for most of the machine learning algorithms since it follows the rule we have no information about testing data while training.
+    Then for centering, one should subtract the mean vector over the training data for both kinds of matrices.
+    For unit norm normalisation, when one calculate $K^{UN}(\mathbf{X}_{test}, \mathbf{X}_{train})$, the two terms inside of square root: $k(\mathbf{x}_i, \mathbf{x}_i)$ is taken from $K(\mathbf{X}_{test}, \mathbf{X}_{test})[i,i]$, and $k(\mathbf{x}_j, \mathbf{x}_j)$ is taken from $K(\mathbf{X}_{train}, \mathbf{X}_{train})[j,j]$.
+
+    \item Normalise training and testing data together, i.e. normalise $K(\mathbf{X}_{train+test}, \mathbf{X}_{train+test})$, then extra the parts we need from the normalised matrix. 
+    This approach is suitable for the case one already know the whole testing features. 
+    For centering, one should subtract the mean vector over the whole matrix $\Phi(\mathbf{X}_{train+test})$. 
+    The unit norm normalisation is the same as the previous case. 
+\end{itemize}
+
+For our experiment, we fix the design space before training, i.e. the testing features are already known before testing. 
+So we choose to normalise the kernel matrix over the training and testing data together,
+by first applying centering and then unit norm normalisation.
 

### 2020-08-24 appendix.tex
add gp.

@@
 \section*{Appendix}
+
+\section{Gaussian Process Regression}
+
+A \textit{Gaussian process} is a collection of random variables, any finite number of which have a joint Gaussian distribution. 
+We define mean function $m(\mathbf{x})$  and covariance function $k(\mathbf{x}, \mathbf{x}^\prime)$ of a real process $f(\mathbf{x})$ as
+\begin{align}
+    m(\mathbf{x}) &= \mathbb{E}[f(\mathbf{x})]\\
+    k(\mathbf{x}, \mathbf{x}^\prime) &= \mathbb{E}[(f(\mathbf{x}) - m(\mathbf{x}))(f(\mathbf{x}^\prime) - m(\mathbf{x}^\prime))].
+\end{align}
+
+A Gaussian process is specified by its mean function and convariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(m(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
+We consider the case where the observations are noisy, i.e. $\{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \sigma_n^2)$. 
+The Gaussian noise is independent identically distributed, and the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\sigma_{n}^{2} \delta_{p q}$,
+where $\delta_{pq}$ is a Kronecker delta which is one iff $p = q$ and zero otherwise.
+It is equivalent to add a diagonal matrix $\sigma_n^2 I$ on the kernel matrix evaluated on the training points.
+
+For $n_\ast$ test points $X_\ast$, assume the prior over the functions values as a random Gaussian vector $\mathbf{f}_\ast \sim \mathcal{N}(\mathbf{0}, K(X_\ast, X_\ast))$.
+Then the joint distribution of the observed target values and the function values at the test points under the prior as 
+\begin{align}
+    \left[\begin{array}{l}\mathbf{y} \\ \mathbf{f}_{*}\end{array}\right] \sim \mathcal{N}\left(\mathbf{0},\left[\begin{array}{cc}K(X, X)+\sigma_{n}^{2} I & K\left(X, X_{*}\right) \\ K\left(X_{*}, X\right) & K\left(X_{*}, X_{*}\right)\end{array}\right]\right)
+\end{align}
+where $K(X, X_\ast)$ denotes the $n \times n_\ast$ covariance/Kernel matrix evaluated at all pairs of training and testing points, similarly for other kernel matrices.
+Then the posterior of the test points (i.e.
+predictive distributions) is given by the conditional distribution $\mathbf{f}_\ast | X, \mathbf{y}, X_\ast \sim \mathcal{N}(\bar{\mathbf{f}}_\ast, cov(\mathbf{f}_\ast))$, where
+\begin{align}
+   \overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\sigma_{n}^{2} I\right]^{-1} \mathbf{y} \\
+   \label{Eq: predicted variance}
+   \operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\sigma_{n}^{2} I\right]^{-1} K\left(X, X_{*}\right) 
+\end{align}
+For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\sigma_n^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq. (\ref{Eq: predicted variance}).
+% We now introduce the marginal likelihood (or evidence) $p(\mathbf{y}|X)$, is which the integral of the likelihood times the prior 
+% \begin{align}
+%     p(\mathbf{y} \mid X)=\int p(\mathbf{y} \mid \mathbf{f}, X) p(\mathbf{f} \mid X) d \mathbf{f}
+% \end{align}
+The choice of covariance function is critical for the performance of Gaussian process regression, we show the choices of string kernel in next section.
 
@@
     $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
-Let $\mathbf{X}, \mathbf{X}^\prime$ be two metrics which includes $n$ sequences, and $\Phi_d^{Spec}(\mathbf{X}) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, the spectrum kernel over metrics is
+Let $X, X^\prime$ be two metrics which includes $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, the spectrum kernel over metrics is
     \begin{align}
-         K_\ell^{\text{Spec}}(\textbf{X}, \textbf{X}^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(\mathbf{X}) \Phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{X}^{\prime}\right)^T.
+         K_\ell^{\text{Spec}}(\textbf{X}, \textbf{X}^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
     \end{align}
@@
     where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
-    sequence, which we set to 1. $S(l)$ determines the shift
+    sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
     range at position $l$.
@@
 We denote $\phi(\mathbf{x}_i)$ as a column feature vector of sequence $\mathbf{x}_i$, 
-where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$. Assume there are totally $n$ sequences in the data $\mathbf{X}$ ($n'$ sequences in the data $\mathbf{X}'$). 
+where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$. Assume there are totally $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
 We illustrate centering and unit norm normalisation below.
@@
     \item Centering. 
-    Define the mean vector $\bar{\Phi}(\mathbf{X}) = \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s) \in \mathbb{R}^d$, the centered feature vector $\phi^C(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    Define the mean vector $\bar{\Phi}(X) = \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s) \in \mathbb{R}^d$, the centered feature vector $\phi^C(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
     \begin{align}
-        \phi^{C}(\mathbf{x}_i) = \phi(\mathbf{x}_i) - \bar{\Phi}(\mathbf{X}) = \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s).
+        \phi^{C}(\mathbf{x}_i) = \phi(\mathbf{x}_i) - \bar{\Phi}(X) = \phi(\mathbf{x}_i) - \frac{1}{n'} \sum_{s = 1}^{n'} \phi(\mathbf{x}_s).
     \end{align}
@@
         k^C(\mathbf{x}_i, \mathbf{x}_j) &= <\phi^C(\mathbf{x}_i), \phi^C(\mathbf{x}_j)>\\
-        &= \left( \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left( \phi(\mathbf{x}_j) - \frac{1}{n} \sum_{s' = 1}^n \phi(\mathbf{x}_{s'})\right)\\
-        &= \phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j) - \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \phi(\mathbf{x}_j) - \phi(\mathbf{x}_i)^T \left(\frac{1}{n} \sum_{s' = 1}^n \phi(\mathbf{x}_{s'})\right) + \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left(\frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        &= \left( \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left( \phi(\mathbf{x}_j) - \frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        &= \phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j) - \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \phi(\mathbf{x}_j) - \phi(\mathbf{x}_i)^T \left(\frac{1}{n} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right) + \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left(\frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
         &= k(\mathbf{x}_i, \mathbf{x}_j) - \frac{1}{n} \sum_{s=1}^n k(\mathbf{x}_s, \mathbf{x}_j) - \frac{1}{n'} \sum_{s'=1}^{n'} k(\mathbf{x}_i, \mathbf{x}_{s'}) + \frac{1}{n^2} \sum_{s = 1}^n \sum_{s'=1}^{n'} k(\mathbf{x}_s, \mathbf{x}_{s'})
@@
     In the following, we show transformations of the unit variance (with centering) normalisation.
-    Define the variance vector ${Var}(\Phi(\mathbf{X})) = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \bar{\Phi}(\mathbf{X})||^2 = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \sum_{s'=1}^n \left(\phi(\mathbf{x}_s')\right)||^2 = \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \in \mathbb{R}$, the unit variance feature vector $\phi^{UV}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    Define the variance vector ${Var}(\Phi(X)) = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \bar{\Phi}(X)||^2 = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \sum_{s'=1}^n \left(\phi(\mathbf{x}_s')\right)||^2 = \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \in \mathbb{R}$, the unit variance feature vector $\phi^{UV}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
     \begin{align}
-        \phi^{UV}(\mathbf{x}_i) = \frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(\mathbf{X}))}}.
+        \phi^{UV}(\mathbf{x}_i) = \frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(X))}}.
     \end{align}
@@
     \begin{align}
-        k^{UV}(\mathbf{x}_i, \mathbf{x}_j) &= <\frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(\mathbf{X}))}}, \frac{\phi(\mathbf{x}_j)}{\sqrt{Var(\Phi(\mathbf{X}'))}}>\\
-        &= \frac{\phi(\mathbf{x}_i)^T \mathbf{x}_j}{\sqrt{Var(\Phi(\mathbf{X})) Var(\Phi(\mathbf{X'}))}}\\
+        k^{UV}(\mathbf{x}_i, \mathbf{x}_j) &= <\frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(X))}}, \frac{\phi(\mathbf{x}_j)}{\sqrt{Var(\Phi(X'))}}>\\
+        &= \frac{\phi(\mathbf{x}_i)^T \mathbf{x}_j}{\sqrt{Var(\Phi(X)) Var(\Phi(\mathbf{X'}))}}\\
         &= \frac{k(\mathbf{x}_i, \mathbf{x}_j)}{\sqrt{ \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \frac{1}{n} \sum_{s'=1}^{n'}  k^C(\mathbf{x}_{s'}, \mathbf{x}_{s'})}}
@@
 \end{itemize}
-For the Gaussian Process regression, we make of use of two kernel matrices: the kernel function between the training data itself, i.e. $K(\mathbf{X}_{train}, \mathbf{X}_{train})$; 
-the kernel function taking the training data and testing data as inputs, i.e. $K(\mathbf{X}_{test}, \mathbf{X}_{train})$. 
+For the Gaussian Process regression, we make of use of two kernel matrices: the kernel function between the training data itself, i.e. $K(X_{train}, X_{train})$; 
+the kernel function taking the training data and testing data as inputs, i.e. $K(X_{test}, X_{train})$. 
 %It is straightforward to normalise a square kernel which same input, i.e. $n = n'$ and $k(\mathbf{x}_i, \mathbf{x}_i),  k(\mathbf{x}_j, \mathbf{x}_j)$ taken from the diagonal of the matrix. 
@@
     Then for centering, one should subtract the mean vector over the training data for both kinds of matrices.
-    For unit norm normalisation, when one calculate $K^{UN}(\mathbf{X}_{test}, \mathbf{X}_{train})$, the two terms inside of square root: $k(\mathbf{x}_i, \mathbf{x}_i)$ is taken from $K(\mathbf{X}_{test}, \mathbf{X}_{test})[i,i]$, and $k(\mathbf{x}_j, \mathbf{x}_j)$ is taken from $K(\mathbf{X}_{train}, \mathbf{X}_{train})[j,j]$.
+    For unit norm normalisation, when one calculate $K^{UN}(X_{test}, X_{train})$, the two terms inside of square root: $k(\mathbf{x}_i, \mathbf{x}_i)$ is taken from $K(X_{test}, X_{test})[i,i]$, and $k(\mathbf{x}_j, \mathbf{x}_j)$ is taken from $K(X_{train}, X_{train})[j,j]$.
 
-    \item Normalise training and testing data together, i.e. normalise $K(\mathbf{X}_{train+test}, \mathbf{X}_{train+test})$, then extra the parts we need from the normalised matrix. 
+    \item Normalise training and testing data together, i.e. normalise $K(X_{train+test}, X_{train+test})$, then extra the parts we need from the normalised matrix. 
     This approach is suitable for the case one already know the whole testing features. 
-    For centering, one should subtract the mean vector over the whole matrix $\Phi(\mathbf{X}_{train+test})$. 
+    For centering, one should subtract the mean vector over the whole matrix $\Phi(X_{train+test})$. 
     The unit norm normalisation is the same as the previous case. 
@@
 
-\section{Design Choices}
+\section{Batch Recommendation}
 
-\begin{itemize}
-    \item Embedding.
-    \item Kernel.
-    \item Regression model.
-    \item Bandit model.
-\end{itemize}
+For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
+%which is based on the \textit{optimism in the face of uncertainty}, 
+selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
+\begin{align}
+\label{Eq: GPUCB}
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
+\end{align}
+where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
+
+Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
+Therefore we consider recommending sequences in batch and use Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
+A key property of Gaussian Process regression is that the predictive variance in Eq.
+(\ref{Eq: predicted variance}) only depends on observed points (i.e.
+features), but not on the labels of those observed points.
+So one can compute the posterior variance without actually observing the labels. 
+The GP-BUCB policy is to select sequences that
+\begin{align}
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
+\end{align}
+where $fb[t]$ ...
 

### 2020-08-24 codes/kernels_for_GPK.py
add compare with onehot, RBF; add one more hyperparameter signal std to kernel

@@ -62,6 +62,7 @@ class String_Kernel(Kernel):
 
     # TODO: change hyperparameters maybe (now set to the same as Dotproduct)
     sigma_0 : float >= 0, default: 0.0
+        signal std
         Parameter controlling the inhomogenity of the kernel. If sigma_0=0,
         the kernel is homogenous.
     sigma_0_bounds : pair of floats >= 0, default: (1e-5, 1e5)
@@ -74,7 +75,7 @@ class String_Kernel(Kernel):
 
     def __init__(self, l=3, features = FEATURES, n_train = None, n_test = None,
                  padding_flag = False, gap_flag = False,
-                 # sigma_0=0.0, sigma_0_bounds=(1e-10,1e10)):
+                 sigma_0= 1 #, sigma_0_bounds=(1e-10,1e10)):
                 ):
         
         self.l = l
@@ -109,7 +110,8 @@ class String_Kernel(Kernel):
             # L = np.linalg.cholesky(self.kernel_all)
             # print('kernel_all is positive definite')
 
-            self.kernel_all_normalised = self.normalisation(self.kernel_all)
+            # TODO: decide whether to use sigma_0
+            self.kernel_all_normalised = self.normalisation(self.kernel_all) * sigma_0
             #print(np.linalg.eigh(self.kernel_all_normalised)[0])
             
             # check positive definite
@@ -485,7 +487,9 @@ class WD_Shift_Kernel(String_Kernel):
                     if s + j <= L:
                         beta = 2 * float(self.l - d + 1)/float(self.l ** 2 + self.l)
                         delta = 1.0/(2 * (s + 1))
-                        K += beta * delta * \
+                        # Todo: adding gamma? RASE: recognition of alternatively spliced exons in C.elegans.
+                        gamma = 1.0/L
+                        K += beta * delta * gamma * \
                             (self.dotproduct_phi(X, Y, l=d, j_X=j+s, j_Y=j, d=d) + 
                             self.dotproduct_phi(X, Y, l=d, j_X=j, j_Y=j+s, d=d))
 

### 2020-08-24 codes/regression.py

@@ -32,7 +32,8 @@ KERNEL_DICT = {
     # 'Mixed_Spectrum_Kernel': Mixed_Spectrum_Kernel,
     # 'WD_Kernel': WeightedDegree_Kernel,
     # 'Sum_Spectrum_Kernel': Sum_Spectrum_Kernel,
-    'WD_Kernel_Shift': WD_Shift_Kernel
+    'WD_Kernel_Shift': WD_Shift_Kernel,
+    'RBF': RBF
     
 }
 
@@ -205,7 +206,9 @@ class GPR_Predictor():
         #     self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, features = self.features, test_size = self.test_size, b = self.b), alpha = self.alpha)
         # else:
         #     self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, features = self.features, test_size = self.test_size,), alpha = self.alpha)
-        
+        elif self.kernel_name == 'RBF':
+            self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(length_scale = 1), alpha = self.alpha, n_restarts_optimizer = 5)
+
         print('gp_reg fit')
         self.gp_reg.fit(X_train,y_train_sample)
         print('gp_reg pred')
@@ -380,9 +383,11 @@ class GPR_Predictor():
                                                             s = s) \
                                                 + WhiteKernel(noise_level=1e-5, noise_level_bounds=(1e-5, 1e+5))
                                 gp_reg = GaussianProcessRegressor(kernel = kernel_instance, alpha = alpha, n_restarts_optimizer = 0)
-                            else:
-                                # TODO: tidy up other cases
-                                gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = l_list, features = self.features, test_size = self.test_size), alpha = alpha)
+                            elif self.kernel_name == 'RBF':
+                                gp_reg = GaussianProcessRegressor(kernel = self.kernel(length_scale = 1), alpha = self.alpha, n_restarts_optimizer = 3)
+                            # else:
+                            #     # TODO: tidy up other cases
+                            #     gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = l_list, features = self.features, test_size = self.test_size), alpha = alpha)
                         
                             gp_reg.fit(X_train, y_train_sample) # train with samples
                             y_train_predict = gp_reg.predict(X_train)

### 2020-08-25 codes/batch_ucb.py
kfold for sigma0; second round design finish.

@@ -70,8 +70,8 @@ class RBS_UCB():
         balance exploration and exploitation
     
     """
-    def __init__(self, df_known, kernel_name='WD_Kernel_Shift', l=6, s=1,
-                embedding='label', alpha=0.1, rec_size=90, beta=1):
+    def __init__(self, df_known, kernel_name='WD_Kernel_Shift', l=6, s=1, sigma_0 = 1,
+                embedding='label', alpha=2, rec_size=90, beta=2):
         self.df_known = df_known
         self.df_known['train_test'] = 'Train'
         self.known_rbs_set = set(self.df_known['RBS'])
@@ -83,6 +83,7 @@ class RBS_UCB():
         self.kernel_name = kernel_name
         self.l = l
         self.s = 1
+        self.sigma_0 = sigma_0 
         
         self.embedding = embedding
         self.alpha = alpha
@@ -96,6 +97,7 @@ class RBS_UCB():
                 embedding=self.embedding,
                 l=self.l, 
                 s = self.s,
+                sigma_0= self.sigma_0,
                 eva_on = 'seq' # for design
                 )
 

### 2020-08-25 codes/kernels_for_GPK.py

@@ -438,7 +438,7 @@ class WD_Shift_Kernel(String_Kernel):
 
     def __init__(self, l=3, features = FEATURES, n_train = None, n_test = None,
                 padding_flag = False, gap_flag = False,
-                #sigma_0=1e-10, sigma_0_bounds=(1e-10,1e10),
+                sigma_0= 1, #, sigma_0_bounds=(1e-10,1e10),
                 s = 0):
         """
 
@@ -449,7 +449,7 @@ class WD_Shift_Kernel(String_Kernel):
         """
         self.s = s
         super().__init__(l, features, n_train, n_test,
-                 padding_flag, gap_flag,) 
+                 padding_flag, gap_flag, sigma_0) 
                  #sigma_0, sigma_0_bounds)
 
     def cal_kernel(self, X, Y=None, eval_gradient=False, print_flag = False, plot_flag = False):

### 2020-08-25 codes/regression.py

@@ -39,8 +39,8 @@ KERNEL_DICT = {
 
 class GPR_Predictor():
     def __init__(self, df, test_size=0.2, train_idx = None, test_idx = None, 
-                 kernel_name='WD_Kernel_Shift', l=3, s = 0, b=0.33, padding_flag=False, gap_flag=False,
-                 alpha=0.5, embedding='label', eva_metric=[r2_score, mean_squared_error],  eva_on = "samples"
+                 kernel_name='WD_Kernel_Shift', l=6, s = 1, b=0.33, sigma_0 = 1, padding_flag=False, gap_flag=False,
+                 alpha=2, embedding='label', eva_metric=[r2_score, mean_squared_error],  eva_on = "samples"
                  ):
         """
         Parameter
@@ -64,6 +64,9 @@ class GPR_Predictor():
             shift length, only used for WD_Kernel_Shift
         b: float
             weight for sum of spectrum kernel
+        sigma_0: float
+            signal std; value multiply to normalised kernel,
+            i.e. https://drafts.distill.pub/gp/#section-4.2
         padding_flag: Boolean, default False
             indicates whether adding padding characters before and after sequences
         gap_flag: Boolean, default False
@@ -92,6 +95,7 @@ class GPR_Predictor():
         self.l = l
         self.s = s
         self.b = b
+        self.sigma_0 = sigma_0
         self.padding_flag = padding_flag
         self.gap_flag = gap_flag
         
@@ -194,7 +198,7 @@ class GPR_Predictor():
             print('create kernel instance')
             kernel_instance = self.kernel(l = self.l, features = self.features, 
                                          n_train=X_train.shape[0], n_test=X_test.shape[0],
-                                         s = self.s) \
+                                         s = self.s, sigma_0=self.sigma_0) \
                             + WhiteKernel(noise_level=1e-5, noise_level_bounds=(1e-5, 1e+5))
             # debug
             self.kernel_instance = kernel_instance
@@ -313,6 +317,7 @@ class GPR_Predictor():
     # TODO: not ready to use
     def Repeated_kfold(self, num_split = 5, num_repeat = 10, 
                        alpha_list = [0.1, 1], l_list = [3], s_list = [0], 
+                       sigma_0_list = [1],
                        eva_on_list = ['samples', 'seq'],
                        eva_metric_list = [mean_squared_error, r2_score]):
         """Repeated kfold for hyparameter choosing.
@@ -330,6 +335,8 @@ class GPR_Predictor():
[…]

### 2020-08-25 sub/SynBio_ML_Writing/appendix.tex

--- 2020-08-24 appendix.tex
@@
 A Gaussian process is specified by its mean function and convariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(m(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
-We consider the case where the observations are noisy, i.e. $\{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \sigma_n^2)$. 
-The Gaussian noise is independent identically distributed, and the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\sigma_{n}^{2} \delta_{p q}$,
+We consider the case where the observations are noisy, i.e. $\{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$. 
+The Gaussian noise is independent identically distributed, and the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
 where $\delta_{pq}$ is a Kronecker delta which is one iff $p = q$ and zero otherwise.
-It is equivalent to add a diagonal matrix $\sigma_n^2 I$ on the kernel matrix evaluated on the training points.
+It is equivalent to add a diagonal matrix $\alpha^2 I$ on the kernel matrix evaluated on the training points.
 
@@
 \begin{align}
-    \left[\begin{array}{l}\mathbf{y} \\ \mathbf{f}_{*}\end{array}\right] \sim \mathcal{N}\left(\mathbf{0},\left[\begin{array}{cc}K(X, X)+\sigma_{n}^{2} I & K\left(X, X_{*}\right) \\ K\left(X_{*}, X\right) & K\left(X_{*}, X_{*}\right)\end{array}\right]\right)
+    \left[\begin{array}{l}\mathbf{y} \\ \mathbf{f}_{*}\end{array}\right] \sim \mathcal{N}\left(\mathbf{0},\left[\begin{array}{cc}K(X, X)+\alpha^{2} I & K\left(X, X_{*}\right) \\ K\left(X_{*}, X\right) & K\left(X_{*}, X_{*}\right)\end{array}\right]\right)
 \end{align}
@@
 \begin{align}
-   \overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\sigma_{n}^{2} I\right]^{-1} \mathbf{y} \\
+   \overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} \mathbf{y} \\
    \label{Eq: predicted variance}
-   \operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\sigma_{n}^{2} I\right]^{-1} K\left(X, X_{*}\right) 
+   \operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} K\left(X, X_{*}\right) 
 \end{align}
-For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\sigma_n^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq. (\ref{Eq: predicted variance}).
+For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq. (\ref{Eq: predicted variance}).
 % We now introduce the marginal likelihood (or evidence) $p(\mathbf{y}|X)$, is which the integral of the likelihood times the prior 
@@
 
+\section{Design Pipeline}
+
+The second round design is based on the first round result, where each sequence has 6 replicates with TIR labels.
+We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively (\textcolor{red}{@Maciej, would you like to describe how raw TIR label is calculated?}).
+After normalisation, each replicate has zero mean and unit variance.
+
+For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
+We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
+We model the covariance matrix using the weighted degree kernel with shift. 
+We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
+The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation.
+We choose $l=6, s= 1, \alpha = 2$ for the second round design.
+
+For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
+The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
+We recommend 90 sequences from the design space.
+
 \section{Intuition behind UCB and visualisation}

### 2020-08-26 appendix.tex
update batch ucb

@@
 Therefore we consider recommending sequences in batch and use Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
+With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
+\begin{align}
+    \mathrm{fb}[t]=\left\{\begin{array}{cl}
+    0 & : t \in\{1, \ldots, B\} \\
+    B & : t \in\{B+1, \ldots, 2 B\} \\
+    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
+    & \vdots
+    \end{array}\right.
+\end{align}
+
 A key property of Gaussian Process regression is that the predictive variance in Eq.
@@
 \begin{align}
-    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right).
 \end{align}
-where $fb[t]$ ...
+And only update $y_{t^{\prime}}=f\left(\boldsymbol{x}_{t^{\prime}}\right)+\varepsilon_{t^{\prime}} \text { for } t^{\prime} \in\{\mathrm{fb}[t]+1, \ldots, \mathrm{fb}[t+1]\}$ at the end of each batch ($\mathrm{fb}[t]<\mathrm{fb}[t+1]$).
+This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right]$, while the posterior variance decreases.
+
+\section{Design Pipeline}
+
+The second round design is based on the first round result, where each sequence has 6 replicates with TIR labels.
+We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively (\textcolor{red}{@Maciej, would you like to describe how raw TIR label is calculated?}).
+After normalisation, each replicate has zero mean and unit variance.
+
+For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
+We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
+We model the covariance matrix using the weighted degree kernel with shift. 
+We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
+The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation.
+We choose $l=6, s= 1, \alpha = 2$ for the second round design.
+
+For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
+The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
+We recommend 90 sequences from the design space.
 

### 2020-09-01 codes/regression.py
add future work

@@ -139,6 +139,7 @@ class GPR_Predictor():
         """
         use_samples_for_train = True
         if use_samples_for_train:
+            # self.df['index'] = self.df.index
             train_df = pd.melt(self.df.loc[self.train_idx], id_vars=['RBS', 'RBS6', 'AVERAGE', 'STD', 'Group'], 
                                value_vars=['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5','Rep6'])
             train_df = train_df.dropna(subset=['RBS', 'AVERAGE', 'value'])
@@ -212,13 +213,25 @@ class GPR_Predictor():
         #     self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(l_list = self.l_list, features = self.features, test_size = self.test_size,), alpha = self.alpha)
         elif self.kernel_name == 'RBF':
             self.gp_reg = GaussianProcessRegressor(kernel = self.kernel(length_scale = 1), alpha = self.alpha, n_restarts_optimizer = 5)
-
+            
         print('gp_reg fit')
         self.gp_reg.fit(X_train,y_train_sample)
-        print('gp_reg pred')
+        print('gp_reg pred') 
+
         y_train_pred_mean, y_train_pred_std = self.gp_reg.predict(X_train, return_std=True)
         # print('regression train pred mean ', y_train_pred_mean)
         y_test_pred_mean, y_test_pred_std = self.gp_reg.predict(X_test, return_std=True)
+
+        # try_hete_gp = True
+        # if try_hete_gp:
+        #     import GPy
+        #     print(np.asarray(self.train_df['index']))
+        #     self.gp_reg = GPy.models.GPHeteroscedasticRegression(X_train,y_train_sample.reshape(-1,1), Y_metadata={'output_index': np.asarray(self.train_df['index'])})
+        #     y_train_pred_mean, y_train_pred_std = self.gp_reg.predict(X_train)
+        #     # print('regression train pred mean ', y_train_pred_mean)
+        #     y_test_pred_mean, y_test_pred_std = self.gp_reg.predict(X_test)
+        #     y_train_pred_std = np.sqrt(y_train_pred_std)
+        #     y_test_pred_std = np.sqrt(y_test_pred_std)
 
         self.train_df['pred mean'] = y_train_pred_mean
         self.test_df['pred mean'] = y_test_pred_mean

### 2020-09-02 methods.tex
Updated methods section.

@@
-\section{Method}
-This will be about the DNA assembly.\\
-\subsection{SynBio methods}
-This will be very specifically talking about the plasmid.
+\section{Methods}
 
-\subsection{Algorithms}
+\subsection{Plasmid Design}
 
-\subsubsection{ Position probability matrix based sampling}
+We have used the pBbB6c-GFP plasmid for all our designs.
+This plasmid comes with GFP mut3b CDS inducible with addition of IPTG.
+The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
+Primers and the assembly strategy have been generated using the Teselagen BUILD software (Teselagen Biotechnology).
 
-\subsubsection{Machine learning experimental design}
+\subsection{PCR}
+PCR amplification of the cloning inserts was done using Q5® High-Fidelity 2X Master Mix (NEB, catalogue no.
+M0492L).
+20µL reactions were prepared by dispensing each of the 10µM reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
+A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 \(\mu\)M forward primer was prepared by and dispensed by hand.
+Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
+Then, samples were incubated at 37°C for 60 minutes, followed by a 20-minute heat inactivation step at 80°C.
+Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
+2µL of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
+ZAG-110-5000; ZAG-130-5000).
+ProSize Data Analysis Software (Agilent Technologies) was used to generated gel images from the sample chromatograms and sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
+
+\subsection{isothermal DNA Assembly}
+Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no.
+E2621L).
+Samples were incubated at 37°C for 60 minutes, followed by a 20-minute heat inactivation step at 80°C.
+Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10µL .
+Assemblies were run in the thermal cycler for 1 hour at 50°C, followed by an infinite hold step at 4°C.
+
+\subsection{\textit{E. coli} transformation}
+The DH5α cell line (Thermo Fisher Scientific, catalogue no.
+18265017) was made chemically competent using the Mix & Go \textit{E.
+coli} Transformation Kit & Buffer Set (Zymo Research, catalogue no.
+T3001).
+20µL of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80°C for later use.
+Plates of cells were thawed on a -20°C cold block before 3µL of the assembly product was added and mixed using the CyBio FeliX liquid handler.
+Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
+***) with 34µg/mL chloramphenicol (Sigma, catalogue no.
+***).
+Multiple dilutions of cells in LB were prepared and plated in parallel on Omnitrays.
+Plates were incubated overnight at 37°C.
+
+\subsection{Automated colony picking and culturing}
+A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates.
+Each selected colony was used to inoculate 1mL of selective medium in a 2mL square well 96 plate.
+They were then cultured overnight in 37°C with shaking (~300rpm).
+
+\subsection{Glycerol stock preparation}
+100µL of sterile 80\% (v/v) glycerol and 100µL of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
+They were then sealed with a 96-well silicon sealing mat and transferred to a -80°C freezer.
+
+\subsection{Cuture analysis}
+Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34µg/mL chloramphenicol with ~2µL of the glycerol stock in a 96 deep (2mL) round well plate.
+Cultures were incubated at 37°C with shaking (~300rpm) for ~17 hours.
+The following morning, 20µL of overnight culture was added to 980µL of fresh selection medium and precultures were grown at 37°C with shaking in 2mL round well 96 plate.
+After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500µM IPTG – one plate of 300µL cultures for flow cytometry analysis (induced with 1.5µL of 0.1M IPTG) and one plate of 200µL cultures for plate reader analysis (induced with 1.0µL of 0.1M IPTG).
+•	Cytation 5 acquisition and incubation/shaking settings
+•	CytoFLEX S acquisition settings
+
+\subsection{Machine learning experimental design}
 
 To automatically design the RBS sequences in batch using machine learning, we need to consider two parts: 
-1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence interval for the prediction. 
-2) Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence interval. 
-Such online learning approach provides the $\textit{exploitation-exploration balance}$. 
-We show the corresponding methods we use in the below.
+1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals for the said predictions. 
+2) Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals. 
+Such online learning approach provides the $\textit{exploitation-exploration balance}$ that we use to control our machine learning process.
 
-We consider our experimental design problem as the problem of sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence point, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+Specifically, since our goal is to find a RBS sequence with the largest possible TIR score after the total number of rounds $N$,  we consider our experimental design problem as one of sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values of each points in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume the noise is under Gaussian distribution with some unknown mean and variance).
-The noise influenced by the accuracy RBS calculator and other experimental interference (e.g. \textcolor{red}{To be added}). 
-Our goal is to pick RBS sequence with the largest possible TIR score after the total number of rounds $N$.
+$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume the noise is following Gaussian distribution with unknown mean and variance).
+This noise is influenced by the accuracy of the RBS predictor and other experimental interference (e.g.
+\textcolor{red}{To be added}).
 
@@
     \item \textit{Gaussian process regression (GPR)}.
-    A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach to regression which provide uncertainty measurements on predictions. 
+    A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach to regression which provides uncertainty measurements on predictions. 
     We model $f$ as a sample from a \textit{Gaussian process} $GP(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.\\
-    We choose to use \textit{spectrum kernel} \cite{leslie2001spectrum} to specify the kernel function of $GP$.  
+    We chose to use \textit{spectrum kernel} \cite{leslie2001spectrum} to specify the kernel function of $GP$.  
     The spectrum kernel is widely used for classifying protein sequences \cite{leslie2001spectrum, ben2008support}, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.
@@
 Denote the number of letters in the alphabet as $|\Sigma|$.
-$\Phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+$\Phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ onto a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
     %Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
@@
     \item \textit{Upper confidence bound (UCB)}.
-Bandit algorithms \cite{lattimore2018bandit} provide various approaches to sequentially design an where an agent adaptively chooses one or more options among several actions based on certain policies.
-We consider a type of bandit algorithms, the upper confidence bound algorithm, which is based on the \textit{optimism in the face of uncertainty}.
-The UCB algorithm, as the name suggested, basically selects RBS sequences with the maximum upper confidence bound, i.e.
+Bandit algorithms \cite{lattimore2018bandit} provide various approaches to sequential design where an agent adaptively chooses one or more options among several actions based on certain policies.
+In our work we used the Upper Confidence Bound version of that algorithm, which is based on the \textit{optimism in the face of uncertainty}.
+The UCB algorithm, as the name suggests, basically selects RBS sequences with the maximum upper confidence bound, i.e.
 $\operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_t(\mathbf{x}_i) + \beta_t \sigma_t(\mathbf{x}_i)\right)$,
-    where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
+    where $\beta_t$ is a hyperparameter balancing the exploitation and exploration, 
     $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$. 
@@
 
-Combine the two parts together, we apply \textit{Gaussian Process Upper Confidence Bound (GPUCB)} algorithm, which has been theoretically analysed by \textcite{srinivas2012information}.
+Combined the two parts together, we apply \textit{Gaussian Process Upper Confidence Bound (GPUCB)} algorithm, which has been theoretically analysed by \textcite{srinivas2012information}.
 We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}.

### 2020-09-02 methods.tex
Update methods.tex

@@
 \subsection{PCR}
-PCR amplification of the cloning inserts was done using Q5® High-Fidelity 2X Master Mix (NEB, catalogue no.
+PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, catalogue no.
 M0492L).
-20µL reactions were prepared by dispensing each of the 10µM reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
-A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 \(\mu\)M forward primer was prepared by and dispensed by hand.
+20 \(\mu\)L reactions were prepared by dispensing each of the 10 \(\mu\)M reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
+A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 forward primer was prepared by and dispensed by hand.
 Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
@@
 Samples were incubated at 37°C for 60 minutes, followed by a 20-minute heat inactivation step at 80°C.
-Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10µL .
+Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10\(\mu\)L .
 Assemblies were run in the thermal cycler for 1 hour at 50°C, followed by an infinite hold step at 4°C.
@@
 T3001).
-20µL of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80°C for later use.
+20\(\mu\)L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80°C for later use.
 Plates of cells were thawed on a -20°C cold block before 3µL of the assembly product was added and mixed using the CyBio FeliX liquid handler.
 Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
-***) with 34µg/mL chloramphenicol (Sigma, catalogue no.
+***) with 34\(\mu\)g/mL chloramphenicol (Sigma, catalogue no.
 ***).
@@
 \subsection{Cuture analysis}
-Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34µg/mL chloramphenicol with ~2µL of the glycerol stock in a 96 deep (2mL) round well plate.
+Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34\(\mu\)g/mL chloramphenicol with ~2\(\mu\)L of the glycerol stock in a 96 deep (2mL) round well plate.
 Cultures were incubated at 37°C with shaking (~300rpm) for ~17 hours.
-The following morning, 20µL of overnight culture was added to 980µL of fresh selection medium and precultures were grown at 37°C with shaking in 2mL round well 96 plate.
-After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500µM IPTG – one plate of 300µL cultures for flow cytometry analysis (induced with 1.5µL of 0.1M IPTG) and one plate of 200µL cultures for plate reader analysis (induced with 1.0µL of 0.1M IPTG).
+The following morning, 20\(\mu\)L of overnight culture was added to 980\(\mu\)L of fresh selection medium and precultures were grown at 37°C with shaking in 2mL round well 96 plate.
+After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500\(\mu\)M IPTG – one plate of 300\(\mu\)L cultures for flow cytometry analysis (induced with 1.5\(\mu\)L of 0.1M IPTG) and one plate of 200\(\mu\)L cultures for plate reader analysis (induced with 1.0\(\mu\)L of 0.1M IPTG).
 •	Cytation 5 acquisition and incubation/shaking settings

### 2020-09-03 methods.tex
Updated methods.tex

@@
 The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
-Primers and the assembly strategy have been generated using the Teselagen BUILD software (Teselagen Biotechnology).
+Primers and the assembly strategy have been generated using the Teselagen DESIGN software (Teselagen Biotechnology).
 
@@
 Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
-Then, samples were incubated at 37°C for 60 minutes, followed by a 20-minute heat inactivation step at 80°C.
+Then, samples were incubated at 37$^{\circ}$C for 60 minutes, followed by a 20-minute heat inactivation step at 80$^{\circ}$C.
 Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
-2µL of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
+2\(\mu\)L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
 ZAG-110-5000; ZAG-130-5000).
@@
 
-\subsection{isothermal DNA Assembly}
+\subsection{Isothermal DNA Assembly}
 Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no.
 E2621L).
-Samples were incubated at 37°C for 60 minutes, followed by a 20-minute heat inactivation step at 80°C.
+Samples were incubated at 37$^{\circ}$C for 60 minutes, followed by a 20-minute heat inactivation step at 80$^{\circ}$C.
 Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10\(\mu\)L .
-Assemblies were run in the thermal cycler for 1 hour at 50°C, followed by an infinite hold step at 4°C.
+Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
 
@@
 T3001).
-20\(\mu\)L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80°C for later use.
-Plates of cells were thawed on a -20°C cold block before 3µL of the assembly product was added and mixed using the CyBio FeliX liquid handler.
+20\(\mu\)L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
+Plates of cells were thawed on a -20$^{\circ}$C cold block before 3µL of the assembly product was added and mixed using the CyBio FeliX liquid handler.
 Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
@@
 ***).
-Multiple dilutions of cells in LB were prepared and plated in parallel on Omnitrays.
-Plates were incubated overnight at 37°C.
+Multiple dilutions of cells in LB were prepared and plated in parallel on Omnitrays in 96 square grid.
+Plates were incubated overnight at 37$^{\circ}$C.
 
@@
 Each selected colony was used to inoculate 1mL of selective medium in a 2mL square well 96 plate.
-They were then cultured overnight in 37°C with shaking (~300rpm).
+They were then cultured overnight in 37$^{\circ}$C with shaking (~300rpm).
 
 \subsection{Glycerol stock preparation}
-100µL of sterile 80\% (v/v) glycerol and 100µL of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
-They were then sealed with a 96-well silicon sealing mat and transferred to a -80°C freezer.
+100\(\mu\)L of sterile 80\% (v/v) glycerol and 100\(\mu\)L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
+They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.
 
@@
 Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34\(\mu\)g/mL chloramphenicol with ~2\(\mu\)L of the glycerol stock in a 96 deep (2mL) round well plate.
-Cultures were incubated at 37°C with shaking (~300rpm) for ~17 hours.
-The following morning, 20\(\mu\)L of overnight culture was added to 980\(\mu\)L of fresh selection medium and precultures were grown at 37°C with shaking in 2mL round well 96 plate.
+Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours.
+The following morning, 20\(\mu\)L of overnight culture was added to 980\(\mu\)L of fresh selection medium and precultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate.
 After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500\(\mu\)M IPTG – one plate of 300\(\mu\)L cultures for flow cytometry analysis (induced with 1.5\(\mu\)L of 0.1M IPTG) and one plate of 200\(\mu\)L cultures for plate reader analysis (induced with 1.0\(\mu\)L of 0.1M IPTG).

### 2020-09-03 appendix.tex
move some ML content to method sec.

@@
 A \textit{Gaussian process} is a collection of random variables, any finite number of which have a joint Gaussian distribution. 
-We define mean function $m(\mathbf{x})$  and covariance function $k(\mathbf{x}, \mathbf{x}^\prime)$ of a real process $f(\mathbf{x})$ as
+We define mean function $\mu(\mathbf{x})$  and covariance function $k(\mathbf{x}, \mathbf{x}^\prime)$ of a real process $f(\mathbf{x})$ as
 \begin{align}
-    m(\mathbf{x}) &= \mathbb{E}[f(\mathbf{x})]\\
-    k(\mathbf{x}, \mathbf{x}^\prime) &= \mathbb{E}[(f(\mathbf{x}) - m(\mathbf{x}))(f(\mathbf{x}^\prime) - m(\mathbf{x}^\prime))].
+    \mu(\mathbf{x}) &= \mathbb{E}[f(\mathbf{x})]\\
+    k(\mathbf{x}, \mathbf{x}^\prime) &= \mathbb{E}[(f(\mathbf{x}) - \mu(\mathbf{x}))(f(\mathbf{x}^\prime) - \mu(\mathbf{x}^\prime))].
 \end{align}
 
-A Gaussian process is specified by its mean function and convariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(m(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
+A Gaussian process is specified by its mean function and convariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(\mu(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
 We consider the case where the observations are noisy, i.e. $\{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$. 
 The Gaussian noise is independent identically distributed, and the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
-where $\delta_{pq}$ is a Kronecker delta which is one iff $p = q$ and zero otherwise.
-It is equivalent to add a diagonal matrix $\alpha^2 I$ on the kernel matrix evaluated on the training points.
+where $\delta_{pq}$ is a Kronecker delta which is one if $p = q$ and zero otherwise.
+It is equivalent to a diagonal matrix $\alpha^2 I$ on the kernel matrix evaluated on the training points.
 
-For $n_\ast$ test points $X_\ast$, assume the prior over the functions values as a random Gaussian vector $\mathbf{f}_\ast \sim \mathcal{N}(\mathbf{0}, K(X_\ast, X_\ast))$.
+For $n_\ast$ test points $X_\ast$, we assume the prior over the functions values as a random Gaussian vector $\mathbf{f}_\ast \sim \mathcal{N}(\mathbf{0}, K(X_\ast, X_\ast))$.
 Then the joint distribution of the observed target values and the function values at the test points under the prior as 
@@
 % \end{align}
-The choice of covariance function is critical for the performance of Gaussian process regression, we show the choices of string kernel in next section.
 
@@
 
-We consider several string kernels below.
+The choice of covariance function is critical for the performance of Gaussian process regression, we show a number of different string kernels tested in this study below:
 
@@
     \begin{align}
-        k_\ell^{\text{Spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
+        k_\ell^{\text{Spec}}(X, X^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
     \end{align}
-     where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. Denote the number of letters in the alphabet as $|\Sigma|$. 
-    $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
-Let $X, X^\prime$ be two metrics which includes $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, the spectrum kernel over metrics is
+     where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
+    $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $X$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+    Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
     \begin{align}
-         K_\ell^{\text{Spec}}(\textbf{X}, \textbf{X}^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
+         K_\ell^{\text{Spec}}(X, X^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
     \end{align}
 
-    \item \textit{Sum of Spectrum Kernel,} considering weighted sum over different part of the string.
+    \item \textit{Sum of Spectrum Kernel,} considers weighted sum over different parts of the string.
 
-    \item \textit{Mixed Spectrum Kernel,} considering weighted sum over different substring length, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$,
+    \item \textit{Mixed Spectrum Kernel,} considers weighted sum over different substring length, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$,
         \begin{align}
@@
         \end{align}
-    \item \textit{Weighted Degree Kernel,} considering positional information. WD kernel counts the match of kmers at corresponding positions in two sequences.
-    For sequences with fixed length $L$ and weighted degree kernel consider substrings starting at each position $l = 1, ..., L$, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$, \\
+    \item \textit{Weighted Degree Kernel,} considers positional information. WD kernel counts the match of kmers at corresponding positions in two sequences.
+    For sequences with fixed length $L$ and weighted degree kernel considers substrings starting at each position $l = 1, ..., L$, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$, \\
     \begin{align}
@@
 the range of all features should be normalised so that each feature contributes approximately proportionately to the predictive model. 
-The kernel matrix is represented by the inner product of the underlying feature vectors, it need to be normalised before using in the downstream regression models. 
+The kernel matrix is represented by the inner product of the underlying feature vectors, it needs to be normalised before being used in the downstream regression models. 
 Up-scaling (down-scaling) features can be understood as down-scaling (up-scaling) regularizers such that they penalise the features less (more).
@@
 We denote $\phi(\mathbf{x}_i)$ as a column feature vector of sequence $\mathbf{x}_i$, 
-where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$. Assume there are totally $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
+where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$. Assume there is total of $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
 We illustrate centering and unit norm normalisation below.
@@
     \item Centering. 
-    Define the mean vector $\bar{\Phi}(X) = \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s) \in \mathbb{R}^d$, the centered feature vector $\phi^C(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    Defining the mean vector as $\bar{\Phi}(X) = \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s) \in \mathbb{R}^d$, the centered feature vector $\phi^C(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
     \begin{align}
@@
     \item Unit Norm.
-Define the ($l_2$) norm of a feature vector $||\phi(\mathbf{x})|| = \sqrt{\sum_{m = 1}^d \phi_d(\mathbf{x})^2} = \sqrt{k(\mathbf{x}, \mathbf{x})} \in \mathbb{R}^+$, then the unit norm feature vector $\phi^{UN}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+Define the ($l_2$) norm of a feature vector as $||\phi(\mathbf{x})|| = \sqrt{\sum_{m = 1}^d \phi_d(\mathbf{x})^2} = \sqrt{k(\mathbf{x}, \mathbf{x})} \in \mathbb{R}^+$, then the unit norm feature vector $\phi^{UN}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
     \begin{align}
@@
 \end{itemize}
-For the Gaussian Process regression, we make of use of two kernel matrices: the kernel function between the training data itself, i.e. $K(X_{train}, X_{train})$; 
+For the Gaussian Process regression, we make of use of two kernel matrices: the kernel function between the training data itself, i.e. $K(X_{train}, X_{train})$; and
 the kernel function taking the training data and testing data as inputs, i.e. $K(X_{test}, X_{train})$. 
@@
 %The second one (between train and test) is a little bit tricky. The different is not only that $n \neq n'$. 
-We will state two ways of normalise those two kind of matrices:
+We will state two ways of normalisation those two kind of matrices:
 \begin{itemize}
     \item Normalise training and testing data separately.
-    This approach is preferred for most of the machine learning algorithms since it follows the rule we have no information about testing data while training.
+    This approach is preferred for most of the machine learning algorithms since it follows the rule that we have no information about testing data while training.
     Then for centering, one should subtract the mean vector over the training data for both kinds of matrices.
-    For unit norm normalisation, when one calculate $K^{UN}(X_{test}, X_{train})$, the two terms inside of square root: $k(\mathbf{x}_i, \mathbf{x}_i)$ is taken from $K(X_{test}, X_{test})[i,i]$, and $k(\mathbf{x}_j, \mathbf{x}_j)$ is taken from $K(X_{train}, X_{train})[j,j]$.
+    For unit norm normalisation, when one calculates $K^{UN}(X_{test}, X_{train})$, the two terms inside of square root: $k(\mathbf{x}_i, \mathbf{x}_i)$ is taken from $K(X_{test}, X_{test})[i,i]$, and $k(\mathbf{x}_j, \mathbf{x}_j)$ is taken from $K(X_{train}, X_{train})[j,j]$.
 
     \item Normalise training and testing data together, i.e. normalise $K(X_{train+test}, X_{train+test})$, then extra the parts we need from the normalised matrix. 
-    This approach is suitable for the case one already know the whole testing features. 
+    This approach is suitable in a case where one already knows the whole of testing features. 
     For centering, one should subtract the mean vector over the whole matrix $\Phi(X_{train+test})$. 
-    The unit norm normalisation is the same as the previous case. 
+    The unit norm normalisation is the same as in the previous case. 
 \end{itemize}
@@
 
-Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
-Therefore we consider recommending sequences in batch and use Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
+Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label after each prediction.
+Therefore we consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
 With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 

### 2020-09-03 methods.tex
move some ML content to method sec.

@@
 
-To automatically design the RBS sequences in batch using machine learning, we need to consider two parts: 
-1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals for the said predictions. 
+To automatically design the RBS sequences in batch using machine learning, we consider two parts: 
+1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals. 
 2) Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals. 
@@
 
-Specifically, since our goal is to find a RBS sequence with the largest possible TIR score after the total number of rounds $N$,  we consider our experimental design problem as one of sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+With the goal of finding RBS sequences with the largest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values of each points in the selected set $\mathcal{S}_t$, i.e.
@@
 This noise is influenced by the accuracy of the RBS predictor and other experimental interference (e.g.
-\textcolor{red}{To be added}).
+time, temperature, operator, etc.).
 
-\begin{itemize}
-    \item \textit{Gaussian process regression (GPR)}.
-    A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach to regression which provides uncertainty measurements on predictions. 
-    We model $f$ as a sample from a \textit{Gaussian process} $GP(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.\\
-    We chose to use \textit{spectrum kernel} \cite{leslie2001spectrum} to specify the kernel function of $GP$.  
-    The spectrum kernel is widely used for classifying protein sequences \cite{leslie2001spectrum, ben2008support}, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.
-More precisely, $k_\ell^{\text{spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\Phi_{\ell}^{\mathrm{spec}}(\mathbf{x}), \Phi_{\ell}^{\mathrm{spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle$, where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$.
-Denote the number of letters in the alphabet as $|\Sigma|$.
-$\Phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ onto a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+For regression model, we consider the \textit{Gaussian Process Regression (GPR)}.
+A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach which provides uncertainty measurements on predictions. 
+We model $f$ as a sample from a \textit{Gaussian process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
+Denote the training testing data as $X, X_{*}$, and the training label as $\mathbf{y}$.
+Then the posterior of the test points (i.e.
+predictive distributions) is given by the conditional distribution $\mathbf{f}_\ast | X, \mathbf{y}, X_\ast \sim \mathcal{N}(\bar{\mathbf{f}}_\ast, cov(\mathbf{f}_\ast))$, where
+\begin{align}
+\overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} \mathbf{y} \\
+\label{Eq: predicted variance in main paper}
+\operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} K\left(X, X_{*}\right) 
+\end{align}
+%For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq.
+(\ref{Eq: predicted variance in main paper}).
+
+To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} \cite{ratsch_rase_2005} to specify the kernel function of $GP$.  
+The weighted degree kernel with shift extends the spectrum kernel \cite{leslie2001spectrum, ben2008support} 
+by considering the positional information and allowing some positional shifts of matching substrings.
+The spectrum kernel and the weighted degree kernel with shifts are string kernels, which take two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.
+We define spectrum kernel firstly,
+\begin{align}
+        k_\ell^{\text{Spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
+    \end{align}
+ where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
+$\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
+    \begin{align}
+         K_\ell^{\text{Spec}}(X, X^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
+    \end{align}
+The weighted degree kernel with shift is defined upon the spectrum kernel, with flexibility of adjusting the substring length $d$, the start position $l$ and the shift length $s$.
+WDS kernel counts the match of kmers at corresponding positions in two sequences as shown in the following,
+\begin{align}
+        k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
+\end{align}
+where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
+sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
+range at position $l$. 
 %Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
@@
 
-    \item \textit{Upper confidence bound (UCB)}.
-Bandit algorithms \cite{lattimore2018bandit} provide various approaches to sequential design where an agent adaptively chooses one or more options among several actions based on certain policies.
+%  For recommendations, we consider the \textit{Upper Confidence Bound (UCB)} type algorithms. 
+%  As one popular type of the bandit algorithms \cite{lattimore2018bandit}, the UCB type of algorithms are based on the \textit{optimism in the face of uncertainty},
+ %provide various approaches to sequential design where an agent adaptively chooses one or more options among several actions based on certain policies.
 In our work we used the Upper Confidence Bound version of that algorithm, which is based on the \textit{optimism in the face of uncertainty}.
-The UCB algorithm, as the name suggests, basically selects RBS sequences with the maximum upper confidence bound, i.e.
+The UCB algorithm, as the name suggests,
+%  which basically select RBS sequences with the maximum upper confidence bound constructed by the sum of the predicted mean and $n$ standard deviation ($n > 0$), i.e.
 $\operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_t(\mathbf{x}_i) + \beta_t \sigma_t(\mathbf{x}_i)\right)$,
-    where $\beta_t$ is a hyperparameter balancing the exploitation and exploration, 
+%     where $\beta_t$ is a hyperparameter balancing the exploitation and exploration, 
+%     $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
+
+For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
+%which is based on the \textit{optimism in the face of uncertainty}, 
+selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
+\begin{align}
+\label{Eq: GPUCB}
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
+\end{align}
+where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
 $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
-\end{itemize}
 
-Combined the two parts together, we apply \textit{Gaussian Process Upper Confidence Bound (GPUCB)} algorithm, which has been theoretically analysed by \textcite{srinivas2012information}.
-We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}.
+Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
+Therefore we consider recommending sequences in batch and use Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
+With batches of size $B$, the feedback mapping is then defined as $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
+\begin{align}
+    \mathrm{fb}[t]=\left\{\begin{array}{cl}
+    0 & : t \in\{1, \ldots, B\} \\
+    B & : t \in\{B+1, \ldots, 2 B\} \\
+    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
+    & \vdots
+    \end{array}\right.
+\end{align}
+
+A key property of Gaussian Process regression is that the predictive variance in Eq.
+(\ref{Eq: predicted variance in main paper}) only depends on observed points (i.e.
+features), but not on the labels of those observed points.
+So one can compute the posterior variance without actually observing the labels. 
+The GP-BUCB policy is to select sequences that
+\begin{align}
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right).
+\end{align}
+And only update $y_{t^{\prime}}=f\left(\boldsymbol{x}_{t^{\prime}}\right)+\varepsilon_{t^{\prime}} \text { for } t^{\prime} \in\{\mathrm{fb}[t]+1, \ldots, \mathrm{fb}[t+1]\}$ at the end of each batch ($\mathrm{fb}[t]<\mathrm{fb}[t+1]$).
+This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right]$, while the posterior variance decreases.
 
@@
     \centering
-    \includegraphics[scale=0.35]{plots/flowchart.pdf}
+    \includegraphics[scale=0.7]{plots/flowchart.pdf}
     \caption{Flowchart of machine learning based experimental design.}

### 2020-09-03 appendix.tex
Updated appendix.

@@
 (\ref{Eq: predicted variance}) only depends on observed points (i.e.
-features), but not on the labels of those observed points.
+features), but not on the labels of these observed points.
 So one can compute the posterior variance without actually observing the labels. 
@@
 
-The second round design is based on the first round result, where each sequence has 6 replicates with TIR labels.
-We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively (\textcolor{red}{@Maciej, would you like to describe how raw TIR label is calculated?}).
+The n + 1 round design is based on the nth round result, where each sequence has 6 replicates with TIR labels.
+We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively, where TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
 After normalisation, each replicate has zero mean and unit variance.

### 2020-09-03 methods.tex
Updated appendix.

@@
 20\(\mu\)L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
-Plates of cells were thawed on a -20$^{\circ}$C cold block before 3µL of the assembly product was added and mixed using the CyBio FeliX liquid handler.
+Plates of cells were thawed on a -20$^{\circ}$C cold block before 3\(\mu\)L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
 Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.

### 2020-09-04 sub/SynBio_ML_Writing/methods.tex

--- 2020-09-03 methods.tex
@@
 M0492L).
-20 \(\mu\)L reactions were prepared by dispensing each of the 10 \(\mu\)M reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
+20 $\mu$L reactions were prepared by dispensing each of the 10 $\mu$M reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
 A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 forward primer was prepared by and dispensed by hand.
@@
 Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
-2\(\mu\)L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
+2$\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
 ZAG-110-5000; ZAG-130-5000).
@@
 The DH5α cell line (Thermo Fisher Scientific, catalogue no.
-18265017) was made chemically competent using the Mix & Go \textit{E.
-coli} Transformation Kit & Buffer Set (Zymo Research, catalogue no.
+18265017) was made chemically competent using the Mix $\&$ Go \textit{E.
+coli} Transformation Kit $\&$ Buffer Set (Zymo Research, catalogue no.
 T3001).
-20\(\mu\)L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
-Plates of cells were thawed on a -20$^{\circ}$C cold block before 3µL of the assembly product was added and mixed using the CyBio FeliX liquid handler.
+20$\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
+Plates of cells were thawed on a -20$^{\circ}$C cold block before 3$\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
 Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
-***) with 34\(\mu\)g/mL chloramphenicol (Sigma, catalogue no.
+***) with 34$\mu$g/mL chloramphenicol (Sigma, catalogue no.
 ***).
@@
 \subsection{Glycerol stock preparation}
-100\(\mu\)L of sterile 80\% (v/v) glycerol and 100\(\mu\)L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
+100$\mu$L of sterile 80\% (v/v) glycerol and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
 They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.
@@
 \subsection{Cuture analysis}
-Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34\(\mu\)g/mL chloramphenicol with ~2\(\mu\)L of the glycerol stock in a 96 deep (2mL) round well plate.
+Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2$\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
 Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours.
-The following morning, 20\(\mu\)L of overnight culture was added to 980\(\mu\)L of fresh selection medium and precultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate.
-After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500\(\mu\)M IPTG – one plate of 300\(\mu\)L cultures for flow cytometry analysis (induced with 1.5\(\mu\)L of 0.1M IPTG) and one plate of 200\(\mu\)L cultures for plate reader analysis (induced with 1.0\(\mu\)L of 0.1M IPTG).
+The following morning, 20$\mu$L of overnight culture was added to 980$\mu$L of fresh selection medium and precultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate.
+After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500$\mu$M IPTG – one plate of 300$\mu$L cultures for flow cytometry analysis (induced with 1.5$\mu$L of 0.1M IPTG) and one plate of 200$\mu$L cultures for plate reader analysis (induced with 1.0$\mu$L of 0.1M IPTG).
 •	Cytation 5 acquisition and incubation/shaking settings

### 2020-09-09 codes/regression.py
generate gpr plot

@@ -197,9 +197,10 @@ class GPR_Predictor():
 
         if self.kernel_name == 'WD_Kernel_Shift':
             print('create kernel instance')
-            kernel_instance = self.kernel(l = self.l, features = self.features, 
+            self.wd_kernel_instance = self.kernel(l = self.l, features = self.features, 
                                          n_train=X_train.shape[0], n_test=X_test.shape[0],
-                                         s = self.s, sigma_0=self.sigma_0) \
+                                         s = self.s, sigma_0=self.sigma_0)
+            kernel_instance = self.wd_kernel_instance \
                             + WhiteKernel(noise_level=1e-5, noise_level_bounds=(1e-5, 1e+5))
             # debug
             self.kernel_instance = kernel_instance

### 2020-09-11 main.tex
Squashed 'sub/SynBio_ML_Writing/' changes from f3cb5d2..19f94fb
19f94fb update ML methods with plots
1d36fa7 fix bugs
9c1aae6 Updated appendix.
60a3ebd move some ML content to method sec.
33d61d0 Updated flowchart
45c8931 Updated methods.tex
c3f267f Update methods.tex
6b82ce7 Updated methods section.
2f52ce4 update batch ucb

git-subtree-dir: sub/SynBio_ML_Writing
git-subtree-split: 19f94fb2cf879238c8fcfd1c1d76d8eb86ce29d7

@@
 
-Our overall experimental goal was to optimise the translation initiation rate by identifying the set of RBS sequences with top TIR scores with minimising the number of DBTL cycle turns that we had to do.
-We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (round 1) or with RBS sequences recommended by the algorithm (subsequent rounds).
+Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
+We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (designated round zero since only preexisting, literature data is used for its design) or with RBS sequences recommended by the algorithm (subsequent rounds).
 The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
-After constructing the plasmids harbouring the new devices are tested in microplate reader and flow cytometer.
-The results are then fed back to the algorithm for it to recommend the next round of designs.\\
-
-\textcolor{red}{DBTL cartoon}
+After constructing, the plasmids harbouring the new genetic devices are tested in microplate reader and flow cytometer.
+The results are then fed back to the algorithm for it to recommend the next round of designs.\\ \textcolor{red}{Figure X} shows the DBTL cycle with the machine learning process emphasised.
+Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase.
+In LEARN phase we use the Gaussian Process algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
+Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the learn phase, which we find a more natural starting point for a machine learning centred workflow.\\
 
@@
 In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space.
-In our case we use the Upper Confidence Bound \textcolor{red}{flavour} of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR).
-Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation.vExploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
+In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR).
+Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation.
+Exploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
 The two approaches can be controlled with the \textbeta\enspace parameter.

### 2020-09-11 methods.tex
Squashed 'sub/SynBio_ML_Writing/' changes from f3cb5d2..19f94fb
19f94fb update ML methods with plots
1d36fa7 fix bugs
9c1aae6 Updated appendix.
60a3ebd move some ML content to method sec.
33d61d0 Updated flowchart
45c8931 Updated methods.tex
c3f267f Update methods.tex
6b82ce7 Updated methods section.
2f52ce4 update batch ucb

git-subtree-dir: sub/SynBio_ML_Writing
git-subtree-split: 19f94fb2cf879238c8fcfd1c1d76d8eb86ce29d7

--- 2020-09-03 methods.tex
@@
 
+We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}. 
 To automatically design the RBS sequences in batch using machine learning, we consider two parts: 
-1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals. 
-2) Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals. 
+1) \textbf{LEARN}: Design a regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals. 
+2) \textbf{DESIGN}: Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals. 
 Such online learning approach provides the $\textit{exploitation-exploration balance}$ that we use to control our machine learning process.
 
-With the goal of finding RBS sequences with the largest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.7]{plots/flowchart.pdf}
+    \caption{Flowchart of machine learning based experimental design.}
+    \label{fig: flowchart of machine learning based experimental design.}
+\end{figure}
+
+\subsubsection{LEARN: Gaussian Process Regression with String Kernel}
+
+To find RBS sequences with the largest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values of each points in the selected set $\mathcal{S}_t$, i.e.
@@
 We model $f$ as a sample from a \textit{Gaussian process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
-Denote the training testing data as $X, X_{*}$, and the training label as $\mathbf{y}$.
-Then the posterior of the test points (i.e.
-predictive distributions) is given by the conditional distribution $\mathbf{f}_\ast | X, \mathbf{y}, X_\ast \sim \mathcal{N}(\bar{\mathbf{f}}_\ast, cov(\mathbf{f}_\ast))$, where
-\begin{align}
-\overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} \mathbf{y} \\
-\label{Eq: predicted variance in main paper}
-\operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} K\left(X, X_{*}\right) 
-\end{align}
-%For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq.
-(\ref{Eq: predicted variance in main paper}).
+As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, GPR can predict both the mean function (solid blue line) and the confidence interval (shaded area). 
+We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0. 
+As we can see, the true function is in the confidence interval when there are data points are observed.
 
-To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} \cite{ratsch_rase_2005} to specify the kernel function of $GP$.  
-The weighted degree kernel with shift extends the spectrum kernel \cite{leslie2001spectrum, ben2008support} 
-by considering the positional information and allowing some positional shifts of matching substrings.
-The spectrum kernel and the weighted degree kernel with shifts are string kernels, which take two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.
-We define spectrum kernel firstly,
-\begin{align}
-        k_\ell^{\text{Spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
-    \end{align}
- where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
-$\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
-Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
-    \begin{align}
-         K_\ell^{\text{Spec}}(X, X^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
-    \end{align}
-The weighted degree kernel with shift is defined upon the spectrum kernel, with flexibility of adjusting the substring length $d$, the start position $l$ and the shift length $s$.
-WDS kernel counts the match of kmers at corresponding positions in two sequences as shown in the following,
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.7]{plots/Prediction with Initial Observations.pdf}
+    \caption{Gaussian Process Regression Example.
+This plot shows the GPR prediction with confidence interval based on 4 initial observations.
+The confidence interval are shown in predicted mean $\pm$ 1.96 standard deviation.}
+    \label{fig: Gaussian Process Regression Example.}
+\end{figure}
+
+The choice of covariance functions is critical for accurate predictions, which controls smoothness and amplitude of the function we model.
+To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005} to specify the kernel function of $GP$.
+WDS is one type of string kernels, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
+DS kernel counts the match of substrings with different length (i.e. kmers) starting for different positions. 
+The WDS considers the positional information and allows positional shifts for matching substrings.
+Let $\mathbf{x}, \mathbf{x}^\prime$ be two RBS sequences. 
+And let $\ell$ be the maximum length of substrings taking into account, $L$ be the length of sequences, and $S(l)$ be the maximum shift length for substrings starting from position $l$.
 \begin{align}
         k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
-        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
-        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        %&= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        %\left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
+        = \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
         \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
 \end{align}
-where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
+where 
+$\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
 sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
 range at position $l$. 
-%Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
+$\mathbb{I}$(A) is the indicator function, which equals 1 if $A$ is true and 0 otherwise.
+
+$\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ basically counts the matches of substrings of length $d$ between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
+Similarly for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
+By having these two terms considering substrings of two sequences with starting positions differing $s$ characters, the WDS can measure the shifts positional information. 
+When $s = 0$, the kernel function counts the matches for no shift between sequences.
 
@@
 
+\subsubsection{DESIGN: Batch UCB}
+
 For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
@@
 where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation (SD) at round $t$ for the sequence $\mathbf{x}_i$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ as \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
+
+A key property of Gaussian Process regression is that the predictive variance only depends on observed points (i.e. features), but not on the labels of those observed points. 
+We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
+As we can see, the predicted confidence interval is the same for the two plots.
+
+\vspace{0.5cm}
+\begin{minipage}{\textwidth}
+  \begin{minipage}[b]{0.49\textwidth}
+    \centering
+    %\includegraphics[scale=0.4]{plots/Hist.pdf}
+    \includegraphics[scale=0.45]{plots/Recommend 1 Data Point (New Observation).pdf}
+     \captionsetup{type=figure}
+     \vspace{-0.5cm}
+    \caption{Recommend 1 Data Point (New Observation)
+    }
+    \label{fig: Recommend 1 Data Point (New Observation)}
+\end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[scale=0.45]{plots/Recommend 1 Data Point (Predicted Mean).pdf}
+     \captionsetup{type=figure}
+     \vspace{-0.5cm}
+    \caption{Recommend 1 Data Point (Predicted Mean)
+    }
+    \label{fig: Recommend 1 Data Point (Predicted Mean)}
+ \end{minipage}
+ \end{minipage}
+ \vspace{0.5cm}
 
 Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
-Therefore we consider recommending sequences in batch and use Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
-With batches of size $B$, the feedback mapping is then defined as $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
-\begin{align}
-    \mathrm{fb}[t]=\left\{\begin{array}{cl}
-    0 & : t \in\{1, \ldots, B\} \\
-    B & : t \in\{B+1, \ldots, 2 B\} \\
-    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
-    & \vdots
-    \end{array}\right.
-\end{align}
+Consider the task of recommending RBS sequences in a batch of size $n$.
+One naive approach is to train GPR on sequences with known TIR labels and recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig: Recommend 2 Data Points: Top 2 UCB.} ($n = 2$).
+This approach may end up recommending similar sequences with high predicted SD.
+However, since we assume similar sequences would have similar labels (e.g.
+by knowing $x=8$ we can gain information of $x=9$ as well), we prefer to not waste time and money on labelling sequences with high similarities in a batch.
 
-A key property of Gaussian Process regression is that the predictive variance in Eq.
-(\ref{Eq: predicted variance in main paper}) only depends on observed points (i.e.
-features), but not on the labels of those observed points.
-So one can compute the posterior variance without actually observing the labels. 
-The GP-BUCB policy is to select sequences that
-\begin{align}
-    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right).
-\end{align}
-And only update $y_{t^{\prime}}=f\left(\boldsymbol{x}_{t^{\prime}}\right)+\varepsilon_{t^{\prime}} \text { for } t^{\prime} \in\{\mathrm{fb}[t]+1, \ldots, \mathrm{fb}[t+1]\}$ at the end of each batch ($\mathrm{fb}[t]<\mathrm{fb}[t+1]$).
-This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right]$, while the posterior variance decreases.
+We can make use of the property that predicted SD does not depend on the labels to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
+That is, recommend sequences sequentially by assuming have seen the previous recommended points and updating the UCB score with the updated predicted SD. 
+Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
+The plot shows the predicted mean and variance after observing 4 data points. 
+2 data points ($x = 8, x = 0$) are recommended in batch. 
+After recommending the sequence with top UCB score ($x = 8$), the algorithm updates the predicted SD (note we do not update the predicted mean since it is influenced by labels) and the second sequence is recommended based on the new UCB score with the updated predicted SD.
+In this case, the predicted SD of sequences around $x = 8$ would decrease (not shown in the plot). 
+So instead of recommending $x = 9$, the BUCB algorithm recommends sequences from another area which potentially contains high labels ($x = 0$).
+By using BUCB, we rule out the sequences with high similarities in the same batch and increase the exploration efficiency.
 
-\begin{figure}[t]
+\vspace{0.5cm}
+\begin{minipage}{\textwidth}
+  \begin{minipage}[b]{0.49\textwidth}
     \centering
-    \includegraphics[scale=0.7]{plots/flowchart.pdf}
-    \caption{Flowchart of machine learning based experimental design.}
-    \label{fig: flowchart of machine learning based experimental design.}
-\end{figure}
+    %\includegraphics[scale=0.4]{plots/Hist.pdf}
+    \includegraphics[scale=0.45]{plots/Recommend 2 Data Points: Top 2 UCB.pdf}
+     \captionsetup{type=figure}
+     \vspace{-0.5cm}
+    \caption{Recommend 2 Data Points: Top 2 UCB
+    }
+    \label{fig: Recommend 2 Data Points: Top 2 UCB.}
+\end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.49\textwidth}
+    \centering
+    \includegraphics[scale=0.45]{plots/Recommend 2 Data Points: GP-BUCB.pdf}
+     \captionsetup{type=figure}
+     \vspace{-0.5cm}
+    \caption{Recommend 2 Data Points: GP-BUCB
+    }
+    \label{fig: Recommend 2 Data Points: GP-BUCB.}
+ \end{minipage}
+ \end{minipage}
+ \vspace{0.5cm}

### 2020-09-13 methods.tex
Squashed 'sub/SynBio_ML_Writing/' changes from 19f94fb..15a9eb6
15a9eb6 update new plots for batch rec

git-subtree-dir: sub/SynBio_ML_Writing
git-subtree-split: 15a9eb67d47e863f18b7085c0cc039c3b1ae8029

@@
 We model $f$ as a sample from a \textit{Gaussian process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
-As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, GPR can predict both the mean function (solid blue line) and the confidence interval (shaded area). 
-We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0. 
-As we can see, the true function is in the confidence interval when there are data points are observed.
+%As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
+GPR can predict both the posterior mean and posterior variance. The posterior variance represents the level of uncertainty for the prediction. 
+% We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0. 
+% As we can see, the true function is in the confidence interval when there are data points are observed.
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.7]{plots/Prediction with Initial Observations.pdf}
-    \caption{Gaussian Process Regression Example.
+% \begin{figure}[t]
+%     \centering
+%     \includegraphics[scale=0.7]{plots/Prediction with Initial Observations.pdf}
+%     \caption{Gaussian Process Regression Example.
 This plot shows the GPR prediction with confidence interval based on 4 initial observations.
 The confidence interval are shown in predicted mean $\pm$ 1.96 standard deviation.}
-    \label{fig: Gaussian Process Regression Example.}
-\end{figure}
+%     \label{fig: Gaussian Process Regression Example.}
+% \end{figure}
 
@@
 
+Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
+Consider the task of recommending RBS sequences in a batch of size $n$. 
+One naive approach is to train GPR on sequences with known TIR labels and recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig: Top UCB rec.} ($n = 2$).
+This approach may end up recommending similar sequences in neighbourhood (e.g. $x = 2, x =2.5$ in this example). 
+However, since we assume similar sequences would have similar labels (e.g.
+by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in a batch.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.6]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}
+    \caption{Batch Recommendation: Top UCB scores.
+We consider the batch size 2, with 5 initial observations.
+The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e.
+{-2, -1.5, -1, ..., 9.5, 10}.
+The recommendations are 2 data points with top UCB scores, constructed with GPR predictions.
+The confidence interval are shown in predicted mean $\pm$ 1.96 standard deviation.
+}
+    \label{fig: Top UCB rec.}
+\end{figure}
+
 A key property of Gaussian Process regression is that the predictive variance only depends on observed points (i.e. features), but not on the labels of those observed points. 
-We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
-As we can see, the predicted confidence interval is the same for the two plots.
+% We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
+% As we can see, the predicted confidence interval is the same for the two plots. 
+We can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
+That is, recommend sequences sequentially by assuming have seen the previous recommended points and updating the UCB score with the updated predicted SD. 
+% Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
+% The plot shows the predicted mean and variance after observing 4 data points. 
+Consider the same setting in Figure \ref{fig: Top UCB rec.} as an toy example, GP-BUCB firstly recommend the data point with maximum UCB scores based on the predictions over initial 5 observations. 
+As shown in Figure \ref{fig: GP-BUCB:_Recommend_1st_Data_Point.}.  
+Then we add the recommended data point ($x = 2$) into training data with the predicted mean of that point as label (note it is not the true label, i.e.
+observation), and update the predicted variance then update the UCB scores.
+We recommend the second data point based on the new UCB scores.   
+As we can see, since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in another area and potentially having high labels ($x = 7.5$).
+By using GP-BUCB, we rule out the sequences with high similarities in the same batch and increase the exploration efficiency. 
+2 data points ($x = 8, x = 0$) are recommended in batch.
 
@@
     %\includegraphics[scale=0.4]{plots/Hist.pdf}
-    \includegraphics[scale=0.45]{plots/Recommend 1 Data Point (New Observation).pdf}
+    \includegraphics[scale=0.45]{plots/GP-BUCB:_Recommend_1st_Data_Point.pdf}
      \captionsetup{type=figure}
      \vspace{-0.5cm}
-    \caption{Recommend 1 Data Point (New Observation)
-    }
-    \label{fig: Recommend 1 Data Point (New Observation)}
+    \caption{GP-BUCB: Recommend 1st Data Point.}
+    \label{fig: GP-BUCB:_Recommend_1st_Data_Point.}
 \end{minipage}
@@
     \centering
-    \includegraphics[scale=0.45]{plots/Recommend 1 Data Point (Predicted Mean).pdf}
+    \includegraphics[scale=0.45]{plots/GP-BUCB:_Recommend_2nd_Data_Point.pdf}
      \captionsetup{type=figure}
      \vspace{-0.5cm}
-    \caption{Recommend 1 Data Point (Predicted Mean)
+    \caption{GP-BUCB: Recommend 2nd Data Point.
     }
-    \label{fig: Recommend 1 Data Point (Predicted Mean)}
+    \label{fig: GP-BUCB:_Recommend_2nd_Data_Point.}
  \end{minipage}
@@
  \vspace{0.5cm}
-
-Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
-Consider the task of recommending RBS sequences in a batch of size $n$.
-One naive approach is to train GPR on sequences with known TIR labels and recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig: Recommend 2 Data Points: Top 2 UCB.} ($n = 2$).
-This approach may end up recommending similar sequences with high predicted SD.
-However, since we assume similar sequences would have similar labels (e.g.
-by knowing $x=8$ we can gain information of $x=9$ as well), we prefer to not waste time and money on labelling sequences with high similarities in a batch.
-
-We can make use of the property that predicted SD does not depend on the labels to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
-That is, recommend sequences sequentially by assuming have seen the previous recommended points and updating the UCB score with the updated predicted SD. 
-Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
-The plot shows the predicted mean and variance after observing 4 data points. 
-2 data points ($x = 8, x = 0$) are recommended in batch. 
-After recommending the sequence with top UCB score ($x = 8$), the algorithm updates the predicted SD (note we do not update the predicted mean since it is influenced by labels) and the second sequence is recommended based on the new UCB score with the updated predicted SD.
-In this case, the predicted SD of sequences around $x = 8$ would decrease (not shown in the plot). 
-So instead of recommending $x = 9$, the BUCB algorithm recommends sequences from another area which potentially contains high labels ($x = 0$).
-By using BUCB, we rule out the sequences with high similarities in the same batch and increase the exploration efficiency.
-
-\vspace{0.5cm}
-\begin{minipage}{\textwidth}
-  \begin{minipage}[b]{0.49\textwidth}
-    \centering
-    %\includegraphics[scale=0.4]{plots/Hist.pdf}
-    \includegraphics[scale=0.45]{plots/Recommend 2 Data Points: Top 2 UCB.pdf}
-     \captionsetup{type=figure}
-     \vspace{-0.5cm}
-    \caption{Recommend 2 Data Points: Top 2 UCB
-    }
-    \label{fig: Recommend 2 Data Points: Top 2 UCB.}
-\end{minipage}
-  \hfill
-  \begin{minipage}[b]{0.49\textwidth}
-    \centering
-    \includegraphics[scale=0.45]{plots/Recommend 2 Data Points: GP-BUCB.pdf}
-     \captionsetup{type=figure}
-     \vspace{-0.5cm}
-    \caption{Recommend 2 Data Points: GP-BUCB
-    }
-    \label{fig: Recommend 2 Data Points: GP-BUCB.}
- \end{minipage}
- \end{minipage}
- \vspace{0.5cm}

### 2020-09-15 methods.tex
update WDS kernel description; add exp-exp explanation.

@@
 
-\subsection{Cuture analysis}
+\subsection{Culture analysis}
 Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2$\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
@@
 We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}. 
-To automatically design the RBS sequences in batch using machine learning, we consider two parts: 
-1) \textbf{LEARN}: Design a regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals. 
-2) \textbf{DESIGN}: Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals. 
-Such online learning approach provides the $\textit{exploitation-exploration balance}$ that we use to control our machine learning process.
+To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
+1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the respective confidence intervals. 
+2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
 
@@
 
-\subsubsection{LEARN: Gaussian Process Regression with String Kernel}
+\subsubsection{LEARN: Gaussian Process Regression with Weighted Degree Kernel}
 
-To find RBS sequences with the largest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
-In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values of each points in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume the noise is following Gaussian distribution with unknown mean and variance).
-This noise is influenced by the accuracy of the RBS predictor and other experimental interference (e.g.
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values at each point in the selected set $\mathcal{S}_t$, i.e.
+$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume that the noise is following Gaussian distribution with unknown mean and variance).
+This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g.
 time, temperature, operator, etc.).
 
-For regression model, we consider the \textit{Gaussian Process Regression (GPR)}.
+For regression model, we have used the \textit{Gaussian Process Regression (GPR)}.
 A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach which provides uncertainty measurements on predictions. 
@@
 
-The choice of covariance functions is critical for accurate predictions, which controls smoothness and amplitude of the function we model.
+The choice of covariance (kernel) function is critical for accurate predictions, since it controls smoothness and amplitude of the function we model.
 To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005} to specify the kernel function of $GP$.
-WDS is one type of string kernels, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
-DS kernel counts the match of substrings with different length (i.e. kmers) starting for different positions. 
-The WDS considers the positional information and allows positional shifts for matching substrings.
-Let $\mathbf{x}, \mathbf{x}^\prime$ be two RBS sequences. 
-And let $\ell$ be the maximum length of substrings taking into account, $L$ be the length of sequences, and $S(l)$ be the maximum shift length for substrings starting from position $l$.
+WDS is a type of a string kernel, which takes two sequences (strings) as inputs and outputs a scalar value which represents the similarities between the two sequences.  
+WDS kernel counts the matches of substrings of a certain length (i.e. kmers).
+The maximum substring length is specified by $\ell$. 
+The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
+Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
+Let $\mathbb{I}$(A) is the indicator function, which equals 1 if $A$ is true and 0 otherwise. 
+Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ counts the matches of substrings of length $d$ between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
+This is similarly done for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
+By having these two terms considering substrings of two sequences with starting positions differing by $s$ characters, the WDS can measure shifted positional information. 
+When $s = 0$, the kernel function counts the matches with no shift between sequences. 
+Let $\mathbf{x}, \mathbf{x}^\prime$ be two RBS sequences with length $L$, the WDS kernel is defined as
 \begin{align}
@@
 where 
-$\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
+$\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting parameter over the position in the
 sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
 range at position $l$.
-$\mathbb{I}$(A) is the indicator function, which equals 1 if $A$ is true and 0 otherwise.
-
-$\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ basically counts the matches of substrings of length $d$ between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
-Similarly for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
-By having these two terms considering substrings of two sequences with starting positions differing $s$ characters, the WDS can measure the shifts positional information. 
-When $s = 0$, the kernel function counts the matches for no shift between sequences.
 
@@
 
-For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
-%which is based on the \textit{optimism in the face of uncertainty}, 
-selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
+For recommendations of RBS sequences that should be experimentally labels next, we have used the \textit{Upper Confidence Bound (UCB)} algorithm.
+On the one hand, we want to exploit in terms of the design space where sequences are believed to have high labels (i.e. high predicted mean); 
+on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted SD).
+The UCB algorithm provides such $\textit{exploitation-exploration balance}$ by balancing the predicted mean and SD.
+More precisely, UCB algorithm selects RBS sequences with the maximum upper confidence bound at round $t$, i.e.
 \begin{align}
@@
 $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation (SD) at round $t$ for the sequence $\mathbf{x}_i$.
-We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ as \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
 
-Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
-Consider the task of recommending RBS sequences in a batch of size $n$. 
-One naive approach is to train GPR on sequences with known TIR labels and recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig: Top UCB rec.} ($n = 2$).
-This approach may end up recommending similar sequences in neighbourhood (e.g. $x = 2, x =2.5$ in this example). 
+Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e.
+one-by-one) and then wait for the label to be tested and used to improve the model.
+Instead, we can recommend RBS sequences in a batch of size $n$. 
+One naive approach is to train GP on sequences with known TIR labels and recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig: Top UCB rec.} ($n = 2$).
+However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
 However, since we assume similar sequences would have similar labels (e.g.
-by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in a batch.
+by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
 
@@
     \caption{Batch Recommendation: Top UCB scores.
-We consider the batch size 2, with 5 initial observations.
+We use the batch size of 2, with 5 initial observations.
 The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e.
 {-2, -1.5, -1, ..., 9.5, 10}.
-The recommendations are 2 data points with top UCB scores, constructed with GPR predictions.
-The confidence interval are shown in predicted mean $\pm$ 1.96 standard deviation.
+The recommendations are 2 data points with top UCB scores, constructed with GP predictions.
+The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
 }
@@
 
-A key property of Gaussian Process regression is that the predictive variance only depends on observed points (i.e. features), but not on the labels of those observed points. 
+A key property of Gaussian Process regression is that the predictive variance  depends only on observed points (i.e. features), but not on the labels of those observed points. 
 % We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
@@
 We can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
-That is, recommend sequences sequentially by assuming have seen the previous recommended points and updating the UCB score with the updated predicted SD. 
+That is, we can recommend sequences sequentially by updating the UCB score with the updated predicted SD with the the previously recommended points.
 % Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
 % The plot shows the predicted mean and variance after observing 4 data points. 
-Consider the same setting in Figure \ref{fig: Top UCB rec.} as an toy example, GP-BUCB firstly recommend the data point with maximum UCB scores based on the predictions over initial 5 observations. 
-As shown in Figure \ref{fig: GP-BUCB:_Recommend_1st_Data_Point.}.  
-Then we add the recommended data point ($x = 2$) into training data with the predicted mean of that point as label (note it is not the true label, i.e.
-observation), and update the predicted variance then update the UCB scores.
-We recommend the second data point based on the new UCB scores.   
-As we can see, since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in another area and potentially having high labels ($x = 7.5$).
-By using GP-BUCB, we rule out the sequences with high similarities in the same batch and increase the exploration efficiency. 
-2 data points ($x = 8, x = 0$) are recommended in batch.
+As an example, consider the setting in Figure \ref{fig: Top UCB rec.}.
+First, GP-BUCB  recommends the data point with maximum UCB score based on the predictions over initial 5 observations as shown in Figure \ref{fig: GP-BUCB:_Recommend_1st_Data_Point.}.
+Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e.
+observation), and update the predicted variance and then we finally update the UCB scores.
+We then recommend the second data point based on the new UCB scores.
+As we can see, since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 7.5$).
+By using GP-BUCB, we can ensure that the sequences with high similarities will not be recommended in the same batch hence increasing the exploration efficiency.
 

### 2020-09-17 methods.tex
revise ML part as EE and MH suggested.

@@
 To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
-1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the respective confidence intervals. 
+1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
@@
 WDS is a type of a string kernel, which takes two sequences (strings) as inputs and outputs a scalar value which represents the similarities between the two sequences.  
-WDS kernel counts the matches of substrings of a certain length (i.e. kmers).
+WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
 The maximum substring length is specified by $\ell$. 
@@
 Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
+For example, for two core sequences \textit{ACCTGA} and \textit{CCTGAA}, there is a common part \textit{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.
+
 Let $\mathbb{I}$(A) is the indicator function, which equals 1 if $A$ is true and 0 otherwise. 
@@
 
-For recommendations of RBS sequences that should be experimentally labels next, we have used the \textit{Upper Confidence Bound (UCB)} algorithm.
-On the one hand, we want to exploit in terms of the design space where sequences are believed to have high labels (i.e. high predicted mean); 
+For recommendations of RBS sequences that should be experimentally labeled next, we have used the \textit{Upper Confidence Bound (UCB)} algorithm.
+On one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
 on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted SD).
@@
 Instead, we can recommend RBS sequences in a batch of size $n$. 
-One naive approach is to train GP on sequences with known TIR labels and recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig: Top UCB rec.} ($n = 2$).
+One naive approach is to 
+%train GP on sequences with known TIR labels and
+recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig:batch rec}(a) ($n = 2$).
 However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
@@
 \centering
-    \includegraphics[scale=0.6]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}
-    \caption{Batch Recommendation: Top UCB scores.
+\subfloat{\label{main:c}\includegraphics[scale=.49]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}}\par\medskip
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:a}\includegraphics[scale=.49]{plots/GP-BUCB:_Recommend_1st_Data_Point.pdf}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:b}\includegraphics[scale=.49]{plots/GP-BUCB:_Recommend_2nd_Data_Point.pdf}}
+\end{minipage}
+\caption{Batch Recommendation.
 We use the batch size of 2, with 5 initial observations.
@@
 {-2, -1.5, -1, ..., 9.5, 10}.
-The recommendations are 2 data points with top UCB scores, constructed with GP predictions.
 The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
-}
-    \label{fig: Top UCB rec.}
+(a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
+(b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
+\label{fig:batch rec}
 \end{figure}
@@
 % The plot shows the predicted mean and variance after observing 4 data points. 
-As an example, consider the setting in Figure \ref{fig: Top UCB rec.}.
-First, GP-BUCB  recommends the data point with maximum UCB score based on the predictions over initial 5 observations as shown in Figure \ref{fig: GP-BUCB:_Recommend_1st_Data_Point.}.
+First, GP-BUCB  recommends the data point with maximum UCB score based on the predictions over initial 5 observations as shown in Figure \ref{fig:batch rec}(b).  
 Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e.
@@
 We then recommend the second data point based on the new UCB scores.
-As we can see, since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 7.5$).
+As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 7.5$).
 By using GP-BUCB, we can ensure that the sequences with high similarities will not be recommended in the same batch hence increasing the exploration efficiency.
-
-\vspace{0.5cm}
-\begin{minipage}{\textwidth}
-  \begin{minipage}[b]{0.49\textwidth}
-    \centering
-    %\includegraphics[scale=0.4]{plots/Hist.pdf}
-    \includegraphics[scale=0.45]{plots/GP-BUCB:_Recommend_1st_Data_Point.pdf}
-     \captionsetup{type=figure}
-     \vspace{-0.5cm}
-    \caption{GP-BUCB: Recommend 1st Data Point.}
-    \label{fig: GP-BUCB:_Recommend_1st_Data_Point.}
-\end{minipage}
-  \hfill
-  \begin{minipage}[b]{0.49\textwidth}
-    \centering
-    \includegraphics[scale=0.45]{plots/GP-BUCB:_Recommend_2nd_Data_Point.pdf}
-     \captionsetup{type=figure}
-     \vspace{-0.5cm}
-    \caption{GP-BUCB: Recommend 2nd Data Point.
-    }
-    \label{fig: GP-BUCB:_Recommend_2nd_Data_Point.}
- \end{minipage}
- \end{minipage}
-\vspace{0.5cm}

### 2020-09-18 sub/SynBio_ML_Writing/methods.tex

@@
 
-To automatically design the RBS sequences in batch using machine learning, we consider two parts: 
-1) Design an regression algorithm which takes the RBS sequences as input and returns the predicted TIR scores and the confidence intervals. 
-2) Design an online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals. 
-Such online learning approach provides the $\textit{exploitation-exploration balance}$ that we use to control our machine learning process.
+We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}. 
+To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
+1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
+2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
 
-With the goal of finding RBS sequences with the largest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequentially optimising an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
-In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values of each points in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume the noise is following Gaussian distribution with unknown mean and variance).
-This noise is influenced by the accuracy of the RBS predictor and other experimental interference (e.g.
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.7]{plots/flowchart.pdf}
+    \caption{Flowchart of machine learning based experimental design.}
+    \label{fig: flowchart of machine learning based experimental design.}
+\end{figure}
+
+\subsubsection{LEARN: Gaussian Process Regression with Weighted Degree Kernel}
+
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values at each point in the selected set $\mathcal{S}_t$, i.e.
+$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume that the noise is following Gaussian distribution with unknown mean and variance).
+This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g.
 time, temperature, operator, etc.).
 
-For regression model, we consider the \textit{Gaussian Process Regression (GPR)}.
+For regression model, we have used the \textit{Gaussian Process Regression (GPR)}.
 A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach which provides uncertainty measurements on predictions. 
 We model $f$ as a sample from a \textit{Gaussian process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
-Denote the training testing data as $X, X_{*}$, and the training label as $\mathbf{y}$.
-Then the posterior of the test points (i.e.
-predictive distributions) is given by the conditional distribution $\mathbf{f}_\ast | X, \mathbf{y}, X_\ast \sim \mathcal{N}(\bar{\mathbf{f}}_\ast, cov(\mathbf{f}_\ast))$, where
-\begin{align}
-\overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} \mathbf{y} \\
-\label{Eq: predicted variance in main paper}
-\operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} K\left(X, X_{*}\right) 
-\end{align}
-%For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq.
-(\ref{Eq: predicted variance in main paper}).
+%As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
+GPR can predict both the posterior mean and posterior variance. The posterior variance represents the level of uncertainty for the prediction. 
+% We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0. 
+% As we can see, the true function is in the confidence interval when there are data points are observed.
 
-To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} \cite{ratsch_rase_2005} to specify the kernel function of $GP$.  
-The weighted degree kernel with shift extends the spectrum kernel \cite{leslie2001spectrum, ben2008support} 
-by considering the positional information and allowing some positional shifts of matching substrings.
-The spectrum kernel and the weighted degree kernel with shifts are string kernels, which take two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.
-We define spectrum kernel firstly,
-\begin{align}
-        k_\ell^{\text{Spec}}(\textbf{x}, \textbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
-    \end{align}
- where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
-$\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\textbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
-Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
-    \begin{align}
-         K_\ell^{\text{Spec}}(X, X^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
-    \end{align}
-The weighted degree kernel with shift is defined upon the spectrum kernel, with flexibility of adjusting the substring length $d$, the start position $l$ and the shift length $s$.
-WDS kernel counts the match of kmers at corresponding positions in two sequences as shown in the following,
+% \begin{figure}[t]
+%     \centering
+%     \includegraphics[scale=0.7]{plots/Prediction with Initial Observations.pdf}
+%     \caption{Gaussian Process Regression Example.
+This plot shows the GPR prediction with confidence interval based on 4 initial observations.
+The confidence interval are shown in predicted mean $\pm$ 1.96 standard deviation.}
+%     \label{fig: Gaussian Process Regression Example.}
+% \end{figure}
+
+The choice of covariance (kernel) function is critical for accurate predictions, since it controls smoothness and amplitude of the function we model.
+To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005} to specify the kernel function of $GP$.
+WDS is a type of a string kernel, which takes two sequences (strings) as inputs and outputs a scalar value which represents the similarities between the two sequences.  
+WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
+The maximum substring length is specified by $\ell$. 
+The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
+Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
+For example, for two core sequences \textit{ACCTGA} and \textit{CCTGAA}, there is a common part \textit{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.
+
+Let $\mathbb{I}$(A) is the indicator function, which equals 1 if $A$ is true and 0 otherwise. 
+Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ counts the matches of substrings of length $d$ between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
+This is similarly done for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
+By having these two terms considering substrings of two sequences with starting positions differing by $s$ characters, the WDS can measure shifted positional information. 
+When $s = 0$, the kernel function counts the matches with no shift between sequences. 
+Let $\mathbf{x}, \mathbf{x}^\prime$ be two RBS sequences with length $L$, the WDS kernel is defined as
 \begin{align}
         k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
-        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
-        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        %&= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        %\left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
+        = \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
         \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
 \end{align}
-where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
+where 
+$\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting parameter over the position in the
 sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
 range at position $l$.
-%Since the sequences in provided data have the pattern that the core area is different from each other, and other areas are similar.
-So the kernel for Gaussian Process we are using is the sum of kernels, for core areas we use spectrum kernel with string as input directly, and for other areas we use one-hot encoding and dot product kernel for simplicity.
 
@@
 
-For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
-%which is based on the \textit{optimism in the face of uncertainty}, 
-selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
+\subsubsection{DESIGN: Batch UCB}
+
+For recommendations of RBS sequences that should be experimentally labeled next, we have used the \textit{Upper Confidence Bound (UCB)} algorithm.
+On one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
+on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted SD).
+The UCB algorithm provides such $\textit{exploitation-exploration balance}$ by balancing the predicted mean and SD.
+More precisely, UCB algorithm selects RBS sequences with the maximum upper confidence bound at round $t$, i.e.
 \begin{align}
@@
 where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation (SD) at round $t$ for the sequence $\mathbf{x}_i$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
 
-Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label for each round.
-Therefore we consider recommending sequences in batch and use Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
-With batches of size $B$, the feedback mapping is then defined as $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
-\begin{align}
-    \mathrm{fb}[t]=\left\{\begin{array}{cl}
-    0 & : t \in\{1, \ldots, B\} \\
-    B & : t \in\{B+1, \ldots, 2 B\} \\
-    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
-    & \vdots
-    \end{array}\right.
-\end{align}
-
-A key property of Gaussian Process regression is that the predictive variance in Eq.
-(\ref{Eq: predicted variance in main paper}) only depends on observed points (i.e.
-features), but not on the labels of those observed points.
-So one can compute the posterior variance without actually observing the labels. 
-The GP-BUCB policy is to select sequences that
-\begin{align}
-    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right).
-\end{align}
-And only update $y_{t^{\prime}}=f\left(\boldsymbol{x}_{t^{\prime}}\right)+\varepsilon_{t^{\prime}} \text { for } t^{\prime} \in\{\mathrm{fb}[t]+1, \ldots, \mathrm{fb}[t+1]\}$ at the end of each batch ($\mathrm{fb}[t]<\mathrm{fb}[t+1]$).
-This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right]$, while the posterior variance decreases.
+Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e.
+one-by-one) and then wait for the label to be tested and used to improve the model.
+Instead, we can recommend RBS sequences in a batch of size $n$. 
+One naive approach is to 
+%train GP on sequences with known TIR labels and
+recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig:batch rec}(a) ($n = 2$).
+However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
+However, since we assume similar sequences would have similar labels (e.g.
+by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
 
@@
 \centering
-    \includegraphics[scale=0.7]{plots/flowchart.pdf}
-    \caption{Flowchart of machine learning based experimental design.}
-    \label{fig: flowchart of machine learning based experimental design.}
+\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}}\par\medskip
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB:_Recommend_1st_Data_Point.pdf}}
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB:_Recommend_2nd_Data_Point.pdf}}
+\end{minipage}
+\caption{Batch Recommendation.
+We use the batch size of 2, with 5 initial observations.
+The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e.
+{-2, -1.5, -1, ..., 9.5, 10}.
+The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
+(a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
+(b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
+\label{fig:batch rec}
 \end{figure}
+
+A key property of Gaussian Process regression is that the predictive variance  depends only on observed points (i.e. features), but not on the labels of those observed points. 
+% We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
+% As we can see, the predicted confidence interval is the same for the two plots. 
+We can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
+That is, we can recommend sequences sequentially by updating the UCB score with the updated predicted SD with the the previously recommended points.
+% Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
+% The plot shows the predicted mean and variance after observing 4 data points. 
+First, GP-BUCB  recommends the data point with maximum UCB score based on the predictions over initial 5 observations as shown in Figure \ref{fig:batch rec}(b).  
+Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e.
+observation), and update the predicted variance and then we finally update the UCB scores.
+We then recommend the second data point based on the new UCB scores.
+As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
+By using GP-BUCB, we can ensure that the sequences with high similarities will not be recommended in the same batch hence increasing the exploration efficiency.

### 2020-09-28 methods.tex
Squashed 'sub/SynBio_ML_Writing/' changes from 15a9eb6..fc97298
fc97298 update plots name
ba377d7 Number of changes to text flow.
639b4df change plots
0dc59c9 revise ML part as EE and MH suggested.
f2eaa9a update WDS kernel description; add exp-exp explanation.

git-subtree-dir: sub/SynBio_ML_Writing
git-subtree-split: fc97298311174519cf4090111f327499dfded902

@@
 
-\subsection{Plasmid Design}
+\subsection{Laboratory experimental design}
 
+\subsubsection{BUILD: Construction of genetic devices}
+
+\textbf{Plasmid Design.}
 We have used the pBbB6c-GFP plasmid for all our designs.
@@
 
-\subsection{PCR}
-PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, catalogue no.
-M0492L).
+\textbf{PCR.}
+PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, catalogue no. M0492L).
 20 $\mu$L reactions were prepared by dispensing each of the 10 $\mu$M reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
@@
 Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
-2$\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
-ZAG-110-5000; ZAG-130-5000).
+2$\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no. ZAG-110-5000; ZAG-130-5000).
 ProSize Data Analysis Software (Agilent Technologies) was used to generated gel images from the sample chromatograms and sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
 
-\subsection{Isothermal DNA Assembly}
-Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no.
-E2621L).
+\textbf{Isothermal DNA Assembly.}
+Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no. E2621L).
 Samples were incubated at 37$^{\circ}$C for 60 minutes, followed by a 20-minute heat inactivation step at 80$^{\circ}$C.
-Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10\(\mu\)L .
+Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10\(\mu\)L.
 Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
 
-\subsection{\textit{E. coli} transformation}
+\textbf{\textit{E. coli} transformation.}
 The DH5α cell line (Thermo Fisher Scientific, catalogue no.
@@
 ***).
-Multiple dilutions of cells in LB were prepared and plated in parallel on Omnitrays in 96 square grid.
-Plates were incubated overnight at 37$^{\circ}$C.
+Multiple dilutions of cells in LB were prepared and plated in parallel on Omnitrays in 96 square grid. Plates were incubated overnight at 37$^{\circ}$C.
 
-\subsection{Automated colony picking and culturing}
+\textbf{Automated colony picking and culturing.}
 A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates.
@@
 
-\subsection{Glycerol stock preparation}
+\textbf{Glycerol stock preparation.}
 100$\mu$L of sterile 80\% (v/v) glycerol and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
@@
 
-\subsection{Culture analysis}
+\subsubsection{TEST: Culture analysis}
+
+\textbf{Test strain culture.}
 Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2$\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
@@
 After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500$\mu$M IPTG – one plate of 300$\mu$L cultures for flow cytometry analysis (induced with 1.5$\mu$L of 0.1M IPTG) and one plate of 200$\mu$L cultures for plate reader analysis (induced with 1.0$\mu$L of 0.1M IPTG).
-•	Cytation 5 acquisition and incubation/shaking settings
-•	CytoFLEX S acquisition settings
+
+\textbf{Microplate spectrophotometry.}
+The plates were tested in Cytation5 microplate reader.
+Cytation 5 acquisition and incubation/shaking settings
+
+\textbf{Flow cytometry.}
+The plates were tested in Beckman Coulter CytoFLEX S flowcytometer.
+CytoFLEX S acquisition settings
 
@@
 \centering
-\subfloat{\label{main:c}\includegraphics[scale=.49]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}}\par\medskip
+\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}}\par\medskip
 \begin{minipage}{.5\linewidth}
 \centering
-\subfloat[]{\label{main:a}\includegraphics[scale=.49]{plots/GP-BUCB:_Recommend_1st_Data_Point.pdf}}
+\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
 \end{minipage}%
@@
 \centering
-\subfloat[]{\label{main:b}\includegraphics[scale=.49]{plots/GP-BUCB:_Recommend_2nd_Data_Point.pdf}}
+\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
 \end{minipage}
@@
 We then recommend the second data point based on the new UCB scores.
-As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 7.5$).
+As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
 By using GP-BUCB, we can ensure that the sequences with high similarities will not be recommended in the same batch hence increasing the exploration efficiency.

### 2020-09-28 sub/SynBio_ML_Writing/methods.tex

@@
 
-\subsection{Glycerol stock preparation}
+\textbf{Glycerol stock preparation.}
 100$\mu$L of sterile 80\% (v/v) glycerol and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
 They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.
+
+\subsubsection{TEST: Culture analysis}
+
+\textbf{Test strain culture.}
+Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2$\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
+Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours.
+The following morning, 20$\mu$L of overnight culture was added to 980$\mu$L of fresh selection medium and precultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate.
+After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500$\mu$M IPTG – one plate of 300$\mu$L cultures for flow cytometry analysis (induced with 1.5$\mu$L of 0.1M IPTG) and one plate of 200$\mu$L cultures for plate reader analysis (induced with 1.0$\mu$L of 0.1M IPTG).
+
+\textbf{Microplate spectrophotometry.}
+The plates were tested in Cytation5 microplate reader.
+Cytation 5 acquisition and incubation/shaking settings
 
@@
 \centering
-\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB:_Recommend_1st_Data_Point.pdf}}
+\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
 \end{minipage}%
@@
 \centering
-\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB:_Recommend_2nd_Data_Point.pdf}}
+\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
 \end{minipage}

### 2020-10-06 sub/SynBio_ML_Writing/main.tex
update

--- 2020-09-11 main.tex
@@
 \usepackage{textgreek}
+\usepackage{caption}
+%\usepackage{subcaption}
 
@@
 
-\title{Machine Learning based TIR predictor}
-\author{}
+\title{Machine Learning based Translation Initiation Rate predictor}
+\author{Zhang M., Holowko M. B., Hayman Zumpe H., Ong, C. S.}
 \date{\today{}}
@@
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
-This normally involves designing the DNA sequence in CAD software and then physically testing it in a laboratory.
+This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory.
 Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used\cite{Yeoh2019,Nielsen2016}.
@@
 \\
-Here we present how machine learning algorithms can be used as a part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimization of protein level expression.
+Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimization of protein level expression.
 RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
 We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
+Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
+We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (designated round zero since only preexisting, literature data is used for its design) or with RBS sequences recommended by the algorithm (subsequent rounds).
+The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
+After constructing, the plasmids harbouring the new genetic devices are tested in microplate reader and flow cytometer.
+The results are then fed back to the algorithm for it to recommend the next round of designs.\\
+>>>>>>> c6ee4d63c55638237c7c41a778b9374cca90cb60
 
@@
 
-Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
-We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (designated round zero since only preexisting, literature data is used for its design) or with RBS sequences recommended by the algorithm (subsequent rounds).
-The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
-After constructing, the plasmids harbouring the new genetic devices are tested in microplate reader and flow cytometer.
-The results are then fed back to the algorithm for it to recommend the next round of designs.\\ \textcolor{red}{Figure X} shows the DBTL cycle with the machine learning process emphasised.
+Figure 1 shows the DBTL cycle with the machine learning process emphasised.
 Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase.
@@
     \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
-The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.
-At the bottom a simplified SBOL diagram of the genetic device is shown.}
+The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
     \label{fig: Anatomy of the randomized sequence.}
@@
 The plasmids were then transformed and the resulting transformants were tested both using microplate reader and flow cytometry.\\
-\textcolor{red}{Figure X} shows the results of the zeroth round.  \\
-
-\textcolor{red}{Figure X} shows the results of the first and subsequent rounds.\\
+Figure X shows the results of the zeroth and all the subsequent rounds.\\
 
@@
 \section{Summary}
+
+\section{Contributions}
+Zhang M.
+and Ong C.
+S.
+designed and implemented the machine learning algorithms and workflow.
+Holowko M.
+B.
+and Hayman Zumpe H.
+have designed and performed the laboratory experiments.
+All authors contributed to and reviewed the manuscript.
 

### 2020-10-07 codes/maciek_code/ODrate.py
Updated code for data processing

@@ -1,41 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Fri Jul 31 15:01:36 2020
-
-@author: hol428
-"""
-
-import pandas as pd
-
-directory='../../data/First_round_results/'
-Filename='Second Plate 310720 (Rep 3 of 3, full).xlsx'
-Path = directory + Filename
-
-
-df = pd.read_excel(Path,sheet_name='ODGFP')
-df2 = pd.read_excel(Path,sheet_name='OD')
-df.drop(["Time"],axis=1,inplace=True)
-df2.drop(["Time"],axis=1,inplace=True)
-
-zupa = df.min(axis=0)
-
-time = [60,120,180,240]
-
-rates = pd.DataFrame(columns=['60','120','180','240','60OD','120OD','180OD','240OD'],index = list(df.columns))
-
-
-for t in time:
-    for column,value in zupa.iteritems():
-        start = df[df[column]==value].index.values.astype(int)[0]
-        end = start + (t/10)
-        rate = (df.loc[start:end, [column]].sum() / t)
-
-        rates.at[column,str(t)] = float(rate[0])
-        
-        rate2 = (df2.loc[start:end, [column]].sum() / t)
-
-        rates.at[column,str(t)+'OD'] = float(rate2[0])        
-
-print(rates)
-
-rates.to_csv(path_or_buf=directory+'rates.csv')

### 2020-10-07 codes/maciek_code/TIR calculator.py

@@ -1,0 +1,63 @@
+# -*- coding: utf-8 -*-
+"""
+TIR calculator
+
+by Maciej Holowko @CSIRO
+"""
+
+import pandas as pd
+
+directory='../../data/First_round_results/'
+Filename='TEST_DATA.xlsx'
+Path = directory + Filename
+
+OD = pd.read_excel(Path,sheet_name='OD')
+GFP = pd.read_excel(Path,sheet_name='GFP')
+GFPOD = OD.copy(deep=True)
+i=0
+
+OD.drop('Time',axis=1,inplace=True)
+
+for col in OD.columns:
+    for ind in GFPOD.index:
+        print(col)
+        print(ind)
+        GFPOD[col][ind]=GFP[col][ind]/OD[col][ind]
+        
+for ind in GFPOD.index:
+    GFPOD['Time'][ind]=i
+    i+=10
+
+GFPOD.set_index('Time',inplace=True)
+
+with pd.ExcelWriter(Path, engine="openpyxl", mode='a') as writer:  
+    GFPOD.to_excel(writer, sheet_name='GFPOD')
+
+GFPOD.reset_index(drop=True,inplace=True)
+
+minp = GFPOD.min(axis=0)
+
+time = [240]
+
+rates = pd.DataFrame(columns=['240','Comment'],index = list(GFPOD.columns))
+
+for t in time:
+    for column,value in minp.iteritems():
+        start = GFPOD[GFPOD[column]==value].index.values.astype(int)[0]
+        end = start + (t/10)
+        rate = (GFPOD.loc[start:end, [column]].sum() / t)
+        # print(start)
+        # print(end)
+        # print(rate)
+        rates.at[column,str(t)] = float(rate[0])
+        if rate[0]<0:
+            rates.at[column,'Comment'] = 'Impossible value'
+        elif rate[0]>100 and OD.max(axis=0)[column]<0.1:          
+            rates.at[column,'Comment'] = 'Possibly an empty well'
+        else:
+            rates.at[column,'Comment'] = 'Correct value'
+
+#print(rates)
+
+with pd.ExcelWriter(Path, engine="openpyxl", mode='a') as writer:  
+    rates.to_excel(writer, sheet_name='TIR')

### 2020-11-22 paper/appendix.tex
create paper folder

--- 2020-09-03 appendix.tex
@@
 \section*{Appendix}
+
+\section{Reproducible Plots}
+In this section, we put reproducible plots and will update the plots when we get full results.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale = 0.4]{plots/swarmplots.pdf}
+    \caption{Swarmplots for different groups, with raw (left) and normalised (right) TIR labels (averaged over 6 replicates). 
+    Group names represent: Consensus (consensus sequence tested in different round); BPS-NC (bps noncore); BPS-C (bps core); UNI (uniformly random); PPM (position-based probability matrix); Bandit-0 (bandit design for round 0); Bandit-1 (bandit design for round 1).}
+    \label{fig:swarmplots.}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.3]{plots/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
+    \includegraphics[scale = 0.3]{plots/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
+    \caption{Kernel Heatmap and Predictions.
+Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265).
+Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right).
+The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
+    \label{fig:my_label}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.25]{plots/clustering.png}
+    \caption{TSNE of RBS sequences in design space with clustering.
+The distance is calculated based on the weighted degree kernel on RBS sequences.
+Colours indicate different groups, shapes indicate different clusters.}
+    \label{fig:my_label}
+\end{figure}
 

### 2020-11-22 codes/data_generating.py
generating plots; adding plot format file.

@@ -24,6 +24,7 @@ data_format = str(args.Format) # str Seq or Sample
 
 # settings
 how_to_normalize = 'roundRep' # choices:  'plateRep', 'roundRep'
+# TODO: if one want to use plateRep, consensus sequence needs to be added for second plate
 COMPLETE_REP_SET = {'1','2','3','4','5','6','7','8','9'}
 sheet_name = 'Microplate' # for masterfile
 Log_flag = True # indicates whether take log label
@@ -110,6 +111,14 @@ def normalize(df, col_name):
     #     return df
 
 
+# rename group names
+def rename_group_names(df):
+    df['Group'] = df['Group'].replace({'reference': 'Consensus', 
+                        'bps_core':'BPS-C', 'bps_noncore': 'BPS-NC', 
+                        'uni random': 'UNI', 'prob random': 'PPM', 
+                        'bandit': 'Bandit-0', 'bandit2': 'Bandit-1'})
+    return df
+
 #-------------------------------------------------------------------------------------------
 # Process first round results file
 # Columns: 
@@ -165,21 +174,21 @@ if normalize_flag == 'True':
     df_new_norm['AVERAGE'] = df_new_norm.loc[: , "Rep1":"Rep6"].mean(axis=1)
     df_new_norm['STD'] = df_new_norm.loc[: , "Rep1":"Rep6"].std(axis=1)
     if data_format == 'Seq':
-        df_new_norm.to_csv(Folder_path + Generated_File_Path)
+        rename_group_names(df_new_norm).to_csv(Folder_path + Generated_File_Path)
     else: # sample
         df_new_norm_melt = pd.melt(df_new_norm, id_vars=['Name', 'RBS', 'RBS6', 'AVERAGE', 'STD', 'Group'], 
                                     value_vars=['Rep2', 'Rep3', 'Rep4', 'Rep5', 'Rep6', 'Rep7', 'Rep8', 'Rep9', 'Rep1'])
         df_new_norm_melt = df_new_norm_melt.rename(columns = {'value': 'label'})
         df_new_norm_melt = df_new_norm_melt.dropna()
-        df_new_norm_melt.to_csv(Folder_path + Generated_File_Path)
+        rename_group_names(df_new_norm_melt).to_csv(Folder_path + Generated_File_Path)
 elif data_format == 'Seq':
     print('seq, no normalises')
[…]

### 2020-11-22 codes/plot_format.py

@@ -1,0 +1,15 @@
+from matplotlib import rcParams
+rcParams['axes.labelsize'] = 10
+rcParams['xtick.labelsize'] = 10
+rcParams['ytick.labelsize'] = 10
+rcParams['legend.fontsize'] = 10
+rcParams['font.family'] = 'serif'
+rcParams['font.serif'] = ['Computer Modern Roman']
+rcParams['text.usetex'] = True
+
+# generate valids names and path for plots
+
+def valid_name(name):
+    return name.replace('_', ' ')  
+def valid_path(path):
+    return path.replace(' ', '_').replace(',', '')

### 2020-11-22 codes/regression.py

@@ -24,6 +24,7 @@ from codes.ucb import GPUCB, Random
 from codes.evaluations import evaluate, plot_eva
 from codes.kernels_for_GPK import *
 from codes.sort_seq import sort_kernel_matrix
+from codes.plot_format import *
 
 from ipywidgets import IntProgress
 
@@ -361,14 +362,17 @@ class GPR_Predictor():
         df = df.loc[order,:].reset_index()
         # axl = plt.subplot(gs[0,0], sharey=ax)
         axb = plt.subplot(gs[1], sharex=ax)
-        axb.plot(df.index, df['pred mean'], label = 'pred')
+        axb.plot(df.index, df['pred mean'], label = 'Pred Mean')
         axb.fill_between(df.index, df['pred mean'] + 1.96 * df['pred std'],df['pred mean'] - 1.96 * df['pred std'], alpha = 0.5)
-        axb.scatter(df[df['train_test'] == 'train'].index, df[df['train_test'] == 'train']['AVERAGE'], s =1, c='green', label = 'train')
-        axb.scatter(df[df['train_test'] == 'test'].index, df[df['train_test'] == 'test']['AVERAGE'], s =1, c='red', label = 'test')
+        axb.scatter(df[df['train_test'] == 'train'].index, df[df['train_test'] == 'train']['AVERAGE'], s =1, c='green', label = 'Train Label')
+        axb.scatter(df[df['train_test'] == 'test'].index, df[df['train_test'] == 'test']['AVERAGE'], s =1, c='red', label = 'Test Label')
         # axb.plot(t, a.mean(0))
+        axb.set_xlabel('RBS Sequences')
+        axb.set_ylabel('TIR Normalised Label')
         axb.legend()
-        ax.set_title(title)
-        plt.show()
+        ax.set_title(valid_name(title))
+        # plt.show()
+        plt.savefig(valid_path(title) + '.pdf', bbox_inches='tight')
 
     def coverage_rate(self, true_label, pred_mean, pred_std):
         """evaluation metric of prediction

### 2020-11-22 codes/sort_seq.py

@@ -4,15 +4,18 @@ from collections import defaultdict
 import math
 from sklearn.cluster import KMeans, AgglomerativeClustering
 
+SORTED_GROUP = ['Consensus', 'BPS-C', 'BPS-NC', 'UNI', 'PPM', 'Bandit-0', 'Bandit-1']
+
 def sort_kernel_matrix(df, feature_kernel, kmeans_based_on='label_distance'):
     group_dict = df.groupby('Group').groups
     groups = list(group_dict.keys())
-    print('groups: ', groups)
     new_ordering = [] # idx ordering
+    intersection = [i for i in SORTED_GROUP if i in set(groups)]
+    print('groups: ', intersection)
 
     frr_seqs = np.asarray(df['RBS'])
 
-    for group in groups:
+    for group in intersection:
         df_group = df[df['Group'] == group]
         num_seqs = len(df_group)
         # num_clusters = int(num_seqs/5) + 1
@@ -39,10 +42,10 @@ def sort_kernel_matrix(df, feature_kernel, kmeans_based_on='label_distance'):
             num_clusters = int(num_seqs/8) + 1
             print('number of clusters: ', num_clusters)
             #kmeans = KMeans(n_clusters = num_clusters, random_state = 0).fit(phi_X[idx[0]: idx[-1] + 1, :])
+            cluster_dict = defaultdict(list) # key: cluster id; value: idx list
             if len(idx) > 1:
                 model = AgglomerativeClustering(n_clusters=num_clusters)
                 model.fit(feature_kernel[idx])
-                cluster_dict = defaultdict(list) # key: cluster id; value: idx list
                 for i, cluster_id in enumerate(model.labels_):
                     cluster_dict[cluster_id].append(idx[i])
             # print('cluster dict: ', cluster_dict)

### 2020-11-23 codes/kernels_for_GPK.py
repeat kfold for kernel normalisation adjusting

@@ -78,6 +78,8 @@ class String_Kernel(Kernel):
                  sigma_0= 1, #, sigma_0_bounds=(1e-10,1e10)):
                  lengthscale_rate = None,
                  kernel_norm_flag = True,
+                 centering_flag = True,
+                 unit_norm_flag = True,
                 ):
         
         self.l = l
@@ -114,7 +116,7 @@ class String_Kernel(Kernel):
 
             # TODO: decide whether to use sigma_0
             if kernel_norm_flag:
-                self.kernel_all_normalised = self.normalisation(self.kernel_all) * sigma_0
+                self.kernel_all_normalised = self.normalisation(self.kernel_all, centering_flag, unit_norm_flag) * sigma_0
             else:
                 print('USE non-normalised kernel!')
                 self.kernel_all_normalised = self.kernel_all * sigma_0
@@ -138,7 +140,7 @@ class String_Kernel(Kernel):
                 # plt.title('Histogram of upper triangle of D^2')
                 lengthscale = np.quantile(self.distance_all**2, lengthscale_rate)
                 print('lengthscale: ', lengthscale)
-                self.kernel_all_normalised = self.normalisation(- self.distance_all**2/lengthscale) * sigma_0
+                self.kernel_all_normalised = self.normalisation(- self.distance_all**2/lengthscale, centering_flag, unit_norm_flag) * sigma_0
             print('init kernel')
             
             type(self).INIT_FLAG = True
@@ -201,23 +203,29 @@ class String_Kernel(Kernel):
         else:
             return K
     
-    def normalisation(self, kernel):
+    def normalisation(self, kernel, centering_flag = True, unit_norm_flag = True):
         # https://jmlr.csail.mit.edu/papers/volume12/kloft11a/kloft11a.pdf
         # Sec 4.2.2
         # First calculate zero mean kernel
         # Then calculate unit variance
         # The the variance (defined as the 1/n trace - kernel mean) is 1 
         # normalise over the whole matrix (train, test)
-        
-        # centering
-        one_mat = np.ones(kernel.shape) # kernel shape, i.e. (n_train+n_test, n_train+n_test)
-        one_vec = np.ones((kernel.shape[0],1)) # (n_train+n_test, 1)
+        print('centering_flag: ', centering_flag)
[…]

### 2020-11-23 codes/regression.py

@@ -45,7 +45,8 @@ class GPR_Predictor():
     def __init__(self, df, test_size=0.2, train_idx = None, test_idx = None, 
                  kernel_name='WD_Kernel_Shift', l=6, s = 1, b=0.33, sigma_0 = 1, padding_flag=False, gap_flag=False,
                  alpha=2, embedding='label', eva_metric=[r2_score, mean_squared_error],  eva_on = "samples",
-                 kernel_norm_flag = True,
+                 kernel_norm_flag = True, centering_flag = True, unit_norm_flag = True,
+                #  kernel_norm_flag = True,　centering_flag = True,　unit_norm_flag = True,
                  ):
         """
         Parameter
@@ -86,6 +87,13 @@ class GPR_Predictor():
             default is r2_score and mean_square_error
         eva_on: string
             indicates evaluating on samples or seqs
+
+        kernel_norm_flag: boolean
+            indicates whether to use kernel normalisation
+        centering_flag: boolean
+            indicates whether to do kernel centering
+        unit_norm_flag: boolean
+            indicates whether to do kernel unit norm normalisation
         """
         self.df = df
         self.test_size = test_size
@@ -96,6 +104,8 @@ class GPR_Predictor():
         self.kernel_name = kernel_name
         self.kernel = KERNEL_DICT[kernel_name]
         self.kernel_norm_flag = kernel_norm_flag
+        self.centering_flag = centering_flag
+        self.unit_norm_flag = unit_norm_flag
         # self.kernel.INIT_FLAG = False 
         # so that kernel will be initialised (cal_kernel)
         self.l = l
@@ -119,7 +129,7 @@ class GPR_Predictor():
         self.test_idx = np.asarray(self.test_idx)
 
     def Train_val_split(self, cv = 5, random_state = 24):
-        kf = KFold(n_splits = cv, shuffle = True)
+        kf = KFold(n_splits = cv, shuffle = True, random_state = random_state)
         return kf.split(range(self.num_data))
 
     def Generate_train_test_data(self):
@@ -205,7 +215,8 @@ class GPR_Predictor():
             print('create kernel instance')
             self.wd_kernel_instance = self.kernel(l = self.l, features = self.features, 
                                          n_train=X_train.shape[0], n_test=X_test.shape[0],
-                                         s = self.s, sigma_0=self.sigma_0, kernel_norm_flag = self.kernel_norm_flag)
[…]

### 2020-11-24 codes/kernels_for_GPK.py
kernel normalisation over all seq,  including unknown seq

@@ -5,6 +5,9 @@ from sklearn import preprocessing
 from sklearn.gaussian_process.kernels import Kernel, Hyperparameter
 import matplotlib.pyplot as plt
 
+from codes.embedding import Embedding
+from codes.environment import Rewards_env
+
 import os
 import sys
 module_path = os.path.abspath(os.path.join('../..'))
@@ -26,10 +29,11 @@ if module_path not in sys.path:
 BASES = ['A','C','G','T']
 
 # global features for all known and design data
-if os.path.exists('../../data/known_design.csv'):
-    FEATURES = np.asarray(pd.read_csv('../../data/known_design.csv')['RBS'])
+if os.path.exists('../../data/unknown_design.csv'):
+    unknown_df = pd.read_csv('../../data/unknown_design.csv') # sequences with unknown label
+    unknown_features = Rewards_env(np.asarray(unknown_df[['RBS', 'RBS6']]), 'label').embedded
 else:
-    FEATURES = None
+    unknown_features = None
     print('No default features for kernel instance. Please specify features.')
 
 class String_Kernel(Kernel):
@@ -73,7 +77,7 @@ class String_Kernel(Kernel):
     KERNEL_ALL_NORM = None
     DISTANCE_ALL = None
 
-    def __init__(self, l=6, features = FEATURES, n_train = None, n_test = None,
+    def __init__(self, l=6, features = None, n_train = None, n_test = None,
                  padding_flag = False, gap_flag = False,
                  sigma_0= 1, #, sigma_0_bounds=(1e-10,1e10)):
                  lengthscale_rate = None,
@@ -99,13 +103,22 @@ class String_Kernel(Kernel):
         self.n_train = n_train
         self.n_test = n_test
         self.features = features # train has to be put at the front
+        print('train size: ', self.n_train)
+        print('test size: ', self.n_test)
+        print('feature size: ', self.features.shape)
+        assert self.n_train + self.n_test == self.features.shape[0]
 
         # global INIT_FLAG
         # global KERNEL_ALL_NORM
         # global DISTANCE_ALL
 
         if not type(self).INIT_FLAG: # to avoid init again in clone (deepcopy) in GPR.fit
-            self.kernel_all = self.cal_kernel(self.features, self.features)
+            if self.features.shape[0] < unknown_features.shape[0]:
+                # use all known + unknown seq for normalisation
+                self.known_unknown_features = np.concatenate((self.features, unknown_features), axis=0)
[…]

### 2020-11-24 codes/environment.py
generate kernel matrix with unique seq

@@ -30,15 +30,16 @@ class Rewards_env():
         """
         self.data = data
         self.num_seq, self.num_col = self.data.shape
-        assert self.num_col == 2
+        # assert self.num_col == 2
         self.embedding_method = embedding_method
 
         self.embedded = self.arm_embedding()
 
-        self.rewards_dict = defaultdict(list)
-        self.create_rewards_dict()
-        self.labels_dict = self.labels_generator()
-        self.arm_features = list(self.labels_dict.keys())
+        # TODO: comment for now, may need to put it back
+        # self.rewards_dict = defaultdict(list)
+        # self.create_rewards_dict()
+        # self.labels_dict = self.labels_generator()
+        # self.arm_features = list(self.labels_dict.keys())
         
     def arm_embedding(self):
         """Embedding biological sequence.

### 2020-11-24 codes/generate_idx_seq.py

@@ -1,0 +1,76 @@
+# Nov.2020 Mengyan Zhang 
+# This file generates a two-way dict (index, sequences) for all design RBS sequences,
+# including 4^6 (core-part) + 3*14 (bps non-core group) = 4138 sequences. 
+# 
+
+import os
+import sys
+module_path = os.path.abspath(os.path.join('../..'))
+if module_path not in sys.path:
+    sys.path.append(module_path)
+import pandas as pd
+import numpy as np
+import itertools
+import pickle
+
+class TwoWayDict(dict):
+    """bijective map between index and sequences.
+    code from Sasha Chedygov's answer https://stackoverflow.com/questions/1456373/two-way-reverse-map
+    """
+    def __setitem__(self, key, value):
+        # Remove any previous connections with these values
+        if key in self:
+            del self[key]
+        if value in self:
+            del self[value]
+        dict.__setitem__(self, key, value)
+        dict.__setitem__(self, value, key)
+
+    def __delitem__(self, key):
+        dict.__delitem__(self, self[key])
+        dict.__delitem__(self, key)
+
+    def __len__(self):
+        """Returns the number of connections"""
+        return dict.__len__(self) // 2 
+
+
+# Setting
+char_sets = ['A','C','G','T']
+design_len = 6
+pre_design = 'TTTAAGA'
+pos_design = 'TATACAT'
+index = 0
+index_list = []
+seq_list = []
+idx_seq_dict = TwoWayDict()
+
+for combo in itertools.product(char_sets, repeat= design_len):
+    combo = pre_design + ''.join(combo) + pos_design
+    index_list.append(index)
+    seq_list.append(combo)
+    idx_seq_dict[index] = combo
+    index += 1
+assert len(idx_seq_dict) == len(char_sets) ** design_len
+
+df = pd.read_excel(os.getcwd()+'/data/Results_Masterfile.xlsx', sheet_name= 'Microplate')
+bps_noncore = df[df['Group'] == 'bps_noncore']['RBS']
+for seq in bps_noncore:
+    index_list.append(index)
+    seq_list.append(seq.upper())
+    idx_seq_dict[index] = seq.upper()
+    index += 1
+assert len(idx_seq_dict) == len(char_sets) ** design_len + (len(char_sets) - 1) * (20-design_len)
+print(len(idx_seq_dict))
+
[…]

### 2020-11-24 codes/kernels_for_GPK.py

@@ -27,14 +27,16 @@ if module_path not in sys.path:
 
 # The all bases for input strings, we use the RBS sequences here
 BASES = ['A','C','G','T']
+SAVED_IDX_SEQ_PATH = '../../data/idx_seq.npz'
+SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.npz'
 
 # global features for all known and design data
-if os.path.exists('../../data/unknown_design.csv'):
-    unknown_df = pd.read_csv('../../data/unknown_design.csv') # sequences with unknown label
-    unknown_features = Rewards_env(np.asarray(unknown_df[['RBS', 'RBS6']]), 'label').embedded
-else:
-    unknown_features = None
-    print('No default features for kernel instance. Please specify features.')
+# if os.path.exists('../../data/unknown_design.csv'):
+#     unknown_df = pd.read_csv('../../data/unknown_design.csv') # sequences with unknown label
+#     unknown_features = Rewards_env(np.asarray(unknown_df[['RBS', 'RBS6']]), 'label').embedded
+# else:
+#     unknown_features = None
+#     print('No default features for kernel instance. Please specify features.')
 
 class String_Kernel(Kernel):
     """Base class for string kernels
@@ -108,64 +110,84 @@ class String_Kernel(Kernel):
         print('feature size: ', self.features.shape)
         assert self.n_train + self.n_test == self.features.shape[0]
 
-        # global INIT_FLAG
-        # global KERNEL_ALL_NORM
-        # global DISTANCE_ALL
+        #----------------------------------------------------------------------------------------
+        # reconstructure
 
-        if not type(self).INIT_FLAG: # to avoid init again in clone (deepcopy) in GPR.fit
-            if self.features.shape[0] < unknown_features.shape[0]:
-                # use all known + unknown seq for normalisation
-                self.known_unknown_features = np.concatenate((self.features, unknown_features), axis=0)
-                print('add unknown seq: ', self.known_unknown_features.shape)
-                
-            self.kernel_all = self.cal_kernel(self.known_unknown_features, self.known_unknown_features)
+        if os.path.exists(SAVED_KERNEL_PATH):
+            print('Load saved kernel matrix...')
+            self.kernel_all_normalised = np.load(SAVED_KERNEL_PATH)['kernel']
+            # TODO: need to check whether the saved kernel is generated with same parameters
+        else:
+            print('No saved kernel matrix. Calculate and save kernel matrix...')
+            seq = np.load(SAVED_IDX_SEQ_PATH)['seqList']
[…]

### 2020-11-24 paper/main.tex
Small changes.

--- 2020-10-06 sub/SynBio_ML_Writing/main.tex
@@
 The results are then fed back to the algorithm for it to recommend the next round of designs.\\
->>>>>>> c6ee4d63c55638237c7c41a778b9374cca90cb60
 
@@
 coli} is AGGAGG.
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure 1).\\
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
+    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
+The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
+    \label{fig:Anatomy}
+\end{figure}
+
 In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant.
@@
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
-    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
-The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
-    \label{fig: Anatomy of the randomized sequence.}
-\end{figure}
-
 \subsection{BUILD and TEST}
@@
 For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
-These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and you have to cover 5-10\% of that space.
+These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and one have to cover 5-10\% of that space.
 And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\

### 2020-11-24 paper/methods.tex
create paper folder
Small changes.

--- 2020-09-28 methods.tex
@@
 We have used the pBbB6c-GFP plasmid for all our designs.
-This plasmid comes with GFP mut3b CDS inducible with addition of IPTG.
+This plasmid comes with GFP mut3b CDS inducible with addition of Isopropyl β-D-1-thiogalactopyranoside (IPTG).
 The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
@@
 They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.
+
+\textbf{Sequencing}
+Sequences that showed TIR on the level of at least 85\% of TIR of the consensus sequence were sequenced using Sanger sequencing either at the Australian Genome Research Facility (Brisbane, QLD, Australia) or at Macrogen (Seoul, Republic of Korea).
 
@@
 \centering
-\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Points:_Top_2_UCB.pdf}}\par\medskip
+\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}}\par\medskip
 \begin{minipage}{.5\linewidth}

### 2020-11-27 codes/config.py
assign each seq an index; kernel normalisation over all 4138 unique seq; run repeated kfold, rank ucb based on new kernel normalisation.

@@ -1,0 +1,14 @@
+# provide configuration, e.g. global variable, path
+
+BASES = ['A','C','G','T'] # embedding
+
+SAVED_IDX_SEQ_PATH = '../../data/idx_seq.pickle'
+# Usage:
+# with open(SAVED_IDX_SEQ_PATH, 'rb') as handle:
+#     saved_idx_seq = pickle.load(handle)
+# seq_list = saved_idx_seq['seq_list']
+# idx_list = saved_idx_seq['idx_list']
+# idx_seq_dict = saved_idx_seq['idx_seq_dict']
+
+SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.pickle'
+

### 2020-11-27 codes/generate_idx_seq.py

@@ -13,26 +13,26 @@ import numpy as np
 import itertools
 import pickle
 
-class TwoWayDict(dict):
-    """bijective map between index and sequences.
-    code from Sasha Chedygov's answer https://stackoverflow.com/questions/1456373/two-way-reverse-map
-    """
-    def __setitem__(self, key, value):
-        # Remove any previous connections with these values
-        if key in self:
-            del self[key]
-        if value in self:
-            del self[value]
-        dict.__setitem__(self, key, value)
-        dict.__setitem__(self, value, key)
+# class TwoWayDict(dict):
+#     """bijective map between index and sequences.
+#     code from Sasha Chedygov's answer https://stackoverflow.com/questions/1456373/two-way-reverse-map
+#     """
+#     def __setitem__(self, key, value):
+#         # Remove any previous connections with these values
+#         if key in self:
+#             del self[key]
+#         if value in self:
+#             del self[value]
+#         dict.__setitem__(self, key, value)
+#         dict.__setitem__(self, value, key)
 
-    def __delitem__(self, key):
-        dict.__delitem__(self, self[key])
-        dict.__delitem__(self, key)
+#     def __delitem__(self, key):
+#         dict.__delitem__(self, self[key])
+#         dict.__delitem__(self, key)
 
-    def __len__(self):
-        """Returns the number of connections"""
-        return dict.__len__(self) // 2 
+#     def __len__(self):
+#         """Returns the number of connections"""
+#         return dict.__len__(self) // 2 
 
 
 # Setting
@@ -43,15 +43,16 @@ pos_design = 'TATACAT'
 index = 0
 index_list = []
 seq_list = []
-idx_seq_dict = TwoWayDict()
+idx_seq_dict = {}
 
 for combo in itertools.product(char_sets, repeat= design_len):
     combo = pre_design + ''.join(combo) + pos_design
     index_list.append(index)
     seq_list.append(combo)
     idx_seq_dict[index] = combo
+    idx_seq_dict[combo] = index
     index += 1
[…]

### 2020-11-27 codes/kernels_for_GPK.py

@@ -13,6 +13,8 @@ import sys
 module_path = os.path.abspath(os.path.join('../..'))
 if module_path not in sys.path:
     sys.path.append(module_path)
+import pickle
+import codes.config as config # provide configuration, e.g. global variable, path
 
 # Aug 2020 Mengyan Zhang 
 # String Kernel Classes, taking strings as input
@@ -26,9 +28,7 @@ if module_path not in sys.path:
 # For Gaussian Process Regression, the specific kernel matrix is sliced from the calculated kernel
 
 # The all bases for input strings, we use the RBS sequences here
-BASES = ['A','C','G','T']
-SAVED_IDX_SEQ_PATH = '../../data/idx_seq.npz'
-SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.npz'
+
 
 # global features for all known and design data
 # if os.path.exists('../../data/unknown_design.csv'):
@@ -79,7 +79,7 @@ class String_Kernel(Kernel):
     KERNEL_ALL_NORM = None
     DISTANCE_ALL = None
 
-    def __init__(self, l=6, features = None, n_train = None, n_test = None,
+    def __init__(self, l=6, 
                  padding_flag = False, gap_flag = False,
                  sigma_0= 1, #, sigma_0_bounds=(1e-10,1e10)):
                  lengthscale_rate = None,
@@ -91,7 +91,13 @@ class String_Kernel(Kernel):
         self.l = l
         self.padding_flag = padding_flag
         self.gap_flag = gap_flag
-        # self.sigma_0 = sigma_0
+        self.sigma_0 = sigma_0
+        self.lengthscale_rate = lengthscale_rate
+
+        self.kernel_norm_flag = kernel_norm_flag
+        self.centering_flag = centering_flag
+        self.unit_norm_flag = unit_norm_flag
+
         # self.sigma_0_bounds = sigma_0_bounds
 
         # calculate the whole kernel matrix for the concatenate train and test data with normalisation
@@ -102,24 +108,32 @@ class String_Kernel(Kernel):
         # features (n_train + n_test) * d
         # kernel_all (n_train + n_test) * (n_train + n_test)
         
-        self.n_train = n_train
-        self.n_test = n_test
-        self.features = features # train has to be put at the front
-        print('train size: ', self.n_train)
-        print('test size: ', self.n_test)
-        print('feature size: ', self.features.shape)
-        assert self.n_train + self.n_test == self.features.shape[0]
+        # self.train_idx = np.asarray(train_idx)
[…]

### 2020-11-27 codes/sort_seq.py

@@ -19,8 +19,8 @@ def sort_kernel_matrix(df, feature_kernel, kmeans_based_on='label_distance'):
         df_group = df[df['Group'] == group]
         num_seqs = len(df_group)
         # num_clusters = int(num_seqs/5) + 1
-        print('Group: ', group)
-        print('Number of sequences: ', num_seqs)
+        # print('Group: ', group)
+        # print('Number of sequences: ', num_seqs)
         # print('number of clusters: ', num_clusters)
         
         idx = np.asarray(group_dict[group])
@@ -30,7 +30,7 @@ def sort_kernel_matrix(df, feature_kernel, kmeans_based_on='label_distance'):
         
         if kmeans_based_on == 'label_distance': # kmeans based on label distances
             num_clusters = int(num_seqs/5) + 1
-            print('number of clusters: ', num_clusters)
+            # print('number of clusters: ', num_clusters)
             kmeans = KMeans(n_clusters = num_clusters, random_state = 0).fit(np.asarray(df['AVERAGE'])[idx].reshape(len(idx),1))
             cluster_dict = defaultdict(list) # key: cluster id; value: idx list
             for i, cluster_id in enumerate(kmeans.labels_):
@@ -40,7 +40,7 @@ def sort_kernel_matrix(df, feature_kernel, kmeans_based_on='label_distance'):
             
         elif kmeans_based_on == 'seq_distance': # kmeans based on spectrum distances 
             num_clusters = int(num_seqs/8) + 1
-            print('number of clusters: ', num_clusters)
+            # print('number of clusters: ', num_clusters)
             #kmeans = KMeans(n_clusters = num_clusters, random_state = 0).fit(phi_X[idx[0]: idx[-1] + 1, :])
             cluster_dict = defaultdict(list) # key: cluster id; value: idx list
             if len(idx) > 1:

### 2020-12-14 paper/appendix.tex
Made changes to paper requested by Cheng Soon. Added statistical test vs core vs non-core group.

@@
 
+\section{Core vs non-core regions of the RBS}
+According to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
+Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS.
+This has also been seen in this work in the 1by1 group of RBSes.
+Change in base in the core region has had a much more profound effect on the TIR than when the base change happened outside the core (Figure \ref{fig:core_vs_noncore}).
+The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.6]{plots/core_vs_noncore.pdf}
+    \caption{\textbf{Comaprison of base change impact on TIR in core versus non-core region.} The core region is highligthed in light green and the lines are rolling averages for each base.
+The TIRs obtained for RBSes where the base change occured outside the core region are significantly less }
+    \label{fig:core_vs_noncore}
+\end{figure}
+
 \section{Intuition behind UCB and visualisation}

### 2020-12-14 paper/main.tex
Made changes to paper requested by Cheng Soon. Added statistical test vs core vs non-core group.

@@
 This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-The label values are between 0 - 100,000 and skewed, which is shown in Figure xx.
-First, we have normalised the label to 0 - 1 using the min-max normalisation.
+This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
+The label values are between 0 - 100,000 and skewed, which is shown in Figure xx. First, we have normalised the label to 0 - 1 using the min-max normalisation. 
 The predictions are shown in \textcolor{red}{Figure Xb}\\
@@
 \subsection{DESIGN of the genetic device}
-There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992}.
+There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992,SHULTZABERGER2001}.
 In \emph{E.

### 2021-03-24 paper/main.tex
Updated abstract, introduction and methods.

@@
 
+\section{Abstract}
+
+Design of small genetic control elements like promoters, binding sites and terminators is currently challenging.
+To a large extent genetic engineers are relying on parts databases and otherwise known sequences to inform designs.
+This means, however, that because the known sequences may not be optimal for the desired output, optimal expression of constructs may not be achieved.
+This problem could be solved by a tailored design of these sequences, however, there are no reliable rules for such design.
+Here, we present a machine learning based workflow for design of one small genetic control element, a bacterial ribosome binding site.
+We have used Gaussian Process for prediction of a relative Transcription Initiation Rates and Upper Confidence Bound Bandit algorithm for recommendation of genetic designs to be tested in vivo.
+By tying use of the algorithms with laboratory automation and high-throughput processes we have created a robust workflow for design of custom bacterial binding sites.
+Results sentence.
+This work has also elucidated some design guidelines, including the required edit distance between two sequences and favourable representation of the analysed DNA sequence.
+
 \section{Introduction}
@@
 Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
-The biggest limitation for this approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve designs based on experimental results.\\
-According to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
+However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
+This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.\\
+The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
+For example, according to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
-\\
+There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}.\\
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
@@
 These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018}.
-In the past few years there was a significant uptake of Machine Learning based approaches in synthetic biology.
+In the past few years there was a significant uptake of Machine Learning based approaches in synthetic biology \cite{LAWSON2021}.
 Jervis \emph{et al.} used support vector machine and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
@@
 There were also successful attempts at using deep learning techniques for analysis of big data-sets \cite{Alipanahi2015,Angermueller2016}.
-However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential.
-\\
+Machine learning has been also used for prediction in proteins \cite{Yang2018}.
+However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. \\
 Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimization of protein level expression.
-RBS being one of the key genetic elements controlling protein expression and at he same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
+RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
 We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
 Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
-We did this by designing a sequential experimental workflow, where we start with either randomised RBS sequences designed to explore the experimental space (designated round zero since only preexisting, literature data is used for its design) or with RBS sequences recommended by the algorithm (subsequent rounds).
+We did this by designing a sequential experimental workflow, where designs in the zeroth round where randomised RBS sequences designed to explore the experimental space and some preliminary machine learning recommended designs based on literature.
+In the subsequent rounds, designs where recommended by the algorithm based on the data obtained in previous rounds. 
 The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
-After constructing, the plasmids harbouring the new genetic devices are tested in microplate reader and flow cytometer.
-The results are then fed back to the algorithm for it to recommend the next round of designs.\\
+After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
+The results were then fed back to the algorithm for it to recommend the next round of designs.\\
 

### 2021-03-24 paper/methods.tex
Updated abstract, introduction and methods.

@@
 \textbf{Plasmid Design.}
-We have used the pBbB6c-GFP plasmid for all our designs.
+The pBbB6c-GFP plasmid has been used for all our designs.
 This plasmid comes with GFP mut3b CDS inducible with addition of Isopropyl β-D-1-thiogalactopyranoside (IPTG).
@@
 20 $\mu$L reactions were prepared by dispensing each of the 10 $\mu$M reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
-A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 forward primer was prepared by and dispensed by hand.
+A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 $\mu$M forward primer was prepared by and dispensed by hand and/or Labcyte Echo.
 Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
-Then, samples were incubated at 37$^{\circ}$C for 60 minutes, followed by a 20-minute heat inactivation step at 80$^{\circ}$C.
 Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
@@
 Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no. E2621L).
-Samples were incubated at 37$^{\circ}$C for 60 minutes, followed by a 20-minute heat inactivation step at 80$^{\circ}$C.
 Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10\(\mu\)L.
 Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
+Finally, samples were incubated with addition of 50 nL of DpnI at 37$^{\circ}$C for 90 minutes.
 
@@
 ***).
-Multiple dilutions of cells in LB were prepared and plated in parallel on Omnitrays in 96 square grid. Plates were incubated overnight at 37$^{\circ}$C.
+Plates were incubated overnight at 37$^{\circ}$C.
 
@@
 \textbf{Sequencing}
-Sequences that showed TIR on the level of at least 85\% of TIR of the consensus sequence were sequenced using Sanger sequencing either at the Australian Genome Research Facility (Brisbane, QLD, Australia) or at Macrogen (Seoul, Republic of Korea).
+Sequences that showed TIR on the level of at least 85\% of TIR of the benchmark sequence were sequenced using Sanger sequencing either at the Australian Genome Research Facility (Brisbane, QLD, Australia) or at Macrogen (Seoul, Republic of Korea).
+Plasmids for sequencing were extracted using the Qiagen plasmid miniprep kit.
 
@@
 \textbf{Test strain culture.}
-Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2$\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
+Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
 Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours. 
-The following morning, 20$\mu$L of overnight culture was added to 980$\mu$L of fresh selection medium and precultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate.
-After 90 minutes, two parallel cultures were prepared in flat-bottom clear polystyrene 96-well plates and were induced with 500$\mu$M IPTG – one plate of 300$\mu$L cultures for flow cytometry analysis (induced with 1.5$\mu$L of 0.1M IPTG) and one plate of 200$\mu$L cultures for plate reader analysis (induced with 1.0$\mu$L of 0.1M IPTG).
+The following morning, 20$\mu$L of overnight cultures were added to 980$\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate. 
+After 90 minutes, 200$\mu$L of each culture (induced with 1.0$\mu$L of 0.1M IPTG) was transferred to a flat-bottom clear polystyrene 96-well plate.
 
@@
 Cytation 5 acquisition and incubation/shaking settings
-
-\textbf{Flow cytometry.}
-The plates were tested in Beckman Coulter CytoFLEX S flowcytometer.
-CytoFLEX S acquisition settings
 

### 2021-03-24 paper/main.tex
Updated references.

@@
 RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
+We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS \cite{desautels2012parallelizing, Rasmussen2004}.
 Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.

### 2021-03-24 paper/methods.tex
Updated references.

@@
 \textbf{Automated colony picking and culturing.}
-A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates.
+A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates using the 490-510nm (cyan) light filter .
 Each selected colony was used to inoculate 1mL of selective medium in a 2mL square well 96 plate.

### 2021-03-25 paper/main.tex
Update results section.

@@
 Here, we present a machine learning based workflow for design of one small genetic control element, a bacterial ribosome binding site.
-We have used Gaussian Process for prediction of a relative Transcription Initiation Rates and Upper Confidence Bound Bandit algorithm for recommendation of genetic designs to be tested in vivo.
+We have used Gaussian Process Regression for prediction of a relative Transcription Initiation Rates and Upper Confidence Bound based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
 By tying use of the algorithms with laboratory automation and high-throughput processes we have created a robust workflow for design of custom bacterial binding sites.
-Results sentence.
+With our workflow, we have generated a novel library of ribosome binding sites, with X of them falling within the +/- \% 10 range from our benchmark translation initiation rate.
 This work has also elucidated some design guidelines, including the required edit distance between two sequences and favourable representation of the analysed DNA sequence.
@@
 
-Figure 1 shows the DBTL cycle with the machine learning process emphasised.
-Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase.
-In LEARN phase we use the Gaussian Process algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
-Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the learn phase, which we find a more natural starting point for a machine learning centred workflow.\\
+Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. 
+In LEARN phase we use the Gaussian Process regression algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
+Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the Learn phase, which we find a more natural starting point for a machine learning centred workflow.\\
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.7]{plots/flowchart.pdf}
+    \caption{Flowchart of machine learning based experimental design.}
+    \label{fig: Flowchart}
+\end{figure}
 
@@
 In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
-Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic.
-\\
-For the zeroth round, since we don't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine} and the Gaussian Process regression method (see \textbf{Methods} section) \cite{srinivas2012information}.
-This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
-The label values are between 0 - 100,000 and skewed, which is shown in Figure xx. First, we have normalised the label to 0 - 1 using the min-max normalisation. 
-The predictions are shown in \textcolor{red}{Figure Xb}\\
-The predictions for next rounds were predicted similarly, but with use of data obtained in previous rounds, visualisation of prediction for the last round of designs is shown in \textcolor{red}{Figure Xc} \\
+Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
 
@@
 There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992,SHULTZABERGER2001}.
-In \emph{E.
-coli} the RBS is usually located in the 20 bases upstream of the start codon.
-The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E.
-coli} is AGGAGG.
+In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
+By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
+Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
+By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
+In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
+The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
 Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\
@@
 
-In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant.
-By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived TIR by measuring fluorescence of cells harbouring plasmid with the device.
-Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1.
-By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
 The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. 
-In our design we focus on randomising of the core -8 to -13 (relative to the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
+In our design we focus on randomising of the core -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
 TTTAAGA + NNNNNN + TATACAT.
-Since for each of the 6 position there are 4 possibilities: A, C, G, T the total variant space is $4^6$ = 4096.\\
+Since for each of the 6 position there are 4 possibilities: A, C, G, T the total experimental (variant) space is $4^6$ = 4096.\\
+To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
+To fit into our automated workflow, we have divided the 450 variants into batches of 90.
+In the zeroth round we have tested two batches of designs, for total of 180 variants. 
+Since at that stage we didn't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine} and the Gaussian Process regression method (see \textbf{Methods} section) \cite{srinivas2012information}.
+This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
+The label values are between 0 - 100,000 and skewed, which is shown in Figure xx. First, we have normalised the label to 0 - 1 using the min-max normalisation. 
+The predictions are shown in \textcolor{red}{Figure Xb}\\
+The predictions for next rounds were predicted similarly, but with use of data obtained in previous rounds, visualisation of prediction for the last round of designs is shown in \textcolor{red}{Figure Xc} \\
+
 Since there was no prior data that we could have used to guide our design for the zeroth round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:

### 2021-03-25 paper/main.tex
Updated results section.

@@
 
-Design of small genetic control elements like promoters, binding sites and terminators is currently challenging.
-To a large extent genetic engineers are relying on parts databases and otherwise known sequences to inform designs.
-This means, however, that because the known sequences may not be optimal for the desired output, optimal expression of constructs may not be achieved.
-This problem could be solved by a tailored design of these sequences, however, there are no reliable rules for such design.
-Here, we present a machine learning based workflow for design of one small genetic control element, a bacterial ribosome binding site.
+We consider the problem of designing DNA sequences corresponding to genetic control elements. 
+In contrast to traditional ways, such as relying on parts already known to genetic designers, we use machine learning to tailor designs for optimal expression.
+To this end, we created a machine learning based workflow for design of one small genetic control element, a bacterial ribosome binding site.
 We have used Gaussian Process Regression for prediction of a relative Transcription Initiation Rates and Upper Confidence Bound based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
 By tying use of the algorithms with laboratory automation and high-throughput processes we have created a robust workflow for design of custom bacterial binding sites.
-With our workflow, we have generated a novel library of ribosome binding sites, with X of them falling within the +/- \% 10 range from our benchmark translation initiation rate.
+With our workflow, we have generated a novel library of ribosome binding sites, with X of them falling within the +/- \% 10 range from our benchmark site's translation initiation rate.
 This work has also elucidated some design guidelines, including the required edit distance between two sequences and favourable representation of the analysed DNA sequence.
@@
 
-\subsection{LEARN - prediction of RBS performance}
-First problem to be tackled was how to embed the sequence to give our RBS sequences numerical form.
-As we will be using a Gaussian Process for regression we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}.
-The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. 
-We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. 
-Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
-More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-This approach allowed for good balance between computational complexity and performance.\\
-The predictions were made using the Gaussian Process regression (GPR) algorithm. 
-In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
-Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
-
 \subsection{DESIGN of the genetic device}
@@
 This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
-The label values are between 0 - 100,000 and skewed, which is shown in Figure xx. First, we have normalised the label to 0 - 1 using the min-max normalisation. 
-The predictions are shown in \textcolor{red}{Figure Xb}\\
-The predictions for next rounds were predicted similarly, but with use of data obtained in previous rounds, visualisation of prediction for the last round of designs is shown in \textcolor{red}{Figure Xc} \\
-
-Since there was no prior data that we could have used to guide our design for the zeroth round of experiments, we have designed 180 (two times 90) RBS sequences based on the consensus sequence that we expected to give good cover of the experimental space:
+This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.\\
+Apart from the abovementioned algorithm recommnded sequences, we have also decided to try some random sequences for the total of 180 sequences:
 
@@
     \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
-    \item 30 RBS sequences randomised based on the position probability matrix (PPM) \textcolor{red}{we need a citation and better explanation for this}  
-    \item 60 RBS sequences recommended by our multi armed bandit recommendation algorithm based on the initial literature, as described below.
+    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occuring RBS sequences in \emph{E. coli} genome
+    \item 60 RBS sequences recommended by our multi armed bandit recommendation algorithm based on the initial literature, as described above.
 \end{enumerate}{}
 
-\begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.7]{plots/TIR_histogram.pdf}
-    \caption{TIR Histogram.}
-    \label{fig: TIR Histogram.}
-\end{figure}
+In the subsequent rounds, data obtained in the previous ones have been used to teach the algorithm. 
+These subsequent rounds were run in batches of 90 each. \\
 
-\begin{enumerate}
-    \item RMSE of predictions.
+\subsection{BUILD and TEST}
 
-    \item Similarity of recommendations. 
-    \begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.7]{plots/similarity_first_round_recommendation.pdf}
-    \caption{TIR Histogram.}
-    \label{fig: TIR Histogram.}
-\end{figure}
-\end{enumerate}{}
+For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
+These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and one have to cover 5-10\% of that space.
+And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
+To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
+This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
+Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
+In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
+The plasmids were then transformed and the resulting transformants were tested using microplate reader.
+Vast majority of reactions were prepared using liquid handling equipment.
+Similarly, colony picking was done by automated colony picker.\\
 
-The bandit algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
+\subsection{LEARN - prediction of RBS performance}
+
+First problem to be tackled was how to embed the sequence to give our RBS sequences numerical form.
+As a Gaussian Process regression has been used, we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}.
+The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. 
+We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. 
+Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
+More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+This approach allowed for good balance between computational complexity and performance.\\
+
+The predictions were made using the Gaussian Process regression (GPR) algorithm. 
+In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
+Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
+
+The recommendations were made by Upper Confidence Bound based Bandit algorithm. 
+This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
 As such, it is very useful in solving problems like the one presented here - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
+
 In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space.
@@
 In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.
-
-\subsection{BUILD and TEST}
-
-For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
-These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and one have to cover 5-10\% of that space.
-And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
-To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
-This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
-Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
-In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly.
-The plasmids were then transformed and the resulting transformants were tested both using microplate reader and flow cytometry.\\
-Figure X shows the results of the zeroth and all the subsequent rounds.\\
 

### 2021-03-25 paper/main.tex
Update Results section.

@@
 
-First problem to be tackled was how to embed the sequence to give our RBS sequences numerical form.
+The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
+Since machine learning is just starting to be used in synthetic biology, there is no agreed approach to any of these problems.\\
+In our case, the problem of feature representation was how to embed the DNA sequence to give our RBS sequences numerical form. 
 As a Gaussian Process regression has been used, we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. 
@@
 
-The recommendations were made by Upper Confidence Bound based Bandit algorithm. 
+The recommendations were made using the Multi-armed Bandit algorithm.
+In short, the this algorithm is a stochastic method of probing of the experimental space. 
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-As such, it is very useful in solving problems like the one presented here - a big pool of potential designs, but only limited time and money that can be used for finding the optimal one.\\
-
-In short, the multi-armed bandit algorithm is a stochastic method of probing of the experimental space.
-In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computer by the prediction algorithm (GPR).
+In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (GPR).
 Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation. 
-Exploration makes the algorithm to recommend designs that will improve the predictions better, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest.
+Exploration makes the algorithm recommend designs that will improve the predictions, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest. 
 The two approaches can be controlled with the \textbeta\enspace parameter. 
@@
 One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). 
-As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.\\
+As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.
 \\
-For rounds beyond the initial, zeroth one, the designs were recommended only by the bandit algorithm based on the data obtained in the previous one, without adding any random designs.
-In total, three more rounds beyond the initial one have been performed, each with designs recommended by the algorithm.
 
@@
 In our case we have performed a total of four rounds, with performance of the RBSes increasing as seen in \textcolor{red}{Figure X}.\\
+
+\subsection{Rules of RBS design}
 

### 2021-03-30 paper/main.tex
Updated abstract.

@@
 
-\title{Machine Learning based Translation Initiation Rate predictor}
+\title{Machine Learning guided workflow for Ribosome Binding Site engineering}
 \author{Zhang M., Holowko M. B., Hayman Zumpe H., Ong, C. S.}
@@
 
-We consider the problem of designing DNA sequences corresponding to genetic control elements. 
-In contrast to traditional ways, such as relying on parts already known to genetic designers, we use machine learning to tailor designs for optimal expression.
-To this end, we created a machine learning based workflow for design of one small genetic control element, a bacterial ribosome binding site.
-We have used Gaussian Process Regression for prediction of a relative Transcription Initiation Rates and Upper Confidence Bound based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
-By tying use of the algorithms with laboratory automation and high-throughput processes we have created a robust workflow for design of custom bacterial binding sites.
-With our workflow, we have generated a novel library of ribosome binding sites, with X of them falling within the +/- \% 10 range from our benchmark site's translation initiation rate.
-This work has also elucidated some design guidelines, including the required edit distance between two sequences and favourable representation of the analysed DNA sequence.
+Fine control of gene expression is extremely important for engineered biological systems.
+Such control can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
+Unfortunately, our understanding of how RBSs perform in vivo is insufficiently well understood to allow predictable, reliable design of RBSs at the level of finesse needed for some applications.
+To address this problem, we have created a machine learning (ML) enabled workflow for the design of bacterial RBSs.
+We used Gaussian Process Regression for prediction of relative transcription initiation rates and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
+The Bandit algorithm addresses the exploration-exploitation trade-off, balancing the search for highly novel sequences with the desire for strong RBSs.
+We have integrated the ML algorithms with laboratory automation and high-throughput processes, creating a robust workflow for the design of custom RBSs.
+By employing these techniques, we were able to increase the reliability and reproducibility of results and increase the confidence in the design process.
+Using our workflow, we generated a novel library of diverse RBSs with a wide range of expression levels.
+Notably, a high number of these sites demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs.
+Additionally, this work elucidated some design guidelines, including the favourable level of difference between two sequences and efficient numerical representation of the analysed DNA/RNA sequence.
+In summary, by employing both machine learning and high-throughput laboratory methods we have created a workflow for creation of small genetic devices enabling efficient generation of parts with required characteristics.
+As a next step, we hope to expand this workflow to more complicated parts including promoters and terminators.
 

### 2021-03-31 paper/main.tex
Updated Results and Summary.

@@
 Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used\cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TRI) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
 However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
@@
 Since machine learning is just starting to be used in synthetic biology, there is no agreed approach to any of these problems.\\
-In our case, the problem of feature representation was how to embed the DNA sequence to give our RBS sequences numerical form. 
-As a Gaussian Process regression has been used, we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. 
-The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. 
-We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. 
-Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
-More specifically, we used a summary of three kernels: spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-This approach allowed for good balance between computational complexity and performance.\\
+In our case, the problem of feature representation was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions. 
+Kernel functions implicitly embed DNA sequences into high-dimensional feature space and capturing similarities between data points.  
+%Gaussian Process regression has been used, we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. 
+% The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. 
+% We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. 
+% Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
+% TODO: citation needed. 
+In our case, we consider string kernels \cite{}, which takes two DNA sequences as input and output a real value which represents the similarity between those two sequences. 
+The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
+We review the design of the above string kernel in Appendix \ref{}.
+For Round 0, we chose to use the spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+For other rounds, we chose to use the weighted degree kernel with shifts. \\
+% TODO: reasoning needed. 
+% This approach allowed for good balance between computational complexity and performance.\\
 
@@
 
-\subsection{Rules of RBS design}
+\subsection{Characteristics of the obtained library and sequences}
+
+A genetic part library should display a number of characteristics to be deemed useful.
+First quality expected from a good library is a wide range of relevant trait values (in our case it is the TIR) so that relevant genetic devices can be appropriately optimised.
+Second important feature of a library is the diversity of offered sequences, which ensures that sequences compatible with different assembly methods and plasmids are available. 
+Additionally, it is very beneficial for the library to show a number of parts with similar numerical values for relevant trait, but with different sequences, so that a designer would have to settle on suboptimal part due to, for example, assembly incompatibility.
+Finally, a good library has to be reliable - if the numbers provided in the library are unreliable designers will be reluctant to use.
+Figure X shows how library conforms to these requirements.
+In short, due to focusing on exploitation distribution of our TIRs is skewed toward strong ones, but there is still a significant amount of RBSs with medium and low strengths.
+High diversity of our sequences is shown by lack of relative lack of bias in position weight matrices for different bins.
+Finally, we were able to keep the average standard deviation for each bin within XX\% with 6 biological replicates for each sample, which we find satisfactory for reliability.\\
+Our library is also showing two characteristics that may be useful in future RBS design efforts.
+One of them is the perceived editing distance between two sequences required for meaningful improvement in the TIR. 
+We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
+Figure X shows what edit distance is required for positive change in TIR for RBS with high and medium TIR.
+For high TIR RBS, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
+For medium TIR RBS, a distance of just 1 is enough to produce a meaningful increase in TIR.
+That means that as the TIR of examined RBSs increases, the algorithm should be "encouraged" to explore sequences which are more dissimilar to the current candidates.
+This finding could also indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs - the low rate of natural mutations will be too slow to discover through more dissimilar sequences \cite{Lee2012}.
+In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or 5 changes in the RBS region might be prohibitively long.
+In such cases, a directed process should be strongly encouraged.\\
+Another feature noticeable in our library is prevalence of specific patterns.
+These include X, Y and Z (Figure X). 
+Interestingly, RBSs containing these patterns are not detected as capable of binding the relevant ribosome anti-sequence by binding energy calculations \cite{Mann2017}.
+At present, there is no strong evidence why these patterns might by permissive for high TIRs, nevertheless, it is important to note that the used algorithm does not require that knowledge to effectively find them.\\
 
 \section{Summary}
+
+In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
+We have used Gaussian Process regression to predict the shape of our function and Upper Confindence Bound Bandit algorithm to recommend sequences to be tested.
+We have investigated a number of methods of digitising the DNA sequence, finally settling on Spectrum Kernel with Shift method, which fit into our regression method.
+We performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
+By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.\\
+There is a number of issues that we would like to address in the future.
+The biggest of them is elucidating the reason for none of our sequences being able to significantly cross the TIR of our benchmark sequence.
+Essentially, there are two possible reasons for this, one is due to the algorithm and one due to nature.
+If the problem is with the algorithm, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
+How to solve this?
+It might be also possible that the TIR seen in the benchamark is simply already the strongest possible and represents a natural ceiling for it. 
+In that case, change in the algorithm will not help.
 
@@
 have designed and performed the laboratory experiments.
-All authors contributed to and reviewed the manuscript.
+All authors analysed the data and contributed to and reviewed the manuscript.
 

### 2021-03-31 paper/methods.tex
Update results section.
Updated Results and Summary.

@@
 The plates were tested in Cytation5 microplate reader.
-Cytation 5 acquisition and incubation/shaking settings
+Cytation 5 acquisition and incubation/shaking settings were as follows: length of run: 8h; interval: 10 min; continuous orbital shake at 237 cpm and slow orbital speed; excitation wavelength: 490/10mm; emission wavelength: 515/10 mm; bottom read; gain: 60; read height: 7mm; read speed: Sweep.
 
@@
 
-We show the flowchart of machine learning based experimental design in Figure \ref{fig: flowchart of machine learning based experimental design.}. 
+The flowchart of machine learning based experimental design is shown in Figure \ref{fig: Flowchart}. 
 To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.7]{plots/flowchart.pdf}
-    \caption{Flowchart of machine learning based experimental design.}
-    \label{fig: flowchart of machine learning based experimental design.}
-\end{figure}
-
-\subsubsection{LEARN: Gaussian Process Regression with Weighted Degree Kernel}
+\subsubsection{LEARN: Gaussian Process Regression with String Kernel}
 
@@
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values at each point in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon_i$, for all $i \in \mathcal{S}$, where $\epsilon_i$ is the noise (we assume that the noise is following Gaussian distribution with unknown mean and variance).
+$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}$, where $\epsilon$ is the noise (we assume that the noise is following Gaussian distribution with unknown mean and variance).
 This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g.

### 2021-04-06 codes/data_generating.py
Round 1-3 pipeline (1/4) data generating reconstruct.

@@ -17,19 +17,23 @@ import pickle
 
 # input parameters
 parser = argparse.ArgumentParser(description='Generate valid data.')
-parser.add_argument('partial_rep_flag', default= 'True', help = 'indicates whether only using partial replicates (indicating in Replicate col). True for using partial reps.')
 parser.add_argument('normalize_flag', default= 'True', help = 'indicates whether normalize labels. True for normalized label.')
-parser.add_argument('Format', default='Seq', help = 'Seq for rows as sequences; Sample for rows as samples')
+parser.add_argument('to_design_round', default= '3', help = 'specify the round to design.')
 
 args = parser.parse_args()
 normalize_flag = str(args.normalize_flag) # str True or False
-Use_partial_rep = str(args.partial_rep_flag) # str True or False
-data_format = str(args.Format) # str Seq or Sample
+to_design_round = str(args.to_design_round) # 0-3
 
-# settings
+# other settings
+Use_partial_rep = True  # Update: for all data points, we uses replicates 1-6
+# parser.add_argument('partial_rep_flag', default= 'True', help = 'indicates whether only using partial replicates (indicating in Replicate col). True for using partial reps.')
+# str(args.partial_rep_flag) 
+data_format = 'Seq' # Update: we only use seq format in downstream tasks, so constrained to 'Seq'
+# str(args.Format) # str Seq or Sample
+# parser.add_argument('Format', default='Seq', help = 'Seq for rows as sequences; Sample for rows as samples')
 how_to_normalize = 'roundRep' # choices:  'plateRep', 'roundRep'
-# TODO: if one want to use plateRep, consensus sequence needs to be added for second plate
-COMPLETE_REP_SET = {'1','2','3','4','5','6','7','8','9'}
+# WARNING: if one want to use plateRep, consensus sequence needs to be in each plate
+COMPLETE_REP_SET = {'1','2','3','4','5','6'}
 sheet_name = 'Microplate' # for masterfile
 Log_flag = True # indicates whether take log label
 Norm_method = 'mean' # indicates how to normalize label (one of 'mean', 'minmax', None)
@@ -42,9 +46,9 @@ Predictions_path = '/data/Designs/design_pred.xlsx'
 idx_seq_path = '/data/idx_seq.pickle'
 
 if normalize_flag == 'True':
[…]

### 2021-04-08 codes/data_generating.py
update data and plots for round 3, rep 1

@@ -22,7 +22,7 @@ parser.add_argument('to_design_round', default= '3', help = 'specify the round t
 
 args = parser.parse_args()
 normalize_flag = str(args.normalize_flag) # str True or False
-to_design_round = str(args.to_design_round) # 0-3
+to_design_round = str(args.to_design_round) # 0-4
 
 # other settings
 Use_partial_rep = True  # Update: for all data points, we uses replicates 1-6
@@ -36,7 +36,7 @@ how_to_normalize = 'roundRep' # choices:  'plateRep', 'roundRep'
 COMPLETE_REP_SET = {'1','2','3','4','5','6'}
 sheet_name = 'Microplate' # for masterfile
 Log_flag = True # indicates whether take log label
-Norm_method = 'mean' # indicates how to normalize label (one of 'mean', 'minmax', None)
+Norm_method = 'minmax' # indicates how to normalize label (one of 'mean', 'minmax', None)
 
 #-------------------------------------------------------------------------------------------------------------
 # path 
@@ -66,7 +66,7 @@ def normalize(df, col_name):
         # When we were about to design round 2, 
         # we observed that the TIR of the same RBS in each round turns out to be very different.
         # So we subtract the mean of the reference sequence (the only one RBS tested repeated in different rounds) in each round.
-        if to_design_round in {'2', '3'}: 
+        if to_design_round in {'2', '3', '4'}: 
             ref_seq_mean = group.loc[group['Group'] == 'reference', col_name].stack().mean()
             print('col name: ', col_name)
             print('round: ', name)
@@ -106,7 +106,7 @@ def rename_group_names(df):
                         'bps_core':'BPS-C', 'bps_noncore': 'BPS-NC', 
                         'uni random': 'UNI', 'prob random': 'PPM', 
                         'bandit': 'Bandit-0', 'bandit2': 'Bandit-1',
-                        'bandit3': 'Bandit-2'})
+                        'bandit3': 'Bandit-2', 'bandit4': 'Bandit-3'})
     return df
 
 # add index

### 2021-04-08 paper/main.tex

@@
 In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
-We have used Gaussian Process regression to predict the shape of our function and Upper Confindence Bound Bandit algorithm to recommend sequences to be tested.
+We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
 We have investigated a number of methods of digitising the DNA sequence, finally settling on Spectrum Kernel with Shift method, which fit into our regression method.
@@
 How to solve this?
-It might be also possible that the TIR seen in the benchamark is simply already the strongest possible and represents a natural ceiling for it. 
+It might be also possible that the TIR seen in the benchmark is simply already the strongest possible and represents a natural ceiling for it. 
 In that case, change in the algorithm will not help.
+Another problem we would like to tackle is the not always good prediction of the TIR of a given RBS.
+In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.\\
+In the future, we hope to extend the algorithm to other, more complicated genetic elements.
+This could include promoters and terminators.
+The complexity of the task quickly increases with the length of the sequence.
+This is because the experimental space grows exponentially with the number of examined positions and so the space becomes increasingly hard to cover with experiments.
+To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
+Finally, we hope to see our RBS library being adapted by the community.
+We believe that its favourable characteristics will make if very useful to designers.
 

### 2021-04-08 paper/main.tex
Added characteristics figure.

@@
 Finally, we were able to keep the average standard deviation for each bin within XX\% with 6 biological replicates for each sample, which we find satisfactory for reliability.\\
+
+\begin{figure}[t]
+\centering
+\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}\par\medskip
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}
+\end{minipage}
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}
+\end{minipage}
+\begin{minipage}{.5\linewidth}
+\centering
+\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}
+\end{minipage}
+\caption{Characteristics of the sequences.}
+\label{fig:characteristics}
+\end{figure}
+
 Our library is also showing two characteristics that may be useful in future RBS design efforts.

### 2021-04-14 codes/maciek_code/TIR_calculator_all.py
Added Rep 2 for Plate 5.

@@ -65,15 +65,15 @@ def create_tir():
 full_report = 'no'
 path = '../../data/Plate_results/*.xlsx'
 ResFile = '../../data/Results_Masterfile.xlsx'
-Results = pd.read_excel(ResFile,sheet_name='Microplate',encoding="utf-8-sig") 
+Results = pd.read_excel(ResFile,sheet_name='Microplate') 
 print(Results.head())
 for File in glob.glob(path):
     
     print(File)
     
     if "Rep" in File :
-        OD = pd.read_excel(File,sheet_name='OD',encoding="utf-8-sig")
-        GFP = pd.read_excel(File,sheet_name='GFP',encoding="utf-8-sig")
+        OD = pd.read_excel(File,sheet_name='OD')
+        GFP = pd.read_excel(File,sheet_name='GFP')
         wb = ox.load_workbook(File, read_only=True)   # open an Excel file and return a workbook
     
         if 'GFPOD' not in wb.sheetnames:

### 2021-04-16 codes/batch_ucb.py
update prediction plots

@@ -77,10 +77,11 @@ class RBS_UCB():
         self.df_known = df_known
         self.df_known['train_test'] = 'Train'
         self.known_rbs_set = set(self.df_known['RBS'])
-        if df_design is None:
-            self.df_design = self.generate_design_space()
-        else:
-            self.df_design = df_design
+        # if df_design is None:
+        #     self.df_design = self.generate_design_space()
+        # else:
+        #     self.df_design = df_design
+        self.df_design = self.generate_design_space(df_design)
         self.df_design['train_test'] = 'Test'
         self.df_train_test = pd.concat([self.df_known, self.df_design], sort = True) #.reset_index()
         self.df_train_test = self.df_train_test.set_index('idx')
@@ -115,29 +116,33 @@ class RBS_UCB():
         self.beta = beta # ucb =  mean + beta * std
         # TODO: the choice of beta is critical
 
-    def generate_design_space(self):
+    def generate_design_space(self, df_design):
         # create all combos
 
-        combos = [] # 20-base
-        combos_6 = [] # 6-base
+        if df_design is None:
+            combos = [] # 20-base
+            combos_6 = [] # 6-base
 
-        # Setting
-        char_sets = ['A', 'G', 'C', 'T']
-        design_len = 6
-        pre_design = 'TTTAAGA'
-        pos_design = 'TATACAT'
+            # Setting
+            char_sets = ['A', 'G', 'C', 'T']
+            design_len = 6
+            pre_design = 'TTTAAGA'
+            pos_design = 'TATACAT'
 
-        for combo in itertools.product(char_sets, repeat= design_len):
-            combo = pre_design + ''.join(combo) + pos_design
-            combos_6.append(''.join(combo))
-            combos.append(combo)
-            
-        assert len(combos) == len(char_sets) ** design_len
+            for combo in itertools.product(char_sets, repeat= design_len):
+                combo = pre_design + ''.join(combo) + pos_design
+                combos_6.append(''.join(combo))
+                combos.append(combo)
+                
+            assert len(combos) == len(char_sets) ** design_len
 
-        df_design = pd.DataFrame()
-        df_design['RBS'] = [x for x in combos if x not in self.known_rbs_set]
+            df_design = pd.DataFrame()
+       
+            df_design['RBS'] = [x for x in combos if x not in self.known_rbs_set]
+        
         df_design['RBS6'] = df_design['RBS'].str[7:13]
[…]

### 2021-04-16 codes/regression.py

@@ -44,7 +44,7 @@ KERNEL_DICT = {
     'RBF': RBF  
 }
 
-REP_LIST = ['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5','Rep6', 'Rep7', 'Rep8','Rep9']
+REP_LIST = ['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5','Rep6']
 
 
 class GPR_Predictor():

### 2021-04-16 codes/sanity_check.py

@@ -1,0 +1,32 @@
+# sanity check for reproducibility 
+# direct to proper path
+import os
+import sys
+module_path = os.path.abspath(os.path.join('../'))
+if module_path not in sys.path:
+    sys.path.append(module_path)
+
+import numpy as np
+import pandas as pd
+
+def check_dataset(data1, data2):
+    for index1, row1 in data1.iterrows():
+        for index2, row2 in data2.iterrows():
+            if row1['RBS'] == row2['RBS'] and row1['RBS'] != 'TTTAAGAAGGAGATATACAT':
+                print(row1['RBS'])
+                # print('ave diff: ', np.abs(row1['AVERAGE'] - row2['AVERAGE']))
+                # print('std diff: ', np.abs(row1['STD'] - row2['STD']))
+                
+                # Instead of using average, we use a single replicate to check
+                # Since AVERAGE calculation in round 1 and 2 were wrong 
+                assert np.abs(row1['Rep6'] - row2['Rep6']) < 1e-3
+                # assert np.abs(row1['STD'] - row2['STD'])< 1e-3
+                break
+    
+source1 = "data/pipeline_data/Results_Microplate_partialTrue_normTrue_mean_roundRep_formatSeq_logTrue_Round2.csv"
+# source2 = "/home/admin-u6015325/ownCloud/SynBio_repo_designs/SynbioML-round1/data/firstRound_Microplate_normTrue_formatSeq_logTrue.csv"
+source2 = "/home/admin-u6015325/ownCloud/SynBio_repo_designs/SynbioML-round2/data/Results_Microplate_partialTrue_normTrue_roundRep_formatSeq_logTrue.csv"
+check_dataset(pd.read_csv(source1), pd.read_csv(source2))
+
+# Log: round 1-3 passed the test (tested rep6 only)
+# REVIEW: need to be careful about round 2, whether used the "correct" set of replicates to train?

### 2021-04-26 paper/main.tex

@@
 \usepackage{algorithmic}
-\usepackage{subfig}
 \usepackage{hyperref}
@@
 \usepackage{caption}
-%\usepackage{subcaption}
+\usepackage{subcaption}
+% \usepackage{subfigure}
 
@@
 Predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
-There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}.\\
+There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} .
+This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .\\
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
@@
 
-\begin{figure}[t]
+\begin{figure}
      \centering
-\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}\par\medskip
-\begin{minipage}{.5\linewidth}
+     \begin{subfigure}[b]{0.49\textwidth}
          \centering
-\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}
-\end{minipage}
-\begin{minipage}{.5\linewidth}
+         \includegraphics[scale=1.2]{plots/logo.pdf}
+         \caption{$y=x$}
+         \label{fig:y equals x}
+     \end{subfigure}
+     \hfill
+     \begin{subfigure}[b]{0.49\textwidth}
          \centering
-\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}
-\end{minipage}
-\begin{minipage}{.5\linewidth}
+         \includegraphics[scale=0.5]{plots/Patterns.png}
+         \caption{$y=3sinx$}
+         \label{fig:three sin x}
+     \end{subfigure}
+     \vskip\baselineskip
+     \begin{subfigure}[b]{0.49\textwidth}
          \centering
-\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/Hd_Heatmap.png}}
-\end{minipage}
-\caption{Characteristics of the sequences.}
-\label{fig:characteristics}
+         \includegraphics[scale=0.5]{plots/Hd_Heatmap.png}
+         \caption{$y=5/x$}
+         \label{fig:five over x}
+     \end{subfigure}
+     \hfill
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \includegraphics[scale=0.5]{plots/Histogram.png}
+         \caption{$y=5/x$}
+         \label{fig:five over x}
+     \end{subfigure}
+        \caption{Three simple graphs}
+        \label{fig:three graphs}
 \end{figure}

### 2021-04-26 paper/main.tex
Updated paper - redid according to Nature Comms requirments, updated Supplementary.

@@
 
-\section{Abstract}
+\section*{Abstract}
 
-Fine control of gene expression is extremely important for engineered biological systems.
-Such control can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
-Unfortunately, our understanding of how RBSs perform in vivo is insufficiently well understood to allow predictable, reliable design of RBSs at the level of finesse needed for some applications.
+Fine control of gene expression can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
+Unfortunately, RBSs are not understood at the level of finesse required for reliable design. 
 To address this problem, we have created a machine learning (ML) enabled workflow for the design of bacterial RBSs.
-We used Gaussian Process Regression for prediction of relative transcription initiation rates and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
-The Bandit algorithm addresses the exploration-exploitation trade-off, balancing the search for highly novel sequences with the desire for strong RBSs.
+We used Gaussian Process Regression for prediction and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
 We have integrated the ML algorithms with laboratory automation and high-throughput processes, creating a robust workflow for the design of custom RBSs.
-By employing these techniques, we were able to increase the reliability and reproducibility of results and increase the confidence in the design process.
 Using our workflow, we generated a novel library of diverse RBSs with a wide range of expression levels.
 Notably, a high number of these sites demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs.
-Additionally, this work elucidated some design guidelines, including the favourable level of difference between two sequences and efficient numerical representation of the analysed DNA/RNA sequence.
-In summary, by employing both machine learning and high-throughput laboratory methods we have created a workflow for creation of small genetic devices enabling efficient generation of parts with required characteristics.
-As a next step, we hope to expand this workflow to more complicated parts including promoters and terminators.
+Additionally, this work elucidated some important RBS design guidelines.
 
@@
 The results were then fed back to the algorithm for it to recommend the next round of designs.\\
-
-\input{methods}
 
@@
 
+\input{methods}
+
 \section{Contributions}
@@
 \appendix
-\input{appendix}
+\input{supplementary}
 \end{document}

### 2021-04-26 paper/supplementary.tex
Updated paper - redid according to Nature Comms requirments, updated Supplementary.

--- 2020-11-22 paper/appendix.tex
@@
-\section*{Appendix}
+\textbf{Supplementary}
 
-\section{Reproducible Plots}
+\section{Extended Machine Learning Discussion}
+
+\subsection{Reproducible Plots}
 In this section, we put reproducible plots and will update the plots when we get full results.
@@
     \centering
-    \includegraphics[scale = 0.4]{plots/swarmplots.pdf}
+    \includegraphics[scale = 0.4]{plots/Supplementary/swarmplots.pdf}
     \caption{Swarmplots for different groups, with raw (left) and normalised (right) TIR labels (averaged over 6 replicates). 
@@
     \centering
-    \includegraphics[scale = 0.3]{plots/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
-    \includegraphics[scale = 0.3]{plots/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
+    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
+    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
     \caption{Kernel Heatmap and Predictions.
@@
     \centering
-    \includegraphics[scale = 0.25]{plots/clustering.png}
+    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
     \caption{TSNE of RBS sequences in design space with clustering.
@@
 
-\section{Gaussian Process Regression}
+\subsection{Gaussian Process Regression}
 
@@
 
-\section{Choices of Kernels}
+\subsection{Choices of Kernels}
 
@@
 
-\subsection{Normalisation of Kernel}
+\subsubsection{Normalisation of Kernel}
 
@@
 
-\section{Batch Recommendation}
+\subsection{Batch Recommendation}
 
@@
 
-\section{Design Pipeline}
+\subsection{Design Pipeline}
 
@@
 
-\section{Intuition behind UCB and visualisation}
+\subsection{Intuition behind UCB and visualisation}
 
@@
 
-\section{Result analysis}
+\subsection{Result analysis}
 
@@
 
-\section{Figures}
+\subsection{Statistics of the core vs non-core regions of the RBS}
+
+In our study, we have tested a set of sequnces designed to confirm the notion that changing bases within the RBS core (6 bases) is statistically more influencing the TIR than the changes made outside the core.
+This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
+For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
+Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS.
+This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
+The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
+    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} The core region is highligthed in light green and the lines are rolling averages for each base.
+The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base, which is colour coded.}
+    \label{fig:core_vs_noncore}
+\end{figure}
+
+\section{Supplementary Figures}
+
+\begin{figure}
+     \centering
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure GGGGGG.pdf}
+         \caption{$y=x$}
+         \label{fig:GGGGGG}
+     \end{subfigure}
+     \hfill
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure GGGGGC.pdf}
+         \caption{$y=3sinx$}
+         \label{fig:GGGGGC}
+     \end{subfigure}
+     \vskip\baselineskip
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure CGGGAC.pdf}
+         \caption{$y=5/x$}
+         \label{fig:CGGGAC}
+     \end{subfigure}
+     \hfill
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure CACGTT.pdf}
+         \caption{$y=5/x$}
+         \label{fig:CACGTT}
+     \end{subfigure}
+        \caption{\textbf{Folding predictionds for four different RBS.} Here we show the predicted, energetically most favourable structures for four of our RBS with the following cores: GGGGGG, GGGGGC, CGGGAC, CACGTT and their immediate upstream and downstream background sequence.
+There is no discerbile and consistent difference between strong ones (the first two) and weak ones (the last two).}
+        \label{fig:structures}
+\end{figure}

### 2021-04-27 paper/main.tex
Updated paper.

@@
 \usepackage{subcaption}
+\usepackage{authblk}
 % \usepackage{subfigure}
@@
 \title{Machine Learning guided workflow for Ribosome Binding Site engineering}
-\author{Zhang M., Holowko M. B., Hayman Zumpe H., Ong, C. S.}
+
+\author[1,2]{Zhang M.}
+\author[3]{Holowko M. B.}
+\author[3]{Hayman Zumpe H.}
+\author[1]{Ong, C. S.}
+\affil[1]{AIML FSP}
+\affil[2]{ANU}
+\affil[3]{CSIRO Synthetic Biology Future Science Platform, CSIRO Land and Water, Brisbane, QLD 4102, Australia}
+
 \date{\today{}}
@@
 
-\section{Summary}
+\section{Discussion}
 
@@
 
-\section{Contributions}
+\section*{Contributions}
 Zhang M.
@@
 
+\section*{Competing interests}
+The authors declare no competing interests
+
+\section*{Acknowledgments}
+The authors would like to acknowlege CSIRO's AI/ML and Synthetic Biology Future Science Platforms for providing funding for this research.
+
 \newpage

### 2021-04-28 codes/data_generating.py
update with all reps in plate 5

@@ -41,7 +41,7 @@ import pickle
 
 # Setting 
 # Specify your choice
-approach = 'c2'
+approach = 'bc2'
 to_design_round = '4'
 
 approach = approach.lower()

### 2021-04-28 codes/data_generating.py
add ratio normalisation, add uni

@@ -41,7 +41,7 @@ import pickle
 
 # Setting 
 # Specify your choice
-approach = 'bc2'
+approach = 'e1'
 to_design_round = '4'
 
 approach = approach.lower()
@@ -60,6 +60,8 @@ if 'c' in approach:
     Norm_method = 'mean' # indicates how to normalize label (one of 'mean', 'minmax', None)
 elif 'd' in approach:
     Norm_method = 'minmax'
+elif 'e' in approach:
+    Norm_method = 'ratio'
 else: # None
     Norm_method = None
 
@@ -140,6 +142,9 @@ def normalize(df, col_name):
             elif Norm_method == 'minmax':
                 # min-max normalization 
                 group[col_name] = (group[col_name] - group[col_name].min())/(group[col_name].max() - group[col_name].min())
+            elif Norm_method == 'ratio':
+                ref_seq_mean = group.loc[group['Group'] == 'reference', 'AVERAGE'].mean()
+                group[col_name] = group[col_name]/ref_seq_mean
             else:
                 assert Norm_method == None
         
@@ -154,6 +159,9 @@ def normalize(df, col_name):
             # min-max normalization 
             normalised_df[col_name] =\
                 (normalised_df[col_name] - normalised_df[col_name].min())/(normalised_df[col_name].max() - group[col_name].min())
+        elif Norm_method == 'ratio':
+            ref_seq_mean = group.loc[group['Group'] == 'reference', 'AVERAGE'].mean()
+            normalised_df[col_name] = normalised_df[col_name]/ref_seq_mean
         else:
             assert Norm_method == None
     

### 2021-04-29 paper/mixture_absGau.tex

\newpage

\section{Mixture of Absolute Gaussian}

The raw TIR scores in each round are positive and  have two modes. 
We assume they are mixture of two absolute Gaussian distributions. 
If we can fit the mixture of absolute Gaussian distributions, we can answer the following questions:
\begin{itemize}
    \item is the left peak decreasing in height with each round?
	\item is the left peak going left?
	\item Is the right peak increasing in height with each round?
	\item is the right peak going right?
\end{itemize}

The density of absolute Gaussian (also called \textit{folded normal distribution}) is 
\begin{align}
    f_{Y}\left(x ; \mu, \sigma^{2}\right)=\frac{1}{\sqrt{2 \pi \sigma^{2}}} e^{-\frac{(z-\mu)^{2}}{2 \sigma^{2}}}+\frac{1}{\sqrt{2 \pi \sigma^{2}}} e^{-\frac{(x+\mu)^{2}}{2 \sigma^{2}}}
\end{align}
for $x \geq 0$ and $0$ everywhere else. 
The mean and variance of the absolute Gaussian distribution is 
\begin{align}
    \mu_{Y} &=\sigma \sqrt{\frac{2}{\pi}} \exp \left(\frac{-\mu^{2}}{2 \sigma^{2}}\right)+\mu \operatorname{erf}\left(\frac{\mu}{\sqrt{2 \sigma^{2}}}\right);\\
    \sigma_{Y}^{2} &=\mu^{2}+\sigma^{2}-\mu_{Y}^{2},
\end{align}
where $\operatorname{erf}$ is the error function 
$\operatorname{erf} z=\frac{2}{\sqrt{\pi}} \int_{0}^{z} e^{-t^{2}} d t$.

We consider the model of the mixture of two absolute Gaussian distributions as

### 2021-05-05 codes/plot_dict.py
add plot_dict

@@ -1,0 +1,70 @@
+# create plot dictionaries:
+# keys are the group name in master file
+# values:
+#   name_dict: name of group show in plot
+#   color_dict: color of group 
+#   loc_dictt: location of group
+
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import cm
+from matplotlib.colors import ListedColormap, LinearSegmentedColormap
+import pickle
+
+name_dict = {}
+color_dict = {}
+loc_dict = {}
+
+name_dict['consensus'] = 'Consensus'
+name_dict['reference'] = 'Reference'
+name_dict['bps_core'] = 'BPS-C'
+name_dict['bps_noncore'] = 'BPS-NC', 
+name_dict['uni random'] = 'UNI'
+name_dict['prob random'] = 'PPM' 
+name_dict['bandit'] = 'Bandit-0'
+name_dict['bandit2'] = 'Bandit-1'
+name_dict['bandit3'] = 'Bandit-2'
+name_dict['bandit4'] = 'Bandit-3'
+
+
+loc_dict['consensus'] = 0
+loc_dict['reference'] = 0
+loc_dict['bps_core'] = 1
+loc_dict['bps_noncore'] = 2 
+loc_dict['uni random'] = 3
+loc_dict['prob random'] = 4
+loc_dict['bandit'] = 5
+loc_dict['bandit2'] = 6
+loc_dict['bandit3'] = 7
+loc_dict['bandit4'] = 8
+
+
+# Follow tutorial https://matplotlib.org/stable/tutorials/colors/colormap-manipulation.html
+viridis = cm.get_cmap('viridis', 5)
+tab10 = cm.get_cmap('tab10', 10)
+
+color_dict['consensus'] = tab10.colors[4]
+color_dict['reference'] = tab10.colors[5]
+color_dict['bps_core'] = tab10.colors[6]
+color_dict['bps_noncore'] = tab10.colors[7]
+color_dict['uni random'] = tab10.colors[8]
+color_dict['prob random'] = tab10.colors[9]
+color_dict['bandit'] = viridis.colors[0]
+color_dict['bandit2'] = viridis.colors[1]
+color_dict['bandit3'] = viridis.colors[2]
+color_dict['bandit4'] = viridis.colors[3]
+
+# for name in name_dict.keys():
+#     plt.scatter(loc_dict[name], loc_dict[name], color = color_dict[name])
[…]

### 2021-05-05 codes/plot_dict.py
pick tab10 to be different from viridis

@@ -38,17 +38,28 @@ loc_dict['bandit2'] = 6
 loc_dict['bandit3'] = 7
 loc_dict['bandit4'] = 8
 
+# loc_dict['consensus'] = 0
+# loc_dict['reference'] = 1
+# loc_dict['bps_core'] = 2
+# loc_dict['bps_noncore'] = 3 
+# loc_dict['uni random'] = 4
+# loc_dict['prob random'] = 5
+# loc_dict['bandit'] = 6
+# loc_dict['bandit2'] = 7
+# loc_dict['bandit3'] = 8
+# loc_dict['bandit4'] = 9
+
 
 # Follow tutorial https://matplotlib.org/stable/tutorials/colors/colormap-manipulation.html
 viridis = cm.get_cmap('viridis', 5)
 tab10 = cm.get_cmap('tab10', 10)
 
-color_dict['consensus'] = tab10.colors[4]
-color_dict['reference'] = tab10.colors[5]
-color_dict['bps_core'] = tab10.colors[6]
-color_dict['bps_noncore'] = tab10.colors[7]
-color_dict['uni random'] = tab10.colors[8]
-color_dict['prob random'] = tab10.colors[9]
+color_dict['consensus'] = tab10.colors[1]
+color_dict['reference'] = tab10.colors[3]
+color_dict['bps_core'] = tab10.colors[5]
+color_dict['bps_noncore'] = tab10.colors[6]
+color_dict['uni random'] = tab10.colors[7]
+color_dict['prob random'] = tab10.colors[8]
 color_dict['bandit'] = viridis.colors[0]
 color_dict['bandit2'] = viridis.colors[1]
 color_dict['bandit3'] = viridis.colors[2]
@@ -56,9 +67,14 @@ color_dict['bandit4'] = viridis.colors[3]
 
 # for name in name_dict.keys():
 #     plt.scatter(loc_dict[name], loc_dict[name], color = color_dict[name])
-# plt.xticks(list(loc_dict.values()), list(name_dict.values()))
+# # plt.xticks(list(loc_dict.values()), list(name_dict.values()))
 # # plt.title('viridis (0-3) + tab 10')
 # plt.show()
+
+# for i in range(10):
+#     plt.scatter(i, i, color = tab10.colors[i])
+# plt.show()
+
 plot_dict={}
 plot_dict['name'] = name_dict
 plot_dict['loc'] = loc_dict

### 2021-05-05 codes/plot_dict.py
generate plots for paper use

@@ -80,7 +80,4 @@ plot_dict['name'] = name_dict
 plot_dict['loc'] = loc_dict
 plot_dict['color'] = color_dict
 
-np.save('plot_dict.npy', plot_dict)
-
-# with open('../data/plot_dict.pickle', 'wb') as handle:
-#     pickle.dump(a, handle, protocol=pickle.HIGHEST_PROTOCOL)
+# np.save('plot_dict.npy', plot_dict)

### 2021-05-06 paper/main.tex
Updates from Overleaf

@@
 Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used\cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
+Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
 However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
-This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.\\
+This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
+The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\
 The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
 For example, according to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
-Predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
+Reported predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
 There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} .
@@
 However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. \\
-Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimization of protein level expression.
+Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimisation of associated protein level expression.
 RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-We have used Gaussian Process-Upper Confidence Bound and multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS \cite{desautels2012parallelizing, Rasmussen2004}.
+In this work We have used Gaussian Process Regression and Upper Confidence Bound multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS \cite{desautels2012parallelizing, Rasmussen2004}.
 Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
 We did this by designing a sequential experimental workflow, where designs in the zeroth round where randomised RBS sequences designed to explore the experimental space and some preliminary machine learning recommended designs based on literature.
-In the subsequent rounds, designs where recommended by the algorithm based on the data obtained in previous rounds. 
+In the subsequent rounds, designs where recommended by the algorithm based on the data obtained in the previous rounds. 
 The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
 After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
-The results were then fed back to the algorithm for it to recommend the next round of designs.\\
+The results were then fed back to the algorithm for it to recommend the next round of designs.
+This way, were able to build an extensive, reliable library of novel RBSs with diverse sequences.
+At the same time we were able to discover new RBS sequences with very high TIRs.\\
 
@@
 Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. 
-In LEARN phase we use the Gaussian Process regression algorithm to predict the behaviour of different RBS sequences and in the DESIGN phase we use the multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
-Conventionally, the Design phase is considered the beginning of the cycle, however we will start description of our results from the Learn phase, which we find a more natural starting point for a machine learning centred workflow.\\
+In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
+The following sections will discuss our machine learning workflow divided into specific parts of the DBTL cycle.\\
 
@@
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same optical density (OD\textsubscript{600}) with addition of IPTG.\\
+By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
 In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
-The RBS usually has a distinguishable, consensus, core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
+Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
 Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\

### 2021-05-06 paper/main.tex
Updated main text.

@@
 The results were then fed back to the algorithm for it to recommend the next round of designs.
-This way, were able to build an extensive, reliable library of novel RBSs with diverse sequences.
+This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
 At the same time we were able to discover new RBS sequences with very high TIRs.\\
@@
 
-Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised. Our machine learning workflow encompasses two phases of the cycle, namely the LEARN and DESIGN phase. 
-In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
-The following sections will discuss our machine learning workflow divided into specific parts of the DBTL cycle.\\
+The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
+We will discuss our workflow in terms of the Design-Build-Learn-Test cycle, where prediction will be discussed in the Design section and feature representation with prediction in the Learn one.\\
+Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised.
+Principally, machine learning is used in the LEARN and DESIGN phase and the other two are driven by choices made by human researchers. 
+In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.\\
 
@@
 \subsection{DESIGN of the genetic device}
-There is a number of factors that impact the protein expression rate, many of them concerned with how the ribosome recognises and binds to the RBS sequence \cite{Chen1994,Vellanoweth1992,SHULTZABERGER2001}.
 In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
@@
 By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
+
 In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
 Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (Figure \ref{fig:Anatomy}).\\
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig:Flowchart}).\\
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.6]{plots/RBS_anatomy.pdf}
-    \caption{\textbf{Diagram of the investigated sequence.} The sequence of the investigated RBS is shown with the randomised core sequence in red.
-The start codon of the GFP coding sequence is also shown with numbers showing relative nucleotide positions.}
-    \label{fig:Anatomy}
-\end{figure}
+Our template RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACA.
+This sequence is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. 
+Since this is the sequence against which new RBS sequenced will be benchmarked we will refer to this sequence as the benchmark sequence from now on.
+In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
+TTTAAGA + NNNNNN + TATACAT.
+We have experimentally confirmed that changing the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials). 
+Since for each of the 6 position there are 4 possibilities: A, C, G, T the total experimental (variant) space is $4^6$ = 4096.\\
 
-The investigated RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACAT, which is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. 
-In our design we focus on randomising of the core -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT.
-Since for each of the 6 position there are 4 possibilities: A, C, G, T the total experimental (variant) space is $4^6$ = 4096.\\
-To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
-To fit into our automated workflow, we have divided the 450 variants into batches of 90.
-In the zeroth round we have tested two batches of designs, for total of 180 variants. 
-Since at that stage we didn't have access to any prior data fitting our design we approximated the predictions using the data from \textcite{jervis2018machine} and the Gaussian Process regression method (see \textbf{Methods} section) \cite{srinivas2012information}.
-This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-This dataset has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.\\
-Apart from the abovementioned algorithm recommnded sequences, we have also decided to try some random sequences for the total of 180 sequences:
-
-\begin{enumerate}
-    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
-This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
-    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occuring RBS sequences in \emph{E. coli} genome
-    \item 60 RBS sequences recommended by our multi armed bandit recommendation algorithm based on the initial literature, as described above.
-\end{enumerate}{}
-
-In the subsequent rounds, data obtained in the previous ones have been used to teach the algorithm. 
-These subsequent rounds were run in batches of 90 each. \\
-
-\subsection{BUILD and TEST}
-
-For the machine learning to work effectively the analysed data-set needs to show two qualities: high relative volume and high quality of data.
-These two elements don't have a specific definitions, but in general the data-set have to be going into at least hundreds possible data points and one have to cover 5-10\% of that space.
-And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
-To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
-This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
-Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
-In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
-The plasmids were then transformed and the resulting transformants were tested using microplate reader.
-Vast majority of reactions were prepared using liquid handling equipment.
-Similarly, colony picking was done by automated colony picker.\\
-
-\subsection{LEARN - prediction of RBS performance}
-
-The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
-Since machine learning is just starting to be used in synthetic biology, there is no agreed approach to any of these problems.\\
-In our case, the problem of feature representation was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions. 
-Kernel functions implicitly embed DNA sequences into high-dimensional feature space and capturing similarities between data points.  
-%Gaussian Process regression has been used, we have investigated use of different types of kernels (also called covariances) for embedding \cite{Ben-Hur2008}. 
-% The additional, and positive, effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier. 
-% We compared performance of Dot Product, RBF and a number of string kernels: spectrum, mixed spectrum, weighted degree and weighted degree with shifting \textcolor{red}{Figure Xa}. 
-% Since we found that the spectrum kernel performed the best, we have used it in subsequent studies.
-% TODO: citation needed. 
-In our case, we consider string kernels \cite{}, which takes two DNA sequences as input and output a real value which represents the similarity between those two sequences. 
-The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
-We review the design of the above string kernel in Appendix \ref{}.
-For Round 0, we chose to use the spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-For other rounds, we chose to use the weighted degree kernel with shifts. \\
-% TODO: reasoning needed. 
-% This approach allowed for good balance between computational complexity and performance.\\
-
-The predictions were made using the Gaussian Process regression (GPR) algorithm. 
-In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
-Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
-
-The recommendations were made using the Multi-armed Bandit algorithm.
+The design recommendations were made using the Multi-armed Bandit algorithm.
 In short, the this algorithm is a stochastic method of probing of the experimental space. 
@@
 \\
+
+To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
+To fit into our automated workflow, we have divided the 450 variants into batches of 90.
+In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:
+
+\begin{enumerate}
+    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
+This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
+    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
+    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome
+    \item 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature, as described below.
+\end{enumerate}{}
+
+To generate algorithm driven designs for the 0th round we have used the data set available in \textcite{jervis2018machine}.
+This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.\\
+In the subsequent rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds. \\
+
+\subsection{BUILD and TEST}
+
+For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
+These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
+And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
+To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
+This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
+Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
+In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
+The plasmids were then transformed and the resulting transformants were tested using microplate reader.
+Vast majority of reactions were prepared using liquid handling equipment.
+Similarly, colony picking was done by an automated colony picker.\\
+
+\subsection{LEARN - prediction of RBS performance}
+
+For our predictions, we had to solve the problem of feature representation first.
+In our case, that problem was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions.\\
+Due to our use of the Gaussian Process Regression for predictions we have chosen kernel (covariance) functions for our embedding.
+Kernel functions implicitly embed DNA sequences into high-dimensional feature space which enables capturing similarities and distances between data points.  
+The additional effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier.\\
+
+There is a number of useful kernel functions that can be used for embedding of DNA.
+In our case, we use string kernels \cite{}, which take two DNA sequences as input and output a real value which represents the similarity between them. 
+The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
+We review the design of the above string kernels in Supplementary materials \ref{}.
+For Round 0, we chose to use the spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+For subsequent rounds, we chose to use the weighted degree kernel with shifts. \\
+% TODO: reasoning needed. 
+% This approach allowed for good balance between computational complexity and performance.\\
+
+The predictions were made using the Gaussian Process regression (GPR) algorithm. 
+In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
+Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
 

### 2021-05-07 paper/main.tex
Updates from Overleaf

@@
 
+\newcommand{\mengyan}[1]{\textcolor{magenta}{#1}}
+\newcommand{\maciej}[1]{\textcolor{blue}{#1}}
+
 \title{Machine Learning guided workflow for Ribosome Binding Site engineering}
@@
 \author[3]{Hayman Zumpe H.}
-\author[1]{Ong, C. S.}
-\affil[1]{AIML FSP}
-\affil[2]{ANU}
-\affil[3]{CSIRO Synthetic Biology Future Science Platform, CSIRO Land and Water, Brisbane, QLD 4102, Australia}
+\author[1,2,4]{Ong, C. S.}
+\affil[1]{Machine Learning and Artificial Intelligence Future Science Platform, CSIRO}
+\affil[2]{Australian National University}
+\affil[3]{CSIRO Synthetic Biology Future Science Platform, CSIRO Land and Water}
+\affil[4]{Data61, CSIRO}
 
@@
 Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig:Flowchart}).\\
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig: Flowchart}).\\
 
@@
 In the zeroth round we have tested two batches of designs, for total of 180 variants split as below: 
+\mengyan{we might want to re-format according to the swarm plot groups, with groups names and in order. }
 
@@
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item 30 RBS sequences that were fully and uniformly randomised (equal probability of choosing either nucleotide for each position) 
-    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome
-    \item 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature, as described below.
-\end{enumerate}{}
+    \item 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
+    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E.
+coli} genome \mengyan{Maciej: can you add reference here?}
+    \item 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+    This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
+\end{enumerate}
 
-To generate algorithm driven designs for the 0th round we have used the data set available in \textcite{jervis2018machine}.
-This set contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.\\
-In the subsequent rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds. \\
+In the subsequent rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds.
 
@@
 These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
+\mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
 And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
@@
 \subsection{LEARN - prediction of RBS performance}
+\mengyan{Maciej: shall we put LEARN before DESIGN? since in DESIGN, we have already used the LEARN result from benchmark data anyway.}
 
-For our predictions, we had to solve the problem of feature representation first.
-In our case, that problem was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions.\\
-Due to our use of the Gaussian Process Regression for predictions we have chosen kernel (covariance) functions for our embedding.
-Kernel functions implicitly embed DNA sequences into high-dimensional feature space which enables capturing similarities and distances between data points.  
-The additional effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier.\\
+Our goal in the LEARN part is to learn a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
+Gaussian Process regression (GPR) is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
+The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences in DESIGN section.
 
-There is a number of useful kernel functions that can be used for embedding of DNA.
-In our case, we use string kernels \cite{}, which take two DNA sequences as input and output a real value which represents the similarity between them. 
-The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
-We review the design of the above string kernels in Supplementary materials \ref{}.
-For Round 0, we chose to use the spectrum kernel to process the core 6bp and dot product kernel to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-For subsequent rounds, we chose to use the weighted degree kernel with shifts. \\
-% TODO: reasoning needed. 
+A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between RBS sequences.
+Kernel functions implicitly embed RBS sequences into high-dimensional feature space which enables capturing similarities between data points and makes the regression process easier.
+% The additional effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier.
+For Round 0, since we only have access to limited number of literature sequences, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+For subsequent rounds, we used the a more powerful kernel function, the \textit{weighted degree kernel with shift} (wds) \cite{ratsch_rase_2005_wds}, which has been shown to have a strong performance in varies prediction tasks \cite{Ben-Hur2008}.
+% The predictions were made using the Gaussian Process regression (GPR) algorithm. 
+% In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, 
+% which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated. 
+% Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
+
+% For our predictions, we had to solve the problem of feature representation first.
+% In our case, that problem was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions.\\
+% Due to our use of the Gaussian Process Regression for predictions we have chosen kernel (covariance) functions for our embedding.
+
+% There is a number of useful kernel functions that can be used for embedding of DNA.
+% In our case, we use string kernels \cite{}, which take two DNA sequences as input and output a real value which represents the similarity between them. 
+% The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
+% We review the design of the above string kernels in Supplementary materials \ref{}.
+
 % This approach allowed for good balance between computational complexity and performance.\\
 
-The predictions were made using the Gaussian Process regression (GPR) algorithm. 
-In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated.
-Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.32]{paper/plots/scatter_abc1_FF.pdf}
+    \caption{Scatterplot.}
+    \label{fig: Scatterplot}
+\end{figure}
 
@@
 There are two potential points where the exit from the cycle should be considered: i) the optimum solution has been found or ii) depletion of available resources (time and/or money).
-In our case we have performed a total of four rounds, with performance of the RBSes increasing as seen in \textcolor{red}{Figure X}.\\
+In our case we have performed a total of four rounds, with performance of the RBSes increasing as seen in Figure \ref{fig: Swarmplot and Quantplot}.\\
+\mengyan{Maciej: shall we show the result (swarmplot) in DESIGN when we introduce different groups?}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.32]{paper/plots/swarmplot.pdf}
+    \includegraphics[scale=0.32]{paper/plots/quantplot.pdf}
+    \caption{Swarmplot and Quantplot.}
+    \label{fig: Swarmplot and Quantplot}
+\end{figure}
 

### 2021-05-12 paper/main.tex

@@
 
-The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
-We will discuss our workflow in terms of the Design-Build-Learn-Test cycle, where prediction will be discussed in the Design section and feature representation with prediction in the Learn one.\\
-Figure \ref{fig: Flowchart} shows the DBTL cycle with the machine learning process emphasised.
+Our experiments followed the Design-Build-Test-Learn cycle methodology.
+Figure \ref{fig: Flowchart} shows our DBTL cycle with the machine learning process emphasised.
 Principally, machine learning is used in the LEARN and DESIGN phase and the other two are driven by choices made by human researchers. 
@@
 \mengyan{Maciej: shall we put LEARN before DESIGN? since in DESIGN, we have already used the LEARN result from benchmark data anyway.}
+
+The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
+We will discuss our workflow in terms of the Design-Build-Learn-Test cycle, where prediction will be discussed in the Design section and feature representation with prediction in the Learn one.\\
 

### 2021-05-12 paper/supplementary.tex

@@
          \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure GGGGGG.pdf}
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGG.pdf}
          \caption{$y=x$}
@@
          \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure GGGGGC.pdf}
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGC.pdf}
          \caption{$y=3sinx$}
@@
          \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure CGGGAC.pdf}
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure_CGGGAC.pdf}
          \caption{$y=5/x$}
@@
          \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure CACGTT.pdf}
+         \includegraphics[scale=0.25]{plots/Supplementary/Structure_CACGTT.pdf}
          \caption{$y=5/x$}
@@
 \end{figure}
+
+\input{paper/mixture_absGau}

### 2021-05-12 paper/main.tex
Updates from Overleaf

@@
 \newcommand{\maciej}[1]{\textcolor{blue}{#1}}
+
+\addbibresource{ref.bib}
 
@@
     \centering
-    \includegraphics[scale=0.7]{plots/flowchart.pdf}
-    \caption{Flowchart of machine learning based experimental design.}
+    \includegraphics[scale=0.7]{plots/Main_Paper/flowchart.pdf}
+    \caption{\textbf{Flowchart of machine learning based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm.
+After generating the recommendations the RBS are built and tested using automated laboratory methods allowing for rapid construction and testing at scale.
+Finally, the obtained results are fed back to the prediction algorithm in the learn phase.
+}
     \label{fig: Flowchart}
@@
 
-\subsection{DESIGN of the genetic device}
+\subsection{Performance of the recommendation algorithm}
 In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
@@
 In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:
-\mengyan{we might want to re-format according to the swarm plot groups, with groups names and in order. }
 
 \begin{enumerate}
-    \item 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
+    \item BPS-NC and BPS-C group - 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
-    \item 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E.
-coli} genome \mengyan{Maciej: can you add reference here?}
-    \item 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+    \item UNI - 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
+    \item PPM - 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}
+    \item Bandit0 - 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
     This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
@@
 
-In the subsequent rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds.
+In the subsequent 3 rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
+\\
 
-\subsection{BUILD and TEST}
+Figure \ref{fig: Swarmplot and Quantplot} shows the results for all the examined groups. 
+Different randomly generated groups were run together with Bandit 0 (literature data taught) group in Round 0.
+Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
+All the randomly generated groups have performed worse then our benchmark sequence in terms of TIR. 
+Similarly, the Bandit 0 group performed poorly, due to being taught using approximate data set.
+However, starting from round 1, where the bandit algorithm was fed data from the round 0 the results become much better, with a number of sequences that perform similarly and in one case better than the benchmark (by 8\%).
+In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
+Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.32]{paper/plots/Main_Paper/swarmplot.pdf}
+    \includegraphics[scale=0.32]{paper/plots/Main_Paper/quantplot.pdf}
+    \caption{\textbf{TIRs of RBS groups examined in this study.} A) Swarm plot showing the obtained TIRs divided into RBS groups.
+BPS-NC: base-by-base changes in the non-core region.
+BPS-C: base-by-base changes in the core region.
+UNI: Randomly generated sequences with uniform distribution.
+PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E.
+coli}.
+Bandit0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
+SD - Shine-Dalgarno sequence.
+B) Plot showing highest TIR obtained in a given quantile of results divided into groups as in A), save the random groups which were shown together due to similar distributions.}
+    \label{fig: Swarmplot and Quantplot}
+\end{figure}
+
+\subsection{Automated RBS variant construction}
 
@@
 To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
-This way we hope to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
+This way we were able to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
 Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
@@
 
-\subsection{LEARN - prediction of RBS performance}
-\mengyan{Maciej: shall we put LEARN before DESIGN? since in DESIGN, we have already used the LEARN result from benchmark data anyway.}
+\subsection{Prediction of RBS performance}
 
-The three main considerations when designing a machine learning workflow for synthetic biology is feature representation, prediction and recommendation algorithms.
-We will discuss our workflow in terms of the Design-Build-Learn-Test cycle, where prediction will be discussed in the Design section and feature representation with prediction in the Learn one.\\
+Our recommendation algorithm selects new designs based on our prediction algorithm.
+This algorithm creates a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
+For this study, we have used the Gaussian Process regression (GPR) as our prediction method.
+GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
+Such a stochastic prediction method fits biological processes naturally, since they are highly stochastic as well. 
+The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
 
-Our goal in the LEARN part is to learn a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
-Gaussian Process regression (GPR) is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
-The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences in DESIGN section.
+A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data point, in our case RBS sequences.
+Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.\\
 
-A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between RBS sequences.
-Kernel functions implicitly embed RBS sequences into high-dimensional feature space which enables capturing similarities between data points and makes the regression process easier.
-% The additional effect of using kernels is that our data will be moved to a higher-dimensional space, which makes the regression process easier.
-For Round 0, since we only have access to limited number of literature sequences, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-For subsequent rounds, we used the a more powerful kernel function, the \textit{weighted degree kernel with shift} (wds) \cite{ratsch_rase_2005_wds}, which has been shown to have a strong performance in varies prediction tasks \cite{Ben-Hur2008}.
-% The predictions were made using the Gaussian Process regression (GPR) algorithm. 
-% In essence, Gaussian Process is a stochastic (Bayesian) predictor of the shape of a function, 
-% which is built based on perceived similarities between data points (kernels), where not only a mean value, but also probability distribution of the mean is calculated. 
-% Such an approach makes GPR well suited to predict biological phenomena which are also highly stochastic. \\
+For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+For subsequent rounds, we used the a more powerful kernel function, the \textit{weighted degree kernel with shift} (wds) \cite{ratsch_rase_2005_wds}, which has been shown to have a strong performance in various prediction tasks \cite{Ben-Hur2008}.
 
-% For our predictions, we had to solve the problem of feature representation first.
-% In our case, that problem was how to embed the DNA sequence to a numerical vector which can be used in machine learning predictions.\\
-% Due to our use of the Gaussian Process Regression for predictions we have chosen kernel (covariance) functions for our embedding.
-
-% There is a number of useful kernel functions that can be used for embedding of DNA.
-% In our case, we use string kernels \cite{}, which take two DNA sequences as input and output a real value which represents the similarity between them. 
-% The commonly used string kernels includes spectrum kernels \cite{}, mixed spectrum kernels \cite{}, weighted degree kernels \cite{} and weighted degree kernels with shifts \cite{}. 
-% We review the design of the above string kernels in Supplementary materials \ref{}.
-
-% This approach allowed for good balance between computational complexity and performance.\\
+Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
+As expected, the predictions in Round 0 were poor due to use of approximated data. 
+The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
+Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
+Interestingly there wasn't much difference in fitness parameters between Round 1-3.
+Another interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
 
@@
     \centering
-    \includegraphics[scale=0.32]{paper/plots/scatter_abc1_FF.pdf}
+    \includegraphics[scale=0.32]{paper/plots/Main_Paper/scatter_abc1_FF.pdf}
     \caption{Scatterplot.}
     \label{fig: Scatterplot}
-\end{figure}
-
-\subsection{EXIT}
-There are two potential points where the exit from the cycle should be considered: i) the optimum solution has been found or ii) depletion of available resources (time and/or money).
-In our case we have performed a total of four rounds, with performance of the RBSes increasing as seen in Figure \ref{fig: Swarmplot and Quantplot}.\\
-\mengyan{Maciej: shall we show the result (swarmplot) in DESIGN when we introduce different groups?}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.32]{paper/plots/swarmplot.pdf}
-    \includegraphics[scale=0.32]{paper/plots/quantplot.pdf}
-    \caption{Swarmplot and Quantplot.}
-    \label{fig: Swarmplot and Quantplot}
 \end{figure}
@@
 A genetic part library should display a number of characteristics to be deemed useful.
-First quality expected from a good library is a wide range of relevant trait values (in our case it is the TIR) so that relevant genetic devices can be appropriately optimised.
-Second important feature of a library is the diversity of offered sequences, which ensures that sequences compatible with different assembly methods and plasmids are available. 
-Additionally, it is very beneficial for the library to show a number of parts with similar numerical values for relevant trait, but with different sequences, so that a designer would have to settle on suboptimal part due to, for example, assembly incompatibility.
-Finally, a good library has to be reliable - if the numbers provided in the library are unreliable designers will be reluctant to use.
-Figure X shows how library conforms to these requirements.
+First quality expected from a good library is a wide range of relevant trait values (in our case it is the TIR) so that designed genetic devices can be appropriately optimised.
+Secondly, it is very beneficial for the library to show a number of parts with similar numerical values for the relevant trait, but with different sequences, so that a designer would not have to settle on suboptimal part due to, for example, assembly incompatibility.
+Finally, a good library has to be reliable - if the numbers provided in the library are unreliable designers will be reluctant to use the data set.
+Figure \ref{fig:Library characteristics} shows how our library conforms to these requirements.
 In short, due to focusing on exploitation distribution of our TIRs is skewed toward strong ones, but there is still a significant amount of RBSs with medium and low strengths.
@@
          \caption{$y=x$}
-         \label{fig:y equals x}
      \end{subfigure}
@@
          \caption{$y=3sinx$}
-         \label{fig:three sin x}
      \end{subfigure}
@@
          \caption{$y=5/x$}
-         \label{fig:five over x}
      \end{subfigure}
@@
         \caption{Three simple graphs}
-        \label{fig:three graphs}
+        \label{fig:Library characteristics}
 \end{figure}
@@
 We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
-Figure X shows what edit distance is required for positive change in TIR for RBS with high and medium TIR.
+Figure \ref{fig:Library characteristics} shows what edit distance is required for positive change in TIR for RBS with high and medium TIR.
 For high TIR RBS, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.

### 2021-05-12 paper/methods.tex

@@
 \textbf{Sequencing}
-Sequences that showed TIR on the level of at least 85\% of TIR of the benchmark sequence were sequenced using Sanger sequencing either at the Australian Genome Research Facility (Brisbane, QLD, Australia) or at Macrogen (Seoul, Republic of Korea).
-Plasmids for sequencing were extracted using the Qiagen plasmid miniprep kit.
+Strains that gave GFP fluorescence intensity readings similar to that of the original RBS were selected for sequence confirmation by capillary electrophoresis sequencing (CES) by Macrogen, Inc.
+(Seoul, South Korea).
+The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 µg/mL).
+Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN) according to the manufacturer’s instructions.
+Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/µL.
+Samples of 20 µL of undiluted plasmid DNA were sequenced using a single primer (5’-CGATATAGGCGCCAGCAA-3’) that binds approximately 150 bp upstream of the RBS.
+Reads were aligned with the template sequence in the Teselagen software (TeselaGen Biotechnology, Inc.).
 
@@
 The choice of covariance (kernel) function is critical for accurate predictions, since it controls smoothness and amplitude of the function we model.
-To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005} to specify the kernel function of $GP$.
+To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.
 WDS is a type of a string kernel, which takes two sequences (strings) as inputs and outputs a scalar value which represents the similarities between the two sequences.  

### 2021-05-12 paper/main.tex
fix bugs for local env

@@
 \usepackage{authblk}
-% \usepackage{subfigure}
+% \usepackage{subfloat}
 
@@
     \centering
-    \includegraphics[scale=0.32]{paper/plots/Main_Paper/swarmplot.pdf}
-    \includegraphics[scale=0.32]{paper/plots/Main_Paper/quantplot.pdf}
+    \includegraphics[scale=0.32]{plots/Main_Paper/swarmplot.pdf}
+    \includegraphics[scale=0.32]{plots/Main_Paper/quantplot.pdf}
     \caption{\textbf{TIRs of RBS groups examined in this study.} A) Swarm plot showing the obtained TIRs divided into RBS groups.
@@
     \centering
-    \includegraphics[scale=0.32]{paper/plots/Main_Paper/scatter_abc1_FF.pdf}
+    \includegraphics[scale=0.32]{plots/Main_Paper/scatter_abc1_FF.pdf}
     \caption{Scatterplot.}
@@
 
-\input{methods}
+\input{methods.tex}
 

### 2021-05-12 paper/methods.tex
fix bugs for local env

@@
 The pBbB6c-GFP plasmid has been used for all our designs.
-This plasmid comes with GFP mut3b CDS inducible with addition of Isopropyl β-D-1-thiogalactopyranoside (IPTG).
+This plasmid comes with GFP mut3b CDS inducible with addition of Isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG).
 The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
@@
 (Seoul, South Korea).
-The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 µg/mL).
+The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
 Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN) according to the manufacturer’s instructions.
-Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/µL.
-Samples of 20 µL of undiluted plasmid DNA were sequenced using a single primer (5’-CGATATAGGCGCCAGCAA-3’) that binds approximately 150 bp upstream of the RBS.
+Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/$\mu$L.
+Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5’-CGATATAGGCGCCAGCAA-3’) that binds approximately 150 bp upstream of the RBS.
 Reads were aligned with the template sequence in the Teselagen software (TeselaGen Biotechnology, Inc.).
@@
 
-\begin{figure}[t]
+% \begin{figure}[t]
+% \centering
+% \subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}}\par\medskip
+% \begin{minipage}{.5\linewidth}
+% \centering
+% \subfloat{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
+% \end{minipage}%
+% \begin{minipage}{.5\linewidth}
+% \centering
+% \subfloat{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
+% \end{minipage}
+% \caption{Batch Recommendation.
+We use the batch size of 2, with 5 initial observations.
+The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e.
+{-2, -1.5, -1, ..., 9.5, 10}.
+% The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
+% (a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
+% (b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
+% \label{fig:batch rec}
+% \end{figure}
+
+\begin{figure}
     \centering
-\subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}}\par\medskip
-\begin{minipage}{.5\linewidth}
+    \begin{subfigure}[b]{0.49\textwidth}
         \centering
-\subfloat[]{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
+        \includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}
+    \end{subfigure}
+    \vskip\baselineskip
+    \begin{subfigure}[b]{0.49\textwidth}
         \centering
-\subfloat[]{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
-\end{minipage}
+        \includegraphics[scale=0.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[scale=0.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}
+    \end{subfigure}
     \caption{Batch Recommendation.

### 2021-05-13 paper/main.tex
update paper plots
Updated Results in Paper.

@@
     This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
+    \mengyan{need some analysis on literature data, to verify 1) our library is "better" 2) it's hard to do ML on available library.
+some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
+    And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
 \end{enumerate}
@@
     \centering
-    \includegraphics[scale=0.32]{plots/Main_Paper/swarmplot.pdf}
-    \includegraphics[scale=0.32]{plots/Main_Paper/quantplot.pdf}
+    \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
+    \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
+    \includegraphics[scale=0.35]{plots/Main_Paper/histogram.pdf}
+    \includegraphics[scale=0.35]{plots/Main_Paper/tsneplot.pdf}
     \caption{\textbf{TIRs of RBS groups examined in this study.} A) Swarm plot showing the obtained TIRs divided into RBS groups.
@@
 SD - Shine-Dalgarno sequence.
-B) Plot showing highest TIR obtained in a given quantile of results divided into groups as in A), save the random groups which were shown together due to similar distributions.}
+Dash line is set 1 and represents the averaged benchmark sequence TIR for that group.
+B) Plot showing highest TIR obtained in a given quantile of results divided into groups as in A), save the random groups which were shown together due to similar distributions.
+C) Histogram with kernel density estimations (KDE) showing distributions of TIRs for different groups.
+D) tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function.
+The size of the circle represents the experimentally obtained (coloured) and predicted (grey) TIR.
+The results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.}
     \label{fig: Swarmplot and Quantplot}
@@
 \subsection{Automated RBS variant construction}
+
+One of the challenging for applying machine learning techniques on synthetic biology is lacking of high-quality labelled data. 
+The available RBS library \cite{jervis2018machine} only cover 56 unique labelled RBS
 
@@
 Similarly, colony picking was done by an automated colony picker.\\
+Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation.
+The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
+Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.\\
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
+    \caption{\textbf{Distribution of standard deviation of tested samples} Join tplot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation.
+TIR is given in terms of raw number.}
+    \label{fig: SD}
+\end{figure}
 
@@
 For subsequent rounds, we used the a more powerful kernel function, the \textit{weighted degree kernel with shift} (wds) \cite{ratsch_rase_2005_wds}, which has been shown to have a strong performance in various prediction tasks \cite{Ben-Hur2008}.
+The relative distances between sequences calculated by our kernel function are shown in figure \ref{fig: Swarmplot and Quantplot} D.
 
@@
     \includegraphics[scale=0.32]{plots/Main_Paper/scatter_abc1_FF.pdf}
-    \caption{Scatterplot.}
+    \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each of the rounds.
+R\textsuperscript{2}, Spearman Correlation coefficient and p value are all provided for each plot.}
     \label{fig: Scatterplot}
@@
 
-\subsection{Characteristics of the obtained library and sequences}
+\subsection{Important characteristics of the evaluated sequences}
 
@@
 
+\section*{Code and data availability}
+
+All code and data required to reproduce the results is available at Github:
+
 \section*{Contributions}

### 2021-05-13 paper/main.tex
add challenges in result sec

@@
 
-Our experiments followed the Design-Build-Test-Learn cycle methodology.
+The challenges of applying machine learning techniques on synthetic biology have two folds.
+On the one hand, there is no large-scale and high-quality labelled data. The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
+One the other hand, it is time-consuming and expensive to measure the TIR for RBS sequences \mengyan{add more details or reference here}. 
+We address these challenges by 
+% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
+% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
+% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
+% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
+apply the Design-Build-Test-Learn (DBTL) cycle methodology.
+Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
+
 Figure \ref{fig: Flowchart} shows our DBTL cycle with the machine learning process emphasised.
-Principally, machine learning is used in the LEARN and DESIGN phase and the other two are driven by choices made by human researchers. 
-In LEARN phase we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences and in the DESIGN phase we use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.\\
+BUILD and TEST are driven by choices made by human researchers.
+Machine learning algorithms are applied in two parts.
+In the DESIGN phase, machine learning based experimental design algorithm (Upper Confidence Bound multi-armed Bandit algorithm) is applied.
+The goal is to design a policy to recommend RBS sequences in batches so that we can identify RBS sequences with highest possible TIR. 
+% We use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
+In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences.
+The goal is to provide predictions about TIR labels and uncertainty to assit the recommendation.\\
 
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
     \centering
-    \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
-    \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
-    \includegraphics[scale=0.35]{plots/Main_Paper/histogram.pdf}
-    \includegraphics[scale=0.35]{plots/Main_Paper/tsneplot.pdf}
+    \includegraphics[scale=0.4]{plots/Main_Paper/swarmplot.pdf}
+    \includegraphics[scale=0.4]{plots/Main_Paper/quantplot.pdf}
+    \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
+    \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
     \caption{\textbf{TIRs of RBS groups examined in this study.} A) Swarm plot showing the obtained TIRs divided into RBS groups.
@@
 
-One of the challenging for applying machine learning techniques on synthetic biology is lacking of high-quality labelled data. 
-The available RBS library \cite{jervis2018machine} only cover 56 unique labelled RBS
-
-For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
-These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
-\mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
-And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
 To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
@@
 Similarly, colony picking was done by an automated colony picker.\\
-Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation.
+Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).
 The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
@@
     \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
-    \caption{\textbf{Distribution of standard deviation of tested samples} Join tplot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation.
+    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
 TIR is given in terms of raw number.}
@@
     \centering
-    \includegraphics[scale=0.32]{plots/Main_Paper/scatter_abc1_FF.pdf}
+    \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF.pdf}
     \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each of the rounds.
@@
 
-\subsection{Important characteristics of the evaluated sequences}
+\subsection{Analysis of RBS positional information}
 
-A genetic part library should display a number of characteristics to be deemed useful.
-First quality expected from a good library is a wide range of relevant trait values (in our case it is the TIR) so that designed genetic devices can be appropriately optimised.
-Secondly, it is very beneficial for the library to show a number of parts with similar numerical values for the relevant trait, but with different sequences, so that a designer would not have to settle on suboptimal part due to, for example, assembly incompatibility.
-Finally, a good library has to be reliable - if the numbers provided in the library are unreliable designers will be reluctant to use the data set.
-Figure \ref{fig:Library characteristics} shows how our library conforms to these requirements.
-In short, due to focusing on exploitation distribution of our TIRs is skewed toward strong ones, but there is still a significant amount of RBSs with medium and low strengths.
-High diversity of our sequences is shown by lack of relative lack of bias in position weight matrices for different bins.
-Finally, we were able to keep the average standard deviation for each bin within XX\% with 6 biological replicates for each sample, which we find satisfactory for reliability.\\
+Figure \ref{fig:Library characteristics} A shows the sequence logo calculated for the Top30 of our sequences.
+The expected bias towards guanine is clearly visible for all positions.
+This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot} D reinforces the notion that our algorithm successfully identified sequences with high TIR probability.\\
 
@@
          \centering
-         \includegraphics[scale=1.2]{plots/logo.pdf}
-         \caption{$y=x$}
+         \includegraphics[scale=1.2]{plots/Main_Paper/TOP30_logo.pdf}
+         \caption{}
      \end{subfigure}
@@
          \centering
-         \includegraphics[scale=0.5]{plots/Patterns.png}
-         \caption{$y=3sinx$}
+         \includegraphics[scale=0.5]{paper/plots/Main_Paper/Hd_Heatmap.png}
+         \caption{}
      \end{subfigure}
@@
          \centering
-         \includegraphics[scale=0.5]{plots/Hd_Heatmap.png}
-         \caption{$y=5/x$}
+         \includegraphics[scale=0.5]{paper/plots/Main_Paper/Patterns.png}
+         \caption{}
      \end{subfigure}
@@
          \includegraphics[scale=0.5]{plots/Histogram.png}
-         \caption{$y=5/x$}
+         \caption{}
          \label{fig:five over x}
      \end{subfigure}
-        \caption{Three simple graphs}
+        \caption{\textbf{Characteristics of the tested RBS sequences.} A) Sequence logo for the Top 30 tested sequences.
+B) Heatmap showing the TIR difference between given RBS (y axis) and a RBS with the highest tested TIR at a given Hamming (edit) distance.
+}
         \label{fig:Library characteristics}
@@
 
-Our library is also showing two characteristics that may be useful in future RBS design efforts.
-One of them is the perceived editing distance between two sequences required for meaningful improvement in the TIR. 
+Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for meaningful improvement in the TIR. 
 We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
-Figure \ref{fig:Library characteristics} shows what edit distance is required for positive change in TIR for RBS with high and medium TIR.
+Figure \ref{fig:Library characteristics} B shows what edit distance is required for positive change in TIR for RBS with high and medium TIR.
 For high TIR RBS, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
@@
 That means that as the TIR of examined RBSs increases, the algorithm should be "encouraged" to explore sequences which are more dissimilar to the current candidates.
-This finding could also indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs - the low rate of natural mutations will be too slow to discover through more dissimilar sequences \cite{Lee2012}.
-In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or 5 changes in the RBS region might be prohibitively long.
+This finding could also indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs - the low rate of natural mutations will be very slow to explore more dissimilar sequences \cite{Lee2012}.
+In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
 In such cases, a directed process should be strongly encouraged.\\
-Another feature noticeable in our library is prevalence of specific patterns.
-These include X, Y and Z (Figure X). 
-Interestingly, RBSs containing these patterns are not detected as capable of binding the relevant ribosome anti-sequence by binding energy calculations \cite{Mann2017}.
-At present, there is no strong evidence why these patterns might by permissive for high TIRs, nevertheless, it is important to note that the used algorithm does not require that knowledge to effectively find them.\\
+
+Finally, our strong sequences did not show neither strong binding to the antisense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
+This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the
 

### 2021-05-13 paper/main.tex
update paper plots

@@
 
-\subsection{Analysis of RBS positional information}
+\subsection{Characteristics of the tested sequences}
 
@@
          \centering
-         \includegraphics[scale=0.5]{paper/plots/Main_Paper/Hd_Heatmap.png}
-         \caption{}
+         \includegraphics[scale=0.5]{plots/Main_Paper/Hd_Heatmap.png}
+         \caption{\textbf{Analysis of RBS positional information} A) Sequence logo calculated for the Top 30 tested sequences.
+B) Heatmap showing what edit (Hamming) distance is required for positive change in TIR for RBS with high and medium TIR.
+The temperature scale shows the difference between a given RBS on y axis and the RBS with strongest TIR at the given distance.}
      \end{subfigure}
-     \vskip\baselineskip
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \includegraphics[scale=0.5]{paper/plots/Main_Paper/Patterns.png}
-         \caption{}
-     \end{subfigure}
-     \hfill
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \includegraphics[scale=0.5]{plots/Histogram.png}
-         \caption{}
-         \label{fig:five over x}
-     \end{subfigure}
-        \caption{\textbf{Characteristics of the tested RBS sequences.} A) Sequence logo for the Top 30 tested sequences.
-B) Heatmap showing the TIR difference between given RBS (y axis) and a RBS with the highest tested TIR at a given Hamming (edit) distance.
-}
-        \label{fig:Library characteristics}
 \end{figure}
@@
 In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
-In such cases, a directed process should be strongly encouraged.\\
+In such cases, a directed process should be strongly encouraged.
+This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed.\\
 
-Finally, our strong sequences did not show neither strong binding to the antisense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
-This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the
+Finally, our strong sequences did not show neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
+This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.
 

### 2021-05-14 paper/main.tex
Updates from Overleaf

@@
 This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
-Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used\cite{Yeoh2019,Nielsen2016}.
+Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
 Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
@@
 This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
-The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\
-The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
-For example, according to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TRI based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
-Reported predictions from all of these models are relatively good ($R^2 >0.8$), but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
+The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.
+\mengyan{Maciej:how about ours? same approach or better?}
+
+The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
+\mengyan{Maciej: can you replace citations like "Reeve \emph{et al.}" to \textcite{Reeve2014}?}
+For example, according to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+Reported predictions from all of these models are relatively good ($R^2 >0.8$), \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g.
+is that because the data?
+So I recommend either remove this or explain it somewhere.}
+but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
 There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} .
-This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .\\
+This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .
+
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
 New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
-These new data-sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018}.
-In the past few years there was a significant uptake of Machine Learning based approaches in synthetic biology \cite{LAWSON2021}.
-Jervis \emph{et al.} used support vector machine and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
+These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018}.
+In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
+Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
-There were also successful attempts at using deep learning techniques for analysis of big data-sets \cite{Alipanahi2015,Angermueller2016}.
+There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.
+
 Machine learning has been also used for prediction in proteins \cite{Yang2018}.
-However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. \\
-Here we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with goal of optimisation of associated protein level expression.
+However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
+The challenges of applying machine learning techniques on synthetic biology have two folds.
+On the one hand, there is no large-scale and high-quality labelled data. The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
+One the other hand, it is time-consuming and expensive to measure the TIR for RBS sequences \mengyan{Maciej: can you add more details or reference here}. 
+We addressed the challenges by applying DBTL workflow. 
+Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
+% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
+% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
+% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
+% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
+\mengyan{why?
+we need to be more precise.
+Mengyan can add it later.
+Make it a paragraph, showing the challenges in learning with small number of noisy data and exploration-exploitation balance, and explain how we address that.}
+
+We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level.
 RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-In this work We have used Gaussian Process Regression and Upper Confidence Bound multi-armed Bandits algorithms for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS \cite{desautels2012parallelizing, Rasmussen2004}.
-Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by identifying the set of RBS sequences with top TIR scores while minimising the number of DBTL cycle turns that we had to do.
-We did this by designing a sequential experimental workflow, where designs in the zeroth round where randomised RBS sequences designed to explore the experimental space and some preliminary machine learning recommended designs based on literature.
-In the subsequent rounds, designs where recommended by the algorithm based on the data obtained in the previous rounds. 
+In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2012parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
+Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
+We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
+In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
+In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
 The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
@@
 This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
-At the same time we were able to discover new RBS sequences with very high TIRs.\\
+At the same time we were able to discover new RBS sequences with very high TIRs \mengyan{Maciej: can you make "very high" concrete?}.
 
@@
 
-The challenges of applying machine learning techniques on synthetic biology have two folds.
-On the one hand, there is no large-scale and high-quality labelled data. The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
-One the other hand, it is time-consuming and expensive to measure the TIR for RBS sequences \mengyan{add more details or reference here}. 
-We address these challenges by 
-% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
-% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
-% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
-% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
-apply the Design-Build-Test-Learn (DBTL) cycle methodology.
-Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
-
-Figure \ref{fig: Flowchart} shows our DBTL cycle with the machine learning process emphasised.
+We show our DBTL workflow in Figure \ref{fig: Flowchart}.
+% with the machine learning process emphasised.
 BUILD and TEST are driven by choices made by human researchers.
 Machine learning algorithms are applied in two parts.
-In the DESIGN phase, machine learning based experimental design algorithm (Upper Confidence Bound multi-armed Bandit algorithm) is applied.
-The goal is to design a policy to recommend RBS sequences in batches so that we can identify RBS sequences with highest possible TIR. 
+In the DESIGN phase, we designed experimental design policy to recommend RBS sequences in batches so that we can identify RBS sequences with highest possible TIR. 
 % We use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
 In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences.
-The goal is to provide predictions about TIR labels and uncertainty to assit the recommendation.\\
+The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation.
 
@@
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
+By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.
 
-In \emph{E. coli} the RBS is usually located in the 20 bases upstream of the start codon. 
-Additionally, there is a consensus RBS core sequence called the Shine-Dalgarno sequence, which in \emph{E. coli} is AGGAGG. 
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig: Flowchart}).\\
+In \emph{E. coli}, the RBS is usually located in the 20 bases upstream of the start codon. 
+Additionally, there is a consensus RBS core sequence called the \textit{Shine-Dalgarno sequence}, which in \emph{E. coli} is \textbf{AGGAGG}. 
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig: Flowchart}). \mengyan{Maciej: what does this sentence mean?}
 
-Our template RBS sequence is 20 bps long with the sequence TTTAAGAAGGAGATATACA.
-This sequence is a known high TIR RBS that comes with the pBb series plasmids \cite{Lee2011}. 
-Since this is the sequence against which new RBS sequenced will be benchmarked we will refer to this sequence as the benchmark sequence from now on.
+Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, 
+which is a known RBS with high \mengyan{Maciej: can we claim "highest"?} TIR that comes with the pBb series plasmids \cite{Lee2011}. 
+Since this is the sequence against which new RBS sequenced will be benchmarked,
+we will refer to this sequence as the \textit{benchmark sequence} from now on.
 In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT.
+TTTAAGA + NNNNNN + TATACAT, where N can be any choices of A, C, G, T.
+The total experimental (variant) space is then $4^6$ = 4096.
 We have experimentally confirmed that changing the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
-Since for each of the 6 position there are 4 possibilities: A, C, G, T the total experimental (variant) space is $4^6$ = 4096.\\
 
@@
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (GPR).
-Another feature of the bandits algorithms is that it balances two approaches: exploration and exploitation. 
-Exploration makes the algorithm recommend designs that will improve the predictions, whereas exploitation will recommend designs that focus on delivering the most efficient design the fastest. 
-The two approaches can be controlled with the \textbeta\enspace parameter. 
-We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. 
+In our case we use the Upper Confidence Bound algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (GPR).
+Another feature of the bandits algorithms is that it balances two parts: the exploration in terms of potential high TIRs in unknown areas, and exploitation in terms of querying areas which is known to give relatively high TIRs.
+% The two approaches can be controlled with the \textbeta\enspace parameter. 
+% We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. 
 One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). 
 As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.
-\\
-
-To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
-To fit into our automated workflow, we have divided the 450 variants into batches of 90.
-In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:
-
-\begin{enumerate}
-    \item BPS-NC and BPS-C group - 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
-This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part.
-    \item UNI - 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
-    \item PPM - 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}
-    \item Bandit0 - 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
-    This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
-    \mengyan{need some analysis on literature data, to verify 1) our library is "better" 2) it's hard to do ML on available library.
-some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
-    And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
-\end{enumerate}
-
-In the subsequent 3 rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
-\\
-
-Figure \ref{fig: Swarmplot and Quantplot} shows the results for all the examined groups. 
-Different randomly generated groups were run together with Bandit 0 (literature data taught) group in Round 0.
-Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
-All the randomly generated groups have performed worse then our benchmark sequence in terms of TIR. 
-Similarly, the Bandit 0 group performed poorly, due to being taught using approximate data set.
-However, starting from round 1, where the bandit algorithm was fed data from the round 0 the results become much better, with a number of sequences that perform similarly and in one case better than the benchmark (by 8\%).
-In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
-Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
 
@@
     \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
-    \caption{\textbf{TIRs of RBS groups examined in this study.} A) Swarm plot showing the obtained TIRs divided into RBS groups.
+    \caption{\textbf{TIRs of RBS groups examined in this study.} 
+    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.
+    \mengyan{Maciej: make the sentence more clear?}
+    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
     BPS-NC: base-by-base changes in the non-core region. 
@@
     UNI: Randomly generated sequences with uniform distribution. 
-PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E.
-coli}.
-Bandit0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
-SD - Shine-Dalgarno sequence.
-Dash line is set 1 and represents the averaged benchmark sequence TIR for that group.
-B) Plot showing highest TIR obtained in a given quantile of results divided into groups as in A), save the random groups which were shown together due to similar distributions.
-C) Histogram with kernel density estimations (KDE) showing distributions of TIRs for different groups.
-D) tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function.
-The size of the circle represents the experimentally obtained (coloured) and predicted (grey) TIR.
-The results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.}
+    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
+    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively. 
+    SD - Shine-Dalgarno sequence. Dash line is set 1 and represents the averaged benchmark sequence TIR for that group. 
+    \mengyan{explain why reference sequences are not all one.}
+    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
+    % save the random groups which were shown together due to similar distributions.
+    UNI and PPM are merged into Random group. 
+    \mengyan{Maciej: maybe need to address why there are different groups in different plot, e.g. why no BPS-NC.}
+    \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
+    \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
+The size of the circle represents the experimentally obtained TIR for measured groups.}
     \label{fig: Swarmplot and Quantplot}
@@
 
+To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
+To fit into our automated workflow, we have divided the 450 variants into batches of 90.
+In the zeroth round we have tested two batches of designs, for total of 180 variants split as below: 
+\begin{itemize}
+    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
+This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (Figure \ref{fig:core_vs_noncore} \mengyan{put it in main paper?}).
+    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
+    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}.
+    \item Bandit-0: 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+    This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
+\end{itemize}
+In the subsequent 3 rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
+
+Figure \ref{fig: Swarmplot and Quantplot} shows the results for all the examined groups. 
+% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
+% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
+All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
+\mengyan{explain the TIR ratio we used.}
+The Bandit-0 group performed poorly, due to being trained based on literature data.
+\mengyan{Maciej: can you add some analysis on literature data?
+to verify 1) our library is "better" 2) it's hard to do ML on available library.
+some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
+And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
+However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
+In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
+Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
+
+\mengyan{the analysis below only focuses on the swarmplot, we need to address other plots as well. Some points to address: 
+1) in swarmplot and histogram, explain the two-mode is due to the exploration and exploitation. 
+2) in quantile plot, address we care about the top TIRs, that's why we look at the quantiles bigger than 0.5.
+3) in tSNE plot, address the coverage of the whole space. Maybe also link to the exploration and exploitation balance.}
+
 \subsection{Automated RBS variant construction}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
+    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
+TIR is given in terms of raw number.}
+    \label{fig: SD}
+\end{figure}
+
+\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this).
+If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
 
@@
 Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).
+\mengyan{Maciej: can you explain what is "given samples standard deviation"?}
 The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
-Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.\\
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
-    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
-TIR is given in terms of raw number.}
-    \label{fig: SD}
-\end{figure}
+Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.
 
@@
 GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
-Such a stochastic prediction method fits biological processes naturally, since they are highly stochastic as well. 
-The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
+% Such a stochastic prediction method fits biological processes naturally, since they are highly stochastic as well. 
+The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.
 
 A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data point, in our case RBS sequences.
-Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.\\
-
+Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.
 For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
@@
 Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
-Interestingly there wasn't much difference in fitness parameters between Round 1-3.
-Another interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
+% Interestingly there wasn't much difference in fitness parameters between Round 1-3.
+One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
+\mengyan{Mengyan to explain more about the prediction v.s. recommendation.}
 
-\begin{figure}[h]
+\mengyan{Some other points need to address: 
+1) the TIR label we used here is different from the TIR ratio in Fig 2. 
+2) the setting of predictor.}
+
+\begin{figure}[!ht]
     \centering
@@
 
-Figure \ref{fig:Library characteristics} A shows the sequence logo calculated for the Top30 of our sequences.
-The expected bias towards guanine is clearly visible for all positions.
-This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot} D reinforces the notion that our algorithm successfully identified sequences with high TIR probability.\\
+Figure \ref{fig:Library characteristics} A shows the sequence logo calculated for the Top 30 \mengyan{Maciej: Top 30 of what? raw TIR? TIR ratio?} of our sequences.
+The expected bias \mengyan{Maciej: what is expected bias?} towards guanine is clearly visible for all positions.
+This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot} D reinforces the notion that our algorithm successfully identified sequences with high TIR probability.
 
@@
          \includegraphics[scale=0.5]{plots/Main_Paper/Hd_Heatmap.png}
+         \caption{}
+     \end{subfigure}
      \caption{\textbf{Analysis of RBS positional information} A) Sequence logo calculated for the Top 30 tested sequences.
@@
 The temperature scale shows the difference between a given RBS on y axis and the RBS with strongest TIR at the given distance.}
-     \end{subfigure}
+     \label{fig:Library characteristics}
 \end{figure}
@@
 Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for meaningful improvement in the TIR. 
-We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
-Figure \ref{fig:Library characteristics} B shows what edit distance is required for positive change in TIR for RBS with high and medium TIR.
-For high TIR RBS, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
-For medium TIR RBS, a distance of just 1 is enough to produce a meaningful increase in TIR.
-That means that as the TIR of examined RBSs increases, the algorithm should be "encouraged" to explore sequences which are more dissimilar to the current candidates.
-This finding could also indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs - the low rate of natural mutations will be very slow to explore more dissimilar sequences \cite{Lee2012}.
+\mengyan{Maciej: what's a "meaningful improvement"}
+We define the editing distance as Hamming distance \mengyan{Maciej: please add citation here}, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
+Figure \ref{fig:Library characteristics} B shows what edit distance is required for positive change in TIR for RBS with high and medium TIR. \mengyan{Maciej: what's the definition of high and medium?}
+For RBS with high TIR, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
+For RBS with medium TIR, a distance of 1 is enough to produce a meaningful increase in TIR.
+That means that as the TIR of examined RBSs increases, 
+exploring sequences which are more dissimilar to the current candidates tends to give meaningful improvement. 
+% the algorithm should be ``encouraged" to explore sequences which are more dissimilar to the current candidates.
+The low rate of natural mutations will be very slow to explore more dissimilar sequences \cite{Lee2012},
+which indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
 In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
 In such cases, a directed process should be strongly encouraged.
-This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed.\\
+This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed.
+\mengyan{Maciej: can you add one or more citations here?}
 
 Finally, our strong sequences did not show neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
-This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.
+This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \mengyan{Maciej: can you add one or more citations here?} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.
+
+\mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any.
+Comparison with existing library, if any.}
 
@@
 We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
-We have investigated a number of methods of digitising the DNA sequence, finally settling on Spectrum Kernel with Shift method, which fit into our regression method.
+We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit into our regression method.
 We performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.\\
+By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.
+
 There is a number of issues that we would like to address in the future.
-The biggest of them is elucidating the reason for none of our sequences being able to significantly cross the TIR of our benchmark sequence.
+The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
 Essentially, there are two possible reasons for this, one is due to the algorithm and one due to nature.
 If the problem is with the algorithm, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
-How to solve this?
-It might be also possible that the TIR seen in the benchmark is simply already the strongest possible and represents a natural ceiling for it. 
-In that case, change in the algorithm will not help.
+It might be also possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
+In that case, changes in the algorithm will not help.
 Another problem we would like to tackle is the not always good prediction of the TIR of a given RBS.
-In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.\\
+\mengyan{Mengyan to address exploration purpose here.}
+In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
+
 In the future, we hope to extend the algorithm to other, more complicated genetic elements.
@@
 This is because the experimental space grows exponentially with the number of examined positions and so the space becomes increasingly hard to cover with experiments.
-To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
+To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.
+
 Finally, we hope to see our RBS library being adapted by the community.
 We believe that its favourable characteristics will make if very useful to designers.
+
+\mengyan{We should try to alter the discussion in terms of our three-sentence story, along with open questions (in a more positive way) and future work.}
 

### 2021-05-14 paper/methods.tex
Updated Results in Paper.
add challenges in result sec
Updates from Overleaf

@@
 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
+The implementation of the machine learning algorithm is in Python 3.7 and scikit-learn library \cite{scikit-learn}.
 
-\subsubsection{LEARN: Gaussian Process Regression with String Kernel}
+\subsubsection{Prediction: Gaussian Process Regression with String Kernel}
 
@@
 
-\subsubsection{DESIGN: Batch UCB}
+\subsubsection{Recommendation: Batch Upper Confidence Bound Bandit}
 

### 2021-05-15 paper/main.tex

@@
 At the same time we were able to discover new RBS sequences with very high TIRs \mengyan{Maciej: can you make "very high" concrete?}.
+
+The \textbf{Story} (3 sentences):\\
+\begin{enumerate}
+    \item we automate the design of sequences for finding high TIR RBS
+    \begin{enumerate}
+        \item DBTL
+    \end{enumerate}
+    \item the ML method discovers biological knowledge (validates existing knowledge), including the highest TIR ever
+    \begin{enumerate}
+        \item high G
+        \item diverse library
+        \item edit distance > 2
+    \end{enumerate}
+    \item BioFoundry helps ML (good data), and ML helps BioFoundry (more automation)
+    \begin{enumerate}
+        \item Sequences have small technical variation
+        \item UCB much more efficient than random
+    \end{enumerate}
+\end{enumerate}
 

### 2021-05-15 paper/methods.tex
fix unicode chat U+2212 error

@@
 Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no. E2621L).
-Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10\(\mu\)L.
+Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10 $\mu$L.
 Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
@@
 \textbf{\textit{E. coli} transformation.}
-The DH5α cell line (Thermo Fisher Scientific, catalogue no.
+The DH5$\alpha$ cell line (Thermo Fisher Scientific, catalogue no.
 18265017) was made chemically competent using the Mix $\&$ Go \textit{E.
@@
 The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
-Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN) according to the manufacturer’s instructions.
+Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN) according to the manufacturer's instructions.
 Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/$\mu$L.
-Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5’-CGATATAGGCGCCAGCAA-3’) that binds approximately 150 bp upstream of the RBS.
+Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5'-CGATATAGGCGCCAGCAA-3') that binds approximately 150 bp upstream of the RBS.
 Reads were aligned with the template sequence in the Teselagen software (TeselaGen Biotechnology, Inc.).

### 2021-05-15 paper/methods.tex
mengyan reviewed sec 4.2

@@
 
-The flowchart of machine learning based experimental design is shown in Figure \ref{fig: Flowchart}. 
-To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
-1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
-2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
-The implementation of the machine learning algorithm is in Python 3.7 and scikit-learn library \cite{scikit-learn}.
+Two types of machine learning algorithms can be applied to drive the experimental design workflow shown in Figure \ref{fig: Flowchart}.
+% To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
+One type of machine learning algorithms is prediction (\textbf{LEARN}), which helps us learn for function of TIR with respect to RBS sequences.
+The other type of machine learning algorithm is recommendation (\textbf{DESIGN}), which recommends RBS sequences to query in each batch based on the predictions from LEARN.
+% 1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
+% 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
+In round $t$, the prediction and design is based on the results obtained in all previous rounds. 
+The implementation of the machine learning algorithms is in Python 3.7 and scikit-learn library \cite{scikit-learn}. 
+In the following, we describe our machine learning pipeline (Bandit 1-3) by the order of data pre-processing, prediction and string kernels, recommendation. 
+\mengyan{The pipeline for Bandit-0 is reported in Supplementary XX.}
+
+\subsubsection{Data pre-processing}
+
+For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
+\mengyan{Table XXX shows an example of our raw data. }
+% We consider both the label (TIR) pre-processing and kernel pre-processing.
+For label pre-processing, we first adjust the TIR values in each round by the round-wise reference values.
+This is because, we found that TIR values in each round for the same sequence can be different (for reference RBS, STD $\approx$ 17.7) due to the measurement noise.
+So we repeatedly measured the reference RBS in each round to rectify the precision of other measured RBS in the same round. 
+More precisely, in round $t$, we subtract the TIR mean of reference RBS measured in round $t$ from all TIR values measured in the same round, for each replicates respectively. 
+We then pre-processed the data by taking a logarithm transformation and standardisation of the adjusted TIR label for each replicates respectively, 
+After normalisation, each replicate has zero mean and unit variance. 
+Furthermore, we also normalised the kernel matrix used for prediction by centering and unit-norm normalisation, which is reviewed in details in Supplementary \ref{supp: Normalisation of Kernel}.
+% Since the design space is known, we consider the transductive approach \cite{} for the kernel pre-processing with centering and unit-norm normalisation.
 
@@
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values at each point in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}$, where $\epsilon$ is the noise (we assume that the noise is following Gaussian distribution with unknown mean and variance).
-This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g.
-time, temperature, operator, etc.).
+$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
+% This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g. time, temperature, operator, etc.). 
+\mengyan{we should discuss measurement noise somewhere, maybe with a sample example for the dataset we used.}
 
-For regression model, we have used the \textit{Gaussian Process Regression (GPR)}.
-A Gaussian process regression model \cite{Rasmussen2004} is a Bayesian approach which provides uncertainty measurements on predictions. 
-We model $f$ as a sample from a \textit{Gaussian process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel (or covariance) function $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
+For regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004}, which provides not only the predicted TIR scores, but also the uncertainty level of the predictions.
+We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
 %As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
-GPR can predict both the posterior mean and posterior variance. The posterior variance represents the level of uncertainty for the prediction. 
+GPR can predict both the posterior mean and standard deviation. The posterior standard deviation represents the level of uncertainty for the prediction. 
+We review the predictions of GPR in Supplementary \ref{supp: GPR}.
 % We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0. 
@@
 
-The choice of covariance (kernel) function is critical for accurate predictions, since it controls smoothness and amplitude of the function we model.
+The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
+\mengyan{maybe we need to introduce spectrum kernel as well since we used it for round 0.}
 To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.
-WDS is a type of a string kernel, which takes two sequences (strings) as inputs and outputs a scalar value which represents the similarities between the two sequences.  
+WDS is a type of string kernels, which take two sequences as inputs and output a scalar value which represents the similarities between the two sequences.  
 WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
@@
 Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
-For example, for two core sequences \textit{ACCTGA} and \textit{CCTGAA}, there is a common part \textit{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.
+This could be useful when there is a shift between two sequences.
+For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. 
+% there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.
 
-Let $\mathbb{I}$(A) is the indicator function, which equals 1 if $A$ is true and 0 otherwise. 
-Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ counts the matches of substrings of length $d$ between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
+We now define WDS kernel. 
+Let $\mathbb{I}$(A) be the indicator function, which equals 1 if $A$ is true and 0 otherwise. 
+Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ indicates whether two substrings of length $d$ matches, between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
 This is similarly done for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
@@
 $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting parameter over the position in the
-sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
-range at position $l$.
+sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the maximum shift
+at position $l$. 
+In our design, the length of RBS sequence $L = 20$. we chose $\ell=6$, and $S(l)= 1$ for all $l$.
 
@@
 
-For recommendations of RBS sequences that should be experimentally labeled next, we have used the \textit{Upper Confidence Bound (UCB)} algorithm.
-On one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
-on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted SD).
-The UCB algorithm provides such $\textit{exploitation-exploration balance}$ by balancing the predicted mean and SD.
+To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides such $\textit{exploitation-exploration balance}$ by balancing the predicted mean and standard deviation.
+On the one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
+on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
 More precisely, UCB algorithm selects RBS sequences with the maximum upper confidence bound at round $t$, i.e.
@@
 where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation (SD) at round $t$ for the sequence $\mathbf{x}_i$.
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
 We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
@@
 one-by-one) and then wait for the label to be tested and used to improve the model.
-Instead, we can recommend RBS sequences in a batch of size $n$. 
+Instead, we can recommend RBS sequences in a batch of size $m$. 
 One naive approach is to 
 %train GP on sequences with known TIR labels and
-recommend sequences in design space with top $n$ UCB scores, as shown in Figure \ref{fig:batch rec}(a) ($n = 2$).
+recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}(a) ($m = 2$).
 However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
-However, since we assume similar sequences would have similar labels (e.g.
+However, since GPR assumes similar sequences would have similar labels (e.g.
 by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
@@
     \end{subfigure}
-    \caption{Batch Recommendation.
+    \caption{Batch Recommendation Illustration.
 We use the batch size of 2, with 5 initial observations.
-The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e.
+The design space is 24 uniformly distributed points in the range [-2,10], i.e.
 {-2, -1.5, -1, ..., 9.5, 10}.
@@
     (a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
-    (b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
+    (b)(c) Batch UCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
     \label{fig:batch rec}
@@
 
-A key property of Gaussian Process regression is that the predictive variance  depends only on observed points (i.e. features), but not on the labels of those observed points. 
+A key property of Gaussian Process regression is that the predictive standard deviation depends only on observed points (i.e. features), but not on the labels of those observed points. 
 % We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
 % As we can see, the predicted confidence interval is the same for the two plots. 
-We can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
-That is, we can recommend sequences sequentially by updating the UCB score with the updated predicted SD with the the previously recommended points.
+One can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
+The strategy is to recommend sequences sequentially by updating the UCB score with the updated predicted standard deviation with the previously recommended points.
 % Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
-% The plot shows the predicted mean and variance after observing 4 data points. 
-First, GP-BUCB  recommends the data point with maximum UCB score based on the predictions over initial 5 observations as shown in Figure \ref{fig:batch rec}(b).  
-Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e.
-observation), and update the predicted variance and then we finally update the UCB scores.
-We then recommend the second data point based on the new UCB scores.
-As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted variance of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
-By using GP-BUCB, we can ensure that the sequences with high similarities will not be recommended in the same batch hence increasing the exploration efficiency.
+% The plot shows the predicted mean and standard deviation after observing 4 data points. 
+As illustrated in Figure \ref{fig:batch rec} (b), the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
+Then we add the recommended data point ($x = 2$) into the training dataset with the predicted mean of that point as label (note it is not the true label, i.e.
+observation), and update the predicted standard deviation and then we finally update the UCB scores.
+The second data point is then recommended based on the new UCB scores.   
+As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
+As summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design area so that the information gain is maximised.

### 2021-05-17 paper/main.tex
fix unicode chat U+2212 error
mengyan reviewed sec 4.2
mengyan edit on several comments

@@
 
+\newcommand{\cheng}[1]{\textcolor{purple}{{\bf Cheng:~}#1}}
 \newcommand{\mengyan}[1]{\textcolor{magenta}{#1}}
@@
 \bibliography{ref.bib}
-\DeclareUnicodeCharacter{2212}{-}
+% \DeclareUnicodeCharacter{2212}{-}
 \begin{document}
@@
 
-Machine learning has been also used for prediction in proteins \cite{Yang2018}.
+% Machine learning has been also used for prediction in proteins \cite{Yang2018}.
 However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
 The challenges of applying machine learning techniques on synthetic biology have two folds.
-On the one hand, there is no large-scale and high-quality labelled data. The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
-One the other hand, it is time-consuming and expensive to measure the TIR for RBS sequences \mengyan{Maciej: can you add more details or reference here}. 
+On the one hand, there is no large-scale and high-quality labelled data. 
+The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
+And it is time-consuming and expensive to measure the TIR for RBS sequences. \mengyan{Maciej: can you add more details or reference here}.
+One the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.
 We addressed the challenges by applying DBTL workflow. 
 Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
+Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. 
 % For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
@@
 % And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
-\mengyan{why?
-we need to be more precise.
-Mengyan can add it later.
-Make it a paragraph, showing the challenges in learning with small number of noisy data and exploration-exploitation balance, and explain how we address that.}
 
@@
 At the same time we were able to discover new RBS sequences with very high TIRs \mengyan{Maciej: can you make "very high" concrete?}.
+\cheng{I suggest reporting everything that is 0.95 and above as "high" (assuming the ratio normalized TIR scale)}
 
-The \textbf{Story} (3 sentences):\\
+The \textbf{Story} (3 sentences):
 \begin{enumerate}
@@
         \item diverse library
-        \item edit distance > 2
+        \item edit distance $>$ 2
     \end{enumerate}
@@
     \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
-    \caption{\textbf{TIRs of RBS groups examined in this study.} 
+    \caption{\cheng{May need to use subfigure to label the figures as A,B,C,D.
+I recall a discussion that C colours with viridis makes it hard to see the different histograms.
+But let's keep to the viridis colours.
+Also, if it is not too hard, make the bin edges [0, 0.1, 0.2, ...].}
+    \textbf{TIRs of RBS groups examined in this study.} 
     The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.
@@
     \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
-The size of the circle represents the experimentally obtained TIR for measured groups.}
+The size \cheng{(area or diameter?, should be area)} of the circle represents the experimentally obtained TIR for measured groups.}
     \label{fig: Swarmplot and Quantplot}
@@
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-\mengyan{explain the TIR ratio we used.}
+We measure the TIR of reference RBS as experimental control. 
+We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of reference sequences in each round.
 The Bandit-0 group performed poorly, due to being trained based on literature data.
@@
 Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
+Note the TIR values reported in Figure \ref{fig: Scatterplot} is normalised TIR based on the standardisation descried in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
@@
 % Interestingly there wasn't much difference in fitness parameters between Round 1-3.
-One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
-\mengyan{Mengyan to explain more about the prediction v.s. recommendation.}
-
-\mengyan{Some other points need to address: 
-1) the TIR label we used here is different from the TIR ratio in Fig 2. 
-2) the setting of predictor.}
+% One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
+One interesting point is that the predictions are also influenced by our recommendation choices. 
+In each round, we selected parts of points for exploration (with high predicted uncertainty), which means the predictors are unlikely to give accurate predicted TIR mean for those points due to lack of observations in data points lying in those areas.
+However, for those explorative points, although the predictors give poor prediction for the predicted TIR, the recommendation choices do allow the predictor to gain new information and learn about those unexplored areas.
 
@@
     \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF.pdf}
-    \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each of the rounds.
+    \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm (GPR with WDS kernel, with maximum kmer as 6 and maximum shift as 1) calculated after each of the rounds.
+    The parameters settings are 
     R\textsuperscript{2}, Spearman Correlation coefficient and p value are all provided for each plot.}
@@
 The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
-Essentially, there are two possible reasons for this, one is due to the algorithm and one due to nature.
-If the problem is with the algorithm, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
-It might be also possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
-In that case, changes in the algorithm will not help.
-Another problem we would like to tackle is the not always good prediction of the TIR of a given RBS.
-\mengyan{Mengyan to address exploration purpose here.}
-In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
+Essentially, there are two possible reasons for this, one due to nature and one is due to the algorithm.
+One the one hand, it might be possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
+In that case, improves in the algorithm will not help.
+On the other hand, the algorithms might be able to be improved to produce better predictions and recommendations. 
+% In terms of the algorithms, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
+% Another problem we would like to tackle is the not prediction of the TIR of a given RBS.
+% In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
+There are open questions need to be addressed for applying machine learning in synthetic biology in general.
+For example, 
+how can we understand the language of RBS sequences better?
+Given a small amount of RBS sequences, how can machine learning algorithm provides more accurate predictions and uncertainty measurement? 
+How should we choose the parameters which controls the exploration-exploitation rate?
 

### 2021-05-17 paper/methods.tex
mengyan edit on several comments

@@
 \subsubsection{Data pre-processing}
+\label{sec: method data pre-procesing}
 
@@
 The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
-\mengyan{maybe we need to introduce spectrum kernel as well since we used it for round 0.}
 To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.

### 2021-05-17 paper/supplementary.tex
fix bugs for local env
add challenges in result sec
update paper plots
Updates from Overleaf
mengyan reviewed sec 4.2
mengyan edit on several comments

@@
 The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
-    \label{fig:my_label}
 \end{figure}
@@
 Colours indicate different groups, shapes indicate different clusters.}
-    \label{fig:my_label}
 \end{figure}
@@
 \subsection{Gaussian Process Regression}
+\label{supp: GPR}
 
@@
 \subsubsection{Normalisation of Kernel}
+\label{supp: Normalisation of Kernel}
 
@@
 \begin{align}
-\label{Eq: GPUCB}
     \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
+    \tag{\ref{Eq: GPUCB}}
 \end{align}
@@
 
-\subsection{Design Pipeline}
+\subsection{Machine Learning Design Pipeline}
+In this section, we report in details about our machine learning design pipeline.
+The raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
 
-The n + 1 round design is based on the nth round result, where each sequence has 6 replicates with TIR labels.
-We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively, where TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
+\mengyan{explain somewhere why we have a different pipeline for round 0.}\\
+\textbf{Bandit-0: } the design of round-0 is based on the literature data \cite{jervis2018machine}.
+We first normalise the raw TIR to values between 0 and 1. 
+We applied the Gaussian Process Regression with noise parameter $\alpha = 1e-10$. 
+We chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+The design size is 60 with UCB parameter $\beta = 1$.\\
+\textbf{Bandit-1 to Bandit-3: }
+the t + 1 round design is based on the $t^th$ ($t \geq 1$) round result, where each sequence has 6 replicates with TIR labels.
+We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively.
 After normalisation, each replicate has zero mean and unit variance. 
-
 For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
@@
 We choose $l=6, s= 1, \alpha = 2$ for the second round design. 
-
 For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
@@
 We recommend 90 sequences from the design space.
-
-\subsection{Intuition behind UCB and visualisation}
-
-\begin{itemize}
-    \item Exploitation and exploration explanation.
-    \item Visualise coverage by clustering plot.
-    \item Table for in-clustering mean and variance.
-\end{itemize}
-
-\subsection{Result analysis}
-
-\begin{itemize}
-    \item violinplot.
-    \item regression performance plot, table.
-    \item kernel matrix plot.
-\end{itemize}
 
@@
 
-\input{paper/mixture_absGau}
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{plots/Supplementary/All_logo.pdf}
+    \caption{\textbf{Sequence logo for all tested sequences.} }
+    \label{fig: All_logo}
+\end{figure}
+
+\input{mixture_absGau}

### 2021-05-17 paper/main.tex
Updated paper to new story.

@@
 
-The \textbf{Story} (3 sentences):
-\begin{enumerate}
-    \item we automate the design of sequences for finding high TIR RBS
-    \begin{enumerate}
-        \item DBTL
-    \end{enumerate}
-    \item the ML method discovers biological knowledge (validates existing knowledge), including the highest TIR ever
-    \begin{enumerate}
-        \item high G
-        \item diverse library
-        \item edit distance $>$ 2
-    \end{enumerate}
-    \item BioFoundry helps ML (good data), and ML helps BioFoundry (more automation)
-    \begin{enumerate}
-        \item Sequences have small technical variation
-        \item UCB much more efficient than random
-    \end{enumerate}
-\end{enumerate}
+\section{Results}
 
-\section{Results}
+\subsection{The experimental workflow}
 
@@
 % with the machine learning process emphasised.
-BUILD and TEST are driven by choices made by human researchers.
-Machine learning algorithms are applied in two parts.
+BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
+Machine learning algorithms are applied in DESIGN AND LEARN.
 In the DESIGN phase, we designed experimental design policy to recommend RBS sequences in batches so that we can identify RBS sequences with highest possible TIR. 
 % We use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
-In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of different RBS sequences.
-The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation.
+In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
+The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation.\\
 
@@
 
-\subsection{Performance of the recommendation algorithm}
+To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
+This way we were able to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
+Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
+In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
+The plasmids were then transformed and the resulting transformants were tested using microplate reader.
+Vast majority of reactions were prepared using liquid handling equipment.
+Similarly, colony picking was done by an automated colony picker.\\
+
+\subsection{Design of the investigated genetic device}
+
 In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
@@
 
+\subsection{Performance of the recommendation algorithm}
+
 The design recommendations were made using the Multi-armed Bandit algorithm.
-In short, the this algorithm is a stochastic method of probing of the experimental space. 
+In short, this algorithm is a stochastic method of probing of the experimental space. 
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
@@
 3) in tSNE plot, address the coverage of the whole space. Maybe also link to the exploration and exploitation balance.}
-
-\subsection{Automated RBS variant construction}
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
-    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
-TIR is given in terms of raw number.}
-    \label{fig: SD}
-\end{figure}
-
-\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this).
-If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
-
-To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
-This way we were able to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
-Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
-In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
-The plasmids were then transformed and the resulting transformants were tested using microplate reader.
-Vast majority of reactions were prepared using liquid handling equipment.
-Similarly, colony picking was done by an automated colony picker.\\
-Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).
-\mengyan{Maciej: can you explain what is "given samples standard deviation"?}
-The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
-Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.
 
@@
 
+\subsection{Advantages of pairing Machine Learning with lab automation}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
+    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
+TIR is given in terms of raw number.}
+    \label{fig: SD}
+\end{figure}
+
+\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this).
+If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
+
+Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).
+\mengyan{Maciej: can you explain what is "given samples standard deviation"?}
+The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
+Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.
+
 \section{Discussion}

### 2021-05-18 paper/main.tex
Updated Paper Results.

@@
 This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
-At the same time we were able to discover new RBS sequences with very high TIRs \mengyan{Maciej: can you make "very high" concrete?}.
-\cheng{I suggest reporting everything that is 0.95 and above as "high" (assuming the ratio normalized TIR scale)}
+At the same time we were able to discover new RBS sequences with very high TIRs from between 95 to 135\% of TIR of our chosen very strong benchmark RBS.
 
@@
 We show our DBTL workflow in Figure \ref{fig: Flowchart}.
-% with the machine learning process emphasised.
 BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
@@
 In the DESIGN phase, we designed experimental design policy to recommend RBS sequences in batches so that we can identify RBS sequences with highest possible TIR. 
-% We use the Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be Built and Tested next.
 In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
@@
 Additionally, there is a consensus RBS core sequence called the \textit{Shine-Dalgarno sequence}, which in \emph{E. coli} is \textbf{AGGAGG}. 
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig: Flowchart}). \mengyan{Maciej: what does this sentence mean?}
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig: Flowchart}).
+\mengyan{Maciej: what does this sentence mean?} \maciej{It means that we consider the RBS to be 20bp, but we will be focusing on the 6bp core region.}
 
 Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, 
-which is a known RBS with high \mengyan{Maciej: can we claim "highest"?} TIR that comes with the pBb series plasmids \cite{Lee2011}. 
+which is a known RBS with high \mengyan{Maciej: can we claim "highest"?} \maciej{Due to poor testing methods that were used for TIR previously we can't cause we don't know how do we compare exactly} TIR that comes with the pBb series plasmids \cite{Lee2011}.
 Since this is the sequence against which new RBS sequenced will be benchmarked,
@@
 Another feature of the bandits algorithms is that it balances two parts: the exploration in terms of potential high TIRs in unknown areas, and exploitation in terms of querying areas which is known to give relatively high TIRs.
-% The two approaches can be controlled with the \textbeta\enspace parameter. 
-% We have decided that in the first iterations of the cycle it would be beneficial to skew the algorithm towards the exploration with exploitation taking increasing role in later iterations. 
 One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). 
@@
     \textbf{TIRs of RBS groups examined in this study.} 
-    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard.
-    \mengyan{Maciej: make the sentence more clear?}
+    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
     \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
@@
     Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
-    SD - Shine-Dalgarno sequence. Dash line is set 1 and represents the averaged benchmark sequence TIR for that group. 
-    \mengyan{explain why reference sequences are not all one.}
+    SD - Shine-Dalgarno sequence.
+    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
+    BN - benchmark sequences for all plates. 
+    They are not all exactly 1 due to them being shown as separate samples rather than averages.
+    \maciej{Can we add a "BN" letters similar to the SD near the red group?.}
     \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
     % save the random groups which were shown together due to similar distributions.
-    UNI and PPM are merged into Random group. 
-    \mengyan{Maciej: maybe need to address why there are different groups in different plot, e.g. why no BPS-NC.}
+    UNI and PPM are merged into Random group and BPS-NC was removed due to changes being made outside the core in this group. 
         \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
@@
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (Figure \ref{fig:core_vs_noncore} \mengyan{put it in main paper?}).
+\maciej{I still feel it is not important enough to put in here.}
     \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
@@
 
-Figure \ref{fig: Swarmplot and Quantplot} shows the results for all the examined groups. 
+Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
 % Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
 % Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
+In each round, we measure the TIR of benchmark RBS as internal standard. 
+We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-We measure the TIR of reference RBS as experimental control. 
-We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of reference sequences in each round.
-The Bandit-0 group performed poorly, due to being trained based on literature data.
+The Bandit-0 group performed poorly, due to being trained on literature data.
+As mentioned before, that data was also exploring the strength of different RBS, however the anatomy of the RBSs in that study was different from ours.
+Additionally, the data set was relatively small and covered very small part of the possible experimental space.
 \mengyan{Maciej: can you add some analysis on literature data?
@@
 And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
+\maciej{I have added some points talking about that literature data, but I feel we can't add too much - it's not that important for the reader and we need to watch the word limit.}
 However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
@@
 Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
-
+Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
+The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
+The decreased result in the 0.9th qunatile compared to the max value for Bandit3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
+Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
+KDE plots have been overlaid to depict the calculated density for each group.
+The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the >0.8 TIR ratio bins.
+Finally, in \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
+Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
+A number of clusters were targeted by our recommendation algorithm.
 \mengyan{the analysis below only focuses on the swarmplot, we need to address other plots as well. Some points to address: 
@@
 3) in tSNE plot, address the coverage of the whole space. Maybe also link to the exploration and exploitation balance.}
+\maciej{I've added description of all the subplots, will add the cluster descriptions after we discuss it on Wednesday.}
 
@@
 
-Figure \ref{fig:Library characteristics} A shows the sequence logo calculated for the Top 30 \mengyan{Maciej: Top 30 of what? raw TIR? TIR ratio?} of our sequences.
-The expected bias \mengyan{Maciej: what is expected bias?} towards guanine is clearly visible for all positions.
-This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot} D reinforces the notion that our algorithm successfully identified sequences with high TIR probability.
+Figure \ref{fig:Library characteristics} A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
+It is generally understood that guanine rich sequences are promoting strong transcription.
+This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
+This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G sequences as ones with high TIR probability.
 

### 2021-05-18 paper/main.tex
mengyan address comments

@@
 Additionally, there is a consensus RBS core sequence called the \textit{Shine-Dalgarno sequence}, which in \emph{E. coli} is \textbf{AGGAGG}. 
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region (see detail in Figure \ref{fig: Flowchart}).
-\mengyan{Maciej: what does this sentence mean?} \maciej{It means that we consider the RBS to be 20bp, but we will be focusing on the 6bp core region.}
+Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region
+(see detail in Figure \ref{fig: Flowchart}).
+% we consider the RBS to be 20bp, but we will be focusing on the 6bp core region.
 
 Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, 
-which is a known RBS with high \mengyan{Maciej: can we claim "highest"?} \maciej{Due to poor testing methods that were used for TIR previously we can't cause we don't know how do we compare exactly} TIR that comes with the pBb series plasmids \cite{Lee2011}.
+which is a known RBS with high TIR that comes with the pBb series plasmids \cite{Lee2011}. 
+% Due to poor testing methods that were used for TIR previously we can't claim "highest" cause we don't know how do we compare exactly
 Since this is the sequence against which new RBS sequenced will be benchmarked,
@@
     \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
-This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (Figure \ref{fig:core_vs_noncore} \mengyan{put it in main paper?}).
-\maciej{I still feel it is not important enough to put in here.}
+This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (Figure \ref{fig:core_vs_noncore}).
+    % \mengyan{put it in main paper?}
     \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
@@
 Additionally, the data set was relatively small and covered very small part of the possible experimental space.
-\mengyan{Maciej: can you add some analysis on literature data?
-to verify 1) our library is "better" 2) it's hard to do ML on available library.
-some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
-And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
-\maciej{I have added some points talking about that literature data, but I feel we can't add too much - it's not that important for the reader and we need to watch the word limit.}
 However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
@@
 The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
-The decreased result in the 0.9th qunatile compared to the max value for Bandit3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
+The decreased result in the 0.9th qunatile compared to the max value for Bandit-3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
 Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
+\mengyan{Maciej: can you rephrase the last sentence?}
 KDE plots have been overlaid to depict the calculated density for each group.
-The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the >0.8 TIR ratio bins.
+The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
+\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
 Finally, in \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
 Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
+We can see the RBSs recommended by Bandits groups have covered majority of the design space. 
 A number of clusters were targeted by our recommendation algorithm.
-\mengyan{the analysis below only focuses on the swarmplot, we need to address other plots as well. Some points to address: 
-1) in swarmplot and histogram, explain the two-mode is due to the exploration and exploitation. 
-2) in quantile plot, address we care about the top TIRs, that's why we look at the quantiles bigger than 0.5.
-3) in tSNE plot, address the coverage of the whole space. Maybe also link to the exploration and exploitation balance.}
-\maciej{I've added description of all the subplots, will add the cluster descriptions after we discuss it on Wednesday.}
+\mengyan{explain the exploration and exploitation balance?}
 
@@
 
-Figure \ref{fig:Library characteristics} A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
+Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
 It is generally understood that guanine rich sequences are promoting strong transcription.
@@
 
-Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for meaningful improvement in the TIR. 
-\mengyan{Maciej: what's a "meaningful improvement"}
-We define the editing distance as Hamming distance \mengyan{Maciej: please add citation here}, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
-Figure \ref{fig:Library characteristics} B shows what edit distance is required for positive change in TIR for RBS with high and medium TIR. \mengyan{Maciej: what's the definition of high and medium?}
+Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
+We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
+Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
 For RBS with high TIR, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
@@
 Finally, our strong sequences did not show neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
-This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \mengyan{Maciej: can you add one or more citations here?} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.
+This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \mengyan{Maciej: can you add one or more citations here?} \maciej{Already included in the introduction.} \mengyan{but shall we cite here again?
+since it is hard to find which literature you mean.} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
 
 \mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any.
-Comparison with existing library, if any.}
+Comparison with existing library, if any.} \maciej{We can't produce a good library figure, but have added a paragraph below.}
 
@@
 
+Taken together our dataset can be seen as a diverse library of well characterised RBS parts.
 Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).

### 2021-05-18 paper/supplementary.tex
mengyan address comments

@@
 
-\mengyan{explain somewhere why we have a different pipeline for round 0.}\\
+\mengyan{explain somewhere why we have a different pipeline for round 0.}
+\mengyan{add some analysis on literature data?
+to verify 1) our library is "better" 2) it's hard to do ML on available library.
+some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
+And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
+
 \textbf{Bandit-0: } the design of round-0 is based on the literature data \cite{jervis2018machine}.

### 2021-05-18 paper/main.tex
update plots

@@
     \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
-    \caption{\cheng{May need to use subfigure to label the figures as A,B,C,D.
-I recall a discussion that C colours with viridis makes it hard to see the different histograms.
-But let's keep to the viridis colours.
-Also, if it is not too hard, make the bin edges [0, 0.1, 0.2, ...].}
+    \caption{\cheng{May need to use subfigure to label the figures as A,B,C,D.}
     \textbf{TIRs of RBS groups examined in this study.} 
@@
     They are not all exactly 1 due to them being shown as separate samples rather than averages.
-    \maciej{Can we add a "BN" letters similar to the SD near the red group?.}
     \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
@@
     \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
-The size \cheng{(area or diameter?, should be area)} of the circle represents the experimentally obtained TIR for measured groups.}
+    The area size of the circle represents the experimentally obtained TIR for measured groups.}
     \label{fig: Swarmplot and Quantplot}

### 2021-05-19 paper/main.tex
maciej's edit

@@
 \mengyan{Maciej:how about ours? same approach or better?}
-
-The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
-\mengyan{Maciej: can you replace citations like "Reeve \emph{et al.}" to \textcite{Reeve2014}?}
-For example, according to Reeve \emph{et al.} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+\maciej{We shouldn't be making comments on ours in the introduction.}
+The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} \maciej{Definitely Learn, Test is only about gathering the data and it has its own limitaions}- there is very limited access to methods and software that can improve and understand designs based on the experimental results.
+For example, according to Reeve \emph{et al.} \textcite{Reeve2014}there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Reported predictions from all of these models are relatively good ($R^2 >0.8$), \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g.
 is that because the data?
-So I recommend either remove this or explain it somewhere.}
+So I recommend either remove this or explain it somewhere.} \maciej{I'll explain it in the discussion}
 but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
@@
 New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
-These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018}.
+These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020}.
 In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
@@
 A number of clusters were targeted by our recommendation algorithm.
+For example, the cluster positioned in the top centre labelled as "G rich" has been actively recommended by the algorithm.
+More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
 \mengyan{explain the exploration and exploitation balance?}
@@
 In such cases, a directed process should be strongly encouraged.
-This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed.
-\mengyan{Maciej: can you add one or more citations here?}
+This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.
 
@@
 If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
+\maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}
 
 Taken together our dataset can be seen as a diverse library of well characterised RBS parts.
-Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given samples standard deviation (coefficient of variation).
-\mengyan{Maciej: can you explain what is "given samples standard deviation"?}
+Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given RBS standard deviation (coefficient of variation).
+We have decided to sample biological over technical replicates as this gives better representation of how different populations of a given RBS variant should perform.
+We also decided to do 6 replicates of each sample rather than the more established 3 because we felt it gives us better representation both in terms of the calculated mean and standard deviation.
 The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
@@
 We performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.
+By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.\\
 
@@
 On the other hand, the algorithms might be able to be improved to produce better predictions and recommendations. 
+Our predictions are robust even though R2 lower than others.
 % In terms of the algorithms, it might due to its inability to predict TIRs higher than the ones it has been already presented with.

### 2021-05-19 paper/main.tex
change colors, setup subfigure caption, add annotation to tsne

@@
 \newcommand{\maciej}[1]{\textcolor{blue}{#1}}
+
+% \captionsetup[subfigure]{position=top, labelfont=bf,textfont=normalfont,singlelinecheck=off,justification=raggedright}
+\captionsetup[subfigure]{font={bf,small}, skip=1pt, singlelinecheck=false}
+\renewcommand{\thesubfigure}{\Alph{subfigure}}
 
@@
     \centering
-    \includegraphics[scale=0.4]{plots/Main_Paper/swarmplot.pdf}
-    \includegraphics[scale=0.4]{plots/Main_Paper/quantplot.pdf}
-    \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
-    \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
-    \caption{\cheng{May need to use subfigure to label the figures as A,B,C,D.}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.3]{plots/Main_Paper/tsneplot.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
+    \end{subfigure}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
+    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
+    \caption{
     \textbf{TIRs of RBS groups examined in this study.} 
@@
     \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
-    The area size of the circle represents the experimentally obtained TIR for measured groups.}
+    The area size of the circle represents the experimentally obtained TIR for measured groups.
+    \textbf{E)} Exploitation v.s. Exploration for Bandit 1-3. Colors approaching to blue represents exploitation, otherwise represents exploration.}
     \label{fig: Swarmplot and Quantplot}
@@
          \centering
+         \caption{}
          \includegraphics[scale=1.2]{plots/Main_Paper/TOP30_logo.pdf}
-         \caption{}
      \end{subfigure}
@@
          \centering
+         \caption{}
          \includegraphics[scale=0.5]{plots/Main_Paper/Hd_Heatmap.png}
-         \caption{}
      \end{subfigure}

### 2021-05-19 paper/methods.tex
change colors, setup subfigure caption, add annotation to tsne

@@
 %train GP on sequences with known TIR labels and
-recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}(a) ($m = 2$).
+recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}(A) ($m = 2$).
 However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
@@
         \centering
+        \caption{}
         \includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}
@@
         \centering
+        \caption{}
         \includegraphics[scale=0.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}
@@
         \centering
+        \caption{}
         \includegraphics[scale=0.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}
@@
     The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
-    (a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
-    (b)(c) Batch UCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
+    (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
+    (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that.}
     \label{fig:batch rec}
@@
 % The plot shows the predicted mean and standard deviation after observing 4 data points. 
-As illustrated in Figure \ref{fig:batch rec} (b), the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
+As illustrated in Figure \ref{fig:batch rec} (B), the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
 Then we add the recommended data point ($x = 2$) into the training dataset with the predicted mean of that point as label (note it is not the true label, i.e.
@@
 The second data point is then recommended based on the new UCB scores.   
-As we can see in Figure \ref{fig:batch rec}(c), since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
+As we can see in Figure \ref{fig:batch rec}(C), since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
 As summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design area so that the information gain is maximised.

### 2021-05-20 paper/main.tex
Updated paper.

@@
 The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.
-\mengyan{Maciej:how about ours? same approach or better?}
-\maciej{We shouldn't be making comments on ours in the introduction.}
-The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} \maciej{Definitely Learn, Test is only about gathering the data and it has its own limitaions}- there is very limited access to methods and software that can improve and understand designs based on the experimental results.
-For example, according to Reeve \emph{et al.} \textcite{Reeve2014}there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
-Reported predictions from all of these models are relatively good ($R^2 >0.8$), \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g.
+
+The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
+For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
+% \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g.
 is that because the data?
@@
     The area size of the circle represents the experimentally obtained TIR for measured groups.
-    \textbf{E)} Exploitation v.s. Exploration for Bandit 1-3. Colors approaching to blue represents exploitation, otherwise represents exploration.}
+    \textbf{E)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration.}
     \label{fig: Swarmplot and Quantplot}
@@
 \mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
-Finally, in \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
+In \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
 Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
@@
 A number of clusters were targeted by our recommendation algorithm.
-For example, the cluster positioned in the top centre labelled as "G rich" has been actively recommended by the algorithm.
+For example, the circled clusters 
+% positioned in the top centre
+labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
 More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
-\mengyan{explain the exploration and exploitation balance?}
+Finally, in Figure \ref{fig: Swarmplot and Quantplot} E) we coloured the swarmplot for Bandit 1-3 groups as the scores representation the exploration-exploitation level.
+The scores are calculated based on the projection of prediction standard deviation and mean pairs (for details see Supplementary).
+We can see the RBSs with high TIRs (blue, with high predicted mean) tend to exploit the design space; the explorative points (red, with high predicted standard deviation) gives relatively low TIR but expands the knowledge of about the unknown design space.
+Another observation is that a higher ratio of the points in Bandit-3 compared the other two groups are exploitative. 
+This is an expected behaviour of our algorithm since we have set a large exploitation rate for the last group (Bandit-3).
 
@@
 
-\mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any.
+% REVIEW:
+% \mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any.
 Comparison with existing library, if any.} \maciej{We can't produce a good library figure, but have added a paragraph below.}
@@
 
-\begin{figure}[!ht]
+\begin{figure}
      \centering
-    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
-    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
-TIR is given in terms of raw number.}
-    \label{fig: SD}
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \caption{}
+         \includegraphics[scale=0.4]{plots/Main_Paper/SD.pdf}
+     \end{subfigure}
+     \hfill
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \caption{}
+         \includegraphics[scale=0.6]{paper/plots/Main_Paper/SDvsTIR.pdf}
+     \end{subfigure}
+     \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
+B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR.
+Standard deviation and average TIR is given in terms of raw number.}
+     \label{fig:Library characteristics}
 \end{figure}
 
-\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this).
+% REVIEW:
+% \mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this).
 If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
-\maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}
+% \maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}
 

### 2021-05-21 paper/methods.tex
mengyan re-format fig 2, create table 1

@@
 Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
-***) with 34$\mu$g/mL chloramphenicol (Sigma, catalogue no.
-***).
+\mengyan{***}) with 34$\mu$g/mL chloramphenicol (Sigma, catalogue no.
+\mengyan{***}).
 Plates were incubated overnight at 37$^{\circ}$C. 
+8
 
@@
 In the following, we describe our machine learning pipeline (Bandit 1-3) by the order of data pre-processing, prediction and string kernels, recommendation. 
-\mengyan{The pipeline for Bandit-0 is reported in Supplementary XX.}
+The pipeline for Bandit-0 is reported in Supplementary \ref{sec: ML design pipeline}.
 
@@
 For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
-\mengyan{Table XXX shows an example of our raw data. }
+% REVIEW:
+% \mengyan{Table XXX shows an example of our raw data. }
 % We consider both the label (TIR) pre-processing and kernel pre-processing.
@@
 % This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g. time, temperature, operator, etc.). 
-\mengyan{we should discuss measurement noise somewhere, maybe with a sample example for the dataset we used.}
+% REVIEW:
+% \mengyan{we should discuss measurement noise somewhere, maybe with a sample example for the dataset we used.}
 
-For regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004}, which provides not only the predicted TIR scores, but also the uncertainty level of the predictions.
+For regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004}.
 We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
@@
 % As we can see, the predicted confidence interval is the same for the two plots. 
-One can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2012parallelizing}.
+One can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
 The strategy is to recommend sequences sequentially by updating the UCB score with the updated predicted standard deviation with the previously recommended points.

### 2021-05-21 paper/main.tex

@@
 \usepackage{authblk}
+% \usepackage[table,xcdraw]{xcolor}
 % \usepackage{subfloat}
@@
 
-\author[1,2]{Zhang M.}
+\author[1,2,4]{Zhang M.}
 \author[3]{Holowko M. B.}
@@
 \affil[1]{Machine Learning and Artificial Intelligence Future Science Platform, CSIRO}
-\affil[2]{Australian National University}
+\affil[2]{Department of Computer Science, Australian National University}
 \affil[3]{CSIRO Synthetic Biology Future Science Platform, CSIRO Land and Water}
@@
 Notably, a high number of these sites demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs.
-Additionally, this work elucidated some important RBS design guidelines.
 
@@
 RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2012parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
+In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
 Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
@@
 BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
-Machine learning algorithms are applied in DESIGN AND LEARN.
-In the DESIGN phase, we designed experimental design policy to recommend RBS sequences in batches so that we can identify RBS sequences with highest possible TIR. 
+Machine learning algorithms are applied in LEARN and DESIGN.
 In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
-The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation.\\
+The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation process.
+In the DESIGN phase, we have used a multi-armed Bandit recommendation algorithm that was selecting RBS designs to be tested in subsequent experimental batches.
+\\
 
@@
 
-To help us obtain reliable and reproducible results we have employed automation-heavy workflow for our experiments.
+To help us obtain reliable and reproducible results we have employed automation-heavy workflow in the BUILD and TEST phases.
 This way we were able to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
-Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down time to prepare our variants.\\
+Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down the time required to prepare our variants.\\
 In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
@@
 (see detail in Figure \ref{fig: Flowchart}).
-% we consider the RBS to be 20bp, but we will be focusing on the 6bp core region.
 
-Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, 
-which is a known RBS with high TIR that comes with the pBb series plasmids \cite{Lee2011}. 
+Our template RBS sequence, is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA.
+This sequence is a known to have high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
 % Due to poor testing methods that were used for TIR previously we can't claim "highest" cause we don't know how do we compare exactly
@@
 we will refer to this sequence as the \textit{benchmark sequence} from now on.
-In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the consensus sequence, i.e.
+In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e.
 TTTAAGA + NNNNNN + TATACAT, where N can be any choices of A, C, G, T.
@@
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-In our case we use the Upper Confidence Bound algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (GPR).
-Another feature of the bandits algorithms is that it balances two parts: the exploration in terms of potential high TIRs in unknown areas, and exploitation in terms of querying areas which is known to give relatively high TIRs.
-One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of given event occurring (in this case RBS having a specific TIR). 
-As such, it pairs naturally with our prediction algorithm, the Gaussian Process, which provides probability based function regression.
+In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (see below).
+Another feature of the bandits algorithm is that it balances two parts: the exploration of the unknown (untested) parts of the design space where high TIR RBS can be hidden, and exploitation which goal is querying areas which are known to give relatively high TIRs.
+One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of a given event occurring (in this case RBS having a specific TIR). 
+As such, it pairs naturally with our prediction algorithm, the Gaussian Process Regression, which provides probability based function regression.
 
@@
     \centering
-    \begin{subfigure}[b]{0.49\textwidth}
+    \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \caption{}
-        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
+        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
     \end{subfigure}
     % \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
+    \begin{subfigure}[b]{0.25\textwidth}
         \centering
         \caption{}
-        \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
+        \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
     \end{subfigure}
     % \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
+    \begin{subfigure}[b]{0.25\textwidth}
         \centering
         \caption{}
-        \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
+        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot_proj.pdf}
     \end{subfigure}
-    \begin{subfigure}[b]{0.49\textwidth}
+    \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \caption{}
-        \includegraphics[scale=0.3]{plots/Main_Paper/tsneplot.pdf}
+        \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
     \end{subfigure}
-    \begin{subfigure}[b]{0.49\textwidth}
+    \begin{subfigure}[b]{0.48\textwidth}
         \centering
         \caption{}
-        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
+        \includegraphics[scale=0.42]{plots/Main_Paper/tsneplot.pdf}
     \end{subfigure}
@@
     UNI and PPM are merged into Random group and BPS-NC was removed due to changes being made outside the core in this group. 
-        \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
-    \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
-    The area size of the circle represents the experimentally obtained TIR for measured groups.
-    \textbf{E)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration.}
+    \textbf{C)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration. 
+    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
+    \textbf{E)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
+    The area of the circle represents the experimentally obtained TIR for measured groups.}
     \label{fig: Swarmplot and Quantplot}
@@
 \begin{itemize}
-    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide changes of all 20 nucleotides of the original, consensus sequence.
+    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide variations of all 20 nucleotides of the original, consensus sequence.
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (Figure \ref{fig:core_vs_noncore}).
     % \mengyan{put it in main paper?}
-    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing either nucleotide for each position. 
+    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. 
     \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}.
@@
 % Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
-In each round, we measure the TIR of benchmark RBS as internal standard. 
+In each round, we measure the TIR of benchmark RBS as the internal standard. 
 We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-The Bandit-0 group performed poorly, due to being trained on literature data.
-As mentioned before, that data was also exploring the strength of different RBS, however the anatomy of the RBSs in that study was different from ours.
-Additionally, the data set was relatively small and covered very small part of the possible experimental space.
-However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
+The Bandit-0 group performed poorly, despite being machine learning driven, due to being trained on literature data.
+However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
 In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
@@
 Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
-The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
-The decreased result in the 0.9th qunatile compared to the max value for Bandit-3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
-Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
-\mengyan{Maciej: can you rephrase the last sentence?}
+The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
+The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
+We see that effect in Figure \ref{fig: Swarmplot and Quantplot} c), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+We can see the RBSs with high TIRs (blue, with high predicted mean) tend to come from exploitation the design space whereas the explorative points (red, with high predicted standard deviation) give relatively low TIR but expand our knowledge about the unknown part of the design space.
+Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
 KDE plots have been overlaid to depict the calculated density for each group.
@@
 \mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
-In \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
+\maciej{I did that in a later section.}
+\mengyan{I recommend to mention the observation and point to where you discuss it.}
+In \ref{fig: Swarmplot and Quantplot}E we show a tSNE plot depicting the experimental space.
 Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
-We can see the RBSs recommended by Bandits groups have covered majority of the design space. 
+We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
 A number of clusters were targeted by our recommendation algorithm.
-For example, the circled clusters 
-% positioned in the top centre
-labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
+For example, the circled clusters labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
 More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
-Finally, in Figure \ref{fig: Swarmplot and Quantplot} E) we coloured the swarmplot for Bandit 1-3 groups as the scores representation the exploration-exploitation level.
-The scores are calculated based on the projection of prediction standard deviation and mean pairs (for details see Supplementary).
-We can see the RBSs with high TIRs (blue, with high predicted mean) tend to exploit the design space; the explorative points (red, with high predicted standard deviation) gives relatively low TIR but expands the knowledge of about the unknown design space.
-Another observation is that a higher ratio of the points in Bandit-3 compared the other two groups are exploitative. 
-This is an expected behaviour of our algorithm since we have set a large exploitation rate for the last group (Bandit-3).
 
@@
 
-Our recommendation algorithm selects new designs based on our prediction algorithm.
+Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
 This algorithm creates a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
@@
 % Such a stochastic prediction method fits biological processes naturally, since they are highly stochastic as well. 
-The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.
-
-A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data point, in our case RBS sequences.
+The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
+A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data points, in our case RBS sequences.
 Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.
 For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-For subsequent rounds, we used the a more powerful kernel function, the \textit{weighted degree kernel with shift} (wds) \cite{ratsch_rase_2005_wds}, which has been shown to have a strong performance in various prediction tasks \cite{Ben-Hur2008}.
-The relative distances between sequences calculated by our kernel function are shown in figure \ref{fig: Swarmplot and Quantplot} D.
+For subsequent rounds, we used 
+% a more powerful \maciej{more powerful in what sense?} kernel function, 
+the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds}, which has been shown to have good performance in various prediction tasks \cite{Ben-Hur2008}.
+The WDS not only counts the matches of substrings of a certain length (i.e. kmers), but also takes into account the positional information and shifting of substrings.
+For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift.
+The relative distances between sequences calculated by our kernel function are shown in figure \ref{fig: Swarmplot and Quantplot}E.\\
 
 Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
-Note the TIR values reported in Figure \ref{fig: Scatterplot} is normalised TIR based on the standardisation descried in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
@@
 % One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
-One interesting point is that the predictions are also influenced by our recommendation choices. 
-In each round, we selected parts of points for exploration (with high predicted uncertainty), which means the predictors are unlikely to give accurate predicted TIR mean for those points due to lack of observations in data points lying in those areas.
-However, for those explorative points, although the predictors give poor prediction for the predicted TIR, the recommendation choices do allow the predictor to gain new information and learn about those unexplored areas.
+One important point to note is that the predictions are also influenced by our recommendation choices. 
+In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
+However, this is still very useful information for future predictions as it allows us to understand the underlying space better.
 
@@
     \centering
-    \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF.pdf}
-    \caption{\textbf{Performances of the predictions.} The scatter plots A-D are showing the performance of our prediction algorithm (GPR with WDS kernel, with maximum kmer as 6 and maximum shift as 1) calculated after each of the rounds.
-    The parameters settings are 
-    R\textsuperscript{2}, Spearman Correlation coefficient and p value are all provided for each plot.}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_0.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_1.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_2.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_3.pdf}
+    \end{subfigure}
+    \caption{\textbf{Predictions generalise to held out data.} The scatter plots A-D are showing the performance of our prediction algorithm (GPR with WDS kernel, with maximum k-mer as 6 and maximum shift as 1) calculated after each of the rounds.
+    Note the TIR values are normalised TIR based on the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
+    The values of 
+    $R^2$ and Spearman correlation coefficient (with corresponding p-value) are all provided for each plot.
+    The p-value decreases with round increases, which shows increasing strong evidence about the Spearman correlation between our prediction and measured TIR. 
+    }
     \label{fig: Scatterplot}
@@
 This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
-This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G sequences as ones with high TIR probability.
+This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as ones with high TIR probability.
 
-\begin{figure}
+\begin{figure}[!t]
      \centering
@@
          \caption{}
-         \includegraphics[scale=0.5]{plots/Main_Paper/Hd_Heatmap.png}
+         \includegraphics[scale=0.5]{plots/Main_Paper/Hd_Heatmap.pdf}
      \end{subfigure}
-     \caption{\textbf{Analysis of RBS positional information} A) Sequence logo calculated for the Top 30 tested sequences.
+     \caption{\textbf{Characteristics of strong RBSs.} A) Sequence logo calculated for the Top 30 tested sequences.
 B) Heatmap showing what edit (Hamming) distance is required for positive change in TIR for RBS with high and medium TIR.
-The temperature scale shows the difference between a given RBS on y axis and the RBS with strongest TIR at the given distance.}
+The temperature scale shows the difference between a given RBS on y axis and the RBS with strongest TIR at the given distance.
+Every second RBS is labelled for increased legibility.
+}
      \label{fig:Library characteristics}
@@
 
+\begin{table}[!h]
+\centering
+\begin{minipage}[c]{0.6\textwidth}
+\centering
+\begin{tabular}{|c|c|}
+\hline
+\textbf{Characteristics of the library}                                                       & \textbf{Statistics} \\ \hline
+Total experimental space                                                                      & 4096                \\ \hline
+Planned constructs                                                                            & 450                 \\ \hline
+Successfully constructed                                                                      & 445                 \\ \hline
+Sequences with CV\textless{}40\%                                                              & 79\%                \\ \hline
+Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
+\begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
+Raw TIR range                                                                                     &      [4.9258, 105.377]               \\ \hline
+TIR ratio range                                                                                     &      [0.0626, 1.3393]               \\ \hline
+\end{tabular}
+\end{minipage}
+\begin{minipage}[c]{0.38\textwidth}
+\centering
+% Please add the following required packages to your document preamble:
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
+\begin{tabular}{|c|c|}
+\hline
+\textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
+GGGGGC                & 1.3393             \\ \hline
+GGGGGT                & 1.1544             \\ \hline
+GGCTAT                & 1.084              \\ \hline
+\textbf{AGGAGA}                & 1                  \\ \hline
+GGCGTT                & 0.9805             \\ \hline
+GGGGGG                & 0.9788             \\ \hline
+GGCGAC                & 0.9758             \\ \hline
+CAGGAG                & 0.9629             \\ \hline
+GGCGAG                & 0.9519             \\ \hline
+\textbf{AGGAGG}                & 0.3941             \\ \hline
+\end{tabular}
+\end{minipage}
+\caption{Characteristics of the library. \mengyan{write caption here}
+For Efficiency of bandit design:
+random = {set of TIR that are in UNI and PPM};
+bandit = {set of TIR that are in bandit 0,1,2,3}.
+The ratio of the max bandit to the max random: $1.34/0.67$ = 2;
+The ratio of the 0.9 quantile bandit to the 0.9 quantile random: $0.82/0.49$ = 1.67.}
+\end{table}
+
 Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
-We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means two completely different sequences).
+We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
 Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
-For RBS with high TIR, the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
-For RBS with medium TIR, a distance of 1 is enough to produce a meaningful increase in TIR.
-That means that as the TIR of examined RBSs increases, 
-exploring sequences which are more dissimilar to the current candidates tends to give meaningful improvement. 
-% the algorithm should be ``encouraged" to explore sequences which are more dissimilar to the current candidates.
-The low rate of natural mutations will be very slow to explore more dissimilar sequences \cite{Lee2012},
+For RBS with high TIR ($>1$), the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
+For RBS with medium TIR ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
+That means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give meaningful improvement. 
+The low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012},
 which indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
@@
 Finally, our strong sequences did not show neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
-This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \mengyan{Maciej: can you add one or more citations here?} \maciej{Already included in the introduction.} \mengyan{but shall we cite here again?
-since it is hard to find which literature you mean.} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
-
-% REVIEW:
-% \mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any.
-Comparison with existing library, if any.} \maciej{We can't produce a good library figure, but have added a paragraph below.}
-
-\subsection{Advantages of pairing Machine Learning with lab automation}
-
-\begin{figure}
-     \centering
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \caption{}
-         \includegraphics[scale=0.4]{plots/Main_Paper/SD.pdf}
-     \end{subfigure}
-     \hfill
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \caption{}
-         \includegraphics[scale=0.6]{paper/plots/Main_Paper/SDvsTIR.pdf}
-     \end{subfigure}
-     \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
-B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR.
-Standard deviation and average TIR is given in terms of raw number.}
-     \label{fig:Library characteristics}
-\end{figure}
-
-% REVIEW:
-% \mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this).
-If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
-% \maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}
-
-Taken together our dataset can be seen as a diverse library of well characterised RBS parts.
-Figure \ref{fig: SD} shows the distribution of relative error for each tested RBS, calculated as the average TIR from 6 biological replicates divided by the given RBS standard deviation (coefficient of variation).
-We have decided to sample biological over technical replicates as this gives better representation of how different populations of a given RBS variant should perform.
-We also decided to do 6 replicates of each sample rather than the more established 3 because we felt it gives us better representation both in terms of the calculated mean and standard deviation.
-The phenomenon of higher volatility of readings in the lower end of fluorescence measurements is well documented and comes from higher influence of small changes on the percentage error in that range.
-Overall, 79\% and 27\% of our samples shows relative error of below 40\% and 20\% respectively which we find satisfactory for the purposes of this study.
+This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
 
@@
 We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
-We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit into our regression method.
-We performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow, we have generated an extensive library of diverse RBS that can be used in the future studies.\\
+We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit well into our prediction method.
+We building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
+By using our workflow, we have found very strong RBS sequences and we have generated an extensive library of diverse RBS that can be used in the future studies.
+We have achieved this despite the relatively low accuracy of our predictions, which means that the presented algorithms are robust and able to identify the right signal in a noisy environment. \\
 
-There is a number of issues that we would like to address in the future.
-The biggest of them is elucidating the reason for not many of our sequences being able to significantly cross the TIR of our benchmark sequence.
-Essentially, there are two possible reasons for this, one due to nature and one is due to the algorithm.
-One the one hand, it might be possible that the TIR seen in the benchmark is simply already one of the strongest possible and represents a natural ceiling for it. 
-In that case, improves in the algorithm will not help.
-On the other hand, the algorithms might be able to be improved to produce better predictions and recommendations. 
-Our predictions are robust even though R2 lower than others.
-% In terms of the algorithms, it might due to its inability to predict TIRs higher than the ones it has been already presented with.
-% Another problem we would like to tackle is the not prediction of the TIR of a given RBS.
-% In principle, the predictions should improve with more data fed into the algorithm, but that would be against the goal of the study which is to minimise the number of data points that have to be obtained.
-There are open questions need to be addressed for applying machine learning in synthetic biology in general.
-For example, 
-how can we understand the language of RBS sequences better?
-Given a small amount of RBS sequences, how can machine learning algorithm provides more accurate predictions and uncertainty measurement? 
-How should we choose the parameters which controls the exploration-exploitation rate?
+There are still open questions that need to be addressed for applying machine learning in synthetic biology.
+For example, how can we understand the language of RBS sequences better?
+In other words, can we extract more biologically important information from the decisions made by our algorithms?
+Another example would be - given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
+Finally, another question with high impact on the performance of the workflow is - how should we choose the parameters which controls the exploration-exploitation rate? \\
+
+We have found our approach bringing machine learning and synthetic biology experts very powerful.
+We envision to that the pairing of machine learning with high-throughput automation will keep delivering a high number of good quality datasets nad improved methods for biological engineering.\\
 
 In the future, we hope to extend the algorithm to other, more complicated genetic elements.
-This could include promoters and terminators.
-The complexity of the task quickly increases with the length of the sequence.
+This could include, for example, promoters and terminators.
+However, its important to note that the complexity of the task quickly increases with the length of the sequence.
 This is because the experimental space grows exponentially with the number of examined positions and so the space becomes increasingly hard to cover with experiments.
-To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.
-
-Finally, we hope to see our RBS library being adapted by the community.
-We believe that its favourable characteristics will make if very useful to designers.
+To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
 
@@
 
-All code and data required to reproduce the results is available at Github:
+All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML}
 
@@
 have designed and performed the laboratory experiments.
-All authors analysed the data and contributed to and reviewed the manuscript.
+Holowko M.
+B.
+and Ong C.
+S.
+conceived and planned the project.
+All authors analysed the data, contributed to and reviewed the manuscript.
 
 \section*{Competing interests}
-The authors declare no competing interests
+The authors declare no competing interests.
 
 \section*{Acknowledgments}
-The authors would like to acknowlege CSIRO's AI/ML and Synthetic Biology Future Science Platforms for providing funding for this research.
+The authors would like to acknowlege CSIRO's Machine Learning and Artificial Intelligence, and Synthetic Biology Future Science Platforms for providing funding for this research.
+The authors would also like to thank CSIRO BioFoundry for help with performing the experiments.
 

### 2021-05-21 paper/supplementary.tex
maciej's edit
change colors, setup subfigure caption, add annotation to tsne
Updated paper.

@@
 Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label after each prediction.
-Therefore we consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2012parallelizing}.
+Therefore we consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2014parallelizing}.
 With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
@@
 \subsection{Machine Learning Design Pipeline}
+\label{sec: ML design pipeline}
 In this section, we report in details about our machine learning design pipeline.
@@
 We choose $l=6, s= 1, \alpha = 2$ for the second round design. 
-For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
+For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2014parallelizing}. 
 The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
@@
 This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
-The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.
+The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.
 
@@
 \end{figure}
+
+\subsection{Other Evaluation}
+
+\textbf{How much more efficient we are by using GP-UCB?}
+
+random = {set of TIR that are in UNI and PPM}
+bandit = {set of TIR that are in bandit 0,1,2,3}
+
+The ratio of the max random to the max bandit: 0.67/1.34 = 0.5
+The ratio of the 0.9 quantile random to the 0.9 quantile bandit: 0.49/0.82 = 0.598
+
+we can also calculate an average regret of TIRs which are higher than SD? The regret for one sequences is calculated based on the difference between the benchmark (1) and TIR.
+
+reciprocal would make a better story. I.e. bandit is 2x better than random.
+
+\textbf{zoom-in tSNE?}
+
+1) supplementary website.
+2) eLife is now offering submissions of interactive papers with Jupyter notebooks. More of a trivia than useful information in our case.
+https://elifesciences.org/labs/ad58f08d/introducing-elife-s-first-computationally-reproducible-article
 
@@
 
+\begin{figure}[!ht]
+    \centering
+     \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF.pdf}
+    \caption{Exploration v.s. Exploitation.}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+     \includegraphics[scale=0.4]{paper/plots/Supplementary/SDhist.pdf}
+    \caption{Histogram of standard deviations.}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \caption{}
+         \includegraphics[scale=0.4]{plots/Supplementary/SD.pdf}
+     \end{subfigure}
+     \hfill
+     \begin{subfigure}[b]{0.49\textwidth}
+         \centering
+         \caption{}
+         \includegraphics[scale=0.6]{plots/Supplementary/SDvsTIR.pdf}
+     \end{subfigure}
+     \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
+B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR.
+Standard deviation and average TIR is given in terms of raw number.}
+     \label{fig:variation of biological replicates}
+\end{figure}
+
 \input{mixture_absGau}

### 2021-05-24 paper/main.tex
Updates from Overleaf

@@
 This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
-The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.
-
+The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\
 The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
@@
 Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
-% \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g.
-is that because the data?
-So I recommend either remove this or explain it somewhere.} \maciej{I'll explain it in the discussion}
 but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
-There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} .
-This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .
-
+There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}.
+This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
@@
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
-There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.
-
-% Machine learning has been also used for prediction in proteins \cite{Yang2018}.
+There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.\\
 However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
@@
 The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
-And it is time-consuming and expensive to measure the TIR for RBS sequences. \mengyan{Maciej: can you add more details or reference here}.
-One the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.
+And it is time-consuming and expensive to measure the TIR for RBS sequences. 
+On the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.
 We addressed the challenges by applying DBTL workflow. 
 Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
-Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. 
-% For the machine learning to work effectively the analysed data set needs to show two qualities: high relative volume and high quality of data.
-% These two elements don't have specific definitions, but in general the data set has to be going into at least hundreds data points and have to cover 5-10\% of that space.
-% \mengyan{Maciej: I think it varies from case to case. We need to either change it or add reference.}
-% And since quality of the obtained data has a direct and strong correlation with quality of the predictions and in effect - recommendations, one need to ensure that the obtained results represent the real value as close as possible.\\
-
+Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. \\
 We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level.
@@
 The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation process.
-In the DESIGN phase, we have used a multi-armed Bandit recommendation algorithm that was selecting RBS designs to be tested in subsequent experimental batches.
-\\
+In the DESIGN phase, we have used a multi-armed Bandit recommendation algorithm that was selecting RBS designs to be tested in subsequent experimental batches.\\
 
@@
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.
-
+By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
 In \emph{E. coli}, the RBS is usually located in the 20 bases upstream of the start codon. 
@@
 Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region
-(see detail in Figure \ref{fig: Flowchart}).
-
+(see detail in Figure \ref{fig: Flowchart}).\\
 Our template RBS sequence, is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA.
 This sequence is a known to have high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
-% Due to poor testing methods that were used for TIR previously we can't claim "highest" cause we don't know how do we compare exactly
 Since this is the sequence against which new RBS sequenced will be benchmarked,
@@
 Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
-% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
-% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
 In each round, we measure the TIR of benchmark RBS as the internal standard. 
@@
 The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
-\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
-\maciej{I did that in a later section.}
-\mengyan{I recommend to mention the observation and point to where you discuss it.}
+Notably, the distributions calculated for all the groups show bivariate distribution, we discuss the possible reasons for that further in the text.\\
 In \ref{fig: Swarmplot and Quantplot}E we show a tSNE plot depicting the experimental space.
@@
 GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
-% Such a stochastic prediction method fits biological processes naturally, since they are highly stochastic as well. 
 The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
@@
 Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
-% Interestingly there wasn't much difference in fitness parameters between Round 1-3.
-% One interesting point is that despite relatively low correlations between predicted and true values our recommendation algorithm performed very well.
 One important point to note is that the predictions are also influenced by our recommendation choices. 
@@
 
+Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in table 1.
 Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
@@
 \end{minipage}
-\caption{Characteristics of the library. \mengyan{write caption here}
-For Efficiency of bandit design:
-random = {set of TIR that are in UNI and PPM};
-bandit = {set of TIR that are in bandit 0,1,2,3}.
-The ratio of the max bandit to the max random: $1.34/0.67$ = 2;
-The ratio of the 0.9 quantile bandit to the 0.9 quantile random: $0.82/0.49$ = 1.67.}
+\caption{\textbf{Characteristics of the library.}
+Left table presents some of the characteristics of our library.
+Right table presents 10 RBS sequences with their corresponding TIR ratios, the first 9 being the strongest sequences including the benchmark sequence (AGGAGA) and the last being the Shine-Dalgarno sequence (AGGAGG).
+CV is coefficient of variation (STD of a sample divided by its mean).
+Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by highest TIR found using random sequence generation. }
 \end{table}
@@
 
-\mengyan{We should try to alter the discussion in terms of our three-sentence story, along with open questions (in a more positive way) and future work.}
-
 \input{methods.tex}

### 2021-05-28 paper/main.tex
Updates from Overleaf

@@
 We see that effect in Figure \ref{fig: Swarmplot and Quantplot} c), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
-We can see the RBSs with high TIRs (blue, with high predicted mean) tend to come from exploitation the design space whereas the explorative points (red, with high predicted standard deviation) give relatively low TIR but expand our knowledge about the unknown part of the design space.
+Those with high predicted mean are coloured blue and represents exploitation, those hued red are with high predicted uncertainty and represent exploration.
+We can see the RBSs with high TIRs tend to come from exploitation the design space whereas the explorative points give relatively low TIR but expand our knowledge about the unknown part of the design space.
+% We can see the RBSs with high TIRs (blue, with high predicted mean) tend to come from exploitation the design space whereas the explorative points (red, with high predicted standard deviation) give relatively low TIR but expand our knowledge about the unknown part of the design space.
 Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
@@
 A number of clusters were targeted by our recommendation algorithm.
-For example, the circled clusters labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
+For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
 More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
@@
     $R^2$ and Spearman correlation coefficient (with corresponding p-value) are all provided for each plot.
+    The p-value is for the hypothesis test whose null hypothesis is that two sets of data are uncorrelated.
     The p-value decreases with round increases, which shows increasing strong evidence about the Spearman correlation between our prediction and measured TIR. 
@@
 \begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
-Raw TIR range                                                                                     &      [4.9258, 105.377]               \\ \hline
-TIR ratio range                                                                                     &      [0.0626, 1.3393]               \\ \hline
+Raw TIR range                                                                                     &      [4.926, 105.377]               \\ \hline
+TIR ratio range                                                                                     &      [0.063, 1.339]               \\ \hline
 \end{tabular}
@@
 \textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
-GGGGGC                & 1.3393             \\ \hline
-GGGGGT                & 1.1544             \\ \hline
+GGGGGC                & 1.339             \\ \hline
+GGGGGT                & 1.154             \\ \hline
 GGCTAT                & 1.084              \\ \hline
 \textbf{AGGAGA}                & 1                  \\ \hline
-GGCGTT                & 0.9805             \\ \hline
-GGGGGG                & 0.9788             \\ \hline
-GGCGAC                & 0.9758             \\ \hline
-CAGGAG                & 0.9629             \\ \hline
-GGCGAG                & 0.9519             \\ \hline
-\textbf{AGGAGG}                & 0.3941             \\ \hline
+GGCGTT                & 0.981            \\ \hline
+GGGGGG                & 0.979             \\ \hline
+GGCGAC                & 0.976             \\ \hline
+CAGGAG                & 0.963             \\ \hline
+GGCGAG                & 0.952             \\ \hline
+\textbf{AGGAGG}                & 0.394            \\ \hline
 \end{tabular}

### 2021-05-28 paper/archive_supp.tex
re-arange supp; plots

--- 2021-05-21 paper/supplementary.tex
@@
-\textbf{Supplementary}
-
-\section{Extended Machine Learning Discussion}
-
-\subsection{Reproducible Plots}
-In this section, we put reproducible plots and will update the plots when we get full results.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.4]{plots/Supplementary/swarmplots.pdf}
-    \caption{Swarmplots for different groups, with raw (left) and normalised (right) TIR labels (averaged over 6 replicates). 
-    Group names represent: Consensus (consensus sequence tested in different round); BPS-NC (bps noncore); BPS-C (bps core); UNI (uniformly random); PPM (position-based probability matrix); Bandit-0 (bandit design for round 0); Bandit-1 (bandit design for round 1).}
-    \label{fig:swarmplots.}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
-    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
-    \caption{Kernel Heatmap and Predictions.
-Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265).
-Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right).
-The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
-    \caption{TSNE of RBS sequences in design space with clustering.
-The distance is calculated based on the weighted degree kernel on RBS sequences.
-Colours indicate different groups, shapes indicate different clusters.}
-\end{figure}
-
 \subsection{Gaussian Process Regression}
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
     \centering
@@
 
-\section{Supplementary Figures}
+\subsection{Mixture of Absolute Gaussian}
 
-\begin{figure}
-     \centering
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGG.pdf}
-         \caption{$y=x$}
-         \label{fig:GGGGGG}
-     \end{subfigure}
-     \hfill
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGC.pdf}
-         \caption{$y=3sinx$}
-         \label{fig:GGGGGC}
-     \end{subfigure}
-     \vskip\baselineskip
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure_CGGGAC.pdf}
-         \caption{$y=5/x$}
-         \label{fig:CGGGAC}
-     \end{subfigure}
-     \hfill
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \includegraphics[scale=0.25]{plots/Supplementary/Structure_CACGTT.pdf}
-         \caption{$y=5/x$}
-         \label{fig:CACGTT}
-     \end{subfigure}
-        \caption{\textbf{Folding predictionds for four different RBS.} Here we show the predicted, energetically most favourable structures for four of our RBS with the following cores: GGGGGG, GGGGGC, CGGGAC, CACGTT and their immediate upstream and downstream background sequence.
-There is no discerbile and consistent difference between strong ones (the first two) and weak ones (the last two).}
-        \label{fig:structures}
-\end{figure}
+The raw TIR scores in each round are positive and  have two modes. 
+We assume they are mixture of two absolute Gaussian distributions. 
+If we can fit the mixture of absolute Gaussian distributions, we can answer the following questions:
+\begin{itemize}
+    \item is the left peak decreasing in height with each round?
+	\item is the left peak going left?
+	\item Is the right peak increasing in height with each round?
+	\item is the right peak going right?
+\end{itemize}
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\textwidth]{plots/Supplementary/All_logo.pdf}
-    \caption{\textbf{Sequence logo for all tested sequences.} }
-    \label{fig: All_logo}
-\end{figure}
+The density of absolute Gaussian (also called \textit{folded normal distribution}) is 
+\begin{align}
+    f_{Y}\left(x ; \mu, \sigma^{2}\right)=\frac{1}{\sqrt{2 \pi \sigma^{2}}} e^{-\frac{(z-\mu)^{2}}{2 \sigma^{2}}}+\frac{1}{\sqrt{2 \pi \sigma^{2}}} e^{-\frac{(x+\mu)^{2}}{2 \sigma^{2}}}
+\end{align}
+for $x \geq 0$ and $0$ everywhere else. 
+The mean and variance of the absolute Gaussian distribution is 
+\begin{align}
+    \mu_{Y} &=\sigma \sqrt{\frac{2}{\pi}} \exp \left(\frac{-\mu^{2}}{2 \sigma^{2}}\right)+\mu \operatorname{erf}\left(\frac{\mu}{\sqrt{2 \sigma^{2}}}\right);\\
+    \sigma_{Y}^{2} &=\mu^{2}+\sigma^{2}-\mu_{Y}^{2},
+\end{align}
+where $\operatorname{erf}$ is the error function 
+$\operatorname{erf} z=\frac{2}{\sqrt{\pi}} \int_{0}^{z} e^{-t^{2}} d t$.
+
+We consider the model of the mixture of two absolute Gaussian distributions as
 
@@
     \centering
-     \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF.pdf}
-    \caption{Exploration v.s. Exploitation.}
+    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
+    \caption{TSNE of RBS sequences in design space with clustering.
+The distance is calculated based on the weighted degree kernel on RBS sequences.
+Colours indicate different groups, shapes indicate different clusters.}
 \end{figure}
-
-\begin{figure}[!ht]
-    \centering
-     \includegraphics[scale=0.4]{paper/plots/Supplementary/SDhist.pdf}
-    \caption{Histogram of standard deviations.}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \caption{}
-         \includegraphics[scale=0.4]{plots/Supplementary/SD.pdf}
-     \end{subfigure}
-     \hfill
-     \begin{subfigure}[b]{0.49\textwidth}
-         \centering
-         \caption{}
-         \includegraphics[scale=0.6]{plots/Supplementary/SDvsTIR.pdf}
-     \end{subfigure}
-     \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
-B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR.
-Standard deviation and average TIR is given in terms of raw number.}
-     \label{fig:variation of biological replicates}
-\end{figure}
-
-\input{mixture_absGau}

### 2021-05-28 paper/methods.tex
Updates from Overleaf
re-arange supp; plots

@@
 %  For recommendations, we consider the \textit{Upper Confidence Bound (UCB)} type algorithms. 
-%  As one popular type of the bandit algorithms \cite{lattimore2018bandit}, the UCB type of algorithms are based on the \textit{optimism in the face of uncertainty},
+%  As one popular type of the bandit algorithms \cite{lattimore2020bandit}, the UCB type of algorithms are based on the \textit{optimism in the face of uncertainty},
  %provide various approaches to sequential design where an agent adaptively chooses one or more options among several actions based on certain policies.
@@
         \caption{}
-        \includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}
+        \includegraphics[scale=.4]{plots/Main_Paper/Recommend_2_Data_Point_Top_2_UCB.pdf}
     \end{subfigure}
@@
         \caption{}
-        \includegraphics[scale=0.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}
+        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_1st_Data_Point.pdf}
     \end{subfigure}
@@
         \caption{}
-        \includegraphics[scale=0.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}
+        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_2nd_Data_Point.pdf}
     \end{subfigure}

### 2021-06-02 paper/main.tex
address colin's comments

@@
 Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
-However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
-This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
-The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\
-The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
+However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\
+The biggest gap in the DTBL cycle, at present, is at the Learn and Design interface - it is hard to translate obtained results into new designs.
 For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
@@
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
-New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
-These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020}.
-In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
-Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
+These new data sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
+For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
-There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.\\
-However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
-The challenges of applying machine learning techniques on synthetic biology have two folds.
-On the one hand, there is no large-scale and high-quality labelled data. 
-The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
-And it is time-consuming and expensive to measure the TIR for RBS sequences. 
-On the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.
-We addressed the challenges by applying DBTL workflow. 
-Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
-Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. \\
-We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level.
+There were also successful attempts at using deep learning techniques for analysis of big synthetic biology data sets \cite{Alipanahi2015,Angermueller2016}.\\
+In this work, we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of bacterial RBS with the goal of optimising associated protein expression level.
 RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
 In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
-Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
+Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
 We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
-In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
-In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
-The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
-After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
-The results were then fed back to the algorithm for it to recommend the next round of designs.
 This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
@@
 We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit well into our prediction method.
-We building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
+In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
 By using our workflow, we have found very strong RBS sequences and we have generated an extensive library of diverse RBS that can be used in the future studies.
@@
 Another example would be - given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
+Next question would be - how generalizable is this method? Can it be succesfully used for design of bigger genetic parts, like promoters?
 Finally, another question with high impact on the performance of the workflow is - how should we choose the parameters which controls the exploration-exploitation rate? \\
@@
 
-All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML}
+All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML} .
+All the processed and raw data is included in the repository.
+Sequences of plasmids and oligos
 

### 2021-06-02 paper/methods.tex
address colin's comments

@@
 Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
-\mengyan{***}) with 34$\mu$g/mL chloramphenicol (Sigma, catalogue no.
-\mengyan{***}).
+244610) and 15g/L agar (Sigma-Adrich, catalogue no.
+A1296) with 34$\mu$g/mL chloramphenicol (Sigma-Aldrich, catalogue no.
+C1919).
 Plates were incubated overnight at 37$^{\circ}$C. 
@@
 \textbf{Glycerol stock preparation.}
-100$\mu$L of sterile 80\% (v/v) glycerol and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
+100$\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, catalogue no.
+GA010) and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
 They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.
@@
 The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
-Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN) according to the manufacturer's instructions.
+Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN, catalogue no. 27106) according to the manufacturer's instructions.
 Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/$\mu$L.
@@
 The following morning, 20$\mu$L of overnight cultures were added to 980$\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate. 
-After 90 minutes, 200$\mu$L of each culture (induced with 1.0$\mu$L of 0.1M IPTG) was transferred to a flat-bottom clear polystyrene 96-well plate.
+After 90 minutes, 300$\mu$L of each culture (induced with 1.0$\mu$L of 0.1M IPTG (Sigma-Aldrich, catalogue no.
+I5502) to a final concentration of 0.5 mM) was transferred to a flat-bottom clear polystyrene 96-well plate.
 

### 2021-06-02 paper/methods.tex
update supp

@@
 $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting parameter over the position in the
-sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the maximum shift
+sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$ ($L = 20$ in our design). $S(l)$ determines the maximum shift
 at position $l$. 
-In our design, the length of RBS sequence $L = 20$. we chose $\ell=6$, and $S(l)= 1$ for all $l$.
+Furthermore, we normalise the kernel with centering and unit norm in terms of all sequences in the design space. 
+% In our design, the length of RBS sequence $L = 20$.
+The hyperparameter for kernel, including maximum substring length $\ell$, maximum shift length $S(l)$, 
+and the standard deviation of GP noise $\alpha$ were chosen based on 10-repeat 5-fold cross validation.
+%  ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).
 
@@
 As summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design area so that the information gain is maximised. 
+In our design, we have set the batch size as 90. We let $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round for more exploitation.

### 2021-06-02 paper/supplementary.tex
update supp

--- 2021-05-17 paper/supplementary.tex
@@
 
-\section{Extended Machine Learning Discussion}
+\section{Machine Learning Methods}
 
-\subsection{Reproducible Plots}
-In this section, we put reproducible plots and will update the plots when we get full results.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale = 0.4]{plots/Supplementary/swarmplots.pdf}
-    \caption{Swarmplots for different groups, with raw (left) and normalised (right) TIR labels (averaged over 6 replicates). 
-    Group names represent: Consensus (consensus sequence tested in different round); BPS-NC (bps noncore); BPS-C (bps core); UNI (uniformly random); PPM (position-based probability matrix); Bandit-0 (bandit design for round 0); Bandit-1 (bandit design for round 1).}
-    \label{fig:swarmplots.}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
-    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
-    \caption{Kernel Heatmap and Predictions.
-Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265).
-Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right).
-The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
-    \caption{TSNE of RBS sequences in design space with clustering.
-The distance is calculated based on the weighted degree kernel on RBS sequences.
-Colours indicate different groups, shapes indicate different clusters.}
-\end{figure}
+In this section, we describe the machine learning methods in details.
 
@@
 
-    \item \textit{Sum of Spectrum Kernel,} considers weighted sum over different parts of the string.
+    % \item \textit{Sum of Spectrum Kernel,} considers weighted sum over different parts of the string.
 
-    \item \textit{Mixed Spectrum Kernel,} considers weighted sum over different substring length, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$,
-        \begin{align}
-            k_\ell^{MixedSpec}(\mathbf{x}, \mathbf{x}^\prime) 
-            = \sum_{d=1}^{\ell} \beta_d k_d^{Spec}(\mathbf{x}, \mathbf{x}^\prime)
-        \end{align}
+    % \item \textit{Mixed Spectrum Kernel,} considers weighted sum over different substring length, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$,
+    %     \begin{align}
+    %         k_\ell^{MixedSpec}(\mathbf{x}, \mathbf{x}^\prime) 
+    %         = \sum_{d=1}^{\ell} \beta_d k_d^{Spec}(\mathbf{x}, \mathbf{x}^\prime)
+    %     \end{align}
     \item \textit{Weighted Degree Kernel,} considers positional information. WD kernel counts the match of kmers at corresponding positions in two sequences.
@@
 
-\textbf{From kernel to distance}:
-$$d(\mathbf{x}, \mathbf{x}^\prime) = \sqrt{k(\mathbf{x}, \mathbf{x}) + k(\mathbf{x}^\prime, \mathbf{x}^\prime) - 2 k(\mathbf{x}, \mathbf{x}^\prime)} $$
+% \textbf{From kernel to distance}:
+% $$d(\mathbf{x}, \mathbf{x}^\prime) = \sqrt{k(\mathbf{x}, \mathbf{x}) + k(\mathbf{x}^\prime, \mathbf{x}^\prime) - 2 k(\mathbf{x}, \mathbf{x}^\prime)} $$
 
@@
 
-\subsection{Machine Learning Design Pipeline}
-In this section, we report in details about our machine learning design pipeline.
-The raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
+\subsection{Bandit-0 Design}
+\label{sec: ML design pipeline}
+% \mengyan{explain somewhere why we have a different pipeline for round 0.}
+% \mengyan{add some analysis on literature data?
+% to verify 1) our library is "better" 2) it's hard to do ML on available library.
+some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
+% And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
 
-\mengyan{explain somewhere why we have a different pipeline for round 0.}\\
-\textbf{Bandit-0: } the design of round-0 is based on the literature data \cite{jervis2018machine}.
+The design of round-0 is based on the literature data \cite{jervis2018machine}.
 We first normalise the raw TIR to values between 0 and 1. 
@@
 We chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-The design size is 60 with UCB parameter $\beta = 1$.\\
-\textbf{Bandit-1 to Bandit-3: }
-the t + 1 round design is based on the $t^th$ ($t \geq 1$) round result, where each sequence has 6 replicates with TIR labels.
-We pre-processed the data by taking a logarithm transformation and standardisation of the raw TIR label for each replicates respectively.
-After normalisation, each replicate has zero mean and unit variance. 
-For prediction, we use Gaussian process regression, with training on all normalised replicates and predicting on the design space (6-base core part design) except known sequences. 
-We assume the observation are noisy, where the noise is under centered normal distribution with standard deviation $\alpha$.  
-We model the covariance matrix using the weighted degree kernel with shift. 
-We normalise the kernel with centering and unit norm in terms of the whole kernel constructed by both first round result and design space. 
-The hyperparameter for kernel, including maximum substring length  $l$, maximum shift length $s$, and the noise standard deviation $\alpha$ of Gaussian process model are choose based on 10-repeat 5-fold cross validation.
-We choose $l=6, s= 1, \alpha = 2$ for the second round design. 
-For recommendation, we use batch upper confidence bound introduced by GP-BUCB algorithm \cite{desautels2012parallelizing}. 
-The upper confidence bound is constructed by predicted mean plus 2 predicted standard deviation.   
-We recommend 90 sequences from the design space.
+The design size is 60 with UCB parameter $\beta = 1$.
 
-\subsection{Statistics of the core vs non-core regions of the RBS}
+\newpage
 
+\section{Supplementary Plots}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale = 0.4]{plots/Supplementary/swarmplot_raw.pdf}
+    \caption{Swarm plot showing the obtained raw TIRs divided into RBS groups.
+    Raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth. 
+    BPS-NC: base-by-base changes in the non-core region. 
+    BPS-C: base-by-base changes in the core region. 
+    UNI: Randomly generated sequences with uniform distribution. 
+    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
+    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
+    SD - Shine-Dalgarno sequence.
+    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
+    BN - benchmark sequences for all plates. 
+    }
+    \label{fig:swarmplots.}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_0.pdf}
+    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_1.pdf}
+    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_2.pdf}
+    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
+    \caption{scatter abc1 TT}
+    \label{fig:scatter abc1 TT.}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
+    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} 
+    The core region is highligthed in light green and the lines are rolling averages for each base. 
+    The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base,
+    which is colour coded.
     In our study, we have tested a set of sequnces designed to confirm the notion that changing bases within the RBS core (6 bases) is statistically more influencing the TIR than the changes made outside the core.
@@
     This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
-The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
-    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} The core region is highligthed in light green and the lines are rolling averages for each base.
-The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base, which is colour coded.}
+    The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.}
     \label{fig:core_vs_noncore}
@@
 
-\section{Supplementary Figures}
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
+    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
+    \caption{Kernel Heatmap and Predictions.
+Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265).
+Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right).
+The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
+\end{figure}
 
-\begin{figure}
+\begin{figure}[!ht]
     \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[!ht]
    \centering
@@
 
-\input{mixture_absGau}
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
+    \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
+    \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
+   \caption{Exploration v.s. Exploitation.}
+\end{figure}
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/SDhist.pdf}
+   \caption{Histogram of standard deviations.}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/SD.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.6]{plots/Supplementary/SDvsTIR.pdf}
+    \end{subfigure}
+    \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
+B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR.
+Standard deviation and average TIR is given in terms of raw number.}
+    \label{fig:variation of biological replicates}
+\end{figure}

### 2021-06-04 paper/main.tex
Updates from Overleaf

@@
 However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\
+
 The biggest gap in the DTBL cycle, at present, is at the Learn and Design interface - it is hard to translate obtained results into new designs.
@@
 This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\
+
 Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
@@
 There were also successful attempts at using deep learning techniques for analysis of big synthetic biology data sets \cite{Alipanahi2015,Angermueller2016}.\\
+
 In this work, we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of bacterial RBS with the goal of optimising associated protein expression level.
@@
 Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
-We did this by designing a sequential experimental workflow shown in Figure \ref{fig: Flowchart}.
 This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
@@
     \includegraphics[scale=0.7]{plots/Main_Paper/flowchart.pdf}
-    \caption{\textbf{Flowchart of machine learning based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm.
+    \caption{\textbf{Flowchart of the machine learning based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm.
 After generating the recommendations the RBS are built and tested using automated laboratory methods allowing for rapid construction and testing at scale.
@@
 \end{figure}
-
 To help us obtain reliable and reproducible results we have employed automation-heavy workflow in the BUILD and TEST phases.
@@
 Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down the time required to prepare our variants.\\
+
 In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
-The plasmids were then transformed and the resulting transformants were tested using microplate reader.
+The plasmids were then transformed and the resulting transformants were tested using a  microplate reader.
 Vast majority of reactions were prepared using liquid handling equipment.
@@
 
-In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP) in its mut3b variant. 
+In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP). 
 By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
@@
 By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
+
 In \emph{E. coli}, the RBS is usually located in the 20 bases upstream of the start codon. 
@@
 (see detail in Figure \ref{fig: Flowchart}).\\
+
 Our template RBS sequence, is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA.
-This sequence is a known to have high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
+This sequence is a known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
 Since this is the sequence against which new RBS sequenced will be benchmarked,
 we will refer to this sequence as the \textit{benchmark sequence} from now on.
-In our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT, where N can be any choices of A, C, G, T.
+We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
+In effect, in our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e.
+TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).
 The total experimental (variant) space is then $4^6$ = 4096.
-We have experimentally confirmed that changing the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
 
@@
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (see below).
-Another feature of the bandits algorithm is that it balances two parts: the exploration of the unknown (untested) parts of the design space where high TIR RBS can be hidden, and exploitation which goal is querying areas which are known to give relatively high TIRs.
-One thing of note is that the bandit algorithm is stochastic, that is it exploits the probabilities of a given event occurring (in this case RBS having a specific TIR). 
-As such, it pairs naturally with our prediction algorithm, the Gaussian Process Regression, which provides probability based function regression.
+In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (see Figure \ref{fig:batch rec}).\\
+
+Another feature of the bandits algorithm is that it balances two aspects: the exploration of the unknown (untested) parts of the design space where high TIR RBS can be hidden, and exploitation which goal is to query areas which are known to give relatively high TIRs.\\
 
@@
 
-To generate the dataset that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
-To fit into our automated workflow, we have divided the 450 variants into batches of 90.
+To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
+To fit into our automated workflow, we have divided the 450 variants into batches of 90.\\
+
 In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:
+
 \begin{itemize}
     \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide variations of all 20 nucleotides of the original, consensus sequence.
-This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (Figure \ref{fig:core_vs_noncore}).
+This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (see Supplementary Figure \ref{fig:core_vs_noncore}).
     % \mengyan{put it in main paper?}
@@
     \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}.
-    \item Bandit-0: 60 RBS sequences recommended by our implementation of recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIR.
+    \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIRs.
     This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
 \end{itemize}
-In the subsequent 3 rounds, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
+
+In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).\\
 
@@
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-The Bandit-0 group performed poorly, despite being machine learning driven, due to being trained on literature data.
+The Bandit-0 group performed poorly, despite being machine learning driven, due to being trained on only approximate data.\\
+
 However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
 In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
-Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
+Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
+
 Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
@@
 The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
-We see that effect in Figure \ref{fig: Swarmplot and Quantplot} c), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
-Those with high predicted mean are coloured blue and represents exploitation, those hued red are with high predicted uncertainty and represent exploration.
-We can see the RBSs with high TIRs tend to come from exploitation the design space whereas the explorative points give relatively low TIR but expand our knowledge about the unknown part of the design space.
-% We can see the RBSs with high TIRs (blue, with high predicted mean) tend to come from exploitation the design space whereas the explorative points (red, with high predicted standard deviation) give relatively low TIR but expand our knowledge about the unknown part of the design space.
+We see that effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+Those with high predicted mean are coloured blue and represent exploitation, those hued red are with high predicted uncertainty and represent exploration.
+We can see the RBSs with high TIRs tend to come from exploitation the design space whereas the explorative points give relatively low TIR but expand our knowledge about the unknown part of the design space.\\
+
 Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
@@
 Notably, the distributions calculated for all the groups show bivariate distribution, we discuss the possible reasons for that further in the text.\\
-In \ref{fig: Swarmplot and Quantplot}E we show a tSNE plot depicting the experimental space.
-Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
+
+In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
+Each RBS is located on the plot according to its distance to other RBSs as calculated by our embedding function.
 We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
@@
 Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
-This algorithm creates a model which takes the RBS sequences as input and predicts the TIR values with uncertainty level about the prediction, based on the experimental data.
-For this study, we have used the Gaussian Process regression (GPR) as our prediction method.
+This algorithm creates a model which takes the RBS sequences as input and predicts the mean TIR values with uncertainty level about that prediction, based on the experimental data.
+For this study, we have used the Gaussian Process Regression (GPR) as our prediction method.
 GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
 The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
-A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between data points, in our case RBS sequences.
-Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process easier.
+
+A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between input data points, in our case RBS sequences.
+Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process possible.
 For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-For subsequent rounds, we used 
-% a more powerful \maciej{more powerful in what sense?} kernel function, 
-the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds}, which has been shown to have good performance in various prediction tasks \cite{Ben-Hur2008}.
+For subsequent rounds, we used the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds}, which has been shown to have good performance in various prediction tasks \cite{Ben-Hur2008}.
 The WDS not only counts the matches of substrings of a certain length (i.e. kmers), but also takes into account the positional information and shifting of substrings.
-For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift.
-The relative distances between sequences calculated by our kernel function are shown in figure \ref{fig: Swarmplot and Quantplot}E.\\
+The relative distances between sequences calculated by our kernel function are shown in Figure \ref{fig: Swarmplot and Quantplot}E.\\
 
-Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
+Figure \ref{fig: Scatterplot} shows our algorithms performance in terms of predictions in each round. 
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
@@
     \end{subfigure}
-    \caption{\textbf{Predictions generalise to held out data.} The scatter plots A-D are showing the performance of our prediction algorithm (GPR with WDS kernel, with maximum k-mer as 6 and maximum shift as 1) calculated after each of the rounds.
-    Note the TIR values are normalised TIR based on the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
-    The values of 
-    $R^2$ and Spearman correlation coefficient (with corresponding p-value) are all provided for each plot.
-    The p-value is for the hypothesis test whose null hypothesis is that two sets of data are uncorrelated.
-    The p-value decreases with round increases, which shows increasing strong evidence about the Spearman correlation between our prediction and measured TIR. 
+    \caption{\textbf{Performance of the prediction algorithm.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each round.
+    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
+    The values of $R^2$ and Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
+    The p-value here is for null hypothesis stating that two sets of data are uncorrelated.
     }
@@
 
-Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in table 1.
-Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 (in terms of TIR ration) of our sequences.
+Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in Table 1.
+Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (in terms of TIR ratio).
 It is generally understood that guanine rich sequences are promoting strong transcription.
@@
 CV is coefficient of variation (STD of a sample divided by its mean).
-Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by highest TIR found using random sequence generation. }
+Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by the highest TIR found using random sequence generation. }
 \end{table}
@@
 Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
-We define the editing distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
+We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
 Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
@@
 For RBS with medium TIR ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
-That means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give meaningful improvement. 
-The low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012},
-which indicate that Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
+That means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
+This also implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.
 In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
 In such cases, a directed process should be strongly encouraged.
-This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.
+This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.\\
 
@@
 
-There are still open questions that need to be addressed for applying machine learning in synthetic biology.
-For example, how can we understand the language of RBS sequences better?
-In other words, can we extract more biologically important information from the decisions made by our algorithms?
-Another example would be - given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
-Next question would be - how generalizable is this method? Can it be succesfully used for design of bigger genetic parts, like promoters?
-Finally, another question with high impact on the performance of the workflow is - how should we choose the parameters which controls the exploration-exploitation rate? \\
-
 We have found our approach bringing machine learning and synthetic biology experts very powerful.
-We envision to that the pairing of machine learning with high-throughput automation will keep delivering a high number of good quality datasets nad improved methods for biological engineering.\\
+We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
+However, there are still some open questions that have to be answered.
+For example, we are still not sure how generalisable are the used methods and if it possible to extract more biological knowledge from the algorithms.
+It is also not clear how to improve the accuracy of predictions, but we believe this will occur through  a combination of bigger data sets and better algorithms.\\
 
@@
 All the processed and raw data is included in the repository.
-Sequences of plasmids and oligos
+Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as separate file.
 
@@
 
+\setcounter{figure}{0}
+\makeatletter 
+\renewcommand{\thefigure}{S\@arabic\c@figure}
+\makeatother
 \appendix

### 2021-06-04 paper/methods.tex
Thinking about the story. Need to refocus results section to not sound like a method section
Updates from Overleaf
Update ML methods.

--- 2021-06-02 paper/methods.tex
@@
 Plates were incubated overnight at 37$^{\circ}$C.
-8
 
@@
 % To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
-One type of machine learning algorithms is prediction (\textbf{LEARN}), which helps us learn for function of TIR with respect to RBS sequences.
-The other type of machine learning algorithm is recommendation (\textbf{DESIGN}), which recommends RBS sequences to query in each batch based on the predictions from LEARN.
+One type of machine learning algorithms is a prediction algorithm (\textbf{LEARN}), which helps us learn the function of TIR with respect to RBS sequences.
+The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each batch based on the predictions from LEARN.
 % 1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
@@
 In round $t$, the prediction and design is based on the results obtained in all previous rounds. 
-The implementation of the machine learning algorithms is in Python 3.7 and scikit-learn library \cite{scikit-learn}. 
-In the following, we describe our machine learning pipeline (Bandit 1-3) by the order of data pre-processing, prediction and string kernels, recommendation. 
+Our implementation of the machine learning algorithms was tested in Python 3.7 and used the scikit-learn library \cite{scikit-learn}. 
+In the following paragraphs, we describe our machine learning pipeline (Bandit 1-3) in the following order: data pre-processing, prediction and kernels and finally recommendation. 
 The pipeline for Bandit-0 is reported in Supplementary \ref{sec: ML design pipeline}.
@@
 
-For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth.
+For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluorescence divided by OD600 of culture over 4h counting from the start of the log phase of growth.
 % REVIEW:
@@
 % We consider both the label (TIR) pre-processing and kernel pre-processing.
-For label pre-processing, we first adjust the TIR values in each round by the round-wise reference values.
-This is because, we found that TIR values in each round for the same sequence can be different (for reference RBS, STD $\approx$ 17.7) due to the measurement noise.
-So we repeatedly measured the reference RBS in each round to rectify the precision of other measured RBS in the same round. 
-More precisely, in round $t$, we subtract the TIR mean of reference RBS measured in round $t$ from all TIR values measured in the same round, for each replicates respectively. 
-We then pre-processed the data by taking a logarithm transformation and standardisation of the adjusted TIR label for each replicates respectively, 
+For label pre-processing, we first adjust the TIR values in each round using the round-wise reference values.
+This reference values are the TIR of the benchmark sequence that is run in triplicate in each round.
+Specifically, in round $t$, we subtract the TIR mean of the benchmark RBS measured in round $t$ from all TIR values measured in the same round, for each replicate separately. 
+We then normalised the data by taking a logarithm transformation and standardisation of the adjusted TIR label for each replicate separately.
 After normalisation, each replicate has zero mean and unit variance. 
-Furthermore, we also normalised the kernel matrix used for prediction by centering and unit-norm normalisation, which is reviewed in details in Supplementary \ref{supp: Normalisation of Kernel}.
+Furthermore, we also normalised the kernel matrix used for prediction by centring and unit-norm normalisation, which is reviewed in details in Supplementary \ref{supp: Normalisation of Kernel}.
 % Since the design space is known, we consider the transductive approach \cite{} for the kernel pre-processing with centering and unit-norm normalisation.
@@
 
-To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
-In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function values at each point in the selected set $\mathcal{S}_t$, i.e.
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
 $y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
@@
 To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.
-WDS is a type of string kernels, which take two sequences as inputs and output a scalar value which represents the similarities between the two sequences.  
+WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
 WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
@@
 Let $\mathbb{I}$(A) be the indicator function, which equals 1 if $A$ is true and 0 otherwise. 
-Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ indicates whether two substrings of length $d$ matches, between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
+Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ indicates whether two substrings of length $d$ match, between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
 This is similarly done for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
@@
 $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting parameter over the position in the
-sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the maximum shift
-at position $l$. 
-In our design, the length of RBS sequence $L = 20$. we chose $\ell=6$, and $S(l)= 1$ for all $l$.
+sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$ ($L = 20$ in our design).
+$S(l)$ determines the maximum shift at position $l$. 
+Furthermore, we normalise the kernel with centring and unit norm in terms of all sequences in the design space. 
+% In our design, the length of RBS sequence $L = 20$.
+The hyperparameters for kernel, including maximum substring length $\ell$, maximum shift length $S(l)$, and the standard deviation of GP noise $\alpha$ were chosen based on 10-repeat 5-fold cross validation.
+%  ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).
 
@@
 
-To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides such $\textit{exploitation-exploration balance}$ by balancing the predicted mean and standard deviation.
+To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$ by balancing the predicted mean and standard deviation.
+\maciej{Not sure what does this mean? The balancing part}
 On the one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
@@
 
-Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e.
+Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequences sequentially (i.e.
 one-by-one) and then wait for the label to be tested and used to improve the model.
@@
 by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
+\maciej{And do what to prevent this waste?}
 
@@
     \end{subfigure}
-    \caption{Batch Recommendation Illustration.
-We use the batch size of 2, with 5 initial observations.
+    \caption{\emph{Batch Recommendation Illustration.} We use the batch size of 2, with 5 initial observations.
 The design space is 24 uniformly distributed points in the range [-2,10], i.e.
@@
     The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
-    (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions. 
+    (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, chosen with GP predictions. 
     (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that.}
@@
 
-A key property of Gaussian Process regression is that the predictive standard deviation depends only on observed points (i.e. features), but not on the labels of those observed points. 
+A key property of Gaussian Process regression is that the predicted standard deviation depends only on observed points (i.e. features), but not on the labels of those observed points. 
 % We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
 % As we can see, the predicted confidence interval is the same for the two plots. 
-One can make use of this property to design batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
+One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
 The strategy is to recommend sequences sequentially by updating the UCB score with the updated predicted standard deviation with the previously recommended points.
@@
 As we can see in Figure \ref{fig:batch rec}(C), since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
-As summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design area so that the information gain is maximised.
+In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 
+In our design, we have set the batch size as 90, to fit the experimental batch.
+We set $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.

### 2021-06-06 codes/data_generating.py
allow kernel normalisation over known features

@@ -6,12 +6,14 @@
 # - A. In each round, substract the mean of every data points by the reference AVERAGE, and then add 100 (to make the values positive).  
 # - B. Take log (base e) transformation for each data points.  
 # - C. Apply z-score normalisation.  
-#     - C.1 on all data, so that the mean and variance of each replicate of all data is zero and one after normalisation. 
-#     - C.2 on each round, so that the mean and variance of each replicate of data in each round is zero and one after normalisation. 
+#     - C.1 on each round, so that the mean and variance of each replicate of data in each round is zero and one after normalisation. 
+#     - C.2 on all data, so that the mean and variance of each replicate of all data is zero and one after normalisation. 
 # - D. Apply min-max normalisation.
-#     - D.1 on all data
-#     - D.2 on each round
+#     - D.1 on each round
+#     - D.2 on all data
 # - E. Apply ratio normalisation. In each round, each data points is devided by the mean of refernce AVERAGE, so that in each round, the reference labels are almost 1. 
+#     - E.1 on each round
+#     - E.2 on all data
 
 # 25/11/2020 Update: add unique index corresponding to 'idx_seq_dict.npz' 
 
@@ -41,8 +43,8 @@ import pickle
 
 # Setting 
 # Specify your choice
-approach = 'e1'
-to_design_round = '4'
+approach = 'd1'
+# to_design_round = '4'
 
 approach = approach.lower()
 
@@ -117,16 +119,16 @@ def normalize(df, col_name):
         # So we subtract the mean of the reference sequence (the only one RBS tested repeated in different rounds) in each round.
         # REVIEW: COMMENT FOR NOW
         if substract_ref_mean:
-            if to_design_round in {'2', '3', '4'}: 
-                ref_seq_mean = group.loc[group['Group'] == 'reference', col_name].stack().mean()
-                print('col name: ', col_name)
-                print('round: ', name)
-                print('before substracting mean: ', group[col_name].mean())
-                print('reference mean: ', ref_seq_mean)
-                # print(group.loc[group['Group'] == 'reference', col_name])
-                # +100 to avoid invalid value in log
-                group[col_name] = group[col_name] - ref_seq_mean + 100
-                print('after substracting mean: ', group[col_name].mean())
+            # if to_design_round in {'2', '3', '4'}: 
+            ref_seq_mean = group.loc[group['Group'] == 'reference', col_name].stack().mean()
[…]

### 2021-06-06 codes/kernels_for_GPK.py

@@ -18,7 +18,7 @@ from codes.environment import Rewards_env
 import pickle
 import codes.config as config # provide configuration, e.g. global variable, path
 
-# Aug 2020 Mengyan Zhang 
+# June 2021 Mengyan Zhang 
 # String Kernel Classes, taking strings as input
 # including spectrum kernel, weighted degree kernel and weighted degree kernel with shift 
 # The implementation is based on Support Vector Machines and Kernels for Computational Biology 
@@ -26,19 +26,11 @@ import codes.config as config # provide configuration, e.g. global variable, pat
 # Inherit from sklearn gaussain process kernel class
 # https://github.com/scikit-learn/scikit-learn/blob/b194674c4/sklearn/gaussian_process/kernels.py#L1213
 
-# When creating an instance, the whole (train + test) kernel matrix is pre-calculated and normalised
-# For Gaussian Process Regression, the specific kernel matrix is sliced from the calculated kernel
+# Two choices of kernel normalisation target:
+# 1. normalisation over the all design space. The kernel matrix is pre-calculated and saved. 
+# 2. normalisation over all currently known RBS sequences. 
+# For Gaussian Process Regression, the specific kernel matrix is sliced from the calculated kernel (in __call__)
 
-# The all bases for input strings, we use the RBS sequences here
-
-
-# global features for all known and design data
-# if os.path.exists('../../data/unknown_design.csv'):
-#     unknown_df = pd.read_csv('../../data/unknown_design.csv') # sequences with unknown label
-#     unknown_features = Rewards_env(np.asarray(unknown_df[['RBS', 'RBS6']]), 'label').embedded
-# else:
-#     unknown_features = None
-#     print('No default features for kernel instance. Please specify features.')
 
 class String_Kernel(Kernel):
     """Base class for string kernels
@@ -58,100 +50,94 @@ class String_Kernel(Kernel):
     kernel_all_normalised: (n_train+n_test, n_train+n_test)
         kernel matrix for all data with centering and unit norm
 
-    * Kernel hyperparameters
-
+    * Kernel hyperparameters *
     l : int, default 3
         number of l-mers (length of 'word')
-    # TODO: delta weight flag
-    padding_flag: Boolean, default False
-        indicates whether adding padding characters before and after sequences
-    gap_flag: Boolean, default False
-        indicates whether generates substrings with gap
-
     # TODO: change hyperparameters maybe (now set to the same as Dotproduct)
[…]

### 2021-06-06 codes/sanity_check.py

@@ -1,32 +1,0 @@
-# sanity check for reproducibility 
-# direct to proper path
-import os
-import sys
-module_path = os.path.abspath(os.path.join('../'))
-if module_path not in sys.path:
-    sys.path.append(module_path)
-
-import numpy as np
-import pandas as pd
-
-def check_dataset(data1, data2):
-    for index1, row1 in data1.iterrows():
-        for index2, row2 in data2.iterrows():
-            if row1['RBS'] == row2['RBS'] and row1['RBS'] != 'TTTAAGAAGGAGATATACAT':
-                print(row1['RBS'])
-                # print('ave diff: ', np.abs(row1['AVERAGE'] - row2['AVERAGE']))
-                # print('std diff: ', np.abs(row1['STD'] - row2['STD']))
-                
-                # Instead of using average, we use a single replicate to check
-                # Since AVERAGE calculation in round 1 and 2 were wrong 
-                assert np.abs(row1['Rep6'] - row2['Rep6']) < 1e-3
-                # assert np.abs(row1['STD'] - row2['STD'])< 1e-3
-                break
-    
-source1 = "data/pipeline_data/Results_Microplate_partialTrue_normTrue_mean_roundRep_formatSeq_logTrue_Round2.csv"
-# source2 = "/home/admin-u6015325/ownCloud/SynBio_repo_designs/SynbioML-round1/data/firstRound_Microplate_normTrue_formatSeq_logTrue.csv"
-source2 = "/home/admin-u6015325/ownCloud/SynBio_repo_designs/SynbioML-round2/data/Results_Microplate_partialTrue_normTrue_roundRep_formatSeq_logTrue.csv"
-check_dataset(pd.read_csv(source1), pd.read_csv(source2))
-
-# Log: round 1-3 passed the test (tested rep6 only)
-# REVIEW: need to be careful about round 2, whether used the "correct" set of replicates to train?

### 2021-06-06 paper/main.tex

--- 2021-06-02 paper/main.tex
@@
 
+We present our RBS optimizing DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
+Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
+in the DESIGN stage (Section~\ref{sec:ucb-results}),
+and we demonstrate that the ML predictions are accurate in the LEARN stage (Section~\ref{sec:gp-results}).
+We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and decribe some interesting
+characteristics of the discovered sequences, 
+as well as show the effectiveness of the automated laboratory workflow stages.
+
 \subsection{The experimental workflow}
+\label{sec:dbtl-workflow}
 
-We show our DBTL workflow in Figure \ref{fig: Flowchart}.
+\cheng{This section (I merged 2.1 and 2.2), feels like a method section. Rewrite to be a results section. 
+This can be much shorter}
+
+We show our DBTL workflow in Figure \ref{fig: Flowchart}, that uses machine learning to optimise protein expression.
 BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
@@
 
-\subsection{Design of the investigated genetic device}
+%\subsection{Design of the investigated genetic device}
 
@@
 
-\subsection{Performance of the recommendation algorithm}
+\subsection{DESIGN: Performance of the recommendation algorithm}
+\label{sec:ucb-results}
 
+\cheng{Write this section to describe Figure 2, in the order of the subfigures (can rearrange if needed). Again, feels like a method section.}
 The design recommendations were made using the Multi-armed Bandit algorithm.
@@
 
-\subsection{Prediction of RBS performance}
+\subsection{LEARN: Prediction of RBS performance}
+\label{sec:gp-results}
 
@@
 
+\cheng{Start this section with these results.}
 Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
@@
 
-\subsection{Characteristics of the tested sequences}
+\subsection{BUILD+TEST: Characteristics of the tested sequences}
+\label{sec:characteristics-of-library}
 
@@
 
+\cheng{Have a look at the paper I wrote with Mengyan, and summarise the first few opportunities and challenges there.}
 There are still open questions that need to be addressed for applying machine learning in synthetic biology.

### 2021-06-07 paper/main.tex

--- 2021-06-04 paper/main.tex
@@
 
-Figure \ref{fig: Scatterplot} shows our algorithms performance in terms of predictions in each round. 
+\cheng{Start this section with these results.}
+Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
@@
 
+\cheng{Have a look at the paper I wrote with Mengyan, and summarise the first few opportunities and challenges there.}
+There are still open questions that need to be addressed for applying machine learning in synthetic biology.
+For example, how can we understand the language of RBS sequences better?
+In other words, can we extract more biologically important information from the decisions made by our algorithms?
+Another example would be - given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
+Next question would be - how generalizable is this method? Can it be succesfully used for design of bigger genetic parts, like promoters?
+Finally, another question with high impact on the performance of the workflow is - how should we choose the parameters which controls the exploration-exploitation rate? \\
+
 We have found our approach bringing machine learning and synthetic biology experts very powerful.

### 2021-06-08 codes/config.py
update code

@@ -1,8 +1,10 @@
 # provide configuration, e.g. global variable, path
 
 BASES = ['A','C','G','T'] # embedding
+DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
 
 # SAVED_IDX_SEQ_PATH = '../../data/idx_seq.pickle'
+# SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
 SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
 
 # Usage:
@@ -13,5 +15,6 @@ SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
 # idx_seq_dict = saved_idx_seq['idx_seq_dict']
 
 # SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.pickle'
-SAVED_KERNEL_PATH = 'data/saved_normalised_kernel.pickle'
+# SAVED_KERNEL_PATH = 'data/saved_normalised_kernel.pickle'
+SAVED_KERNEL_PATH = DRIVE_PATH+ 'saved_normalised_kernel.pickle'
 

### 2021-06-08 codes/kernels_for_GPK.py

@@ -85,6 +85,10 @@ class String_Kernel(Kernel):
         
         self.kernel_over_all_flag = kernel_over_all_flag
 
+        self.feature = feature
+        self.n_train = n_train
+        self.n_test = n_test
+
         #----------------------------------------------------------------------------------------
         # reconstructure: use all 4138 unique seqs for kernel normalisation 
         if self.kernel_over_all_flag:
@@ -126,9 +130,6 @@ class String_Kernel(Kernel):
             # train n_train * d
             # test n_test * d
             # kernel_all (n_train + n_test) * (n_train + n_test)
-            self.feature = feature
-            self.n_train = n_train
-            self.n_test = n_test
             assert self.n_train + self.n_test == self.feature.shape[0]
 
             print('normalise over known features')

### 2021-06-08 codes/regression.py

@@ -233,7 +233,7 @@ class GPR_Predictor():
             print('create kernel instance')
             self.wd_kernel_instance = self.kernel(l = self.l, s = self.s, sigma_0=self.sigma_0, 
                                         kernel_norm_flag = self.kernel_norm_flag, centering_flag = self.centering_flag, unit_norm_flag = self.unit_norm_flag,
-                                        feature=feature, kernel_over_all_flag=False,
+                                        feature=feature, kernel_over_all_flag=True,
                                         n_train=X_train.shape[0], n_test=X_test.shape[0],)
             kernel_instance = self.wd_kernel_instance \
                             + WhiteKernel(noise_level=1e-5, noise_level_bounds=(1e-5, 1e+5))

### 2021-06-08 test/sanity_check.py

@@ -11,6 +11,9 @@
 import numpy as np
 import pandas as pd
 from codes.batch_ucb import *
+import codes.config
+
+DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
 
 
 design_round = 2
@@ -54,13 +57,13 @@ def check_dataset(data1, data2, design_round):
 #     source2 = "SynBio_repo_designs/SynbioML-round0/data/firstRound_Microplate_normTrue_formatSeq_logTrue.csv"
 if design_round == 1:  
     source1 = "data/pipeline_data/Results_bc1.csv"
-    source2 = "SynBio_repo_designs/SynbioML-round1/data/firstRound_Microplate_normTrue_formatSeq_logTrue.csv"
+    source2 = DRIVE_PATH + "SynBio_repo_designs/SynbioML-round1/data/firstRound_Microplate_normTrue_formatSeq_logTrue.csv"
 elif design_round == 2:
     source1 = "data/pipeline_data/Results_abc1.csv"
-    source2 = "SynBio_repo_designs/SynbioML-round2/data/Results_Microplate_partialTrue_normTrue_roundRep_formatSeq_logTrue.csv"
+    source2 = DRIVE_PATH + "SynBio_repo_designs/SynbioML-round2/data/Results_Microplate_partialTrue_normTrue_roundRep_formatSeq_logTrue.csv"
 elif design_round == 3:
     source1 = "data/pipeline_data/Results_abc1.csv"
-    source2 = "SynBio_repo_designs/SynbioML-round3/data/Results_Microplate_partialTrue_normTrue_mean_roundRep_formatSeq_logTrue.csv"
+    source2 = DRIVE_PATH + "SynBio_repo_designs/SynbioML-round3/data/Results_Microplate_partialTrue_normTrue_mean_roundRep_formatSeq_logTrue.csv"
 
 df_source1 = pd.read_csv(source1)
 df_source2 = pd.read_csv(source2)
@@ -94,4 +97,4 @@ if check_rec_flag:
 
 # Log: round 3 rec passed the test (overlap 90)
 # round 2 overlap 74
-# round 1 overlap 74
+# round 1 overlap 74 -> 88 (kernel normalisation over known features)

### 2021-06-08 paper/methods.tex

@@
+%! TEX root=./main.tex
 \section{Methods}
@@
 
-To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$ by balancing the predicted mean and standard deviation.
-\maciej{Not sure what does this mean? The balancing part}
+To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$.
 On the one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
@@
 by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
-\maciej{And do what to prevent this waste?}
+Batch recommendation strategies are designed to avoid recommendation such high-similarity sequences in the same batch.
 

### 2021-06-08 test/sanity_check.py
add reproducble log. Round 1-3 DONE.

@@ -95,6 +95,6 @@ if check_rec_flag:
     num_overlap = len(new_rec.intersection(lib_rec))
     print('The overlap for round ' + str(design_round) + ' is ' + str(num_overlap))
 
-# Log: round 3 rec passed the test (overlap 90)
-# round 2 overlap 74
+# Log: round 3 rec passed the test (overlap 90, gpbucb with beta = 0/top-n)
+# round 2 overlap 74 -> 90 (kernel normalisation over all design space, gpbucb with beta = 2)
 # round 1 overlap 74 -> 88 (kernel normalisation over known features)

### 2021-06-08 paper/methods.tex
Another round of cuts in the text.

@@
 
-For regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004}.
+For regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004,srinivas2012information, romero_navigating_2013}.
 We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
@@
 The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
-To represent the RBS sequences and formulate the similarity between sequences, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds} to specify the kernel function of $GP$.
+For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+To represent the RBS sequences in subsequent rounds, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds,Ben-Hur2008} to specify the kernel function of $GP$.  
 WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
@@
 To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$.
-On the one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
+On one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
 on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).

### 2021-06-08 codes/batch_ucb.py
add kernel over all flag para

@@ -73,7 +73,8 @@ class RBS_UCB():
     """
     def __init__(self, df_known, kernel_name='WD_Kernel_Shift', l=6, s=1, sigma_0 = 1,
                 embedding='label', alpha=2, rec_size=90, beta=2, kernel_norm_flag = True,
-                centering_flag = True, unit_norm_flag = True, df_design = None):
+                centering_flag = True, unit_norm_flag = True, kernel_over_all_flag = True,
+                df_design = None):
         self.df_known = df_known
         self.df_known['train_test'] = 'Train'
         self.known_rbs_set = set(self.df_known['RBS'])
@@ -108,7 +109,8 @@ class RBS_UCB():
                 eva_on = 'seq', # for design
                 kernel_norm_flag= kernel_norm_flag,
                 centering_flag=centering_flag,
-                unit_norm_flag=unit_norm_flag
+                unit_norm_flag=unit_norm_flag,
+                kernel_over_all_flag = kernel_over_all_flag
                 )
 
         # initialization for ucb parameters

### 2021-06-08 codes/config.py

@@ -5,7 +5,7 @@ DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
 
 # SAVED_IDX_SEQ_PATH = '../../data/idx_seq.pickle'
 # SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
-SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
+SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
 
 # Usage:
 # with open(SAVED_IDX_SEQ_PATH, 'rb') as handle:

### 2021-06-08 codes/regression.py

@@ -53,7 +53,7 @@ class GPR_Predictor():
                  kernel_name='WD_Kernel_Shift', l=6, s = 1, b=0.33, sigma_0 = 1, padding_flag=False, gap_flag=False,
                  alpha=2, embedding='label', eva_metric=[r2_score, mean_squared_error],  eva_on = "samples",
                  kernel_norm_flag = True, centering_flag = True, unit_norm_flag = True,
-                #  kernel_norm_flag = True,　centering_flag = True,　unit_norm_flag = True,
+                 kernel_over_all_flag = True
                  ):
         """
         Parameter
@@ -113,6 +113,7 @@ class GPR_Predictor():
         self.kernel_norm_flag = kernel_norm_flag
         self.centering_flag = centering_flag
         self.unit_norm_flag = unit_norm_flag
+        self.kernel_over_all_flag = kernel_over_all_flag
         # self.kernel.INIT_FLAG = False 
         # so that kernel will be initialised (cal_kernel)
         self.l = l
@@ -233,7 +234,7 @@ class GPR_Predictor():
             print('create kernel instance')
             self.wd_kernel_instance = self.kernel(l = self.l, s = self.s, sigma_0=self.sigma_0, 
                                         kernel_norm_flag = self.kernel_norm_flag, centering_flag = self.centering_flag, unit_norm_flag = self.unit_norm_flag,
-                                        feature=feature, kernel_over_all_flag=True,
+                                        feature=feature, kernel_over_all_flag=self.kernel_over_all_flag,
                                         n_train=X_train.shape[0], n_test=X_test.shape[0],)
             kernel_instance = self.wd_kernel_instance \
                             + WhiteKernel(noise_level=1e-5, noise_level_bounds=(1e-5, 1e+5))

### 2021-06-08 test/sanity_check.py

@@ -16,7 +16,7 @@ import codes.config
 DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
 
 
-design_round = 2
+design_round = 1
 
 rec_size = 90
 l = 6
@@ -82,11 +82,16 @@ if design_round == 3:
 else:
     beta = 2
 
+if design_round == 1: 
+    kernel_over_all_flag = False
+else:
+    kernel_over_all_flag = True
+
 if check_rec_flag:
     gpbucb = GP_BUCB(df_source1[df_source1['Round'] < design_round], kernel_name=kernel, l=l, s=s, sigma_0=sigma_0,
                     embedding=embedding, alpha=alpha, rec_size=rec_size, beta=beta, 
                     kernel_norm_flag=kernel_norm_flag, centering_flag = centering_flag,              
-                    unit_norm_flag=unit_norm_flag)
+                    unit_norm_flag=unit_norm_flag, kernel_over_all_flag = kernel_over_all_flag)
 
     gpbucb_rec_df = gpbucb.run_experiment()
     new_rec = set(np.asarray(gpbucb_rec_df['RBS6']))

### 2021-06-08 synbio_rbs/src/batch_ucb.py
add reproducible subfolder

@@ -1,0 +1,276 @@
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+import itertools
+from collections import defaultdict
+import math
+import json
+
+from sklearn.gaussian_process import GaussianProcessRegressor
+from sklearn.gaussian_process.kernels import PairwiseKernel, DotProduct, RBF 
+from sklearn.kernel_ridge import KernelRidge
+from sklearn.metrics import r2_score, mean_squared_error, make_scorer
+from sklearn.model_selection import KFold
+from sklearn_extra.cluster import KMedoids
+
+from src.embedding import Embedding
+from src.environment import Rewards_env
+from src.ucb import GPUCB, Random
+from src.evaluations import evaluate, plot_eva
+from src.regression import *
+from src.kernels_for_GPK import *
+import src.config
+
+from ipywidgets import IntProgress
+
+# Aug 2020 Mengyan Zhang
+# Implement batch ucb, with three methods:
+# 1. Return Top n arm directly (non-batch)
+# 2. Clustering 
+# 3. GP-BUCB
+
+KERNEL_DICT = {
+    # 'Spectrum_Kernel': Spectrum_Kernel,
+    # 'WD_Kernel': WeightedDegree_Kernel,
+    # 'Sum_Spectrum_Kernel': Sum_Spectrum_Kernel,
+    # 'Mixed_Spectrum_Kernel': Mixed_Spectrum_Kernel,
+    'WD_Kernel_Shift': WD_Shift_Kernel   
+}
+
+
+class RBS_UCB():
+    """GPUCB for RBS sequences design.
+
+    Attributes
+    -----------------------------------------------
+    df_known: pandas dataframe
+        known RBS sequences with labels
+        train data
+    df_design: pandas dataframe
+        core part design (4^6) expect those already in df_known
+        test data
+    df_train_test: pandas dataframe
+        concat of df_known and df_design
+
+    kernel_name: string
+        indicates kernel to use 
+    l: int
+        lmer
+    s: int
+        shift number
+    embedding: string
+        label or onehot
+    alpha: float
+        parameter for GPR, value added to kernel diagonal
+    
+    rec_size: int
+        recommendation size
+    beta: float
+        parameter for UCB, mean + beta * std
+        balance exploration and exploitation
+    
+    """
+    def __init__(self, df_known, kernel_name='WD_Kernel_Shift', l=6, s=1, sigma_0 = 1,
[…]

### 2021-06-08 synbio_rbs/src/config.py

@@ -1,0 +1,20 @@
+# provide configuration, e.g. global variable, path
+
+BASES = ['A','C','G','T'] # embedding
+DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
+
+# SAVED_IDX_SEQ_PATH = '../../data/idx_seq.pickle'
+# SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
+SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
+
+# Usage:
+# with open(SAVED_IDX_SEQ_PATH, 'rb') as handle:
+#     saved_idx_seq = pickle.load(handle)
+# seq_list = saved_idx_seq['seq_list']
+# idx_list = saved_idx_seq['idx_list']
+# idx_seq_dict = saved_idx_seq['idx_seq_dict']
+
+# SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.pickle'
+# SAVED_KERNEL_PATH = 'data/saved_normalised_kernel.pickle'
+SAVED_KERNEL_PATH = DRIVE_PATH+ 'saved_normalised_kernel.pickle'
+

### 2021-06-08 synbio_rbs/src/data_generating.py

@@ -1,0 +1,311 @@
+# Logs:
+
+# 23/04/2021 Update: add step names in comments
+
+# Define the following steps on each replicate:  
+# - a. In each round, substract the mean of every data points by the reference AVERAGE, and then add 100 (to make the values positive).  
+# - b. Take log (base e) transformation for each data points.  
+# - c. Apply z-score normalisation.  
+#     - c.1 on each round, so that the mean and variance of each replicate of data in each round is zero and one after normalisation. 
+#     - c.2 on all data, so that the mean and variance of each replicate of all data is zero and one after normalisation. 
+# - d. Apply min-max normalisation.
+#     - d.1 on each round
+#     - d.2 on all data
+# - e. Apply ratio normalisation. In each round, each data points is devided by the mean of refernce AVERAGE, so that in each round, the reference labels are almost 1. 
+#     - e.1 on each round
+#     - e.2 on all data
+
+# 25/11/2020 Update: add unique index corresponding to 'idx_seq_dict.npz' 
+
+# 23/10/2020 Update: normalisation for each plate each replicate, 
+# substracting mean of reference sequence in each plate
+
+# 04/11/2020 Update: add Rep7,8,9
+#-----------------------------------------------------------------------------------
+
+# This code is used to generate the ready to use data 
+# based on the first round result
+import numpy as np
+import pandas as pd
+import os
+import argparse
+import pickle
+
+# # input parameters
+# parser = argparse.ArgumentParser(description='Generate valid data.')
+# parser.add_argument('normalize_flag', default= 'True', help = 'indicates whether normalize labels. True for normalized label.')
+# parser.add_argument('to_design_round', default= '3', help = 'specify the round to design.')
+
+# args = parser.parse_args()
+# normalize_flag = str(args.normalize_flag) # str True or False
+# to_design_round = str(args.to_design_round) # 0-4
+
+
+# Setting 
+# Specify your choice
+approach = 'd1'
+# to_design_round = '4'
+folder_path = '../data/'
+
+def normalize(df, col_name):
+    if how_to_normalize == 'plateRep': # normalise for each plate each replicate
+        by_column = 'Plate'
+    elif how_to_normalize  == 'roundRep':
+        by_column  = 'Round'
+    else:
+        print('Unknown normalized type. Use roundRep instead')
+        by_column = 'Round'
+
[…]

### 2021-06-08 synbio_rbs/src/embedding.py

@@ -1,0 +1,102 @@
+import numpy as np
+from sklearn.feature_extraction.text import CountVectorizer
+from sklearn.preprocessing import LabelEncoder, normalize
+import src.config as config # provide configuration, e.g. global variable, path
+
+# 2nd Nov 2019
+# Mengyan Zhang
+
+class Embedding():
+    """Embedding for biological sequences.
+    """
+    def __init__(self, data = None):
+        """
+        Parameters
+        --------------------------------------
+        data: ndarray 
+            (num_seq, ) (each element is a biological sequence)
+        """
+        self.data = data
+        # self.num_seq, self.num_bases = data.shape
+        self.num_seq,  = data.shape
+        self.bases = config.BASES
+
+    def label(self):
+        """Label encoder from sklearn.
+        Encode each row from string into numeric label.
+        For example, from 'GACAAA' to [0 1 2 0 0 0]
+
+        Note when use inverse_transform (in kernel), the order
+        of bases (fitting) should be the same.
+
+        Returns
+        ----------------------------------------------------
+        embedded_data: array
+            num_seq * num_bases
+        """
+        le = LabelEncoder()
+        le.fit(self.bases)
+        # if data is None: # if no input data, return the encoder
+        #     return le
+        # else:
+        embedded_data = np.array([le.transform(list(self.data[i]))\
+                                    for i in range(self.num_seq)])
+        return embedded_data
+
+    def onehot(self):
+        """One-hot embedding.
+
+        Returns
+        --------------------------------------------
+        embedded_data: ndarray
+            {0, 1}^{num_seq x num_bases * 4}
+        """
+        
+        base_dict = dict(zip(self.bases,range(4))) # {'A' : 0, 'C' : 1, 'G' : 2, 'T' : 3}
+        self.num_bases = len(self.data[0])
+        embedded_data = np.zeros((self.num_seq, self.num_bases * 4))
+
+        # loop through the array of sequences to create a feature array 
+        for i in range(self.num_seq):
+            seq = self.data[i]
+            # loop through each individual sequence, from the 5' to 3' end
+            for b in range(self.num_bases):
+                embedded_data[i, b * 4 + base_dict[seq[b]]] = 1
+
+        embedded_data -= np.nanmean(embedded_data, axis = 0)
+        normalised_embedded_data = normalize(embedded_data, norm = 'l2')
+        return normalised_embedded_data
+
+    def kmer(self, size = 3):
[…]

### 2021-06-08 synbio_rbs/src/environment.py

@@ -1,0 +1,117 @@
+import numpy as np
+from collections import defaultdict, OrderedDict
+import operator
+from src.embedding import Embedding
+
+class Rewards_env():
+    """Rewards environment for biology sequence.
+
+    Attributes
+    -------------------------------------------------
+    rewards_dict: dict of list
+        keys: string of embedded sequence 
+            e.g. '001000100010001000100010'
+        values: list of available labels
+    labels_dict: dict of list
+        keys: string of embedded sequence 
+            e.g. '001000100010001000100010'
+        values: label (expectation of rewards list)
+    """
+    def __init__(self, data, embedding_method = 'onehot'):
+        """
+        Parameters
+        ---------------------------------------------
+        data: ndarray 
+            num_data * 2
+            two columns: biology sequence; score (label)
+        embedding_method: string
+            specify embedding method
+            default is 'onehot'
+        """
+        self.data = data
+        self.num_seq, self.num_col = self.data.shape
+        # assert self.num_col == 2
+        self.embedding_method = embedding_method
+
+        self.embedded = self.arm_embedding()
+
+        # TODO: comment for now, may need to put it back
+        # self.rewards_dict = defaultdict(list)
+        # self.create_rewards_dict()
+        # self.labels_dict = self.labels_generator()
+        # self.arm_features = list(self.labels_dict.keys())
+        
+    def arm_embedding(self):
+        """Embedding biological sequence.
+        """
+        # encoder = Embedding(self.data[:, 0].reshape(self.num_seq, 1))
+        encoder = Embedding(self.data[:, 0])
+        if self.embedding_method == 'label':
+            embedded= encoder.label()
+        elif self.embedding_method == 'onehot':
+            embedded = encoder.onehot()
+        elif self.embedding_method == 'kmer':
+            embedded = encoder.kmer()
+        else:
+            print('To be added.')
+            embedded = self.data[:, 0]
+        return embedded
+
+    def create_rewards_dict(self):
+        """Create reward dict based on the embedded features.
+        """
+        for i, d in enumerate(self.data):
+            key = self.to_string(self.embedded[i])
+            self.rewards_dict[key].append(d[1])
+
+        # remove all sort
+        # sort the rewards dict by keys
+        # self.rewards_dict = dict(sorted(self.rewards_dict.items()))
+        self.rewards_dict = dict(self.rewards_dict.items())
+
+    def labels_generator(self):
+        """Generate labels for each arm.
+
+        Return 
+        -----------------------------------------
[…]

### 2021-06-08 synbio_rbs/src/evaluations.py

@@ -1,0 +1,72 @@
+import matplotlib.pyplot as plt
+import numpy as np
+from collections import defaultdict
+
+ylabel_dict = {'sd': 'suboptimal draws',
+                'r': 'cumulative regrets',
+            }
+
+def evaluate(sds, rs):
+    """Calculated expected evaluation metrics (suboptimal draws, regret)
+
+    Parameters
+    ----------------------------------------------------
+    sds: list 
+        len: num_exper; list of sub-optimal draws for each experiment
+    rs: list
+        len: num_exper; list of cumulative regrets for each experiment
+
+    Returns
+    ----------------------------------------------------
+    eva_dict: dict
+        key: name of evaluation metrics
+            one of 'sd' and 'r'
+        value: array with shape 1 * num_rounds
+    """
+    eva_dict =  defaultdict(dict)
+    for key, values in sds.items():
+        eva_dict[key]['sd'] = defaultdict(list)
+        eva_dict[key]['sd'] = np.mean(np.asarray(values), axis = 0)
+    for key, values in rs.items():
+        eva_dict[key]['r'] = defaultdict(list)
+        eva_dict[key]['r'] = np.mean(np.asarray(values), axis = 0)
+    return eva_dict
+
+def plot_eva(results, eva_method, paper_flag = True):
+    """Plot method for evaluations
+
+    Parameters
+    -----------------------------------------------------------
+    results: dict
+        keys: 'env name + num_exper + num_rounds'
+        values: dict
+            keys: 'est_var + hyperpara' or 'bound'
+            values: dict
+                keys: 'sd', 'r'
+                values: list of result  
+    eva_method: str
+        options ('sd', 'r')
+    paper_flag: boolean
+        indicates whether plotting for paper.
+        True is for paper; False is not.
+    """
+    plt.figure(figsize=(5, 5))
+       
+    # setup title
+    title_name = 'Cumulative Regret vs. Round'
+    plt.title(title_name)
+
+    plt.xlabel('Iteration')
+    plt.ylabel('Expected ' + ylabel_dict[eva_method])
+    
+    for key, value in results.items():
+        plt.plot(range(len(value[eva_method])), 
+                value[eva_method],  
+                marker = 'o', 
+                markevery = 10,
+                markersize = 5,
+                label = key)
+
+    plt.legend(loc="upper left")
+    #file_name = 'Exper_' + str(eva_method) + '_' + name + '_' + subname + '.pdf'
+    #plt.savefig(file_name, bbox_inches='tight')

### 2021-06-08 synbio_rbs/src/generate_idx_seq.py

@@ -1,0 +1,101 @@
+# Nov.2020 Mengyan Zhang 
+# This file generates a two-way dict (index, sequences) for all design RBS sequences,
+# including 4^6 (core-part) + 3*14 (bps non-core group) = 4138 sequences. 
+
+# usage:
+# with open(Path, 'rb') as handle:
+#    idx_seq = pickle.load(handle)
+
+# idx_seq has three keys
+# dict_keys(['idx_seq_dict', 'idx_list', 'seq_list'])
+
+# idx_list and seq_list gives you the idx/seq in the same order
+# idx_seq_dict is a two way dict, 
+# you can do idx_seq_dict[20-long-RBS] = idx or idx_seq_dict[idx] = 20-long-RBS
+
+import os
+import sys
+module_path = os.path.abspath(os.path.join('../..'))
+if module_path not in sys.path:
+    sys.path.append(module_path)
+import pandas as pd
+import numpy as np
+import itertools
+import pickle
+
+# class TwoWayDict(dict):
+#     """bijective map between index and sequences.
+#     code from Sasha Chedygov's answer https://stackoverflow.com/questions/1456373/two-way-reverse-map
+#     """
+#     def __setitem__(self, key, value):
+#         # Remove any previous connections with these values
+#         if key in self:
+#             del self[key]
+#         if value in self:
+#             del self[value]
+#         dict.__setitem__(self, key, value)
+#         dict.__setitem__(self, value, key)
+
+#     def __delitem__(self, key):
+#         dict.__delitem__(self, self[key])
+#         dict.__delitem__(self, key)
+
+#     def __len__(self):
+#         """Returns the number of connections"""
+#         return dict.__len__(self) // 2 
+
+
+# Setting
+char_sets = ['A','C','G','T']
+design_len = 6
+pre_design = 'TTTAAGA'
+pos_design = 'TATACAT'
+index = 0
+index_list = []
+seq_list = []
+idx_seq_dict = {}
+
+for combo in itertools.product(char_sets, repeat= design_len):
+    combo = pre_design + ''.join(combo) + pos_design
+    index_list.append(index)
+    seq_list.append(combo)
+    idx_seq_dict[index] = combo
[…]

### 2021-06-08 synbio_rbs/src/kernels_for_GPK.py

@@ -1,0 +1,598 @@
+import numpy as np
+import pandas as pd
+from sklearn.feature_extraction.text import CountVectorizer
+from sklearn import preprocessing
+from sklearn.gaussian_process.kernels import Kernel, Hyperparameter
+import matplotlib.pyplot as plt
+import os
+import sys
+module_path = os.path.abspath(os.path.join(''))
+print (module_path)
+print(sys.path)
+if module_path not in sys.path:
+    sys.path.append(module_path)
+from src.embedding import Embedding
+from src.environment import Rewards_env
+
+
+import pickle
+import src.config as config # provide configuration, e.g. global variable, path
+
+# June 2021 Mengyan Zhang 
+# String Kernel Classes, taking strings as input
+# including spectrum kernel, weighted degree kernel and weighted degree kernel with shift 
+# The implementation is based on Support Vector Machines and Kernels for Computational Biology 
+# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2547983/
+# Inherit from sklearn gaussain process kernel class
+# https://github.com/scikit-learn/scikit-learn/blob/b194674c4/sklearn/gaussian_process/kernels.py#L1213
+
+# Two choices of kernel normalisation target:
+# 1. normalisation over the all design space. The kernel matrix is pre-calculated and saved. 
+# 2. normalisation over all currently known RBS sequences. 
+# For Gaussian Process Regression, the specific kernel matrix is sliced from the calculated kernel (in __call__)
+
+
+class String_Kernel(Kernel):
+    """Base class for string kernels
+    
+    Parameters
+    ------------------
+    * For initialisation, calculating the whole kernel
+
+    features: array
+        list of features for all training and testing data
+    n_train: int
+        number of training data
+    n_test: int 
+        number of testing data
+    kernel_all: (n_train+n_test, n_train+n_test)
+        kernel matrix for all data
+    kernel_all_normalised: (n_train+n_test, n_train+n_test)
+        kernel matrix for all data with centering and unit norm
+
+    * Kernel hyperparameters *
+    l : int, default 3
+        number of l-mers (length of 'word')
+    # TODO: change hyperparameters maybe (now set to the same as Dotproduct)
+    sigma_0 : float >= 0, default: 0.0
+        signal std
+        Parameter controlling the inhomogenity of the kernel. If sigma_0=0,
+        the kernel is homogenous.
+    """
+    
+    INIT_FLAG = False
[…]

### 2021-06-08 synbio_rbs/src/regression.py

@@ -1,0 +1,673 @@
+# update: 25/Nov/2020
+# use unique seq for kernel normalisation
+
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from matplotlib import gridspec
+import plotly
+import plotly.graph_objs as go
+import seaborn as sns
+import itertools
+from collections import defaultdict
+import math
+import json
+import xarray as xr  
+
+from sklearn.gaussian_process import GaussianProcessRegressor
+from sklearn.gaussian_process.kernels import PairwiseKernel, DotProduct, RBF, WhiteKernel
+from sklearn.kernel_ridge import KernelRidge
+from sklearn.metrics import r2_score, mean_squared_error, make_scorer
+from sklearn.model_selection import KFold
+from sklearn.cluster import KMeans, AgglomerativeClustering
+from scipy.stats import spearmanr
+
+from src.embedding import Embedding
+from src.environment import Rewards_env
+from src.evaluations import evaluate, plot_eva
+from src.kernels_for_GPK import *
+from src.sort_seq import sort_kernel_matrix
+
+from ipywidgets import IntProgress
+import src.config # provide configuration, e.g. global variable, path
+
+# Aug 2020 Mengyan Zhang
+# Implement predictors based on Gaussian Process Regression
+
+KERNEL_DICT = {
+    # 'Spectrum_Kernel': Spectrum_Kernel,
+    # 'Mixed_Spectrum_Kernel': Mixed_Spectrum_Kernel,
+    # 'WD_Kernel': WeightedDegree_Kernel,
+    # 'Sum_Spectrum_Kernel': Sum_Spectrum_Kernel,
+    'WD_Kernel_Shift': WD_Shift_Kernel,
+    'RBF': RBF  
+}
+
+REP_LIST = ['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5','Rep6']
+
+def valid_name(name):
+    return name.replace('_', ' ')  
+def valid_path(path):
+    return path.replace(' ', '_').replace(',', '')
+
+
+class GPR_Predictor():
+    def __init__(self, df, test_size=0.2, train_idx = None, test_idx = None, 
+                 kernel_name='WD_Kernel_Shift', l=6, s = 1, b=0.33, sigma_0 = 1, padding_flag=False, gap_flag=False,
+                 alpha=2, embedding='label', eva_metric=[r2_score, mean_squared_error],  eva_on = "samples",
[…]

### 2021-06-08 synbio_rbs/src/sort_seq.py

@@ -1,0 +1,88 @@
+import numpy as np
+import pandas as pd
+from collections import defaultdict
+import math
+from sklearn.cluster import KMeans, AgglomerativeClustering
+
+SORTED_GROUP = ['Consensus', 'BPS-C', 'BPS-NC', 'UNI', 'PPM', 'Bandit-0', 'Bandit-1']
+
+def sort_kernel_matrix(df, feature_kernel, kmeans_based_on='label_distance'):
+    group_dict = df.groupby('Group').groups
+    groups = list(group_dict.keys())
+    new_ordering = [] # idx ordering
+    intersection = [i for i in SORTED_GROUP if i in set(groups)]
+    print('groups: ', intersection)
+
+    frr_seqs = np.asarray(df['RBS'])
+
+    for group in intersection:
+        df_group = df[df['Group'] == group]
+        num_seqs = len(df_group)
+        # num_clusters = int(num_seqs/5) + 1
+        # print('Group: ', group)
+        # print('Number of sequences: ', num_seqs)
+        # print('number of clusters: ', num_clusters)
+        
+        idx = np.asarray(group_dict[group])
+        # print('idx: ', idx)
+        #print(label_distance[idx[0]: idx[-1], idx[0]: idx[-1]])
+        
+        
+        if kmeans_based_on == 'label_distance': # kmeans based on label distances
+            num_clusters = int(num_seqs/5) + 1
+            # print('number of clusters: ', num_clusters)
+            kmeans = KMeans(n_clusters = num_clusters, random_state = 0).fit(np.asarray(df['AVERAGE'])[idx].reshape(len(idx),1))
+            cluster_dict = defaultdict(list) # key: cluster id; value: idx list
+            for i, cluster_id in enumerate(kmeans.labels_):
+                cluster_dict[cluster_id].append(idx[i])
+            # print('cluster dict: ', cluster_dict)
+            # print('kmeans labels: ', kmeans.labels_)
+            
+        elif kmeans_based_on == 'seq_distance': # kmeans based on spectrum distances 
+            num_clusters = int(num_seqs/8) + 1
+            # print('number of clusters: ', num_clusters)
+            #kmeans = KMeans(n_clusters = num_clusters, random_state = 0).fit(phi_X[idx[0]: idx[-1] + 1, :])
+            cluster_dict = defaultdict(list) # key: cluster id; value: idx list
+            if len(idx) > 1:
[…]

### 2021-06-08 paper/main.tex
address Maciej's comment in method sec

--- 2021-06-04 paper/main.tex
@@
 
-\title{Machine Learning guided workflow for Ribosome Binding Site engineering}
+\title{Machine Learning guided workflow\\
+for Ribosome Binding Site engineering}
 
@@
 
+We present our RBS optimizing DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
+Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
+in the DESIGN stage (Section~\ref{sec:ucb-results}),
+and we demonstrate that the ML predictions are accurate in the LEARN stage (Section~\ref{sec:gp-results}).
+We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and describe some interesting
+characteristics of the discovered sequences, 
+as well as show the effectiveness of the automated laboratory workflow stages.
+
 \subsection{The experimental workflow}
+\label{sec:dbtl-workflow}
 
-We show our DBTL workflow in Figure \ref{fig: Flowchart}.
+We show our DBTL workflow in Figure \ref{fig: Flowchart}, that uses machine learning to optimise protein expression.
 BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
 Machine learning algorithms are applied in LEARN and DESIGN.
-In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
-The goal is to provide predictions about TIR labels and uncertainty to assist the recommendation process.
-In the DESIGN phase, we have used a multi-armed Bandit recommendation algorithm that was selecting RBS designs to be tested in subsequent experimental batches.\\
+In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.\\
 
@@
 \end{figure}
-To help us obtain reliable and reproducible results we have employed automation-heavy workflow in the BUILD and TEST phases.
-This way we were able to eliminate a big part of sample-to-sample variation as well as human-introduced variation.
-Additionally, performing all the procedures directly in 96-well microplate format enabled us to significantly cut down the time required to prepare our variants.\\
-
-In short, the genetic variations of the RBS were introduced to the plasmids with combination of PCR and isothermal assembly. 
-The plasmids were then transformed and the resulting transformants were tested using a  microplate reader.
-Vast majority of reactions were prepared using liquid handling equipment.
-Similarly, colony picking was done by an automated colony picker.\\
-
-\subsection{Design of the investigated genetic device}
 
@@
 
-In \emph{E. coli}, the RBS is usually located in the 20 bases upstream of the start codon. 
-Additionally, there is a consensus RBS core sequence called the \textit{Shine-Dalgarno sequence}, which in \emph{E. coli} is \textbf{AGGAGG}. 
-Here, we put that 20 bp long sequence into focus with main emphasis being put on the 6bp core region
-(see detail in Figure \ref{fig: Flowchart}).\\
-
-Our template RBS sequence, is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA.
+Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
 This sequence is a known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
-Since this is the sequence against which new RBS sequenced will be benchmarked,
+Since this is the sequence against which new RBS sequences will be benchmarked,
 we will refer to this sequence as the \textit{benchmark sequence} from now on.
 We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
-In effect, in our design we focus on randomising of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e.
+In effect, in our design we focus on modification of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e.
 TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).
@@
 
-\subsection{Performance of the recommendation algorithm}
+\subsection{DESIGN: Performance of the recommendation algorithm}
+\label{sec:ucb-results}
 
@@
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-In our case we use the Upper Confidence Bound version of the algorithm, which focuses its recommendations on sequences that should give highest TIR based on the probabilities computed by the prediction algorithm (see Figure \ref{fig:batch rec}).\\
+\\
 
-Another feature of the bandits algorithm is that it balances two aspects: the exploration of the unknown (untested) parts of the design space where high TIR RBS can be hidden, and exploitation which goal is to query areas which are known to give relatively high TIRs.\\
+Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
+In each round, we measure the TIR of benchmark RBS as the internal standard. 
+We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\
 
@@
 
-Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
-In each round, we measure the TIR of benchmark RBS as the internal standard. 
-We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
@@
 
-\subsection{Prediction of RBS performance}
+\subsection{LEARN: Prediction of RBS performance}
+\label{sec:gp-results}
 
-Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
-This algorithm creates a model which takes the RBS sequences as input and predicts the mean TIR values with uncertainty level about that prediction, based on the experimental data.
-For this study, we have used the Gaussian Process Regression (GPR) as our prediction method.
-GPR is a Bayesian approach and has been widely used for experimental design \cite{srinivas2012information, romero_navigating_2013}.
-The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
-
-A crucial ingredient in a Gaussian Process predictor \cite{Rasmussen2004} is the kernel (covariance) function, which captures the similarity between input data points, in our case RBS sequences.
-Specifically, kernel function implicitly embeds RBS sequences into high-dimensional feature space which makes the regression process possible.
-For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
-For subsequent rounds, we used the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds}, which has been shown to have good performance in various prediction tasks \cite{Ben-Hur2008}.
-The WDS not only counts the matches of substrings of a certain length (i.e. kmers), but also takes into account the positional information and shifting of substrings.
-The relative distances between sequences calculated by our kernel function are shown in Figure \ref{fig: Swarmplot and Quantplot}E.\\
-
-Figure \ref{fig: Scatterplot} shows our algorithms performance in terms of predictions in each round. 
+Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
@@
 In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
-However, this is still very useful information for future predictions as it allows us to understand the underlying space better.
+However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\
+
+Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
+This algorithm creates a model which takes the RBS sequences as input and predicts the mean TIR values with uncertainty level about that prediction, based on the experimental data.
+For this study, we have used the Gaussian Process Regression (GPR) as our prediction method.
+The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
 
@@
 
+There are still open questions that need to be addressed for applying machine learning in synthetic biology.
+For one, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
+Secondly, given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
+Thirdly, the generalisability of the method is unknown.
+We believe that the method would be useful for use in desgin of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence.
+Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.\\
+
 We have found our approach bringing machine learning and synthetic biology experts very powerful.

### 2021-06-09 synbio_rbs/src/batch_ucb.py
finish repro pipeline round 1-3

@@ -16,7 +16,6 @@ from sklearn_extra.cluster import KMedoids
 
 from src.embedding import Embedding
 from src.environment import Rewards_env
-from src.ucb import GPUCB, Random
 from src.evaluations import evaluate, plot_eva
 from src.regression import *
 from src.kernels_for_GPK import *

### 2021-06-09 synbio_rbs/src/data_generating.py

@@ -43,11 +43,19 @@ import pickle
 
 # Setting 
 # Specify your choice
-approach = 'd1'
 # to_design_round = '4'
-folder_path = '../data/'
+Folder_path = '../data/'
 
-def normalize(df, col_name):
+#-------------------------------------------------------------------------------------------------------------
+# Read and write path
+
+# Folder_path = os.getcwd() # folder path might need to change for different devices
+Results_path = 'Results_Masterfile.xlsx'
+Predictions_path = 'Designs/design_pred.xlsx'
+idx_seq_path = 'idx_seq.pickle'
+
+
+def normalize(df, col_name, how_to_normalize, substract_ref_mean, Log_flag, Norm_method):
     if how_to_normalize == 'plateRep': # normalise for each plate each replicate
         by_column = 'Plate'
     elif how_to_normalize  == 'roundRep':
@@ -132,7 +140,7 @@ def assign_idx(df):
     with open(Folder_path + idx_seq_path, 'rb') as handle:
         idx_seq_dict = pickle.load(handle)['idx_seq_dict']
     for df_index, row in df.iterrows():
-        print(idx_seq_dict[str(row['RBS']).upper()])
+        # print(idx_seq_dict[str(row['RBS']).upper()])
         df.loc[df_index,'idx'] = idx_seq_dict[str(row['RBS']).upper()]
     
     # put index at first col
@@ -143,6 +151,7 @@ def assign_idx(df):
 
 def generate_data(approach = 'n'):
     approach = approach.lower()
+    Generated_File_Path = 'Results_' + approach + '.csv'
 
     if 'a' in approach:
         substract_ref_mean = True
@@ -179,15 +188,6 @@ def generate_data(approach = 'n'):
     # str(args.Format) # str Seq or Sample
     # parser.add_argument('Format', default='Seq', help = 'Seq for rows as sequences; Sample for rows as samples')
     sheet_name = 'Microplate' # for masterfile
-
-    #-------------------------------------------------------------------------------------------------------------
-    # Read and write path
-
-    Folder_path = os.getcwd() # folder path might need to change for different devices
-    Results_path = folder_path + 'Results_Masterfile.xlsx'
-    Predictions_path = folder_path + 'Designs/design_pred.xlsx'
-    idx_seq_path = folder_path + 'idx_seq.pickle'
[…]

### 2021-06-09 paper/main.tex
Updates from Overleaf

@@
 
-One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
+One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
@@
 
-We present our RBS optimizing DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
+We present our RBS optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
 Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
@@
 
-We show our DBTL workflow in Figure \ref{fig: Flowchart}, that uses machine learning to optimise protein expression.
+We show our DBTL workflow, that uses machine learning to optimise protein expression, in Figure \ref{fig: Flowchart}.
 BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
 Machine learning algorithms are applied in LEARN and DESIGN.
-In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.\\
+In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
+In DESIGN phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS designs based on the predictions from LEARN.\\
 
@@
 Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
-This sequence is a known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
+This sequence is known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
 Since this is the sequence against which new RBS sequences will be benchmarked,
@@
 
+\mengyan{Maciej: do we want to put the group description into Fig 2 as we discussed last Fri?}
+\maciej{I think we have decided to put it in the text, but within the description of Fig 2A as it is now. Otherwise the caption would be suuuppppppeeeerrrr long}
 Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
-In each round, we measure the TIR of benchmark RBS as the internal standard. 
+In each experimental round, we measure the TIR of benchmark RBS as the internal standard. 
 We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\
@@
     \textbf{TIRs of RBS groups examined in this study.} 
-    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
     \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
@@
     BN - benchmark sequences for all plates. 
-    They are not all exactly 1 due to them being shown as separate samples rather than averages.
+    They are not all exactly 1 due to them being shown as separate samples rather than per round averages.
     \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
     % save the random groups which were shown together due to similar distributions.
-    UNI and PPM are merged into Random group and BPS-NC was removed due to changes being made outside the core in this group. 
+    UNI and PPM are merged into Random group and BPS-NC is not shown due to changes being made outside the core in that group. 
     \textbf{C)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration. 
     \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
-    \textbf{E)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
-    The area of the circle represents the experimentally obtained TIR for measured groups.}
+    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
+    The area of the circle corresponds to the the experimentally obtained TIR value.
+    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
+}
     \label{fig: Swarmplot and Quantplot}
@@
 
-However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
+However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results improve significantly, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
 In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
@@
 Those with high predicted mean are coloured blue and represent exploitation, those hued red are with high predicted uncertainty and represent exploration.
-We can see the RBSs with high TIRs tend to come from exploitation the design space whereas the explorative points give relatively low TIR but expand our knowledge about the unknown part of the design space.\\
+We can see that the RBSs with high TIRs tend to come from exploitation of the design space whereas the explorative points give relatively low TIR, but expand our knowledge about the unknown part of the design space.\\
 
@@
 We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
-A number of clusters were targeted by our recommendation algorithm.
+Additionally, a number of clusters were especially targeted by our recommendation algorithm.
 For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
@@
 
-Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
+Figure \ref{fig: Scatterplot} shows how our implemetation of the Gaussian Process algorithm performed in terms of predictions in each round. 
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
 The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
-Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
+Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.\\
+
 One important point to note is that the predictions are also influenced by our recommendation choices. 
@@
 However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\
-
-Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
-This algorithm creates a model which takes the RBS sequences as input and predicts the mean TIR values with uncertainty level about that prediction, based on the experimental data.
-For this study, we have used the Gaussian Process Regression (GPR) as our prediction method.
-The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\
 
@@
 
-\subsection{BUILD+TEST: Characteristics of the tested sequences}
+\subsection{BUILD \& TEST: Characteristics of the tested sequences}
 \label{sec:characteristics-of-library}
 
-Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in Table 1.
+Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in Table 1.\\
+
 Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (in terms of TIR ratio).
@@
 This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
-This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as ones with high TIR probability.
+This result combined with the Bandits' algorithm bias towards the G rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as the ones with high TIR probability.\\
 
@@
 
+We have found our approach bringing machine learning and synthetic biology experts very powerful.
+We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.\\
+
 There are still open questions that need to be addressed for applying machine learning in synthetic biology.
@@
 Thirdly, the generalisability of the method is unknown.
-We believe that the method would be useful for use in desgin of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence.
-Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.\\
-
-We have found our approach bringing machine learning and synthetic biology experts very powerful.
-We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
-However, there are still some open questions that have to be answered.
-For example, we are still not sure how generalisable are the used methods and if it possible to extract more biological knowledge from the algorithms.
-It is also not clear how to improve the accuracy of predictions, but we believe this will occur through  a combination of bigger data sets and better algorithms.\\
+We believe that the method would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
+Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.
+In this research we have decided to gradually move the balance from exploration to exploitation over the rounds, but it is not yet clear what is the method protocol for this change.\\
 
@@
 All the processed and raw data is included in the repository.
-Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as separate file.
+Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.
 

### 2021-06-09 paper/methods.tex

@@
 In round $t$, the prediction and design is based on the results obtained in all previous rounds. 
-Our implementation of the machine learning algorithms was tested in Python 3.7 and used the scikit-learn library \cite{scikit-learn}. 
+Our implementation of the machine learning algorithms was tested in Python 3.6 and used the scikit-learn library \cite{scikit-learn}. 
 In the following paragraphs, we describe our machine learning pipeline (Bandit 1-3) in the following order: data pre-processing, prediction and kernels and finally recommendation. 
@@
 To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+\\
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
 $y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
+\\
 % This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g. time, temperature, operator, etc.). 
@@
 GPR can predict both the posterior mean and standard deviation. The posterior standard deviation represents the level of uncertainty for the prediction. 
-We review the predictions of GPR in Supplementary \ref{supp: GPR}.
+We review the predictions of GPR in Supplementary \ref{supp: GPR}.\\
 % We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0. 
@@
 This could be useful when there is a shift between two sequences.
-For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. 
+For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. \\
 % there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.
@@
 % In our design, the length of RBS sequence $L = 20$.
-The hyperparameters for kernel, including maximum substring length $\ell$, maximum shift length $S(l)$, and the standard deviation of GP noise $\alpha$ were chosen based on 10-repeat 5-fold cross validation.
+The hyperparameters for kernel, including maximum substring length $\ell$, maximum shift length $S(l)$, and the standard deviation of GP noise $\alpha$ were chosen based on 10-repeat 5-fold cross validation.\\
 %  ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).
@@
 $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
-We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.\\
 
@@
 by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
-Batch recommendation strategies are designed to avoid recommendation such high-similarity sequences in the same batch.
+Batch recommendation strategies are designed to avoid recommendation such high-similarity sequences in the same batch. \\
 
@@
 One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
-The strategy is to recommend sequences sequentially by updating the UCB score with the updated predicted standard deviation with the previously recommended points.
+The strategy is to recommend sequences sequentially by updating the UCB score with the updated predicted standard deviation with the previously recommended points.\\
 % Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
 % The plot shows the predicted mean and standard deviation after observing 4 data points. 
-As illustrated in Figure \ref{fig:batch rec} (B), the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
+As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
 Then we add the recommended data point ($x = 2$) into the training dataset with the predicted mean of that point as label (note it is not the true label, i.e.
@@
 The second data point is then recommended based on the new UCB scores.   
-As we can see in Figure \ref{fig:batch rec}(C), since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
+As we can see in Figure \ref{fig:batch rec}C, since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
 In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 

### 2021-06-09 synbio_rbs/src/RBS_extractor.py
Updated the RBS design example and added other useful code.

@@ -1,0 +1,51 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jan  9 12:24:10 2020
+
+@author: hol428
+"""
+
+import sys
+from Bio import SeqIO
+import xlwt
+
+seq_record = SeqIO.read('Citrobacter freundii CFNIH1, complete genome.gb','gb')
+cds_list_plus = []
+cds_list_minus = []
+for feature in seq_record.features:
+    if feature.type == 'CDS':
+        if feature.strand == -1:
+            mystart = feature.location.end + 0
+            myend = feature.location.end + 20 
+            cds_list_minus.append((mystart,myend,-1))
+        elif feature.strand == 1:
+            mystart = feature.location.start - 20
+            myend = feature.location.start + 0
+            cds_list_plus.append((mystart,myend,1))
+        else:
+            sys.stderr.write("No strand indicated %d-%d. Assuming +\n" %
+                              (mystart, myend))
+            cds_list_plus.append((mystart,myend,1))
+
+
+rbsplus = []
+rbsminus = []
+
+for item in cds_list_plus:
+    rbsplus.append(str((seq_record.seq[item[0]:item[1]])))
+    
+for item in cds_list_plus:
+    rbsminus.append(str((seq_record.seq[item[0]:item[1]].reverse_complement())))
+
+rbs_full = rbsplus + rbsminus
+print(rbs_full)
+
+book = xlwt.Workbook()
+sh = book.add_sheet('RBS')
+
+for el,row in zip(rbs_full,range(0,len(rbs_full))):
+    sh.write(row,0,'GenomicRBS' + '_' + str(row))
+    sh.write(row,1,str(el))
+
+book.save('Citrobacter freundii CFNIH1, complete genome' + '.xls')
+    

### 2021-06-09 synbio_rbs/src/RandomRBSgenerator.py

@@ -1,0 +1,43 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed 20 Sep 14:16:13 2019
+@author: hol428
+Script to generate random sequences of DNA
+"""
+
+import random
+import xlwt
+
+#define function that generates a stretch of DNA of n length 
+def nrandom (length=6,char=['A','T','G','C']):
+    return ''.join(random.choice(['A','T','G','C']) for _ in range(length))
+
+sequences = []
+tester = []
+seqN = 4096
+
+#sequence generator
+for seq in range (0,seqN):
+    #regeneration of random sequences on each pass
+    
+    seq2 = nrandom(6)
+    
+    while seq2 in tester:
+        seq2 = nrandom(6)
+    
+    tester.append(seq2)
+    
+    sequences.append(nrandom(6))
+    print('Generating sequence number: ' + str(seq))
+    
+#create Excel workbook to save the sequences to
+book = xlwt.Workbook()
+sh = book.add_sheet('Sequences')
+
+#generate the sequences, however many you like
+for el,row in zip(tester,range(0,seqN)):
+    sh.write(row,0,'Random_RBS' + '_' + str(row))
+    sh.write(row,1,str(el))
+
+#save the workbook
+book.save('RBSseq' + '.xls')

### 2021-06-09 synbio_rbs/src/TIR_calculator.py

@@ -1,0 +1,119 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Oct  8 13:45:20 2020
+
+@author: hol428
+"""
+
+import pandas as pd
+import glob
+import re
+import openpyxl as ox
+
+def create_gfpod():
+    GFPOD = OD.copy(deep=True)
+    i=0
+    
+    OD.drop('Time',axis=1,inplace=True)
+    
+    for col in OD.columns:
+        for ind in GFPOD.index:
+            # print(col)
+            # print(ind)
+            GFPOD[col][ind]=GFP[col][ind]/OD[col][ind]
+            
+    for ind in GFPOD.index:
+        GFPOD['Time'][ind]=i
+        i+=10
+    
+    GFPOD.set_index('Time',inplace=True)
+    
+    with pd.ExcelWriter(File, engine="openpyxl", mode='a') as writer:  
+        GFPOD.to_excel(writer, sheet_name='GFPOD')
+
+def create_tir():
+    GFPOD.drop('Time',axis=1,inplace=True)
+    # GFPOD.reset_index(drop=True,inplace=True)
+        
+    minp = GFPOD.min(axis=0)
+    
+    time = [240]
+    
+    rates = pd.DataFrame(columns=['240','Comment'],index = list(GFPOD.columns))
+    
+    for t in time:
+        for column,value in minp.iteritems():
+            start = GFPOD[GFPOD[column]==value].index.values.astype(int)[0]
+            end = start + (t/10)
+            rate = (GFPOD.loc[start:end, [column]].sum() / t)
+            # print(start)
+            # print(end)
+            # print(rate)
+            rates.at[column,str(t)] = float(rate[0])
+            if rate[0]<0:
+                rates.at[column,'Comment'] = 'Impossible value'
+            elif rate[0]>100 and OD.max(axis=0)[column]<0.1:          
+                rates.at[column,'Comment'] = 'Possibly an empty well'
+            else:
+                rates.at[column,'Comment'] = 'Correct value'
+    
+    print(rates)
+    
+    with pd.ExcelWriter(File, engine="openpyxl", mode='a') as writer:  
+        rates.to_excel(writer, sheet_name='TIR')
+
+full_report = 'no'
+path = '../../data/Plate_results/*.xlsx'
+ResFile = '../../data/Results_Masterfile.xlsx'
+Results = pd.read_excel(ResFile,sheet_name='Microplate') 
[…]

### 2021-06-10 paper/main.tex

@@
 After generating the recommendations the RBS are built and tested using automated laboratory methods allowing for rapid construction and testing at scale.
-Finally, the obtained results are fed back to the prediction algorithm in the learn phase.
+Finally, the obtained results are fed back to the Gaussian Process Regression prediction algorithm in the learn phase.
 }
@@
 This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-\\
+It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBS with high predicted TIR.\\
 
@@
 \maciej{I think we have decided to put it in the text, but within the description of Fig 2A as it is now. Otherwise the caption would be suuuppppppeeeerrrr long}
-Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
-In each experimental round, we measure the TIR of benchmark RBS as the internal standard. 
-We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\
+Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the RBS groups tested experimentally. 
+In each experimental round, apart from the new RBS designs, we measure the TIR of benchmark RBS as the internal standard. 
+We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR of a new design and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\
 
@@
 To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
-To fit into our automated workflow, we have divided the 450 variants into batches of 90.\\
+To fit into our automated workflow, we have divided the 450 variants into batches of 90 split into 4 design rounds.\\
 
@@
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-The Bandit-0 group performed poorly, despite being machine learning driven, due to being trained on only approximate data.\\
+The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR by the changes made outside the RBS core. 
+The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on only approximate data.\\
 
-However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results improve significantly, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
+However, starting from Round 1, where the prediction and recommendation algorithms were fed data from the Round 0 the results improve significantly, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
 In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
@@
 The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
-Notably, the distributions calculated for all the groups show bivariate distribution, we discuss the possible reasons for that further in the text.\\
+Notably, the distributions calculated for all the groups show bimodal distribution, we discuss the possible reasons for that further in the text.\\
 
@@
 
-Figure \ref{fig: Scatterplot} shows how our implemetation of the Gaussian Process algorithm performed in terms of predictions in each round. 
+Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round. 
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
@@
 We have found our approach bringing machine learning and synthetic biology experts very powerful.
-We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.\\
+We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
+Synthetic biology is promoting standardised and normalised testing in biology which naturally pairs with machine learning which can leverage the high quality biological data sets generated when the correct design rules are followed.
+The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
+On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.\\
+
+Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require a substantial time and experiments to uncover.
+For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR of such sequences.
+This gives hope that the continuing rise in use of machine learning in synthetic biology will enable to uncover new biologically relevant rules of design.\\
 
@@
 For one, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
+We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the obtained results.
 Secondly, given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
 Thirdly, the generalisability of the method is unknown.
-We believe that the method would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
+We believe that the method described here would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
+
 Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.
-In this research we have decided to gradually move the balance from exploration to exploitation over the rounds, but it is not yet clear what is the method protocol for this change.\\
+In this research we have decided to gradually move the balance from exploration to exploitation over the rounds, but it is not yet clear what is the optimal way of conducting this change.\\
 
@@
 This could include, for example, promoters and terminators.
-However, its important to note that the complexity of the task quickly increases with the length of the sequence.
+However, it's important to reiterate that the complexity of the task quickly increases with the length of the sequence.
 This is because the experimental space grows exponentially with the number of examined positions and so the space becomes increasingly hard to cover with experiments.

### 2021-06-10 paper/main.tex
address comments in method sec by iadine

@@
 
-\title{Machine Learning guided workflow\\
-for Ribosome Binding Site engineering}
+\title{Machine Learning guided engineering\\
+of Ribosome Binding Sites}
 
@@
 Unfortunately, RBSs are not understood at the level of finesse required for reliable design. 
-To address this problem, we have created a machine learning (ML) enabled workflow for the design of bacterial RBSs.
+To address this problem, we have created a machine learning (ML) guided engineering approach for the design of bacterial RBSs.
 We used Gaussian Process Regression for prediction and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
-We have integrated the ML algorithms with laboratory automation and high-throughput processes, creating a robust workflow for the design of custom RBSs.
-Using our workflow, we generated a novel library of diverse RBSs with a wide range of expression levels.
-Notably, a high number of these sites demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs.
+We have integrated these ML algorithms with laboratory automation and high-throughput processes for more reliable data generation.
+Our approach has shown that the algorithms were able to identify rules conducive to high protein expression.
+Notably, a high number of designed RBSs demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
+Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for engineering of more complicated genetic devices.
 
@@
 
-\mengyan{Maciej: do we want to put the group description into Fig 2 as we discussed last Fri?}
-\maciej{I think we have decided to put it in the text, but within the description of Fig 2A as it is now. Otherwise the caption would be suuuppppppeeeerrrr long}
 Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the RBS groups tested experimentally. 

### 2021-06-10 paper/supplementary.tex
address comments in method sec by iadine

@@
+%! TEX root=./main.tex
 \textbf{Supplementary}
@@
 
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
-    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
-    \caption{Kernel Heatmap and Predictions.
+% \begin{figure}[!ht]
+%     \centering
+%     \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
+%     \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
+%     \caption{Kernel Heatmap and Predictions.
 Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265).
@@
 The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
-\end{figure}
+% \end{figure}
 

### 2021-06-11 codes/batch_ucb.py
rename folder

@@ -1,276 +1,0 @@
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-import itertools
-from collections import defaultdict
-import math
-import json
-
-from sklearn.gaussian_process import GaussianProcessRegressor
-from sklearn.gaussian_process.kernels import PairwiseKernel, DotProduct, RBF 
-from sklearn.kernel_ridge import KernelRidge
-from sklearn.metrics import r2_score, mean_squared_error, make_scorer
-from sklearn.model_selection import KFold
-from sklearn_extra.cluster import KMedoids
-
-from codes.embedding import Embedding
-from codes.environment import Rewards_env
-from codes.ucb import GPUCB, Random
-from codes.evaluations import evaluate, plot_eva
-from codes.regression import *
-from codes.kernels_for_GPK import *
-import codes.config
-
-from ipywidgets import IntProgress
-
-# Aug 2020 Mengyan Zhang
-# Implement batch ucb, with three methods:
-# 1. Return Top n arm directly (non-batch)
-# 2. Clustering 
-# 3. GP-BUCB
-
-KERNEL_DICT = {
-    # 'Spectrum_Kernel': Spectrum_Kernel,
-    # 'WD_Kernel': WeightedDegree_Kernel,
-    # 'Sum_Spectrum_Kernel': Sum_Spectrum_Kernel,
-    # 'Mixed_Spectrum_Kernel': Mixed_Spectrum_Kernel,
-    'WD_Kernel_Shift': WD_Shift_Kernel   
-}
-
-
-class RBS_UCB():
-    """GPUCB for RBS sequences design.
-
-    Attributes
-    -----------------------------------------------
-    df_known: pandas dataframe
-        known RBS sequences with labels
-        train data
-    df_design: pandas dataframe
-        core part design (4^6) expect those already in df_known
-        test data
-    df_train_test: pandas dataframe
-        concat of df_known and df_design
-
-    kernel_name: string
-        indicates kernel to use 
-    l: int
-        lmer
-    s: int
-        shift number
-    embedding: string
-        label or onehot
-    alpha: float
-        parameter for GPR, value added to kernel diagonal
-    
-    rec_size: int
-        recommendation size
-    beta: float
-        parameter for UCB, mean + beta * std
-        balance exploration and exploitation
-    
-    """
-    def __init__(self, df_known, kernel_name='WD_Kernel_Shift', l=6, s=1, sigma_0 = 1,
[…]

### 2021-06-11 codes/config.py

@@ -1,20 +1,0 @@
-# provide configuration, e.g. global variable, path
-
-BASES = ['A','C','G','T'] # embedding
-DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
-
-# SAVED_IDX_SEQ_PATH = '../../data/idx_seq.pickle'
-# SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
-SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
-
-# Usage:
-# with open(SAVED_IDX_SEQ_PATH, 'rb') as handle:
-#     saved_idx_seq = pickle.load(handle)
-# seq_list = saved_idx_seq['seq_list']
-# idx_list = saved_idx_seq['idx_list']
-# idx_seq_dict = saved_idx_seq['idx_seq_dict']
-
-# SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.pickle'
-# SAVED_KERNEL_PATH = 'data/saved_normalised_kernel.pickle'
-SAVED_KERNEL_PATH = DRIVE_PATH+ 'saved_normalised_kernel.pickle'
-

### 2021-06-11 codes/data_generating.py

@@ -1,309 +1,0 @@
-# Logs:
-
-# 23/04/2021 Update: add step names in comments
-
-# Define the following steps on each replicate:  
-# - A. In each round, substract the mean of every data points by the reference AVERAGE, and then add 100 (to make the values positive).  
-# - B. Take log (base e) transformation for each data points.  
-# - C. Apply z-score normalisation.  
-#     - C.1 on each round, so that the mean and variance of each replicate of data in each round is zero and one after normalisation. 
-#     - C.2 on all data, so that the mean and variance of each replicate of all data is zero and one after normalisation. 
-# - D. Apply min-max normalisation.
-#     - D.1 on each round
-#     - D.2 on all data
-# - E. Apply ratio normalisation. In each round, each data points is devided by the mean of refernce AVERAGE, so that in each round, the reference labels are almost 1. 
-#     - E.1 on each round
-#     - E.2 on all data
-
-# 25/11/2020 Update: add unique index corresponding to 'idx_seq_dict.npz' 
-
-# 23/10/2020 Update: normalisation for each plate each replicate, 
-# substracting mean of reference sequence in each plate
-
-# 04/11/2020 Update: add Rep7,8,9
-#-----------------------------------------------------------------------------------
-
-# This code is used to generate the ready to use data 
-# based on the first round result
-import numpy as np
-import pandas as pd
-import os
-import argparse
-import pickle
-
-# # input parameters
-# parser = argparse.ArgumentParser(description='Generate valid data.')
-# parser.add_argument('normalize_flag', default= 'True', help = 'indicates whether normalize labels. True for normalized label.')
-# parser.add_argument('to_design_round', default= '3', help = 'specify the round to design.')
-
-# args = parser.parse_args()
-# normalize_flag = str(args.normalize_flag) # str True or False
-# to_design_round = str(args.to_design_round) # 0-4
-
-
-# Setting 
-# Specify your choice
-approach = 'd1'
-# to_design_round = '4'
-
-approach = approach.lower()
-
-if 'a' in approach:
-    substract_ref_mean = True
-else:
-    substract_ref_mean = False
-
-if 'b' in approach:
-    Log_flag = True # indicates whether take log label
-else:
-    Log_flag = False
-
-if 'c' in approach:
[…]

### 2021-06-11 codes/embedding.py

@@ -1,102 +1,0 @@
-import numpy as np
-from sklearn.feature_extraction.text import CountVectorizer
-from sklearn.preprocessing import LabelEncoder, normalize
-import codes.config as config # provide configuration, e.g. global variable, path
-
-# 2nd Nov 2019
-# Mengyan Zhang
-
-class Embedding():
-    """Embedding for biological sequences.
-    """
-    def __init__(self, data = None):
-        """
-        Parameters
-        --------------------------------------
-        data: ndarray 
-            (num_seq, ) (each element is a biological sequence)
-        """
-        self.data = data
-        # self.num_seq, self.num_bases = data.shape
-        self.num_seq,  = data.shape
-        self.bases = config.BASES
-
-    def label(self):
-        """Label encoder from sklearn.
-        Encode each row from string into numeric label.
-        For example, from 'GACAAA' to [0 1 2 0 0 0]
-
-        Note when use inverse_transform (in kernel), the order
-        of bases (fitting) should be the same.
-
-        Returns
-        ----------------------------------------------------
-        embedded_data: array
-            num_seq * num_bases
-        """
-        le = LabelEncoder()
-        le.fit(self.bases)
-        # if data is None: # if no input data, return the encoder
-        #     return le
-        # else:
-        embedded_data = np.array([le.transform(list(self.data[i]))\
-                                    for i in range(self.num_seq)])
-        return embedded_data
-
-    def onehot(self):
-        """One-hot embedding.
-
-        Returns
-        --------------------------------------------
-        embedded_data: ndarray
-            {0, 1}^{num_seq x num_bases * 4}
-        """
-        
-        base_dict = dict(zip(self.bases,range(4))) # {'A' : 0, 'C' : 1, 'G' : 2, 'T' : 3}
-        self.num_bases = len(self.data[0])
-        embedded_data = np.zeros((self.num_seq, self.num_bases * 4))
-
-        # loop through the array of sequences to create a feature array 
-        for i in range(self.num_seq):
-            seq = self.data[i]
-            # loop through each individual sequence, from the 5' to 3' end
-            for b in range(self.num_bases):
-                embedded_data[i, b * 4 + base_dict[seq[b]]] = 1
-
-        embedded_data -= np.nanmean(embedded_data, axis = 0)
-        normalised_embedded_data = normalize(embedded_data, norm = 'l2')
-        return normalised_embedded_data
-
-    def kmer(self, size = 3):
[…]

### 2021-06-11 codes/environment.py

@@ -1,117 +1,0 @@
-import numpy as np
-from collections import defaultdict, OrderedDict
-import operator
-from codes.embedding import Embedding
-
-class Rewards_env():
-    """Rewards environment for biology sequence.
-
-    Attributes
-    -------------------------------------------------
-    rewards_dict: dict of list
-        keys: string of embedded sequence 
-            e.g. '001000100010001000100010'
-        values: list of available labels
-    labels_dict: dict of list
-        keys: string of embedded sequence 
-            e.g. '001000100010001000100010'
-        values: label (expectation of rewards list)
-    """
-    def __init__(self, data, embedding_method = 'onehot'):
-        """
-        Parameters
-        ---------------------------------------------
-        data: ndarray 
-            num_data * 2
-            two columns: biology sequence; score (label)
-        embedding_method: string
-            specify embedding method
-            default is 'onehot'
-        """
-        self.data = data
-        self.num_seq, self.num_col = self.data.shape
-        # assert self.num_col == 2
-        self.embedding_method = embedding_method
-
-        self.embedded = self.arm_embedding()
-
-        # TODO: comment for now, may need to put it back
-        # self.rewards_dict = defaultdict(list)
-        # self.create_rewards_dict()
-        # self.labels_dict = self.labels_generator()
-        # self.arm_features = list(self.labels_dict.keys())
-        
-    def arm_embedding(self):
-        """Embedding biological sequence.
-        """
-        # encoder = Embedding(self.data[:, 0].reshape(self.num_seq, 1))
-        encoder = Embedding(self.data[:, 0])
-        if self.embedding_method == 'label':
-            embedded= encoder.label()
-        elif self.embedding_method == 'onehot':
-            embedded = encoder.onehot()
-        elif self.embedding_method == 'kmer':
-            embedded = encoder.kmer()
-        else:
-            print('To be added.')
-            embedded = self.data[:, 0]
-        return embedded
-
-    def create_rewards_dict(self):
-        """Create reward dict based on the embedded features.
-        """
-        for i, d in enumerate(self.data):
-            key = self.to_string(self.embedded[i])
-            self.rewards_dict[key].append(d[1])
-
-        # remove all sort
-        # sort the rewards dict by keys
-        # self.rewards_dict = dict(sorted(self.rewards_dict.items()))
-        self.rewards_dict = dict(self.rewards_dict.items())
-
-    def labels_generator(self):
-        """Generate labels for each arm.
-
-        Return 
-        -----------------------------------------
[…]

### 2021-06-11 codes/evaluations.py

@@ -1,72 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-from collections import defaultdict
-
-ylabel_dict = {'sd': 'suboptimal draws',
-                'r': 'cumulative regrets',
-            }
-
-def evaluate(sds, rs):
-    """Calculated expected evaluation metrics (suboptimal draws, regret)
-
-    Parameters
-    ----------------------------------------------------
-    sds: list 
-        len: num_exper; list of sub-optimal draws for each experiment
-    rs: list
-        len: num_exper; list of cumulative regrets for each experiment
-
-    Returns
-    ----------------------------------------------------
-    eva_dict: dict
-        key: name of evaluation metrics
-            one of 'sd' and 'r'
-        value: array with shape 1 * num_rounds
-    """
-    eva_dict =  defaultdict(dict)
-    for key, values in sds.items():
-        eva_dict[key]['sd'] = defaultdict(list)
-        eva_dict[key]['sd'] = np.mean(np.asarray(values), axis = 0)
-    for key, values in rs.items():
-        eva_dict[key]['r'] = defaultdict(list)
-        eva_dict[key]['r'] = np.mean(np.asarray(values), axis = 0)
-    return eva_dict
-
-def plot_eva(results, eva_method, paper_flag = True):
-    """Plot method for evaluations
-
-    Parameters
-    -----------------------------------------------------------
-    results: dict
-        keys: 'env name + num_exper + num_rounds'
-        values: dict
-            keys: 'est_var + hyperpara' or 'bound'
-            values: dict
-                keys: 'sd', 'r'
-                values: list of result  
-    eva_method: str
-        options ('sd', 'r')
-    paper_flag: boolean
-        indicates whether plotting for paper.
-        True is for paper; False is not.
-    """
-    plt.figure(figsize=(5, 5))
-       
-    # setup title
-    title_name = 'Cumulative Regret vs. Round'
-    plt.title(title_name)
-
-    plt.xlabel('Iteration')
-    plt.ylabel('Expected ' + ylabel_dict[eva_method])
-    
-    for key, value in results.items():
-        plt.plot(range(len(value[eva_method])), 
-                value[eva_method],  
-                marker = 'o', 
-                markevery = 10,
-                markersize = 5,
-                label = key)
-
-    plt.legend(loc="upper left")
-    #file_name = 'Exper_' + str(eva_method) + '_' + name + '_' + subname + '.pdf'
-    #plt.savefig(file_name, bbox_inches='tight')

### 2021-06-11 codes/generate_idx_seq.py

@@ -1,101 +1,0 @@
-# Nov.2020 Mengyan Zhang 
-# This file generates a two-way dict (index, sequences) for all design RBS sequences,
-# including 4^6 (core-part) + 3*14 (bps non-core group) = 4138 sequences. 
-
-# usage:
-# with open(Path, 'rb') as handle:
-#    idx_seq = pickle.load(handle)
-
-# idx_seq has three keys
-# dict_keys(['idx_seq_dict', 'idx_list', 'seq_list'])
-
-# idx_list and seq_list gives you the idx/seq in the same order
-# idx_seq_dict is a two way dict, 
-# you can do idx_seq_dict[20-long-RBS] = idx or idx_seq_dict[idx] = 20-long-RBS
-
-import os
-import sys
-module_path = os.path.abspath(os.path.join('../..'))
-if module_path not in sys.path:
-    sys.path.append(module_path)
-import pandas as pd
-import numpy as np
-import itertools
-import pickle
-
-# class TwoWayDict(dict):
-#     """bijective map between index and sequences.
-#     code from Sasha Chedygov's answer https://stackoverflow.com/questions/1456373/two-way-reverse-map
-#     """
-#     def __setitem__(self, key, value):
-#         # Remove any previous connections with these values
-#         if key in self:
-#             del self[key]
-#         if value in self:
-#             del self[value]
-#         dict.__setitem__(self, key, value)
-#         dict.__setitem__(self, value, key)
-
-#     def __delitem__(self, key):
-#         dict.__delitem__(self, self[key])
-#         dict.__delitem__(self, key)
-
-#     def __len__(self):
-#         """Returns the number of connections"""
-#         return dict.__len__(self) // 2 
-
-
-# Setting
-char_sets = ['A','C','G','T']
-design_len = 6
-pre_design = 'TTTAAGA'
-pos_design = 'TATACAT'
-index = 0
-index_list = []
-seq_list = []
-idx_seq_dict = {}
-
-for combo in itertools.product(char_sets, repeat= design_len):
-    combo = pre_design + ''.join(combo) + pos_design
-    index_list.append(index)
-    seq_list.append(combo)
-    idx_seq_dict[index] = combo
[…]

### 2021-06-11 codes/kernels_for_GPK.py

@@ -1,598 +1,0 @@
-import numpy as np
-import pandas as pd
-from sklearn.feature_extraction.text import CountVectorizer
-from sklearn import preprocessing
-from sklearn.gaussian_process.kernels import Kernel, Hyperparameter
-import matplotlib.pyplot as plt
-import os
-import sys
-module_path = os.path.abspath(os.path.join(''))
-print (module_path)
-print(sys.path)
-if module_path not in sys.path:
-    sys.path.append(module_path)
-from codes.embedding import Embedding
-from codes.environment import Rewards_env
-
-
-import pickle
-import codes.config as config # provide configuration, e.g. global variable, path
-
-# June 2021 Mengyan Zhang 
-# String Kernel Classes, taking strings as input
-# including spectrum kernel, weighted degree kernel and weighted degree kernel with shift 
-# The implementation is based on Support Vector Machines and Kernels for Computational Biology 
-# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2547983/
-# Inherit from sklearn gaussain process kernel class
-# https://github.com/scikit-learn/scikit-learn/blob/b194674c4/sklearn/gaussian_process/kernels.py#L1213
-
-# Two choices of kernel normalisation target:
-# 1. normalisation over the all design space. The kernel matrix is pre-calculated and saved. 
-# 2. normalisation over all currently known RBS sequences. 
-# For Gaussian Process Regression, the specific kernel matrix is sliced from the calculated kernel (in __call__)
-
-
-class String_Kernel(Kernel):
-    """Base class for string kernels
-    
-    Parameters
-    ------------------
-    * For initialisation, calculating the whole kernel
-
-    features: array
-        list of features for all training and testing data
-    n_train: int
-        number of training data
-    n_test: int 
-        number of testing data
-    kernel_all: (n_train+n_test, n_train+n_test)
-        kernel matrix for all data
-    kernel_all_normalised: (n_train+n_test, n_train+n_test)
-        kernel matrix for all data with centering and unit norm
-
-    * Kernel hyperparameters *
-    l : int, default 3
-        number of l-mers (length of 'word')
-    # TODO: change hyperparameters maybe (now set to the same as Dotproduct)
-    sigma_0 : float >= 0, default: 0.0
-        signal std
-        Parameter controlling the inhomogenity of the kernel. If sigma_0=0,
-        the kernel is homogenous.
-    """
-    
-    INIT_FLAG = False
[…]

### 2021-06-11 codes/regression.py

@@ -1,670 +1,0 @@
-# update: 25/Nov/2020
-# use unique seq for kernel normalisation
-
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-from matplotlib import gridspec
-import plotly
-import plotly.graph_objs as go
-import seaborn as sns
-import itertools
-from collections import defaultdict
-import math
-import json
-import xarray as xr  
-
-from sklearn.gaussian_process import GaussianProcessRegressor
-from sklearn.gaussian_process.kernels import PairwiseKernel, DotProduct, RBF, WhiteKernel
-from sklearn.kernel_ridge import KernelRidge
-from sklearn.metrics import r2_score, mean_squared_error, make_scorer
-from sklearn.model_selection import KFold
-from sklearn.cluster import KMeans, AgglomerativeClustering
-from scipy.stats import spearmanr
-
-from codes.embedding import Embedding
-from codes.environment import Rewards_env
-from codes.ucb import GPUCB, Random
-from codes.evaluations import evaluate, plot_eva
-from codes.kernels_for_GPK import *
-from codes.sort_seq import sort_kernel_matrix
-from codes.plot_format import *
-
-from ipywidgets import IntProgress
-import codes.config # provide configuration, e.g. global variable, path
-
-# Aug 2020 Mengyan Zhang
-# Implement predictors based on Gaussian Process Regression
-
-KERNEL_DICT = {
-    # 'Spectrum_Kernel': Spectrum_Kernel,
-    # 'Mixed_Spectrum_Kernel': Mixed_Spectrum_Kernel,
-    # 'WD_Kernel': WeightedDegree_Kernel,
-    # 'Sum_Spectrum_Kernel': Sum_Spectrum_Kernel,
-    'WD_Kernel_Shift': WD_Shift_Kernel,
-    'RBF': RBF  
-}
-
-REP_LIST = ['Rep1', 'Rep2', 'Rep3', 'Rep4', 'Rep5','Rep6']
-
-
-class GPR_Predictor():
-    def __init__(self, df, test_size=0.2, train_idx = None, test_idx = None, 
-                 kernel_name='WD_Kernel_Shift', l=6, s = 1, b=0.33, sigma_0 = 1, padding_flag=False, gap_flag=False,
-                 alpha=2, embedding='label', eva_metric=[r2_score, mean_squared_error],  eva_on = "samples",
-                 kernel_norm_flag = True, centering_flag = True, unit_norm_flag = True,
[…]

### 2021-06-11 paper/methods.tex

--- 2021-06-08 paper/methods.tex
@@
 The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each batch based on the predictions from LEARN.
+\\
+
 % 1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
@@
 Specifically, in round $t$, we subtract the TIR mean of the benchmark RBS measured in round $t$ from all TIR values measured in the same round, for each replicate separately. 
-We then normalised the data by taking a logarithm transformation and standardisation of the adjusted TIR label for each replicate separately.
+We then normalised the data by performing a logarithm transformation and standardisation on the adjusted TIR label for each replicate separately.
 After normalisation, each replicate has zero mean and unit variance. 
@@
 
-To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points, and $f(\mathbf{x})$ is the TIR score at $\mathbf{x}$.
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ is the TIR score at 20-base RBS sequence $\mathbf{x} \in \{A,C,G,T\}^{20}$.
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
+$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
+\\
 % This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g. time, temperature, operator, etc.). 
@@
         = \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
+        \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + \mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
 \end{align}
@@
 To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$.
-On one hand, we want to exploit the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
-on the other hand, we also want to explore the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
-More precisely, UCB algorithm selects RBS sequences with the maximum upper confidence bound at round $t$, i.e.
+On one hand, UCB exploits the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
+on the other hand, UCB explores the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
+More precisely, UCB algorithm selects RBS sequence $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at round $t$, i.e.
 \begin{align}
@@
 $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
-We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t-1$.\\
 
@@
 However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
-However, since GPR assumes similar sequences would have similar labels (e.g.
+Since GPR assumes similar sequences would have similar labels (e.g.
 by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
-Batch recommendation strategies are designed to avoid recommendation such high-similarity sequences in the same batch.
+To counter this, we use a batch recommendation strategy that is designed to avoid recommending such highly similar sequences in the same batch, described below. \\
 
@@
 
-A key property of Gaussian Process regression is that the predicted standard deviation depends only on observed points (i.e. features), but not on the labels of those observed points. 
+A key property of Gaussian Process regression is that the predicted standard deviation depends only on observed points (i.e. features), not on the labels of those observed points. 
 % We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
@@
 One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
-The strategy is to recommend sequences sequentially by updating the UCB score with the updated predicted standard deviation with the previously recommended points.
+The strategy here is to recommend sequences sequentially by updating the UCB score with the predicted standard deviation of the previously recommended points.\\
 % Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
 % The plot shows the predicted mean and standard deviation after observing 4 data points. 
-As illustrated in Figure \ref{fig:batch rec} (B), the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
+As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
 Then we add the recommended data point ($x = 2$) into the training dataset with the predicted mean of that point as label (note it is not the true label, i.e.
@@
 The second data point is then recommended based on the new UCB scores.   
-As we can see in Figure \ref{fig:batch rec}(C), since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
+Figure \ref{fig:batch rec}C shows that since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
+\mengyan{need clarification (commented by iadine)}
 In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 

### 2021-06-11 synbio_rbs/example/Round0/codes/embedding.py
reproducible round 0

@@ -1,0 +1,98 @@
+import numpy as np
+from sklearn.feature_extraction.text import CountVectorizer
+from sklearn.preprocessing import LabelEncoder
+
+# 2nd Nov 2019
+# Mengyan Zhang
+
+class Embedding():
+    """Embedding for biological sequences.
+    """
+    def __init__(self, data):
+        """
+        Parameters
+        --------------------------------------
+        data: ndarray 
+            (num_seq, ) (each element is a biological sequence)
+        """
+        self.data = data
+        # self.num_seq, self.num_bases = data.shape
+        self.num_seq,  = data.shape
+        self.num_bases = len(data[0])
+        self.bases = ['A','C','G','T']
+
+    def label(self):
+        """Label encoder from sklearn.
+        Encode each row from string into numeric label.
+        For example, from 'GACAAA' to [0 1 2 0 0 0]
+
+        Note when use inverse_transform (in kernel), the order
+        of bases (fitting) should be the same.
+
+        Returns
+        ----------------------------------------------------
+        embedded_data: array
+            num_seq * num_bases
+        """
+        le = LabelEncoder()
+        le.fit(self.bases)
+        embedded_data = np.array([le.transform(list(self.data[i]))\
+                                     for i in range(self.num_seq)])
+        return embedded_data
+
+
+
+    def onehot(self):
+        """One-hot embedding.
+
+        Returns
+        --------------------------------------------
+        embedded_data: ndarray
+            {0, 1}^{num_seq x num_bases * 4}
+        """
+        
+        base_dict = dict(zip(self.bases,range(4))) # {'A' : 0, 'C' : 1, 'G' : 2, 'T' : 3}
+
+        embedded_data = np.zeros((self.num_seq, self.num_bases * 4))
+
+        # loop through the array of sequences to create a feature array 
+        for i in range(self.num_seq):
+            seq = self.data[i]
+            # loop through each individual sequence, from the 5' to 3' end
+            for b in range(self.num_bases):
+                embedded_data[i, b * 4 + base_dict[seq[b]]] = 1
+        return embedded_data
+
+    def kmer(self, size = 3):
+        """k-merization embedding.
+           See https://en.wikipedia.org/wiki/K-mer.
+           Tutorial (https://www.kaggle.com/thomasnelson/working-with-dna-sequence-data-for-ml#Machine-learning-with-DNA-sequence-data
+
+        Parameters
+        --------------------------------------------------------------------
+        size: int 
[…]

### 2021-06-11 synbio_rbs/example/Round0/codes/environment.py

@@ -1,0 +1,121 @@
+import numpy as np
+from collections import defaultdict, OrderedDict
+import operator
+from codes.embedding import Embedding
+
+class Rewards_env():
+    """Rewards environment for biology sequence.
+
+    Attributes
+    -------------------------------------------------
+    rewards_dict: dict of list
+        keys: string of embedded sequence 
+            e.g. '001000100010001000100010'
+        values: list of available labels
+    labels_dict: dict of list
+        keys: string of embedded sequence 
+            e.g. '001000100010001000100010'
+        values: label (expectation of rewards list)
+    """
+    def __init__(self, data, embedding_method = 'onehot'):
+        """
+        Parameters
+        ---------------------------------------------
+        data: ndarray 
+            num_data * 2
+            two columns: biology sequence; score (label)
+        embedding_method: string
+            specify embedding method
+            default is 'onehot'
+        """
+        self.data = data
+        self.num_seq, self.num_col = self.data.shape
+        assert self.num_col == 2
+        self.embedding_method = embedding_method
+
+        self.embedded = self.arm_embedding()
+
+        self.rewards_dict = defaultdict(list)
+        self.create_rewards_dict()
+        self.labels_dict = self.labels_generator()
+        self.arm_features = self.sort_arms()
+        
+    def sort_arms(self):
+        # remove all sort
+        #self.labels_dict = OrderedDict(sorted(self.labels_dict.items(), key=operator.itemgetter(1)))
+        return list(self.labels_dict.keys())
+        
+    def arm_embedding(self):
+        """Embedding biological sequence.
+        """
+        # encoder = Embedding(self.data[:, 0].reshape(self.num_seq, 1))
+        encoder = Embedding(self.data[:, 0])
+        if self.embedding_method == 'label':
+            embedded= encoder.label()
+        elif self.embedding_method == 'onehot':
+            embedded = encoder.onehot()
+        elif self.embedding_method == 'kmer':
+            embedded = encoder.kmer()
+        else:
+            print('To be added.')
+            embedded = self.data[:, 0]
+        return embedded
+
+    def create_rewards_dict(self):
+        """Create reward dict based on the embedded features.
+        """
+        for i, d in enumerate(self.data):
+            key = self.to_string(self.embedded[i])
+            self.rewards_dict[key].append(d[1])
+
+        # remove all sort
+        # sort the rewards dict by keys
+        # self.rewards_dict = dict(sorted(self.rewards_dict.items()))
+        self.rewards_dict = dict(self.rewards_dict.items())
+
[…]

### 2021-06-11 synbio_rbs/example/Round0/codes/kernels.py

@@ -1,0 +1,254 @@
+import numpy as np
+from sklearn.feature_extraction.text import CountVectorizer
+from sklearn.metrics.pairwise import check_pairwise_arrays
+from sklearn import preprocessing
+
+def Phi(X, Y, l, j_X, j_Y, d):
+    """Calculate spectrum features for spectrum kernel.
+
+    Phi is a mapping of the matrix X into a |alphabet|^l
+    dimensional feature space. For each sequence in X,
+    each dimension corresponds to one of the |alphabet|^l 
+    possible strings s of length l and is the count of 
+    the number of occurrance of s in x. 
+
+    Paramters
+    ---------------------------------------------------
+    X : array of shape (n_samples_X, )
+        each row is a sequence (string)
+    Y : array of shape (n_samples_Y, )
+        each row is a sequence (string)
+    l : int, default 3
+        number of l-mers (length of 'word') 
+    j_X : int
+        start position of sequence in X
+    j_Y : int
+        start position of sequence in Y
+    d : int
+        the length of analysed sequence
+        j + d is end position of sequence 
+
+    Returns
+    ----------------------------------------------------
+    embedded matrix of X: array
+        num_X * num_embedded_features
+    embedded matrix of Y
+        num_Y * num_embedded_features
+    """
+    num_X = X.shape[0]
+    num_Y = Y.shape[0]
+
+    sentences = []
+
+    for i in range(num_X):
+        sequence= X[i][j_X:j_X + d]
+        words = [sequence[a:a+l] for a in range(len(sequence) - l + 1)]
+        sentence = ' '.join(words)
+        sentences.append(sentence)
+
+    for i in range(num_Y):
+        sequence= Y[i][j_Y:j_Y + d]
+        words = [sequence[a:a+l] for a in range(len(sequence) - l + 1)]
+        sentence = ' '.join(words)
+        sentences.append(sentence)
+    #print(sentences)
+    cv = CountVectorizer(analyzer='word',token_pattern=u"(?u)\\b\\w+\\b")
+    #cv =  CountVectorizer()
+    embedded = cv.fit_transform(sentences).toarray()
+    #print(embedded)
+    #print(cv.get_feature_names())
+    normalised_embedded = preprocessing.normalize(embedded, norm = 'l2')
[…]

### 2021-06-11 synbio_rbs/example/Round0/codes/kernels_pairwise.py

@@ -1,0 +1,443 @@
+import numpy as np
+from sklearn.feature_extraction.text import CountVectorizer
+from sklearn.preprocessing import LabelEncoder, OneHotEncoder, normalize
+from codes.embedding import Embedding
+
+def phi(x, y, l, j_x, j_y, d):
+    """Calculate spectrum features for spectrum kernel.
+
+    phi is a mapping of a row of matrix x into a |alphabet|^l
+    dimensional feature space. For each sequence in x,
+    each dimension corresponds to one of the |alphabet|^l 
+    possible strings s of length l and is the count of 
+    the number of occurrance of s in x. 
+
+    Paramters
+    ---------------------------------------------------
+    x : string
+        a row of the data matrix
+    y : string
+        a row of the data matrix
+    l : int, default 3
+        number of l-mers (length of 'word') 
+    j_x : int
+        start position of sequence in x
+    j_y : int
+        start position of sequence in y
+    d : int
+        the length of analysed sequence
+        j + d is end position of sequence 
+
+    Returns
+    ----------------------------------------------------
+    embedded_x: array
+        1 * num_embedded_features
+    embedded_y: array
+        1 * num_embedded_features
+    """
+
+    sentences = []
+
+    sequence_x= x[j_x:j_x + d]
+    words_x = [sequence_x[a:a+l] for a in range(len(sequence_x) - l + 1)]
+    sentence_x = ' '.join(words_x)
+    sentences.append(sentence_x)
+
+    sequence_y= y[j_y:j_y + d]
+    words_y = [sequence_y[a:a+l] for a in range(len(sequence_y) - l + 1)]
+    sentence_y = ' '.join(words_y)
+    sentences.append(sentence_y)
+    cv = CountVectorizer(analyzer='word',token_pattern=u"(?u)\\b\\w+\\b")
+    #cv = CountVectorizer()
+    embedded = cv.fit_transform(sentences).toarray()
+    #print(embedded)
+    normalised_embedded = normalize(embedded, norm = 'l2')
+
+    return embedded[0], embedded[1]
+
+def inverse_label(x):
+    """convert_to_string
+    """
+    le = LabelEncoder()
+    bases = ['A','C','G','T']
+    le.fit(bases)
+    
+    int_x = []
[…]

### 2021-06-11 synbio_rbs/example/Round0/codes/regression.py

@@ -1,0 +1,149 @@
+import numpy as np
+from collections import defaultdict, OrderedDict
+import operator
+import matplotlib.pyplot as plt
+
+from sklearn.linear_model import LinearRegression
+from sklearn.kernel_ridge import KernelRidge
+from sklearn.gaussian_process import GaussianProcessRegressor
+from sklearn.gaussian_process.kernels import DotProduct
+from sklearn.gaussian_process.kernels import RBF
+from sklearn.model_selection import train_test_split, cross_val_score, ShuffleSplit
+from sklearn.metrics import r2_score, mean_squared_error
+
+from codes.environment import Rewards_env
+from codes.kernels import spectrum_kernel, mixed_spectrum_kernel, WD_kernel, WD_shift_kernel
+
+KERNEL_TYPE = {'spectrum': spectrum_kernel,
+               'mixed_spectrum': mixed_spectrum_kernel,
+               'WD': WD_kernel,
+               'WD_shift': WD_shift_kernel
+            }
+
+class Regression():
+    """Regression.
+
+    Attributes
+    --------------------------------------------------------
+    model: instance of regression class (from sklearn)
+            attribute: kernel (if 'precomputed', use precomputed kernel matrix)
+    X: array
+        features array (num_samples, ) 
+        first column of data, each element is a string
+    Y: array
+        labels array (num_samples, )
+        second column of data, each element is a float/int
+    """
+    def __init__(self, model, data, embedding_method = None, 
+                 precomputed_kernel = None):
+        """
+        Paramters:
+        ------------------------------------------------------
+        model: instance of regression class (from sklearn)
+            attribute: kernel (if 'precomputed', use precomputed kernel matrix)
+        data: ndarray 
+            num_data * 2
+            two columns: biology sequence; score (label)
+        embedding_method: string, default is None
+            if None, no embedding is performed and set X to the first column of data
+        precomputed_kernel: string, default is None
+            must be the key of KERNEL_TYPE dict
+        """
+        self.model = model
+        
+        if embedding_method is not None:
+            self.my_env = Rewards_env(data, embedding_method)
+            self.X = self.my_env.embedded
+        else:
+            self.X = data[:, 0]
+        self.Y = data[:, 1]
+        self.split_data()
+
[…]

### 2021-06-11 synbio_rbs/example/Round0/codes/ucb.py

@@ -1,0 +1,370 @@
+from abc import ABC, abstractmethod
+import numpy as np
+from collections import defaultdict
+
+from sklearn.gaussian_process import GaussianProcessRegressor
+from sklearn.gaussian_process.kernels import DotProduct
+
+import matplotlib.pyplot as plt
+
+from codes.environment import Rewards_env
+
+class Bandits_discrete(ABC):
+    """Base class for bandit algorithms of a finite number of arms in discrete space.
+
+    Arguments
+    -----------------------------------------------------------------------------
+    env: instance of Rewards_env
+        Attributes: rewards_dict, labels_dict, embedded, arm_features
+    num_rounds: int
+        total number of rounds. 
+    num_arms: int
+        number of arms
+    num_init: int 
+        number of initialization set 
+    init_list: list
+        list of idx of init arms
+
+    rewards_dict: dict of list
+        rewards environment
+        keys: string of embedded sequence 
+            e.g. '001000100010001000100010'
+        values: list of available labels
+    labels_dict: dict of list
+        label environment
+        keys: string of embedded sequence 
+            e.g. '001000100010001000100010'
+        values: label (expectation of rewards list)
+        sorted according to the values 
+    embedded: ndarray
+        num_arms x num_features
+    arm_features: list
+        list of arm features (string, e.g. '001000100010001000100010')
+        same order as the labels dict
+
+    bestarm_idx: int
+        the arm with maximum label (assume only one arm is the best)
+        TODO: consider the case for multiple best arms
+    
+    sample_features: list
+        list of embedding features for selected arms
+    sample_labels: list
+        list of labels for selected arms
+    sample_idx: list
+        list of idx for selected arms
+
+    ---------------For Evaluation-----------------------------------------  
+    sd: int
+        suboptimal draws
+    r: float
+        cumulative regret
+    suboptimalDraws: list
+        sum of draws of each sub-optimal arm 
+    cumulativeRegrets: list
+        Depends on the regret definition of each algorithm
+        express the difference (of rewards) between drawing optimal arm 
+        all the time and drawing arms based on the designed policy.  
+
+    """
+    def __init__(self, env, num_rounds, init_list=None, init_per=0.2, 
+                 num_rec = 1, model = None, arm_features = None):
+        """
+        Parameters
[…]

### 2021-06-11 synbio_rbs/src/config.py
create saved kernel matrix and update readme

@@ -1,10 +1,7 @@
 # provide configuration, e.g. global variable, path
 
 BASES = ['A','C','G','T'] # embedding
-DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
 
-# SAVED_IDX_SEQ_PATH = '../../data/idx_seq.pickle'
-# SAVED_IDX_SEQ_PATH = 'data/idx_seq.pickle'
 SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
 
 # Usage:
@@ -14,7 +11,13 @@ SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
 # idx_list = saved_idx_seq['idx_list']
 # idx_seq_dict = saved_idx_seq['idx_seq_dict']
 
-# SAVED_KERNEL_PATH = '../../data/saved_normalised_kernel.pickle'
-# SAVED_KERNEL_PATH = 'data/saved_normalised_kernel.pickle'
-SAVED_KERNEL_PATH = DRIVE_PATH+ 'saved_normalised_kernel.pickle'
+SAVED_KERNEL_PATH = '../data/saved_kernel/'
+
+# ---------------------------------------------------------------------
+
+# DRIVE_PATH = '/localdata/u6015325/SynbioML_drive/'
+# SAVED_KERNEL_PATH = DRIVE_PATH+ 'saved_normalised_kernel.pickle'
+
+# with open('../data/saved_normalised_kernel.pickle', 'rb') as handle:
+#     kernel = pickle.load(handle)
 

### 2021-06-11 synbio_rbs/src/kernels_for_GPK.py

@@ -31,6 +31,15 @@ import src.config as config # provide configuration, e.g. global variable, path
 # 2. normalisation over all currently known RBS sequences. 
 # For Gaussian Process Regression, the specific kernel matrix is sliced from the calculated kernel (in __call__)
 
+def generate_triu(kernel_matrix):
+    return kernel_matrix[np.triu_indices(kernel_matrix.shape[0], k = 0)]
+
+def putback_kernel_matrix(triu, size = 4138):
+    X = np.zeros((size,size))
+    X[np.triu_indices(X.shape[0], k = 0)] = triu
+    X = X + X.T - np.diag(np.diag(X))
+    return X
+
 
 class String_Kernel(Kernel):
     """Base class for string kernels
@@ -92,17 +101,24 @@ class String_Kernel(Kernel):
         #----------------------------------------------------------------------------------------
         # reconstructure: use all 4138 unique seqs for kernel normalisation 
         if self.kernel_over_all_flag:
-            if os.path.exists(config.SAVED_KERNEL_PATH):
-                with open(config.SAVED_KERNEL_PATH, 'rb') as handle:
-                    SAVED_KERNELS = pickle.load(handle)
-                    # print('available kernels: ', list(SAVED_KERNELS.keys()))
-            else:
-                print('No saved kernel dictionary found. Create one...')
-                SAVED_KERNELS = {} # key: kernel name (with parameters); value: kernel matrix
-            print(self.kernel_name_para)
-            if self.kernel_name_para in list(SAVED_KERNELS.keys()):
+            # if os.path.exists(config.SAVED_KERNEL_PATH):
+            #     with open(config.SAVED_KERNEL_PATH, 'rb') as handle:
+            #         SAVED_KERNELS = pickle.load(handle)
+            #         # print('available kernels: ', list(SAVED_KERNELS.keys()))
+            # else:
+            #     print('No saved kernel dictionary found. Create one...')
+            #     SAVED_KERNELS = {} # key: kernel name (with parameters); value: kernel matrix
+            # print(self.kernel_name_para)
+
+            kernel_path = config.SAVED_KERNEL_PATH + self.kernel_name_para + '.pickle'
+            # SAVED_KERNEL_PATH = '../data/saved_kernel/' is specified in config.py
+
+            if os.path.exists(kernel_path):
                 print('Load saved kernel matrix...')
-                self.kernel_all_normalised = SAVED_KERNELS[self.kernel_name_para]
[…]

### 2021-06-11 synbio_rbs/src/config.py
rename examples to demo

@@ -20,4 +20,11 @@ SAVED_KERNEL_PATH = '../data/saved_kernel/'
 
 # with open('../data/saved_normalised_kernel.pickle', 'rb') as handle:
 #     kernel = pickle.load(handle)
+    
+# new_kernel = {}
+
+# for key in kernel.keys():
+    
+#     with open('../data/' + key + '.pickle', 'wb') as handle:
+#         pickle.dump(generate_triu(kernel[key]), handle, protocol=pickle.HIGHEST_PROTOCOL)
 

### 2021-06-15 paper/main.tex
Updates from Overleaf

@@
 \usepackage[utf8]{inputenc}
-\usepackage[maxcitenames=1,style=numeric]{biblatex}
+\usepackage[maxcitenames=1,style=numeric, sorting = none]{biblatex}
 \usepackage{amsmath}
@@
 
-\title{Machine Learning guided engineering\\
+\title{Machine Learning guided Design\\
 of Ribosome Binding Sites}
@@
 
-Fine control of gene expression can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
-Unfortunately, RBSs are not understood at the level of finesse required for reliable design. 
-To address this problem, we have created a machine learning (ML) guided engineering approach for the design of bacterial RBSs.
-We used Gaussian Process Regression for prediction and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
+% Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
+Ribosome binding site (RBS), as one of the engineering transcription and translation control elements, allows fine control of gene expression.
+% Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
+We consider the problem of designing RBSs in nucleotides level to optimise the gene expression with fixed budget. 
+To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
+We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
 We have integrated these ML algorithms with laboratory automation and high-throughput processes for more reliable data generation.
 Our approach has shown that the algorithms were able to identify rules conducive to high protein expression.
-Notably, a high number of designed RBSs demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
-Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for engineering of more complicated genetic devices.
+Notably, by testing only 450 RBS variants, we have discovered 
+% designed and experimentally-validated 
+RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
+Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for the experimental design of more complicated genetic devices.
 
@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
-This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
-Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
-Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
-However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\
+This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
+This cycle involves designing the DNA sequence in computer-aided design (CAD) software and then physically testing it in a laboratory. 
+Additionally, computer modelling can be used to predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
+Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
+However, de-novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.\\
 
-The biggest gap in the DTBL cycle, at present, is at the Learn and Design interface - it is hard to translate obtained results into new designs.
+The biggest gap in the DTBL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
 For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
-but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}.
-There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}.
+but they come with a number of caveats: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which increases perceived prediction error \cite{Goss1998}.
+There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}.
 This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\
 
-Synthetic biology is currently going through a phase of exponential increase in volume of data produced during experiments \cite{Freemont2019}. 
-These new data sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
-For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.
+Currently, the volume of experimental data in synthetic biology is increasing exponentially. \cite{Freemont2019}. 
+Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
+For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoids in \emph{Esherichia coli} \cite{Jervis2019}.
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
-There were also successful attempts at using deep learning techniques for analysis of big synthetic biology data sets \cite{Alipanahi2015,Angermueller2016}.\\
+Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.\\
 
-In this work, we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of bacterial RBS with the goal of optimising associated protein expression level.
-RBS being one of the key genetic elements controlling protein expression and at the same time having a relatively short sequence is a perfect target for establishing workflows that can be later translated to more complicated systems.
-In this work we have used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS .
-Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.
-This way, we were able to build an extensive, reliable library of novel RBSs with diverse sequences.
-At the same time we were able to discover new RBS sequences with very high TIRs from between 95 to 135\% of TIR of our chosen very strong benchmark RBS.
+In this work, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level.
+We used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
+As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that can be later translated to more complicated systems.
+Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
+This way, we have built an extensive and reliable library of novel RBSs with diverse sequences.
+We discovered new RBS sequences with very high TIRs, between 95 to 135\% of our very strong benchmark RBS.
 
@@
 
-We present our RBS optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
-Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
-in the DESIGN stage (Section~\ref{sec:ucb-results}),
-and we demonstrate that the ML predictions are accurate in the LEARN stage (Section~\ref{sec:gp-results}).
+We present our RBS-optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
+Machine learning is used in two different ways: i) we show the efficacy of the ML recommendations
+in the Design stage (Section~\ref{sec:ucb-results}),
+and ii) we demonstrate that the ML predictions are accurate in the Learn stage (Section~\ref{sec:gp-results}).
 We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and describe some interesting
 characteristics of the discovered sequences, 
-as well as show the effectiveness of the automated laboratory workflow stages.
+as well as show the effectiveness of the automated laboratory workflow.
 
@@
 
-We show our DBTL workflow, that uses machine learning to optimise protein expression, in Figure \ref{fig: Flowchart}.
-BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
-Machine learning algorithms are applied in LEARN and DESIGN.
-In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
-In DESIGN phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS designs based on the predictions from LEARN.\\
+Our DBTL workflow, which uses machine learning to optimise protein expression, is shown in Figure \ref{fig: Flowchart}.
+Build and Test are driven chiefly by choices made by human researchers and the use of automated methods.
+Machine learning algorithms are applied in Learn and Design.
+In the Learn phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
+In the Design phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS sequences based on the predictions from Learn.\\
 
@@
     \includegraphics[scale=0.7]{plots/Main_Paper/flowchart.pdf}
-    \caption{\textbf{Flowchart of the machine learning based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm.
-After generating the recommendations the RBS are built and tested using automated laboratory methods allowing for rapid construction and testing at scale.
-Finally, the obtained results are fed back to the Gaussian Process Regression prediction algorithm in the learn phase.
-}
+    \caption{\textbf{Flowchart of the machine learning-based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm.
+After generating the recommendations, the RBSs are built and tested using automated laboratory methods allowing for rapid construction and testing at scale.
+Finally, the obtained results are fed back to the Gaussian Process Regression prediction algorithm in the Learn phase.
+n is the current design round and k is the maximum number of rounds allowed by time and/or money.}
     \label{fig: Flowchart}
@@
 
-In our genetic design, the investigated RBS controls expression of the Green Fluorescent Protein (GFP). 
-By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring plasmid with the device over time.
+In our genetic design, the investigated RBS controls expression of green fluorescent protein (GFP). 
+By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring the expression vector over time.
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-By making the whole device inducible we can synchronise the start of the expression of the GFP in all the cultures by inducing them at the same time with addition of IPTG.\\
+Inducible expression allows one to synchronise the start of the GFP expression in all the cultures upon addition of IPTG.\\
 
-Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
-This sequence is known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
+Our template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
+This sequence is known to have a very high TIR and is present in the pBb series of plasmids \cite{Lee2011}. 
 Since this is the sequence against which new RBS sequences will be benchmarked,
-we will refer to this sequence as the \textit{benchmark sequence} from now on.
+we will refer to this sequence as the \textit{benchmark sequence} hereafter.
 We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
-In effect, in our design we focus on modification of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e.
+In our design, we focus on modification of the core at nucelotide positions -8 to -13 (relative to the start codon of the GFP) of the RBS and keep other positions the same as the benchmark sequence, i.e.
 TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).
@@
 In short, this algorithm is a stochastic method of probing of the experimental space. 
-This algorithm aims at maximising the reward (output) from testing a limited number of instances from a big pool which cannot be wholly tested due to limited resources (time, computational power, capital).
-It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBS with high predicted TIR.\\
+It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
+It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\
 
 Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the RBS groups tested experimentally. 
-In each experimental round, apart from the new RBS designs, we measure the TIR of benchmark RBS as the internal standard. 
+In each experimental round, in addition to the new RBS designs, we measure the TIR of the benchmark RBS as the internal standard. 
 We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR of a new design and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\
@@
     BN - benchmark sequences for all plates. 
-    They are not all exactly 1 due to them being shown as separate samples rather than per round averages.
+    (not all are exactly 1 due to them being shown as separate samples rather than per round averages.)
     \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
@@
 
-To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, which constitutes a little over 10\% of the whole experimental space. 
+To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, little over 10\% of the whole experimental space. 
 To fit into our automated workflow, we have divided the 450 variants into batches of 90 split into 4 design rounds.\\
 
-In the zeroth round we have tested two batches of designs, for total of 180 variants split as below:
+In the zeroth round we have tested two batches of designs, giving a total of 180 variants split as below:
 
@@
     \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. 
-    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E. coli} genome \cite{Stormo1982}.
+    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E.
+coli} genome \cite{barrick1994quantitative}.
+    % \cite{Stormo1982}.
     \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIRs.
@@
 All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR by the changes made outside the RBS core. 
-The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on only approximate data.\\
+The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR of changes made outside the RBS core. 
+The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on approximate data.\\
 
-However, starting from Round 1, where the prediction and recommendation algorithms were fed data from the Round 0 the results improve significantly, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
-In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
-Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
+However, starting from Round 1, where the prediction and recommendation algorithms were fed data from Round 0, the results improve significantly, with a number of sequences which perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
+In Round 2 we have observed further improvement by obtaining more sequences with TIRs similar to our benchmark sequence.
+Finally, in Round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
 
-Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
-The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
-The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
-We see that effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
-Those with high predicted mean are coloured blue and represent exploitation, those hued red are with high predicted uncertainty and represent exploration.
-We can see that the RBSs with high TIRs tend to come from exploitation of the design space whereas the explorative points give relatively low TIR, but expand our knowledge about the unknown part of the design space.\\
+Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group shows the highest TIR for that quantile.
+The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
+The decreased result in the 0.9th quantile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
+We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
+We can see that the RBSs with high TIRs tend to come from exploitation of the design space, whereas the explorative points give relatively low TIRs, but expand our knowledge of the unknown part of the design space.\\
 
-Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins of width of 0.1 TIR ratio.
+Figure \ref{fig: Swarmplot and Quantplot}D shows the TIRs of RBSs tested in the Bandit groups divided into bins with width equal to a TIR ratio of 0.1.
 KDE plots have been overlaid to depict the calculated density for each group.
-The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
-Notably, the distributions calculated for all the groups show bimodal distribution, we discuss the possible reasons for that further in the text.\\
+The increase in prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3, constituting the bulk of results in the $>0.8$ TIR ratio bins.
+Notably, the distributions calculated for all the groups are bimodal - we discuss the possible reasons for that later in the text.\\
 
 In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
-Each RBS is located on the plot according to its distance to other RBSs as calculated by our embedding function.
-We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
+Each RBS is located on the plot according to its distance from other RBSs as calculated by our embedding function.
+The RBSs recommended by Bandit groups have covered majority of the design space. 
 Additionally, a number of clusters were especially targeted by our recommendation algorithm.
@@
 
-In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
-We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
-We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit well into our prediction method.
+In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
+We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
+% We have investigated a number of methods of digitising the DNA sequence, finally settling on 
+To represent RBSs and capture the similarities between them, we chose to use the
+Weighted Degree Kernel with Shift method, which fit well into our prediction method.
 In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow, we have found very strong RBS sequences and we have generated an extensive library of diverse RBS that can be used in the future studies.
-We have achieved this despite the relatively low accuracy of our predictions, which means that the presented algorithms are robust and able to identify the right signal in a noisy environment. \\
+By using our workflow and testing only 450 RBS variants, we have discovered 
+% designed and experimentally-validated 
+RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
+Furthermore, we have generated an extensive library of diverse RBS that can be used in the future studies.\\
+
+Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require a substantial time and experiments to uncover.
+For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR of such sequences.
+We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
+This gives hope that the continuing rise in use of machine learning in synthetic biology will enable to uncover new biologically relevant rules of design.\\
 
@@
 
-Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require a substantial time and experiments to uncover.
-For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR of such sequences.
-This gives hope that the continuing rise in use of machine learning in synthetic biology will enable to uncover new biologically relevant rules of design.\\
+In this study we have limited the number of design rounds to four.
+There was a number of reasons for this, including time and money limitations, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
+There is a possibility that increasing number of experimental rounds would enable us to improve the results further, however this has to be put in the context of the limited resources.
+For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would need unreasonable use of resources.
+Our approach also compares favourably with approaches using deep learning \cite{Hollerer2020}, where the data sets required for accurate predictions are 3 orders of magnitude bigger than ours. \\
 
 There are still open questions that need to be addressed for applying machine learning in synthetic biology.
-For one, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
+Firstly, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
 We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the obtained results.
@@
 We believe that the method described here would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
-
 Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.
@@
 
-All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML} .
+All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/Solaris/tree/master/synbio_rbs} .
 All the processed and raw data is included in the repository.

### 2021-06-15 paper/methods.tex
Updates from Overleaf

@@
 \subsubsection{Prediction: Gaussian Process Regression with String Kernel}
+\label{sec: method prediction with kernel}
 
@@
 The second data point is then recommended based on the new UCB scores.   
-Figure \ref{fig:batch rec}C shows that since we assume we have observed $x = 2$, then the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, we recommend the data point which is in a different maximum area with potentially high labels ($x = 8$).
-\mengyan{need clarification (commented by iadine)}
+Figure \ref{fig:batch rec}C shows that since we assume we have observed $x = 2$, the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, the algorithm recommends $x = 8$, which is in another local maximum design area.
+In this way, the batch recommendation can potentially cover more local maximum areas than the sequential design. 
 In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 

### 2021-06-15 paper/supplementary.tex

@@
 
-The choice of covariance function is critical for the performance of Gaussian process regression, we show a number of different string kernels tested in this study below:
+The choice of covariance function is critical for the performance of Gaussian process regression. 
+Except the weighted degree kernel with shift described in Section \ref{sec: method prediction with kernel}, we show two closely related string kernels tested in this study below.
 
@@
 
-    \item \textit{Weighted Degree Kernel With Shift.}
-    \begin{align}
-        k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
-        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
-        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
-        \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
-    \end{align}
-    where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
-    sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
-    range at position $l$.
+    % \item \textit{Weighted Degree Kernel With Shift.}
+    % \begin{align}
+    %     k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
+    %     &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+    %     \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
+    %     &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+    %     \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
+    % \end{align}
+    % where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
+    % sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
+    % range at position $l$.
 \end{itemize}
@@
 For our experiment, we fix the design space before training, i.e. the testing features are already known before testing. 
-So we choose to normalise the kernel matrix over the training and testing data together,
+So we choose to normalise the kernel matrix over the training and testing data together (for Round 2-3, normalise over all RBS sequence in the design space),
 by first applying centering and then unit norm normalisation.
@@
 
-For recommending RBS sequences to label, we consider the Upper Confidence Bound (UCB) algorithm, 
-%which is based on the \textit{optimism in the face of uncertainty}, 
-selecting RBS sequences with the maximum upper confidence bound at round $t$, i.e.
-\begin{align}
-    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
-    \tag{\ref{Eq: GPUCB}}
-\end{align}
-where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
-
-Since labelling sequences is time-consuming, it is unrealistic to recommend sequence sequentially (i.e. one-by-one) and waiting for the label after each prediction.
-Therefore we consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2014parallelizing}.
+We consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2014parallelizing}.
+We show technical details of GP-BUCB in the following.
 With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
@@
 
-\newpage
+% \newpage
 
@@
     \includegraphics[scale = 0.4]{plots/Supplementary/swarmplot_raw.pdf}
-    \caption{Swarm plot showing the obtained raw TIRs divided into RBS groups.
+    \caption{
+    \textbf{Raw TIRs of RBS groups examined in this study.}
     Raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth. 
@@
     \centering
-    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_0.pdf}
-    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_1.pdf}
-    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_2.pdf}
-    \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
-    \caption{scatter abc1 TT}
+    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_bc1_TTF_0.pdf}
+    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_1.pdf}
+    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_2.pdf}
+    % \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
+    \caption{\textbf{Predictions in design pipeline in Round 1-3.} For round 1, the kernel matrix is normalised based on known RBS sequences in Round 0.
+For round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.}
     \label{fig:scatter abc1 TT.}
@@
    \centering
-   \includegraphics[width=\textwidth]{plots/Supplementary/All_logo.pdf}
+   \includegraphics[width=0.5\textwidth]{plots/Supplementary/All_logo.pdf}
    \caption{\textbf{Sequence logo for all tested sequences.} }
@@
    \centering
-    \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
-    \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
-    \includegraphics[scale=0.4]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{Exploration v.s. Exploitation.}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} The orange points show the predicted STD and TIR (with the setting of Figure \ref{fig: Scatterplot}) for test data points (Recommendation in Round 1-3), which are projected to the diagonal line.
+The projected values are the exploration-exploitation scores shown in Figure \ref{fig: Swarmplot and Quantplot}C.}
 \end{figure}

### 2021-06-15 paper/main.tex
mengyan edit on abstract and discussion

@@
 For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
-More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
+More specifically, sequences with 4 or more guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
 

### 2021-06-15 paper/main.tex
Updates from Overleaf

@@
 % Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
-Ribosome binding site (RBS), as one of the engineering transcription and translation control elements, allows fine control of gene expression.
+The ribosome binding site (RBS), as one of the transcription and translation control elements, allows fine control of gene expression.
 % Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
-We consider the problem of designing RBSs in nucleotides level to optimise the gene expression with fixed budget. 
+We consider the problem of designing RBSs at the nucleotide level to optimise gene expression with a fixed budget. 
 To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
 We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
-We have integrated these ML algorithms with laboratory automation and high-throughput processes for more reliable data generation.
-Our approach has shown that the algorithms were able to identify rules conducive to high protein expression.
+We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
+The algorithms were able to identify RBS sequences rules conducive to high protein expression.
 Notably, by testing only 450 RBS variants, we have discovered 
@@
 RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
-Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for the experimental design of more complicated genetic devices.
+Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.
 
@@
 \label{sec:ucb-results}
-
-The design recommendations were made using the Multi-armed Bandit algorithm.
-In short, this algorithm is a stochastic method of probing of the experimental space. 
-It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
-It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\
 
@@
 
-In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).\\
+In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
+The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
+% In short, this algorithm is a stochastic method of probing of the experimental space. 
+% It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
+It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\
 
@@
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
-The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
-Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.\\
+The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.067 for round 0 to R\textsuperscript{2} of 0.213 for round 3.
+Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.\\
 
 One important point to note is that the predictions are also influenced by our recommendation choices. 
-In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
+In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having a real mean different from what was predicted.
 However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\
@@
 \begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
-Raw TIR range                                                                                     &      [4.926, 105.377]               \\ \hline
-TIR ratio range                                                                                     &      [0.063, 1.339]               \\ \hline
+Raw TIR range                                                                                     &      [4.93, 105.38]               \\ \hline
+TIR ratio range                                                                                     &      [0.06, 1.34]               \\ \hline
 \end{tabular}
@@
 \textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
-GGGGGC                & 1.339             \\ \hline
-GGGGGT                & 1.154             \\ \hline
-GGCTAT                & 1.084              \\ \hline
+GGGGGC                & 1.34             \\ \hline
+GGGGGT                & 1.15             \\ \hline
+GGCTAT                & 1.08              \\ \hline
 \textbf{AGGAGA}                & 1                  \\ \hline
-GGCGTT                & 0.981            \\ \hline
-GGGGGG                & 0.979             \\ \hline
-GGCGAC                & 0.976             \\ \hline
-CAGGAG                & 0.963             \\ \hline
-GGCGAG                & 0.952             \\ \hline
-\textbf{AGGAGG}                & 0.394            \\ \hline
+GGCGTT                & 0.98            \\ \hline
+GGGGGG                & 0.98             \\ \hline
+GGCGAC                & 0.98             \\ \hline
+CAGGAG                & 0.96             \\ \hline
+GGCGAG                & 0.95             \\ \hline
+\textbf{AGGAGG}       & 0.39            \\ \hline
 \end{tabular}
@@
 For RBS with high TIR ($>1$), the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
-For RBS with medium TIR ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
-That means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
+For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
+This means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
 This also implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.
@@
 
-Finally, our strong sequences did not show neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
+Finally, our strong sequences showed neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS, nor any obvious secondary structures that could explain their TIRs (see Supplementary).
 This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
@@
 Weighted Degree Kernel with Shift method, which fit well into our prediction method.
-In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
+In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
 By using our workflow and testing only 450 RBS variants, we have discovered 
@@
 RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
-Furthermore, we have generated an extensive library of diverse RBS that can be used in the future studies.\\
+Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\
 
-Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require a substantial time and experiments to uncover.
-For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR of such sequences.
+Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
+For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
 We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
-This gives hope that the continuing rise in use of machine learning in synthetic biology will enable to uncover new biologically relevant rules of design.\\
+This gives hope that the continuing rise in use of machine learning in synthetic biology will enable us to uncover new biologically relevant rules of design.\\
 
-We have found our approach bringing machine learning and synthetic biology experts very powerful.
-We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
-Synthetic biology is promoting standardised and normalised testing in biology which naturally pairs with machine learning which can leverage the high quality biological data sets generated when the correct design rules are followed.
+We have found our approach of bringing machine learning and synthetic biology experts together very powerful.
+We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
+Synthetic biology is promoting standardised and normalised testing in biology which naturally pairs with machine learning which can leverage the high quality biological data sets generated when the correct design rules are observed.
 The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
@@
 There was a number of reasons for this, including time and money limitations, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
-There is a possibility that increasing number of experimental rounds would enable us to improve the results further, however this has to be put in the context of the limited resources.
-For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would need unreasonable use of resources.
+There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however this has to be put in the context of the limited resources.
+For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
 Our approach also compares favourably with approaches using deep learning \cite{Hollerer2020}, where the data sets required for accurate predictions are 3 orders of magnitude bigger than ours. \\
@@
 There are still open questions that need to be addressed for applying machine learning in synthetic biology.
-Firstly, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
+Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
 We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the obtained results.
-Secondly, given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
+Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurement? 
 Thirdly, the generalisability of the method is unknown.
-We believe that the method described here would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
-Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.
-In this research we have decided to gradually move the balance from exploration to exploitation over the rounds, but it is not yet clear what is the optimal way of conducting this change.\\
+We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
+Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
+In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.\\
 
-In the future, we hope to extend the algorithm to other, more complicated genetic elements.
-This could include, for example, promoters and terminators.
-However, it's important to reiterate that the complexity of the task quickly increases with the length of the sequence.
-This is because the experimental space grows exponentially with the number of examined positions and so the space becomes increasingly hard to cover with experiments.
-To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
+In the future, we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
+However, it is important to reiterate that the complexity of the task quickly increases with the length of the sequence.
+This is because the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments.
+To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
 

### 2021-06-15 paper/main.tex
Final paper update.

@@
 
-% Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
-The ribosome binding site (RBS), as one of the transcription and translation control elements, allows fine control of gene expression.
-% Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
-We consider the problem of designing RBSs at the nucleotide level to optimise gene expression with a fixed budget. 
+Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
+Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
 To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
@@
 The algorithms were able to identify RBS sequences rules conducive to high protein expression.
-Notably, by testing only 450 RBS variants, we have discovered 
-% designed and experimentally-validated 
-RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
+Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
 Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.
@@
 \label{sec:ucb-results}
+
+The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
+In short, this algorithm is a stochastic method of probing of the experimental space. 
+It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
+It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\
 
@@
 
-In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
-The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
-% In short, this algorithm is a stochastic method of probing of the experimental space. 
-% It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
-It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\
+In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).\\
 
@@
 We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
-% We have investigated a number of methods of digitising the DNA sequence, finally settling on 
 To represent RBSs and capture the similarities between them, we chose to use the
-Weighted Degree Kernel with Shift method, which fit well into our prediction method.
+Weighted Degree Kernel with Shift method, which fits well into our prediction method.
 In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow and testing only 450 RBS variants, we have discovered 
-% designed and experimentally-validated 
-RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
+By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
 Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\

### 2021-06-15 paper/methods.tex
Final paper update.

@@
 %train GP on sequences with known TIR labels and
-recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}(A) ($m = 2$).
+recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}A ($m = 2$).
 However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
 Since GPR assumes similar sequences would have similar labels (e.g.
-by knowing $x=2$ we can gain information of $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
+by knowing $x=2$ we can gain information about $x=2.5$ as well), we prefer to not to waste time and money on labelling sequences with high similarities in the same batch.
 To counter this, we use a batch recommendation strategy that is designed to avoid recommending such highly similar sequences in the same batch, described below. \\
@@
 {-2, -1.5, -1, ..., 9.5, 10}.
-    The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
+    The confidence intervals are shown with predicted mean of $\pm$ 1.96 standard deviation.
     (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, chosen with GP predictions. 
-    (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that.}
+    (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that updated interval.}
     \label{fig:batch rec}
@@
 As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
-Then we add the recommended data point ($x = 2$) into the training dataset with the predicted mean of that point as label (note it is not the true label, i.e.
+Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e.
 observation), and update the predicted standard deviation and then we finally update the UCB scores.
@@
 In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 
-In our design, we have set the batch size as 90, to fit the experimental batch.
-We set $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.
+In our design, we have set the batch size to 90, to fit the experimental batch.
+Finally, we set $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.

### 2021-08-04 paper/main.tex
Updates from Overleaf

@@
 \usepackage{authblk}
+\usepackage{soul}
+
 % \usepackage[table,xcdraw]{xcolor}
@@
 
-\author[1,2,4]{Zhang M.}
-\author[3]{Holowko M. B.}
-\author[3]{Hayman Zumpe H.}
-\author[1,2,4]{Ong, C. S.}
+\author[1,2,4]{Zhang Mengyan}
+\author[3]{Holowko Maciej Bartosz}
+\author[3]{Hayman Zumpe Huw}
+\author[1,2,4,*]{Ong Cheng Soon}
 \affil[1]{Machine Learning and Artificial Intelligence Future Science Platform, CSIRO}
 \affil[2]{Department of Computer Science, Australian National University}
-\affil[3]{CSIRO Synthetic Biology Future Science Platform, CSIRO Land and Water}
+\affil[3]{Synthetic Biology Future Science Platform, CSIRO}
 \affil[4]{Data61, CSIRO}
+\affil[*]{email: cheng-soon.ong@data61.csiro.au}
 
@@
 Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
-Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.
+Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\
+
+Keywords: machine learning, genetic part design, ribosome binding site
+
+\newpage
 
@@
 
-The biggest gap in the DTBL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
-For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+The biggest gap in the DBTL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
+For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009,Reis2020}.
 Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
-but they come with a number of caveats: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which increases perceived prediction error \cite{Goss1998}.
+but they come with a number of caveats: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}.
 There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}.
@@
 Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
-For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoids in \emph{Esherichia coli} \cite{Jervis2019}.
+For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoids in \emph{Escherichia coli} \cite{Jervis2019}.
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}.
@@
 We used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
-As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that can be later translated to more complicated systems.
+As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that \hl{could be potentially translated to more complicated systems}.
 Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
@@
 
+\input{methods.tex}
+
 \section{Results}
@@
 We present our RBS-optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}. 
+hl{In this case, the \hl{DBTL} paradigm is applied directly to the design of the RBS itself, rather than general strain optimisation.}
 Machine learning is used in two different ways: i) we show the efficacy of the ML recommendations
@@
 Finally, the obtained results are fed back to the Gaussian Process Regression prediction algorithm in the Learn phase.
-n is the current design round and k is the maximum number of rounds allowed by time and/or money.}
+n is the current design round and k is the maximum number of rounds allowed by time and/or money.
+\hl{TIR(n)=TIR(n-1) denotes "lack of TIR improvement following the used optimisation function".}}
     \label{fig: Flowchart}
@@
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-Inducible expression allows one to synchronise the start of the GFP expression in all the cultures upon addition of IPTG.\\
+Inducible expression allows one to synchronise the start of the GFP expression in all the cultures upon addition of IPTG.
+\hl{Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier} \cite{Beal2021}.\\
 
-Our template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
-This sequence is known to have a very high TIR and is present in the pBb series of plasmids \cite{Lee2011}. 
+Our template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
+This sequence is known to have a very high TIR \hl{in a similar context of expressing fluorescent protein} and is present in the pBb series of plasmids \cite{Lee2011}.
 Since this is the sequence against which new RBS sequences will be benchmarked,
 we will refer to this sequence as the \textit{benchmark sequence} hereafter.
-We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
-In our design, we focus on modification of the core at nucelotide positions -8 to -13 (relative to the start codon of the GFP) of the RBS and keep other positions the same as the benchmark sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).
+\hl{Additionally, we have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it }(see Figure \ref{fig:core_vs_noncore}).
+\hl{This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.}
+\hl{For example, according to} \cite{SHULTZABERGER2001} \hl{there is a strong bias towards A and G bases in the core region of the RBS.}
+\hl{Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overall TIR of a given RBS.}
+\hl{Notably, focusing on the 6 to 8bp long core sequence is a common RBS design approach} \cite{Jeschek2016}.
+Hence, in our design, we focus on modification of the core at nucleotide positions -8 to -13 (relative to the start codon of the GFP; \hl{ this is where the consensus Shine-Dalgarno AGGAGG sequence is usually found in wild type \textit{E.
+coli}) of the RBS and keep other positions the same as the benchmark sequence, i.e.
+TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).}
 The total experimental (variant) space is then $4^6$ = 4096.
@@
 In each experimental round, in addition to the new RBS designs, we measure the TIR of the benchmark RBS as the internal standard. 
-We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR of a new design and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\
+We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR of a new design and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
+Figure \ref{fig:rawswarmplots.} shows these results in terms of raw TIRs.\\
 
@@
     UNI: Randomly generated sequences with uniform distribution. 
-    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
+    PPM: Randomly generated sequences with distribution following the \hl{Position Probability Matrix} for all natural RBS in \emph{E. coli}. 
     Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
@@
     \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
-    The area of the circle corresponds to the the experimentally obtained TIR value.
+    The area of the circle corresponds to the experimentally obtained TIR value.
     The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
@@
 \begin{itemize}
-    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide variations of all 20 nucleotides of the original, consensus sequence.
+    \item BPS-NC and BPS-C group: 60 RBS sequences which are subsequent single nucleotide variations of all 20 nucleotides of the original, benchmark sequence.
 This batch is designed to show us influence of such single nucleotide changes on the overall performance of the RBS and the potential impact of changes made beyond the core part (see Supplementary Figure \ref{fig:core_vs_noncore}).
@@
 The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR of changes made outside the RBS core. 
-The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on approximate data.\\
+The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on approximate data.
+\hl{These results show that generating a strong RBS sequence by random mutations is a non-trivial task, when the tested data set is relatively small.}\\
 
@@
 The decreased result in the 0.9th quantile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
-We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown if Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
@@
 
-In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
+In \ref{fig: Swarmplot and Quantplot}E we show a \hl{t-distributed stochastic neighbour embedding (t-SNE)} plot depicting the experimental space.
 Each RBS is located on the plot according to its distance from other RBSs as calculated by our embedding function.
@@
 Additionally, a number of clusters were especially targeted by our recommendation algorithm.
-For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
+For example, the circled clusters labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
 More specifically, sequences with 4 or more guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
@@
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
-The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.067 for round 0 to R\textsuperscript{2} of 0.213 for round 3.
-Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.\\
+The predictions improved for the subsequent rounds, from R\textsuperscript{2} (coefficient of determination) of 0.067 for round 0 to R\textsuperscript{2} of 0.213 for round 3.
+Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
+Figure \ref{fig:scatter abc1 TT.} shows performance of predictions for individual rounds.\\
 
-One important point to note is that the predictions are also influenced by our recommendation choices. 
-In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having a real mean different from what was predicted.
-However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\
+\hl{The obtained values of the coefficient of determination and Spearman correlation coefficient are influenced by a number of factors}
+\hl{Primary one of them is our choice of the exploitation to exploration balance.}
+\hl{In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having a real mean different from what was predicted, thus decreasing the R\textsuperscript{2} values.}
+\hl{This is, however, a very useful information for future predictions as it allows us to understand the underlying space better.}
+\hl{Effect of the better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.}
+\mengyan{Learn and test only on exploitation?}\\
 
@@
 
-Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in Table 1.\\
+Our data set taken together can be viewed as a \hl{reliable (low coefficients of variation) and extensive (high number of entries)} library of RBS sequences for \emph{E.
+coli}, some characteristics of which are shown in Table 1.\\
 
-Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (in terms of TIR ratio).
+Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows logo generated for all tested sequences).
 It is generally understood that guanine rich sequences are promoting strong transcription.
@@
 Right table presents 10 RBS sequences with their corresponding TIR ratios, the first 9 being the strongest sequences including the benchmark sequence (AGGAGA) and the last being the Shine-Dalgarno sequence (AGGAGG).
-CV is coefficient of variation (STD of a sample divided by its mean).
+CV is coefficient of variation (standard deviation (STD) of a sample divided by its mean; see details in Figures \ref{fig: SDhist} and \ref{fig:variation of biological replicates}).
 Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by the highest TIR found using random sequence generation. }
@@
 In such cases, a directed process should be strongly encouraged.
-This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.\\
+This is in l \cite{Jackel2008}.\\
 
-Finally, our strong sequences showed neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS, nor any obvious secondary structures that could explain their TIRs (see Supplementary).
+Finally, our strong sequences showed neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS, nor any obvious secondary structures that could explain their TIRs (see Figure \ref{fig:structures}).
 This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
@@
 Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\
+
+\hl{In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.}
+\hl{This approach may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.}
+\hl{This change could be achieved within the Bayesian optimisation framework by altering the objective of the algorithm to e.g.
+maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objectives method \mbox{\cite{shu2020new}}.}\\
 
@@
 For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
-Our approach also compares favourably with approaches using deep learning \cite{Hollerer2020}, where the data sets required for accurate predictions are 3 orders of magnitude bigger than ours. \\
+\hl{Compared to solutions like the one reported by Hollerer} \cite{Hollerer2020}, \hl{our solution can be used when a high volume method for data-generation is not available, while still providing the required results (optimised part).} \\
 
@@
 We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
+\hl{Additionally, our method would not be appropriate for more complex cases involving multi-part devices/systems.}
+\hl{In such cases, significantly higher number of data points and deep learning techniques would be required since the number of parameters to optimise would be also significantly higher.}
 Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
@@
 
-\input{methods.tex}
-
-\section*{Code and data availability}
+\section*{Code, data and material availability}
 
@@
 Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.
+The pBbB6c plasmid is available on Addgene. Other strains and plasmids are available on request from the authors; MTA will need to be negotiated between the parties.
 

### 2021-08-04 paper/methods.tex
Updates from Overleaf

@@
 The pBbB6c-GFP plasmid has been used for all our designs.
-This plasmid comes with GFP mut3b CDS inducible with addition of Isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG).
+This plasmid comes with GFP mut3b CDS inducible with addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG).
 The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
@@
 PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, catalogue no. M0492L).
-20 $\mu$L reactions were prepared by dispensing each of the 10 $\mu$M reverse primers into a well of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
-A mastermix consisting of polymerase premix, plasmid DNA template, and the single 10 $\mu$M forward primer was prepared by and dispensed by hand and/or Labcyte Echo.
+20 $\mu$L reactions were prepared by dispensing \hl{1 $\mu$L of each 10 $\mu$M reverse primer into the wells} of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
+A mastermix consisting of polymerase premix, plasmid DNA template \hl{(pBbB6C, 5-10 ng per reaction)}, and the single 10 $\mu$M forward primer was prepared by and dispensed \hl{using the FeliX liquid handler or electronic multi-channel pipette}.
 Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
 Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
-2$\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no. ZAG-110-5000; ZAG-130-5000).
+2 $\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
+ZAG-110-5000; ZAG-130-5000).
+\hl{ZAG sample plates were prepared using the Perkin Elmer Sciclone G3 liquid handler.}
 ProSize Data Analysis Software (Agilent Technologies) was used to generated gel images from the sample chromatograms and sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
@@
 Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no. E2621L).
-Reactions consisting of the common fragment and the variable fragment were prepared using the Echo acoustic liquid handler, to a final volume of 5 or 10 $\mu$L.
+Reactions consisting of \hl{approximately equal amounts} of the common fragment and the variable fragment were prepared \hl{using the FeliX liquid handler or electronic multi-channel pipette}, to a final volume of 5 or 10 $\mu$L.
 Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
-Finally, samples were incubated with addition of 50 nL of DpnI at 37$^{\circ}$C for 90 minutes.
+Finally, samples were \hl{incubated with of 50 nL of DpnI (NEB, catalogue no. R0176S) at 37$^{\circ}$C for 90 minutes to remove any residual template DNA.}
 
@@
 T3001).
-20$\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
-Plates of cells were thawed on a -20$^{\circ}$C cold block before 3$\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
-Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no.
-244610) and 15g/L agar (Sigma-Adrich, catalogue no.
-A1296) with 34$\mu$g/mL chloramphenicol (Sigma-Aldrich, catalogue no.
+20 $\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
+Plates of cells were thawed on a -20$^{\circ}$C cold block before 3 $\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
+Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 \hl{(12 x 8)} grid on Omnitrays containing LB (BD, catalogue no.
+244610) and 15 g/L agar (Sigma-Aldrich, catalogue no.
+A1296) with 34 $\mu$g/mL chloramphenicol (Sigma-Aldrich, catalogue no.
 C1919).
-Plates were incubated overnight at 37$^{\circ}$C.
+Plates were incubated overnight at 37$^{\circ}$C. \hl{Cells were plated using the CyBio FeliX liquid handler.}
 
 \textbf{Automated colony picking and culturing.}
-A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates using the 490-510nm (cyan) light filter .
-Each selected colony was used to inoculate 1mL of selective medium in a 2mL square well 96 plate.
-They were then cultured overnight in 37$^{\circ}$C with shaking (~300rpm).
+A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates using the 490-510 nm (cyan) light filter.
+Each selected colony was used to inoculate 1 mL of selective medium in a 2 mL square well 96 plate.
+They were then cultured overnight in 37$^{\circ}$C with shaking (~300 rpm).
 
 \textbf{Glycerol stock preparation.}
-100$\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, catalogue no.
-GA010) and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
-They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.
+100 $\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, catalogue no.
+GA010) and 100 $\mu$L of overnight culture were combined in the wells of a 96 deep (2 mL) round well plate \hl{using the CyBio FeliX liquid handler or electronic multichannel pipette}
+They were then sealed with a 96-well silicone sealing mat and transferred to a -80$^{\circ}$C freezer.
 
@@
 \textbf{Test strain culture.}
-Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
-Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours. 
-The following morning, 20$\mu$L of overnight cultures were added to 980$\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate. 
-After 90 minutes, 300$\mu$L of each culture (induced with 1.0$\mu$L of 0.1M IPTG (Sigma-Aldrich, catalogue no.
-I5502) to a final concentration of 0.5 mM) was transferred to a flat-bottom clear polystyrene 96-well plate.
+Overnight cultures were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, \hl{using the CyBio FeliX liquid handler or electronic multichannel pipette}
+Cultures were incubated at 37$^{\circ}$C with shaking (~300 rpm) for ~17 hours. 
+The following morning, 20 $\mu$L of overnight cultures were added to 980 $\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2 mL round well 96 plate. 
+\hl{After 90 minutes, cultures were induced with IPTG (Sigma-Aldrich, catalogue no.
+I5502) to a final concentration of 0.5 mM.
+This was done by transferring 1.0 $\mu$L of 0.1M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo acoustic liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multichannel pipette}
 
 \textbf{Microplate spectrophotometry.}
-The plates were tested in Cytation5 microplate reader.
-Cytation 5 acquisition and incubation/shaking settings were as follows: length of run: 8h; interval: 10 min; continuous orbital shake at 237 cpm and slow orbital speed; excitation wavelength: 490/10mm; emission wavelength: 515/10 mm; bottom read; gain: 60; read height: 7mm; read speed: Sweep.
+The plates were monitored in the Cytation 5 microplate reader immediately after the addition of IPTG.
+Cytation 5 acquisition and incubation/shaking settings were as follows: length of run: \hl{8 hours}; interval: 10 min; continuous orbital shake at 237 rpm and slow orbital speed; excitation wavelength: 490/10 nm; emission wavelength: 515/10 nm; bottom read; gain: 60; read height: 7 mm; read speed: Sweep.
 
@@
 One type of machine learning algorithms is a prediction algorithm (\textbf{LEARN}), which helps us learn the function of TIR with respect to RBS sequences.
-The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each batch based on the predictions from LEARN.
+The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each round (in the sense of a single pass through the DBTL cycle) based on the predictions from LEARN.
 \\
@@
 Our implementation of the machine learning algorithms was tested in Python 3.6 and used the scikit-learn library \cite{scikit-learn}. 
-In the following paragraphs, we describe our machine learning pipeline (Bandit 1-3) in the following order: data pre-processing, prediction and kernels and finally recommendation. 
-The pipeline for Bandit-0 is reported in Supplementary \ref{sec: ML design pipeline}.
+In the following paragraphs, we describe our machine learning pipeline \hl{(which is applicable to design rounds 1-3, denoted as Bandit 1-3)} in the following order: data pre-processing, prediction and kernels and finally recommendation.
+\hl{The pipeline for our 0th round of designs, denoted as Bandit-0, is reported in Supplementary} \ref{sec: ML design pipeline}.
 
@@
 
-For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluorescence divided by OD600 of culture over 4h counting from the start of the log phase of growth.
+\hl{For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluorescence divided by OD600 of culture over time, following the formula:}
+
+$$\mathnormal{TIR=\int_{t_{min}}^{t_{max}}\frac{GFP}{OD_{600}}dt}$$
+
+\hl{where ${t_{min}}$ is time at which $\frac{GFP}{OD_{600}}$ value is minimum denoting start of the log growth phase and ${t_{max}}$ is time at which the logarithmic phase ends (on average, for our samples that was at 4h from ${t_{min}}$ and so that was the time point used for TIR calculations for all samples).}\\
+
 % REVIEW:
@@
 
-To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ is the TIR score at 20-base RBS sequence $\mathbf{x} \in \{A,C,G,T\}^{20}$.
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ \hl{is the TIR score of the 6-base core sequence considered in its wider 20-base full RBS sequence $\mathbf{x} \in \{A,C,G,T\}^{20}$}.
 In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
@@
 %As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
-GPR can predict both the posterior mean and standard deviation. The posterior standard deviation represents the level of uncertainty for the prediction. 
+GPR can predict both the posterior mean and standard deviation \hl{of the RBS sequences in the design space}. The posterior standard deviation represents the level of uncertainty for the prediction. 
 We review the predictions of GPR in Supplementary \ref{supp: GPR}.\\
@@
 {-2, -1.5, -1, ..., 9.5, 10}.
-    The confidence intervals are shown with predicted mean of $\pm$ 1.96 standard deviation.
+    The confidence intervals are shown with predicted mean $\pm$ 1.96 standard deviation.
     (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, chosen with GP predictions. 
@@
 
-A key property of Gaussian Process regression is that the predicted standard deviation depends only on observed points (i.e. features), not on the labels of those observed points. 
+A key property of Gaussian Process regression is that the predicted standard deviation depends only on \hl{features}, not on the \hl{labels}. 
 % We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
@@
 One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
-The strategy here is to recommend sequences sequentially by updating the UCB score with the predicted standard deviation of the previously recommended points.\\
+The strategy here is to recommend \hl{RBS sequences one-by-one by sequentially adding the newly recommended RBS sequences into the training data and updating UCB scores.}\\
 % Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
@@
 As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
-Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label, i.e.
-observation), and update the predicted standard deviation and then we finally update the UCB scores.
+Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label), and update the predicted standard deviation and then we finally update the UCB scores.
 The second data point is then recommended based on the new UCB scores.   

### 2021-08-04 paper/supplementary.tex
Updates from Overleaf

@@
 
-A Gaussian process is specified by its mean function and convariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(\mu(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
-We consider the case where the observations are noisy, i.e. $\{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$. 
-The Gaussian noise is independent identically distributed, and the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
+A Gaussian process is specified by its mean function and \hl{covariance} function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(\mu(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
+We consider the case where the observations are noisy, i.e.
+$(X, \mathbf{y}) = \{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$.
+The Gaussian noise is independent identically distributed, and the \hl{covariance of the prior on the} noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
 where $\delta_{pq}$ is a Kronecker delta which is one if $p = q$ and zero otherwise.
-It is equivalent to a diagonal matrix $\alpha^2 I$ on the kernel matrix evaluated on the training points.
+It is equivalent to a diagonal matrix $\alpha^2 I$ \hl{added} on the kernel matrix evaluated on the training points.
 
@@
     \begin{align}
-        k_\ell^{\text{Spec}}(X, X^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
+        k_\ell^{\text{Spec}}(\mathbf{x}, \mathbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
     \end{align}
      where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
-    $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $X$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
-    Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_d^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
+     \hl{
+     In our case, $\Sigma = \{A,C,G,T\}$ and $|\Sigma| = 4$.
+     }
+    $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\mathbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+    \hl{
+    For example, if $\mathbf{x} = ACACAG$ and $\ell = 3$, then $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps $\mathbf{x}$ into $4^3$ dimensional feature space, where each dimension corresponds to $AAA, AAC, AAT, AAG, ACA, ..., TTT$.
+Among those, the dimension corresponds to $ACA, CAC, CAG$ are 2,1,1 and others are 0.
+    }
+    Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_\ell^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
     \begin{align}
@@
 As defined before, consider $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$.
-We denote $\phi(\mathbf{x}_i)$ as a column feature vector of sequence $\mathbf{x}_i$, 
-where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$. Assume there is total of $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
+We denote $\phi(\mathbf{x})$ as a column feature vector of sequence $\mathbf{x}$, 
+where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$,
+\hl{with $d$ as the dimension of features.
+One example of $\phi$ can be $\phi_\ell^{spec}(\mathbf{x})$ in the spectrum kernel.
+Recall the corresponding kernel can be defined as $k(\mathbf{x}, \mathbf{x}^\prime) = \phi(\mathbf{x})^T \phi(\mathbf{x}).$
+}
+Assume there is total of $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
 We illustrate centering and unit norm normalisation below.
@@
     \item Unit Norm.
-Define the ($l_2$) norm of a feature vector as $||\phi(\mathbf{x})|| = \sqrt{\sum_{m = 1}^d \phi_d(\mathbf{x})^2} = \sqrt{k(\mathbf{x}, \mathbf{x})} \in \mathbb{R}^+$, then the unit norm feature vector $\phi^{UN}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+Define the ($l_2$) norm of a feature vector as $||\phi(\mathbf{x})|| = \sqrt{\sum_{m = 1}^d \phi_m(\mathbf{x})^2} = \sqrt{k(\mathbf{x}, \mathbf{x})} \in \mathbb{R}^+$, then the unit norm feature vector $\phi^{UN}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
     \begin{align}
@@
     \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
+    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} 
+    \hl{The core region is highlighted in light green and the lines are rolling averages (in this case mean of the 5 points centred over the indicated position) for each base.}
+    The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base, which is colour coded.
+    \hl{Since only changes from the original base at each position are shown, the dotted lines start and end at different positions depending on how many changes to the respective base are at positions adjacent to the indicated one.}
+    The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.}
+    \label{fig:core_vs_noncore}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
     \includegraphics[scale = 0.4]{plots/Supplementary/swarmplot_raw.pdf}
@@
     }
-    \label{fig:swarmplots.}
+    \label{fig:rawswarmplots.}
+\end{figure}
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} The orange points show the predicted STD and TIR (with the setting of Figure \ref{fig: Scatterplot}) for test data points (Recommendation in Round 1-3), which are projected to the diagonal line.
+The projected values are the exploration-exploitation scores shown in Figure \ref{fig: Swarmplot and Quantplot}C.}
+   \label{fig:exploitationvsexplorationdetail}
 \end{figure}
@@
     \label{fig:scatter abc1 TT.}
-\end{figure}
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
-    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} 
-    The core region is highligthed in light green and the lines are rolling averages for each base. 
-    The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base,
-    which is colour coded.
-    In our study, we have tested a set of sequnces designed to confirm the notion that changing bases within the RBS core (6 bases) is statistically more influencing the TIR than the changes made outside the core.
-    This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
-    For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
-    Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS.
-    This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
-    The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.}
-    \label{fig:core_vs_noncore}
 \end{figure}
@@
    \centering
+   \includegraphics[width=0.5\textwidth]{plots/Supplementary/All_logo.pdf}
+   \caption{\textbf{Sequence logo for all tested sequences.} Compared to \ref{fig:Library characteristics}A there are no significant biases at each position.}
+   \label{fig: All_logo}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
     \begin{subfigure}[b]{0.49\textwidth}
@@
         \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGG.pdf}
-        \caption{$y=x$}
         \label{fig:GGGGGG}
@@
         \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGC.pdf}
-        \caption{$y=3sinx$}
         \label{fig:GGGGGC}
@@
         \includegraphics[scale=0.25]{plots/Supplementary/Structure_CGGGAC.pdf}
-        \caption{$y=5/x$}
         \label{fig:CGGGAC}
@@
         \includegraphics[scale=0.25]{plots/Supplementary/Structure_CACGTT.pdf}
-        \caption{$y=5/x$}
         \label{fig:CACGTT}
@@
    \centering
-   \includegraphics[width=0.5\textwidth]{plots/Supplementary/All_logo.pdf}
-   \caption{\textbf{Sequence logo for all tested sequences.} }
-   \label{fig: All_logo}
-\end{figure}
-
-\begin{figure}[!ht]
-   \centering
-    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
-    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
-    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{\textbf{Exploration-Exploitation scores illustration.} The orange points show the predicted STD and TIR (with the setting of Figure \ref{fig: Scatterplot}) for test data points (Recommendation in Round 1-3), which are projected to the diagonal line.
-The projected values are the exploration-exploitation scores shown in Figure \ref{fig: Swarmplot and Quantplot}C.}
-\end{figure}
-
-\begin{figure}[!ht]
-   \centering
     \includegraphics[scale=0.4]{plots/Supplementary/SDhist.pdf}
-   \caption{Histogram of standard deviations.}
+   \caption{\textbf{Histogram of standard deviations.} Average raw standard deviations for each of the tested RBSs are  shown.}
+   \label{fig: SDhist}
 \end{figure}

### 2021-08-04 paper/main.tex
make plots font larger

@@
     (not all are exactly 1 due to them being shown as separate samples rather than per round averages.)
-    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
+    \textbf{B)} Line plot showing TIR obtained in a given quantile (Q) of results divided into groups as in A).
     % save the random groups which were shown together due to similar distributions.
@@
          \caption{}
-         \includegraphics[scale=0.5]{plots/Main_Paper/Hd_Heatmap.pdf}
+         \includegraphics[scale=0.43]{plots/Main_Paper/heatmap.pdf}
      \end{subfigure}

### 2021-08-31 paper/methods.tex
make plots font larger
Updated text of the manuscript.

@@
 The pBbB6c-GFP plasmid has been used for all our designs.
-This plasmid comes with GFP mut3b CDS inducible with addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG).
-The original RBS for the GFP CDS was replaced with combination of PCR and isothermal assembly.
-Primers and the assembly strategy have been generated using the Teselagen DESIGN software (Teselagen Biotechnology).
+This plasmid \hl{contains the GFP mut3b CDS, expression of which can be induced by the addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG)}.
+The original RBS for the GFP CDS was replaced \hl{using a combination} of PCR and isothermal assembly.
+\hl{Primer sequences and the assembly strategy were generated using the Teselagen DESIGN software (Teselagen Biotechnology)}.
 
@@
 20 $\mu$L reactions were prepared by dispensing \hl{1 $\mu$L of each 10 $\mu$M reverse primer into the wells} of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
-A mastermix consisting of polymerase premix, plasmid DNA template \hl{(pBbB6C, 5-10 ng per reaction)}, and the single 10 $\mu$M forward primer was prepared by and dispensed \hl{using the FeliX liquid handler or electronic multi-channel pipette}.
+A mastermix consisting of polymerase premix, plasmid DNA template \hl{(pBbB6c, 5-10 ng per reaction)}, and the single 10 $\mu$M forward primer was prepared by and dispensed \hl{using the FeliX liquid handler or electronic multi-channel pipette}.
 Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
@@
 \textbf{Test strain culture.}
-Overnight cultures were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, \hl{using the CyBio FeliX liquid handler or electronic multichannel pipette}
+Overnight cultures \hl{(total of six replicates for each batch)} were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, \hl{using the CyBio FeliX liquid handler or electronic multichannel pipette.}
 Cultures were incubated at 37$^{\circ}$C with shaking (~300 rpm) for ~17 hours. 
@@
 I5502) to a final concentration of 0.5 mM.
-This was done by transferring 1.0 $\mu$L of 0.1M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo acoustic liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multichannel pipette}
+This was done by transferring 1.0 $\mu$L of 0.1M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo acoustic liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multichannel pipette.}
 
@@
 
-Two types of machine learning algorithms can be applied to drive the experimental design workflow shown in Figure \ref{fig: Flowchart}.
+Two types of machine learning algorithms have to be applied to drive the experimental design workflow shown in Figure \ref{fig: Flowchart}.
 % To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
@@
 % 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
-In round $t$, the prediction and design is based on the results obtained in all previous rounds. 
+In Round $t$, the prediction and design are based on the results obtained in all previous rounds. 
 Our implementation of the machine learning algorithms was tested in Python 3.6 and used the scikit-learn library \cite{scikit-learn}. 
-In the following paragraphs, we describe our machine learning pipeline \hl{(which is applicable to design rounds 1-3, denoted as Bandit 1-3)} in the following order: data pre-processing, prediction and kernels and finally recommendation.
-\hl{The pipeline for our 0th round of designs, denoted as Bandit-0, is reported in Supplementary} \ref{sec: ML design pipeline}.
+In the following paragraphs, we describe our machine learning pipeline \hl{(which is applicable to design Rounds 1-3, denoted as Bandit 1-3)} in the following order: data pre-processing, prediction and kernels and finally recommendation.
+\hl{The pipeline for our 0th Round of machine learning designs, denoted as Bandit-0, is reported in Supplementary} \ref{sec: ML design pipeline}.
 
@@
 
-\hl{For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as a derivative of GFP fluorescence divided by OD600 of culture over time, following the formula:}
+\hl{For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as an averaged sum of GFP fluorescence divided by Optical Density at 600nm (OD600) of culture over time (calculated using values from all the measured data points), following the formula:}
 
-$$\mathnormal{TIR=\int_{t_{min}}^{t_{max}}\frac{GFP}{OD_{600}}dt}$$
+$$\mathnormal{TIR=\frac{\sum_{t_{min}}^{t_{max}}\frac{GFP(t)}{OD_{600}(t)}}{\Delta t}}$$
 
-\hl{where ${t_{min}}$ is time at which $\frac{GFP}{OD_{600}}$ value is minimum denoting start of the log growth phase and ${t_{max}}$ is time at which the logarithmic phase ends (on average, for our samples that was at 4h from ${t_{min}}$ and so that was the time point used for TIR calculations for all samples).}\\
+\hl{where ${t_{min}}$ is time at which $\frac{GFP}{OD_{600}}$ value is at its minimum denoting start of the log growth phase, ${t_{max}}$ is time at which the logarithmic phase ends (on average, for our samples that was at 4h from ${t_{min}}$ and so that was the time point used for TIR calculations for all samples), ${\Delta t}$ is the difference between ${t_{max}}$ and ${t_{min}}$.}\\
 
@@
 This reference values are the TIR of the benchmark sequence that is run in triplicate in each round.
-Specifically, in round $t$, we subtract the TIR mean of the benchmark RBS measured in round $t$ from all TIR values measured in the same round, for each replicate separately. 
+Specifically, in Round $t$, we subtract the TIR mean of the benchmark RBS measured in Round $t$ from all TIR values measured in the same round, for each replicate separately. 
 We then normalised the data by performing a logarithm transformation and standardisation on the adjusted TIR label for each replicate separately.
@@
 To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ \hl{is the TIR score of the 6-base core sequence considered in its wider 20-base full RBS sequence $\mathbf{x} \in \{A,C,G,T\}^{20}$}.
-In each round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
+In each Round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
 $y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
@@
 on the other hand, UCB explores the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
-More precisely, UCB algorithm selects RBS sequence $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at round $t$, i.e.
+More precisely, UCB algorithm selects RBS sequence $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at Round $t$, i.e.
 \begin{align}
@@
 where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
-We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at round $t-1$.\\
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at Round $t$ for the sequence $\mathbf{x}_i$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at Round $t-1$.\\
 
@@
         \caption{}
-        \includegraphics[scale=.4]{plots/Main_Paper/Recommend_2_Data_Point_Top_2_UCB.pdf}
+        \includegraphics[scale=.4]{plots/Main_Paper/Recommend_2_Data_Points_Top_2_UCB.pdf}
     \end{subfigure}
@@
 One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
-The strategy here is to recommend \hl{RBS sequences one-by-one by sequentially adding the newly recommended RBS sequences into the training data and updating UCB scores.}\\
+The strategy here is to recommend \hl{RBS sequences one-by-one by sequentially adding the newly recommended RBSs predicted means (without testing them) into the training data and updating UCB scores.}\\
+
 % Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
@@
 In our design, we have set the batch size to 90, to fit the experimental batch.
-Finally, we set $\beta_t = 2$ for round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.
+Finally, we set $\beta_t = 2$ for Round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.

### 2021-08-31 paper/main.tex

@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
+\hl{Part design is a central problem of synthetic biology as it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.}
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
@@
 The biggest gap in the DBTL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
-For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009,Reis2020}.
-Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
-but they come with a number of caveats: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}.
+For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+\hl{Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set } \cite{Reis2020}.
+This may be due to number of reasons: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}.
 There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. 
@@
 In this work, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level.
+\hl{Importantly, this work focuses on the RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.}
 We used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
@@
 We present our RBS-optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
-hl{In this case, the \hl{DBTL} paradigm is applied directly to the design of the RBS itself, rather than general strain optimisation.}
+\hl{In this case, the {DBTL} paradigm is applied directly to the design of the RBS part itself, rather than general strain optimisation.}
 Machine learning is used in two different ways: i) we show the efficacy of the ML recommendations
@@
 n is the current design round and k is the maximum number of rounds allowed by time and/or money.
-\hl{TIR(n)=TIR(n-1) denotes "lack of TIR improvement following the used optimisation function".}}
+\hl{In regards to the "Goal Met?" condition, the goal in our case was to find sequences with TIR higher than benchmark, but the goal can be generalised to fit the requirements of the user.} }
     \label{fig: Flowchart}
@@
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-Inducible expression allows one to synchronise the start of the GFP expression in all the cultures upon addition of IPTG.
+Inducible expression allows one to synchronise the start of the GFP expression in all the cultures with addition of IPTG.
 \hl{Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier} \cite{Beal2021}.\\
@@
 
-In the zeroth round we have tested two batches of designs, giving a total of 180 variants split as below:
+In the zeroth Round we have tested two batches of designs, giving a total of 180 variants split as below:
 
@@
     % \mengyan{put it in main paper?}
-    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. 
+    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e.
+equal probability of choosing any nucleotide for each position.
+\hl{This group shows what is the performance of RBSs generated randomly}.
     \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E.
 coli} genome \cite{barrick1994quantitative}.
-    % \cite{Stormo1982}.
+\hl{This group will shows what is the performance of RBSs generated randomly, but following the natural nucleotide distribution}.
     \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIRs.
@@
 The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on approximate data.
-\hl{These results show that generating a strong RBS sequence by random mutations is a non-trivial task, when the tested data set is relatively small.}\\
-
-However, starting from Round 1, where the prediction and recommendation algorithms were fed data from Round 0, the results improve significantly, with a number of sequences which perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
+% \hl{These results show that generating a strong RBS sequence by random mutations is a non-trivial task, when the tested data set is relatively small.}\\
+Starting from Round 1, where the prediction and recommendation algorithms were fed data from Round 0, the results improve significantly, with a number of sequences which perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
 In Round 2 we have observed further improvement by obtaining more sequences with TIRs similar to our benchmark sequence.
 Finally, in Round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
+
+\hl{In summary, in our study, out of 450 tested sequences around 40\% (BPS-NC, BPS-C, UNI, PPM) have been created by some kind of sequence randomisation.
+Out of these randomised sequences, only a few got close to benchmark sequences TIR and were still 20\% weaker than it is (Figure 3A).
+In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which were statistically shown not to be significantly impactful on the TIR (see Supplementary Figure S1 and, for example, work by \mbox{\textcite{Jeschek2016}}).
+More representative would be sequences from other random groups (BPS-C, UNI, PPM), from which the best one achieved only about 65\% of benchmark TIR.
+These results show that generating a strong RBS sequence by random mutations is a non-trivial task, when the tested data set is relatively small.
+Contrasted with this, non-randomized, bandit driven design batch has shown much better results, with RBSs getting close to benchmark performance and even exceeding it. }\\
 
@@
 
-In \ref{fig: Swarmplot and Quantplot}E we show a \hl{t-distributed stochastic neighbour embedding (t-SNE)} plot depicting the experimental space.
-Each RBS is located on the plot according to its distance from other RBSs as calculated by our embedding function.
+In Figure \ref{fig: Swarmplot and Quantplot}E we show a \hl{t-distributed stochastic neighbour embedding (t-SNE)} plot depicting the experimental space.
+Each RBS is located on the plot according to its distance from other RBSs as calculated by our \hl{weighted degree kernel with shift (see Section \mbox{\ref{sec: method prediction with kernel}})}.
 The RBSs recommended by Bandit groups have covered majority of the design space. 
@@
 As expected, the predictions in Round 0 were poor due to use of approximated data. 
-The predictions improved for the subsequent rounds, from R\textsuperscript{2} (coefficient of determination) of 0.067 for round 0 to R\textsuperscript{2} of 0.213 for round 3.
+The predictions improved for the subsequent rounds, from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
 Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
@@
 
-\hl{The obtained values of the coefficient of determination and Spearman correlation coefficient are influenced by a number of factors}
+\hl{The obtained values of the coefficient of determination and Spearman correlation coefficient are influenced by a number of factors.}
 \hl{Primary one of them is our choice of the exploitation to exploration balance.}
@@
 \hl{This is, however, a very useful information for future predictions as it allows us to understand the underlying space better.}
-\hl{Effect of the better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.}
-\mengyan{Learn and test only on exploitation?}\\
+\hl{Effect of the better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.}\\
 
@@
     \end{subfigure}
-    \caption{\textbf{Performance of the prediction algorithm.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each round.
+    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D are showing the performance of our prediction algorithm calculated after each round.
     Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
@@
 
-Our data set taken together can be viewed as a \hl{reliable (low coefficients of variation) and extensive (high number of entries)} library of RBS sequences for \emph{E.
+Our data set taken together can be viewed as a \hl{reliable (low coefficients of variation) and extensive (high number of entries, compared to e.g.
+widely used iGEM RBS collections} \cite{IGEM}) library of RBS sequences for \emph{E.
 coli}, some characteristics of which are shown in Table 1.\\
@@
 This means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
-This also implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.
+\hl{As long as this does not impact targeted methods like machine learning guided design}, it does imply that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.
 In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
-In such cases, a directed process should be strongly encouraged.
-This is in l \cite{Jackel2008}.\\
+In such cases, a directed process, like the one described here, should be strongly encouraged.
+\hl{This observation is in line with approaches seen in other disciplines, e.g. protein engineering, where more directed changes yield better results than wide random changes} \cite{Jackel2008}.\\
 
@@
 \hl{This change could be achieved within the Bayesian optimisation framework by altering the objective of the algorithm to e.g.
-maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objectives method \mbox{\cite{shu2020new}}.}\\
+maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objectives method \mbox{\cite{shu2020new}}.}
+% \hl{
+% While we focus our optimisation framework on the RBS part optimisation task, our framework has good potential to be generalised to multi-gene pathway design as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}.
+For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current GPR prediction model could be updated to a deep GPR approach and the current bandits algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\cite{shahriari2015taking}}.
+% }
+\\
 
@@
 For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
-We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
-This gives hope that the continuing rise in use of machine learning in synthetic biology will enable us to uncover new biologically relevant rules of design.\\
+\hl{We have achieved this despite the relatively low R\textsuperscript{2} values for our predictions.}
+\hl{Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low R\textsuperscript{2} values} \cite{Radivojevic2020, Opgenorth2019}.
+\hl{Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric}
+\cite{Schober2018,  Kang2019}.
+\hl{Finally, our predictor GPR, compared with previously described calculators using deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our Bandit algorithm to give better recommendations.}\\
 
@@
 We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
-\hl{Additionally, our method would not be appropriate for more complex cases involving multi-part devices/systems.}
-\hl{In such cases, significantly higher number of data points and deep learning techniques would be required since the number of parameters to optimise would be also significantly higher.}
+\hl{Additionally, while we focus our optimisation framework on the RBS part optimisation task, this framework has good potential to be generalised to multi-gene pathway design.}
+\hl{Since such optimisation would be a large-scale data task, the current GPR prediction model could be updated to a deep GPR approach and the current bandits algorithm could be optimised towards querying large design space to reduce computational complexity } \cite{Shahriari2016}.
 Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.

### 2021-08-31 paper/supplementary.tex

@@
 \end{align}
-For noisy test targets $\mathbf{y}_\ast$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq. (\ref{Eq: predicted variance}).
+For noisy test targets $\mathbf{y}_\ast = f(\mathbf{x}_\ast) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq.
+(\ref{Eq: predicted variance}),
+\hl{
+i.e. $\overline{\mathbf{y}}_{*} = \overline{\mathbf{f}}_{*}$, $\operatorname{cov}\left(\mathbf{y}_{*}\right) = \operatorname{cov}\left(\mathbf{f}_{*}\right) + \alpha^2 I$.
+}
+
 % We now introduce the marginal likelihood (or evidence) $p(\mathbf{y}|X)$, is which the integral of the likelihood times the prior 
@@
 
-The design of round-0 is based on the literature data \cite{jervis2018machine}.
+The design of Round-0 is based on the literature data \cite{jervis2018machine}.
 We first normalise the raw TIR to values between 0 and 1. 
@@
     SD - Shine-Dalgarno sequence.
-    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
     BN - benchmark sequences for all plates. 
@@
     \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{\textbf{Exploration-Exploitation scores illustration.} The orange points show the predicted STD and TIR (with the setting of Figure \ref{fig: Scatterplot}) for test data points (Recommendation in Round 1-3), which are projected to the diagonal line.
-The projected values are the exploration-exploitation scores shown in Figure \ref{fig: Swarmplot and Quantplot}C.}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} The orange points show the predicted STD and TIR (with the setting of Figure \ref{fig: Scatterplot}) for test data points (recommendations in Round 1-3).
+   \hl{
+   We project the orange points to the diagonal line crossing (0,1) and (1,0).
+We colour those projected points using RdBu palette, which are exploration-exploitation colours/scores shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
+   The hued red colour corresponds to points that are predicted to have high STD (uncertainty), which are recommended for \textit{exploration}; the hued blue colour corresponds to points that are predicted to have high TIR, which are recommended for \textit{exploitation}.}}
    \label{fig:exploitationvsexplorationdetail}
@@
     % \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
-    \caption{\textbf{Predictions in design pipeline in Round 1-3.} For round 1, the kernel matrix is normalised based on known RBS sequences in Round 0.
-For round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.}
+    \caption{\textbf{Predictions in design pipeline in Round 1-3.} For Round 1, the kernel matrix is normalised based on known RBS sequences in Round 0.
+For Round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.
+    \hl{
+    Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation shown in Section \mbox{\ref{supp: Normalisation of Kernel}}.
+Although the $R^2$ scores decrease a lot compared with Figure \mbox{\ref{fig: Scatterplot}}, the Spearman correlation remains similar.
+    }
+    }
     \label{fig:scatter abc1 TT.}
@@
    \includegraphics[width=0.5\textwidth]{plots/Supplementary/All_logo.pdf}
-   \caption{\textbf{Sequence logo for all tested sequences.} Compared to \ref{fig:Library characteristics}A there are no significant biases at each position.}
+   \caption{\textbf{Sequence logo for all tested sequences.} Compared to Figure \ref{fig:Library characteristics}A there are no significant biases at each position.}
    \label{fig: All_logo}
@@
     \includegraphics[scale=0.4]{plots/Supplementary/SDhist.pdf}
-   \caption{\textbf{Histogram of standard deviations.} Average raw standard deviations for each of the tested RBSs are  shown.}
+   \caption{\textbf{Histogram of standard deviations.} \hl{
+   For each RBS sequence, we tested 6 biological replicates. This plot shows the histogram of the STD of raw TIR values of the 6 replicates for each RBS sequence. }}
    \label{fig: SDhist}

### 2021-09-17 paper/main.tex
Minor edits

@@
 
-Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
-Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
+Fine control of gene expression can be achieved through engineering transcription and translation control elements.
+We consider the problem of designing a particular genetic part,
+the ribosome binding site (RBS), which
+are not yet understood at the level of detail required for reliable sequence design.
 To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
-We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
-We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
-The algorithms were able to identify RBS sequences rules conducive to high protein expression.
+We use Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
+We integrate these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
+The algorithms identified patterns of RBS sequence that are conducive to high protein expression.
 Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
@@
 The decreased result in the 0.9th quantile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
-We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown if Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
-We can see that the RBSs with high TIRs tend to come from exploitation of the design space, whereas the explorative points give relatively low TIRs, but expand our knowledge of the unknown part of the design space.\\
+\hl{Since we expect that an RBS sequence chosen at random will have low TIR
+(as shown by UNI and PPM), this implies that most of the exploration will result in low TIR.
+Our results confirm that
+RBSs with high TIRs tend to come from exploitation of the design space,
+whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge of unknown parts of the design space.}\\
 
@@
 Finally, our strong sequences showed neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS, nor any obvious secondary structures that could explain their TIRs (see Figure \ref{fig:structures}).
-This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
+This result combined with the unexpectedly bimodal nature of densities in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 
@@
 To represent RBSs and capture the similarities between them, we chose to use the
-Weighted Degree Kernel with Shift method, which fits well into our prediction method.
+Weighted Degree with Shift kernel, which fits well into our prediction method.
 In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
@@
 Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.
-The pBbB6c plasmid is available on Addgene. Other strains and plasmids are available on request from the authors; MTA will need to be negotiated between the parties.
+The pBbB6c plasmid is available on Addgene. Other strains and plasmids are available on request from the authors; a material transfer agreement will need to be negotiated between the parties.
 

### 2021-09-17 paper/methods.tex
Minor edits

@@
 % To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts:
-One type of machine learning algorithms is a prediction algorithm (\textbf{LEARN}), which helps us learn the function of TIR with respect to RBS sequences.
+One type of machine learning algorithm is a prediction algorithm (\textbf{LEARN}), which helps us learn the function that maps RBS sequences to TIR values.
 The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each round (in the sense of a single pass through the DBTL cycle) based on the predictions from LEARN.
@@
 % 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
-In Round $t$, the prediction and design are based on the results obtained in all previous rounds. 
-Our implementation of the machine learning algorithms was tested in Python 3.6 and used the scikit-learn library \cite{scikit-learn}. 
+In Round $t$, the prediction and recommendation are based on the results obtained in all previous rounds.
+Our implementation of the machine learning algorithms was tested in Python 3.6 and builds upon the scikit-learn library \cite{scikit-learn}.
 In the following paragraphs, we describe our machine learning pipeline \hl{(which is applicable to design Rounds 1-3, denoted as Bandit 1-3)} in the following order: data pre-processing, prediction and kernels and finally recommendation.
@@
 %As illustrated in Figure \ref{fig: Gaussian Process Regression Example.},
-GPR can predict both the posterior mean and standard deviation \hl{of the RBS sequences in the design space}. The posterior standard deviation represents the level of uncertainty for the prediction. 
+GPR can predict both the posterior mean and standard deviation \hl{of the TIR values corresponding to RBS sequences in the design space}.
+The posterior standard deviation represents the level of uncertainty for the prediction.
 We review the predictions of GPR in Supplementary \ref{supp: GPR}.\\
@@
 
-To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$.
+To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm,
+which comes from a class of so-called Bandit algorithms \cite{srinivas2012information}.
+\hl{The UCB algorithm takes the posterior mean and variance (of TIR as predicted by GPR),
+and balances the \textit{exploration-exploitation trade-off}.}
 On one hand, UCB exploits the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean);

### 2021-11-15 paper/main.tex
Updates from Overleaf

--- 2021-08-04 paper/main.tex
@@
 \usepackage{soul}
+\usepackage{bm}
 
@@
 
-\title{Machine Learning guided Design\\
-of Ribosome Binding Sites}
+%\title{Machine Learning guided Design\\
+%of Ribosome Binding Sites}
+
+\title{Multi-armed bandit \mengyan{(Bayesian?)} optimisation\\
+of bacterial Ribosome Binding Site}
 
@@
 
+\mengyan{We consider the problem of design gene sequences in batches to optimise the protein expression level, by replacing nucleotides from the known strong gene sequences.}
 Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
-Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
-To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
-We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
-We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
-The algorithms were able to identify RBS sequences rules conducive to high protein expression.
-Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
-Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\
+Unfortunately, RBSs are not understood at the level of finesse required for reliable design \hl {in every context}.\mengyan{what does that mean?}
+To address this problem \mengyan{make "this problem" -- challenges more clear.
+From my understanding, the problem we want to address is the search space is too large to enumerate + experiment is expensive}, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs \hl{to enable reliable part design for specific context}.\mengyan{(?) Please explain it to me.
+If the readers include people from ML field, I recommend to make it more clear.}
+We used Gaussian Process Regression for \mengyan{Learn} and the Upper Confidence Bound multi-armed Bandit algorithm for  \mengyan{Design} of genetic designs to be tested in vivo.
+We have integrated these machine learning algorithms with laboratory automation \mengyan{(Build)} and high-throughput processes for more reliable data generation \mengyan{(Test)}.
+The algorithms were able to identify RBS sequence rules conducive to high protein expression.
+Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in our specific genetic context} by up to 34\%.
+Overall, our results show that machine learning is a very \mengyan{avoid using words like "very"} powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\
 
-Keywords: machine learning, genetic part design, ribosome binding site
+Keywords: machine learning, \mengyan{optimisation? DBTL cycle?} genetic part design, ribosome binding site
 
@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
+Part design is a central problem of synthetic biology as it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
@@
 Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
-However, de-novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.\\
+However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.
+\\
 
 The biggest gap in the DBTL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
-For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009,Reis2020}.
-Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
-but they come with a number of caveats: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}.
+For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
+This may be due to number of reasons: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}.
 There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. 
-This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.\\
+This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.
+\hl{This suggests that there is a number of interactions determining the mRNA-ribosome binding which can be hard to capture in a single, universal model}.\\
 
@@
 Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
-For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoids in \emph{Escherichia coli} \cite{Jervis2019}.
+For example, Jervis \emph{et al.} used support vector machines and neural networks to optimise production of monoterpenoids in \emph{Escherichia coli} \cite{Jervis2019}.
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}. 
@@
 
-In this work, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level.
-We used Gaussian Process Regression \cite{Rasmussen2004} and Upper Confidence Bound multi-armed Bandits algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBS.
-As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that \hl{could be potentially translated to more complicated systems}.
-Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
-This way, we have built an extensive and reliable library of novel RBSs with diverse sequences. 
-We discovered new RBS sequences with very high TIRs, between 95 to 135\% of our very strong benchmark RBS.
+\hl{In genetic design practice, the designers often have to fine tune characteristics of the used parts to meet the specific metrics of their strains.
+For instance, if the goal is to produce a protein at high yield, the designer might want to increase the TIR of the RBS of the said protein.
+To do that, one could use the existing RBS predictors, but they can be unreliable in the given genetic context.}
+\hl{To address this problem}, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context}.
+Importantly, this work focuses on RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
+We used Gaussian Process Regression \cite{Rasmussen2004} and  Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBSs.
+As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that could be potentially translated to more complicated systems.
+Our overall experimental goal was to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
+\hl{This way, we discovered new RBS sequences, increasing production of our target protein compared to our very strong benchmark RBS by up to 35\%}.
 
@@
 We present our RBS-optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
-hl{In this case, the \hl{DBTL} paradigm is applied directly to the design of the RBS itself, rather than general strain optimisation.}
+In this case, the {DBTL} paradigm is applied directly to the design of the RBS part itself, rather than general strain optimisation.
 Machine learning is used in two different ways: i) we show the efficacy of the ML recommendations
@@
 In the Learn phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
-In the Design phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS sequences based on the predictions from Learn.\\
+In the Design phase, we use the Bandit algorithm to recommend new RBS sequences based on the predictions from Learn.\\
 
@@
     \includegraphics[scale=0.7]{plots/Main_Paper/flowchart.pdf}
-    \caption{\textbf{Flowchart of the machine learning-based experimental design.} The RBS design is recommended by the Upper Confidence Bound Bandit algorithm.
+    \caption{\textbf{Flowchart of the machine learning-based experimental design.} The RBS design is recommended by the Upper Confidence Bound multi-armed Bandit algorithm.
 After generating the recommendations, the RBSs are built and tested using automated laboratory methods allowing for rapid construction and testing at scale.
@@
 n is the current design round and k is the maximum number of rounds allowed by time and/or money.
-\hl{TIR(n)=TIR(n-1) denotes "lack of TIR improvement following the used optimisation function".}}
+In regards to the "Goal Met?" condition, the goal in our case was to find sequences with TIR significantly higher than benchmark, but the goal can be generalised to fit the requirements of the user.}
     \label{fig: Flowchart}
@@
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1. 
-Inducible expression allows one to synchronise the start of the GFP expression in all the cultures upon addition of IPTG.
-\hl{Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier} \cite{Beal2021}.\\
+Inducible expression allows one to synchronise the start of the GFP expression in all the cultures with addition of IPTG.
+Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier \cite{Beal2021}.\\
 
 Our template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
-This sequence is known to have a very high TIR \hl{in a similar context of expressing fluorescent protein} and is present in the pBb series of plasmids \cite{Lee2011}.
+This sequence is known to have a very high TIR in a similar context of expressing fluorescent protein and is present in the pBb series of plasmids \cite{Lee2011}.
 Since this is the sequence against which new RBS sequences will be benchmarked,
 we will refer to this sequence as the \textit{benchmark sequence} hereafter.
-\hl{Additionally, we have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it }(see Figure \ref{fig:core_vs_noncore}).
-\hl{This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.}
-\hl{For example, according to} \cite{SHULTZABERGER2001} \hl{there is a strong bias towards A and G bases in the core region of the RBS.}
-\hl{Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overall TIR of a given RBS.}
-\hl{Notably, focusing on the 6 to 8bp long core sequence is a common RBS design approach} \cite{Jeschek2016}.
-Hence, in our design, we focus on modification of the core at nucleotide positions -8 to -13 (relative to the start codon of the GFP; \hl{ this is where the consensus Shine-Dalgarno AGGAGG sequence is usually found in wild type \textit{E.
-coli}) of the RBS and keep other positions the same as the benchmark sequence, i.e.
-TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).}
+Additionally, we have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Figure \ref{fig:core_vs_noncore}).
+This hypothesis has been built based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
+For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
+Similarly, outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overall TIR of a given RBS.
+Focusing on the 6 to 8 bp core sequence is a common RBS design approach \cite{Jeschek2016}.
+Hence, in our design, we focus on modification of the core at nucleotide positions -8 to -13 (relative to the start codon of the GFP; this is where the consensus Shine-Dalgarno AGGAGG sequence is usually found in wild type \textit{E.
+coli}) of the RBS and we keep other positions the same as the benchmark sequence, i.e.
+TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).
 The total experimental (variant) space is then $4^6$ = 4096.
@@
 
-The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
-In short, this algorithm is a stochastic method of probing of the experimental space. 
+The Bandit recommendations were made using the batch Upper Confidence Bound multi-armed Bandit algorithm.
+In short, this algorithm is a stochastic method of probing the experimental space. 
 It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
@@
     UNI: Randomly generated sequences with uniform distribution. 
-    PPM: Randomly generated sequences with distribution following the \hl{Position Probability Matrix} for all natural RBS in \emph{E. coli}. 
+    PPM: Randomly generated sequences with distribution following the Position Probability Matrix for all natural RBS in \emph{E. coli}. 
     Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
     SD - Shine-Dalgarno sequence.
-    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
+    Dashed line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
     BN - benchmark sequences for all plates. 
@@
     \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
-    The area of the circle corresponds to the experimentally obtained TIR value.
-    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
+    The area of the circle corresponds to the experimentally-obtained TIR value.
+    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard; the TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
 }
@@
 
-To generate the data set that the algorithm could learn from we have decided to characterise a total of 450 RBS variants, little over 10\% of the whole experimental space. 
-To fit into our automated workflow, we have divided the 450 variants into batches of 90 split into 4 design rounds.\\
+To generate the data set from which the algorithm would learn, we decided to characterise a total of 450 RBS variants, little over 10\% of the whole experimental space. 
+To fit our automated workflow, we divided the 450 variants into batches of 90, split into 4 design rounds.\\
 
-In the zeroth round we have tested two batches of designs, giving a total of 180 variants split as below:
+In the zeroth Round we tested two batches of designs, giving a total of 180 variants split as below:
 
@@
     % \mengyan{put it in main paper?}
-    \item UNI group: 30 RBS sequences that were  uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. 
-    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally occurring RBS sequences in \emph{E.
+    \item UNI group: 30 RBS sequences that were uniformly randomised, i.e. equal probability of choosing any nucleotide for each position. This group shows the performance of RBSs generated randomly.
+    \item PPM group: 30 RBS sequences randomised based on the position probability matrix (PPM) generated from all the naturally-occurring RBS sequences in  the \emph{E.
 coli} genome \cite{barrick1994quantitative}.
-    % \cite{Stormo1982}.
-    \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with the respective TIRs.
-    This data set has been used due to perceived similarity of its goal to the one of this work - prediction of TIR based on phenotypic output.
+This group shows the performance of RBSs generated randomly, but following the natural nucleotide distribution.
+    \item Bandit-0: 60 RBS sequences recommended by our implementation of the recommendation algorithm based on a data set obtained from literature \cite{jervis2018machine}, which contains 113 non-repeated records for 56 unique RBS sequences with their respective TIRs.
+    This data set has been used due to the perceived similarity of its goal to that of this work - prediction of TIR based on phenotypic output.
 \end{itemize}
@@
 
-All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
-The best performing of these groups is the BPS-NC, which is explained by relatively small impact on the TIR of changes made outside the RBS core. 
-The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning driven, due to being trained on approximate data.
-\hl{These results show that generating a strong RBS sequence by random mutations is a non-trivial task, when the tested data set is relatively small.}\\
+All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) performed worse than our benchmark sequence in terms of TIR. 
+The best-performing group was the BPS-NC, which is explained by the relatively small impact on the TIR of changes made outside the RBS core. 
+The Bandit-0 group's performance is similar to randomly generated designs, despite being machine learning-driven, due to being trained on approximate data.
+% \hl{These results show that generating a strong RBS sequence by random mutations is a non-trivial task, when the tested data set is relatively small.}\\
+Starting from Round 1, where the prediction and recommendation algorithms were fed data from Round 0, the results improved significantly, with a number of sequences performing better than the consensus Shine-Dalgarno sequence and in one case, better than the benchmark (by 8\%).
+In Round 2 we observed further improvement by obtaining more sequences with TIRs similar to our benchmark sequence.
+Finally, in Round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
 
-However, starting from Round 1, where the prediction and recommendation algorithms were fed data from Round 0, the results improve significantly, with a number of sequences which perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
-In Round 2 we have observed further improvement by obtaining more sequences with TIRs similar to our benchmark sequence.
-Finally, in Round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\
+In summary, out of 450 tested sequences, around 40\% (BPS-NC, BPS-C, UNI, PPM) were created through some kind of sequence randomisation.
+Out of these randomised sequences, only a few got close to the benchmark sequence's TIR and were still 20\% weaker than it is (Figure 3A).
+In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (see Supplementary Figure S1 and, for example, work by \mbox{\textcite{Jeschek2016}}).
+More representative would be sequences from other random groups (BPS-C, UNI, PPM), from which the best sequence achieved only about 65\% of the benchmark TIR.
+These results show that generating a strong RBS sequence by random mutation is a non-trivial task, when the tested data set is relatively small.
+Contrasted with this, non-randomized, our Bandit-driven design batch gave much better results, with RBSs getting close to benchmark performance and even exceeding it.\\
 
@@
 The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
-The decreased result in the 0.9th quantile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
-We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown if Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+The decreased result in the 0.9th quantile compared to the maximum value for the Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
+We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
@@
 KDE plots have been overlaid to depict the calculated density for each group.
-The increase in prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3, constituting the bulk of results in the $>0.8$ TIR ratio bins.
+The increase in prevalence of later Bandit groups in the higher bins is evident, especially for Bandit 2 and 3, constituting the bulk of results in the $>0.8$ TIR ratio bins.
 Notably, the distributions calculated for all the groups are bimodal - we discuss the possible reasons for that later in the text.\\
 
-In \ref{fig: Swarmplot and Quantplot}E we show a \hl{t-distributed stochastic neighbour embedding (t-SNE)} plot depicting the experimental space.
-Each RBS is located on the plot according to its distance from other RBSs as calculated by our embedding function.
-The RBSs recommended by Bandit groups have covered majority of the design space. 
+In Figure \ref{fig: Swarmplot and Quantplot}E we show a t-distributed stochastic neighbour embedding (t-SNE) \mbox{\cite{tsne2008}} plot depicting the experimental space.
+Each RBS is located on the plot according to its distance from other RBSs as calculated by our weighted degree kernel with shift (see Section \mbox{\ref{sec: method prediction with kernel}}).
+The RBSs recommended by Bandit groups have covered the majority of the design space. 
 Additionally, a number of clusters were especially targeted by our recommendation algorithm.
@@
 Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round. 
-As expected, the predictions in Round 0 were poor due to use of approximated data. 
-The predictions improved for the subsequent rounds, from R\textsuperscript{2} (coefficient of determination) of 0.067 for round 0 to R\textsuperscript{2} of 0.213 for round 3.
+As expected, the predictions in Round 0 were poor due to use of approximate data. 
+The predictions improved for the subsequent rounds, from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
 Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
-Figure \ref{fig:scatter abc1 TT.} shows performance of predictions for individual rounds.\\
+Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.\\
 
-\hl{The obtained values of the coefficient of determination and Spearman correlation coefficient are influenced by a number of factors}
-\hl{Primary one of them is our choice of the exploitation to exploration balance.}
-\hl{In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having a real mean different from what was predicted, thus decreasing the R\textsuperscript{2} values.}
-\hl{This is, however, a very useful information for future predictions as it allows us to understand the underlying space better.}
-\hl{Effect of the better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.}
-\mengyan{Learn and test only on exploitation?}\\
+The obtained values of the coefficient of determination and Spearman correlation coefficient are influenced by a number of factors.
+The primary one is our choice of the exploration-exploitation balance.
+In each round, we select a number of data points for exploration, which means that when tested, they have a high chance of having a experimentally-obtained mean different from what was predicted, thus decreasing the R\textsuperscript{2} values.
+However, this is very useful information for future predictions as it allows us to understand the underlying space better.
+The effect of better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.\\
 
@@
     \end{subfigure}
-    \caption{\textbf{Performance of the prediction algorithm.} The scatter plots A-D are showing the performance of our prediction algorithm calculated after each round.
-    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in the Figure \ref{fig: Swarmplot and Quantplot}.
-    The values of $R^2$ and Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
-    The p-value here is for null hypothesis stating that two sets of data are uncorrelated.
+    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of our prediction algorithm calculated after each round.
+    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
+    The $R^2$ value and Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
+    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
     }
@@
 
-Our data set taken together can be viewed as a \hl{reliable (low coefficients of variation) and extensive (high number of entries)} library of RBS sequences for \emph{E.
+% Taken together, our data can be viewed as a reliable (low coefficients of variation) and extensive (high number of entries compared with widely used iGEM RBS collections, for example \cite{IGEM}) library of RBS sequences for \emph{E.
 coli}, some characteristics of which are shown in Table 1.\\
 
-Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows logo generated for all tested sequences).
-It is generally understood that guanine rich sequences are promoting strong transcription.
+\hl{We present some important characteristics of the tested RBSs in Table 1}.
+Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows the logo generated for all tested sequences).
+It is generally understood that guanine-rich sequences promote strong transcription.
 This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
-This result combined with the Bandits' algorithm bias towards the G rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as the ones with high TIR probability.\\
+This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with high TIR probability.\\
 
@@
      \caption{\textbf{Characteristics of strong RBSs.} A) Sequence logo calculated for the Top 30 tested sequences.
-B) Heatmap showing what edit (Hamming) distance is required for positive change in TIR for RBS with high and medium TIR.
-The temperature scale shows the difference between a given RBS on y axis and the RBS with strongest TIR at the given distance.
-Every second RBS is labelled for increased legibility.
-}
+B) Heatmap showing the edit (Hamming) distance required for positive change in TIR for RBSs with high and medium TIRs.
+The temperature scale shows the difference between a given RBS on the y-axis and the RBS with the strongest TIR at the given distance.
+Every second RBS is labelled for increased legibility.}
      \label{fig:Library characteristics}
@@
 Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
-\begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
+\begin{tabular}[c]{@{}c@{}}Efficiency of Bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
 Raw TIR range                                                                                     &      [4.93, 105.38]               \\ \hline
@@
 Left table presents some of the characteristics of our library.
-Right table presents 10 RBS sequences with their corresponding TIR ratios, the first 9 being the strongest sequences including the benchmark sequence (AGGAGA) and the last being the Shine-Dalgarno sequence (AGGAGG).
+Right table presents 10 RBS sequences with their corresponding TIR ratios; the first 9 are the strongest sequences including the benchmark sequence (AGGAGA) and the last is the Shine-Dalgarno sequence (AGGAGG).
 CV is coefficient of variation (standard deviation (STD) of a sample divided by its mean; see details in Figures \ref{fig: SDhist} and \ref{fig:variation of biological replicates}).
-Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by the highest TIR found using random sequence generation. }
+Efficiency of the Bandit design is calculated by dividing the highest TIR found using machine learning by the highest TIR found using random sequence generation.}
 \end{table}
@@
 We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
-Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
-For RBS with high TIR ($>1$), the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
+Figure \ref{fig:Library characteristics}B shows the edit distance is required for positive change in TIR for an RBS with TIR $>0.75$.
+For RBSs with high TIRs ($>1$), the minimum distance that is required to increase the TIR is 2, with edit distances between 2 and 5 giving similar results.
 For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
-This means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
-This also implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.
-In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
-In such cases, a directed process should be strongly encouraged.
-This is in l \cite{Jackel2008}.\\
+This means that as the TIRs of examined RBSs increase, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
+As long as this does not impact targeted methods like machine learning-guided design, it does imply that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
+In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
+In such cases, a directed process, like the one described here, should be strongly encouraged.
+This observation is in line with approaches seen in other disciplines, e.g. protein engineering, where more directed changes yield better results than wide random changes \cite{Jackel2008}.\\
 
-Finally, our strong sequences showed neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS, nor any obvious secondary structures that could explain their TIRs (see Figure \ref{fig:structures}).
-This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
+\hl{Finally, while our strong sequences showed some affinity towards the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs} (see Figure \ref{fig:structures}).
+This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion, based on the previously reported literature \cite{Saito2020,EspahBorujeni2016}, that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
 
@@
 
-In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
-We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
-To represent RBSs and capture the similarities between them, we chose to use the
-Weighted Degree Kernel with Shift method, which fits well into our prediction method.
+In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBSs.
+We have used Gaussian Process regression to predict the TIR function to be optimised and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested.
+To represent RBSs and capture the similarities between them, we chose to use the Weighted Degree Kernel with Shift method, which fits well into our prediction method.
 In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
-Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\
+By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in this context} by up to 34\%.
+Furthermore, we have generated an extensive library of diverse RBSs that can be used \hl{as a basis} in future studies.\\
 
-\hl{In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.}
-\hl{This approach may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.}
-\hl{This change could be achieved within the Bayesian optimisation framework by altering the objective of the algorithm to e.g.
-maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objectives method \mbox{\cite{shu2020new}}.}\\
+In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.
+This approach may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
+This change could be achieved within the Bayesian optimisation framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objective method, for example \mbox{\cite{shu2020new}}.\\
+% \hl{
+% While we focus our optimisation framework on the RBS optimisation task, our framework has good potential to be generalised to multi-gene pathway design, as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}.
+For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process Regression (GPR) prediction model could be updated to a deep GPR approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\cite{shahriari2015taking}}.
+% }
+
+It is important to make a distinction between works focusing on the part itself like \mbox{\textcite{Hollerer2020}} and works where the part is optimised with a wider goal of strain optimisation like in \mbox{\textcite{Jervis2019}}.
+The part-centric optimisation, like the one considered in this paper, is an important task in synthetic biology because metabolic pathways are made up of individual parts and understanding their performance in separation allows for better strain design.
+While we focus our optimisation framework on the RBS optimisation task, this framework has good potential to be generalised to multi-gene pathway design as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}.
+For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\textcite{Shahriari2016}}.
+\\
 
@@
 For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
-We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
-This gives hope that the continuing rise in use of machine learning in synthetic biology will enable us to uncover new biologically relevant rules of design.\\
+We have achieved this despite the relatively low R\textsuperscript{2} values for our predictions.
+Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low R\textsuperscript{2} values \cite{Radivojevic2020, Opgenorth2019}.
+Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric
+\cite{Schober2018,  Kang2019}.
+Finally, our predictor Gaussian Process regression model, compared with previously described calculators using a deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our Bandit algorithm to give better recommendations.\\
 
@@
 We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
-Synthetic biology is promoting standardised and normalised testing in biology which naturally pairs with machine learning which can leverage the high quality biological data sets generated when the correct design rules are observed.
+Synthetic biology is promoting standardised and normalised testing in biology and naturally pairs with machine learning, which can leverage the high quality biological data sets generated when the correct design rules are observed.
 The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
@@
 In this study we have limited the number of design rounds to four.
-There was a number of reasons for this, including time and money limitations, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
-There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however this has to be put in the context of the limited resources.
+There were a number of reasons for this, including limitations on time and money, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
+There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however this has to be put in the context of limited resources.
 For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
-\hl{Compared to solutions like the one reported by Hollerer} \cite{Hollerer2020}, \hl{our solution can be used when a high volume method for data-generation is not available, while still providing the required results (optimised part).} \\
+Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available, while still providing the required results (optimised part).
+\\
 
@@
 Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
-We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the obtained results.
-Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurement? 
+We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
+Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurements? 
 Thirdly, the generalisability of the method is unknown.
+\hl{Since the model has been taught on data relating to this specific genetic context its applicability for other context will be limited and new data will have to be obtained.}
 We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
-\hl{Additionally, our method would not be appropriate for more complex cases involving multi-part devices/systems.}
-\hl{In such cases, significantly higher number of data points and deep learning techniques would be required since the number of parameters to optimise would be also significantly higher.}
+% \hl{Additionally, while we focus our optimisation framework on the RBS part optimisation task, this framework has good potential to be generalised to multi-gene pathway design.}
+% \hl{Since such optimisation would be a large-scale data task, the current GPR prediction model could be updated to a deep GPR approach and the current bandits algorithm could be optimised towards querying large design space to reduce computational complexity } \cite{Shahriari2016}.
 Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
@@
 
-All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/Solaris/tree/master/synbio_rbs} .
-All the processed and raw data is included in the repository.
-Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.
-The pBbB6c plasmid is available on Addgene. Other strains and plasmids are available on request from the authors; MTA will need to be negotiated between the parties.
+All code and data required to reproduce the results are available at Github: \url{https://github.com/mholowko/Solaris/tree/master/synbio_rbs}.
+All the processed and raw data are included in the repository.
+Sequences of plasmids and oligos and assembly reports used in this study are available in the supplementary information as a separate file.
+The pBbB6c plasmid is available on Addgene. Other strains and plasmids are available on request from the authors; MTAs will need to be negotiated between the parties.
 

### 2021-11-15 paper/methods.tex
Updates from Overleaf

--- 2021-08-31 paper/methods.tex
@@
 The pBbB6c-GFP plasmid has been used for all our designs.
-This plasmid \hl{contains the GFP mut3b CDS, expression of which can be induced by the addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG)}.
-The original RBS for the GFP CDS was replaced \hl{using a combination} of PCR and isothermal assembly.
-\hl{Primer sequences and the assembly strategy were generated using the Teselagen DESIGN software (Teselagen Biotechnology)}.
+This plasmid contains the GFP mut3b CDS, expression of which can be induced by the addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG).
+The original RBS for the GFP CDS was replaced using a combination of PCR and isothermal assembly.
+Primer sequences and the assembly strategy were generated using the Teselagen DESIGN software (Teselagen Biotechnology).
 
@@
 PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, catalogue no. M0492L).
-20 $\mu$L reactions were prepared by dispensing \hl{1 $\mu$L of each 10 $\mu$M reverse primer into the wells} of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
-A mastermix consisting of polymerase premix, plasmid DNA template \hl{(pBbB6c, 5-10 ng per reaction)}, and the single 10 $\mu$M forward primer was prepared by and dispensed \hl{using the FeliX liquid handler or electronic multi-channel pipette}.
+20 $\mu$L reactions were prepared by dispensing 1 $\mu$L of each 10 $\mu$M reverse primer into the wells of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
+A mastermix consisting of polymerase premix, plasmid DNA template (pBbB6c, 5-10 ng per reaction), and the single 10 $\mu$M forward primer was prepared and dispensed using the FeliX liquid handler or electronic multi-channel pipette.
 Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
@@
 ZAG-110-5000; ZAG-130-5000).
-\hl{ZAG sample plates were prepared using the Perkin Elmer Sciclone G3 liquid handler.}
-ProSize Data Analysis Software (Agilent Technologies) was used to generated gel images from the sample chromatograms and sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
+ZAG sample plates were prepared using the Perkin Elmer Sciclone G3 liquid handler.
+ProSize Data Analysis Software (Agilent Technologies) was used to generate gel images from the sample chromatograms, and amplicon sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
 
@@
 Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no. E2621L).
-Reactions consisting of \hl{approximately equal amounts} of the common fragment and the variable fragment were prepared \hl{using the FeliX liquid handler or electronic multi-channel pipette}, to a final volume of 5 or 10 $\mu$L.
+Reactions consisting of approximately equal amounts of the common fragment and the variable fragment were prepared using the FeliX liquid handler or electronic multi-channel pipette, to a final volume of 5 or 10 $\mu$L.
 Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
-Finally, samples were \hl{incubated with of 50 nL of DpnI (NEB, catalogue no. R0176S) at 37$^{\circ}$C for 90 minutes to remove any residual template DNA.}
+Finally, samples were incubated with of 50 nL of DpnI (NEB, catalogue no. R0176S) at 37$^{\circ}$C for 90 minutes to degrade any residual template DNA.
 
@@
 Plates of cells were thawed on a -20$^{\circ}$C cold block before 3 $\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
-Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 \hl{(12 x 8)} grid on Omnitrays containing LB (BD, catalogue no.
+Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 (12 x 8) grid on Omnitrays containing LB (BD, catalogue no.
 244610) and 15 g/L agar (Sigma-Aldrich, catalogue no.
@@
 C1919).
-Plates were incubated overnight at 37$^{\circ}$C. \hl{Cells were plated using the CyBio FeliX liquid handler.}
+Plates were incubated overnight at 37$^{\circ}$C. Cells were plated using the CyBio FeliX liquid handler.
 
@@
 Each selected colony was used to inoculate 1 mL of selective medium in a 2 mL square well 96 plate.
-They were then cultured overnight in 37$^{\circ}$C with shaking (~300 rpm).
+They were then cultured overnight at 37$^{\circ}$C with shaking (300 rpm).
 
@@
 100 $\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, catalogue no.
-GA010) and 100 $\mu$L of overnight culture were combined in the wells of a 96 deep (2 mL) round well plate \hl{using the CyBio FeliX liquid handler or electronic multichannel pipette}
+GA010) and 100 $\mu$L of overnight culture were combined in the wells of a 96 deep (2 mL) round well plate using the CyBio FeliX liquid handler or electronic multichannel pipette
 They were then sealed with a 96-well silicone sealing mat and transferred to a -80$^{\circ}$C freezer.
@@
 \textbf{Test strain culture.}
-Overnight cultures \hl{(total of six replicates for each batch)} were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, \hl{using the CyBio FeliX liquid handler or electronic multichannel pipette.}
-Cultures were incubated at 37$^{\circ}$C with shaking (~300 rpm) for ~17 hours. 
-The following morning, 20 $\mu$L of overnight cultures were added to 980 $\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2 mL round well 96 plate. 
-\hl{After 90 minutes, cultures were induced with IPTG (Sigma-Aldrich, catalogue no.
+Overnight cultures (six replicates for each batch) were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with 2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, using the CyBio FeliX liquid handler or electronic multichannel pipette.
+Cultures were incubated at 37$^{\circ}$C with shaking (300 rpm) for 17 hours. 
+The following morning, 20 $\mu$L of each overnight culture was added to 980 $\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in a 96 deep (2 mL) round well plate.
+After 90 minutes, cultures were induced with IPTG (Sigma-Aldrich, catalogue no.
 I5502) to a final concentration of 0.5 mM.
-This was done by transferring 1.0 $\mu$L of 0.1M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo acoustic liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multichannel pipette.}
+This was done by transferring 1.0 $\mu$L of 0.1 M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo acoustic liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multichannel pipette.
 
@@
 The plates were monitored in the Cytation 5 microplate reader immediately after the addition of IPTG.
-Cytation 5 acquisition and incubation/shaking settings were as follows: length of run: \hl{8 hours}; interval: 10 min; continuous orbital shake at 237 rpm and slow orbital speed; excitation wavelength: 490/10 nm; emission wavelength: 515/10 nm; bottom read; gain: 60; read height: 7 mm; read speed: Sweep.
+Cytation 5 acquisition and incubation/shaking settings were as follows: length of run: 8 hours; interval: 10 min; continuous orbital shake at 237 rpm and slow orbital speed; excitation wavelength: 490/10 nm; emission wavelength: 515/10 nm; bottom read; gain: 60; read height: 7 mm; read speed: Sweep.
 
@@
 % To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
-One type of machine learning algorithms is a prediction algorithm (\textbf{LEARN}), which helps us learn the function of TIR with respect to RBS sequences.
+One type of machine learning algorithm is a prediction algorithm (\textbf{LEARN}), which helps us learn the function of TIR with respect to RBS sequence.
 The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each round (in the sense of a single pass through the DBTL cycle) based on the predictions from LEARN.
@@
 % 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
-In Round $t$, the prediction and design are based on the results obtained in all previous rounds. 
+In Round $t$, prediction and design are based on the results obtained in all previous rounds. 
 Our implementation of the machine learning algorithms was tested in Python 3.6 and used the scikit-learn library \cite{scikit-learn}. 
-In the following paragraphs, we describe our machine learning pipeline \hl{(which is applicable to design Rounds 1-3, denoted as Bandit 1-3)} in the following order: data pre-processing, prediction and kernels and finally recommendation.
-\hl{The pipeline for our 0th Round of machine learning designs, denoted as Bandit-0, is reported in Supplementary} \ref{sec: ML design pipeline}.
+In the following paragraphs, we describe our machine learning pipeline (which is applicable to design Rounds 1-3, denoted as Bandit 1-3) in the following order: data pre-processing, prediction and kernels, and finally recommendation.
+The pipeline for our 0th Round of machine learning designs, denoted as Bandit-0, is reported in Supplementary \ref{sec: ML design pipeline}.
 
@@
 
-\hl{For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as an averaged sum of GFP fluorescence divided by Optical Density at 600nm (OD600) of culture over time (calculated using values from all the measured data points), following the formula:}
+For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as an averaged sum of GFP fluorescence divided by Optical Density at 600 nm (OD600) of the culture over time (calculated using values from all the measured data points), following the formula:
 
@@
 
-\hl{where ${t_{min}}$ is time at which $\frac{GFP}{OD_{600}}$ value is at its minimum denoting start of the log growth phase, ${t_{max}}$ is time at which the logarithmic phase ends (on average, for our samples that was at 4h from ${t_{min}}$ and so that was the time point used for TIR calculations for all samples), ${\Delta t}$ is the difference between ${t_{max}}$ and ${t_{min}}$.}\\
+where ${t_{min}}$ is the time at which the $\frac{GFP}{OD_{600}}$ value is at its minimum denoting the start of the log growth phase, ${t_{max}}$ is the time at which the logarithmic phase ends (on average, for our samples that was 4 hours from ${t_{min}}$ so that was the time point used for TIR calculations for all samples), ${\Delta t}$ is the difference between ${t_{max}}$ and ${t_{min}}$.\\
 
@@
 For label pre-processing, we first adjust the TIR values in each round using the round-wise reference values.
-This reference values are the TIR of the benchmark sequence that is run in triplicate in each round.
+The reference value is the TIR of the benchmark sequence that is run in triplicate in each round.
 Specifically, in Round $t$, we subtract the TIR mean of the benchmark RBS measured in Round $t$ from all TIR values measured in the same round, for each replicate separately. 
@@
 
-To find RBS sequences with the highest possible TIR score after a total number of rounds $N$,  we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ \hl{is the TIR score of the 6-base core sequence considered in its wider 20-base full RBS sequence $\mathbf{x} \in \{A,C,G,T\}^{20}$}.
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$, we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ is the TIR score of the 6-base core sequence \hl{ of the RBS} $\mathbf{x} \in \{A,C,G,T\}^{6}$.
 In each Round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and variance.
+$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and \hl{standard deviation}.
 \\
@@
 
-For regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004,srinivas2012information, romero_navigating_2013}.
+For the regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004,srinivas2012information, romero_navigating_2013}.
 We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
 %As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
-GPR can predict both the posterior mean and standard deviation \hl{of the RBS sequences in the design space}. The posterior standard deviation represents the level of uncertainty for the prediction. 
+GPR can predict both the posterior mean and standard deviation of the RBS sequences in the design space. The posterior standard deviation represents the level of uncertainty for the prediction. 
 We review the predictions of GPR in Supplementary \ref{supp: GPR}.\\
@@
 The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
-For Bandit designs in Round 0, since we only had access to limited number of data points from literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+For Bandit designs in Round 0, since we only had access to a limited number of data points from the literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6 bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7 bp flanking sequences both upstream and downstream of the core sequence.
 To represent the RBS sequences in subsequent rounds, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds,Ben-Hur2008} to specify the kernel function of $GP$.  
@@
 $S(l)$ determines the maximum shift at position $l$. 
-Furthermore, we normalise the kernel with centring and unit norm in terms of all sequences in the design space. 
+Furthermore, we normalise the kernel with centring and unit-norm in terms of all sequences in the design space. 
 % In our design, the length of RBS sequence $L = 20$.
@@
 
-\subsubsection{Recommendation: Batch Upper Confidence Bound Bandit}
+\subsubsection{Recommendation: Upper Confidence Bound multi-armed Bandit algorithm}
 
-To recommend the RBS sequences to query in next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm \cite{srinivas2012information}, which provides  $\textit{exploitation-exploration balance}$.
+To recommend the RBS sequences to query in the next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm
+\cite{lattimore2020bandit}
+% \cite{srinivas2012information}
+, which provides  $\textit{exploration-exploitation balance}$.
 On one hand, UCB exploits the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
 on the other hand, UCB explores the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
-More precisely, UCB algorithm selects RBS sequence $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at Round $t$, i.e.
+More precisely, the UCB algorithm selects RBS sequences $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at Round $t$, i.e.
 \begin{align}
@@
 \end{align}
-where $\beta_t$ is a hyperparmeter balancing the exploitation and exploration, 
+where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
 $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at Round $t$ for the sequence $\mathbf{x}_i$.
@@
 Since GPR assumes similar sequences would have similar labels (e.g.
-by knowing $x=2$ we can gain information about $x=2.5$ as well), we prefer to not to waste time and money on labelling sequences with high similarities in the same batch.
+by knowing $x=2$ we can gain information about $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
 To counter this, we use a batch recommendation strategy that is designed to avoid recommending such highly similar sequences in the same batch, described below. \\
@@
 
-A key property of Gaussian Process regression is that the predicted standard deviation depends only on \hl{features}, not on the \hl{labels}. 
+A key property of Gaussian Process regression is that the predicted standard deviation depends only on features, not on the labels. 
 % We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
@@
 One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
-The strategy here is to recommend \hl{RBS sequences one-by-one by sequentially adding the newly recommended RBSs predicted means (without testing them) into the training data and updating UCB scores.}\\
+The strategy here is to recommend RBS sequences one-by-one by sequentially adding the newly recommended RBSs' predicted means (without testing them) into the training data and updating UCB scores.\\
 
@@
 As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
-Then we add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as label (note it is not the true label), and update the predicted standard deviation and then we finally update the UCB scores.
+We then add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as the label (note it is not the true label), and update the predicted standard deviation, then we finally update the UCB scores.
 The second data point is then recommended based on the new UCB scores.   

### 2021-11-15 paper/main.tex

@@
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
-We can see that the RBSs with high TIRs tend to come from exploitation of the design space, whereas the explorative points give relatively low TIRs, but expand our knowledge of the unknown part of the design space.\\
+\hl{Since we expect that an RBS sequence chosen at random will have low TIR
+(as shown by UNI and PPM), this implies that most of the exploration will result in low TIR.
+Our results confirm that
+RBSs with high TIRs tend to come from exploitation of the design space,
+whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge of unknown parts of the design space.}\\
 
@@
 For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
+
 This means that as the TIRs of examined RBSs increase, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
@@
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 

### 2021-11-15 paper/supplementary.tex

@@
 
-A Gaussian process is specified by its mean function and \hl{covariance} function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(\mu(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
+A Gaussian process is specified by its mean function and covariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(\mu(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
 We consider the case where the observations are noisy, i.e.
-$(X, \mathbf{y}) = \{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$.
-The Gaussian noise is independent identically distributed, and the \hl{covariance of the prior on the} noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
+$(X, \mathbf{y}) = \{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where label $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$.
+The Gaussian noise is independent identically distributed, and the covariance of the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
 where $\delta_{pq}$ is a Kronecker delta which is one if $p = q$ and zero otherwise.
-It is equivalent to a diagonal matrix $\alpha^2 I$ \hl{added} on the kernel matrix evaluated on the training points.
+It is equivalent to a diagonal matrix $\alpha^2 I$ added on the kernel matrix evaluated on the training points.
 
@@
 \begin{align}
-    \left[\begin{array}{l}\mathbf{y} \\ \mathbf{f}_{*}\end{array}\right] \sim \mathcal{N}\left(\mathbf{0},\left[\begin{array}{cc}K(X, X)+\alpha^{2} I & K\left(X, X_{*}\right) \\ K\left(X_{*}, X\right) & K\left(X_{*}, X_{*}\right)\end{array}\right]\right)
+    \left[\begin{array}{l}\mathbf{y} \\ \mathbf{f}_{*}\end{array}\right] \sim \mathcal{N}\left(
+        \left[\begin{array}{cc}
+        \bm{\mu}  \\
+        \bm{\mu}_\ast
+        \end{array}\right],
+    \left[\begin{array}{cc}K(X, X)+\alpha^{2} I & K\left(X, X_{*}\right) \\ K\left(X_{*}, X\right) & K\left(X_{*}, X_{*}\right)\end{array}\right]\right)
 \end{align}
-where $K(X, X_\ast)$ denotes the $n \times n_\ast$ covariance/Kernel matrix evaluated at all pairs of training and testing points, similarly for other kernel matrices.
+where
+$\bm{\mu}, \bm{\mu}_\ast$ are mean vectors for $\mathbf{y}, \mathbf{f}_{*}$ respectively,
+$K(X, X_\ast)$ denotes the $n \times n_\ast$ covariance/Kernel matrix evaluated at all pairs of training and testing points, similarly for other kernel matrices.
 Then the posterior of the test points (i.e.
@@
 \begin{align}
-   \overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} \mathbf{y} \\
+   \overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=
+   \bm{\mu}_\ast + K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} (\mathbf{y} - \bm{\mu})\\
    \label{Eq: predicted variance}
@@
 (\ref{Eq: predicted variance}),
-\hl{
 i.e. $\overline{\mathbf{y}}_{*} = \overline{\mathbf{f}}_{*}$, $\operatorname{cov}\left(\mathbf{y}_{*}\right) = \operatorname{cov}\left(\mathbf{f}_{*}\right) + \alpha^2 I$.
-}
 
@@
      where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
-     \hl{
      In our case, $\Sigma = \{A,C,G,T\}$ and $|\Sigma| = 4$.
-     }
     $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\mathbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
-    \hl{
     For example, if $\mathbf{x} = ACACAG$ and $\ell = 3$, then $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps $\mathbf{x}$ into $4^3$ dimensional feature space, where each dimension corresponds to $AAA, AAC, AAT, AAG, ACA, ..., TTT$.
 Among those, the dimension corresponds to $ACA, CAC, CAG$ are 2,1,1 and others are 0.
-    }
     Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_\ell^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
@@
 where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$,
-\hl{with $d$ as the dimension of features.
+with $d$ as the dimension of features.
 One example of $\phi$ can be $\phi_\ell^{spec}(\mathbf{x})$ in the spectrum kernel.
 Recall the corresponding kernel can be defined as $k(\mathbf{x}, \mathbf{x}^\prime) = \phi(\mathbf{x})^T \phi(\mathbf{x}).$
-}
+
 Assume there is total of $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
@@
     \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} 
-    \hl{The core region is highlighted in light green and the lines are rolling averages (in this case mean of the 5 points centred over the indicated position) for each base.}
+    The core region is highlighted in light green and the lines are rolling averages (in this case mean of the 5 points centred over the indicated position) for each base.
     The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base, which is colour coded.
-    \hl{Since only changes from the original base at each position are shown, the dotted lines start and end at different positions depending on how many changes to the respective base are at positions adjacent to the indicated one.}
+    Since only changes from the original base at each position are shown, the dotted lines start and end at different positions depending on how many changes to the respective base are at positions adjacent to the indicated one.
     The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.}
@@
     \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{\textbf{Exploration-Exploitation scores illustration.} The orange points show the predicted STD and TIR (with the setting of Figure \ref{fig: Scatterplot}) for test data points (recommendations in Round 1-3).
-   \hl{
-   We project the orange points to the diagonal line crossing (0,1) and (1,0).
-We colour those projected points using RdBu palette, which are exploration-exploitation colours/scores shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
-   The hued red colour corresponds to points that are predicted to have high STD (uncertainty), which are recommended for \textit{exploration}; the hued blue colour corresponds to points that are predicted to have high TIR, which are recommended for \textit{exploitation}.}}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}).
+   For a given round all the tested sequences are then projected along the diagonal line crossing (0,1) and (1,0) according to their predicted STD and mean TIR (the higher the predicted mean the closer the point is to the (1,0) coordinate and the higher the predicted STD the closer the point is to the (0,1) coordinate).
+   Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
+   We do this to transform our predictions into a common basis (where the basis is the diagonal) and so by calculating where a given point falls on the diagonal we can show if its mean TIR or STD is higher.
+   Finally, we colour those projected points using the RdBu spectrum where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue, and these correspond to the points shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
+   }
    \label{fig:exploitationvsexplorationdetail}
@@
     For Round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.
-    \hl{
-    Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation shown in Section \mbox{\ref{supp: Normalisation of Kernel}}.
-Although the $R^2$ scores decrease a lot compared with Figure \mbox{\ref{fig: Scatterplot}}, the Spearman correlation remains similar.
-    }
+    Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation described in Section \mbox{\ref{supp: Normalisation of Kernel}}.
+Although the $R^2$ scores decrease considerably compared with Figure \mbox{\ref{fig: Scatterplot}}, the Spearman correlation remains similar.
     }
@@
     \includegraphics[scale=0.4]{plots/Supplementary/SDhist.pdf}
-   \caption{\textbf{Histogram of standard deviations.} \hl{
-   For each RBS sequence, we tested 6 biological replicates. This plot shows the histogram of the STD of raw TIR values of the 6 replicates for each RBS sequence. }}
+   \caption{\textbf{Histogram of standard deviations.}
+   For each RBS sequence, we tested 6 biological replicates. This plot shows the histogram of the STD of raw TIR values of the 6 replicates for each RBS sequence. }
    \label{fig: SDhist}

### 2021-11-16 paper/main.tex
mengyan comments - intro

@@
 \usepackage{bm}
+\usepackage{lineno}
 
@@
 
+\linenumbers
 \topmargin -.5in
@@
 
-\author[1,2,4]{Zhang Mengyan}
-\author[3]{Holowko Maciej Bartosz}
-\author[3]{Hayman Zumpe Huw}
-\author[1,2,4,*]{Ong Cheng Soon}
+\author[1,2,3]{Zhang Mengyan}
+\author[4,5]{Holowko Maciej Bartosz}
+\author[4,5]{Hayman Zumpe Huw}
+\author[1,2,3,*]{Ong Cheng Soon}
 \affil[1]{Machine Learning and Artificial Intelligence Future Science Platform, CSIRO}
 \affil[2]{Department of Computer Science, Australian National University}
-\affil[3]{Synthetic Biology Future Science Platform, CSIRO}
-\affil[4]{Data61, CSIRO}
+\affil[3]{Data61, CSIRO}
+\affil[4]{Synthetic Biology Future Science Platform, CSIRO}
+\affil[5]{Land and Water, CSIRO}
 \affil[*]{email: cheng-soon.ong@data61.csiro.au}
@@
 
-\mengyan{We consider the problem of design gene sequences in batches to optimise the protein expression level, by replacing nucleotides from the known strong gene sequences.}
-Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
-Unfortunately, RBSs are not understood at the level of finesse required for reliable design \hl {in every context}.\mengyan{what does that mean?}
-To address this problem \mengyan{make "this problem" -- challenges more clear.
-From my understanding, the problem we want to address is the search space is too large to enumerate + experiment is expensive}, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs \hl{to enable reliable part design for specific context}.\mengyan{(?) Please explain it to me.
-If the readers include people from ML field, I recommend to make it more clear.}
-We used Gaussian Process Regression for \mengyan{Learn} and the Upper Confidence Bound multi-armed Bandit algorithm for  \mengyan{Design} of genetic designs to be tested in vivo.
-We have integrated these machine learning algorithms with laboratory automation \mengyan{(Build)} and high-throughput processes for more reliable data generation \mengyan{(Test)}.
+Optimisation of gene expression levels is an essential part of the organism design process.
+Fine control of this process can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
+Unfortunately, design of specific genetic parts can still be challenging due to lack of reliable design methods.
+\mengyan{I still feel "reliable design methods" is too vague. e.g. what does reliable mean? why lack such methods? what are the challenging parts exactly?}
+To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed.
+We used Gaussian Process Regression for the Learn part of cycle and the Upper Confidence Bound multi-armed Bandit algorithm for the Design of genetic designs to be tested in vivo.
+We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for a more reliable data generation .
 The algorithms were able to identify RBS sequence rules conducive to high protein expression.
-Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in our specific genetic context} by up to 34\%.
-Overall, our results show that machine learning is a very \mengyan{avoid using words like "very"} powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\
+Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
+Overall, our results show that machine learning is a powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\
 
-Keywords: machine learning, \mengyan{optimisation? DBTL cycle?} genetic part design, ribosome binding site
+Keywords: machine learning, optimisation, genetic part design, ribosome binding site
 
@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-Part design is a central problem of synthetic biology as it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
+Part design \mengyan{what is part design?} is a central problem of synthetic biology as it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
+\mengyan{make the previous sentence shorter?}
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
@@
 
-The biggest gap in the DBTL cycle, at present, is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
+The biggest gap in the DBTL cycle is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
+\mengyan{what does "translate" mean?
+it feels like recommendation, but in the following we are discussing predictions.}
 For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
-This may be due to number of reasons: i) they rely on calculations of free energies which can be difficult to estimate with high precision, ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}.
+This may be due to number of reasons: 
+i) they rely on calculations of free energies which can be difficult to estimate with high precision, 
+ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and
+iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}. 
 There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. 
@@
 
-Currently, the volume of experimental data in synthetic biology is increasing exponentially. \cite{Freemont2019}. 
+The volume of experimental data in synthetic biology is increasing exponentially. \cite{Freemont2019}. 
+\mengyan{this is not the case for our study. The amount of data we have is quite small in ML view. And the small amount of data motivates the DBTL approach. }
 Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
@@
 
-\hl{In genetic design practice, the designers often have to fine tune characteristics of the used parts to meet the specific metrics of their strains.
-For instance, if the goal is to produce a protein at high yield, the designer might want to increase the TIR of the RBS of the said protein.
-To do that, one could use the existing RBS predictors, but they can be unreliable in the given genetic context.}
+\hl{In genetic design practice, the designers often have to fine tune characteristics of the used parts to meet the specific metrics of the designed strains.
+For instance, if the goal is to produce a protein at a high yield, the designer might want to increase the TIR of the RBS responsible for translation of the said protein.
+To do that, one could use the existing RBS predictors, but they can be unreliable in the investigated genetic context.}
 \hl{To address this problem}, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context}.
+\mengyan{what does "specific genetic context" mean?}
 Importantly, this work focuses on RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
+
 We used Gaussian Process Regression \cite{Rasmussen2004} and  Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBSs.
-As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a perfect target for establishing a workflow that could be potentially translated to more complicated systems.
+\mengyan{Mengyan to extend this to one paragraph: emphases two ML tasks.}
+
+As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.
 Our overall experimental goal was to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
-\hl{This way, we discovered new RBS sequences, increasing production of our target protein compared to our very strong benchmark RBS by up to 35\%}.
-
-\input{methods.tex}
+\hl{This way, we were able to rationally increase production of our target protein compared to our very strong benchmark RBS by up to 35\%}.
 
@@
 We present our RBS-optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
-In this case, the {DBTL} paradigm is applied directly to the design of the RBS part itself, rather than general strain optimisation.
 Machine learning is used in two different ways: i) we show the efficacy of the ML recommendations
@@
 
-In our genetic design, the investigated RBS controls expression of green fluorescent protein (GFP).
-By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring the expression vector over time.
-Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1.
-Inducible expression allows one to synchronise the start of the GFP expression in all the cultures with addition of IPTG.
-Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier \cite{Beal2021}.\\
-
-Our template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
-This sequence is known to have a very high TIR in a similar context of expressing fluorescent protein and is present in the pBb series of plasmids \cite{Lee2011}.
+\hl{The exact position of the RBS in the sequence upstream of the CDS can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.} 
+In our investigation, we are using an RBS which is known to have a very high TIR when expressing GFP and is present in the pBb series of plasmids \cite{Lee2011}.
+This template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
 Since this is the sequence against which new RBS sequences will be benchmarked,
@@
 TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T).
-The total experimental (variant) space is then $4^6$ = 4096.
+The total experimental (variant) space is then $4^6$ = 4096.\\
+
+In our genetic design, the investigated RBS controls expression of green fluorescent protein (GFP).
+By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring the expression vector over time.
+Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1.
+Inducible expression allows one to synchronise the start of the GFP expression in all the cultures with addition of IPTG.
+Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier \cite{Beal2021}.\\
 
@@
 However, this is very useful information for future predictions as it allows us to understand the underlying space better.
+\hl{In other words, we are intentionally sacrificing accuracy of our predictions in each round to improve them in the future rounds}. 
 The effect of better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.\\
@@
 For RBSs with high TIRs ($>1$), the minimum distance that is required to increase the TIR is 2, with edit distances between 2 and 5 giving similar results.
-For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
+For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.\\
 
@@
 
-In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBSs.
+In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising TIR of a bacterial RBS}.
 We have used Gaussian Process regression to predict the TIR function to be optimised and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested.
@@
 In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.
-This approach may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
+\hl{As long as maximisation could be the appropriate function for optimising RBSs or other small parts}, it may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
 This change could be achieved within the Bayesian optimisation framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objective method, for example \mbox{\cite{shu2020new}}.\\
@@
 
-It is important to make a distinction between works focusing on the part itself like \mbox{\textcite{Hollerer2020}} and works where the part is optimised with a wider goal of strain optimisation like in \mbox{\textcite{Jervis2019}}.
-The part-centric optimisation, like the one considered in this paper, is an important task in synthetic biology because metabolic pathways are made up of individual parts and understanding their performance in separation allows for better strain design.
+Further, it is important to make a distinction between works focusing on the part itself like \mbox{\textcite{Hollerer2020}} and works where the part is optimised with a wider goal of strain optimisation like in \mbox{\textcite{Jervis2019}}.
+The part-centric optimisation, like the one considered in this paper, is an important task in synthetic biology since metabolic pathways are made up of individual parts and understanding their performance in separation allows for better strain design.
 While we focus our optimisation framework on the RBS optimisation task, this framework has good potential to be generalised to multi-gene pathway design as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}.
@@
 To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
+
+\input{methods.tex}
 

### 2021-11-17 paper/main.tex
mengyan's edit on intro

@@
 To do that, one could use the existing RBS predictors, but they can be unreliable in the investigated genetic context.}
-\hl{To address this problem}, we demonstrate how machine learning algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context}.
-\mengyan{what does "specific genetic context" mean?}
 Importantly, this work focuses on RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
+As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\
 
-We used Gaussian Process Regression \cite{Rasmussen2004} and  Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} for prediction and recommendation respectively to analyse and optimise the initiation rates of the designed RBSs.
-\mengyan{Mengyan to extend this to one paragraph: emphases two ML tasks.}
-
-As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.
 Our overall experimental goal was to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
-\hl{This way, we were able to rationally increase production of our target protein compared to our very strong benchmark RBS by up to 35\%}.
+We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context (i.e.
+with these specific bases upstream and downstream of the investigated RBS sequences)}.
+Two types of ML algorithms are applied in the DBTL cycle. 
+For the Learn part, the goal is design a predictor to predict the protein expression level by learning from the logged data. 
+We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression \cite{Rasmussen2004} in our pipeline. 
+For the Design part, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget. 
+A variant of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} was applied in our pipeline to guide our recommendations. 
+The two types of algorithms cooperate with each other and provides powerful tools for DBTL cycle in genetic parts design.
+\hl{This way, we were able to rationally increase production of our target protein compared to our very strong benchmark RBS by up to 35\%}. \\
 

### 2021-11-18 paper/main.tex
mengyan edit on sec 2.3

@@
 \mengyan{I still feel "reliable design methods" is too vague. e.g. what does reliable mean? why lack such methods? what are the challenging parts exactly?}
+\maciej {We don't have to be that specific in the abstract, remember that we only have 150 words.}
 To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed.
@@
 it feels like recommendation, but in the following we are discussing predictions.}
+\maciej{In this case translate means making recommendations from previous data.}
 For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
@@
 \mengyan{this is not the case for our study. The amount of data we have is quite small in ML view. And the small amount of data motivates the DBTL approach. }
+\maciej{It is, our data set is still considered pretty big :)}
 Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
@@
 
+To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design part, we select a regression algorithm, Gaussian Process regression (GPR),  which can generate not only predictions, but also uncertainty estimations.
+GPR, as a non-parametric Bayesian regression algorithm, assumes the reward function is sampled from a Gaussian Process and has been shown with good performance with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}.
+Additionally, GPR has been studied with bandits algorithms, such as UCB and its variants, with both theoretical and empirical good recommendation performance (i.e.
+sub-linear regret upper bound) \cite{srinivas2012information, desautels2014parallelizing}.
+\\
+
+The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks \cite{}.
+At the early iterations, we have a limited number of data points and relatively high variants due to the measurement noise. 
+Fortunately, since the predictions are only used for recommendations in the Design part, instead of precise predictions, we only need to provide a valid ranking for both predicted TIR and uncertainties, and thus the ranking of UCB scores.
+So we provide the evaluation of Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown as a more suitable evaluation metric when the prediction is used for recommendation tasks than $R^2$ or Pearman correlation coefficient
+\cite{Schober2018,  Kang2019}.
+\mengyan{Do we even want to put R2 score in?}\\
+
 Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.  
-As expected, the predictions in Round 0 were poor due to use of approximate data. 
-The predictions improved for the subsequent rounds, from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
-Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
-Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.\\
+In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design part (i.e.
+test data, blue).
+We reported the Spearman and $R^2$ score in each plot's title on test data only. 
+X-axis and y-axis are true TIR from our Test part of the DBTL cycle and the predicted TIR from our predictor.
+In (D), we train our predictor on randomly chosen 80\% of all available data, and test on the rest 20\% data. 
+As expected, the predictions in Round 0 were poor due to the use of approximate data. 
+The predictions improved for the subsequent rounds, 
+% from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
+the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
+\mengyan{We also show $R_2$ score for the reader's interest, which quantifies value-based prediction evaluation. Similarly, the $R_2$ increases along with rounds increase. }
+% Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
+\\
 
-The obtained values of the coefficient of determination and Spearman correlation coefficient are influenced by a number of factors.
-The primary one is our choice of the exploration-exploitation balance.
-In each round, we select a number of data points for exploration, which means that when tested, they have a high chance of having a experimentally-obtained mean different from what was predicted, thus decreasing the R\textsuperscript{2} values.
-However, this is very useful information for future predictions as it allows us to understand the underlying space better.
-\hl{In other words, we are intentionally sacrificing accuracy of our predictions in each round to improve them in the future rounds}. 
-The effect of better exploration of the space is the ability to find high TIR RBSs even with relatively low R\textsuperscript{2} values.\\
+The 
+% obtained values of the coefficient of determination and Spearman correlation coefficient 
+performance of Learn part 
+is also influenced by the exploration-exploitation balance in the Design part.
+In each round, we select some data points for exploration on the areas we have few or none data points.
+For those exploration points, since the predictor never learns their label distributions and patterns, there is no chance for the predictor to provide accurate predictions of TIR values,
+% which means that when tested, they have a high chance of having a experimentally-obtained mean different from what was predicted, 
+thus hurting the prediction performance in the short term.
+However, this is useful information for future predictions as it allows us to understand the whole underlying space instead of focusing on local sub-optimal data points.
+\hl{In other words, we are intentionally sacrificing the accuracy of our predictions in each round to improve them in the future rounds}. 
+The effect of exploration of the space is the ability to find high TIR RBSs even with relatively low prediction performance.\\
 
@@
 
-All code and data required to reproduce the results are available at Github: \url{https://github.com/mholowko/Solaris/tree/master/synbio_rbs}.
+All code and data required to reproduce the results are available at Github (San Francisco, CA): \url{https://github.com/mholowko/Solaris/tree/master/synbio_rbs}.
 All the processed and raw data are included in the repository.
 Sequences of plasmids and oligos and assembly reports used in this study are available in the supplementary information as a separate file.
-The pBbB6c plasmid is available on Addgene. Other strains and plasmids are available on request from the authors; MTAs will need to be negotiated between the parties.
+The pBbB6c plasmid is available on Addgene (Watertown, MA). Other strains and plasmids are available on request from the authors; MTAs will need to be negotiated between the parties.
 

### 2021-11-18 paper/methods.tex
mengyan edit on sec 2.3

@@
 The pBbB6c-GFP plasmid has been used for all our designs.
-This plasmid contains the GFP mut3b CDS, expression of which can be induced by the addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG).
+This plasmid contains the GFP mut3b CDS, expression of which can be induced by the addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG) (Merck, Darmstadt, Germany, catalogue no. I5502).
 The original RBS for the GFP CDS was replaced using a combination of PCR and isothermal assembly.
-Primer sequences and the assembly strategy were generated using the Teselagen DESIGN software (Teselagen Biotechnology).
+Primer sequences and the assembly strategy were generated using the Teselagen DESIGN software (Teselagen Biotechnology, San Francisco, CA).
 
 \textbf{PCR.}
-PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, catalogue no. M0492L).
-20 $\mu$L reactions were prepared by dispensing 1 $\mu$L of each 10 $\mu$M reverse primer into the wells of a 96-well PCR plate using the Labcyte Echo Liquid Handler.
-A mastermix consisting of polymerase premix, plasmid DNA template (pBbB6c, 5-10 ng per reaction), and the single 10 $\mu$M forward primer was prepared and dispensed using the FeliX liquid handler or electronic multi-channel pipette.
-Reactions were run using Touchdown PCR or standard PCR cycling methods in BioRad C1000 thermal cyclers.
-Capillary electrophoresis of PCR products was performed using the Agilent Technologies ZAG DNA Analyzer system.
-2 $\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000bp) or ZAG 110 dsDNA Kit (35-5000bp) (Agilent Technologies, catalogue no.
+PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, Ipswich, MA, catalogue no. M0492L).
+20 $\mu$L reactions were prepared by dispensing 1 $\mu$L of each 10 $\mu$M reverse primer into the wells of a 96-well PCR plate using the Echo liquid handler (Beckman Coulter, Brea, CA).
+A mastermix consisting of polymerase premix, plasmid DNA template (pBbB6c, 5-10 ng per reaction), and the single 10 $\mu$M forward primer was prepared and dispensed using the FeliX liquid handler (Analytik Jena, Jena, Germany) or electronic multi-channel pipette.
+Reactions were run using Touchdown PCR or standard PCR cycling methods in C1000 thermal cyclers (Bio-Rad, Hercules, CA).
+Capillary electrophoresis of PCR products was performed using the ZAG DNA Analyzer system (Agilent Technologies, Santa Clara, CA).
+2 $\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000 bp) or ZAG 110 dsDNA Kit (35-5000 bp) (Agilent Technologies, catalogue no.
 ZAG-110-5000; ZAG-130-5000).
-ZAG sample plates were prepared using the Perkin Elmer Sciclone G3 liquid handler.
+ZAG sample plates were prepared using the Sciclone G3 liquid handler (Perkin Elmer, Waltham, MA).
 ProSize Data Analysis Software (Agilent Technologies) was used to generate gel images from the sample chromatograms, and amplicon sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
@@
 \textbf{\textit{E. coli} transformation.}
-The DH5$\alpha$ cell line (Thermo Fisher Scientific, catalogue no.
+The DH5$\alpha$ cell line (Thermo Fisher Scientific, Waltham, MA, catalogue no.
 18265017) was made chemically competent using the Mix $\&$ Go \textit{E.
-coli} Transformation Kit $\&$ Buffer Set (Zymo Research, catalogue no.
+coli} Transformation Kit $\&$ Buffer Set (Zymo Research, Irvine, CA, catalogue no.
 T3001).
 20 $\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
-Plates of cells were thawed on a -20$^{\circ}$C cold block before 3 $\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
-Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 (12 x 8) grid on Omnitrays containing LB (BD, catalogue no.
-244610) and 15 g/L agar (Sigma-Aldrich, catalogue no.
-A1296) with 34 $\mu$g/mL chloramphenicol (Sigma-Aldrich, catalogue no.
+Plates of cells were thawed on a -20$^{\circ}$C cold block before 3 $\mu$L of the assembly product was added and mixed using the FeliX liquid handler.
+Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 (12 x 8) grid on Omnitrays containing LB (BD, Franklin Lakes, NJ, catalogue no.
+244610) and 15 g/L agar (Merck, catalogue no.
+A1296) with 34 $\mu$g/mL chloramphenicol (Merck, catalogue no.
 C1919).
-Plates were incubated overnight at 37$^{\circ}$C. Cells were plated using the CyBio FeliX liquid handler.
+Plates were incubated overnight at 37$^{\circ}$C. Cells were plated using the FeliX liquid handler.
 
 \textbf{Automated colony picking and culturing.}
-A Singer Instruments PIXL colony picker was used to select individual colonies from the transformation plates using the 490-510 nm (cyan) light filter.
+A PIXL colony picker (Singer Instruments, Roadwater, United Kingdom) was used to select individual colonies from the transformation plates using the 490-510 nm (cyan) light filter.
 Each selected colony was used to inoculate 1 mL of selective medium in a 2 mL square well 96 plate.
@@
 \textbf{Glycerol stock preparation.}
-100 $\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, catalogue no.
-GA010) and 100 $\mu$L of overnight culture were combined in the wells of a 96 deep (2 mL) round well plate using the CyBio FeliX liquid handler or electronic multichannel pipette
+100 $\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, Gillman, Australia, catalogue no.
+GA010) and 100 $\mu$L of overnight culture were combined in the wells of a 96 deep (2 mL) round well plate using the FeliX liquid handler or electronic multi-channel pipette
 They were then sealed with a 96-well silicone sealing mat and transferred to a -80$^{\circ}$C freezer.
@@
 \textbf{Sequencing}
-Strains that gave GFP fluorescence intensity readings similar to that of the original RBS were selected for sequence confirmation by capillary electrophoresis sequencing (CES) by Macrogen, Inc.
-(Seoul, South Korea).
+Strains that gave GFP fluorescence intensity readings similar to that of the original RBS were selected for sequence confirmation by capillary electrophoresis sequencing (CES) (Macrogen, Inc., Seoul, South Korea).
 The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
-Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN, catalogue no. 27106) according to the manufacturer's instructions.
-Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/$\mu$L.
+Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN, Hilden, Germany, catalogue no. 27106) according to the manufacturer's instructions.
+Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek, Winooski, VT) and all fell in the range of 100-200 ng/$\mu$L.
 Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5'-CGATATAGGCGCCAGCAA-3') that binds approximately 150 bp upstream of the RBS.
-Reads were aligned with the template sequence in the Teselagen software (TeselaGen Biotechnology, Inc.).
+Reads were aligned with the template sequence in the Teselagen software.
 
@@
 \textbf{Test strain culture.}
-Overnight cultures (six replicates for each batch) were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with 2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, using the CyBio FeliX liquid handler or electronic multichannel pipette.
+Overnight cultures (six biological (restarted every time from glycerol stock) replicates for each batch) were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with 2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, using the FeliX liquid handler or electronic multi-channel pipette.
 Cultures were incubated at 37$^{\circ}$C with shaking (300 rpm) for 17 hours. 
 The following morning, 20 $\mu$L of each overnight culture was added to 980 $\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in a 96 deep (2 mL) round well plate.
-After 90 minutes, cultures were induced with IPTG (Sigma-Aldrich, catalogue no.
-I5502) to a final concentration of 0.5 mM.
-This was done by transferring 1.0 $\mu$L of 0.1 M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo acoustic liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multichannel pipette.
+After 90 minutes, cultures were induced with IPTG to a final concentration of 0.5 mM.
+This was done by transferring 1.0 $\mu$L of 0.1 M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multi-channel pipette.
 
@@
 For label pre-processing, we first adjust the TIR values in each round using the round-wise reference values.
+
 The reference value is the TIR of the benchmark sequence that is run in triplicate in each round.
@@
 GPR can predict both the posterior mean and standard deviation of the RBS sequences in the design space. The posterior standard deviation represents the level of uncertainty for the prediction.
+
 We review the predictions of GPR in Supplementary \ref{supp: GPR}.\\
@@
 sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$ ($L = 20$ in our design).
+
 $S(l)$ determines the maximum shift at position $l$. 
@@
 %train GP on sequences with known TIR labels and
-recommend sequences in design space with top $m$ UCB scores, as shown in Figure \ref{fig:batch rec}A ($m = 2$).
 However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 

### 2021-11-21 paper/main.tex
mengyan edits on r2 and ee colors

@@
 
-\title{Multi-armed bandit \mengyan{(Bayesian?)} optimisation\\
-of bacterial Ribosome Binding Site}
+\title{Machine learning guided batched design\\
+of a bacterial Ribosome Binding Site}
 
@@
 The decreased result in the 0.9th quantile compared to the maximum value for the Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
-We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
+We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Supplementary \ref{subsec: Exploration-exploitation visualisation} and Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
@@
 RBSs with high TIRs tend to come from exploitation of the design space,
-whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge of unknown parts of the design space.}\\
+whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge to unknown parts of the design space.}\\
 
@@
 \label{sec:gp-results}
+\label{subsec: results LEARN}
 
@@
 
-The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks \cite{}.
+The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks.
 At the early iterations, we have a limited number of data points and relatively high variants due to the measurement noise. 
@@
 So we provide the evaluation of Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown as a more suitable evaluation metric when the prediction is used for recommendation tasks than $R^2$ or Pearman correlation coefficient
-\cite{Schober2018,  Kang2019}.
-\mengyan{Do we even want to put R2 score in?}\\
+\cite{Schober2018,  Kang2019}.\\
 
@@
 test data, blue).
-We reported the Spearman and $R^2$ score in each plot's title on test data only. 
+We reported the Spearman score in each plot's title on test data only. 
 X-axis and y-axis are true TIR from our Test part of the DBTL cycle and the predicted TIR from our predictor.
@@
 the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
-\mengyan{We also show $R_2$ score for the reader's interest, which quantifies value-based prediction evaluation. Similarly, the $R_2$ increases along with rounds increase. }
 % Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
@@
     Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
-    The $R^2$ value and Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
+    The Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
     The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
@@
 For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
-We have achieved this despite the relatively low R\textsuperscript{2} values for our predictions.
-Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low R\textsuperscript{2} values \cite{Radivojevic2020, Opgenorth2019}.
-Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric
-\cite{Schober2018,  Kang2019}.
+We have achieved this despite the relatively low Spearman scores for our predictions.
+Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low prediction scores \cite{Radivojevic2020, Opgenorth2019}.
+% Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric
+% \cite{Schober2018,  Kang2019}.
 Finally, our predictor Gaussian Process regression model, compared with previously described calculators using a deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our Bandit algorithm to give better recommendations.\\

### 2021-11-21 paper/supplementary.tex
mengyan edits on r2 and ee colors

@@
 
+\subsection{Addition Results}
+
+As stated in Section \ref{subsec: results LEARN}, we merely aimed to provide a valid ranking in Learn part due to the small amount of noisy data, which can be captured by the Spearman correlation coefficient.
+We show $R^2$ score for the reader's interest, which quantifies value-based prediction evaluation. The $R^2$ increases along with rounds increase.
+
+\begin{table}[h!]
+\caption{$R^2$ score for predictions in Figure \ref{fig: Scatterplot}.}
+\centering
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+                  & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0\\ Test 1\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0-1\\ Test 2\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0-2\\ Test 3\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 80\%\\ Test 20\%\end{tabular}} \\ \hline
+\textbf{$R^2$}       & 0.067                                                             & 0.226                                                               & 0.213                                                               & 0.269                                                                   \\ \hline
+\textbf{Spearman} & 0.269                                                             & 0.468                                                               & 0.546                                                               & 0.479                                                                   \\ \hline
+\end{tabular}
+\end{table}
+
+\subsection{Exploration-exploitation visualisation}
+\label{subsec: Exploration-exploitation visualisation}
+
+In this section, we show our method of exploration-exploitation visualisation displayed in Figure \ref{fig: Swarmplot and Quantplot}C.
+Recall the UCB score of a sequence $\mathbf{x}_i$ at Round $t$ is defined as 
+\begin{align*}
+% \label{Eq: UCB}
+    \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i),
+\end{align*}
+where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
+$\mu_{t-1}(\mathbf{x}_i), \sigma_{t-1}(\mathbf{x}_i)$ are the predicted TIR and standard deviation (STD) at Round $t-1$ for the sequence $\mathbf{x}_i$.
+The sequences with high predicted TIR address exploitation, while the sequences with high predicted STD address exploration. 
+We show the predicted TIR (x-axis) v.s. predicted STD (y-axis) in Figure \ref{fig:exploitationvsexplorationdetail} in orange points.
+
+To compare the predicted TIR and STD in the same dimension (basis), we project the (predicted TIR, predicted STD) pair to the diagonal line (the grey dashline) crossing (0,1) and (1,0).
+The slope of the diagonal line is selected to weight the predicted TIR and STD equally. 
+Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
+As a result, the higher the predicted TIR the closer the point is to the (1,0) coordinate, and the higher the predicted STD the closer the point is to the (0,1) coordinate.
+Finally, we colour those projected points using the RdBu spectrum (according to either of the two dimension of the projected points), where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
+These colors are correspond to ones shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
+
+% \mengyan{mengyan to re-write it}
+% For a given round all the tested sequences are then projected along the diagonal line crossing (0,1) and (1,0) according to their predicted STD and mean TIR (the higher the predicted mean the closer the point is to the (1,0) coordinate and the higher the predicted STD the closer the point is to the (0,1) coordinate).
+% Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
+% We do this to transform our predictions into a common basis (where the basis is the diagonal) and so by calculating where a given point falls on the diagonal we can show if its mean TIR or STD is higher.
+% Finally, we colour those projected points using the RdBu spectrum where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}).
+    We explain how the colors are generated in Section \ref{subsec: Exploration-exploitation visualisation}, which are correspond to the colors shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C.}
+   }
+   \label{fig:exploitationvsexplorationdetail}
+\end{figure}
+
 % \newpage
@@
     \centering
-    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
-    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
-    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}).
-   For a given round all the tested sequences are then projected along the diagonal line crossing (0,1) and (1,0) according to their predicted STD and mean TIR (the higher the predicted mean the closer the point is to the (1,0) coordinate and the higher the predicted STD the closer the point is to the (0,1) coordinate).
-   Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
-   We do this to transform our predictions into a common basis (where the basis is the diagonal) and so by calculating where a given point falls on the diagonal we can show if its mean TIR or STD is higher.
-   Finally, we colour those projected points using the RdBu spectrum where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue, and these correspond to the points shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
-   }
-   \label{fig:exploitationvsexplorationdetail}
-\end{figure}
-
-\begin{figure}[!ht]
-    \centering
     \includegraphics[scale = 0.35]{plots/Supplementary/scatter_bc1_TTF_0.pdf}
@@
     Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation described in Section \mbox{\ref{supp: Normalisation of Kernel}}.
-Although the $R^2$ scores decrease considerably compared with Figure \mbox{\ref{fig: Scatterplot}}, the Spearman correlation remains similar.
+Although the precise value prediction looks different from Figure \mbox{\ref{fig: Scatterplot}}, the rankings provided in two figures are similar, which can be reflected by the similar Spearman correlation coefficients.
     }

### 2021-11-22 paper/main.tex
Update with requested changes (paper).

@@
 Unfortunately, design of specific genetic parts can still be challenging due to lack of reliable design methods.
-\mengyan{I still feel "reliable design methods" is too vague. e.g. what does reliable mean? why lack such methods? what are the challenging parts exactly?}
-\maciej {We don't have to be that specific in the abstract, remember that we only have 150 words.}
-To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed.
-We used Gaussian Process Regression for the Learn part of cycle and the Upper Confidence Bound multi-armed Bandit algorithm for the Design of genetic designs to be tested in vivo.
-We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for a more reliable data generation .
-The algorithms were able to identify RBS sequence rules conducive to high protein expression.
+To address this problem, we have created a machine learning guided Design-Build-Test-Learn cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed \hl{using relatively small, high-quality data sets}.
+We used Gaussian Process Regression for the Learn part of cycle and the Upper Confidence Bound multi-armed Bandit algorithm for the Design of genetic variants to be tested in vivo.
+We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for a more reliable data generation.
 Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-Part design \mengyan{what is part design?} is a central problem of synthetic biology as it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
-\mengyan{make the previous sentence shorter?}
+\hl{Part design, understood as modifying the sequence of a part for it to meet specific requirements, is a central problem of synthetic biology.
+This is} because, it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
@@
 
-The biggest gap in the DBTL cycle is at the interface between Learn and Design - it is hard to translate obtained results into new designs.
-\mengyan{what does "translate" mean?
-it feels like recommendation, but in the following we are discussing predictions.}
-\maciej{In this case translate means making recommendations from previous data.}
-For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+\hl{For instance}, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
@@
 
-The volume of experimental data in synthetic biology is increasing exponentially. \cite{Freemont2019}. 
-\mengyan{this is not the case for our study. The amount of data we have is quite small in ML view. And the small amount of data motivates the DBTL approach. }
-\maciej{It is, our data set is still considered pretty big :)}
+The volume of experimental data in synthetic biology is increasing exponentially \cite{Freemont2019}. 
 Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
@@
 Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}. 
-Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.\\
+Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.
+\hl{However, such big data sets are still generated relatively rarely and there is a need to develop methods operating on smaller data sets.}\\
 
@@
 For instance, if the goal is to produce a protein at a high yield, the designer might want to increase the TIR of the RBS responsible for translation of the said protein.
-To do that, one could use the existing RBS predictors, but they can be unreliable in the investigated genetic context.}
+To do that, one could use the existing RBS predictors, but they can be unreliable in the investigated genetic context.
+On the other hand, the designer could use machine learning methods to make the predictions, but these usually require large data sets.
+To address this problem, we have created a machine learning guided Design-Build-Test-Learn cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.}
 Importantly, this work focuses on RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
@@
 Our overall experimental goal was to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
+\maciej{I've decided not to put the "other than maximisation" sentence here, I think it's appropriate to have it in discussion.}
 We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context (i.e.
-with these specific bases upstream and downstream of the investigated RBS sequences)}.
+with the specific bases upstream and downstream of the investigated RBS sequences)}.
 Two types of ML algorithms are applied in the DBTL cycle. 
@@
 We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression \cite{Rasmussen2004} in our pipeline. 
+\hl{GPR, as a non-parametric Bayesian regression algorithm, assumes that the reward function is sampled from a Gaussian Process and has been shown to  perform well with a small amount of data in biological prediction tasks} \cite{Rasmussen2004, desautels2014parallelizing}.
 For the Design part, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget. 
 A variant of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} was applied in our pipeline to guide our recommendations. 
-The two types of algorithms cooperate with each other and provides powerful tools for DBTL cycle in genetic parts design.
-\hl{This way, we were able to rationally increase production of our target protein compared to our very strong benchmark RBS by up to 35\%}. \\
+The two types of algorithms cooperate with each other and provides powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
+\hl{This way, we were able to rationally increase production of our target protein compared to the very strong benchmark RBS by up to 35\%}. \\
 
@@
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
-\hl{Since we expect that an RBS sequence chosen at random will have low TIR
-(as shown by UNI and PPM), this implies that most of the exploration will result in low TIR.
-Our results confirm that
-RBSs with high TIRs tend to come from exploitation of the design space,
-whereas the exploration points give relatively low TIRs. Note that the exploration is necessary to expand our knowledge to unknown parts of the design space.}\\
+\hl{The implication of the fact that an RBS sequence chosen at random will have low TIR
+(as shown by UNI and PPM), is that most of the exploration will result in low TIR.
+Our results confirm that RBSs with high TIRs tend to come from exploitation of the design space,
+whereas the exploration points give relatively low TIRs.
+Note that the exploration is necessary to expand our knowledge to the unknown parts of the design space and in effect allow us to exploit it better.}\\
 
@@
 
-To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design part, we select a regression algorithm, Gaussian Process regression (GPR),  which can generate not only predictions, but also uncertainty estimations.
-GPR, as a non-parametric Bayesian regression algorithm, assumes the reward function is sampled from a Gaussian Process and has been shown with good performance with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}.
-Additionally, GPR has been studied with bandits algorithms, such as UCB and its variants, with both theoretical and empirical good recommendation performance (i.e.
-sub-linear regret upper bound) \cite{srinivas2012information, desautels2014parallelizing}.
-\\
-
-The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks.
-At the early iterations, we have a limited number of data points and relatively high variants due to the measurement noise. 
-Fortunately, since the predictions are only used for recommendations in the Design part, instead of precise predictions, we only need to provide a valid ranking for both predicted TIR and uncertainties, and thus the ranking of UCB scores.
-So we provide the evaluation of Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown as a more suitable evaluation metric when the prediction is used for recommendation tasks than $R^2$ or Pearman correlation coefficient
-\cite{Schober2018,  Kang2019}.\\
-
-Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.  
-In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design part (i.e.
-test data, blue).
-We reported the Spearman score in each plot's title on test data only. 
-X-axis and y-axis are true TIR from our Test part of the DBTL cycle and the predicted TIR from our predictor.
-In (D), we train our predictor on randomly chosen 80\% of all available data, and test on the rest 20\% data. 
+\hl{To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design part, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points.
+}
+Figure \ref{fig: Scatterplot} \hl{shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.
 As expected, the predictions in Round 0 were poor due to the use of approximate data.
 The predictions improved for the subsequent rounds, 
-% from R\textsuperscript{2} (coefficient of determination) of 0.067 for Round 0 to R\textsuperscript{2} of 0.213 for Round 3.
-the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
+the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.}
 % Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
@@
 
-The 
+\hl{The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks.
+In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise. 
+However, since the predictions are only used for recommendations in the Design part, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
+This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
+Because of that, we provide the evaluation of the Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearman correlation coefficient}
+\cite{Schober2018,  Kang2019}.\\
+
+Additionally, the 
 % obtained values of the coefficient of determination and Spearman correlation coefficient 
@@
 is also influenced by the exploration-exploitation balance in the Design part.
-In each round, we select some data points for exploration on the areas we have few or none data points.
+In each round, we select some data points for exploration of the areas in which we have few or none tested data points.
 For those exploration points, since the predictor never learns their label distributions and patterns, there is no chance for the predictor to provide accurate predictions of TIR values,
@@
     Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
-    The Spearman correlation coefficient (with corresponding p-value) are provided for each plot.
+    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided for each plot.
+    X-axis and y-axis are the true, tested TIR and the predicted TIR respectively.
+    In D), we train our predictor on randomly chosen 80\% of all available data, and test on the remaining 20\% data. 
     The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
@@
 
-\section{Discussion and Conclusion}
+\section{Discussion}
 
@@
 Thirdly, the generalisability of the method is unknown.
-\hl{Since the model has been taught on data relating to this specific genetic context its applicability for other context will be limited and new data will have to be obtained.}
 We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
-% \hl{Additionally, while we focus our optimisation framework on the RBS part optimisation task, this framework has good potential to be generalised to multi-gene pathway design.}
-% \hl{Since such optimisation would be a large-scale data task, the current GPR prediction model could be updated to a deep GPR approach and the current bandits algorithm could be optimised towards querying large design space to reduce computational complexity } \cite{Shahriari2016}.
+\hl{Similarly, the re-usability of the obtained data set is currently unknown. 
+The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures} \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} \hl{respectively) studies would indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
+For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.}
 Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.

### 2021-12-18 paper/main.tex
cheng's changes in intro

@@
 Unfortunately, design of specific genetic parts can still be challenging due to lack of reliable design methods.
-To address this problem, we have created a machine learning guided Design-Build-Test-Learn cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed \hl{using relatively small, high-quality data sets}.
-We used Gaussian Process Regression for the Learn part of cycle and the Upper Confidence Bound multi-armed Bandit algorithm for the Design of genetic variants to be tested in vivo.
-We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for a more reliable data generation.
-Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
-Overall, our results show that machine learning is a powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\
+To address this problem, we have created a machine learning guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed using relatively small, high-quality data sets.
+We used Gaussian Process Regression for the Learn phase of cycle and the Upper Confidence Bound multi-armed bandit algorithm for the Design of genetic variants to be tested in vivo.
+We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for reliable data generation.
+Notably, by Testing a total of 450 RBS variants in four DBTL cycles, we experimentally validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
+Overall, our results show that machine learning is a powerful tool for designing RBSs, and they pave the way towards more complicated genetic devices.\\
 
@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-\hl{Part design, understood as modifying the sequence of a part for it to meet specific requirements, is a central problem of synthetic biology.
-This is} because, it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
+Part design, understood as modifying the sequence of a part for it to meet specific requirements, is a central challenge in synthetic biology.
+Genetic parts are the units which are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
@@
 Additionally, computer modelling can be used to predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
-Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
+Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBSs) \cite{Xia1998,Chen2013,Reeve2014}.
 However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.
@@
 
-\hl{For instance}, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+In the Design phase of the DBTL cycle, designers often have to fine tune characteristics of parts to give the resulting strains their desired properties.
+The ability to predict a characteristic of a genetic part (for example using a machine
+learning method) only solves half of the problem. The designer still needs to choose from the large number of 
+possible variants to Build.
+For instance, if the goal is to produce a protein at a high yield, the designer might increase the TIR of the RBS responsible for translation of that protein.
+Hence the Design phase of TIR has two steps: first to predict the TIR value for each potential RBS sequence,
+and second to select a small set of RBS sequences based on their predicted TIR values.\\
+
+To achieve the first step, one could use the existing RBS predictors.
+\textcite{Reeve2014} surveys three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
 Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
@@
 This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.
-\hl{This suggests that there is a number of interactions determining the mRNA-ribosome binding which can be hard to capture in a single, universal model}.\\
+This suggests that multiple interactions determine the mRNA-ribosome binding, and predicting TIR from the genomic sequence is still challenging.\\
 
-The volume of experimental data in synthetic biology is increasing exponentially \cite{Freemont2019}. 
-Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
-For example, Jervis \emph{et al.} used support vector machines and neural networks to optimise production of monoterpenoids in \emph{Escherichia coli} \cite{Jervis2019}.
-Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}. 
+Recent work has explored the use of machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
+This leverages the exponential increase in experimental data from synthetic biology~\cite{Freemont2019}
+to improve predictors in the Learn phase of the DBTL cycle.
+For example, \textcite{Jervis2019} used support vector machines and neural networks to optimise production of monoterpenoids in \emph{Escherichia coli}.
+Similarly, \textcite{Costello2018} have used a number of machine learning approaches to analyse time-series multi-omics data to predict metabolic pathway behaviour. 
 Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.
-\hl{However, such big data sets are still generated relatively rarely and there is a need to develop methods operating on smaller data sets.}\\
+However large data sets may not be available for a genetic part of interest, for instance the RBS.
+Hence there is a need to develop methods for training predictors on smaller data sets in the Learn phase of DBTL.
+The trained predictors
+can then be used in the first step of the Design phase to predict characteristics of interest for a potential variant.\\
 
-\hl{In genetic design practice, the designers often have to fine tune characteristics of the used parts to meet the specific metrics of the designed strains.
-For instance, if the goal is to produce a protein at a high yield, the designer might want to increase the TIR of the RBS responsible for translation of the said protein.
-To do that, one could use the existing RBS predictors, but they can be unreliable in the investigated genetic context.
-On the other hand, the designer could use machine learning methods to make the predictions, but these usually require large data sets.
-To address this problem, we have created a machine learning guided Design-Build-Test-Learn cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.}
-Importantly, this work focuses on RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
-As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\
+In this paper, we propose a machine learning guided Design-Build-Test-Learn cycle 
+for the experimental design of bacterial RBSs. We show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.
+In particular, we develop a method for the Learn phase, and use the learned predictor for the first step of the Design phase.
+Furthermore we develop a method for the second step of the Design phase,
+that recommends batches of genetic designs that optimize the characteristic of interest.
+This work focuses on RBS part design and TIR prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
+As the RBS is one of the key genetic elements controlling protein expression in bacteria,
+it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\
 
-Our overall experimental goal was to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
-\maciej{I've decided not to put the "other than maximisation" sentence here, I think it's appropriate to have it in discussion.}
-We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context (i.e.
-with the specific bases upstream and downstream of the investigated RBS sequences)}.
+Our overall experimental goal is to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
+%\maciej{I've decided not to put the "other than maximisation" sentence here, I think it's appropriate to have it in discussion.}
+We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level in the specific genetic context (i.e.
+with specified bases upstream and downstream of the investigated RBS sequences).
 Two types of ML algorithms are applied in the DBTL cycle. 
-For the Learn part, the goal is design a predictor to predict the protein expression level by learning from the logged data. 
-We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression \cite{Rasmussen2004} in our pipeline. 
-\hl{GPR, as a non-parametric Bayesian regression algorithm, assumes that the reward function is sampled from a Gaussian Process and has been shown to  perform well with a small amount of data in biological prediction tasks} \cite{Rasmussen2004, desautels2014parallelizing}.
-For the Design part, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget. 
+For the Learn phase, the goal is to train a predictor that will predict the protein expression level by learning from the logged data. 
+We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression (GPR) \cite{Rasmussen2004} in our pipeline. 
+GPR has been shown to  perform well with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}. 
+For the Design phase, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget.
+The policy uses the outputs of the GPR predictions to recommend useful batches for the Build and Test phases.
 A variant of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} was applied in our pipeline to guide our recommendations. 
-The two types of algorithms cooperate with each other and provides powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
-\hl{This way, we were able to rationally increase production of our target protein compared to the very strong benchmark RBS by up to 35\%}. \\
+The two types of algorithms cooperate with each other and provide powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
+Using this approach, as summarized in Figure~\ref{fig: Flowchart}, we were able to increase expression of our target protein by up to 35\%, as compared to the very strong benchmark RBS. \\
 
@@
 
-\hl{The exact position of the RBS in the sequence upstream of the CDS can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.} 
+\hl{The exact position of the RBS in the sequence upstream of the protein coding sequence (CDS) can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.}
 In our investigation, we are using an RBS which is known to have a very high TIR when expressing GFP and is present in the pBb series of plasmids \cite{Lee2011}.
@@
 \hl{The implication of the fact that an RBS sequence chosen at random will have low TIR
-(as shown by UNI and PPM), is that most of the exploration will result in low TIR.
+(as shown by UNI and PPM), is that most of the exploration will result in low TIRs.
 Our results confirm that RBSs with high TIRs tend to come from exploitation of the design space,
@@
 
-\hl{To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design part, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points.
+\hl{To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design phase, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points.
 }
+% GPR, as a non-parametric Bayesian regression algorithm, assumes the reward function is sampled from a Gaussian Process and has been shown with good performance with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}.
+Additionally, GPR has been studied with bandits algorithms, such as UCB and its variants, with both theoretical and empirical good recommendation performance (i.e.
+sub-linear regret upper bound) \cite{srinivas2012information, desautels2014parallelizing}.
 Figure \ref{fig: Scatterplot} \hl{shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.
@@
 
-\hl{The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks.
+\hl{The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
 In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise. 
-However, since the predictions are only used for recommendations in the Design part, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
+However, since the predictions are only used for recommendations in the Design phase, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
 This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
-Because of that, we provide the evaluation of the Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearman correlation coefficient}
+Because of that, we provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient}
 \cite{Schober2018,  Kang2019}.\\
@@
 % obtained values of the coefficient of determination and Spearman correlation coefficient 
-performance of Learn part 
-is also influenced by the exploration-exploitation balance in the Design part.
+performance of the Learn phase 
+is also influenced by the exploration-exploitation balance in the Design phase.
 In each round, we select some data points for exploration of the areas in which we have few or none tested data points.
@@
     Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
-    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided for each plot.
     X-axis and y-axis are the true, tested TIR and the predicted TIR respectively.
+    In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design phase (i.e.
+test data, blue).
     In D), we train our predictor on randomly chosen 80\% of all available data, and test on the remaining 20\% data. 
+    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided in each plot's title on test data only.
     The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
@@
 
-\hl{Finally, while our strong sequences showed some affinity towards the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs} (see Figure \ref{fig:structures}).
+\hl{Finally, while our strong sequences showed some affinity for the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs} (see Figure \ref{fig:structures}).
 This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion, based on the previously reported literature \cite{Saito2020,EspahBorujeni2016}, that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
@@
 
-In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising TIR of a bacterial RBS}.
+In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising the TIR of a bacterial RBS}.
 We have used Gaussian Process regression to predict the TIR function to be optimised and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested.
@@
 In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in this context} by up to 34\%.
-Furthermore, we have generated an extensive library of diverse RBSs that can be used \hl{as a basis} in future studies.\\
+By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in this genetic context} by up to 34\%.
+Furthermore, we have generated an extensive library of diverse RBSs that can be used \hl{as a basis} for future studies.\\
 
 In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.
-\hl{As long as maximisation could be the appropriate function for optimising RBSs or other small parts}, it may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
+\hl{While maximisation could be the appropriate function for optimising RBSs or other small parts}, it may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
 This change could be achieved within the Bayesian optimisation framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objective method, for example \mbox{\cite{shu2020new}}.\\
@@
 \hl{Similarly, the re-usability of the obtained data set is currently unknown. 
-The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures} \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} \hl{respectively) studies would indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
+The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures} \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} \hl{respectively) studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
 For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.}
@@
 \end{document}
+However large data sets may not be available for a genetic part of interest, for instance the RBS.
+Hence there is a need to develop methods for training predictors on smaller data sets in the Learn phase of DBTL.\\

### 2021-12-18 paper/supplementary.tex
cheng's changes in intro

@@
 In this section, we show our method of exploration-exploitation visualisation displayed in Figure \ref{fig: Swarmplot and Quantplot}C.
-Recall the UCB score of a sequence $\mathbf{x}_i$ at Round $t$ is defined as 
+The UCB score of a sequence $\mathbf{x}_i$ at Round $t$ is defined as 
 \begin{align*}
@@
 The sequences with high predicted TIR address exploitation, while the sequences with high predicted STD address exploration. 
-We show the predicted TIR (x-axis) v.s. predicted STD (y-axis) in Figure \ref{fig:exploitationvsexplorationdetail} in orange points.
+We show the predicted TIR (x-axis) v.s. predicted STD (y-axis) in Figure \ref{fig:exploitationvsexplorationdetail} as orange points.
 
-To compare the predicted TIR and STD in the same dimension (basis), we project the (predicted TIR, predicted STD) pair to the diagonal line (the grey dashline) crossing (0,1) and (1,0).
+To compare the predicted TIR and STD in the same dimension (basis), we project the (predicted TIR, predicted STD) pair to the diagonal line (the grey dashline) set between (0,1) and (1,0) coordinates.
 The slope of the diagonal line is selected to weight the predicted TIR and STD equally. 
@@
 As a result, the higher the predicted TIR the closer the point is to the (1,0) coordinate, and the higher the predicted STD the closer the point is to the (0,1) coordinate.
-Finally, we colour those projected points using the RdBu spectrum (according to either of the two dimension of the projected points), where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
-These colors are correspond to ones shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
+Finally, we colour those projected points using the RdBu spectrum, where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
+Finally, these colours correspond to ones shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
 
@@
     \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}).
-    We explain how the colors are generated in Section \ref{subsec: Exploration-exploitation visualisation}, which are correspond to the colors shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C.}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}) and their corresponding projections to (0,1), (0,1) diagonal coloured in RdBu spectrum.
+    We explain in details how the colours are generated in Section \ref{subsec: Exploration-exploitation visualisation} .
    }

### 2021-12-18 paper/main.tex
mengyan's changes in intro

@@
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
-This cycle involves designing the DNA sequence in computer-aided design (CAD) software and then physically testing it in a laboratory.
-Additionally, computer modelling can be used to predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
+This cycle involves designing (D) new DNA sequences to achieve a desired property in computer-aided design (CAD) software, then physically building (B) new constructed variants and testing (T) through microplate reader in a laboratory.
+Computer modelling can be used to learn (L) and predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
 Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBSs) \cite{Xia1998,Chen2013,Reeve2014}.
@@
 The ability to predict a characteristic of a genetic part (for example using a machine
-learning method) only solves half of the problem. The designer still needs to choose from the large number of 
+learning method) only provides inputs to the problem. The designer still needs to choose from the large number of 
 possible variants to Build.
 For instance, if the goal is to produce a protein at a high yield, the designer might increase the TIR of the RBS responsible for translation of that protein.
-Hence the Design phase of TIR has two steps: first to predict the TIR value for each potential RBS sequence,
-and second to select a small set of RBS sequences based on their predicted TIR values.\\
+% Hence the Design phase of TIR has two steps: 
+% first to predict the TIR value for each potential RBS sequence,
+Hence the goal for the Design phase of TIR
+is to select a small set of RBS sequences from the design space (i.e. all possible DNA sequences) based on the predictions from Learn phase.
+Not only the predicted TIR mean values are needed for exploitation, the uncertainty levels of predictions, such as predicted standard deviation, are also needed for exploration. 
+One main challenge of the Design phase is to balance the exploitation and exploration, 
+which can be addressed by one type of machine learning approaches called multi-armed bandits \cite{lattimore2020bandit}.
+% We address this challenge by applying a multi-armed bandits algorithm named Upper Confidence Bound (UCB) \cite{lattimore2020bandit}.
+\\
 
-To achieve the first step, one could use the existing RBS predictors.
+% To achieve the first step, 
+One way to generate TIR mean value predictions is to use the existing RBS calculators.
 \textcite{Reeve2014} surveys three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
@@
 This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.
-This suggests that multiple interactions determine the mRNA-ribosome binding, and predicting TIR from the genomic sequence is still challenging.\\
+This suggests that multiple interactions determine the mRNA-ribosome binding, and predicting TIR from the genomic sequence is still challenging.
+Furthermore, the existing RBS calculators cannot predict the uncertainty levels to guide exploration.\\
 
-Recent work has explored the use of machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
+Recent work has explored the use of machine learning predictors to learn from history data and generate predictions for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
 This leverages the exponential increase in experimental data from synthetic biology~\cite{Freemont2019}
@@
 However large data sets may not be available for a genetic part of interest, for instance the RBS.
-Hence there is a need to develop methods for training predictors on smaller data sets in the Learn phase of DBTL.
-The trained predictors
-can then be used in the first step of the Design phase to predict characteristics of interest for a potential variant.\\
+Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both mean values and uncertainties in the Learn phase of DBTL. The predictions
+can then be used in the Design phase to predict characteristics of interest for a potential variant.
+% We apply a Bayesian model called Gaussian Process Regression \cite{Rasmussen2004} to learn from data in previous rounds and predict in design space.
+\\
 
@@
 for the experimental design of bacterial RBSs. We show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.
-In particular, we develop a method for the Learn phase, and use the learned predictor for the first step of the Design phase.
-Furthermore we develop a method for the second step of the Design phase,
-that recommends batches of genetic designs that optimize the characteristic of interest.
+% In particular, we develop a method and use the learned predictor to generate both mean and uncertainty predictions for the Learn phase.
+% Furthermore we develop a method the Design phase
+% that recommends batches of genetic designs that optimize the characteristic of interest.
 This work focuses on RBS part design and TIR prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
@@
 In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising the TIR of a bacterial RBS}.
-We have used Gaussian Process regression to predict the TIR function to be optimised and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested.
+We have used Gaussian Process regression to predict the TIR mean and standard deviation, and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
 To represent RBSs and capture the similarities between them, we chose to use the Weighted Degree Kernel with Shift method, which fits well into our prediction method.

### 2021-12-23 paper/main.tex
Editing, the conclusion needs a bit more work

@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-Part design, understood as modifying the sequence of a part for it to meet specific requirements, is a central challenge in synthetic biology.
-Genetic parts are the units which are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms.
+A central challenge is part design, understood as modifying the sequence of a genetic part for it to meet specific requirements.
+Genetic parts are the units which are ultimately combined into more complex genetic circuits that produce desired functions in the target organisms.
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
-This cycle involves designing (D) new DNA sequences to achieve a desired property in computer-aided design (CAD) software, then physically building (B) new constructed variants and testing (T) through microplate reader in a laboratory.
+This cycle involves designing (D) new DNA sequences to achieve a desired property in computer-aided design software, then physically building (B) new constructed variants and testing (T) through a microplate reader in a laboratory.
 Computer modelling can be used to learn (L) and predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
 Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBSs) \cite{Xia1998,Chen2013,Reeve2014}.
-However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.
-\\
+However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.\\
+
+In this paper, we propose a machine learning guided Design-Build-Test-Learn cycle
+for the experimental design of bacterial RBSs. This consists of two distinct
+types of machine learning methods, one in the Learn phase and a second in the
+Design phase.
+We show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.
+% In particular, we develop a method and use the learned predictor to generate both mean and uncertainty predictions for the Learn phase.
+% Furthermore we develop a method the Design phase
+% that recommends batches of genetic designs that optimize the characteristic of interest.
+This work focuses on RBS part design and TIR prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
+As the RBS is one of the key genetic elements controlling protein expression in bacteria,
+it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\
 
@@
 possible variants to Build.
-For instance, if the goal is to produce a protein at a high yield, the designer might increase the TIR of the RBS responsible for translation of that protein.
+For instance, if the goal is to increase the yield of a protein, the designer might increase the TIR of the RBS responsible for translation of that protein.
 % Hence the Design phase of TIR has two steps:
@@
 is to select a small set of RBS sequences from the design space (i.e. all possible DNA sequences) based on the predictions from Learn phase.
-Not only the predicted TIR mean values are needed for exploitation, the uncertainty levels of predictions, such as predicted standard deviation, are also needed for exploration. 
-One main challenge of the Design phase is to balance the exploitation and exploration, 
-which can be addressed by one type of machine learning approaches called multi-armed bandits \cite{lattimore2020bandit}.
-% We address this challenge by applying a multi-armed bandits algorithm named Upper Confidence Bound (UCB) \cite{lattimore2020bandit}.
+However since the computational predictions are not perfectly accurate,
+there is uncertainty about which potential RBS sequence has the desired
+property (for example highest protein yield).
+The designer could use the mean predicted TIR values and choose the best
+ones to exploit the knowledge modelled by the computational predictions.
+But the designer may also want to explore regions of DNA design space
+where the predictions are highly uncertain (e.g. as measured by
+the predicted TIR standard deviation) and could yield high payoff.
+Hence a major challenge of the Design phase is to balance
+the exploitation and exploration,
+which is addressed by one class of machine learning approaches
+called multi-armed bandits \cite{lattimore2020bandit}.
+As we will see in this paper, multi-armed bandits are well suited
+to solving the challenge of recommending a small set of RBS sequences
+in the Design phase of the DBTL cycle.
 \\
 
-% To achieve the first step, 
-One way to generate TIR mean value predictions is to use the existing RBS calculators.
+One way to generate predicted TIR values is to use the existing RBS calculators.
 \textcite{Reeve2014} surveys three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
@@
 This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.
-This suggests that multiple interactions determine the mRNA-ribosome binding, and predicting TIR from the genomic sequence is still challenging.
-Furthermore, the existing RBS calculators cannot predict the uncertainty levels to guide exploration.\\
+This suggests that multiple interactions determine the mRNA-ribosome binding, and predicting TIR from the genomic sequence is still challenging.\\
 
-Recent work has explored the use of machine learning predictors to learn from history data and generate predictions for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
+Recent work has explored the use of machine learning predictors to learn
+from historical data and generate predictions for use in synthetic biology,
+vastly improving the DBTL cycle's performance~\cite{Camacho2018,Radivojevic2020,LAWSON2021}.
 This leverages the exponential increase in experimental data from synthetic biology~\cite{Freemont2019}
@@
 Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.
-However large data sets may not be available for a genetic part of interest, for instance the RBS.
-Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both mean values and uncertainties in the Learn phase of DBTL. The predictions
-can then be used in the Design phase to predict characteristics of interest for a potential variant.
+Recall that the multi-armed bandit approach balances the exploration-exploitation
+trade off by using predictive uncertainty. The machine learning predictors
+mentioned, as well as existing RBS calculators,
+do not yet provide uncertainty levels to guide exploration.
+Furthermore large data sets may not be available for a genetic part of interest, for instance the RBS.
+Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both mean values and uncertainties in the Learn phase of DBTL.
+The predictions of characteristics of interest for
+each potential variant
+can then be used in the Design phase.
 % We apply a Bayesian model called Gaussian Process Regression \cite{Rasmussen2004} to learn from data in previous rounds and predict in design space.
@@
 
-In this paper, we propose a machine learning guided Design-Build-Test-Learn cycle 
-for the experimental design of bacterial RBSs. We show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.
-% In particular, we develop a method and use the learned predictor to generate both mean and uncertainty predictions for the Learn phase.
-% Furthermore we develop a method the Design phase
-% that recommends batches of genetic designs that optimize the characteristic of interest.
-This work focuses on RBS part design and TIR prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
-As the RBS is one of the key genetic elements controlling protein expression in bacteria,
-it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\
-
-Our overall experimental goal is to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
+Our overall experimental goal is to maximise the TIR by building and testing batches of RBS sequences with only a small number of DBTL cycle iterations.
 %\maciej{I've decided not to put the "other than maximisation" sentence here, I think it's appropriate to have it in discussion.}
-We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level in the specific genetic context (i.e.
-with specified bases upstream and downstream of the investigated RBS sequences).
+We demonstrate how machine learning (ML) algorithms can be incorporated
+into the DBTL cycle to predict (Learn) and recommend (Design) variants
+of a bacterial RBS with the goal of optimising associated protein expression
+level in the specific genetic context
+(i.e. with specified bases upstream and downstream of the investigated RBS sequences).
+Our proposed machine learning guided DBTL cycle is summarized in Figure~\ref{fig: Flowchart}.
 Two types of ML algorithms are applied in the DBTL cycle.
 For the Learn phase, the goal is to train a predictor that will predict the protein expression level by learning from the logged data.
-We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression (GPR) \cite{Rasmussen2004} in our pipeline. 
-GPR has been shown to  perform well with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}. 
-For the Design phase, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget.
+We used a Bayesian, non-parametric regression algorithm, called
+Gaussian Process Regression (GPR) \cite{Rasmussen2004} in our pipeline.
+In addition to providing a predicted mean TIR, GPR also provides
+uncertainty estimates via its predicted standard deviation.
+GPR has been shown to  perform well with small amounts of trainig data in
+biological prediction tasks~\cite{Rasmussen2004, desautels2014parallelizing}.
+For the Design phase, the goal is to find a policy to recommend RBS sequences
+to query in batches so that we can identify the optimized RBS sequences
+within a given budget.
+We use a version of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} to learn the policy.
 The policy uses the outputs of the GPR predictions to recommend useful batches for the Build and Test phases.
-A variant of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} was applied in our pipeline to guide our recommendations. 
+We demonstrate that laboratory automation methods in the Build and Test phases
+result in high quality TIR data, that is well suited for training
+machine learning methods in the subsequent Learn phase.
 The two types of algorithms cooperate with each other and provide powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
-Using this approach, as summarized in Figure~\ref{fig: Flowchart}, we were able to increase expression of our target protein by up to 35\%, as compared to the very strong benchmark RBS. \\
+Using our proposed machine learning guided DBTL cycle
+(Figure~\ref{fig: Flowchart}), we were able to increase expression of our
+target protein by up to 35\%, as compared to the very strong benchmark RBS. \\
 
@@
 
-The Bandit recommendations were made using the batch Upper Confidence Bound multi-armed Bandit algorithm.
-In short, this algorithm is a stochastic method of probing the experimental space. 
+The Bandit recommendations were made using the batch Upper Confidence Bound multi-armed Bandit algorithm~\cite{desautels2014parallelizing}.
+In short, this algorithm is a stochastic method of probing the experimental space, and is sometimes referred to as Bayesian optimization.
 It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
@@
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 
-In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising the TIR of a bacterial RBS}.
-We have used Gaussian Process regression to predict the TIR mean and standard deviation, and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
-To represent RBSs and capture the similarities between them, we chose to use the Weighted Degree Kernel with Shift method, which fits well into our prediction method.
-In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
-By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in this genetic context} by up to 34\%.
-Furthermore, we have generated an extensive library of diverse RBSs that can be used \hl{as a basis} for future studies.\\
+In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
+In the Learn phase, we used Gaussian process regression to predict the TIR mean and standard deviation, and in the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
+To represent RBSs and capture the similarities between them, we choose to use
+the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
+In building and testing, we performed our experiments using laboratory
+automation to increase their speed, reliability and reproducibility.
+Using our proposed workflow and testing only 450 RBS variants in 4 DBTL cycles,
+we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
+Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.\\
 
-In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.
-\hl{While maximisation could be the appropriate function for optimising RBSs or other small parts}, it may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
-This change could be achieved within the Bayesian optimisation framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objective method, for example \mbox{\cite{shu2020new}}.\\
-% \hl{
-% While we focus our optimisation framework on the RBS optimisation task, our framework has good potential to be generalised to multi-gene pathway design, as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}.
-For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process Regression (GPR) prediction model could be updated to a deep GPR approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\cite{shahriari2015taking}}.
-% }
+In this section we discuss the lessons learned for each phase of the
+DBTL cycle (Figure~\ref{fig: Flowchart}).
+In the Design phase, we focus on maximising the TIR of RBSs, by gradually moving our emphasis from exploration to exploitation as we progress through the design rounds.
+While maximisation could be the appropriate payoff function for optimising RBSs
+or other small parts, other payoffs may be better in more complex cases.
+For example when considering multi-gene metabolic engineering, maximising expression of individual genes may result in excessive metabolic burden.
+This change could be achieved within the bandit framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold~\cite{gotovos2013active}, or combining different goals into a multi-objective method~\cite{shu2020new}.\\
 
-Further, it is important to make a distinction between works focusing on the part itself like \mbox{\textcite{Hollerer2020}} and works where the part is optimised with a wider goal of strain optimisation like in \mbox{\textcite{Jervis2019}}.
-The part-centric optimisation, like the one considered in this paper, is an important task in synthetic biology since metabolic pathways are made up of individual parts and understanding their performance in separation allows for better strain design.
-While we focus our optimisation framework on the RBS optimisation task, this framework has good potential to be generalised to multi-gene pathway design as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}.
-For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\textcite{Shahriari2016}}.
+It is important to make a distinction between works focusing on the part itself like \textcite{Hollerer2020} and works where the part is optimised with a wider goal of strain optimisation like in \textcite{Jervis2019}.
+The part-centric optimisation, like the one considered in this paper, is an important task in synthetic biology since metabolic pathways are made up of individual parts and understanding their performance by themselves allows for better strain design.
+While we focus our machine learning guided DBTL cycle on RBS optimisation,
+this framework has good potential to be generalised to multi-gene pathway
+design as proposed by \textcite{hamedirad2019towards} and recently reviewed by \textcite{LAWSON2021}.
+For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity~\cite{Shahriari2016}.
 \\
 
-Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
-For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
+Our approach has shown that this combination of the two machine learning algorithms is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
+In the Learn phase, our approach has correctly identified the correlation of high guanine content in the RBS with high TIR.
 We have achieved this despite the relatively low Spearman scores for our predictions.
-Notably, other machine learning methods used in synthetic biology on small data sets were also shown to produce useful recommendations despite low prediction scores \cite{Radivojevic2020, Opgenorth2019}.
+This observation of useful recommendations despite low prediction scores corroborates recent evidence from other studies~\cite{Radivojevic2020, Opgenorth2019}.
 % Additionally, it has been shown that when the prediction is used for recommendation tasks like the one considered in this paper, ranking-based metric like Spearman correlation coefficient is a better evaluation metric
 % \cite{Schober2018,  Kang2019}.
-Finally, our predictor Gaussian Process regression model, compared with previously described calculators using a deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our Bandit algorithm to give better recommendations.\\
-
-We have found our approach of bringing machine learning and synthetic biology experts together very powerful.
-We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
-Synthetic biology is promoting standardised and normalised testing in biology and naturally pairs with machine learning, which can leverage the high quality biological data sets generated when the correct design rules are observed.
-The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
-On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.\\
+Finally, our predictor Gaussian Process regression model, compared with previously described calculators using a deterministic thermodynamic approach, is able to show the uncertainty of the predictions, which can be used by our bandit algorithm to give better recommendations.\\
 
@@
 
+We have found our approach of bringing machine learning and synthetic biology experts together very fruitful.
+We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
+Synthetic biology is promoting standardised and normalised testing in biology and naturally pairs with machine learning, which can leverage the high quality biological data sets generated when the correct design rules are observed.
+The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
+On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.\\
+
 There are still open questions that need to be addressed for applying machine learning in synthetic biology.
@@
 We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
-\hl{Similarly, the re-usability of the obtained data set is currently unknown. 
-The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures} \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} \hl{respectively) studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
-For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.}
-Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
-In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.\\
+Similarly, the re-usability of the obtained data set is currently unknown.
+The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} respectively).
+Studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
+For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.
+Finally, the optimal exploration-exploitation balance between rounds and samples is still an open theoretical question.
+%In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.
+\\
 
-In the future, we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
-However, it is important to reiterate that the complexity of the task quickly increases with the length of the sequence.
-This is because the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments.
+In future we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
+However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments.
 To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\

### 2021-12-27 paper/main.tex
mengyan's edit on discussion and conclusion

@@
 
-\section{Discussion and Conclusion}
+\section{Discussion}
 
-In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
-In the Learn phase, we used Gaussian process regression to predict the TIR mean and standard deviation, and in the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
-To represent RBSs and capture the similarities between them, we choose to use
-the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
-In building and testing, we performed our experiments using laboratory
-automation to increase their speed, reliability and reproducibility.
-Using our proposed workflow and testing only 450 RBS variants in 4 DBTL cycles,
-we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
-Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.\\
+In this section, we first revisit our overall experimental goal and challenges encountered in each phase of the
+DBTL cycle (Figure~\ref{fig: Flowchart}).
+We then link our proposed methods with related work and discuss the potential generalisation of our framework.
+% Then we will discuss the lessons learned for 
+We further discuss our design choices and open questions in this line of research.\\
 
-In this section we discuss the lessons learned for each phase of the
-DBTL cycle (Figure~\ref{fig: Flowchart}).
+Our goal is to show the power of the machine learning guided DBTL cycle on RBS optimisation.
+Our approach has shown that this combination of the two machine learning algorithms is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
+We focus on the part-centric optimisation \textcite{Hollerer2020}, which is an important task in synthetic biology. 
+Understanding how to design the individual parts also allows us to extend the framework to better strain design \textcite{Jervis2019}, where the part is optimised with a wider goal of strain optimisation.
+% metabolic pathways are made up of individual parts and understanding their performance by themselves allows for better strain design.
+% It is important to make a distinction between works focusing on the part itself like \textcite{Hollerer2020} and works where the part is optimised with a wider goal of strain optimisation like in \textcite{Jervis2019}
+\\
+
+The machine learning guided DBTL framework has good potential to be generalised to multi-gene pathway
+design as proposed by \textcite{hamedirad2019towards} and recently reviewed by \textcite{LAWSON2021}.
+For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity~\cite{Shahriari2016}.
+
 In the Design phase, we focus on maximising the TIR of RBSs, by gradually moving our emphasis from exploration to exploitation as we progress through the design rounds.
@@
 
-It is important to make a distinction between works focusing on the part itself like \textcite{Hollerer2020} and works where the part is optimised with a wider goal of strain optimisation like in \textcite{Jervis2019}.
-The part-centric optimisation, like the one considered in this paper, is an important task in synthetic biology since metabolic pathways are made up of individual parts and understanding their performance by themselves allows for better strain design.
-While we focus our machine learning guided DBTL cycle on RBS optimisation,
-this framework has good potential to be generalised to multi-gene pathway
-design as proposed by \textcite{hamedirad2019towards} and recently reviewed by \textcite{LAWSON2021}.
-For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity~\cite{Shahriari2016}.
-\\
-
-Our approach has shown that this combination of the two machine learning algorithms is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
 In the Learn phase, our approach has correctly identified the correlation of high guanine content in the RBS with high TIR.
@@
 
-We have found our approach of bringing machine learning and synthetic biology experts together very fruitful.
-We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
-Synthetic biology is promoting standardised and normalised testing in biology and naturally pairs with machine learning, which can leverage the high quality biological data sets generated when the correct design rules are observed.
-The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
-On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.\\
-
 There are still open questions that need to be addressed for applying machine learning in synthetic biology.
@@
 For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.
-Finally, the optimal exploration-exploitation balance between rounds and samples is still an open theoretical question.
+Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open theoretical question.
 %In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.
@@
 To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
+
+\section{Conclusion}
+
+In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
+In the Learn phase, we used Gaussian process regression to predict the TIR mean and standard deviation.
+To represent RBSs and capture the similarities between them, we choose to use
+the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
+In the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
+In the Build and Test phase, we performed our experiments using laboratory
+automation to increase their speed, reliability and reproducibility.
+Using our proposed workflow and testing 450 RBS variants in 4 DBTL cycles,
+we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
+Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.\\
+
+We have found our approach of bringing machine learning and synthetic biology experts together very fruitful.
+Synthetic biology is promoting standardised and normalised testing in biology and naturally pairs with machine learning, which can leverage the high quality biological data sets generated when the correct design rules are observed.
+The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
+On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.
+We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.\\
 

### 2021-12-28 paper/main.tex
add ref

@@
 % Then we will discuss the lessons learned for 
-We further discuss our design choices and open questions in this line of research.\\
+We further discuss our design choices and open questions in this line of research.
+We refer \cite{zhang_synbio2021} as more detailed discussion.\\
 
@@
 design as proposed by \textcite{hamedirad2019towards} and recently reviewed by \textcite{LAWSON2021}.
-For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity~\cite{Shahriari2016}.
+For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process regression prediction model could be updated to a deep Gaussian Process regression approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity~\cite{Shahriari2016}.\\
 
@@
 or other small parts, other payoffs may be better in more complex cases.
-For example when considering multi-gene metabolic engineering, maximising expression of individual genes may result in excessive metabolic burden.
-This change could be achieved within the bandit framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold~\cite{gotovos2013active}, or combining different goals into a multi-objective method~\cite{shu2020new}.\\
+For example when considering multi-gene metabolic engineering, maximising expression of individual genes may result in excessive metabolic burden, 
+which could be achieved within the bandit framework by 
+% changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold~\cite{gotovos2013active}, or 
+combining different goals into a multi-objective method~\cite{shu2020new}.
+We constrain our design space over 6-core parts of RBS sequences in this study, the design space will increase exponentially when designing a longer sequence. 
+The computational complexity of calculating and sorting acquisition functions (e.g.
+UCB scores) can be reduced by discretizing the space adaptively and hierarchically \cite{shekhar2018GPBanditDiscre, zhang2021GPOO}.\\
 
@@
 For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.
-Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open theoretical question.
+Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open question.
 %In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.

### 2022-01-05 paper/main.tex
Remove highlights, fix typos.

@@
 This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
-This cycle involves designing (D) new DNA sequences to achieve a desired property in computer-aided design software, then physically building (B) new constructed variants and testing (T) through a microplate reader in a laboratory.
+This cycle involves designing (D) new DNA sequences to achieve a desired property in computer-aided design software, then physically building (B) new constructed variants and testing (T) using an analytical instrument in a laboratory.
 Computer modelling can be used to learn (L) and predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
@@
 
-\hl{The exact position of the RBS in the sequence upstream of the protein coding sequence (CDS) can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.}
+The exact position of the RBS in the sequence upstream of the protein coding sequence (CDS) can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.
 In our investigation, we are using an RBS which is known to have a very high TIR when expressing GFP and is present in the pBb series of plasmids \cite{Lee2011}.
@@
 Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
-\hl{The implication of the fact that an RBS sequence chosen at random will have low TIR
+The implication of the fact that an RBS sequence chosen at random will have low TIR
 (as shown by UNI and PPM), is that most of the exploration will result in low TIRs.
@@
 whereas the exploration points give relatively low TIRs.
-Note that the exploration is necessary to expand our knowledge to the unknown parts of the design space and in effect allow us to exploit it better.}\\
+Note that the exploration is necessary to expand our knowledge to the unknown parts of the design space and in effect allow us to exploit it better.\\
 
@@
 
-\hl{To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design phase, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points.
-}
+To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design phase, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points.
 % GPR, as a non-parametric Bayesian regression algorithm, assumes the reward function is sampled from a Gaussian Process and has been shown with good performance with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}.
@@
 sub-linear regret upper bound) \cite{srinivas2012information, desautels2014parallelizing}.
-Figure \ref{fig: Scatterplot} \hl{shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.
+Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.
 As expected, the predictions in Round 0 were poor due to the use of approximate data.
 The predictions improved for the subsequent rounds,
-the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.}
+the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
 % Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
@@
 
-\hl{The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
+The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
 In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise.
@@
 This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
-Because of that, we provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient}
+Because of that, we provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient
 \cite{Schober2018,  Kang2019}.\\
@@
 However, this is useful information for future predictions as it allows us to understand the whole underlying space instead of focusing on local sub-optimal data points.
-\hl{In other words, we are intentionally sacrificing the accuracy of our predictions in each round to improve them in the future rounds}.
+In other words, we are intentionally sacrificing the accuracy of our predictions in each round to improve them in the future rounds.
 The effect of exploration of the space is the ability to find high TIR RBSs even with relatively low prediction performance.\\
@@
 
-\hl{We present some important characteristics of the tested RBSs in Table 1}.
+We present some important characteristics of the tested RBSs in Table 1.
 Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows the logo generated for all tested sequences).
@@
 
-\hl{Finally, while our strong sequences showed some affinity for the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs} (see Figure \ref{fig:structures}).
+Finally, while our strong sequences showed some affinity for the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs (see Figure \ref{fig:structures}).
 This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion, based on the previously reported literature \cite{Saito2020,EspahBorujeni2016}, that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\
@@
 Our approach has shown that this combination of the two machine learning algorithms is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
-We focus on the part-centric optimisation \textcite{Hollerer2020}, which is an important task in synthetic biology. 
-Understanding how to design the individual parts also allows us to extend the framework to better strain design \textcite{Jervis2019}, where the part is optimised with a wider goal of strain optimisation.
+We focus on the part-centric optimisation \cite{Hollerer2020}, which is an important task in synthetic biology. 
+Understanding how to design the individual parts also allows us to extend the framework to better strain design \cite{Jervis2019}, where the part is optimised with a wider goal of strain optimisation.
 % metabolic pathways are made up of individual parts and understanding their performance by themselves allows for better strain design.
@@
 Similarly, the re-usability of the obtained data set is currently unknown.
-The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} respectively).
-Studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
+The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} respectively) studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
 For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.

### 2022-01-05 paper/main.tex
combine discussion and conclusion together

@@
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 
-In this section, we first revisit our overall experimental goal and challenges encountered in each phase of the
+In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
+In the Learn phase, we used Gaussian process regression to predict the TIR mean and standard deviation.
+To represent RBSs and capture the similarities between them, we choose to use
+the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
+In the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
+In the Build and Test phase, we performed our experiments using laboratory
+automation to increase their speed, reliability and reproducibility.
+Using our proposed workflow and testing 450 RBS variants in 4 DBTL cycles,
+we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
+Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.
+In the rest of the section, we first revisit our overall experimental goal and challenges encountered in each phase of the
 DBTL cycle (Figure~\ref{fig: Flowchart}).
@@
 
-\section{Conclusion}
-
-In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
-In the Learn phase, we used Gaussian process regression to predict the TIR mean and standard deviation.
-To represent RBSs and capture the similarities between them, we choose to use
-the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
-In the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
-In the Build and Test phase, we performed our experiments using laboratory
-automation to increase their speed, reliability and reproducibility.
-Using our proposed workflow and testing 450 RBS variants in 4 DBTL cycles,
-we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
-Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.\\
+% \section{Conclusion}
 

### 2022-01-05 paper/methods.tex
cheng's changes in intro
Swapped final paragraphs.

@@
 %! TEX root=./main.tex
-\section{Methods}
+\section{Materials and Methods}
 
@@
 
-To find RBS sequences with the highest possible TIR score after a total number of rounds $N$, we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ is the TIR score of the 6-base core sequence \hl{ of the RBS} $\mathbf{x} \in \{A,C,G,T\}^{6}$.
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$, we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ is the TIR score of the 6-base core sequence of the RBS $\mathbf{x} \in \{A,C,G,T\}^{6}$.
 In each Round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
-$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and \hl{standard deviation}.
+$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and standard deviation.
 \\

### 2022-03-15 synbio_rbs/sim/run_sim.py
change to ratio_tir, add g rich clusters on tsne

@@ -149,7 +149,7 @@ with open(config.SAVED_IDX_SEQ_PATH, 'rb') as handle:
 
 idx_seq_dict = idx_seq['idx_seq_dict']
 
-for i in range(n_repeat):
+for i in range(80, n_repeat):
     rec_dfs = []
 
     df['Round'] = total_round

### 2022-03-15 paper/main.tex
Swapped final paragraphs.
Updated paper.
Updates from Overleaf

@@
 mentioned, as well as existing RBS calculators,
-do not yet provide uncertainty levels to guide exploration.
+do not yet provide uncertainty levels to guide exploration \hl{with the notable exception of ART} \cite{Radivojevic2020} .
 Furthermore large data sets may not be available for a genetic part of interest, for instance the RBS.
-Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both mean values and uncertainties in the Learn phase of DBTL.
+Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both their mean values and uncertainties \hl{of the predictions} in the Learn phase of DBTL.
 The predictions of characteristics of interest for
@@
 Out of these randomised sequences, only a few got close to the benchmark sequence's TIR and were still 20\% weaker than it is (Figure 3A).
-In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (see Supplementary Figure S1 and, for example, work by \mbox{\textcite{Jeschek2016}}).
+In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (\hl{Note, this applies only to the 7 bps on either side of the core; }see Supplementary Figure S1 and, for example, work by \mbox{\textcite{Jeschek2016}}).
 More representative would be sequences from other random groups (BPS-C, UNI, PPM), from which the best sequence achieved only about 65\% of the benchmark TIR.
@@
 
+\hl{We provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient}
+\cite{Schober2018,  Kang2019}.
 The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
@@
 However, since the predictions are only used for recommendations in the Design phase, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
+\hl{
+This is because, in the Design part, we select RBSs based on the ranks of the TIR predictions instead of numerical predictions.
+% in this task, we are maximising the TIR of RBS and so we are looking for RBS with high rank 1, regardless of its numerical TIR.
+}
 This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
-Because of that, we provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient
-\cite{Schober2018,  Kang2019}.\\
+\\
 
@@
 is also influenced by the exploration-exploitation balance in the Design phase.
-In each round, we select some data points for exploration of the areas in which we have few or none tested data points.
+In each round, we select some data points for exploration of the areas in which we have few or no tested data points.
 For those exploration points, since the predictor never learns their label distributions and patterns, there is no chance for the predictor to provide accurate predictions of TIR values,
@@
 The effect of exploration of the space is the ability to find high TIR RBSs even with relatively low prediction performance.\\
+
+\hl{To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator} \cite{Salis2009} \hl{and used our approach to simulate the search for the strongest designs in limited number of rounds.
+We show these results in Supplementary Figures} \ref{fig: Swarmplot and Quantplot Salis}, \hl{but in short, our approach was able to identify all the sequences with TIR higher than the benchmark's within the first 2 simulated rounds of experiments in 81 out of 100 runs and in 3 simulated rounds of experiments in 99 out of 100 runs.
+This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
+Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.}
 
@@
 This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
-This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with high TIR probability.\\
+This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with \hl{higher probability of having a high TIR value}.\\
 
@@
 
-Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high.
+Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high \hl{(we assume here our data set includes most of the highest RBSs for each distance - sequence combination)}.
 We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
-Figure \ref{fig:Library characteristics}B shows the edit distance is required for positive change in TIR for an RBS with TIR $>0.75$.
+Figure \ref{fig:Library characteristics}B shows the edit distance that is required for positive change in TIR for an RBS with TIR $>0.75$.
 For RBSs with high TIRs ($>1$), the minimum distance that is required to increase the TIR is 2, with edit distances between 2 and 5 giving similar results.
-For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.\\
+For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
+\\
 
@@
 As long as this does not impact targeted methods like machine learning-guided design, it does imply that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
-In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
+In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long \hl{(Figure} \ref{fig: Swarmplot and Quantplot}A \hl{shows comparison between efficiency of random versus directed sequence generation)}.
 In such cases, a directed process, like the one described here, should be strongly encouraged.
@@
 
-\section{Discussion and Conclusion}
+\section{Discussion}
 
 In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
-In the Learn phase, we used Gaussian process regression to predict the TIR mean and standard deviation.
+In the Learn phase, we used Gaussian process regression to predict the TIR mean and \hl{uncertainty of that prediction}.
 To represent RBSs and capture the similarities between them, we choose to use
@@
 For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
-Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available, while still providing the required results (optimised part).
+Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available \hl{(for example where there is no fluorescent readout available)}, while still providing the required results (optimised part).
 \\
@@
 We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
-Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurements?
+Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty \hl{quantification}?
 Thirdly, the generalisability of the method is unknown.
@@
 Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open question.
+\hl{In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance} \cite{Radivojevic2020}.
 %In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.
 \\
-
-In future we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
-However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments.
-To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
-
 % \section{Conclusion}
@@
 We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.\\
+
+In future we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
+However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments \hl{and so the percentage of the space that is measured will inevitably decrease, which might impact the expected recommendation results}.
+To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
 

### 2022-03-15 paper/methods.tex
Updates from Overleaf

@@
 %As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
-GPR can predict both the posterior mean and standard deviation of the RBS sequences in the design space. The posterior standard deviation represents the level of uncertainty for the prediction.
+\hl{GPR can predict the posterior mean of the RBS sequences in the design space.
+At the same time, it can provide the posterior standard deviation, which represents the level of uncertainty for the prediction.}
 
@@
 % In our design, the length of RBS sequence $L = 20$.
-The hyperparameters for kernel, including maximum substring length $\ell$, maximum shift length $S(l)$, and the standard deviation of GP noise $\alpha$ were chosen based on 10-repeat 5-fold cross validation.
+\hl{
+The hyperparameters for kernel, including maximum substring length $\ell = 6$, maximum shift length $S(l) = 1$, and the standard deviation of GP noise $\alpha = 1$ were chosen based on 10-repeat 5-fold cross validation.
+The values of hyperparameters considered were $\ell \in \{3,4,5,6\}$, $S(l) = \{0,1,2\}$, $\alpha = [0.5, 1, 2]$. }
 % ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).
@@
 where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at Round $t$ for the sequence $\mathbf{x}_i$.
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation \hl{(that is uncertainty of the prediction)} at Round $t$ for the sequence $\mathbf{x}_i$.
 We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at Round $t-1$.\\
@@
 Instead, we can recommend RBS sequences in a batch of size $m$.
-One naive approach is to
+%One naive approach is to
 %train GP on sequences with known TIR labels and

### 2022-03-15 paper/supplementary.tex
Updates from Overleaf

@@
 
-\subsection{Addition Results}
+\subsection{Additional Results}
 
@@
     \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
-   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}) and their corresponding projections to (0,1), (0,1) diagonal coloured in RdBu spectrum.
+   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}) and their corresponding projections to (0,1), (1,0) diagonal coloured in RdBu spectrum.
     We explain in details how the colours are generated in Section \ref{subsec: Exploration-exploitation visualisation} .
@@
    \label{fig:exploitationvsexplorationdetail}
+\end{figure}
+
+\subsection{Model evaluation on RBS calculator data}
+\label{subsec: Model evaluation on RBS calculator data}
+
+\hl{We have evaluated our model using data generated using the RBS calculator.} \cite{Salis2009} 
+\hl{To do this, we have generated TIR predictions for all 4096 RBS sequences using the RBS calculator ver. 2.0. 
+Next, we have applied our approach to this data set - first we have chosen 90 sequences at random and after teaching the algorithm using the obtained numbers we have generated 90 further sequences to be tested (using UCB for recommendations) and did this for a total of 5 rounds.
+We have repeated this experiment 100 times to increase our confidence in the results, by doing so we have found that the approach can identify all sequences with TIR above the benchmark in 2 rounds in 81 cases and in 3 rounds in 99 cases out of 100.}
+Figure \ref{fig: Swarmplot and Quantplot Salis} \hl{shows plot analogues to these in Figure} \ref{fig: Swarmplot and Quantplot} \hl{ but for one of the runs using the RBS calculator data.
+The apparent drop in TIR of the recommended sequences after round 1 is showing our processes efficacy - the best sequences are recommended early. }
+
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.35]{plots/Supplementary/swarmplot_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.25\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.35]{plots/Supplementary/quantplot_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.25\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.35]{plots/Supplementary/swarmplot_proj_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/histogram_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.42]{plots/Supplementary/tsneplot_salis.pdf}
+    \end{subfigure}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
+    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
+    \caption{
+    \textbf{Model evaluation on RBS calculator data.}
+    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
+    Random: randomly chosen sequences.
+    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
+    Dashed line is set to TIR of the benchmark sequence.
+    \textbf{B)} Line plot showing TIR obtained in a given quantile (Q) of results divided into groups as in A)
+    \textbf{C)} Exploitation v.s. Exploration for Bandit 0-3. Blue-hued points represent exploitation, those hued red represent exploration.
+    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
+    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
+    The area of the circle corresponds to the TIR value.
+    The TIR results in all subplots are shown normalised to the  benchmark sequence.}
+    \label{fig: Swarmplot and Quantplot Salis}
 \end{figure}
@@
     \label{fig:rawswarmplots.}
+\end{figure}
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.5]{plots/Supplementary/tsneplot_hamming.pdf}
+   \caption{\textbf{t-SNE plot based on Hamming distances.} This t-SNE plot has been drawn based on Hamming distances between sequences. 
+    }
+   \label{fig: TSNE hamming}
 \end{figure}

### 2022-04-08 paper/main.tex
fix ACS format

@@
-\documentclass{article}
+\documentclass{scrartcl}[2013/05/29]% Need at least an early state of KOMA-Script 3.12
 \usepackage[utf8]{inputenc}
-\usepackage[maxcitenames=1,style=numeric, sorting = none]{biblatex}
+\usepackage[mincitenames=2, maxcitenames=2,style=numeric, sorting = none]{biblatex}
 \usepackage{amsmath}
@@
 \renewcommand{\thesubfigure}{\Alph{subfigure}}
+\renewcommand\Authfont{\fontsize{14}{14.4}\selectfont}
+\renewcommand\Affilfont{\fontsize{11}{10.8}\itshape}
 
@@
 
-\author[1,2,3]{Zhang Mengyan}
-\author[4,5]{Holowko Maciej Bartosz}
-\author[4,5]{Hayman Zumpe Huw}
-\author[1,2,3,*]{Ong Cheng Soon}
-\affil[1]{Machine Learning and Artificial Intelligence Future Science Platform, CSIRO}
-\affil[2]{Department of Computer Science, Australian National University}
-\affil[3]{Data61, CSIRO}
-\affil[4]{Synthetic Biology Future Science Platform, CSIRO}
-\affil[5]{Land and Water, CSIRO}
+\author[1,2,3]{Mengyan Zhang}
+\author[4,5]{Maciej Bartosz Holowko}
+\author[4,5]{Huw Hayman Zumpe}
+\author[1,2,3,*]{Cheng Soon Ong}
+\affil[1]{Machine Learning and Artificial Intelligence Future Science Platform, CSIRO, Canberra ACT 2601}
+\affil[2]{Department of Computer Science, Australian National University, Canberra ACT 2601}
+\affil[3]{Data61, CSIRO, Canberra ACT 2601}
+\affil[4]{Synthetic Biology Future Science Platform, CSIRO, Canberra ACT 2601}
+\affil[5]{Land and Water, CSIRO, Canberra ACT 2601}
 \affil[*]{email: cheng-soon.ong@data61.csiro.au}
@@
 In short, this algorithm is a stochastic method of probing the experimental space, and is sometimes referred to as Bayesian optimization.
-It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
+It maximises the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
 It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\
@@
 
-To generate the data set from which the algorithm would learn, we decided to characterise a total of 450 RBS variants, little over 10\% of the whole experimental space.
+To generate the data set from which the algorithm would learn, we decided to characterise a total of 450 RBS variants, a little over 10\% of the whole experimental space. 
+About a quarter of the designs are experimental controls to provide a baseline for comparison.
 To fit our automated workflow, we divided the 450 variants into batches of 90, split into 4 design rounds.\\
@@
 
-In summary, out of 450 tested sequences, around 40\% (BPS-NC, BPS-C, UNI, PPM) were created through some kind of sequence randomisation.
-Out of these randomised sequences, only a few got close to the benchmark sequence's TIR and were still 20\% weaker than it is (Figure 3A).
-In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (\hl{Note, this applies only to the 7 bps on either side of the core; }see Supplementary Figure S1 and, for example, work by \mbox{\textcite{Jeschek2016}}).
-More representative would be sequences from other random groups (BPS-C, UNI, PPM), from which the best sequence achieved only about 65\% of the benchmark TIR.
+In summary, 120 out of 450 sequences (BPS-NC, BPS-C, UNI, PPM) are experimental controls created by sequence randomisation.
+Only a few of the random sequences were promising, but they were still 20\% weaker than the benchmark sequence's TIR (Figure 3A).
+In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (\hl{Note, this applies only to the 7 bps on either side of the core; }see Supplementary Figure S1 and work by \mbox{\textcite{Jeschek2016}}).
+Sequences from randomization of the core RBS (BPS-C, UNI, PPM) are more appropriate controls, from which the best sequence achieved only about 65\% of the benchmark TIR.
 These results show that generating a strong RBS sequence by random mutation is a non-trivial task, when the tested data set is relatively small.
-Contrasted with this, non-randomized, our Bandit-driven design batch gave much better results, with RBSs getting close to benchmark performance and even exceeding it.\\
+In contrast, our Bandit-driven design gives much better results, with RBSs getting close to benchmark performance and even exceeding it.\\
 
@@
 
-To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design phase, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points.
+In the Learn phase, we use a popular regression algorithm, Gaussian Process Regression (GPR).
+GPR has two benefits: (1) It provides predicted TIR and the confidence interval needed for the UCB algorithm in the Design phase, and (2) it provides a natural way to calculate similarities between genomic sequences via kernels.
 % GPR, as a non-parametric Bayesian regression algorithm, assumes the reward function is sampled from a Gaussian Process and has been shown with good performance with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}.
@@
 sub-linear regret upper bound) \cite{srinivas2012information, desautels2014parallelizing}.
-Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round.
+Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions (on an hold out test set) in each round.
 As expected, the predictions in Round 0 were poor due to the use of approximate data.
@@
 
-\hl{To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator} \cite{Salis2009} \hl{and used our approach to simulate the search for the strongest designs in limited number of rounds.
-We show these results in Supplementary Figures} \ref{fig: Swarmplot and Quantplot Salis}, \hl{but in short, our approach was able to identify all the sequences with TIR higher than the benchmark's within the first 2 simulated rounds of experiments in 81 out of 100 runs and in 3 simulated rounds of experiments in 99 out of 100 runs.
-This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
-Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.}
-
 \begin{figure}[!ht]
@@
     Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
-    X-axis and y-axis are the true, tested TIR and the predicted TIR respectively.
+    X-axis and y-axis are respectively: the true measured TIR, and the predicted TIR.
     In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design phase (i.e.
@@
     The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided in each plot's title on test data only.
-    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
+    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated, providing strong evidence that the predicted TIR are accurate.
     }
@@
 
-We present some important characteristics of the tested RBSs in Table 1.
+We present some important characteristics of the tested RBSs in Table 1, demonstrating the effectiveness of our Build and Test phases.
 Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows the logo generated for all tested sequences).
@@
 Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
-\begin{tabular}[c]{@{}c@{}}Efficiency of Bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
+\begin{tabular}[c]{@{}c@{}}Efficiency ratio of Bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
 Raw TIR range                                                                                     &      [4.93, 105.38]               \\ \hline
@@
 
-Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high \hl{(we assume here our data set includes most of the highest RBSs for each distance - sequence combination)}.
+We discovered an interesting phenomenon of large edit distances between RBSs with high TIRs.
+In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR \hl{(we assume here our data set includes most of the highest RBSs for each distance - sequence combination)}.
 We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
@@
 In such cases, a directed process, like the one described here, should be strongly encouraged.
-This observation is in line with approaches seen in other disciplines, e.g. protein engineering, where more directed changes yield better results than wide random changes \cite{Jackel2008}.\\
+This observation is in line with approaches seen in other disciplines, e.g. protein engineering, where more directed changes yield better results than random changes \cite{Jackel2008}.\\
 
@@
 
+\hl{To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator} \cite{Salis2009} \hl{and used our approach to simulate the search for the strongest designs in limited number of rounds.
+We show these results in Supplementary Figure} \ref{fig: Swarmplot and Quantplot Salis}, \hl{but in short, our approach was able to  identify above 75\% of the sequences stronger than the benchmark in up to 2 rounds and 99\% of them in up to 3 rounds.
+This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
+Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.}\\
+
 In this study we have limited the number of design rounds to four.
@@
 Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open question.
-\hl{In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance} \cite{Radivojevic2020}.
+\hl{In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance} \mbox{\cite{Radivojevic2020}}.
+%While Radivojevic \textcite{Radivojevic2020} provides a Bayesian ensemble framework, we focus on Gaussian process and provides methods to design batch recommendation better.}
 %In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.
@@
 \section*{Contributions}
-Zhang M.
-and Ong C.
+M.
+Zhang and C.
 S.
-designed and implemented the machine learning algorithms and workflow.
-Holowko M.
+Ong  designed and implemented the machine learning algorithms and workflow.
+M.
 B.
-and Hayman Zumpe H.
-have designed and performed the laboratory experiments.
-Holowko M.
+Holowko and H.
+Hayman Zumpe have designed and performed the laboratory experiments.
+M.
 B.
-and Ong C.
+Holowko and C.
 S.
-conceived and planned the project.
+Ong conceived and planned the project.
 All authors analysed the data, contributed to and reviewed the manuscript.
@@
 The authors declare no competing interests.
+
+\section*{Support Information}
+\begin{itemize}
+    \item A detailed text description of machine learning methods, including Gaussian Process regression, choices of kernels, batch recommendation, experimental settings, additional results and model evaluation on RBS calculator data.
+    \item Figure S1-S12.
+    \item Table S1-S2.
+\end{itemize}
 
@@
 \setcounter{figure}{0}
+\setcounter{table}{0}
 \makeatletter
 \renewcommand{\thefigure}{S\@arabic\c@figure}
+\renewcommand{\thetable}{S\@arabic\c@table}
 \makeatother
+\pagenumbering{arabic}% resets `page` counter to 1
+\renewcommand*{\thepage}{S\arabic{page}}
 \appendix
@@
 \end{document}
-However large data sets may not be available for a genetic part of interest, for instance the RBS.
+% However large data sets may not be available for a genetic part of interest, for instance the RBS.
 Hence there is a need to develop methods for training predictors on smaller data sets in the Learn phase of DBTL.\\

### 2022-04-08 paper/methods.tex
fix ACS format

@@
 For Bandit designs in Round 0, since we only had access to a limited number of data points from the literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6 bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7 bp flanking sequences both upstream and downstream of the core sequence.
-To represent the RBS sequences in subsequent rounds, we use the \textit{weighted degree kernel with shift} (WDS) \cite{ratsch_rase_2005_wds,Ben-Hur2008} to specify the kernel function of $GP$.  
+\hl{
+Among various string kernels which have been studied by \mbox{\textcite{Ben-Hur2008}} and we will introduce briefly in Supplementary \mbox{\ref{sec: Choices of Kernels}}, we chose the one with the best reported performance, the \textit{weighted degree kernel with shift} (WDS) \mbox{\cite{ratsch_rase_2005_wds, Ben-Hur2008}},
+to represent the RBS sequences in subsequent rounds and specify the kernel function of $GP$.  }
 WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  

### 2022-04-08 paper/supplementary.tex
fix ACS format

@@
 %! TEX root=./main.tex
-\textbf{Supplementary}
+\title{Supplementary of Machine learning guided\\
+batched design of a bacterial Ribosome Binding Site}
+\maketitle
 
@@
 \subsection{Choices of Kernels}
+\label{sec: Choices of Kernels}
 
@@
 
-\hl{We have evaluated our model using data generated using the RBS calculator.} \cite{Salis2009} 
-\hl{To do this, we have generated TIR predictions for all 4096 RBS sequences using the RBS calculator ver. 2.0. 
-Next, we have applied our approach to this data set - first we have chosen 90 sequences at random and after teaching the algorithm using the obtained numbers we have generated 90 further sequences to be tested (using UCB for recommendations) and did this for a total of 5 rounds.
-We have repeated this experiment 100 times to increase our confidence in the results, by doing so we have found that the approach can identify all sequences with TIR above the benchmark in 2 rounds in 81 cases and in 3 rounds in 99 cases out of 100.}
-Figure \ref{fig: Swarmplot and Quantplot Salis} \hl{shows plot analogues to these in Figure} \ref{fig: Swarmplot and Quantplot} \hl{ but for one of the runs using the RBS calculator data.
-The apparent drop in TIR of the recommended sequences after round 1 is showing our processes efficacy - the best sequences are recommended early. }
+\hl{To test the reliability of our proposed model, we have generated TIR predictions for all 4096 RBS sequences using the RBS  calculator ver.
+2.0 \mbox{\cite{Salis2009}} and evaluated our model using that data.
+We normalised the TIR values by dividing them by the TIR of the benchmark sequence, and show the TIR ratio mapping between the RBS calculator and TIRs obtained experimentally for this work in Figure \mbox{\ref{fig:TIR ratio salis micro}}.
+We observe that the RBS calculator values are not well aligned to the measured values, which underlines the need for development of new prediction and design methods. 
+Among all 4096 RBS sequences, 38 RBS were predicted to have higher TIR than the benchmark sequence by the RBS calculator.
+We call those sequences the \textit{optimal set}, and the goal is to find as many sequences from the optimal set as permitted by the budget (450).
+
+To do this, we have applied our approach to this data set - first we have chosen 90 sequences at random and after teaching the algorithm using the obtained numbers we have generated 90 further sequences to be tested (using UCB for recommendations) and did this for a total of 4 rounds.
+We have repeated this experiment 100 times to increase our confidence in the results, by doing so we have found that our model can identify above 75\% of the optimal set in up to 2 rounds and 99\% of them in up to 3 rounds.
+We reported the detailed statistics for the 100 runs in Table \mbox{\ref{table: Important statistics for the 100 synthetic experiments}}.
+Figure \mbox{\ref{fig: Swarmplot and Quantplot Salis}} shows plot analogues to these in Figure \mbox{\ref{fig: Swarmplot and Quantplot}}  but for one of the runs using the RBS calculator data.
+Our algorithm has successfully found most of the sequences in optimal set in the first and second rounds, thus the recommended TIRs have an apparent drop in the following rounds (as there are no good sequences left to recommend).
+This might suggest that it might be valuable to add an additional stopping condition for the DBTL cycle - decreasing round performance.}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{paper/plots/Supplementary/TIR_ratio_salis_micro.png}
+    \caption{TIR ratio of RBS calculator ver. 2.0 \cite{Salis2009} plotted against TIR ratios obtained experimentally in this work. The red point represents the benchmark sequence.}
+    \label{fig:TIR ratio salis micro}
+\end{figure}
 
@@
 
-% \newpage
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_0_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_1_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_2_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_3_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/salis_scatter_abc1_FF_4.pdf}
+    \end{subfigure}
+    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation) on the RBS calculator data}.
+The scatter plots A-D show the performance of our prediction algorithm calculated after each round using the RBS calculator data.
+    E shows the performance on all collected data.
+    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
+    X-axis and y-axis are the TIR predicted by the RBS calculator and the TIR predicted by our model respectively.
+    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided in each plot's title on test data only.
+    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
+    }
+    \label{fig: Scatterplot_Salis}
+\end{figure}
+
+\begin{table}[h!]
+\caption{\textbf{Important statistics for the 100 synthetic experiments run using the RBS calculator data.}
+Each round contains 90 samples, where Round 0 is uniformly random sampled from the whole space, and Rounds 1-4 follow the bandits algorithm. 
+In column \textit{Percent of the Optimal Set Found} and \textit{Standard Deviation}, we reported the average percent of the \textit{optimal set} found in each round and the standard deviation of this value averaged over the 100 experiments.
+In the last column, we reported the average percent of the (\textit{optimal set}) found in total after each round. 
+We can observe that our algorithm can find the majority of the (\textit{optimal set}) sequences within three rounds.
+}
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+                  \textbf{Round} & \textbf{\begin{tabular}[c]{@{}c@{}}Percent of\\ the Optimal Set Found\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Standard\\ Deviation\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Total Percent of\\ the Optimal Set Found\end{tabular}} \\ \hline
+ \textbf{0}  & 0.022  & 0.026 & 0.022  \\ \hline
+  \textbf{1}  & 0.744  & 0.085 & 0.766  \\ \hline
+   \textbf{2}  & 0.222  & 0.077 & 0.988  \\ \hline
+    \textbf{3}  & 0.012  & 0.026 & 1.000  \\ \hline
+     \textbf{4}  & 0.000  & 0.000 & 1.000  \\ \hline
+
+\end{tabular}
+\label{table: Important statistics for the 100 synthetic experiments}
+\end{table}
+
+\newpage
 

### 2022-04-08 synbio_rbs/sim_lit/run_sim.py
lit result: rec size 40

@@ -123,7 +123,7 @@ else:
 
 # setting
 
-rec_size = 20 # in each round, we recommend 90 RBS sequences
+rec_size = 40 # in each round, we recommend 90 RBS sequences
 l = 6 # maximum kmer as 6
 s = 1 # maximum shift as 1
 alpha = 2 # GPR noise parameter, get from cross validation
@@ -141,12 +141,12 @@ kernel_over_all_flag = True  # we keep the setting the same as our last round
 #     kernel_over_all_flag = True
 #     df = df_round23
 
-n_repeat = 2
+n_repeat = 100
 total_round = 5
 
 all_recs = []
 # save_folder_path = './sim_results'
-save_folder_path = '/home/v-mezhang/blob/Solaris/sim_results/lit/topucb'
+save_folder_path = '/data4/u6015325/Solaris/sim_results/lit/topucb/'
 print('save folder path: ', save_folder_path)
 if not os.path.exists(save_folder_path):
     os.makedirs(save_folder_path)
@@ -158,6 +158,7 @@ for i in range(whole_size):
     index_list.append(i)
     seq_list.append(df.loc[i, 'RBS'])
     idx_seq_dict[df.loc[i, 'RBS']] = i
+    idx_seq_dict[i] = df.loc[i, 'RBS']
     df.loc[i, 'idx'] = i
 # print(idx_seq_dict)
 save_dict = {}
@@ -188,9 +189,10 @@ for i in range(1,n_repeat):
 
         if design_round == 0:
             # random sample
-            print('Design round {}: randomly generate {} recommendations.'.format(design_round, rec_size))
+            random_rec_size = rec_size # * 4
+            print('Design round {}: randomly generate {} recommendations.'.format(design_round, random_rec_size))
 
-            rec_idxs = np.random.choice(range(whole_size), size = rec_size, replace=False)
+            rec_idxs = np.random.choice(range(whole_size), size = random_rec_size, replace=False)
             # rec first rec size
             # rec_idxs = range((0 + i) * rec_size, (1+i)* rec_size)
             # print(rec_idxs)

### 2022-04-08 synbio_rbs/src/kernels_for_GPK.py

@@ -115,18 +115,19 @@ class String_Kernel(Kernel):
             kernel_path = config.SAVED_KERNEL_PATH + self.kernel_name_para + '.pickle'
             # SAVED_KERNEL_PATH = '../data/saved_kernel/' is specified in config.py
 
+            with open(config.SAVED_IDX_SEQ_PATH, 'rb') as handle:
+                seq = pickle.load(handle)['seq_list']
+
             if os.path.exists(kernel_path):
                 print('Load saved kernel matrix...')
                 print(self.kernel_name_para)
                 with open(kernel_path, 'rb') as handle:
                     SAVED_KERNEL = pickle.load(handle)
-                self.kernel_all_normalised = putback_kernel_matrix(SAVED_KERNEL)
+                self.kernel_all_normalised = putback_kernel_matrix(SAVED_KERNEL, size = len(seq))
                 # TODO: need to check whether the saved kernel is generated with same parameters
             else:
                 print('No saved kernel matrix. Calculate and save kernel matrix...')
                 # seq = np.load(SAVED_IDX_SEQ_PATH)['seqList']
-                with open(config.SAVED_IDX_SEQ_PATH, 'rb') as handle:
-                    seq = pickle.load(handle)['seq_list']
                 features = Rewards_env(np.asarray(seq).reshape(len(seq), 1), 'label').embedded
 
                 self.kernel_all = self.cal_kernel(features, features)

### 2022-04-13 paper/main.tex
Updated paper.

@@
 We discovered an interesting phenomenon of large edit distances between RBSs with high TIRs.
-In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR \hl{(we assume here our data set includes most of the highest RBSs for each distance - sequence combination)}.
+In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR \hl{(we assume here that our data set includes most of the strongest RBSs for each distance - sequence combination)}.
 We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
@@
 
-\section*{Support Information}
-\begin{itemize}
-    \item A detailed text description of machine learning methods, including Gaussian Process regression, choices of kernels, batch recommendation, experimental settings, additional results and model evaluation on RBS calculator data.
-    \item Figure S1-S12.
-    \item Table S1-S2.
-\end{itemize}
+\section*{Supporting Information}
+
+A detailed text description of the machine learning methods, including Gaussian Process regression, choices of kernels, batch recommendation, experimental settings, additional results and model evaluation on RBS calculator data; Figures S1-S12; Tables S1-S2.
 
@@
 \printbibliography
+
+\newpage
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[scale=0.5]{paper/plots/Supplementary/TOC.pdf}
+    \caption{\textbf{For Table of Contents Only.} }
+    \label{fig: TOC}
+\end{figure}
 

### 2022-04-19 paper/main.tex
Final commit of paper with highlights.

@@
 \usepackage[utf8]{inputenc}
-\usepackage[mincitenames=2, maxcitenames=2,style=numeric, sorting = none]{biblatex}
+\usepackage[mincitenames=2, maxcitenames=2, maxbibnames=99, style=numeric, sorting = none]{biblatex}
 \usepackage{amsmath}
@@
 Optimisation of gene expression levels is an essential part of the organism design process.
-Fine control of this process can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
-Unfortunately, design of specific genetic parts can still be challenging due to lack of reliable design methods.
-To address this problem, we have created a machine learning guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed using relatively small, high-quality data sets.
-We used Gaussian Process Regression for the Learn phase of cycle and the Upper Confidence Bound multi-armed bandit algorithm for the Design of genetic variants to be tested in vivo.
+Fine control of this process can be achieved by engineering transcription and translation control elements, including the ribosome binding site (RBS).
+Unfortunately, the design of specific genetic parts remains to be challenging due to the lack of reliable design methods.
+To address this problem, we have created a machine learning guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to demonstrate how small genetic parts can be reliably designed using relatively small, high-quality data sets.
+We used Gaussian Process Regression for the Learn phase of the cycle and the Upper Confidence Bound multi-armed bandit algorithm for the Design of genetic variants to be tested in vivo.
 We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for reliable data generation.
-Notably, by Testing a total of 450 RBS variants in four DBTL cycles, we experimentally validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
+Notably, by Testing a total of 450 RBS variants in four DBTL cycles, we have experimentally validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
 Overall, our results show that machine learning is a powerful tool for designing RBSs, and they pave the way towards more complicated genetic devices.\\
@@
 One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
-A central challenge is part design, understood as modifying the sequence of a genetic part for it to meet specific requirements.
+A central challenge is part design, which is understood as modifying the sequence of a genetic part for it to meet specific requirements.
 Genetic parts are the units which are ultimately combined into more complex genetic circuits that produce desired functions in the target organisms.
-This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
-This cycle involves designing (D) new DNA sequences to achieve a desired property in computer-aided design software, then physically building (B) new constructed variants and testing (T) using an analytical instrument in a laboratory.
-Computer modelling can be used to learn (L) and predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
+This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where a given genetic part or organism is continually improved through an iterative process.
+This cycle involves Designing (D) new DNA sequences to achieve a desired property in computer-aided design software, then physically Building (B) new constructed variants and Testing (T) using an analytical instrument in a laboratory.
+Computer modelling can be used to Learn (L) and predict the characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
 Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBSs) \cite{Xia1998,Chen2013,Reeve2014}.
-However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.\\
+However, de novo design of small genetic elements is challenging due to unknown relationships between their sequence and performance.\\
 
@@
 
-In the Design phase of the DBTL cycle, designers often have to fine tune characteristics of parts to give the resulting strains their desired properties.
+In the Design phase of the DBTL cycle, designers often fine tune the characteristics of parts to give the resulting strains their desired properties.
 The ability to predict a characteristic of a genetic part (for example using a machine
@@
 possible variants to Build.
-For instance, if the goal is to increase the yield of a protein, the designer might increase the TIR of the RBS responsible for translation of that protein.
+For instance, to increase the yield of a protein, increasing the TIR of the RBS responsible for translation of that protein could be targeted.
 % Hence the Design phase of TIR has two steps:
 % first to predict the TIR value for each potential RBS sequence,
-Hence the goal for the Design phase of TIR
-is to select a small set of RBS sequences from the design space (i.e. all possible DNA sequences) based on the predictions from Learn phase.
-However since the computational predictions are not perfectly accurate,
+Hence, the goal of the Design phase of TIR is to select a small set of RBS sequences from the design space (i.e. all possible DNA sequences) based on the predictions from Learn phase.
+However, since the computational predictions are not perfectly accurate,
 there is uncertainty about which potential RBS sequence has the desired
@@
 ones to exploit the knowledge modelled by the computational predictions.
-But the designer may also want to explore regions of DNA design space
-where the predictions are highly uncertain (e.g. as measured by
+But the designer may also want to explore regions of DNA design space where the predictions are highly uncertain (e.g. as measured by
 the predicted TIR standard deviation) and could yield high payoff.
-Hence a major challenge of the Design phase is to balance
-the exploitation and exploration,
-which is addressed by one class of machine learning approaches
-called multi-armed bandits \cite{lattimore2020bandit}.
-As we will see in this paper, multi-armed bandits are well suited
-to solving the challenge of recommending a small set of RBS sequences
-in the Design phase of the DBTL cycle.
+Therefore, a major challenge of the Design phase is to balance the exploitation and exploration, which is addressed by one class of machine learning approaches called multi-armed bandits \cite{lattimore2020bandit}.
+As we will see in this paper, multi-armed bandits are well suited to solving the challenge of recommending a small set of RBS sequences in the Design phase of the DBTL cycle.
 \\
 
-One way to generate predicted TIR values is to use the existing RBS calculators.
-\textcite{Reeve2014} surveys three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
-Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
-This may be due to number of reasons:
+One way to generate predicted TIR values is to use existing RBS calculators.
+\textcite{Reeve2014} surveyed three main RBS calculators, all of which predict the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}.
+Predictions reported from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
+This may be caused by:
 i) they rely on calculations of free energies which can be difficult to estimate with high precision,
@@
 vastly improving the DBTL cycle's performance~\cite{Camacho2018,Radivojevic2020,LAWSON2021}.
-This leverages the exponential increase in experimental data from synthetic biology~\cite{Freemont2019}
+This work leverages the exponential increase in experimental data produced in synthetic biology~\cite{Freemont2019}
 to improve predictors in the Learn phase of the DBTL cycle.
@@
 Two types of ML algorithms are applied in the DBTL cycle.
-For the Learn phase, the goal is to train a predictor that will predict the protein expression level by learning from the logged data.
-We used a Bayesian, non-parametric regression algorithm, called
-Gaussian Process Regression (GPR) \cite{Rasmussen2004} in our pipeline.
-In addition to providing a predicted mean TIR, GPR also provides
-uncertainty estimates via its predicted standard deviation.
-GPR has been shown to  perform well with small amounts of trainig data in
-biological prediction tasks~\cite{Rasmussen2004, desautels2014parallelizing}.
-For the Design phase, the goal is to find a policy to recommend RBS sequences
-to query in batches so that we can identify the optimized RBS sequences
-within a given budget.
+In the Learn phase, the goal is to train a predictor that will predict the protein expression level by learning from the logged data.
+We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression (GPR) \cite{Rasmussen2004} in our pipeline.
+In addition to providing a predicted mean TIR, GPR also provides uncertainty estimates via its predicted standard deviation.
+GPR has been shown to  perform well with small amounts of trainig data in biological prediction tasks~\cite{Rasmussen2004, desautels2014parallelizing}.
+The goal of the Design phase is to find a policy to recommend RBS sequences to query in batches so that we can identify the optimized RBS sequences within a given budget.
 We use a version of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} to learn the policy.
-The policy uses the outputs of the GPR predictions to recommend useful batches for the Build and Test phases.
-We demonstrate that laboratory automation methods in the Build and Test phases
-result in high quality TIR data, that is well suited for training
-machine learning methods in the subsequent Learn phase.
+The policy uses the outputs from the GPR predictions to recommend useful batches for the Build and Test phases.
+We demonstrate that laboratory automation methods in the Build and Test phases result in high quality TIR data, that is well suited for training machine learning methods in the subsequent Learn phase.
 The two types of algorithms cooperate with each other and provide powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
@@
 Our DBTL workflow, which uses machine learning to optimise protein expression, is shown in Figure \ref{fig: Flowchart}.
-Build and Test are driven chiefly by choices made by human researchers and the use of automated methods.
+The Build and Test phases are driven chiefly by choices made by human researchers and the use of automated methods.
 Machine learning algorithms are applied in Learn and Design.
@@
 
-The exact position of the RBS in the sequence upstream of the protein coding sequence (CDS) can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.
+The exact position of the RBS in the sequence upstream of the protein coding sequence (CDS) can be hard to pinpoint.
+However, most previous studies place the RBS in the 20 bp directly preceding the coding sequence.
 In our investigation, we are using an RBS which is known to have a very high TIR when expressing GFP and is present in the pBb series of plasmids \cite{Lee2011}.
 This template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
-Since this is the sequence against which new RBS sequences will be benchmarked,
+As this is the sequence against which new RBS sequences will be benchmarked,
 we will refer to this sequence as the \textit{benchmark sequence} hereafter.
@@
 
-In our genetic design, the investigated RBS controls expression of green fluorescent protein (GFP).
+In our genetic design, the investigated RBS controls the expression of green fluorescent protein (GFP).
 By controlling expression of a fluorescent protein with the RBS we can quickly assess the perceived relative TIR by measuring fluorescence of cells harbouring the expression vector over time.
 Finally, the mRNA is transcribed from an IPTG-inducible promoter pLlacO-1.
-Inducible expression allows one to synchronise the start of the GFP expression in all the cultures with addition of IPTG.
+Inducible expression allows synchronization of the start of the GFP expression in all the cultures by addition of IPTG.
 Since standardisation and comparative studies should be done in as similar genetic context as possible, the design of this device has been deliberately kept simple to make such studies easier \cite{Beal2021}.\\
@@
 (as shown by UNI and PPM), is that most of the exploration will result in low TIRs.
-Our results confirm that RBSs with high TIRs tend to come from exploitation of the design space,
+These results confirm that RBSs with high TIRs tend to come from exploitation of the design space,
 whereas the exploration points give relatively low TIRs.
@@
 Each RBS is located on the plot according to its distance from other RBSs as calculated by our weighted degree kernel with shift (see Section \mbox{\ref{sec: method prediction with kernel}}).
-The RBSs recommended by Bandit groups have covered the majority of the design space.
-Additionally, a number of clusters were especially targeted by our recommendation algorithm.
+The RBSs recommended by Bandit groups cover the majority of the design space.
+Additionally, a number of clusters were especially targeted by the recommendation algorithm.
 For example, the circled clusters labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
@@
 Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions (on an hold out test set) in each round. 
-As expected, the predictions in Round 0 were poor due to the use of approximate data.
-The predictions improved for the subsequent rounds,
-the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
+As expected, the use of approximate data lead to poor predictions in Round 0.
+The predictions improved for the subsequent rounds, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.
 % Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
@@
 
-Additionally, the
+The
 % obtained values of the coefficient of determination and Spearman correlation coefficient
@@
 is also influenced by the exploration-exploitation balance in the Design phase.
-In each round, we select some data points for exploration of the areas in which we have few or no tested data points.
+In each round, some data points were selected for exploration of the areas in which we have few or no tested data points.
 For those exploration points, since the predictor never learns their label distributions and patterns, there is no chance for the predictor to provide accurate predictions of TIR values,
@@
     \end{subfigure}
-    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of our prediction algorithm calculated after each round.
+    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of the prediction algorithm calculated after each round.
     Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
@@
 test data, blue).
-    In D), we train our predictor on randomly chosen 80\% of all available data, and test on the remaining 20\% data.
+    In D), the predictor was trained on a randomly chosen 80\% subset of all available data, and tested on the remaining 20\% data.
     The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided in each plot's title on test data only.
@@
 
-This means that as the TIRs of examined RBSs increase, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement.
-As long as this does not impact targeted methods like machine learning-guided design, it does imply that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
+This means that as the TIRs of examined RBSs increase, exploring sequences which are increasignly dissimilar to the current candidates tends to give more meaningful improvement.
+As long as this does not impact targeted methods like machine learning-guided design, this implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
 In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long \hl{(Figure} \ref{fig: Swarmplot and Quantplot}A \hl{shows comparison between efficiency of random versus directed sequence generation)}.
@@
 or other small parts, other payoffs may be better in more complex cases.
-For example when considering multi-gene metabolic engineering, maximising expression of individual genes may result in excessive metabolic burden, 
+For example, when considering multi-gene metabolic engineering, maximising expression of individual genes may result in excessive metabolic burden, 
 which could be achieved within the bandit framework by 
@@
 
-In the Learn phase, our approach has correctly identified the correlation of high guanine content in the RBS with high TIR.
+In the Learn phase, our approach has correctly identified the correlation between high guanine content in the RBS and high TIR.
 We have achieved this despite the relatively low Spearman scores for our predictions.
@@
 In this study we have limited the number of design rounds to four.
-There were a number of reasons for this, including limitations on time and money, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
-There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however this has to be put in the context of limited resources.
+There were a number of reasons for this, including limitations on time and money, but also because the results obtained showed that we have achieved our goal of generating very strong RBS designs.
+There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however, this has to be put in the context of limited resources.
 For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
@@
 
-There are still open questions that need to be addressed for applying machine learning in synthetic biology.
+There are still open questions that need to be addressed for the application of machine learning in synthetic biology.
 Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.

### 2022-04-19 paper/supplementary.tex
Final commit of paper with highlights.

@@
     \begin{align}
-        k^C(\mathbf{x}_i, \mathbf{x}_j) &= <\phi^C(\mathbf{x}_i), \phi^C(\mathbf{x}_j)>\\
-        &= \left( \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left( \phi(\mathbf{x}_j) - \frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
-        &= \phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j) - \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \phi(\mathbf{x}_j) - \phi(\mathbf{x}_i)^T \left(\frac{1}{n} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right) + \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left(\frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
-        &= k(\mathbf{x}_i, \mathbf{x}_j) - \frac{1}{n} \sum_{s=1}^n k(\mathbf{x}_s, \mathbf{x}_j) - \frac{1}{n'} \sum_{s'=1}^{n'} k(\mathbf{x}_i, \mathbf{x}_{s'}) + \frac{1}{n^2} \sum_{s = 1}^n \sum_{s'=1}^{n'} k(\mathbf{x}_s, \mathbf{x}_{s'})
+        k^C(\mathbf{x}_i, \mathbf{x}_j) =& <\phi^C(\mathbf{x}_i), \phi^C(\mathbf{x}_j)>\\
+        =& \left( \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left( \phi(\mathbf{x}_j) - \frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        =& \phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j) - \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \phi(\mathbf{x}_j) - \phi(\mathbf{x}_i)^T \left(\frac{1}{n} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right) +\\
+        &\left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left(\frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        =& k(\mathbf{x}_i, \mathbf{x}_j) - \frac{1}{n} \sum_{s=1}^n k(\mathbf{x}_s, \mathbf{x}_j) - \frac{1}{n'} \sum_{s'=1}^{n'} k(\mathbf{x}_i, \mathbf{x}_{s'}) + \frac{1}{n^2} \sum_{s = 1}^n \sum_{s'=1}^{n'} k(\mathbf{x}_s, \mathbf{x}_{s'})
     \end{align}
@@
 And only update $y_{t^{\prime}}=f\left(\boldsymbol{x}_{t^{\prime}}\right)+\varepsilon_{t^{\prime}} \text { for } t^{\prime} \in\{\mathrm{fb}[t]+1, \ldots, \mathrm{fb}[t+1]\}$ at the end of each batch ($\mathrm{fb}[t]<\mathrm{fb}[t+1]$).
-This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right]$, while the posterior variance decreases.
+This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $$\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right],$$
+while the posterior variance decreases.
 

### 2022-04-20 paper/main.tex
fit acs bib, merge main, method, supp into one file

@@
 \documentclass{scrartcl}[2013/05/29]% Need at least an early state of KOMA-Script 3.12
+% \documentclass[journal=jacsat,manuscript=article]{achemso}
 \usepackage[utf8]{inputenc}
-\usepackage[mincitenames=2, maxcitenames=2, maxbibnames=99, style=numeric, sorting = none]{biblatex}
+\usepackage[mincitenames=2, maxcitenames=2, maxbibnames=99, style=chem-acs, sorting = none]{biblatex}
 \usepackage{amsmath}
@@
 trade off by using predictive uncertainty. The machine learning predictors
-mentioned, as well as existing RBS calculators,
-do not yet provide uncertainty levels to guide exploration \hl{with the notable exception of ART} \cite{Radivojevic2020} .
+mentioned, as well as existing RBS calculators,do not yet provide uncertainty levels to guide exploration with the notable exception of ART \cite{Radivojevic2020} .
 Furthermore large data sets may not be available for a genetic part of interest, for instance the RBS.
-Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both their mean values and uncertainties \hl{of the predictions} in the Learn phase of DBTL.
+Hence there is a need to develop methods for training predictors on smaller data sets and generate predictions for both their mean values and uncertainties of the predictions in the Learn phase of DBTL.
 The predictions of characteristics of interest for
@@
 
-\input{methods.tex}
+%! TEX root=./main.tex
+\section{Materials and Methods}
+
+\subsection{Laboratory experimental design}
+
+\subsubsection{BUILD: Construction of genetic devices}
+
+\textbf{Plasmid Design.}
+The pBbB6c-GFP plasmid was used for all our designs.
+This plasmid contains the GFP mut3b CDS, expression of which can be induced by the addition of isopropyl $\beta$-D-1-thiogalactopyranoside (IPTG) (Merck, Darmstadt, Germany, catalogue no. I5502).
+The original RBS for the GFP CDS was replaced using a combination of PCR and isothermal assembly.
+Primer sequences and the assembly strategy were generated using the Teselagen DESIGN software (Teselagen Biotechnology, San Francisco, CA).
+
+\textbf{PCR.}
+PCR amplification of the cloning inserts was done using Q5 High-Fidelity 2X Master Mix (NEB, Ipswich, MA, catalogue no. M0492L).
+20 $\mu$L reactions were prepared by dispensing 1 $\mu$L of each 10 $\mu$M reverse primer into the wells of a 96-well PCR plate using the Echo liquid handler (Beckman Coulter, Brea, CA).
+A mastermix consisting of polymerase premix, plasmid DNA template (pBbB6c, 5-10 ng per reaction), and the single 10 $\mu$M forward primer was prepared and dispensed using the FeliX liquid handler (Analytik Jena, Jena, Germany) or electronic multi-channel pipette.
+Reactions were run using Touchdown PCR or standard PCR cycling methods in C1000 thermal cyclers (Bio-Rad, Hercules, CA).
+Capillary electrophoresis of PCR products was performed using the ZAG DNA Analyzer system (Agilent Technologies, Santa Clara, CA).
+2 $\mu$L of each PCR reaction was electrophoresed using the ZAG 130 dsDNA Kit (75-20000 bp) or ZAG 110 dsDNA Kit (35-5000 bp) (Agilent Technologies, catalogue no.
+ZAG-110-5000; ZAG-130-5000).
+ZAG sample plates were prepared using the Sciclone G3 liquid handler (Perkin Elmer, Waltham, MA).
+ProSize Data Analysis Software (Agilent Technologies) was used to generate gel images from the sample chromatograms, and amplicon sizes were estimated by reference to the upper and lower DNA markers spiked into each sample and a DNA ladder run in well H12 of each sample plate.
+
+\textbf{Isothermal DNA Assembly.}
+Constructs were assembled using NEBuilder HiFi DNA Assembly Master Mix (NEB, catalogue no. E2621L).
+Reactions consisting of approximately equal amounts of the common fragment and the variable fragment were prepared using the FeliX liquid handler or electronic multi-channel pipette, to a final volume of 5 or 10 $\mu$L.
+Assemblies were run in the thermal cycler for 1 hour at 50$^{\circ}$C, followed by an infinite hold step at 4$^{\circ}$C.
+Finally, samples were incubated with of 50 nL of DpnI (NEB, catalogue no. R0176S) at 37$^{\circ}$C for 90 minutes to degrade any residual template DNA.
+
+\textbf{\textit{E. coli} transformation.}
+The DH5$\alpha$ cell line (Thermo Fisher Scientific, Waltham, MA, catalogue no.
+18265017) was made chemically competent using the Mix $\&$ Go \textit{E.
+coli} Transformation Kit $\&$ Buffer Set (Zymo Research, Irvine, CA, catalogue no.
+T3001).
+20 $\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
+Plates of cells were thawed on a -20$^{\circ}$C cold block before 3 $\mu$L of the assembly product was added and mixed using the FeliX liquid handler.
+Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 (12 x 8) grid on Omnitrays containing LB (BD, Franklin Lakes, NJ, catalogue no.
+244610) and 15 g/L agar (Merck, catalogue no.
+A1296) with 34 $\mu$g/mL chloramphenicol (Merck, catalogue no.
+C1919).
+Plates were incubated overnight at 37$^{\circ}$C. Cells were plated using the FeliX liquid handler.
+
+\textbf{Automated colony picking and culturing.}
+A PIXL colony picker (Singer Instruments, Roadwater, United Kingdom) was used to select individual colonies from the transformation plates using the 490-510 nm (cyan) light filter.
+Each selected colony was used to inoculate 1 mL of selective medium in a 2 mL square well 96 plate.
+They were then cultured overnight at 37$^{\circ}$C with shaking (300 rpm).
+
+\textbf{Glycerol stock preparation.}
+100 $\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, Gillman, Australia, catalogue no.
+GA010) and 100 $\mu$L of overnight culture were combined in the wells of a 96 deep (2 mL) round well plate using the FeliX liquid handler or electronic multi-channel pipette
+They were then sealed with a 96-well silicone sealing mat and transferred to a -80$^{\circ}$C freezer.
+
+\textbf{Sequencing}
+Strains that gave GFP fluorescence intensity readings similar to that of the original RBS were selected for sequence confirmation by capillary electrophoresis sequencing (CES) (Macrogen, Inc., Seoul, South Korea).
+The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
+Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN, Hilden, Germany, catalogue no. 27106) according to the manufacturer's instructions.
+Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek, Winooski, VT) and all fell in the range of 100-200 ng/$\mu$L.
+Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5'-CGATATAGGCGCCAGCAA-3') that binds approximately 150 bp upstream of the RBS.
+Reads were aligned with the template sequence in the Teselagen software.
+
+\subsubsection{TEST: Culture analysis}
+
+\textbf{Test strain culture.}
+Overnight cultures (six biological (restarted every time from glycerol stock) replicates for each batch) were started by inoculating 1 mL of LB medium supplemented with 34 $\mu$g/mL chloramphenicol with 2 $\mu$L of the glycerol stock in a 96 deep (2 mL) round well plate, using the FeliX liquid handler or electronic multi-channel pipette.
+Cultures were incubated at 37$^{\circ}$C with shaking (300 rpm) for 17 hours. 
+The following morning, 20 $\mu$L of each overnight culture was added to 980 $\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in a 96 deep (2 mL) round well plate.
+After 90 minutes, cultures were induced with IPTG to a final concentration of 0.5 mM.
+This was done by transferring 1.0 $\mu$L of 0.1 M IPTG to each well of a flat-bottom clear polystyrene 96-well plate using the Echo liquid handler, then adding 300 $\mu$L of culture to each well using an electronic multi-channel pipette.
+
+\textbf{Microplate spectrophotometry.}
+The plates were monitored in the Cytation 5 microplate reader immediately after the addition of IPTG.
+Cytation 5 acquisition and incubation/shaking settings were as follows: length of run: 8 hours; interval: 10 min; continuous orbital shake at 237 rpm and slow orbital speed; excitation wavelength: 490/10 nm; emission wavelength: 515/10 nm; bottom read; gain: 60; read height: 7 mm; read speed: Sweep.
+
+\subsection{Machine learning experimental design}
+
+Two types of machine learning algorithms have to be applied to drive the experimental design workflow shown in Figure \ref{fig: Flowchart}.
+% To automatically design the RBS sequences in batch using machine learning, we can logically divided the workflow into two parts: 
+One type of machine learning algorithm is a prediction algorithm (\textbf{LEARN}), which helps us learn the function of TIR with respect to RBS sequence.
+The other type of machine learning algorithm is a recommendation algorithm (\textbf{DESIGN}), which recommends RBS sequences to query (test) in each round (in the sense of a single pass through the DBTL cycle) based on the predictions from LEARN.
+\\
+
+% 1) \textbf{LEARN}: A regression algorithm which takes the RBS sequences as input features and TIR scores as labels, trains on sequences with known labels and returns the predicted TIR scores and the respective confidence intervals.
+% 2) \textbf{DESIGN}: An online learning approach which recommends the RBS sequences based on the predicted TIR scores and confidence intervals.
+In Round $t$, prediction and design are based on the results obtained in all previous rounds. 
+Our implementation of the machine learning algorithms was tested in Python 3.6 and used the scikit-learn library \cite{scikit-learn}. 
+In the following paragraphs, we describe our machine learning pipeline (which is applicable to design Rounds 1-3, denoted as Bandit 1-3) in the following order: data pre-processing, prediction and kernels, and finally recommendation.
+The pipeline for our 0th Round of machine learning designs, denoted as Bandit-0, is reported in Supplementary \ref{sec: ML design pipeline}.
+
+\subsubsection{Data pre-processing}
+\label{sec: method data pre-procesing}
+
+For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as an averaged sum of GFP fluorescence divided by Optical Density at 600 nm (OD600) of the culture over time (calculated using values from all the measured data points), following the formula:
+
+$$\mathnormal{TIR=\frac{\sum_{t_{min}}^{t_{max}}\frac{GFP(t)}{OD_{600}(t)}}{\Delta t}}$$
+
+where ${t_{min}}$ is the time at which the $\frac{GFP}{OD_{600}}$ value is at its minimum denoting the start of the log growth phase, ${t_{max}}$ is the time at which the logarithmic phase ends (on average, for our samples that was 4 hours from ${t_{min}}$ so that was the time point used for TIR calculations for all samples), ${\Delta t}$ is the difference between ${t_{max}}$ and ${t_{min}}$.\\
+
+% REVIEW:
+% \mengyan{Table XXX shows an example of our raw data. }
+% We consider both the label (TIR) pre-processing and kernel pre-processing.
+For label pre-processing, we first adjust the TIR values in each round using the round-wise reference values.
+
+The reference value is the TIR of the benchmark sequence that is run in triplicate in each round.
+Specifically, in Round $t$, we subtract the TIR mean of the benchmark RBS measured in Round $t$ from all TIR values measured in the same round, for each replicate separately. 
+We then normalised the data by performing a logarithm transformation and standardisation on the adjusted TIR label for each replicate separately.
+After normalisation, each replicate has zero mean and unit variance.
+Furthermore, we also normalised the kernel matrix used for prediction by centring and unit-norm normalisation, which is reviewed in details in Supplementary \ref{supp: Normalisation of Kernel}.
+% Since the design space is known, we consider the transductive approach \cite{} for the kernel pre-processing with centering and unit-norm normalisation.
+
+\subsubsection{Prediction: Gaussian Process Regression with String Kernel}
+\label{sec: method prediction with kernel}
+
+To find RBS sequences with the highest possible TIR score after a total number of rounds $N$, we consider our experimental design problem as a sequential optimisation of an unknown reward function $f: \mathcal{D} \rightarrow \mathbb{R}$, where $\mathcal{D}$ is the set containing all RBS sequence points in the design space, and $f(\mathbf{x})$ is the TIR score of the 6-base core sequence of the RBS $\mathbf{x} \in \{A,C,G,T\}^{6}$.
+In each Round $t$, we choose a set of $m$ points $\mathcal{S}_t \subset \mathcal{D}$ and observe the function value at each point in the selected set $\mathcal{S}_t$, i.e.
+$y_i = f(\mathbf{x}_i) + \epsilon$, for all $i \in \mathcal{S}_t$, where $\epsilon$ is the Gaussian random noise with unknown mean and standard deviation.
+\\
+% This noise is influenced by the accuracy of the RBS predictor and other experimental sources of interference (e.g. time, temperature, operator, etc.). 
+% REVIEW:
+% \mengyan{we should discuss measurement noise somewhere, maybe with a sample example for the dataset we used.}
+
+For the regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004,srinivas2012information, romero_navigating_2013}.
+We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
+%As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
+\hl{GPR can predict the posterior mean of the RBS sequences in the design space.
+At the same time, it can provide the posterior standard deviation, which represents the level of uncertainty for the prediction.}
+
+We review the predictions of GPR in Supplementary \ref{supp: GPR}.\\
+% We assume the observations are noisy, so even when a data point in observed then confidence interval is still bigger than 0.
+% As we can see, the true function is in the confidence interval when there are data points are observed.
+
+% \begin{figure}[t]
+%     \centering
+%     \includegraphics[scale=0.7]{plots/Prediction with Initial Observations.pdf}
+%     \caption{Gaussian Process Regression Example.
+This plot shows the GPR prediction with confidence interval based on 4 initial observations.
+The confidence interval are shown in predicted mean $\pm$ 1.96 standard deviation.}
+%     \label{fig: Gaussian Process Regression Example.}
+% \end{figure}
+
+The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
+For Bandit designs in Round 0, since we only had access to a limited number of data points from the literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6 bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7 bp flanking sequences both upstream and downstream of the core sequence.
+\hl{
+Among various string kernels which have been studied by \mbox{\textcite{Ben-Hur2008}} and we will introduce briefly in Supplementary \mbox{\ref{sec: Choices of Kernels}}, we chose the one with the best reported performance, the \textit{weighted degree kernel with shift} (WDS) \mbox{\cite{ratsch_rase_2005_wds, Ben-Hur2008}},
+to represent the RBS sequences in subsequent rounds and specify the kernel function of $GP$.  }
+WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
+WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
+The maximum substring length is specified by $\ell$.
+The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
+Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
+This could be useful when there is a shift between two sequences.
+For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. \\
+% there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.
+
+We now define WDS kernel.
+Let $\mathbb{I}$(A) be the indicator function, which equals 1 if $A$ is true and 0 otherwise.
+Then $\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime)$ indicates whether two substrings of length $d$ match, between $\mathbf{x}$ starting from position $l+s$ and $\mathbf{x}^\prime$ starting from position $l$.
+This is similarly done for $\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)$.
+By having these two terms considering substrings of two sequences with starting positions differing by $s$ characters, the WDS can measure shifted positional information.
+When $s = 0$, the kernel function counts the matches with no shift between sequences.
+Let $\mathbf{x}, \mathbf{x}^\prime$ be two RBS sequences with length $L$, the WDS kernel is defined as
+\begin{align}
+        k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime)
+        %&= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        %\left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
+        = \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+        \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + \mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
+\end{align}
+where
+$\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting parameter over the position in the
+sequence, where we chose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$ ($L = 20$ in our design).
+
+$S(l)$ determines the maximum shift at position $l$. 
+Furthermore, we normalise the kernel with centring and unit-norm in terms of all sequences in the design space. 
+% In our design, the length of RBS sequence $L = 20$.
+\hl{
+The hyperparameters for kernel, including maximum substring length $\ell = 6$, maximum shift length $S(l) = 1$, and the standard deviation of GP noise $\alpha = 1$ were chosen based on 10-repeat 5-fold cross validation.
+The values of hyperparameters considered were $\ell \in \{3,4,5,6\}$, $S(l) = \{0,1,2\}$, $\alpha = [0.5, 1, 2]$. }
+% ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).
+
+%  For recommendations, we consider the \textit{Upper Confidence Bound (UCB)} type algorithms.
+%  As one popular type of the bandit algorithms \cite{lattimore2020bandit}, the UCB type of algorithms are based on the \textit{optimism in the face of uncertainty},
+ %provide various approaches to sequential design where an agent adaptively chooses one or more options among several actions based on certain policies.
+In our work we used the Upper Confidence Bound version of that algorithm, which is based on the \textit{optimism in the face of uncertainty}.
+The UCB algorithm, as the name suggests,
+%  which basically select RBS sequences with the maximum upper confidence bound constructed by the sum of the predicted mean and $n$ standard deviation ($n > 0$), i.e.
+$\operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_t(\mathbf{x}_i) + \beta_t \sigma_t(\mathbf{x}_i)\right)$,
+%     where $\beta_t$ is a hyperparameter balancing the exploitation and exploration,
+%     $\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation at round $t$ for the sequence $\mathbf{x}_i$.
+
+\subsubsection{Recommendation: Upper Confidence Bound multi-armed Bandit algorithm}
+
+To recommend the RBS sequences to query in the next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm
+\cite{lattimore2020bandit}
+% \cite{srinivas2012information}
+, which provides  $\textit{exploration-exploitation balance}$.
+On one hand, UCB exploits the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
+on the other hand, UCB explores the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
+More precisely, the UCB algorithm selects RBS sequences $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at Round $t$, i.e.
+\begin{align}
+\label{Eq: GPUCB}
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
+\end{align}
+where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation \hl{(that is uncertainty of the prediction)} at Round $t$ for the sequence $\mathbf{x}_i$.
+We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at Round $t-1$.\\
+
+Since experimentally labelling sequences is time-consuming, it is unrealistic to recommend sequences sequentially (i.e.
+one-by-one) and then wait for the label to be tested and used to improve the model.
+Instead, we can recommend RBS sequences in a batch of size $m$.
+%One naive approach is to
+%train GP on sequences with known TIR labels and
+However, this approach may end up recommending similar sequences in the same local maximum (e.g. $x = 2, x =2.5$ in this example). 
+Since GPR assumes similar sequences would have similar labels (e.g.
+by knowing $x=2$ we can gain information about $x=2.5$ as well), we prefer to not waste time and money on labelling sequences with high similarities in the same batch.
+To counter this, we use a batch recommendation strategy that is designed to avoid recommending such highly similar sequences in the same batch, described below. \\
+
+% \begin{figure}[t]
+% \centering
+% \subfloat{\label{main:c}\includegraphics[scale=.4]{plots/Recommend_2_Data_Point_Top_2_UCB.pdf}}\par\medskip
+% \begin{minipage}{.5\linewidth}
+% \centering
+% \subfloat{\label{main:a}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_1st_Data_Point.pdf}}
+% \end{minipage}%
+% \begin{minipage}{.5\linewidth}
+% \centering
+% \subfloat{\label{main:b}\includegraphics[scale=.4]{plots/GP-BUCB_Recommend_2nd_Data_Point.pdf}}
+% \end{minipage}
+% \caption{Batch Recommendation.
+We use the batch size of 2, with 5 initial observations.
+The design (recommendation) space is 24 uniformly distributed points in the range [-2,10], i.e.
+{-2, -1.5, -1, ..., 9.5, 10}.
+% The confidence interval are shown with predicted mean of $\pm$ 1.96 standard deviation.
+% (a) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, constructed with GP predictions.
+% (b)(c) GP-BUCB recommendations. (b) shows the first recommended sequence, (c) shows the new predicted confidence interval and the second recommendation based on that.}
+% \label{fig:batch rec}
+% \end{figure}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=.4]{plots/Main_Paper/Recommend_2_Data_Points_Top_2_UCB.pdf}
+    \end{subfigure}
+    \vskip\baselineskip
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_1st_Data_Point.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Main_Paper/GP-BUCB_Recommend_2nd_Data_Point.pdf}
+    \end{subfigure}
+    \caption{\emph{Batch Recommendation Illustration.} We use the batch size of 2, with 5 initial observations.
+The design space is 24 uniformly distributed points in the range [-2,10], i.e.
+{-2, -1.5, -1, ..., 9.5, 10}.
+    The confidence intervals are shown with predicted mean $\pm$ 1.96 standard deviation.
+    (A) Top UCB recommendations. The recommendations are 2 data points with top UCB scores, chosen with GP predictions.
+    (B)(C) Batch UCB recommendations. (B) shows the first recommended sequence, (C) shows the new predicted confidence interval and the second recommendation based on that updated interval.}
+    \label{fig:batch rec}
+\end{figure}
+
+A key property of Gaussian Process regression is that the predicted standard deviation depends only on features, not on the labels. 
+% We illustrate this fact by comparing the prediction with the true function value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (New Observation)}) and the prediction with the predicted value of the new recommended point (Figure \ref{fig: Recommend 1 Data Point (Predicted Mean)}).
+% As we can see, the predicted confidence interval is the same for the two plots.
+One can make use of this property to design a batch upper confidence bound (BUCB) algorithm \cite{desautels2014parallelizing}.
+The strategy here is to recommend RBS sequences one-by-one by sequentially adding the newly recommended RBSs' predicted means (without testing them) into the training data and updating UCB scores.\\
+
+% Figure \ref{fig: Recommend 2 Data Points: GP-BUCB.} shows an example of batch UCB recommendation. 
+% The plot shows the predicted mean and standard deviation after observing 4 data points. 
+As illustrated in Figure \ref{fig:batch rec}B, the algorithm recommends the data point with maximum UCB score based on the predictions over initial 5 observations.  
+We then add the recommended data point ($x = 2$) into the training data set with the predicted mean of that point as the label (note it is not the true label), and update the predicted standard deviation, then we finally update the UCB scores.
+The second data point is then recommended based on the new UCB scores.   
+Figure \ref{fig:batch rec}C shows that since we assume we have observed $x = 2$, the new predicted standard deviation of the data points in design space around $x =2$  decreases, so instead of recommending a similar data point $x = 2.5$, the algorithm recommends $x = 8$, which is in another local maximum design area.
+In this way, the batch recommendation can potentially cover more local maximum areas than the sequential design. 
+In summary, the exploration efficiency is improved since the recommended sequences in one batch will tend to be in different design areas so that the information gain is maximised. 
+In our design, we have set the batch size to 90, to fit the experimental batch.
+Finally, we set $\beta_t = 2$ for Round 0-3 and $\beta_t = 0$ for the last round to allow for more exploitation.
 
@@
 
-\section*{Contributions}
+\section*{Author Contribution}
 M.
@@
 \appendix
-\input{supplementary}
+
+%! TEX root=./main.tex
+\title{Supplementary of Machine learning guided\\
+batched design of a bacterial Ribosome Binding Site}
+\maketitle
+
+\section{Machine Learning Methods}
+
+In this section, we describe the machine learning methods in details.
+
+\subsection{Gaussian Process Regression}
+\label{supp: GPR}
+
+A \textit{Gaussian process} is a collection of random variables, any finite number of which have a joint Gaussian distribution. 
+We define mean function $\mu(\mathbf{x})$  and covariance function $k(\mathbf{x}, \mathbf{x}^\prime)$ of a real process $f(\mathbf{x})$ as
+\begin{align}
+    \mu(\mathbf{x}) &= \mathbb{E}[f(\mathbf{x})]\\
+    k(\mathbf{x}, \mathbf{x}^\prime) &= \mathbb{E}[(f(\mathbf{x}) - \mu(\mathbf{x}))(f(\mathbf{x}^\prime) - \mu(\mathbf{x}^\prime))].
+\end{align}
+
+A Gaussian process is specified by its mean function and covariance function as $f(\mathbf{x}) \sim \mathcal{G} \mathcal{P}\left(\mu(\mathbf{x}), k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)\right)$.
+We consider the case where the observations are noisy, i.e.
+$(X, \mathbf{y}) = \{(\mathbf{x}_i, y_i)| i = 1, \dots, n\}$, where label $y_i = f(\mathbf{x}_i) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$.
+The Gaussian noise is independent identically distributed, and the covariance of the prior on the noisy observations is then $\operatorname{cov}\left(y_{p}, y_{q}\right)=k\left(\mathbf{x}_{p}, \mathbf{x}_{q}\right)+\alpha^{2} \delta_{p q}$,
+where $\delta_{pq}$ is a Kronecker delta which is one if $p = q$ and zero otherwise.
+It is equivalent to a diagonal matrix $\alpha^2 I$ added on the kernel matrix evaluated on the training points.
+
+For $n_\ast$ test points $X_\ast$, we assume the prior over the functions values as a random Gaussian vector $\mathbf{f}_\ast \sim \mathcal{N}(\mathbf{0}, K(X_\ast, X_\ast))$.
+Then the joint distribution of the observed target values and the function values at the test points under the prior as 
+\begin{align}
+    \left[\begin{array}{l}\mathbf{y} \\ \mathbf{f}_{*}\end{array}\right] \sim \mathcal{N}\left(
+        \left[\begin{array}{cc}
+        \bm{\mu}  \\
+        \bm{\mu}_\ast
+        \end{array}\right],
+    \left[\begin{array}{cc}K(X, X)+\alpha^{2} I & K\left(X, X_{*}\right) \\ K\left(X_{*}, X\right) & K\left(X_{*}, X_{*}\right)\end{array}\right]\right)
+\end{align}
+where
+$\bm{\mu}, \bm{\mu}_\ast$ are mean vectors for $\mathbf{y}, \mathbf{f}_{*}$ respectively,
+$K(X, X_\ast)$ denotes the $n \times n_\ast$ covariance/Kernel matrix evaluated at all pairs of training and testing points, similarly for other kernel matrices.
+Then the posterior of the test points (i.e.
+predictive distributions) is given by the conditional distribution $\mathbf{f}_\ast | X, \mathbf{y}, X_\ast \sim \mathcal{N}(\bar{\mathbf{f}}_\ast, cov(\mathbf{f}_\ast))$, where
+\begin{align}
+   \overline{\mathbf{f}}_{*} & \triangleq \mathbb{E}\left[\mathbf{f}_{*} \mid X, \mathbf{y}, X_{*}\right]=
+   \bm{\mu}_\ast + K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} (\mathbf{y} - \bm{\mu})\\
+   \label{Eq: predicted variance}
+   \operatorname{cov}\left(\mathbf{f}_{*}\right) &=K\left(X_{*}, X_{*}\right)-K\left(X_{*}, X\right)\left[K(X, X)+\alpha^{2} I\right]^{-1} K\left(X, X_{*}\right) 
+\end{align}
+For noisy test targets $\mathbf{y}_\ast = f(\mathbf{x}_\ast) + \epsilon$ with $\epsilon \sim \mathcal{N}(0, \alpha^2)$, we can compute the predictive distribution by adding $\alpha^2 I$ to the variance term $cov(\mathbf{f}_\ast)$ in Eq.
+(\ref{Eq: predicted variance}),
+i.e. $\overline{\mathbf{y}}_{*} = \overline{\mathbf{f}}_{*}$, $\operatorname{cov}\left(\mathbf{y}_{*}\right) = \operatorname{cov}\left(\mathbf{f}_{*}\right) + \alpha^2 I$.
+
+% We now introduce the marginal likelihood (or evidence) $p(\mathbf{y}|X)$, is which the integral of the likelihood times the prior 
+% \begin{align}
+%     p(\mathbf{y} \mid X)=\int p(\mathbf{y} \mid \mathbf{f}, X) p(\mathbf{f} \mid X) d \mathbf{f}
+% \end{align}
+
+\subsection{Choices of Kernels}
+\label{sec: Choices of Kernels}
+
+The choice of covariance function is critical for the performance of Gaussian process regression. 
+Except the weighted degree kernel with shift described in Section \ref{sec: method prediction with kernel}, we show two closely related string kernels tested in this study below.
+
+\begin{itemize}
+    \item \textit{Spectrum Kernel.}
+    \begin{align}
+        k_\ell^{\text{Spec}}(\mathbf{x}, \mathbf{x}^\prime) =\left\langle\phi_{\ell}^{\mathrm{Spec}}(\mathbf{x}), \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right)\right\rangle = \phi_{\ell}^{\mathrm{Spec}}(\mathbf{x})^T \phi_{\ell}^{\mathrm{Spec}}\left(\mathbf{x}^{\prime}\right).
+    \end{align}
+     where $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$. We denote the number of letters in the alphabet as $|\Sigma|$. 
+     In our case, $\Sigma = \{A,C,G,T\}$ and $|\Sigma| = 4$.
+    $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps the sequence $\mathbf{x}$ into a $|\Sigma|^\ell$ dimensional feature space, where each dimension is the count of the number of one of the $|\Sigma|^\ell$ possible strings $s$ of length $\ell$.
+    For example, if $\mathbf{x} = ACACAG$ and $\ell = 3$, then $\phi_{\ell}^{\mathrm{spec}}(\mathbf{x})$ maps $\mathbf{x}$ into $4^3$ dimensional feature space, where each dimension corresponds to $AAA, AAC, AAT, AAG, ACA, ..., TTT$.
+Among those, the dimension corresponds to $ACA, CAC, CAG$ are 2,1,1 and others are 0.
+    Let $X, X^\prime$ be two metrics which include $n$ sequences, and $\Phi_\ell^{Spec}(X) \in \mathbb{R}^{n \times |\Sigma|^{\ell}}$, then the spectrum kernel over metrics is 
+    \begin{align}
+         K_\ell^{\text{Spec}}(X, X^\prime) = \Phi_{\ell}^{\mathrm{Spec}}(X) \Phi_{\ell}^{\mathrm{Spec}}\left(X^{\prime}\right)^T.
+    \end{align}
+
+    % \item \textit{Sum of Spectrum Kernel,} considers weighted sum over different parts of the string.
+
+    % \item \textit{Mixed Spectrum Kernel,} considers weighted sum over different substring length, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$,
+    %     \begin{align}
+    %         k_\ell^{MixedSpec}(\mathbf{x}, \mathbf{x}^\prime) 
+    %         = \sum_{d=1}^{\ell} \beta_d k_d^{Spec}(\mathbf{x}, \mathbf{x}^\prime)
+    %     \end{align}
+    \item \textit{Weighted Degree Kernel,} considers positional information. WD kernel counts the match of kmers at corresponding positions in two sequences.
+    For sequences with fixed length $L$ and weighted degree kernel considers substrings starting at each position $l = 1, ..., L$, with $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}$, \\
+    \begin{align}
+        k_\ell^{WD}(\mathbf{x}, \mathbf{x}^\prime) 
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l:l+d]}^\prime)\\
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \phi_d^{Spec}(\mathbf{x}_{[l:l+d]})^T \phi_d^{Spec}(\mathbf{x}_{[l:l+d]}^\prime)\\
+        &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \mathbb{I}(\mathbf{x}_{[l:l+d]} = \mathbf{x}_{[l:l+d]}^\prime),
+    \end{align}
+    where $\mathbb{I}(\text{true}) = 1$ and 0 otherwise.
+
+    % \item \textit{Weighted Degree Kernel With Shift.}
+    % \begin{align}
+    %     k_\ell^{WDS}(\mathbf{x}, \mathbf{x}^\prime) 
+    %     &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+    %     \left(k_d^{Spec}(\mathbf{x}_{[l+s:l+s+d]}, \mathbf{x}_{[l:l+d]}^\prime) + (k_d^{Spec}(\mathbf{x}_{[l:l+d]}, \mathbf{x}_{[l+s:l+s+d]}^\prime)\right)\\
+    %     &= \sum_{d=1}^{\ell} \beta_d \sum_{l=1}^{L-d+1} \gamma_l \sum_{s = 0, s + l \leq L}^{S(l)} \delta_s
+    %     \left(\mathbb{I}(\mathbf{x}_{[l+s:l+s+d]} = \mathbf{x}_{[l:l+d]}^\prime) + (\mathbb{I}(\mathbf{x}_{[l:l+d]}= \mathbf{x}_{[l+s:l+s+d]}^\prime)\right),
+    % \end{align}
+    % where $\beta_d = \frac{2(\ell - d + 1)}{\ell(\ell+1)}, \delta_s = \frac{1}{2(s+1)}$, $\gamma_l$ is a weighting over the position in the
+    % sequence, where we choose to use a uniform weighting over the sequences, i.e. $\gamma_l = 1/L$. $S(l)$ determines the shift
+    % range at position $l$.
+\end{itemize}
+
+% \textbf{From kernel to distance}:
+% $$d(\mathbf{x}, \mathbf{x}^\prime) = \sqrt{k(\mathbf{x}, \mathbf{x}) + k(\mathbf{x}^\prime, \mathbf{x}^\prime) - 2 k(\mathbf{x}, \mathbf{x}^\prime)} $$
+
+\subsubsection{Normalisation of Kernel}
+\label{supp: Normalisation of Kernel}
+
+As part of data pre-processing,
+the range of all features should be normalised so that each feature contributes approximately proportionately to the predictive model. 
+The kernel matrix is represented by the inner product of the underlying feature vectors, it needs to be normalised before being used in the downstream regression models. 
+Up-scaling (down-scaling) features can be understood as down-scaling (up-scaling) regularizers such that they penalise the features less (more).
+
+Here we consider two approaches for kernel normalisation: centering and unit norm. 
+We will show how to convert the normalisation in terms of feature vectors to normalisation in terms of kernel matrices. 
+As defined before, consider $\mathbf{x}, \mathbf{x}^\prime$ are two RBS sequences in $\mathcal{D}$ over an alphabet $\Sigma$.
+We denote $\phi(\mathbf{x})$ as a column feature vector of sequence $\mathbf{x}$, 
+where a feature function $\phi: \mathbf{x} \rightarrow \mathbb{R}^d$,
+with $d$ as the dimension of features.
+One example of $\phi$ can be $\phi_\ell^{spec}(\mathbf{x})$ in the spectrum kernel.
+Recall the corresponding kernel can be defined as $k(\mathbf{x}, \mathbf{x}^\prime) = \phi(\mathbf{x})^T \phi(\mathbf{x}).$
+
+Assume there is total of $n$ sequences in the data $X$ ($n'$ sequences in the data $X'$). 
+We illustrate centering and unit norm normalisation below.
+
+\begin{itemize}
+    \item Centering. 
+    Defining the mean vector as $\bar{\Phi}(X) = \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s) \in \mathbb{R}^d$, the centered feature vector $\phi^C(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    \begin{align}
+        \phi^{C}(\mathbf{x}_i) = \phi(\mathbf{x}_i) - \bar{\Phi}(X) = \phi(\mathbf{x}_i) - \frac{1}{n'} \sum_{s = 1}^{n'} \phi(\mathbf{x}_s).
+    \end{align}
+    The corresponding centering kernel value between $\mathbf{x}_i$ and $\mathbf{x}_j$ is then 
+    \begin{align}
+        k^C(\mathbf{x}_i, \mathbf{x}_j) =& <\phi^C(\mathbf{x}_i), \phi^C(\mathbf{x}_j)>\\
+        =& \left( \phi(\mathbf{x}_i) - \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left( \phi(\mathbf{x}_j) - \frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        =& \phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j) - \left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \phi(\mathbf{x}_j) - \phi(\mathbf{x}_i)^T \left(\frac{1}{n} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right) +\\
+        &\left( \frac{1}{n} \sum_{s = 1}^n \phi(\mathbf{x}_s)\right)^T \left(\frac{1}{n'} \sum_{s' = 1}^{n'} \phi(\mathbf{x}_{s'})\right)\\
+        =& k(\mathbf{x}_i, \mathbf{x}_j) - \frac{1}{n} \sum_{s=1}^n k(\mathbf{x}_s, \mathbf{x}_j) - \frac{1}{n'} \sum_{s'=1}^{n'} k(\mathbf{x}_i, \mathbf{x}_{s'}) + \frac{1}{n^2} \sum_{s = 1}^n \sum_{s'=1}^{n'} k(\mathbf{x}_s, \mathbf{x}_{s'})
+    \end{align}
+
+    \item Unit Norm.
+Define the ($l_2$) norm of a feature vector as $||\phi(\mathbf{x})|| = \sqrt{\sum_{m = 1}^d \phi_m(\mathbf{x})^2} = \sqrt{k(\mathbf{x}, \mathbf{x})} \in \mathbb{R}^+$, then the unit norm feature vector $\phi^{UN}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    \begin{align}
+        \phi^{UN}(\mathbf{x}_i) = \frac{\phi(\mathbf{x}_i)}{||\phi(\mathbf{x}_i)||}.
+    \end{align}
+    The corresponding unit norm kernel value between $\mathbf{x}_i$ and $\mathbf{x}_j$ is then 
+    \begin{align}
+        k^{UN}(\mathbf{x}_i, \mathbf{x}_j) &= <\frac{\phi(\mathbf{x}_i)}{||\phi(\mathbf{x}_i)||}, \frac{\phi(\mathbf{x}_j)}{||\phi(\mathbf{x}_j)||}>\\
+        &= \frac{\phi(\mathbf{x}_i)^T \phi(\mathbf{x}_j)}{||\phi(\mathbf{x}_i)|| \times ||\phi(\mathbf{x}_j)||}\\
+        &= \frac{k(\mathbf{x}_i, \mathbf{x}_j)}{\sqrt{k(\mathbf{x}_i, \mathbf{x}_i)  k(\mathbf{x}_j, \mathbf{x}_j)}}
+    \end{align}
+
+     \item Unit Variance. 
+    After the centering and unit norm normalisation, the kernel matrix is unit variance as well. 
+    In the following, we show transformations of the unit variance (with centering) normalisation.
+    Define the variance vector ${Var}(\Phi(X)) = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \bar{\Phi}(X)||^2 = \frac{1}{n} \sum_{s=1}^n ||\phi(\mathbf{x}_s) - \sum_{s'=1}^n \left(\phi(\mathbf{x}_s')\right)||^2 = \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \in \mathbb{R}$, the unit variance feature vector $\phi^{UV}(\mathbf{x}_i) \in \mathbb{R}^d$ of $\mathbf{x}_i$ is
+    \begin{align}
+        \phi^{UV}(\mathbf{x}_i) = \frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(X))}}.
+    \end{align}
+    The corresponding kernel representation is 
+    \begin{align}
+        k^{UV}(\mathbf{x}_i, \mathbf{x}_j) &= <\frac{\phi(\mathbf{x}_i)}{\sqrt{Var(\Phi(X))}}, \frac{\phi(\mathbf{x}_j)}{\sqrt{Var(\Phi(X'))}}>\\
+        &= \frac{\phi(\mathbf{x}_i)^T \mathbf{x}_j}{\sqrt{Var(\Phi(X)) Var(\Phi(\mathbf{X'}))}}\\
+        &= \frac{k(\mathbf{x}_i, \mathbf{x}_j)}{\sqrt{ \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s)  \frac{1}{n} \sum_{s'=1}^{n'}  k^C(\mathbf{x}_{s'}, \mathbf{x}_{s'})}}
+    \end{align}
+    After centering and unit norm, $ \frac{1}{n} \sum_{s=1}^n  k^C(\mathbf{x}_s, \mathbf{x}_s) = k(\mathbf{x}_i, \mathbf{x}_i)$, which implies that after centering and unit norm, the kernel matrix is already unit variance normalised.
+\end{itemize}
+For the Gaussian Process regression, we make of use of two kernel matrices: the kernel function between the training data itself, i.e. $K(X_{train}, X_{train})$; and
+the kernel function taking the training data and testing data as inputs, i.e. $K(X_{test}, X_{train})$. 
+%It is straightforward to normalise a square kernel which same input, i.e. $n = n'$ and $k(\mathbf{x}_i, \mathbf{x}_i),  k(\mathbf{x}_j, \mathbf{x}_j)$ taken from the diagonal of the matrix. 
+%The second one (between train and test) is a little bit tricky. The different is not only that $n \neq n'$. 
+We will state two ways of normalisation those two kind of matrices:
+\begin{itemize}
+    \item Normalise training and testing data separately.
+    This approach is preferred for most of the machine learning algorithms since it follows the rule that we have no information about testing data while training.
+    Then for centering, one should subtract the mean vector over the training data for both kinds of matrices.
+    For unit norm normalisation, when one calculates $K^{UN}(X_{test}, X_{train})$, the two terms inside of square root: $k(\mathbf{x}_i, \mathbf{x}_i)$ is taken from $K(X_{test}, X_{test})[i,i]$, and $k(\mathbf{x}_j, \mathbf{x}_j)$ is taken from $K(X_{train}, X_{train})[j,j]$.
+
+    \item Normalise training and testing data together, i.e. normalise $K(X_{train+test}, X_{train+test})$, then extra the parts we need from the normalised matrix. 
+    This approach is suitable in a case where one already knows the whole of testing features. 
+    For centering, one should subtract the mean vector over the whole matrix $\Phi(X_{train+test})$. 
+    The unit norm normalisation is the same as in the previous case. 
+\end{itemize}
+
+For our experiment, we fix the design space before training, i.e. the testing features are already known before testing. 
+So we choose to normalise the kernel matrix over the training and testing data together (for Round 2-3, normalise over all RBS sequence in the design space),
+by first applying centering and then unit norm normalisation.
+
+\subsection{Batch Recommendation}
+
+We consider recommending sequences in batch and using Gaussian Process Batch Upper Confidence Bound (GP-BUCB) algorithm  \cite{desautels2014parallelizing}.
+We show technical details of GP-BUCB in the following.
+With batches of size $B$, the feedback mapping $fb[t] = \lfloor(t-1) / B\rfloor B$, i.e. 
+\begin{align}
+    \mathrm{fb}[t]=\left\{\begin{array}{cl}
+    0 & : t \in\{1, \ldots, B\} \\
+    B & : t \in\{B+1, \ldots, 2 B\} \\
+    2 B & : t \in\{2 B+1, \ldots, 3 B\} \\
+    & \vdots
+    \end{array}\right.
+\end{align}
+
+A key property of Gaussian Process regression is that the predictive variance in Eq.
+(\ref{Eq: predicted variance}) only depends on observed points (i.e.
+features), but not on the labels of these observed points.
+So one can compute the posterior variance without actually observing the labels. 
+The GP-BUCB policy is to select sequences that
+\begin{align}
+    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{fb[t]}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right).
+\end{align}
+And only update $y_{t^{\prime}}=f\left(\boldsymbol{x}_{t^{\prime}}\right)+\varepsilon_{t^{\prime}} \text { for } t^{\prime} \in\{\mathrm{fb}[t]+1, \ldots, \mathrm{fb}[t+1]\}$ at the end of each batch ($\mathrm{fb}[t]<\mathrm{fb}[t+1]$).
+This is equivalent to sequential GP-UCB with \textit{hallucinated observations} $$\boldsymbol{y}_{\mathrm{fb}[t]+1: t-1}=\left[\mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{\mathrm{fb}[t]+1}\right), \ldots, \mu_{\mathrm{fb}[t]}\left(\boldsymbol{x}_{t-1}\right)\right],$$
+while the posterior variance decreases.
+
+\subsection{Bandit-0 Design}
+\label{sec: ML design pipeline}
+% \mengyan{explain somewhere why we have a different pipeline for round 0.}
+% \mengyan{add some analysis on literature data?
+% to verify 1) our library is "better" 2) it's hard to do ML on available library.
+some analysis in my mind (for literature data): number of data points, number of replicates (I remembered there only one replicates, if not, then std), label histogram, coverage of RBS.
+% And we need to address the label in literature data and ours is "different", in the way that our pipeline is different for bandit-0 design}
+
+The design of Round-0 is based on the literature data \cite{jervis2018machine}.
+We first normalise the raw TIR to values between 0 and 1. 
+We applied the Gaussian Process Regression with noise parameter $\alpha = 1e-10$. 
+We chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7bp flanking sequences both upstream and downstream of the core sequence.
+The design size is 60 with UCB parameter $\beta = 1$.
+
+\subsection{Additional Results}
+
+As stated in Section \ref{subsec: results LEARN}, we merely aimed to provide a valid ranking in Learn part due to the small amount of noisy data, which can be captured by the Spearman correlation coefficient.
+We show $R^2$ score for the reader's interest, which quantifies value-based prediction evaluation. The $R^2$ increases along with rounds increase.
+
+\begin{table}[h!]
+\caption{$R^2$ score for predictions in Figure \ref{fig: Scatterplot}.}
+\centering
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+                  & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0\\ Test 1\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0-1\\ Test 2\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 0-2\\ Test 3\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Train 80\%\\ Test 20\%\end{tabular}} \\ \hline
+\textbf{$R^2$}       & 0.067                                                             & 0.226                                                               & 0.213                                                               & 0.269                                                                   \\ \hline
+\textbf{Spearman} & 0.269                                                             & 0.468                                                               & 0.546                                                               & 0.479                                                                   \\ \hline
+\end{tabular}
+\end{table}
+
+\subsection{Exploration-exploitation visualisation}
+\label{subsec: Exploration-exploitation visualisation}
+
+In this section, we show our method of exploration-exploitation visualisation displayed in Figure \ref{fig: Swarmplot and Quantplot}C.
+The UCB score of a sequence $\mathbf{x}_i$ at Round $t$ is defined as 
+\begin{align*}
+% \label{Eq: UCB}
+    \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i),
+\end{align*}
+where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
+$\mu_{t-1}(\mathbf{x}_i), \sigma_{t-1}(\mathbf{x}_i)$ are the predicted TIR and standard deviation (STD) at Round $t-1$ for the sequence $\mathbf{x}_i$.
+The sequences with high predicted TIR address exploitation, while the sequences with high predicted STD address exploration. 
+We show the predicted TIR (x-axis) v.s. predicted STD (y-axis) in Figure \ref{fig:exploitationvsexplorationdetail} as orange points.
+
+To compare the predicted TIR and STD in the same dimension (basis), we project the (predicted TIR, predicted STD) pair to the diagonal line (the grey dashline) set between (0,1) and (1,0) coordinates.
+The slope of the diagonal line is selected to weight the predicted TIR and STD equally. 
+Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
+As a result, the higher the predicted TIR the closer the point is to the (1,0) coordinate, and the higher the predicted STD the closer the point is to the (0,1) coordinate.
+Finally, we colour those projected points using the RdBu spectrum, where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
+Finally, these colours correspond to ones shown in Figure \mbox{\ref{fig: Swarmplot and Quantplot}C}.
+
+% \mengyan{mengyan to re-write it}
+% For a given round all the tested sequences are then projected along the diagonal line crossing (0,1) and (1,0) according to their predicted STD and mean TIR (the higher the predicted mean the closer the point is to the (1,0) coordinate and the higher the predicted STD the closer the point is to the (0,1) coordinate).
+% Each point on the diagonal then, corresponds to an orthogonal projection (known as \emph{vector projection}) of an orange point.
+% We do this to transform our predictions into a common basis (where the basis is the diagonal) and so by calculating where a given point falls on the diagonal we can show if its mean TIR or STD is higher.
+% Finally, we colour those projected points using the RdBu spectrum where the points with high STD (exploration) are shown in red and those with high mean TIR (exploitation) are shown in blue.
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_0.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_1.pdf}
+    \includegraphics[scale=0.35]{plots/Supplementary/proj_scatter_abc1_FF_2.pdf}
+   \caption{\textbf{Exploration-Exploitation scores illustration.} We show the predictions for each sequence from Round 1-3 in orange (with the same settings as Figure \mbox{\ref{fig: Scatterplot}}) and their corresponding projections to (0,1), (1,0) diagonal coloured in RdBu spectrum.
+    We explain in details how the colours are generated in Section \ref{subsec: Exploration-exploitation visualisation} .
+   }
+   \label{fig:exploitationvsexplorationdetail}
+\end{figure}
+
+\subsection{Model evaluation on RBS calculator data}
+\label{subsec: Model evaluation on RBS calculator data}
+
+\hl{To test the reliability of our proposed model, we have generated TIR predictions for all 4096 RBS sequences using the RBS  calculator ver.
+2.0 \mbox{\cite{Salis2009}} and evaluated our model using that data.
+We normalised the TIR values by dividing them by the TIR of the benchmark sequence, and show the TIR ratio mapping between the RBS calculator and TIRs obtained experimentally for this work in Figure \mbox{\ref{fig:TIR ratio salis micro}}.
+We observe that the RBS calculator values are not well aligned to the measured values, which underlines the need for development of new prediction and design methods. 
+Among all 4096 RBS sequences, 38 RBS were predicted to have higher TIR than the benchmark sequence by the RBS calculator.
+We call those sequences the \textit{optimal set}, and the goal is to find as many sequences from the optimal set as permitted by the budget (450).
+
+To do this, we have applied our approach to this data set - first we have chosen 90 sequences at random and after teaching the algorithm using the obtained numbers we have generated 90 further sequences to be tested (using UCB for recommendations) and did this for a total of 4 rounds.
+We have repeated this experiment 100 times to increase our confidence in the results, by doing so we have found that our model can identify above 75\% of the optimal set in up to 2 rounds and 99\% of them in up to 3 rounds.
+We reported the detailed statistics for the 100 runs in Table \mbox{\ref{table: Important statistics for the 100 synthetic experiments}}.
+Figure \mbox{\ref{fig: Swarmplot and Quantplot Salis}} shows plot analogues to these in Figure \mbox{\ref{fig: Swarmplot and Quantplot}}  but for one of the runs using the RBS calculator data.
+Our algorithm has successfully found most of the sequences in optimal set in the first and second rounds, thus the recommended TIRs have an apparent drop in the following rounds (as there are no good sequences left to recommend).
+This might suggest that it might be valuable to add an additional stopping condition for the DBTL cycle - decreasing round performance.}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale = 0.6]{paper/plots/Supplementary/TIR_ratio_salis_micro.png}
+    \caption{TIR ratio of RBS calculator ver. 2.0 \cite{Salis2009} plotted against TIR ratios obtained experimentally in this work. The red point represents the benchmark sequence.}
+    \label{fig:TIR ratio salis micro}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.35]{plots/Supplementary/swarmplot_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.25\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.35]{plots/Supplementary/quantplot_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.25\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.35]{plots/Supplementary/swarmplot_proj_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/histogram_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.48\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.42]{plots/Supplementary/tsneplot_salis.pdf}
+    \end{subfigure}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
+    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
+    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
+    \caption{
+    \textbf{Model evaluation on RBS calculator data.}
+    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
+    Random: randomly chosen sequences.
+    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
+    Dashed line is set to TIR of the benchmark sequence.
+    \textbf{B)} Line plot showing TIR obtained in a given quantile (Q) of results divided into groups as in A)
+    \textbf{C)} Exploitation v.s. Exploration for Bandit 0-3. Blue-hued points represent exploitation, those hued red represent exploration.
+    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
+    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
+    The area of the circle corresponds to the TIR value.
+    The TIR results in all subplots are shown normalised to the  benchmark sequence.}
+    \label{fig: Swarmplot and Quantplot Salis}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_0_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_1_salis.pdf}
+    \end{subfigure}
+    % \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_2_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/scatter_abc1_FF_3_salis.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/salis_scatter_abc1_FF_4.pdf}
+    \end{subfigure}
+    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation) on the RBS calculator data}.
+The scatter plots A-D show the performance of our prediction algorithm calculated after each round using the RBS calculator data.
+    E shows the performance on all collected data.
+    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
+    X-axis and y-axis are the TIR predicted by the RBS calculator and the TIR predicted by our model respectively.
+    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided in each plot's title on test data only.
+    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
+    }
+    \label{fig: Scatterplot_Salis}
+\end{figure}
+
+\begin{table}[h!]
+\caption{\textbf{Important statistics for the 100 synthetic experiments run using the RBS calculator data.}
+Each round contains 90 samples, where Round 0 is uniformly random sampled from the whole space, and Rounds 1-4 follow the bandits algorithm. 
+In column \textit{Percent of the Optimal Set Found} and \textit{Standard Deviation}, we reported the average percent of the \textit{optimal set} found in each round and the standard deviation of this value averaged over the 100 experiments.
+In the last column, we reported the average percent of the (\textit{optimal set}) found in total after each round. 
+We can observe that our algorithm can find the majority of the (\textit{optimal set}) sequences within three rounds.
+}
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+                  \textbf{Round} & \textbf{\begin{tabular}[c]{@{}c@{}}Percent of\\ the Optimal Set Found\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Standard\\ Deviation\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Total Percent of\\ the Optimal Set Found\end{tabular}} \\ \hline
+ \textbf{0}  & 0.022  & 0.026 & 0.022  \\ \hline
+  \textbf{1}  & 0.744  & 0.085 & 0.766  \\ \hline
+   \textbf{2}  & 0.222  & 0.077 & 0.988  \\ \hline
+    \textbf{3}  & 0.012  & 0.026 & 1.000  \\ \hline
+     \textbf{4}  & 0.000  & 0.000 & 1.000  \\ \hline
+
+\end{tabular}
+\label{table: Important statistics for the 100 synthetic experiments}
+\end{table}
+
+\newpage
+
+\section{Supplementary Plots}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
+    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} 
+    The core region is highlighted in light green and the lines are rolling averages (in this case mean of the 5 points centred over the indicated position) for each base.
+    The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base, which is colour coded.
+    Since only changes from the original base at each position are shown, the dotted lines start and end at different positions depending on how many changes to the respective base are at positions adjacent to the indicated one.
+    The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value $<$ 0.0001 and 34 degrees of Freedom.}
+    \label{fig:core_vs_noncore}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale = 0.4]{plots/Supplementary/swarmplot_raw.pdf}
+    \caption{
+    \textbf{Raw TIRs of RBS groups examined in this study.}
+    Raw TIR is calculated as a derivative of GFP fluoresence divided by OD600 of culture over 4h counting from the start of log phase of growth. 
+    BPS-NC: base-by-base changes in the non-core region. 
+    BPS-C: base-by-base changes in the core region. 
+    UNI: Randomly generated sequences with uniform distribution. 
+    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
+    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
+    SD - Shine-Dalgarno sequence.
+    BN - benchmark sequences for all plates. 
+    }
+    \label{fig:rawswarmplots.}
+\end{figure}
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.5]{plots/Supplementary/tsneplot_hamming.pdf}
+   \caption{\textbf{t-SNE plot based on Hamming distances.} This t-SNE plot has been drawn based on Hamming distances between sequences. 
+    }
+   \label{fig: TSNE hamming}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_bc1_TTF_0.pdf}
+    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_1.pdf}
+    \includegraphics[scale = 0.35]{plots/Supplementary/scatter_abc1_TT_2.pdf}
+    % \includegraphics[scale = 0.4]{plots/Supplementary/scatter_abc1_TT_3.pdf}
+    \caption{\textbf{Predictions in design pipeline in Round 1-3.} For Round 1, the kernel matrix is normalised based on known RBS sequences in Round 0.
+    For Round 2-3, the kernel matrix is normalised based on all RBS sequences in the design space.
+    Compared to the results shown in Figure \mbox{\ref{fig: Scatterplot}}, here we have applied kernel normalisation described in Section \mbox{\ref{supp: Normalisation of Kernel}}.
+Although the precise value prediction looks different from Figure \mbox{\ref{fig: Scatterplot}}, the rankings provided in two figures are similar, which can be reflected by the similar Spearman correlation coefficients.
+    }
+    \label{fig:scatter abc1 TT.}
+\end{figure}
+
+% \begin{figure}[!ht]
+%     \centering
+%     \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
+%     \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
+%     \caption{Kernel Heatmap and Predictions.
+Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265).
+Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right).
+The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
+% \end{figure}
+
+\begin{figure}[!ht]
+   \centering
+   \includegraphics[width=0.5\textwidth]{plots/Supplementary/All_logo.pdf}
+   \caption{\textbf{Sequence logo for all tested sequences.} Compared to Figure \ref{fig:Library characteristics}A there are no significant biases at each position.}
+   \label{fig: All_logo}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGG.pdf}
+        \label{fig:GGGGGG}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[scale=0.25]{plots/Supplementary/Structure_GGGGGC.pdf}
+        \label{fig:GGGGGC}
+    \end{subfigure}
+    \vskip\baselineskip
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[scale=0.25]{plots/Supplementary/Structure_CGGGAC.pdf}
+        \label{fig:CGGGAC}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[scale=0.25]{plots/Supplementary/Structure_CACGTT.pdf}
+        \label{fig:CACGTT}
+    \end{subfigure}
+       \caption{\textbf{Folding predictionds for four different RBS.} Here we show the predicted, energetically most favourable structures for four of our RBS with the following cores: GGGGGG, GGGGGC, CGGGAC, CACGTT and their immediate upstream and downstream background sequence.
+There is no discerbile and consistent difference between strong ones (the first two) and weak ones (the last two).}
+       \label{fig:structures}
+\end{figure}
+
+\begin{figure}[!ht]
+   \centering
+    \includegraphics[scale=0.4]{plots/Supplementary/SDhist.pdf}
+   \caption{\textbf{Histogram of standard deviations.}
+   For each RBS sequence, we tested 6 biological replicates. This plot shows the histogram of the STD of raw TIR values of the 6 replicates for each RBS sequence. }
+   \label{fig: SDhist}
+\end{figure}
+
+\begin{figure}[!ht]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.4]{plots/Supplementary/SD.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{}
+        \includegraphics[scale=0.6]{plots/Supplementary/SDvsTIR.pdf}
+    \end{subfigure}
+    \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation).
+B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR.
+Standard deviation and average TIR is given in terms of raw number.}
+    \label{fig:variation of biological replicates}
+\end{figure}
+
 \end{document}

### 2022-04-20 paper/main.tex
Final paper with highlights.

@@
 trade off by using predictive uncertainty. The machine learning predictors
-mentioned, as well as existing RBS calculators,do not yet provide uncertainty levels to guide exploration with the notable exception of ART \cite{Radivojevic2020} .
+mentioned, as well as existing RBS calculators,do not yet provide uncertainty levels to guide exploration \hl{with the notable exception of ART} \cite{Radivojevic2020} .
 Furthermore large data sets may not be available for a genetic part of interest, for instance the RBS.
@@
 All the processed and raw data are included in the repository.
-Sequences of plasmids and oligos and assembly reports used in this study are available in the supplementary information as a separate file.
 The pBbB6c plasmid is available on Addgene (Watertown, MA). Other strains and plasmids are available on request from the authors; MTAs will need to be negotiated between the parties.
@@
 A detailed text description of the machine learning methods, including Gaussian Process regression, choices of kernels, batch recommendation, experimental settings, additional results and model evaluation on RBS calculator data; Figures S1-S12; Tables S1-S2.
+Sequences of plasmids and oligos and assembly reports used in this study.
 

### 2022-04-20 paper/main.tex
Final paper no highlights.

--- 2022-04-20 paper/main.tex
@@
 Only a few of the random sequences were promising, but they were still 20\% weaker than the benchmark sequence's TIR (Figure 3A).
-In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (\hl{Note, this applies only to the 7 bps on either side of the core; }see Supplementary Figure S1 and work by \mbox{\textcite{Jeschek2016}}).
+In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (Note, this applies only to the 7 bps on either side of the core; see Supplementary Figure S1 and work by \mbox{\textcite{Jeschek2016}}).
 Sequences from randomization of the core RBS (BPS-C, UNI, PPM) are more appropriate controls, from which the best sequence achieved only about 65\% of the benchmark TIR.
@@
 
-\hl{We provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient}
+We provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient
 \cite{Schober2018,  Kang2019}.
@@
 However, since the predictions are only used for recommendations in the Design phase, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
-\hl{
 This is because, in the Design part, we select RBSs based on the ranks of the TIR predictions instead of numerical predictions.
 % in this task, we are maximising the TIR of RBS and so we are looking for RBS with high rank 1, regardless of its numerical TIR.
-}
 This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
@@
 This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
-This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with \hl{higher probability of having a high TIR value}.\\
+This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with higher probability of having a high TIR value.\\
 
@@
 We discovered an interesting phenomenon of large edit distances between RBSs with high TIRs.
-In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR \hl{(we assume here that our data set includes most of the strongest RBSs for each distance - sequence combination)}.
+In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR (we assume here that our data set includes most of the strongest RBSs for each distance - sequence combination).
 We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
@@
 As long as this does not impact targeted methods like machine learning-guided design, this implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
-In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long \hl{(Figure} \ref{fig: Swarmplot and Quantplot}A \hl{shows comparison between efficiency of random versus directed sequence generation)}.
+In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long (Figure \ref{fig: Swarmplot and Quantplot}A shows comparison between efficiency of random versus directed sequence generation).
 In such cases, a directed process, like the one described here, should be strongly encouraged.
@@
 In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
-In the Learn phase, we used Gaussian process regression to predict the TIR mean and \hl{uncertainty of that prediction}.
+In the Learn phase, we used Gaussian process regression to predict the TIR mean and uncertainty of that prediction.
 To represent RBSs and capture the similarities between them, we choose to use
@@
 
-\hl{To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator} \cite{Salis2009} \hl{and used our approach to simulate the search for the strongest designs in limited number of rounds.
-We show these results in Supplementary Figure} \ref{fig: Swarmplot and Quantplot Salis}, \hl{but in short, our approach was able to  identify above 75\% of the sequences stronger than the benchmark in up to 2 rounds and 99\% of them in up to 3 rounds.
+To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator \cite{Salis2009} and used our approach to simulate the search for the strongest designs in limited number of rounds.
+We show these results in Supplementary Figure \ref{fig: Swarmplot and Quantplot Salis}, but in short, our approach was able to  identify above 75\% of the sequences stronger than the benchmark in up to 2 rounds and 99\% of them in up to 3 rounds.
 This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
-Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.}\\
+Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.\\
 
@@
 For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
-Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available \hl{(for example where there is no fluorescent readout available)}, while still providing the required results (optimised part).
+Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available (for example where there is no fluorescent readout available), while still providing the required results (optimised part).
 \\
@@
 We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
-Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty \hl{quantification}?
+Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty quantification?
 Thirdly, the generalisability of the method is unknown.
@@
 Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open question.
-\hl{In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance} \mbox{\cite{Radivojevic2020}}.
+In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance \mbox{\cite{Radivojevic2020}}.
 %While Radivojevic \textcite{Radivojevic2020} provides a Bayesian ensemble framework, we focus on Gaussian process and provides methods to design batch recommendation better.}
@@
 In future we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
-However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments \hl{and so the percentage of the space that is measured will inevitably decrease, which might impact the expected recommendation results}.
+However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments and so the percentage of the space that is measured will inevitably decrease, which might impact the expected recommendation results.
 To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
@@
 %As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
-\hl{GPR can predict the posterior mean of the RBS sequences in the design space.
-At the same time, it can provide the posterior standard deviation, which represents the level of uncertainty for the prediction.}
+GPR can predict the posterior mean of the RBS sequences in the design space.
+At the same time, it can provide the posterior standard deviation, which represents the level of uncertainty for the prediction.
 
@@
 For Bandit designs in Round 0, since we only had access to a limited number of data points from the literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6 bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7 bp flanking sequences both upstream and downstream of the core sequence.
-\hl{
 Among various string kernels which have been studied by \mbox{\textcite{Ben-Hur2008}} and we will introduce briefly in Supplementary \mbox{\ref{sec: Choices of Kernels}}, we chose the one with the best reported performance, the \textit{weighted degree kernel with shift} (WDS) \mbox{\cite{ratsch_rase_2005_wds, Ben-Hur2008}},
-to represent the RBS sequences in subsequent rounds and specify the kernel function of $GP$.  }
+to represent the RBS sequences in subsequent rounds and specify the kernel function of $GP$. 
 WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
@@
 % In our design, the length of RBS sequence $L = 20$.
-\hl{
+
 The hyperparameters for kernel, including maximum substring length $\ell = 6$, maximum shift length $S(l) = 1$, and the standard deviation of GP noise $\alpha = 1$ were chosen based on 10-repeat 5-fold cross validation.
-The values of hyperparameters considered were $\ell \in \{3,4,5,6\}$, $S(l) = \{0,1,2\}$, $\alpha = [0.5, 1, 2]$. }
+The values of hyperparameters considered were $\ell \in \{3,4,5,6\}$, $S(l) = \{0,1,2\}$, $\alpha = [0.5, 1, 2]$.
 % ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).
@@
 where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
-$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation \hl{(that is uncertainty of the prediction)} at Round $t$ for the sequence $\mathbf{x}_i$.
+$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation (that is uncertainty of the prediction) at Round $t$ for the sequence $\mathbf{x}_i$.
 We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at Round $t-1$.\\
@@
 
-\hl{To test the reliability of our proposed model, we have generated TIR predictions for all 4096 RBS sequences using the RBS  calculator ver.
+To test the reliability of our proposed model, we have generated TIR predictions for all 4096 RBS sequences using the RBS  calculator ver.
 2.0 \mbox{\cite{Salis2009}} and evaluated our model using that data.
@@
 Our algorithm has successfully found most of the sequences in optimal set in the first and second rounds, thus the recommended TIRs have an apparent drop in the following rounds (as there are no good sequences left to recommend).
-This might suggest that it might be valuable to add an additional stopping condition for the DBTL cycle - decreasing round performance.}
+This might suggest that it might be valuable to add an additional stopping condition for the DBTL cycle - decreasing round performance.
 

### 2022-05-25 paper/main.tex
update tsne plot

@@
     \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
+    The Unlabeled represents the RBS sequences in design space but have not been tested.
     The area of the circle corresponds to the experimentally-obtained TIR value.
@@
 We show these results in Supplementary Figure \ref{fig: Swarmplot and Quantplot Salis}, but in short, our approach was able to  identify above 75\% of the sequences stronger than the benchmark in up to 2 rounds and 99\% of them in up to 3 rounds.
-This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
+This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmically are easier to predict by our algorithm than real world data.
 Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.\\
@@
 
-For each RBS sequence, we measured the TIR of 6 biological replicates, where TIR is calculated as an averaged sum of GFP fluorescence divided by Optical Density at 600 nm (OD600) of the culture over time (calculated using values from all the measured data points), following the formula:
-
-$$\mathnormal{TIR=\frac{\sum_{t_{min}}^{t_{max}}\frac{GFP(t)}{OD_{600}(t)}}{\Delta t}}$$
-
-where ${t_{min}}$ is the time at which the $\frac{GFP}{OD_{600}}$ value is at its minimum denoting the start of the log growth phase, ${t_{max}}$ is the time at which the logarithmic phase ends (on average, for our samples that was 4 hours from ${t_{min}}$ so that was the time point used for TIR calculations for all samples), ${\Delta t}$ is the difference between ${t_{max}}$ and ${t_{min}}$.\\
-
-% REVIEW:
-% \mengyan{Table XXX shows an example of our raw data. }
-% We consider both the label (TIR) pre-processing and kernel pre-processing.
+For each RBS sequence, we measured the TIR of 6 biological replicates.
+We measure GFP fluorescence of the culture in the log phase as it is the most representative one.
+% \textit{Log growth phase} is a phase in the growth of bacterial culture, the culture in general starts slow which is called \textit{lag phase}, then it grows exponentially which is called log phase after which it plateaus.
+We also measure the Optical Density at 600 nm (OD600) of the culture every 10 minutes, which reflects how dense (how many cells) the culture is.
+Denote our measuring time as $t \in \{t_{min}, \ldots, t_{max}\}$, where ${t_{min}}$ start and end of the log growth phase.
+At $t_{min}$, the $\frac{GFP}{OD_{600}}$ value is at its minimum.
+We choose the time range to be 4 hours, i.e. $t_{max} - t_{min} = 4h$.
+In our experiment, TIR is calculated as an averaged sum of GFP fluorescence divided by OD600 of the culture over time (calculated using values from all the measured data points),
+$$TIR=\frac{1}{t_{max} - t_{min}} \sum_{t = t_{min}}^{t_{max}}\frac{GFP(t)}{OD_{600}(t)}.$$
+
+% where TIR is calculated as an averaged sum of GFP fluorescence divided by Optical Density at 600 nm (OD600) of the culture over time (calculated using values from all the measured data points), following the formula:
+
+% $$\mathnormal{TIR=\frac{\sum_{t_{min}}^{t_{max}}\frac{GFP(t)}{OD_{600}(t)}}{\Delta t}}$$
+
+% where ${t_{min}}$ is the time at which the $\frac{GFP}{OD_{600}}$ value is at its minimum denoting the start of the log growth phase, ${t_{max}}$ is the time at which the logarithmic phase ends (on average, for our samples that was 4 hours from ${t_{min}}$ so that was the time point used for TIR calculations for all samples), ${\Delta t}$ is the difference between ${t_{max}}$ and ${t_{min}}$.\\
+
 For label pre-processing, we first adjust the TIR values in each round using the round-wise reference values.
-
 The reference value is the TIR of the benchmark sequence that is run in triplicate in each round.
@@
 
-    \item Normalise training and testing data together, i.e. normalise $K(X_{train+test}, X_{train+test})$, then extra the parts we need from the normalised matrix. 
+    \item Normalise training and testing data together, i.e. normalise $K(X_{train+test}, X_{train+test})$, then extract the parts we need from the normalised matrix. 
     This approach is suitable in a case where one already knows the whole of testing features. 
@@
     \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
+    The Unlabeled represents the RBS sequences in design space but have not been tested.
     The area of the circle corresponds to the TIR value.
@@
     \end{subfigure}
-       \caption{\textbf{Folding predictionds for four different RBS.} Here we show the predicted, energetically most favourable structures for four of our RBS with the following cores: GGGGGG, GGGGGC, CGGGAC, CACGTT and their immediate upstream and downstream background sequence.
+       \caption{\textbf{Folding predictions for four different RBS.} Here we show the predicted, energetically most favourable structures for four of our RBS with the following cores: GGGGGG, GGGGGC, CGGGAC, CACGTT and their immediate upstream and downstream background sequence.
 There is no discerbile and consistent difference between strong ones (the first two) and weak ones (the last two).}

### 2022-05-25 synbio_rbs/src/config.py
update tsne plot

@@ -3,6 +3,7 @@
 BASES = ['A','C','G','T'] # embedding
 
 SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
+# SAVED_IDX_SEQ_PATH = '../data/idx_seq_lit.pickle'
 
 # Usage:
 # with open(SAVED_IDX_SEQ_PATH, 'rb') as handle:
@@ -12,6 +13,7 @@ SAVED_IDX_SEQ_PATH = '../data/idx_seq.pickle'
 # idx_seq_dict = saved_idx_seq['idx_seq_dict']
 
 SAVED_KERNEL_PATH = '../data/saved_kernel/'
+# SAVED_KERNEL_PATH = '/data4/u6015325/Solaris/saved_kernel_lit/'
 
 # ---------------------------------------------------------------------
 

### 2022-05-25 synbio_rbs/src/kernels_for_GPK.py

@@ -16,6 +16,7 @@ from src.environment import Rewards_env
 
 
 import pickle
+# import pickle5 as pickle
 import src.config as config # provide configuration, e.g. global variable, path
 
 # June 2021 Mengyan Zhang 
@@ -116,7 +117,9 @@ class String_Kernel(Kernel):
             # SAVED_KERNEL_PATH = '../data/saved_kernel/' is specified in config.py
 
             with open(config.SAVED_IDX_SEQ_PATH, 'rb') as handle:
-                seq = pickle.load(handle)['seq_list']
+                seq = pickle.load(handle)
+#                 seq.to_pickle(config.SAVED_IDX_SEQ_PATH)
+                seq = seq['seq_list']
 
             if os.path.exists(kernel_path):
                 print('Load saved kernel matrix...')

### 2022-05-26 paper/main.tex
Updated captions in paper.

@@
     \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
-    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
-    The Unlabeled represents the RBS sequences in design space but have not been tested.
-    The area of the circle corresponds to the experimentally-obtained TIR value.
+    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design space as calculated by our kernel function (weighted degree kernel with shift).
+    The Unlabeled points represent the RBS sequences in the design space that have not been tested.
+    The area of the marker corresponds to the experimentally-obtained TIR value.
     The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard; the TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate.
@@
     \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
-    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift).
-    The Unlabeled represents the RBS sequences in design space but have not been tested.
-    The area of the circle corresponds to the TIR value.
+    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design space as calculated by our kernel function (weighted degree kernel with shift).
+    The Unlabeled points represent the RBS sequences in the design space that have not been tested.
+    The area of the marker corresponds to the TIR value.
     The TIR results in all subplots are shown normalised to the  benchmark sequence.}
